### 2014-11-10 docs/papers/shallow_performance/shallow_perf.tex
Import of skeleton for paper on serial performance

%% shallow_perf.tex
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{./IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/tex-archive/info/epslatex/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
% Axel Sommerfeldt. This package may be useful when used in conjunction with 
% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
% submissions have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.4.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/url/
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The abstract goes here.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Performance, Code-generation, Finite-difference
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
for IEEE journal papers produced under \LaTeX\ using
IEEEtran.cls version 1.8 and later.

\subsection{The 'shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourney.  Originally written
by Paul Schwartrauber of NCAR.  Organic development - optimised for
previous hardware.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

\subsubsection{Subsubsection Heading Here}
Subsubsection text here.

\section{Results}

We have results for the Gnu, Intel, Cray and IBM compilers.
We've run on Intel Ivy Bridge, Intel Haswell, and IBM BG/Q.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

\section{Conclusion}
The conclusion goes here.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank...

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}

%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

%\end{thebibliography}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiographynophoto}{A.~R.~Porter}
Biography text here.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{R.~Ford}
Biography text here.
\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

\begin{IEEEbiographynophoto}{G.~Riley}
Biography text here.
\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2014-11-10 shallow_perf.tex
Starting to draft the steps taken in the serial work

--- 2014-11-10 docs/papers/shallow_performance/shallow_perf.tex
@@
 
-\subsubsection{Subsubsection Heading Here}
-Subsubsection text here.
+\subsection{The {PS}y{KA}l Approach}
+Proliferation of lightweight cores is resulting in increased
+complexity in the programming models required to achieve performance.
+Combine with the complexity of e.g. a model simulating circulation in
+the global ocean. Significant challenge in software engineering.
+Propose an approach for meeting this challenge by separating the
+computational science demands from the e.g. oceanographic demands.
+
+\begin{figure}
+\centering
+\caption{The {PS}y{KA}l Approach.}
+\end{figure}
+
+While encapsulation makes it easier to produce more robust software,
+it can often come with a performance penalty. In this work we
+investigate the origin of this penalty and the steps required to
+remove it.
 
@@
 We've run on Intel Ivy Bridge, Intel Haswell, and IBM BG/Q.
+
+\begin{itemize}
+
+\item Performance of the 'original' version of shallow; i.e. the version we
+were supplied with at the start of this work.
+
+\item Performance of the vanilla, unoptimised version after it has been
+re-structured following the {PS}y{KA}l approach.
+
+\end{itemize}
 

### 2014-11-10 SEQ/original/shallow_base.f90
Remove use of netcdf from original

@@ -36,8 +36,6 @@
       USE timing_mod
       IMPLICIT NONE
 
-      INCLUDE 'netcdf.inc'
-
       INTEGER :: m, n    ! global domain size
       INTEGER :: itmax   ! number of timesteps
       INTEGER :: mprint  ! frequency of output    
@@ -427,7 +425,8 @@
 
       IF (l_out) THEN 
 
-         iret = nf_close(ncid)
+         iret = 0
+         !iret = nf_close(ncid)
          call check_err(iret)
       ENDIF
 
@@ -458,9 +457,10 @@
 
       subroutine check_err(iret)
       integer iret
-      include 'netcdf.inc'
-      if(iret .ne. NF_NOERR) then
-         print *, nf_strerror(iret)
+!      include 'netcdf.inc'
+      if(iret .ne. 0) then
+      !if(iret .ne. NF_NOERR) then
+       !  print *, nf_strerror(iret)
          stop
       endif
       end subroutine
@@ -472,7 +472,7 @@
       character(len=*) file
       integer m,n
 !     declarations for netCDF library
-      include 'netcdf.inc'
+!      include 'netcdf.inc'
 !     error status return
       integer iret
 !     netCDF id
@@ -499,37 +499,38 @@
       integer istart(p_rank)
       integer icount(p_rank)
       
+      iret = 0
 !     enter define mode
-      iret = nf_create(file, NF_CLOBBER,ncid)
+      !iret = nf_create(file, NF_CLOBBER,ncid)
       call check_err(iret)
 !     define dimensions
-      iret = nf_def_dim(ncid, 'm', m, m_dim)
+      !iret = nf_def_dim(ncid, 'm', m, m_dim)
       call check_err(iret)
-      iret = nf_def_dim(ncid, 'n', n, n_dim)
+      !iret = nf_def_dim(ncid, 'n', n, n_dim)
       call check_err(iret)
 !     time is an unlimited dimension so that any number of
 !     records can be added
-      iret = nf_def_dim(ncid, 'time', NF_UNLIMITED, time_dim)
+      !iret = nf_def_dim(ncid, 'time', NF_UNLIMITED, time_dim)
       call check_err(iret)
 !     define coordinate variable for time      
       t_dims(1) = time_dim
-      iret = nf_def_var(ncid, 'time', NF_INT, 1, t_dims, t_id)
[…]

### 2014-11-11 docs/papers/shallow_performance/shallow_perf.tex
Add list of transformations that have been applied to the code

@@
 
-\subsubsection{Subsubsection Heading Here}
-Subsubsection text here.
+\subsection{The {PS}y{KA}l Approach}
+Proliferation of lightweight cores is resulting in increased
+complexity in the programming models required to achieve performance.
+Combine with the complexity of e.g. a model simulating circulation in
+the global ocean. Significant challenge in software engineering.
+Propose an approach for meeting this challenge by separating the
+computational science demands from the e.g. oceanographic demands.
+
+\begin{figure}
+\centering
+\caption{The {PS}y{KA}l Approach.}
+\end{figure}
+
+While encapsulation makes it easier to produce more robust software,
+it can often come with a performance penalty. In this work we
+investigate in detail the hardware and compiler dependencies of this
+penalty and the steps required to remove it.
 
@@
 We have results for the Gnu, Intel, Cray and IBM compilers.
-We've run on Intel Ivy Bridge, Intel Haswell, and IBM BG/Q.
+We've run on Intel Ivy Bridge, Intel Haswell, IBM BG/Q, IBM P8.
+
+\begin{itemize}
+
+\item Performance of the 'original' version of shallow; i.e. the version we
+were supplied with at the start of this work.
+
+\item Performance of the vanilla, unoptimised version after it has been
+re-structured following the {PS}y{KA}l approach.
+
+\end{itemize}
+
+Beginning with the vanilla {PS}y{KA}l version, we apply a series of
+optimisations with the aim of recovering, as much as is possible, the
+structure and thus performance of the original version of the code.
+
+\begin{enumerate}
+
+\item Pass explicit array bounds to middle layer (rather than using
+  assumed-size array arguments)
+
+\item Move kernel routines into same module as middle layer
+
+\item Loop fusion
+
+\item In-lined field-copy operation
+
+\item Fused field-copy operation
+
+\end{enumerate}
+
+These optimisations do not always result in improved performance (AND
+HENCE WE SHOULD CALL THEM TRANSFORMATIONS PERHAPS?). Whether or not
+they do so depends both on the compiler and the problem size.
 

### 2014-11-11 docs/papers/shallow_performance/shallow_perf.tex
Add makefile, slowdown figure and bibliography

@@
 
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
+\usepackage{cite}
 
@@
 \ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
+  \usepackage[pdftex]{graphicx}
   % declare the path(s) where your graphic files are
@@
   % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
 \else
@@
   % driver is specified.
-  % \usepackage[dvips]{graphicx}
+  \usepackage[dvips]{graphicx}
   % declare the path(s) where your graphic files are
@@
   % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
+  \DeclareGraphicsExtensions{.eps}
 \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/tex-archive/info/epslatex/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
 
@@
 %
-%\usepackage{url}
+\usepackage{url}
 % url.sty was written by Donald Arseneau. It provides better support for
@@
 \hyphenation{op-tical net-works semi-conduc-tor}
+
+\newcommand{\psykal}{{PS}y{KA}l}
 
@@
 
-\subsection{The 'shallow' Program}
+\subsection{The `shallow' Program}
 For this work we have used a benchmarking program that solves the
 shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourney.  Originally written
-by Paul Schwartrauber of NCAR.  Organic development - optimised for
-previous hardware.
+finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
+Originally written by Paul Schwartrauber of NCAR.  Organic development
+- optimised for previous hardware.
 
@@
 
-\subsection{The {PS}y{KA}l Approach}
+\subsection{The \psykal Approach}
 Proliferation of lightweight cores is resulting in increased
@@
 \centering
-\caption{The {PS}y{KA}l Approach.}
+\caption{The \psykal Approach.}
 \end{figure}
@@
 \item Performance of the vanilla, unoptimised version after it has been
-re-structured following the {PS}y{KA}l approach.
+re-structured following the \psykal approach.
 
@@
 
-Beginning with the vanilla {PS}y{KA}l version, we apply a series of
+Beginning with the vanilla \psykal version, we apply a series of
 optimisations with the aim of recovering, as much as is possible, the
@@
 
+In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
+original version of the shallow code for the range of compilers and
+hardware considered here.
+
 % An example of a floating figure using the graphicx package.
@@
 %
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.8in]{orig_summary}
 % where an .eps filename suffix will be assumed under latex, 
@@
 % via \DeclareGraphicsExtensions.
-%\caption{Simulation Results.}
-%\label{fig_sim}
-%\end{figure}
+\caption{Summary of the performance of the original shallow code on 
+the range of compilers and hardware under consideration.}
+\label{FIG_orig_perf_summary}
+\end{figure}
 
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.8in]{best_psykal_summary}
+\caption{Summary of the best performance achieved by any \psykal 
+version of shallow with each of the compilers and CPUs under 
+consideration.}
+\label{FIG_best_psykal_perf_summary}
+\end{figure}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.8in]{slowdown_summary}
+\caption{Summary of the performance achieved by any \psykal
+version as a percentage of that achieved by the original version of the code.}
+\label{FIG_slowdown_summary}
+\end{figure}
 
@@
 
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
 % http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
@@
 % http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
+\bibliographystyle{IEEEtran}
 % argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
+\bibliography{IEEEabrv,shallow_perf}
 %
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
 
@@
 Biography text here.
-\end{IEEEbiography}
+\end{IEEEbiographynophoto}
 

### 2014-11-11 shallow_perf.tex
Add tables on hardware and compilers

@@
 
-\subsection{The 'shallow' Program}
+\subsection{The `shallow' Program}
 For this work we have used a benchmarking program that solves the
 shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourney.  Originally written
-by Paul Schwartrauber of NCAR.  Organic development - optimised for
-previous hardware.
+finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
+Originally written by Paul Schwartrauber of NCAR.  Organic development
+- optimised for previous hardware.
+
+In what follows we consider square domains of dimension 64, 128, 256,
+512 and 1024.
 
@@
 it can often come with a performance penalty. In this work we
-investigate the origin of this penalty and the steps required to
-remove it.
+investigate in detail the hardware and compiler dependencies of this
+penalty and the steps required to remove it.
 
-\section{Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Methodology(?)}
 
-We have results for the Gnu, Intel, Cray and IBM compilers.
-We've run on Intel Ivy Bridge, Intel Haswell, and IBM BG/Q.
+Our aim in this work is to achieve portable performance. Consequently,
+we have performed tests with the hardware and compilers listed in
+Table~\ref{TABLE_compilers}. Where a compiler is available on a given
+piece of hardware, the version number used in this work is specified.
+The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
+3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
+system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
+cache of 30~MB. The CPU in the IBM BG/Q system is built around the
+PowerPC A2 core with a clock speed of 1.6~GHz. The chip has 32~MB of
+last-level cache.
+
+For completeness, we give the compiler flags used with each compiler
+in Table~\ref{TABLE_compiler_flags}. These flags are particularly
+important for the \psykal versions of the code. The performance of the
+original version of the code is very much less sensitive to the
+compiler options. This is because it consists of a single source file
+and a single routine (excluding those for I/O).
+
+We do retain I/O because any realistic code has to output results to disk.
+
+% An example of a floating table. Note that, for IEEE style tables, the 
+% \caption command should come BEFORE the table. Table text will default to
+% \footnotesize as IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+\begin{table}[!t]
+% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{The compilers and CPUs used in this work.}
+\label{TABLE_compilers}
+\centering
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
+\hline
+Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
+Intel Ivy Bridge &       & 14.0.1.106  & 8.3.3   &          \\
+IBM A2 (BGQ)     &       &             &         & 14.1     \\
+IBM Power 8      &       &             &         &          \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[!t]
+% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{The compiler flags used in this work.}
+\label{TABLE_compiler_flags}
+\centering
+\begin{tabular}{|l|c|}
+\hline
+Compiler  &  Flags \\
+\hline
+Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
+Intel     & -O3 -fast -fno-inline-factor    \\
+Cray      & -O3 -O ipa5 -h wp    \\
+IBM bgxlf & -O5 -qinline=auto:level=10    \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
+Beginning with the vanilla \psykal version, we apply a series of
+optimisations with the aim of recovering, as much as is possible, the
+structure and thus performance of the original version of the code.
+
+\begin{enumerate}
+
+\item Pass explicit array bounds to middle layer (rather than using
+  assumed-size array arguments)
+
+\item Move kernel routines into same module as middle layer
+
+\item Loop fusion
+
+\item In-lined field-copy operation
+
+\item Fused field-copy operation
+
+\end{enumerate}
+
+These optimisations do not always result in improved performance (AND
+HENCE WE SHOULD CALL THEM TRANSFORMATIONS PERHAPS?). Whether or not
+they do so depends both on the compiler and the problem size.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
+
+In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
+original version of the shallow code for the range of compilers and
+hardware considered here. This summary demonstrates the effect of the
+larger last-level cache of the Intel Ivy Bridge system compared to
+that of the Haswell system; note the significant drop in performance
+in going from a problem size of $256^{2}$ to a problem size of
+$512^{2}$ for the black and yellow bars. The performance of the Ivy
+Bridge system (red and purple bars) only drops to this level when the
+domain size is increased to $1024^{2}$. At this point, the working set
+no longer fits within cache and performance is dominated by the
+bandwidth to main memory making it relatively insensitive to the
+choice of compiler.
+
+For problem sizes that fit within cache there is
+little to choose between the performance obtained from the Intel and
+Gnu compilers.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=2.8in]{orig_summary}
 % where an .eps filename suffix will be assumed under latex, 
@@
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
 % Note that IEEE does not put floats in the very first column - or typically
@@
 
-\section{Conclusion}
+\section{Conclusions}
 The conclusion goes here.
@@
 % use section* for acknowledgement
-\section*{Acknowledgment}
+\section*{Acknowledgments}
 

### 2014-11-12 psykal_vanilla/shallow_base.f90
Alter timing of psykal version to include first time-step

@@ -95,6 +95,14 @@ program shallow
   ! Write intial values of p, u, and v into a netCDF file   
   CALL model_write(0, p, u, v)
 
+  !===================================================
+  ! Start timer. Tell the timing system that 
+  ! this single timed region actually contains itmax
+  ! repeats (albeit with the first time-step slightly
+  ! different from the rest because of the lack of
+  ! values from a previous step).
+  CALL timer_start('Time-stepping', idxt0, itmax)
+
   !====================================
   ! Perform the first time step
   CALL invoke_compute_cu(CU, P, U)
@@ -117,12 +125,7 @@ program shallow
   CALL copy_field(VNEW, V)
   CALL copy_field(PNEW, P)
 
-  !===================================================
-  ! Start timer. Tell the timing system that 
-  ! this single timed region actually contains itmax-1
-  ! repeats.
-  CALL timer_start('Time-stepping', idxt0, itmax-1)
-
+  !====================================
   !  ** Start of time-stepping loop proper ** 
   DO ncycle=2,itmax
     

### 2014-11-12 docs/papers/shallow_performance/shallow_perf.tex
Add table of summary perf no.s for 256 case

@@
 %
-% \usepackage{dblfloatfix}
+\usepackage{dblfloatfix}
 % The latest version can be found at:
@@
 
-\newcommand{\psykal}{{PS}y{KA}l}
+\newcommand{\psykal}{{PS}y{KA}l\ }
 
@@
 
+In what follows we consider square domains of dimension 64, 128, 256,
+512 and 1024.
+
 % needed in second column of first page if using \IEEEpubid
@@
 
-\section{Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Methodology(?)}
 
-We have results for the Gnu, Intel, Cray and IBM compilers.
-We've run on Intel Ivy Bridge, Intel Haswell, IBM BG/Q, IBM P8.
+Our aim in this work is to achieve portable performance. Consequently,
+we have performed tests with the hardware and compilers listed in
+Table~\ref{TABLE_compilers}. Where a compiler is available on a given
+piece of hardware, the version number used in this work is specified.
+The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
+3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
+system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
+cache of 30~MB. The CPU in the IBM BG/Q system is built around the
+PowerPC A2 core with a clock speed of 1.6~GHz. The chip has 32~MB of
+last-level cache.
+
+For completeness, we give the compiler flags used with each compiler
+in Table~\ref{TABLE_compiler_flags}. These flags are particularly
+important for the \psykal versions of the code. The performance of the
+original version of the code is very much less sensitive to the
+compiler options. This is because it consists of a single source file
+and a single routine (excluding those for I/O).
+
+We do retain I/O because any realistic code has to output results to disk.
+
+% An example of a floating table. Note that, for IEEE style tables, the 
+% \caption command should come BEFORE the table. Table text will default to
+% \footnotesize as IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+\begin{table}[!t]
+% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{The matrix of compilers and CPUs used in this work. The use
+  of a compiler on a given CPU is indicated by the specification of
+  the version of the compiler in the relevent element. No entry
+  indicates that a compiler was not available/used on that CPU.}
+\label{TABLE_compilers}
+\centering
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+                 & \multicolumn{4}{c|}{Compiler}             \\
+\hline
+                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
+\hline
+Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
+Intel Ivy Bridge &       & 14.0.1.106  & 8.3.3   &          \\
+IBM A2 (BGQ)     &       &             &         & 14.1     \\
+IBM Power 8      &       &             &         &          \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[!t]
+% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{The compiler flags used in this work.}
+\label{TABLE_compiler_flags}
+\centering
+\begin{tabular}{l|c}
+\hline
+Compiler  &  Flags \\
+\hline
+Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
+Intel     & -O3 -fast -fno-inline-factor    \\
+Cray      & -O3 -O ipa5 -h wp    \\
+IBM bgxlf & -O5 -qinline=auto:level=10    \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
+
 In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
 original version of the shallow code for the range of compilers and
-hardware considered here.
+hardware considered here. This summary demonstrates the effect of the
+larger last-level cache of the Intel Ivy Bridge system compared to
+that of the Haswell system; note the significant drop in performance
+in going from a problem size of $256^{2}$ to a problem size of
+$512^{2}$ for the black and yellow bars. The performance of the Ivy
+Bridge system (red and purple bars) only drops to this level when the
+domain size is increased to $1024^{2}$. At this point, the working set
+no longer fits within cache and performance is dominated by the
+bandwidth to main memory making it relatively insensitive to the
+choice of compiler.
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
+For problem sizes that fit within cache there is
+little to choose between the performance obtained from the Intel and
+Gnu compilers.
+
 \begin{figure}[!t]
@@
 \includegraphics[width=2.8in]{slowdown_summary}
-\caption{Summary of the performance achieved by any \psykal
-version as a percentage of that achieved by the original version of the code.}
+\caption{Comparison of the performance of the best \psykal
+version with that of the original version of the code. A negative value 
+indicates that the \psykal version is faster than the original.}
 \label{FIG_slowdown_summary}
 \end{figure}
+
+\begin{table*}[!t]
+% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
+code transformation.}
+\label{TABLE_opt_breakdown}
+\centering
+\begin{tabular}{l|c|c|c|c|c}
+\hline
+Compiler:                  &  Gnu    & Intel & Cray & Intel & IBM bgxlf \\
+\hline
+CPU:                       & Haswell & Haswell & Ivy Brige & Ivy Bridge & PowerPC A2 \\
+\hline
+Original                   &  1.49   & 1.41   & 1.24  & 1.57  &        \\
+Vanilla \psykal            & 25.27   & 1.73   & 1.67  & 1.96  &        \\
+Explicitly-sized arrays    & 25.28   & 1.72   & 1.39  & 1.93  & 113.74 \\
+Kernels copied into module &  2.20   & 1.73   & 1.42  & 1.93  &  68.28 \\
+Loops fused                &  2.18   & 1.58   & 1.47  & 1.76  &  71.91 \\
+In-lined field-copy        &  2.17   & 1.58   & 1.46  & 1.75  &  83.36 \\
+Fused field-copy           &  2.18   & 1.85   & 1.30  & 2.07  &  72.62 \\
+\hline
+\%-slow-down c.f. original &  45.3   & 11.7   & 5.0   & 11.0  &        \\
+\hline
+\end{tabular}
+\end{table*}
 
@@
 
-\section{Conclusion}
+\section{Conclusions}
 The conclusion goes here.
 
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
 % use section* for acknowledgement
-\section*{Acknowledgment}
+\section*{Acknowledgments}
 

### 2014-11-12 shallow/SEQ/original/shallow_base.f90
Correct bounds used in checksum calculations

@@ -417,11 +417,11 @@
       CALL timer_stop(idxt0)
 
       WRITE(6,"('P CHECKSUM after ',I6,' steps = ',E15.7)") &
-           itmax, SUM(PNEW(:,:))
+           itmax, SUM(PNEW(1:M,1:N))
       WRITE(6,"('U CHECKSUM after ',I6,' steps = ',E15.7)") &
-           itmax,SUM(UNEW(:,:))
+           itmax,SUM(UNEW(2:M+1,1:N))
       WRITE(6,"('V CHECKSUM after ',I6,' steps = ',E15.7)") &
-           itmax,SUM(VNEW(:,:))
+           itmax,SUM(VNEW(1:M,2:N+1))
 
  !     Close the netCDF file
 

### 2014-11-12 shallow/SEQ/shallow_base.f90

@@ -172,15 +172,15 @@ program shallow
 
   CALL timer_stop(idxt0)
 
-  CALL compute_checksum(pnew, csum)
+  CALL compute_checksum(pnew(1:m,1:n), csum)
   CALL model_write_log("('P CHECKSUM after ',I6,' steps = ',E15.7)", &
                        itmax, csum)
 
-  CALL compute_checksum(unew, csum)
+  CALL compute_checksum(unew(2:m+1,1:n), csum)
   CALL model_write_log("('U CHECKSUM after ',I6,' steps = ',E15.7)", &
                        itmax, csum)
 
-  CALL compute_checksum(vnew, csum)
+  CALL compute_checksum(vnew(1:m,2:n+1), csum)
   CALL model_write_log("('V CHECKSUM after ',I6,' steps = ',E15.7)", &
                        itmax, csum)
 

### 2014-11-12 shallow/SEQ/shallow_gocean.f90

@@ -171,15 +171,15 @@ PROGRAM shallow
 
   CALL timer_stop(idxt0)
 
-  CALL compute_checksum(pnew, csum)
+  CALL compute_checksum(pnew(1:m,1:n), csum)
   CALL model_write_log("('P CHECKSUM after ',I6,' steps = ',E15.7)", &
                        itmax, csum)
 
-  CALL compute_checksum(unew, csum)
+  CALL compute_checksum(unew(2:m+1,1:n), csum)
   CALL model_write_log("('U CHECKSUM after ',I6,' steps = ',E15.7)", &
                        itmax, csum)
 
-  CALL compute_checksum(vnew, csum)
+  CALL compute_checksum(vnew(1:m,2:n+1), csum)
   CALL model_write_log("('V CHECKSUM after ',I6,' steps = ',E15.7)", &
                        itmax, csum)
 

### 2014-11-12 shallow/SEQ/apply_bcs_cf_mod.f90
Remove manual_ prefix and re-structure with single invoke in time-stepping loop

@@ -9,7 +9,7 @@ module apply_bcs_cf_mod
 
   private
 
-  public manual_invoke_apply_bcs_cf
+  public invoke_apply_bcs_cf
   public apply_bcs_cf_type, apply_bcs_cf_code
 
   type, extends(kernel_type) :: apply_bcs_cf_type
@@ -29,7 +29,7 @@ contains
 
   !> Manual implementation of the code needed to invoke
   !! apply_bcs_cf_code().
-  subroutine manual_invoke_apply_bcs_cf(field)
+  subroutine invoke_apply_bcs_cf(field)
     use field_mod,    only: copy_field
     use topology_mod, only: cf
     implicit none
@@ -74,7 +74,7 @@ contains
 
     end do
 
-  end subroutine manual_invoke_apply_bcs_cf
+  end subroutine invoke_apply_bcs_cf
 
   !===================================================
 

### 2014-11-12 shallow/SEQ/apply_bcs_cu_mod.f90

@@ -10,7 +10,7 @@ module apply_bcs_cu_mod
 
   private
 
-  public manual_invoke_apply_bcs_cu
+  public invoke_apply_bcs_cu
   public apply_bcs_cu_type, apply_bcs_cu_code
 
   type, extends(kernel_type) :: apply_bcs_cu_type
@@ -30,7 +30,7 @@ contains
 
   !> Manual implementation of the code needed to invoke
   !! apply_bcs_cu_code().
-  subroutine manual_invoke_apply_bcs_cu(fld)
+  subroutine invoke_apply_bcs_cu(fld)
     use field_mod,    only: copy_field
     use topology_mod, only: cu
     implicit none
@@ -73,7 +73,7 @@ contains
        CALL copy_field(fld, cu%halo(ihalo)%src, cu%halo(ihalo)%dest)
     end do
 
-  end subroutine manual_invoke_apply_bcs_cu
+  end subroutine invoke_apply_bcs_cu
 
   !===================================================
 

### 2014-11-12 shallow/SEQ/compute_cv_mod.f90

@@ -9,7 +9,7 @@ module compute_cv_mod
 
   private
 
-  public manual_invoke_compute_cv
+  public invoke_compute_cv
   public compute_cv_type, compute_cv_code
 
   type, extends(kernel_type) :: compute_cv_type
@@ -31,7 +31,7 @@ contains
 
   !> Manual implementation of the code needed to invoke
   !! compute_cv_code().
-  subroutine manual_invoke_compute_cv(cvfld, p, v)
+  subroutine invoke_compute_cv(cvfld, p, v)
     use topology_mod, only: cv
     implicit none
     real(wp), intent(out), dimension(:,:) :: cvfld
@@ -87,7 +87,7 @@ contains
        end do
     end do
 
-  end subroutine manual_invoke_compute_cv
+  end subroutine invoke_compute_cv
 
   !===================================================
 

### 2014-11-12 shallow/SEQ/compute_h_mod.f90

@@ -6,7 +6,7 @@ MODULE compute_h_mod
 
   PRIVATE
 
-  PUBLIC manual_invoke_compute_h
+  PUBLIC invoke_compute_h
   PUBLIC compute_h_type, compute_h_code
 
   TYPE, EXTENDS(kernel_type) :: compute_h_type
@@ -27,7 +27,7 @@ CONTAINS
 
   !===================================================
 
-  subroutine manual_invoke_compute_h(h, p, u, v)
+  subroutine invoke_compute_h(h, p, u, v)
     use topology_mod, only: ct
     implicit none
     real(wp), intent(out), dimension(:,:) :: h
@@ -87,7 +87,7 @@ CONTAINS
        END DO
     END DO
 
-  END SUBROUTINE manual_invoke_compute_h
+  END SUBROUTINE invoke_compute_h
 
   !===================================================
 

### 2014-11-12 shallow/SEQ/compute_z_mod.f90

@@ -9,7 +9,7 @@ module compute_z_mod
 
   private
 
-  public manual_invoke_compute_z
+  public invoke_compute_z
   public compute_z_type, compute_z_code
 
   type, extends(kernel_type) :: compute_z_type
@@ -32,7 +32,7 @@ contains
 
   !> Manual implementation of the code needed to invoke
   !! compute_z_code().
-  subroutine manual_invoke_compute_z(z, p, u, v)
+  subroutine invoke_compute_z(z, p, u, v)
     use topology_mod, only: cf
     implicit none
     real(wp), intent(out), dimension(:,:) :: z
@@ -91,7 +91,7 @@ contains
        end do
     end do
 
-  end subroutine manual_invoke_compute_z
+  end subroutine invoke_compute_z
 
   !===================================================
 

### 2014-11-12 shallow/SEQ/manual_invoke_apply_bcs_mod.F90

@@ -3,42 +3,42 @@ MODULE manual_invoke_apply_bcs_mod
   IMPLICIT none
   PRIVATE
 
-  PUBLIC manual_invoke_apply_bcs_uvtf
-  PUBLIC manual_invoke_apply_bcs_uvt
+  PUBLIC invoke_apply_bcs_uvtf
+  PUBLIC invoke_apply_bcs_uvt
 
 CONTAINS
 
   !===================================================
 
-  SUBROUTINE manual_invoke_apply_bcs_uvtf(ufield, vfield, tfield, ffield)
-    USE apply_bcs_cf_mod, ONLY: manual_invoke_apply_bcs_cf
-    use apply_bcs_ct_mod, only: manual_invoke_apply_bcs_ct
-    use apply_bcs_cu_mod, only: manual_invoke_apply_bcs_cu
-    use apply_bcs_cv_mod, only: manual_invoke_apply_bcs_cv
+  SUBROUTINE invoke_apply_bcs_uvtf(ufield, vfield, tfield, ffield)
+    USE apply_bcs_cf_mod, ONLY: invoke_apply_bcs_cf
+    use apply_bcs_ct_mod, only: invoke_apply_bcs_ct
+    use apply_bcs_cu_mod, only: invoke_apply_bcs_cu
+    use apply_bcs_cv_mod, only: invoke_apply_bcs_cv
     IMPLICIT none
     REAL(wp), INTENT(inout), DIMENSION(:,:) :: ufield, vfield, tfield, ffield
 
-    call manual_invoke_apply_bcs_cu(ufield)
-    call manual_invoke_apply_bcs_ct(tfield)
-    call manual_invoke_apply_bcs_cv(vfield)
-    call manual_invoke_apply_bcs_cf(ffield)
+    call invoke_apply_bcs_cu(ufield)
+    call invoke_apply_bcs_ct(tfield)
+    call invoke_apply_bcs_cv(vfield)
+    call invoke_apply_bcs_cf(ffield)
 
-  END SUBROUTINE manual_invoke_apply_bcs_uvtf
+  END SUBROUTINE invoke_apply_bcs_uvtf
 
   !===================================================
 
-  subroutine manual_invoke_apply_bcs_uvt(ufield, vfield, tfield)
-    use apply_bcs_ct_mod, only: manual_invoke_apply_bcs_ct
-    use apply_bcs_cu_mod, only: manual_invoke_apply_bcs_cu
[…]

### 2014-11-12 shallow/SEQ/manual_invoke_compute_new_fields_mod.F90

@@ -3,11 +3,11 @@ MODULE manual_invoke_compute_new_fields_mod
   IMPLICIT none
   PRIVATE
 
-  PUBLIC manual_invoke_compute_new_fields
+  PUBLIC invoke_compute_new_fields
 
 CONTAINS
 
-  SUBROUTINE manual_invoke_compute_new_fields(unew, uold, &
+  SUBROUTINE invoke_compute_new_fields(unew, uold, &
                                               vnew, vold, &
                                               pnew, pold, &
                                               z, cufld, cvfld, hfld, tdt)
@@ -49,6 +49,6 @@ CONTAINS
        END DO
     END DO
 
-  END SUBROUTINE manual_invoke_compute_new_fields
+  END SUBROUTINE invoke_compute_new_fields
 
 END MODULE manual_invoke_compute_new_fields_mod

### 2014-11-12 shallow/SEQ/shallow_base.f90

@@ -40,16 +40,16 @@ program shallow
   use timing_mod
   use model_mod
   use initial_conditions_mod
-  use time_smooth_mod,  ONLY: manual_invoke_time_smooth
-  use apply_bcs_cu_mod, ONLY: manual_invoke_apply_bcs_cu
-  use apply_bcs_cv_mod, ONLY: manual_invoke_apply_bcs_cv
-  use manual_invoke_apply_bcs_mod, ONLY: manual_invoke_apply_bcs_uvtf
-  use manual_invoke_apply_bcs_mod, ONLY: manual_invoke_apply_bcs_uvt
-  use compute_cu_mod, ONLY: manual_invoke_compute_cu
-  use compute_cv_mod, ONLY: manual_invoke_compute_cv
-  use compute_z_mod,  ONLY: manual_invoke_compute_z
-  use compute_h_mod,  ONLY: manual_invoke_compute_h
-  use manual_invoke_compute_new_fields_mod, ONLY: manual_invoke_compute_new_fields
+  use time_smooth_mod,  ONLY: invoke_time_smooth
+  use apply_bcs_cu_mod, ONLY: invoke_apply_bcs_cu
+  use apply_bcs_cv_mod, ONLY: invoke_apply_bcs_cv
+  use manual_invoke_apply_bcs_mod, ONLY: invoke_apply_bcs_uvtf
+  use manual_invoke_apply_bcs_mod, ONLY: invoke_apply_bcs_uvt
+  use compute_cu_mod, ONLY: invoke_compute_cu
+  use compute_cv_mod, ONLY: invoke_compute_cv
+  use compute_z_mod,  ONLY: invoke_compute_z
+  use compute_h_mod,  ONLY: invoke_compute_h
+  use manual_invoke_compute_new_fields_mod, ONLY: invoke_compute_new_fields
   implicit none
 
   !type(field_type) :: pressure
@@ -88,8 +88,8 @@ program shallow
   CALL init_velocity_v(v, psi, m, n)
 
   !     PERIODIC CONTINUATION
-  CALL manual_invoke_apply_bcs_cu(U)
-  CALL manual_invoke_apply_bcs_cv(V)
+  CALL invoke_apply_bcs_cu(U)
+  CALL invoke_apply_bcs_cv(V)
 
   ! Initialise fields that will hold data at previous time step
   CALL copy_field(U, UOLD)
@@ -109,30 +109,30 @@ program shallow
 
[…]

### 2014-11-12 shallow/SEQ/shallow_gocean.f90

@@ -40,11 +40,12 @@ PROGRAM shallow
   use model_mod
   use initial_conditions_mod
   use time_smooth_mod,  only: time_smooth_type
-  use apply_bcs_cf_mod, only: manual_invoke_apply_bcs_cf
-  use apply_bcs_ct_mod, only: manual_invoke_apply_bcs_ct
-  use apply_bcs_cu_mod, only: manual_invoke_apply_bcs_cu
-  use apply_bcs_cv_mod, only: manual_invoke_apply_bcs_cv
-  use manual_invoke_apply_bcs_mod, only: manual_invoke_apply_bcs_uvtf
+  use apply_bcs_cf_mod, only: invoke_apply_bcs_cf
+  use apply_bcs_ct_mod, only: invoke_apply_bcs_ct
+  use apply_bcs_cu_mod, only: invoke_apply_bcs_cu
+  use apply_bcs_cv_mod, only: invoke_apply_bcs_cv
+  use manual_invoke_apply_bcs_mod, only: invoke_apply_bcs_uvtf, &
+                                         invoke_apply_bcs_uvt
   use compute_cu_mod, only: compute_cu_type
   use compute_cv_mod, only: compute_cv_type
   use compute_z_mod, only: compute_z_type
@@ -61,7 +62,7 @@ PROGRAM shallow
   INTEGER :: ncycle
    
   !> Integer tags for timers
-  INTEGER :: idxt0, idxt1
+  INTEGER :: idxt0
 
   !  ** Initialisations of model parameters (dt etc) ** 
   CALL model_init()
@@ -84,86 +85,71 @@ PROGRAM shallow
   CALL init_velocity_v(v, psi, m, n)
 
   !     PERIODIC CONTINUATION
-  CALL manual_invoke_apply_bcs_cu(U)
-  CALL manual_invoke_apply_bcs_cv(V)
+  CALL invoke_apply_bcs_cu(U)
+  CALL invoke_apply_bcs_cv(V)
 
   ! Initialise fields that will hold data at previous time step
   CALL copy_field(U, UOLD)
   CALL copy_field(V, VOLD)
   CALL copy_field(P, POLD)
      
+  !====================================
   ! Write intial values of p, u, and v into a netCDF file   
   CALL model_write(0, p, u, v)
 
+  !====================================
   !     Start timer
   CALL timer_start('Time-stepping',idxt0)
 
-  !  ** Start of time loop ** 
-  DO ncycle=1,itmax
[…]

### 2014-11-12 shallow/SEQ/time_smooth_mod.f90

@@ -6,7 +6,7 @@ MODULE time_smooth_mod
 
   PRIVATE
 
-  PUBLIC time_smooth_init, manual_invoke_time_smooth
+  PUBLIC time_smooth_init, invoke_time_smooth
   PUBLIC time_smooth_type, time_smooth_code
 
   !> Parameter for time smoothing
@@ -48,7 +48,7 @@ CONTAINS
 
   !> Manual implementation of code to invoke the time-smoothing
   !! kernel
-  SUBROUTINE manual_invoke_time_smooth(field, field_new, field_old)
+  SUBROUTINE invoke_time_smooth(field, field_new, field_old)
     IMPLICIT none
     REAL(wp), INTENT(in), DIMENSION(:,:) :: field
     REAL(wp), INTENT(in), DIMENSION(:,:) :: field_new
@@ -69,7 +69,7 @@ CONTAINS
       END DO
     END DO
 
-  END SUBROUTINE manual_invoke_time_smooth
+  END SUBROUTINE invoke_time_smooth
 
   !===================================================
 

### 2014-11-12 shallow_perf.tex
Move to expecting IEEEtran.cls to already be installed. Include bargraph.pl with the source for this paper

@@
 %
-\documentclass[journal]{./IEEEtran}
+\documentclass[journal]{IEEEtran}
 %
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{The compilers and CPUs used in this work.}
+\caption{The matrix of compilers and CPUs used in this work. The use
+  of a compiler on a given CPU is indicated by the specification of
+  the version of the compiler in the relevent element. No entry
+  indicates that a compiler was not available/used on that CPU.}
 \label{TABLE_compilers}
@@
 \begin{tabular}{|l|c|c|c|c|}
+\hline
+                 & \multicolumn{4}{c|}{Compiler}             \\
 \hline
@@
 \centering
-\begin{tabular}{|l|c|}
+\begin{tabular}{l|c}
 \hline
@@
 The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
 

### 2014-11-12 shallow_perf.tex
Beginnings of detailed discussion of results for 256 case

@@
 finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
-Originally written by Paul Schwartrauber of NCAR.  Organic development
-- optimised for previous hardware.
+Originally written by Paul Swarztrauber of the National Center for
+Atmospheric Research, US.  Organic development - optimised for
+previous hardware. In describing our work, we term the version of the
+shallow code obtained at the beginning of this work the `original'
+version.
 
@@
 
-For completeness, we give the compiler flags used with each compiler
-in Table~\ref{TABLE_compiler_flags}. These flags are particularly
-important for the \psykal versions of the code. The performance of the
-original version of the code is very much less sensitive to the
-compiler options. This is because it consists of a single source file
-and a single routine (excluding those for I/O).
+In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
+used with each compiler. These flags are particularly important for
+the \psykal versions of the code. The performance of the original
+version of the code is very much less sensitive to the compiler
+options. This is because it consists of a single source file
+containing effectively a single routine (excluding those for I/O).
 
@@
 no longer fits within cache and performance is dominated by the
-bandwidth to main memory making it relatively insensitive to the
+bandwidth to main memory, making it relatively insensitive to the
 choice of compiler.
 
-For problem sizes that fit within cache there is
-little to choose between the performance obtained from the Intel and
-Gnu compilers.
+For problem sizes that fit within cache there is little to choose
+between the performance obtained from the Intel and Gnu compilers on
+the Intel Haswell CPU.
 
@@
 \includegraphics[width=2.8in]{orig_summary}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
 \caption{Summary of the performance of the original shallow code on 
@@
 
+In Table~\ref{TABLE_opt_breakdown} we show detailed timings of the
+code after each of the transformations has been applied. For the Gnu
+compiler, the vanilla \psykal version of shallow is some 20 times
+slower than the original. Copying the code for each of the kernels
+into the module from which they are called has the most dramatic
+effect on the performance of the compiled code. None of the subsequent
+transformations succeed in significantly improving the performance of
+the compiled executable which is still some 45\% slower than the
+original.
+{\bf What is losing us this 45\%?}
+
 \begin{table*}[!t]
@@
 \hline
-CPU:                       & Haswell & Haswell & Ivy Brige & Ivy Bridge & PowerPC A2 \\
+CPU:                       & Haswell & Haswell & Ivy Bridge & Ivy Bridge & PowerPC A2 \\
 \hline
@@
 \hline
-\%-slow-down c.f. original &  45.3   & 11.7   & 5.0   & 11.0  &        \\
+\parbox{2.5cm}{\%-slow-down of best c.f. original} &  45.3   & 11.7   & 5.0   & 11.0  &        \\
 \hline
@@
 \end{table*}
+
+% 0.32/1.41 = 23% and 0.39/1.57 = 25%
+Considering now the Intel compiler, Table~\ref{TABLE_opt_breakdown}
+shows that even the vanilla \psykal version of shallow is only
+23--25\% slower than the original version. The only code
+transformation that produced a significant performance improvement for
+this case was the fusion of the computational loops. Further
+investigation of the slow-down caused by attempting to fuse the
+field-copy operation with the preceeding time-smoothing loop revealed
+that the compiler was unable to vectorise the loop after it had
+in-lined the body of the time-smoothing kernel. Once this in-lining
+was done by hand, the compiler was able to vectorise the loop and the
+performance of the original version was recovered.
+
+\begin{figure}
+\begin{verbatim}
+! Time smoothing
+DO J=1,N+1
+  DO I=1,M+1
+    CALL time_smooth_code(i,j,ufld,unew,uold)
+!    uold(i,j) = ufld(i,j) + alpha* &
+!        (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
+
+    CALL time_smooth_code(i,j,vfld,vnew,vold)
+!    vold(i,j) = vfld(i,j) + alpha* &
+!        (vnew(i,j) - 2.*vfld(i,j) + vold(i,j))
+
+    CALL time_smooth_code(i,j,pfld,pnew,pold)
+!    pold(i,j) = pfld(i,j) + alpha* &
+!        (pnew(i,j) - 2.*pfld(i,j) + pold(i,j))
+  END DO
+END DO
+
+! Copy new fields into current fields
+DO J=1,N+1
+  DO I=1,M+1
+    Ufld(I,J) = UNEW(I,J)
+    Vfld(I,J) = VNEW(I,J)
+    Pfld(I,J) = PNEW(I,J)
+  END DO
+END DO
+\end{verbatim}
+\caption{Example of the coding structure that required manual
+  in-lining of the kernel body (as indicated by commented-out lines)
+  to retrieve performance of original shallow with the Intel
+  compiler.}
+\label{FIG_time_smooth_code}
+\end{figure}
 

### 2014-11-13 SEQ/time_step_mod.f90
In-line time-smooth and compute_c{u,v}, compute_{z,h} kernels

@@ -11,6 +11,7 @@ contains
                               pfld, pnew, pold, &
                               hfld, zfld, tdt)
     use time_smooth_mod, only: alpha
+    use mesh_mod, only: fsdx, fsdy ! Properties of the grid (4.0/d{x,y})
     implicit none
     real(wp), dimension(M+1,N+1), intent(inout) :: cufld, cvfld
     real(wp), dimension(M+1,N+1), intent(inout) :: unew, vnew, pnew
@@ -31,18 +32,28 @@ contains
     do J= 1, N, 1
        do I = 1, M, 1
 
-          call compute_cu_code(i+1, j, cufld, pfld, ufld)
+          !call compute_cu_code(i+1, j, cufld, pfld, ufld)
+          CUfld(I+1,J) = .5*(Pfld(I+1,J)+Pfld(I,J))*Ufld(I+1,J)
 
-          call compute_cv_code(i, j+1, cvfld, pfld, vfld)
+          !call compute_cv_code(i, j+1, cvfld, pfld, vfld)
+          CVfld(I,J+1) = .5*(Pfld(I,J+1)+Pfld(I,J))*Vfld(I,J+1)
+
        end do
     end do
 
     do J= 1, N, 1
        do I= 1, M, 1
 
-          call compute_z_code(i+1, j+1, zfld, pfld, ufld, vfld)
+          !call compute_z_code(i+1, j+1, zfld, pfld, ufld, vfld)
+          Zfld(I+1,J+1) =(FSDX*(Vfld(I+1,J+1)-Vfld(I,J+1))- &
+                          FSDY*(Ufld(I+1,J+1)-Ufld(I+1,J)))/ &
+                   (pfld(I,J)+Pfld(I+1,J)+Pfld(I+1,J+1)+Pfld(I,J+1))
 
-          call compute_h_code(i, j, hfld, pfld, ufld, vfld)
+          !call compute_h_code(i, j, hfld, pfld, ufld, vfld)
+          Hfld(I,J) = Pfld(I,J) + &
+                      .25*(Ufld(I+1,J)*Ufld(I+1,J)+Ufld(I,J)*Ufld(I,J) & 
+                       + Vfld(I,J+1)*Vfld(I,J+1)+Vfld(I,J)*Vfld(I,J))
+
        END DO
     END DO
 
@@ -117,22 +128,15 @@ contains
     ! Loop over 'columns'
[…]

### 2014-11-13 SEQ/time_step_mod.f90
In-line remaining kernel code into invoke_time_step

@@ -11,7 +11,7 @@ contains
                               pfld, pnew, pold, &
                               hfld, zfld, tdt)
     use time_smooth_mod, only: alpha
-    use mesh_mod, only: fsdx, fsdy ! Properties of the grid (4.0/d{x,y})
+    use mesh_mod, only: dx, dy, fsdx, fsdy ! Properties of the grid (4.0/d{x,y})
     implicit none
     real(wp), dimension(M+1,N+1), intent(inout) :: cufld, cvfld
     real(wp), dimension(M+1,N+1), intent(inout) :: unew, vnew, pnew
@@ -23,6 +23,7 @@ contains
     real(wp),                     intent(in) :: tdt
     ! Locals
     integer :: I, J, idxt
+    REAL(wp) :: tdts8, tdtsdy, tdtsdx
 
     !============================================
     ! COMPUTE CAPITAL U, CAPITAL V, Z AND H
@@ -87,17 +88,30 @@ contains
     ! COMPUTE NEW VALUES U,V AND P
     call timer_start('Compute {U,V,P}NEW',idxt)
 
+    tdtsdx = tdt/dx
+    tdtsdy = tdt/dy
+    tdts8 = tdt/8.0d0
+
     DO J=1, N, 1
        DO I= 1, M, 1
 
-          CALL compute_unew_code(i+1, j, unew, uold, &
-                                 zfld, cvfld, hfld, tdt)
+          !CALL compute_unew_code(i+1, j, unew, uold, &
+          !                       zfld, cvfld, hfld, tdt)
+          UNEW(I,J) = UOLD(I,J) +                                 &
+                      TDTS8*(Zfld(I,J+1)+Zfld(I,J)) *                   &
+                      (CVfld(I,J+1)+CVfld(I-1,J+1)+CVfld(I-1,J)+CVfld(I,J)) - &
+                       TDTSDX*(Hfld(I,J)-Hfld(I-1,J))
 
-          CALL compute_vnew_code(i, j+1, vnew, vold, &
-                                 zfld, cufld, hfld, tdt)
+          !CALL compute_vnew_code(i, j+1, vnew, vold, &
+          !                       zfld, cufld, hfld, tdt)
+          VNEW(I,J) = VOLD(I,J)-TDTS8*(Zfld(I+1,J)+Zfld(I,J))           &
[…]

### 2014-11-17 docs/papers/shallow_performance/shallow_perf.tex
Update table of results and add plot of same data

@@
 %%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
 % Also note that the "draftcls" or "draftclsnofoot", not "draft", option
@@
 
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
 % Also, note that IEEEtran.cls V1.7 and later provides a builtin
@@
 Propose an approach for meeting this challenge by separating the
-computational science demands from the e.g. oceanographic demands.
+computational science demands from the {\it e.g.} oceanographic
+demands.
 
@@
 Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
-Intel Ivy Bridge &       & 14.0.1.106  & 8.3.3   &          \\
+Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
 IBM A2 (BGQ)     &       &             &         & 14.1     \\
@@
 
+In Table~\ref{TABLE_opt_breakdown} we show detailed timings for the
+$256^{2}$ case after each of the transformations has been applied to
+the code. For the Gnu compiler, the vanilla \psykal version of shallow
+is almost 20 times slower than the original. Copying the code for each
+of the kernels into the module from which they are called has the most
+dramatic effect on the performance of the compiled code. The next most
+significant transformation is that of (manually) in-lining the kernel
+code into the bodies of the loops from which they are called. This
+gets the \psykal version to within 8\% of the performance of the
+original.
+
 \begin{table*}[!t]
@@
 \centering
-\begin{tabular}{l|c|c|c|c|c}
+\begin{tabular}{l|c|c|c|c|c|c}
 \hline
-Compiler:                  &  Gnu    & Intel & Cray & Intel & IBM bgxlf \\
+Compiler:          &  Gnu    & Intel   & Gnu & Cray & Intel & IBM bgxlf \\
 \hline
-CPU:                       & Haswell & Haswell & Ivy Brige & Ivy Bridge & PowerPC A2 \\
+CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
 \hline
-Original                   &  1.49   & 1.41   & 1.24  & 1.57  &        \\
-Vanilla \psykal            & 25.27   & 1.73   & 1.67  & 1.96  &        \\
-Explicitly-sized arrays    & 25.28   & 1.72   & 1.39  & 1.93  & 113.74 \\
-Kernels copied into module &  2.20   & 1.73   & 1.42  & 1.93  &  68.28 \\
-Loops fused                &  2.18   & 1.58   & 1.47  & 1.76  &  71.91 \\
-In-lined field-copy        &  2.17   & 1.58   & 1.46  & 1.75  &  83.36 \\
-Fused field-copy           &  2.18   & 1.85   & 1.30  & 2.07  &  72.62 \\
+Original                             & 1.49 & 1.41 & 1.54 & 1.23 & 1.57  &        \\
+Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.67 & 1.96  &        \\
+Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.39 & 1.93  & 113.74 \\
+Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.42 & 1.93  &  68.28 \\
+Loops fused                          & 2.18 & 1.58 & 1.79 & 1.47 & 1.76  &  71.91 \\
+In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.46 & 1.75  &  83.36 \\
+Fused field-copy                     & 2.18 & 1.85 & 1.99 & 1.30 & 2.07  &  72.62 \\
+In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
+Fully fuse 1st loop nest             & 1.62 & ADD  & 1.55 & 1.25 & 1.55  &        \\
+Explicit, constant array/loop bounds & 1.61 & ADD  & 1.56 & 1.23 & 1.54  &        \\
 \hline
-\%-slow-down c.f. original &  45.3   & 11.7   & 5.0   & 11.0  &        \\
+\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 1.2 & 0.6 & 0.0 & -2.0 &     \\
 \hline
@@
 \end{table*}
+
+% 0.32/1.41 = 23% and 0.39/1.57 = 25%
+Considering now the Intel compiler, Table~\ref{TABLE_opt_breakdown}
+shows that even the vanilla \psykal version of shallow is only
+23--25\% slower than the original version. The only code
+transformation that produced a significant performance improvement for
+this case was the fusion of the computational loops. Further
+investigation of the slow-down caused by attempting to fuse the
+field-copy operation with the preceeding time-smoothing loop revealed
+that the compiler was unable to vectorise the loop after it had
+in-lined the body of the time-smoothing kernel. Once this in-lining
+was done by hand, the compiler was able to vectorise the loop and the
+performance of the original version was recovered.
+
+\begin{figure}
+\begin{verbatim}
+! Time smoothing
+DO J=1,N+1
+  DO I=1,M+1
+    CALL time_smooth_code(i,j,ufld,unew,uold)
+!    uold(i,j) = ufld(i,j) + alpha* &
+!        (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
+
+    CALL time_smooth_code(i,j,vfld,vnew,vold)
+!    vold(i,j) = vfld(i,j) + alpha* &
+!        (vnew(i,j) - 2.*vfld(i,j) + vold(i,j))
+
+    CALL time_smooth_code(i,j,pfld,pnew,pold)
+!    pold(i,j) = pfld(i,j) + alpha* &
+!        (pnew(i,j) - 2.*pfld(i,j) + pold(i,j))
+  END DO
+END DO
+
+! Copy new fields into current fields
+DO J=1,N+1
+  DO I=1,M+1
+    Ufld(I,J) = UNEW(I,J)
+    Vfld(I,J) = VNEW(I,J)
+    Pfld(I,J) = PNEW(I,J)
+  END DO
+END DO
+\end{verbatim}
+\caption{Example of the coding structure that required manual
+  in-lining of the kernel body (as indicated by commented-out lines)
+  to retrieve performance of original shallow with the Intel
+  compiler.}
+\label{FIG_time_smooth_code}
+\end{figure}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=85mm]{opt_stages_256}
+\caption{The absolute performance of the code at each stage of 
+optimisation for the $256^{2}$ domain.}
+\label{FIG_opt_stages_256}
+\end{figure}
 

### 2014-11-18 docs/papers/shallow_performance/shallow_perf.tex
Extending discussion of results.

@@
 Beginning with the vanilla \psykal version, we apply a series of
-optimisations with the aim of recovering, as much as is possible, the
-structure and thus performance of the original version of the code.
-
+optimisations while obeying the \psykal separation of concerns, {\it
+  i.e.} optimisation is restricted to the `middle layer' and leaves
+the kernel and algorithm layers unchanged. The aim of these
+optimisations is to recover, as much as is possible, the structure and
+thus performance of the original version of the code. The
+optimisations we have performed are as follows:
 \begin{enumerate}
@@
 
+\item Manually in-line kernel bodies into middle-layer code
+
+\item Fully fuse first loop (only the outer loop of this loop nest was
+  fused in the first pass)
+
+\item Make the array/loop bounds explicitly constant for the
+      duration of the time-stepping loop.
 \end{enumerate}
 
-These optimisations do not always result in improved performance (AND
-HENCE WE SHOULD CALL THEM TRANSFORMATIONS PERHAPS?). Whether or not
-they do so depends both on the compiler and the problem size.
+We shall see that these optimisations do not always result in improved
+performance. Whether or not they do so depends both on the compiler
+and the problem size. We also emphasise that the aim of these
+optimisations is to recover, as far as is possible, the structure of
+the original version of the code. It may well be that transforming the
+code into some other structure would result in better performance on a
+particular architecture. However, exploring this optimisation space is
+beyond the scope of the present work.
 
@@
 
-In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
+In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
 original version of the shallow code for the range of compilers and
@@
 in going from a problem size of $256^{2}$ to a problem size of
-$512^{2}$ for the black and yellow bars. The performance of the Ivy
-Bridge system (red and purple bars) only drops to this level when the
-domain size is increased to $1024^{2}$. At this point, the working set
-no longer fits within cache and performance is dominated by the
-bandwidth to main memory, making it relatively insensitive to the
-choice of compiler.
-
-For problem sizes that fit within cache there is little to choose
-between the performance obtained from the Intel and Gnu compilers on
-the Intel Haswell CPU.
+$512^{2}$ for the black and green bars. The performance of the Ivy
+Bridge system with the Cray or Intel compiler (red and purple bars)
+only drops to this level when the domain size is increased to
+$1024^{2}$. At this point, the working set no longer fits within cache
+and performance is dominated by the bandwidth to main memory, making
+it relatively insensitive to the choice of compiler.
 
@@
 
+For this compiler-friendly form of the code, there is little
+performance difference between the executables produced by the Gnu and
+Intel compilers. However, the executable produced by the Cray compiler
+generally performs significantly better. {\bf Why is this; can we work
+out what it's doing that the others aren't?}
+
+Note that the results for the Gnu compiler on Ivy Bridge show the same
+drop-off (as problem size is increased) as those on the Haswell
+system. This may be because Archer's login nodes seem to be running on
+older hardware (E5-2650 with 20MB LLC) during the Archer upgrade. {\bf
+  Need to re-run these cases once Archer is back up.}
+
+Moving now the to the \psykal version of shallow,
+Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
+fastest \psykal version for each of the compiler/CPU combinations. The
+only significant difference between this and the plot of the
+performance of the original version of the code in
+Figure~\ref{FIG_orig_perf_summary} is that the difference between the
+Gnu- and Intel-compiled executables is slightly more marked. This is
+clarified in Figure~\ref{FIG_slowdown_summary} which plots the
+percentage difference between the performance of the original and the
+(best) \psykal versions of shallow for each compiler/CPU combination.
+It is clear from this plot that the difference in the performance of
+the Gnu- and Intel-compiled excutables on the Haswell system is due to
+the Gnu compiler being unable to recover the performance achieved by
+the original code. In contrast there is no such problem for the
+Gnu-compiled executable on the Ivy Bridge system. We attribute this to
+the more recent version of the compiler on the latter system (4.9.1 as
+opposed to 4.8.2).
+
 \begin{figure}[!t]
 \centering
-\includegraphics[width=2.8in]{best_psykal_summary}
+\includegraphics[width=7.5cm]{best_psykal_summary}
 \caption{Summary of the best performance achieved by any \psykal 
@@
 
-In Table~\ref{TABLE_opt_breakdown} we show detailed timings for the
-$256^{2}$ case after each of the transformations has been applied to
-the code. For the Gnu compiler, the vanilla \psykal version of shallow
-is almost 20 times slower than the original. Copying the code for each
-of the kernels into the module from which they are called has the most
-dramatic effect on the performance of the compiled code. The next most
+In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
+\psykal version of shallow is never more than 2\% slower than the
+original and in some cases is faster.  This demonstrates that we can
+reliably recover the performance of the original version of the code,
+despite the significant restructuring required by the \psykal
+approach. {\bf We need to re-run/check the Intel results on Ivy Bridge
+  once Archer is back from its upgrade.}
+
+Having show that, in general, we can recover and occasionally improve
+upon the performance of the original version of shallow, the next
+logical step is to examine the necessary code transformations.  We do
+this for the $256^{2}$ case since this fits within cache on both the
+Haswell and Ivy Bridge CPUs.  Table~\ref{TABLE_opt_breakdown} shows
+detailed timings for this case after each transformation has been
+applied to the code. The same data is visualised in
+Figure~\ref{FIG_opt_stages_256} where the performance of the best
+\psykal version for a given compiler/CPU is given relative to the
+performance of the original with the same compiler/CPU. For the Gnu
+compiler, the vanilla \psykal version of shallow is almost 20 times
+slower than the original. Copying the code for each of the kernels
+into the module from which they are called has the most dramatic
+effect on the performance of the compiled code. The next most
 significant transformation is that of (manually) in-lining the kernel
@@
 In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
-Fully fuse 1st loop nest             & 1.62 & ADD  & 1.55 & 1.25 & 1.55  &        \\
-Explicit, constant array/loop bounds & 1.61 & ADD  & 1.56 & 1.23 & 1.54  &        \\
+Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.25 & 1.55  &        \\
+Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.23 & 1.54  &        \\
 \hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 1.2 & 0.6 & 0.0 & -2.0 &     \\
+\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
 \hline
@@
 \centering
-\includegraphics[width=85mm]{opt_stages_256}
-\caption{The absolute performance of the code at each stage of 
-optimisation for the $256^{2}$ domain.}
+\includegraphics[width=90mm]{opt_stages_256}
+\caption{The performance of the \psykal version of shallow for the
+  $256^{2}$ domain at each stage of optimisation. Results are given as
+  a percentage of the performance of the original code for each
+  compiler/CPU combination. A figure greater than 100\% indicates that
+  the \psykal version performs better than the original.}
 \label{FIG_opt_stages_256}
@@
 
-The authors would like to thank...
+This work made use of the ARCHER UK National Supercomputing Service
+(\url{http://www.archer.ac.uk}).
 

### 2014-11-18 docs/papers/shallow_performance/shallow_perf.tex
Add detailed discussion of Gnu, Intel and Cray perf evolution

@@
 this for the $256^{2}$ case since this fits within cache on both the
-Haswell and Ivy Bridge CPUs.  Table~\ref{TABLE_opt_breakdown} shows
-detailed timings for this case after each transformation has been
-applied to the code. The same data is visualised in
-Figure~\ref{FIG_opt_stages_256} where the performance of the best
-\psykal version for a given compiler/CPU is given relative to the
-performance of the original with the same compiler/CPU. For the Gnu
-compiler, the vanilla \psykal version of shallow is almost 20 times
-slower than the original. Copying the code for each of the kernels
-into the module from which they are called has the most dramatic
-effect on the performance of the compiled code. The next most
-significant transformation is that of (manually) in-lining the kernel
-code into the bodies of the loops from which they are called. This
-gets the \psykal version to within 8\% of the performance of the
-original.
+Haswell and Ivy Bridge CPUs we are using here.
+Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
+after each transformation has been applied to the code. The same data
+is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
+of the \psykal version for a given compiler/CPU is given relative to
+the performance of the original with the same compiler/CPU.
 
@@
 \hline
-Compiler:          &  Gnu    & Intel   & Gnu & Cray & Intel & IBM bgxlf \\
+Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
 \hline
@@
 \hline
-Original                             & 1.49 & 1.41 & 1.54 & 1.23 & 1.57  &        \\
-Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.67 & 1.96  &        \\
-Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.39 & 1.93  & 113.74 \\
-Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.42 & 1.93  &  68.28 \\
-Loops fused                          & 2.18 & 1.58 & 1.79 & 1.47 & 1.76  &  71.91 \\
-In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.46 & 1.75  &  83.36 \\
-Fused field-copy                     & 2.18 & 1.85 & 1.99 & 1.30 & 2.07  &  72.62 \\
-In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
-Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.25 & 1.55  &        \\
-Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.23 & 1.54  &        \\
+Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
+Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
+Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
+Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
+Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
+In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
+Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
+In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
+Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
+Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
 \hline
@@
 
-% 0.32/1.41 = 23% and 0.39/1.57 = 25%
-Considering now the Intel compiler, Table~\ref{TABLE_opt_breakdown}
-shows that even the vanilla \psykal version of shallow is only
-23--25\% slower than the original version. The only code
-transformation that produced a significant performance improvement for
-this case was the fusion of the computational loops. Further
-investigation of the slow-down caused by attempting to fuse the
-field-copy operation with the preceeding time-smoothing loop revealed
-that the compiler was unable to vectorise the loop after it had
-in-lined the body of the time-smoothing kernel. Once this in-lining
-was done by hand, the compiler was able to vectorise the loop and the
-performance of the original version was recovered.
+With the Gnu compiler (first and third clusters in
+Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
+shallow only achieves ~5-6\% of the performance of the
+original. Simply copying the code for each of the kernel subroutines
+into the module from which they are called has the most dramatic
+effect on the performance of the compiled code; it now achieves almost
+80\% of the performance of the original. The next most significant
+transformation is that of (manually) in-lining the kernel code itself into
+the bodies of the loops from which they are called. This gets the
+\psykal version to within 8\% of the performance of the original.
+{\bf What's losing us 8\% of performance with the Gnu compiler?}
+
+With the Intel compiler, the key transformations to recover
+performance are different (second and fourth clusters in
+Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
+shows that even the vanilla \psykal version of shallow achieves some
+80\% of the performance of the original version. Only two code
+transformations were required to increase this to 100\%. The first of
+these was the fusion of the computational loops. However, we found
+that performance suffered (purple bar) when this was extended to
+include the loop required for field copies at the end of a time-step
+(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
+slow-down revealed that the compiler was unable to SIMD vectorise the
+newly-fused loop after it had in-lined the body of the kernel called
+from within it. Once this in-lining was done by hand (dark-blue bar),
+the compiler was able to vectorise the loop and the performance of the
+original version of shallow was recovered.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=95mm]{opt_stages_256}
+\caption{The performance of the \psykal version of shallow for the
+  $256^{2}$ domain at each stage of optimisation. Results are given as
+  a percentage of the performance of the original code for each
+  compiler/CPU combination. A figure greater than 100\% indicates that
+  the \psykal version performs better than the original.}
+\label{FIG_opt_stages_256}
+\end{figure}
 
@@
 !     (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
-
-    CALL time_smooth_code(i,j,vfld,vnew,vold)
-!    vold(i,j) = vfld(i,j) + alpha* &
-!        (vnew(i,j) - 2.*vfld(i,j) + vold(i,j))
-
-    CALL time_smooth_code(i,j,pfld,pnew,pold)
-!    pold(i,j) = pfld(i,j) + alpha* &
-!        (pnew(i,j) - 2.*pfld(i,j) + pold(i,j))
   END DO
@@
     Ufld(I,J) = UNEW(I,J)
-    Vfld(I,J) = VNEW(I,J)
-    Pfld(I,J) = PNEW(I,J)
   END DO
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=90mm]{opt_stages_256}
-\caption{The performance of the \psykal version of shallow for the
-  $256^{2}$ domain at each stage of optimisation. Results are given as
-  a percentage of the performance of the original code for each
-  compiler/CPU combination. A figure greater than 100\% indicates that
-  the \psykal version performs better than the original.}
-\label{FIG_opt_stages_256}
-\end{figure}
+The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
+evolution of the performance of the \psykal version of shallow with
+the Cray compiler. Again, this has significant differences from the
+behaviour seen with the Gnu and Intel compilers. As with the Intel
+compiler, the performance of the Vanilla \psykal version is fairly
+good at 74\% of that of the original. In contrast to the other
+compilers consisdered so far, the first significant transformation is
+simply to specify the bounds of the arrays being used within the
+middle layer using module variables (as opposed to specifying them as
+assumed size). This tells the compiler that all of the arrays used in
+the middle layer have the same extent and also that all of the
+computational loops are over almost all elements of these arrays. This
+simple change gives a \psykal version that achieves 88\% of the
+performance of the original.
+
+Subsequent transformations actually harm performance until the
+field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
+with the preceeding loop. The resulting \psykal version now achieves
+94\% of the original performance. Two further steps are required to
+match the latter. First, the first loop nest has to be fully fused
+(both inner and outer loops fused) and second, the array/loop bounds
+are passed as an argument to the middle layer. This is done in order
+to specify to the compiler that these bounds remain constant for the
+duration of the time-stepping loop.
 

### 2014-11-19 docs/papers/shallow_performance/shallow_perf.tex
Add schematic of vanilla psykal shallow

@@
 \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/
 
@@
 
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
 \usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/url/
-% Basically, \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 % Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
-for IEEE journal papers produced under \LaTeX\ using
-IEEEtran.cls version 1.8 and later.
+% and "HE" in caps to complete the first word.
+\IEEEPARstart{T}{he} challenge presented to scientific software developers by
+the drive towards Exascale computing is considerable.
 
@@
 shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
+finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
 Originally written by Paul Swarztrauber of the National Center for
@@
 
-In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
-used with each compiler. These flags are particularly important for
-the \psykal versions of the code. The performance of the original
-version of the code is very much less sensitive to the compiler
-options. This is because it consists of a single source file
-containing effectively a single routine (excluding those for I/O).
-
-We do retain I/O because any realistic code has to output results to disk.
-
 % An example of a floating table. Note that, for IEEE style tables, the 
@@
 
+In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
+used with each compiler. These flags are particularly important for
+the \psykal versions of the code. The performance of the original
+version of the code is very much less sensitive to the compiler
+options. This is because it consists of a single source file
+containing effectively a single routine (excluding those for I/O).
+
+Since any real oceanographic computational model must output results,
+we ensure that any \psykal version of shallow retains the Input/Output
+capability of the original. This aids in limiting the optimisations
+that can be performed on the \psykal version to those that should also
+be applicable to full oceanographic models. Note that although we
+retain the I/O functionality, all of the results presented in this work
+carefully exclude the effects of I/O since it is compute performance
+that interests us here.
+
 \begin{table}[!t]
@@
 \end{table}
+
+In order to maximise the flexibility (and thus potential for
+optimisation) of the \psykal version of shallow, we made the kernels
+as fine-grained as possible. In this case, this resulted in eight
+distinct kernels, each of which operated on a single field at a single
+point (since we have chosen to use point-wise kernels).
+
+\begin{figure}
+\centering
+\includegraphics[width=85mm]{psykal_shallow}
+\caption{A schematic of the vanilla \psykal version of the shallow code.}
+\label{FIG_psykal_shallow_structure}
+\end{figure}
 
@@
 80\% of the performance of the original. The next most significant
-transformation is that of (manually) in-lining the kernel code itself into
-the bodies of the loops from which they are called. This gets the
-\psykal version to within 8\% of the performance of the original.
-{\bf What's losing us 8\% of performance with the Gnu compiler?}
+transformation is that of (manually) in-lining the kernel code itself
+into the bodies of the loops from which they are called. This gets the
+\psykal version to within 8\% of the performance of the original.  The
+dramatic performance improvement obtained from the first
+transformation demonstrates that the Gnu compiler is unable to
+optimise over separate files. Therefore, the fact that the
+time-stepping loop in the \psykal version still retains a call to an
+I/O routine and a call to the middle-layer, both of which are in
+different files, is likely to account for this remaining performance
+penalty of 8\%.
 
@@
   in-lining of the kernel body (as indicated by commented-out lines)
-  to retrieve performance of original shallow with the Intel
-  compiler.}
+  to retrieve the performance of the original version of shallow with
+  the Intel compiler.}
 \label{FIG_time_smooth_code}

### 2014-11-19 docs/papers/shallow_performance/shallow_perf.tex
More text on the psykal structure of shallow

@@
 distinct kernels, each of which operated on a single field at a single
-point (since we have chosen to use point-wise kernels).
+point (since we have chosen to use point-wise kernels). With a little
+bit of tidying/re-structuring, we found it was possible to express the
+contents of the main time-stepping loop as a single invoke (a call to
+the PSy layer) and a call to the I/O system
+(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
+routine then consists of applying each of the kernels to all of the
+points on the model mesh requiring an update.
+
+{\bf How we handle the PBCs.}
 

### 2014-11-19 docs/papers/shallow_performance/shallow_perf.tex
More work describing vanilla psykal

@@
 % can use linebreaks \\ within to get better formatting as desired
-% Do not put math or special symbols in the title.
 \title{Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)}
 %
-%
 % author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
 % use \thanks{} to gain access to the first footnote area
@@
 % Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
+% you are supposed to have spaces between the names.
 
@@
 % and "HE" in caps to complete the first word.
-\IEEEPARstart{T}{he} challenge presented to scientific software developers by
-the drive towards Exascale computing is considerable.
-
-\subsection{The `shallow' Program}
-For this work we have used a benchmarking program that solves the
-shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
-Originally written by Paul Swarztrauber of the National Center for
-Atmospheric Research, US.  Organic development - optimised for
-previous hardware. In describing our work, we term the version of the
-shallow code obtained at the beginning of this work the `original'
-version.
-
-In what follows we consider square domains of dimension 64, 128, 256,
-512 and 1024.
+\IEEEPARstart{T}{he} challenge presented to the developers of
+scientific software by the drive towards Exascale computing is
+considerable.
 
@@
 penalty and the steps required to remove it.
+
+\subsection{The `shallow' Program}
+For this work we have used a benchmarking program that solves the
+shallow-water equations on a bi-periodic plane following the
+finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
+Originally written by Paul Swarztrauber of the National Center for
+Atmospheric Research, US.  Organic development - optimised for
+previous hardware. In describing our work, we term the version of the
+shallow code obtained at the beginning of this work the `original'
+version.
+
+In what follows we consider square domains of dimension 64, 128, 256,
+512 and 1024.
+
+Since any real oceanographic computational model must output results,
+we ensure that any \psykal version of shallow retains the Input/Output
+capability of the original. This aids in limiting the optimisations
+that can be performed on the \psykal version to those that should also
+be applicable to full oceanographic models. Note that although we
+retain the I/O functionality, all of the results presented in this work
+carefully exclude the effects of I/O since it is compute performance
+that interests us here.
+
+In order to maximise the flexibility (and thus potential for
+optimisation) of the \psykal version of shallow, we made the kernels
+as fine-grained as possible. In this case, this resulted in eight
+distinct kernels, each of which operated on a single field at a single
+point (since we have chosen to use point-wise kernels). With a little
+bit of tidying/re-structuring, we found it was possible to express the
+contents of the main time-stepping loop as a single invoke (a call to
+the PSy layer) and a call to the I/O system
+(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
+routine then consists of applying each of the kernels to all of the
+points on the model mesh requiring an update. In its basic,
+unoptimised (`vanilla') form, this PSy-layer routine then contains a
+doubly-nested loop around each kernel call, as indicated by the
+pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.
+
+\begin{figure}
+\centering
+\includegraphics[width=85mm]{psykal_shallow}
+\caption{A schematic of the vanilla \psykal version of the shallow code.}
+\label{FIG_psykal_shallow_structure}
+\end{figure}
+
+As with any full oceanographic model, boundary conditions must be
+applied at the edges of the model domain. In the case of Shallow we
+have periodic boundary conditions in both the $x$ and $y$ dimensions.
+These are simply implemented by having additional (`halo')
+rows/columns at the edges of the domain and ensuring the data in them
+is up-to-date before it is read. These rows/columns are updated by
+copying data from the corresponding row/column on the opposite side of
+the domain. In the current work, the lines of code to do these copies
+is manually inserted in the PSy-layer routine as required.
+
+{\bf Auto-generated BC code by reasoning about meta-data. But then,
+  that brings us on to meta-data...}
 
@@
 
-Since any real oceanographic computational model must output results,
-we ensure that any \psykal version of shallow retains the Input/Output
-capability of the original. This aids in limiting the optimisations
-that can be performed on the \psykal version to those that should also
-be applicable to full oceanographic models. Note that although we
-retain the I/O functionality, all of the results presented in this work
-carefully exclude the effects of I/O since it is compute performance
-that interests us here.
-
 \begin{table}[!t]
@@
 \end{table}
-
-In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of shallow, we made the kernels
-as fine-grained as possible. In this case, this resulted in eight
-distinct kernels, each of which operated on a single field at a single
-point (since we have chosen to use point-wise kernels). With a little
-bit of tidying/re-structuring, we found it was possible to express the
-contents of the main time-stepping loop as a single invoke (a call to
-the PSy layer) and a call to the I/O system
-(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
-routine then consists of applying each of the kernels to all of the
-points on the model mesh requiring an update.
-
-{\bf How we handle the PBCs.}
-
-\begin{figure}
-\centering
-\includegraphics[width=85mm]{psykal_shallow}
-\caption{A schematic of the vanilla \psykal version of the shallow code.}
-\label{FIG_psykal_shallow_structure}
-\end{figure}
 

### 2014-11-20 shallow_perf.tex
Work-up the description of Shallow and the work done with it

--- 2014-11-12 shallow_perf.tex
@@
 % Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
-for IEEE journal papers produced under \LaTeX\ using
-IEEEtran.cls version 1.8 and later.
-
-\subsection{The `shallow' Program}
-For this work we have used a benchmarking program that solves the
-shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
-Originally written by Paul Schwartrauber of NCAR.  Organic development
-- optimised for previous hardware.
-
-In what follows we consider square domains of dimension 64, 128, 256,
-512 and 1024.
+% and "HE" in caps to complete the first word.
+\IEEEPARstart{T}{he} challenge presented to the developers of
+scientific software by the drive towards Exascale computing is
+considerable.
 
@@
 penalty and the steps required to remove it.
+
+\subsection{The `shallow' Program}
+For this work we have used a benchmarking program that solves the
+shallow-water equations on a bi-periodic plane following the
+finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
+This software was originally written in 1984 by Paul Swarztrauber of
+the National Center for Atmospheric Research, US.  However, in common
+with many scientific codes, it has subsequently undergone some sort of
+evolutionary development with subsequent people making various changes
+and optimising it for previous generations of hardware.  In describing
+our work, we term the version of the Shallow program obtained at the
+beginning of this project the `original' version.
+
+Shallow is a very good test case for our purposes since the original
+version is short (some 600 lines) and contained within a single source
+file. This makes it relatively straightforward for a compiler to
+optimise. Its performance is thus quite a demanding target for our
+modified versions of the code to reproduce.
+
+Since any real oceanographic computational model must output results,
+we ensure that any \psykal version of shallow retains the Input/Output
+capability of the original. This aids in limiting the optimisations
+that can be performed on the \psykal version to those that should also
+be applicable to full oceanographic models. Note that although we
+retain the I/O functionality, all of the results presented in this work
+carefully exclude the effects of I/O since it is compute performance
+that interests us here.
+
+In order to maximise the flexibility (and thus potential for
+optimisation) of the \psykal version of shallow, we made the kernels
+as fine-grained as possible. In this case, this resulted in eight
+distinct kernels, each of which operated on a single field at a single
+point (since we have chosen to use point-wise kernels). With a little
+bit of tidying/re-structuring, we found it was possible to express the
+contents of the main time-stepping loop as a single invoke (a call to
+the PSy layer) and a call to the I/O system
+(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
+routine then consists of applying each of the kernels to all of the
+points on the model mesh requiring an update. In its basic,
+unoptimised (`vanilla') form, this PSy-layer routine then contains a
+doubly-nested loop around each kernel call, as indicated by the
+pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.
+
+\begin{figure}
+\centering
+\includegraphics[width=85mm]{psykal_shallow}
+\caption{A schematic of the vanilla \psykal version of the shallow code.}
+\label{FIG_psykal_shallow_structure}
+\end{figure}
+
+As with any full oceanographic model, boundary conditions must be
+applied at the edges of the model domain. In the case of Shallow we
+have periodic boundary conditions in both the $x$ and $y$ dimensions.
+These are simply implemented by having additional (`halo')
+rows/columns at the edges of the domain and ensuring the data in them
+is up-to-date before it is read. These rows/columns are updated by
+copying data from the corresponding row/column on the opposite side of
+the domain. In the current work, the lines of code to do these copies
+is manually inserted in the PSy-layer routine as required.
+
+{\bf Auto-generated BC code by reasoning about meta-data. But then,
+  that brings us on to meta-data...}
 
@@
 
-For completeness, we give the compiler flags used with each compiler
-in Table~\ref{TABLE_compiler_flags}. These flags are particularly
-important for the \psykal versions of the code. The performance of the
-original version of the code is very much less sensitive to the
-compiler options. This is because it consists of a single source file
-and a single routine (excluding those for I/O).
-
-We do retain I/O because any realistic code has to output results to disk.
-
 % An example of a floating table. Note that, for IEEE style tables, the 
@@
 \renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
 \caption{The matrix of compilers and CPUs used in this work. The use
@@
 
+In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
+used with each compiler. These flags are particularly important for
+the \psykal versions of the code. The performance of the original
+version of the code is much less sensitive to the compiler
+options. This is because it consists of a single source file
+containing effectively a single routine (excluding those for I/O).
+
 \begin{table}[!t]
@@
 \renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
 \caption{The compiler flags used in this work.}
@@
 
-\begin{itemize}
+Before applying any code transformations, we benchmark the original
+version of the code. We also benchmark the vanilla, unoptimised
+version after it has been re-structured following the \psykal
+approach. These two versions of the code effectively provide upper and
+lower bounds, respectively, on the performance we expect to achieve.
 
-\item Performance of the 'original' version of shallow; i.e. the version we
-were supplied with at the start of this work.
-
-\item Performance of the vanilla, unoptimised version after it has been
-re-structured following the \psykal approach.
-
-\end{itemize}
-
-Beginning with the vanilla \psykal version, we apply a series of
-optimisations with the aim of recovering, as much as is possible, the
-structure and thus performance of the original version of the code.
-
+Beginning with the vanilla \psykal version, we then apply a series of
+code transformations while obeying the \psykal separation of concerns,
+{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
+the kernel and algorithm layers unchanged. The aim of these
+optimisations is to recover, as much as is possible, the structure and
+thus performance of the original version of the code. The
+transformations we have performed are as follows:
 \begin{enumerate}
@@
 
+\item Manually in-line kernel bodies into middle-layer code
+
+\item Fully fuse first loop (only the outer loop of this loop nest was
+  fused in the first pass)
+
+\item Make the array/loop bounds explicitly constant for the
+      duration of the time-stepping loop.
 \end{enumerate}
 
-These optimisations do not always result in improved performance (AND
-HENCE WE SHOULD CALL THEM TRANSFORMATIONS PERHAPS?). Whether or not
-they do so depends both on the compiler and the problem size.
+We shall see that these optimisations do not always result in improved
+performance. Whether or not they do so depends both on the compiler
+and the problem size. We also emphasise that the aim of these
+optimisations is to recover, as far as is possible, the structure of
+the original version of the code. It may well be that transforming the
+code into some other structure would result in better performance on a
+particular architecture. However, exploring this optimisation space is
+beyond the scope of the present work.
+
+We explore the extent to which performance depends upon the problem
+size by using square domains of dimension 64, 128, 256, 512 and
+1024. This range allows us to investigate what happens when cache is
+exhausted as well as giving us some insight into the decisions that
+different compilers make when optimising the code.
 
@@
 in going from a problem size of $256^{2}$ to a problem size of
-$512^{2}$ for the black and yellow bars. The performance of the Ivy
-Bridge system (red and purple bars) only drops to this level when the
-domain size is increased to $1024^{2}$. At this point, the working set
-no longer fits within cache and performance is dominated by the
-bandwidth to main memory making it relatively insensitive to the
-choice of compiler.
-
-For problem sizes that fit within cache there is
-little to choose between the performance obtained from the Intel and
-Gnu compilers.
+$512^{2}$ for the black and green bars. The performance of the Ivy
+Bridge system with the Cray or Intel compiler (red and purple bars)
+only drops to this level when the domain size is increased to
+$1024^{2}$. At this point, the working set no longer fits within cache
+and performance is dominated by the bandwidth to main memory, making
+it relatively insensitive to the choice of compiler.
 
@@
 
+In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
+\psykal version of shallow is never more than 2\% slower than the
+original and in some cases is faster.  This demonstrates that we can
+reliably recover the performance of the original version of the code,
+despite the significant restructuring required by the \psykal
+approach. {\bf We need to re-run/check the Intel results on Ivy Bridge
+  once Archer is back from its upgrade.}
+
+Having show that, in general, we can recover and occasionally improve
+upon the performance of the original version of shallow, the next
+logical step is to examine the necessary code transformations.  We do
+this for the $256^{2}$ case since this fits within cache on both the
+Haswell and Ivy Bridge CPUs we are using here.
+Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
+after each transformation has been applied to the code. The same data
+is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
+of the \psykal version for a given compiler/CPU is given relative to
+the performance of the original with the same compiler/CPU.
+
 \begin{table*}[!t]
@@
 \centering
-\begin{tabular}{l|c|c|c|c|c}
+\begin{tabular}{l|c|c|c|c|c|c}
 \hline
-Compiler:                  &  Gnu    & Intel & Cray & Intel & IBM bgxlf \\
+Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
 \hline
-CPU:                       & Haswell & Haswell & Ivy Brige & Ivy Bridge & PowerPC A2 \\
+CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
 \hline
-Original                   &  1.49   & 1.41   & 1.24  & 1.57  &        \\
-Vanilla \psykal            & 25.27   & 1.73   & 1.67  & 1.96  &        \\
-Explicitly-sized arrays    & 25.28   & 1.72   & 1.39  & 1.93  & 113.74 \\
-Kernels copied into module &  2.20   & 1.73   & 1.42  & 1.93  &  68.28 \\
-Loops fused                &  2.18   & 1.58   & 1.47  & 1.76  &  71.91 \\
-In-lined field-copy        &  2.17   & 1.58   & 1.46  & 1.75  &  83.36 \\
-Fused field-copy           &  2.18   & 1.85   & 1.30  & 2.07  &  72.62 \\
+Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
+Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
+Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
+Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
+Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
+In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
+Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
+In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
+Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
+Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
 \hline
-\%-slow-down c.f. original &  45.3   & 11.7   & 5.0   & 11.0  &        \\
+\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
 \hline
@@
 \end{table*}
+
+With the Gnu compiler (first and third clusters in
+Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
+shallow only achieves ~5-6\% of the performance of the
+original. Simply copying the code for each of the kernel subroutines
+into the module from which they are called has the most dramatic
+effect on the performance of the compiled code; it now achieves almost
+80\% of the performance of the original. The next most significant
+transformation is that of (manually) in-lining the kernel code itself
+into the bodies of the loops from which they are called. This gets the
+\psykal version to within 8\% of the performance of the original.  The
+dramatic performance improvement obtained from the first
+transformation demonstrates that the Gnu compiler is unable to
+optimise over separate files. Therefore, the fact that the
+time-stepping loop in the \psykal version still retains a call to an
+I/O routine and a call to the middle-layer, both of which are in
+different files, is likely to account for this remaining performance
+penalty of 8\%.
+
+With the Intel compiler, the key transformations to recover
+performance are different (second and fourth clusters in
+Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
+shows that even the vanilla \psykal version of shallow achieves some
+80\% of the performance of the original version. Only two code
+transformations were required to increase this to 100\%. The first of
+these was the fusion of the computational loops. However, we found
+that performance suffered (purple bar) when this was extended to
+include the loop required for field copies at the end of a time-step
+(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
+slow-down revealed that the compiler was unable to SIMD vectorise the
+newly-fused loop after it had in-lined the body of the kernel called
+from within it. Once this in-lining was done by hand (dark-blue bar),
+the compiler was able to vectorise the loop and the performance of the
+original version of shallow was recovered.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=95mm]{opt_stages_256}
+\caption{The performance of the \psykal version of shallow for the
+  $256^{2}$ domain at each stage of optimisation. Results are given as
+  a percentage of the performance of the original code for each
+  compiler/CPU combination. A figure greater than 100\% indicates that
+  the \psykal version performs better than the original.}
+\label{FIG_opt_stages_256}
+\end{figure}
+
+\begin{figure}
+\begin{verbatim}
+! Time smoothing
+DO J=1,N+1
+  DO I=1,M+1
+    CALL time_smooth_code(i,j,ufld,unew,uold)
+!    uold(i,j) = ufld(i,j) + alpha* &
+!     (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
+  END DO
+END DO
+
+! Copy new fields into current fields
+DO J=1,N+1
+  DO I=1,M+1
+    Ufld(I,J) = UNEW(I,J)
+  END DO
+END DO
+\end{verbatim}
+\caption{Example of the coding structure that required manual
+  in-lining of the kernel body (as indicated by commented-out lines)
+  to retrieve the performance of the original version of shallow with
+  the Intel compiler.}
+\label{FIG_time_smooth_code}
+\end{figure}
+
+The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
+evolution of the performance of the \psykal version of shallow with
+the Cray compiler. Again, this has significant differences from the
+behaviour seen with the Gnu and Intel compilers. As with the Intel
+compiler, the performance of the Vanilla \psykal version is fairly
+good at 74\% of that of the original. In contrast to the other
+compilers consisdered so far, the first significant transformation is
+simply to specify the bounds of the arrays being used within the
+middle layer using module variables (as opposed to specifying them as
+assumed size). This tells the compiler that all of the arrays used in
+the middle layer have the same extent and also that all of the
+computational loops are over almost all elements of these arrays. This
+simple change gives a \psykal version that achieves 88\% of the
+performance of the original.
+
+Subsequent transformations actually harm performance until the
+field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
+with the preceeding loop. The resulting \psykal version now achieves
+94\% of the original performance. Two further steps are required to
+match the latter. First, the first loop nest has to be fully fused
+(both inner and outer loops fused) and second, the array/loop bounds
+are passed as an argument to the middle layer. This is done in order
+to specify to the compiler that these bounds remain constant for the
+duration of the time-stepping loop.
 

### 2014-11-20 shallow_perf.tex
WIP on the first paragraph

@@
 scientific software by the drive towards Exascale computing is
-considerable.
+considerable. With power consumption becoming the overriding design
+constraint, CPU clock speeds are falling and the complex,
+multi-purpose compute core is being replaced multiple, simpler
+cores. This philosophy can be seen at work in the rise of so-called
+accelerator based machines in the Top 500 List~\cite{top500} of
+supercomputers: five of the top-ten machines in the November 2014 list
+make use of Intel Xeon Phi's or NVIDIA GPUs. Four of the remaining
+five machines in the top ten are IBM BlueGene/Qs, the CPU of which has
+hardware support for running 64 threads.
 
@@
 
-We shall see that these optimisations do not always result in improved
+We shall see that these transformations do not always result in improved
 performance. Whether or not they do so depends both on the compiler

### 2014-11-21 shallow_perf.tex
Add text to Intro and Conclusions

@@
 constraint, CPU clock speeds are falling and the complex,
-multi-purpose compute core is being replaced multiple, simpler
+multi-purpose compute core is being replaced by multiple, simpler
 cores. This philosophy can be seen at work in the rise of so-called
@@
 hardware support for running 64 threads.
+
+Achieving good performance on large numbers of light-weight cores
+requires exploiting as much parallelism in an application as possible
+and this results in increased complexity in the programming models
+that must be used. This in turn increases the burden of code
+maintenance and code development, in part because two specialisms are
+required: that of the scientific domain which a code is modelling
+({\it e.g.} oceanography) and that of computational science. The
+situation is currently complicated still further by the existence of
+competing hardware technology; if one was to begin writing a major
+scientific application today it is unclear whether one would target
+GPU, Xeon Phi, traditional CPU, FPGA or something else entirely. This
+is a problem because, generally speaking, these different technologies
+require different programming approaches.
 
@@
 despite the significant restructuring required by the \psykal
-approach. {\bf We need to re-run/check the Intel results on Ivy Bridge
-  once Archer is back from its upgrade.}
+approach.
 
@@
 % line do not exceed the text width or a line break will occur.
+
 %
@@
 \section{Conclusions}
-The conclusion goes here.
+
+We have investigated the application of the \psykal separation of
+concerns approach to the domain of finite-difference ocean
+models. This approach enables the computational science (performance)
+related aspects of a computer model to be kept separate from the
+natural (oceanographic) science aspects. In this work we have shown
+that the \psykal approach has the potential to improve performance and
+to improve code structure/maintainability.
+
+The application of code transformations to the middle/PSy layer is key
+to the performance of the \psykal version of a code.
 

### 2014-11-21 shallow/SEQ/shallow_base.f90
Move 0.2 API shallow to single invoke

@@ -52,6 +52,7 @@ program shallow
   use compute_unew_mod, only: invoke_compute_unew
   use compute_vnew_mod, only: invoke_compute_vnew
   use compute_pnew_mod, only: invoke_compute_pnew
+  use time_step_mod,    only: invoke_time_step
   implicit none
 
   type(grid_type), target :: model_grid
@@ -74,7 +75,7 @@ program shallow
   INTEGER :: ncycle
    
   !> Integer tags for timers
-  INTEGER :: idxt0, idxt1
+  INTEGER :: idxt0
 
   ! Create the model grid
   model_grid = grid_type(ARAKAWA_C,                           &
@@ -154,79 +155,46 @@ program shallow
   !     Start timer
   CALL timer_start('Time-stepping',idxt0)
 
-  !  ** Start of time loop ** 
-  DO ncycle=1,itmax
+  !====================================
+  ! Perform the first time step
+  CALL invoke_compute_cu(CU_fld, P_fld, U_fld)
+  CALL invoke_compute_cv(CV_fld, P_fld, V_fld)
+  CALL invoke_compute_z(z_fld, P_fld, U_fld, V_fld)
+  CALL invoke_compute_h(h_fld, P_fld, U_fld, V_fld)
+
+  CALL invoke_apply_bcs(cu_fld)
+  CALL invoke_apply_bcs(cv_fld)
+  CALL invoke_apply_bcs(h_fld)
+  CALL invoke_apply_bcs(z_fld)
+
+  CALL invoke_compute_unew(unew_fld, uold_fld,  z_fld, cv_fld, &
+                           h_fld, tdt%data)
+  CALL invoke_compute_vnew(vnew_fld, vold_fld,  z_fld, cu_fld, &
+                           h_fld, tdt%data)
+  CALL invoke_compute_pnew(pnew_fld, pold_fld, cu_fld, cv_fld, &
+                           tdt%data)
+
+  CALL invoke_apply_bcs(UNEW_fld)
+  CALL invoke_apply_bcs(VNEW_fld)
+  CALL invoke_apply_bcs(PNEW_fld)
+
+  ! Set tdt to = 2*dt
+  CALL increment_field(tdt, tdt)
+
+  CALL copy_field(UNEW_fld, U_fld)
+  CALL copy_field(VNEW_fld, V_fld)
+  CALL copy_field(PNEW_fld, P_fld)
+
+  !====================================
+  !  ** Start of time-stepping loop proper ** 
[…]

### 2014-11-21 shallow/SEQ/time_step_mod.f90

@@ -1,0 +1,172 @@
+module time_step_mod
+  use kind_params_mod
+  use field_mod
+  implicit none
+
+contains
+
+  subroutine invoke_time_step(cufld, cvfld, ufld, unew, uold, &
+                              vfld, vnew, vold, &
+                              pfld, pnew, pold, &
+                              hfld, zfld, tdt)
+    use compute_cu_mod, only: compute_cu_code
+    use compute_cv_mod, only: compute_cv_code
+    use compute_z_mod,  only: compute_z_code
+    use compute_h_mod,  only: compute_h_code
+    use compute_unew_mod, only: compute_unew_code
+    use compute_vnew_mod, only: compute_vnew_code
+    use compute_pnew_mod, only: compute_pnew_code
+    use time_smooth_mod, only: time_smooth_code
+    use apply_bcs_mod, only: invoke_apply_bcs
+    implicit none
+    type(r2d_field), intent(inout) :: cufld, cvfld
+    type(r2d_field), intent(inout) :: unew, vnew, pnew
+    type(r2d_field), intent(inout) :: hfld, zfld, pfld, &
+                                               ufld, vfld
+    type(r2d_field), intent(inout) :: uold, vold, pold
+    type(scalar_field), intent(in) :: tdt
+
+    ! Locals
+    integer :: I, J, idim1, idim2
+    real(wp) :: dx, dy
+
+    ! COMPUTE CAPITAL U, CAPITAL V, Z AND H
+
+    do J=cufld%internal%ystart, cufld%internal%ystop
+       do I=cufld%internal%xstart, cufld%internal%xstop
+
+          call compute_cu_code(i, j, cufld%data, pfld%data, ufld%data)
+       end do
+    end do
+    do J=cvfld%internal%ystart, cvfld%internal%ystop
+       do I=cvfld%internal%xstart, cvfld%internal%xstop
+
+          call compute_cv_code(i, j, cvfld%data, pfld%data, vfld%data)
+       end do
+    end do
+    dx = zfld%grid%dx
+    dy = zfld%grid%dy
+
+    do J=zfld%internal%ystart, zfld%internal%ystop, 1
+       do I=zfld%internal%xstart, zfld%internal%xstop, 1
+
+          call compute_z_code(i, j,      &
+                              zfld%data, &
+                              pfld%data, &
+                              ufld%data, &
+                              vfld%data, &
+                              dx, dy)
+
[…]

### 2014-11-21 direct_psykal_manual/time_step_mod.f90
WIP putting nemolite into single-invoke form

@@ -1,0 +1,51 @@
+module time_step_mod
+  implicit none
+
+  private
+
+  public invoke_time_step
+
+contains
+
+  subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
+                              sshn, sshn_u, sshn_v, &
+                              hu, hv, ht, ua, va, un, vn)
+    use field_mod
+    use momentum_mod, only: invoke_momentum_u, invoke_momentum_v
+    use continuity_mod, only: invoke_continuity
+    use time_update_mod, only: invoke_next_sshu, invoke_next_sshv
+    use boundary_conditions_mod
+    implicit none
+    integer,         intent(in) :: istp
+    type(r2d_field), intent(inout) :: un, vn, sshn, sshn_u, sshn_v
+    type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
+    type(r2d_field), intent(in) :: hu, hv, ht
+
+  CALL invoke_continuity(ssha, sshn, sshn_u, sshn_v, &
+                         hu, hv, un, vn)
+
+  CALL invoke_momentum_u(ua, un, vn, &
+                         hu, hv, ht, &
+                         ssha_u, sshn, sshn_u, sshn_v)
+
+  CALL invoke_momentum_v(va, un, vn, &
+                         hu, hv, ht, &
+                         ssha_v, sshn, sshn_u, sshn_v)
+
+  ! Apply open and solid boundary conditions
+  CALL invoke_bc_ssh(istp, ssha)
+  CALL invoke_bc_solid_u(ua)
+  CALL invoke_bc_solid_v(va)
+  CALL invoke_bc_flather_u(ua, hu, sshn_u)
+  CALL invoke_bc_flather_v(va, hv, sshn_v)
+
+  ! Time update of fields
+  call copy_field(ua, un)
+  call copy_field(va, vn)
+  call copy_field(ssha, sshn)
+  call invoke_next_sshu(sshn_u, sshn)
+  call invoke_next_sshv(sshn_v, sshn)
+
+  end subroutine invoke_time_step
+
+end module time_step_mod

### 2014-11-21 gocean2d_direct.f90
WIP putting nemolite into single-invoke form

@@ -35,7 +35,8 @@ program gocean2d
   ! Create the model grid. We use a NE offset (i.e. the U, V and F
   ! points immediately to the North and East of a T point all have the
   ! same i,j index).  This is the same offset scheme as used by NEMO.
-  model_grid = grid_type(ARAKAWA_C,                           &
+  model_grid = grid_type(ARAKAWA_C, &
+  !  BC_PERIODIC, BC_NON_PERIODIC ??
                          (/BC_EXTERNAL,BC_EXTERNAL,BC_NONE/), &
                          OFFSET_NE)
 
@@ -108,10 +109,7 @@ subroutine step(grid, istp, &
   use kind_params_mod
   use grid_mod
   use field_mod
-  use momentum_mod, only: invoke_momentum_u, invoke_momentum_v
-  use continuity_mod, only: invoke_continuity
-  use time_update_mod, only: invoke_next_sshu, invoke_next_sshv
-  use boundary_conditions_mod
+  use time_step_mod, only: invoke_time_step
   use gocean2d_io_mod, only: model_write
   implicit none
   type(grid_type), intent(in) :: grid
@@ -120,30 +118,10 @@ subroutine step(grid, istp, &
   type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
   type(r2d_field), intent(in) :: hu, hv, ht
 
-  CALL invoke_continuity(ssha, sshn, sshn_u, sshn_v, &
-                         hu, hv, un, vn)
+  call invoke_time_step(istp, ssha, ssha_u, ssha_v, &
+                        sshn, sshn_u, sshn_v, &
+                        hu, hv, ht, ua, va, un, vn)
 
-  CALL invoke_momentum_u(ua, un, vn, &
-                         hu, hv, ht, &
-                         ssha_u, sshn, sshn_u, sshn_v)
-
-  CALL invoke_momentum_v(va, un, vn, &
-                         hu, hv, ht, &
-                         ssha_v, sshn, sshn_u, sshn_v)
-
-  ! Apply open and solid boundary conditions
-  CALL invoke_bc_ssh(istp, ssha)
-  CALL invoke_bc_solid_u(ua)
-  CALL invoke_bc_solid_v(va)
-  CALL invoke_bc_flather_u(ua, hu, sshn_u)
-  CALL invoke_bc_flather_v(va, hv, sshn_v)
-
-  ! Time update of fields
[…]

### 2014-11-21 gocean2d_direct.f90
Call kernels directly from go2d single invoke

@@ -113,6 +113,7 @@ subroutine step(grid, istp, &
   use gocean2d_io_mod, only: model_write
   implicit none
   type(grid_type), intent(in) :: grid
+  !> The current time step
   integer,         intent(in) :: istp
   type(r2d_field), intent(inout) :: un, vn, sshn, sshn_u, sshn_v
   type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v

### 2014-11-21 direct_psykal_manual/boundary_conditions_mod.f90
Call kernels directly from go2d single invoke

@@ -12,6 +12,8 @@ module boundary_conditions_mod
   public invoke_bc_solid_u,   invoke_bc_solid_v
   public invoke_bc_flather_u, invoke_bc_flather_v
   public invoke_bc_ssh
+  public bc_ssh_code, bc_solid_u_code, bc_solid_v_code
+  public bc_flather_u_code, bc_flather_v_code
 
   !=======================================
 
@@ -152,14 +154,7 @@ contains
     integer,            intent(in)    :: istep
     type(r2d_field),    intent(inout) :: ssha
     ! Locals
-    real(wp) :: amp_tide, omega_tide
     integer  :: ji, jj
-
-    amp_tide   = 0.2_wp
-    omega_tide = 2.0_wp * 3.14159_wp / (12.42_wp * 3600._wp)
-
-    !        DO jj = 1, jpj  
-    !          DO ji = 1, jpi 
 
     DO jj = ssha%internal%ystart, ssha%internal%ystop
        DO ji = ssha%internal%xstart, ssha%internal%xstop

### 2014-11-21 direct_psykal_manual/momentum_mod.f90

@@ -10,6 +10,7 @@ module momentum_mod
   private
 
   public invoke_momentum_u, invoke_momentum_v
+  public momentum_u_code, momentum_v_code
 
   !=======================================
 

### 2014-11-21 direct_psykal_manual/time_step_mod.f90

@@ -8,43 +8,130 @@ module time_step_mod
 contains
 
   subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
-                              sshn, sshn_u, sshn_v, &
+                              sshn_t, sshn_u, sshn_v, &
                               hu, hv, ht, ua, va, un, vn)
     use field_mod
-    use momentum_mod, only: invoke_momentum_u, invoke_momentum_v
-    use continuity_mod, only: invoke_continuity
-    use time_update_mod, only: invoke_next_sshu, invoke_next_sshv
+    use model_mod, only: rdt
+    use momentum_mod, only: momentum_u_code, momentum_v_code
+    use continuity_mod, only: continuity_code
+    use time_update_mod, only: next_sshu_code, next_sshv_code
     use boundary_conditions_mod
     implicit none
-    integer,         intent(in) :: istp
-    type(r2d_field), intent(inout) :: un, vn, sshn, sshn_u, sshn_v
+    integer,         intent(in)    :: istp
+    type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
     type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
-    type(r2d_field), intent(in) :: hu, hv, ht
+    type(r2d_field), intent(in)    :: hu, hv, ht
+    ! Locals
+    integer :: ji, jj
 
-  CALL invoke_continuity(ssha, sshn, sshn_u, sshn_v, &
-                         hu, hv, un, vn)
+    do jj = ssha%internal%ystart, ssha%internal%ystop, 1
+      do ji = ssha%internal%xstart, ssha%internal%xstop, 1
 
-  CALL invoke_momentum_u(ua, un, vn, &
-                         hu, hv, ht, &
-                         ssha_u, sshn, sshn_u, sshn_v)
+        call continuity_code(ji, jj,                             &
+                             ssha%data, sshn_t%data,             &
+                             sshn_u%data, sshn_v%data,           &
+                             hu%data, hv%data, un%data, vn%data, &
+                             rdt, ssha%grid%area_t)
+      end do
+    end do
 
-  CALL invoke_momentum_v(va, un, vn, &
-                         hu, hv, ht, &
-                         ssha_v, sshn, sshn_u, sshn_v)
+    do jj = ua%internal%ystart, ua%internal%ystop, 1
+      do ji = ua%internal%xstart, ua%internal%xstop, 1
 
[…]

### 2015-02-10 api_v0.2/field_mod.f90
Add tile setup to field initialisation

@@ -41,16 +41,11 @@ module field_mod
   end TYPE scalar_field
 
   type, public, extends(field_type) :: r2d_field
-     !> Array holding the actual field values
-     real(wp), dimension(:,:), allocatable :: data
-  end type r2d_field
-
-  type, public, extends(field_type) :: tiled_r2d_field
      integer :: ntiles
      type(tile_type), dimension(:), allocatable :: tile
      !> Array holding the actual field values
      real(wp), dimension(:,:), allocatable :: data
-  end type tiled_r2d_field
+  end type r2d_field
 
   !interface set_field
   !   module procedure set_scalar_field
@@ -79,14 +74,14 @@ module field_mod
   end interface r2d_field
 
   ! User-defined constructor for tiled_r2d_field type objects
-  interface tiled_r2d_field
-     module procedure tiled_r2d_field_constructor
-  end interface tiled_r2d_field
+!  interface tiled_r2d_field
+!     module procedure tiled_r2d_field_constructor
+!  end interface tiled_r2d_field
 
   !> Interface for the field checksum operation. Overloaded to take either
   !! a field object or a 2D, real(wp) array.
   interface field_checksum
-     module procedure tiled_fld_checksum, fld_checksum, array_checksum
+     module procedure fld_checksum, array_checksum
   end interface field_checksum
 
   !> Info on the tile sizes
@@ -130,6 +125,9 @@ module field_mod
   !! \todo remove this parameter and determine it from
   !! the supplied T mask.
   integer, public, parameter :: NBOUNDARY = 1
+
+  !> Whether or not to 'tile' (sub-divide) field arrays
+  logical, public, parameter :: TILED_FIELDS = .TRUE.
 
 contains
 
@@ -175,7 +173,9 @@ contains
 
     ! Set-up the limits of the 'internal' region of this field
     !
-    call set_field_bounds(self,grid_points)
+    call set_field_bounds(self,fld_type,grid_points)
+
+    call tile_setup(self)
 
     ! We allocate all fields to have the same extent as that
     ! with the greatest extents. This enables the (Cray) compiler
@@ -228,84 +228,7 @@ contains
 
   !===================================================
 
-  function tiled_r2d_field_constructor(grid,    &
-                                       grid_points) result(self)
-    implicit none
-    ! Arguments
[…]

### 2015-02-10 api_v0.2/omp_tiling_mod.f90

@@ -7,7 +7,7 @@ module omp_tiling_mod
   TYPE :: tile_type
      type(region_type) :: internal
      type(region_type) :: whole
-     real(wp), dimension(:,:), allocatable :: data
+     !real(wp), dimension(:,:), allocatable :: data
   END TYPE tile_type
 
   INTEGER, SAVE                                    :: ntiles
@@ -21,7 +21,7 @@ module omp_tiling_mod
   INTEGER, SAVE                                    :: max_tile_height
 
   public openmp_grid_init
-  public tile_type, ntiles, tile
+  public tile_type
 
 contains
 
@@ -235,7 +235,8 @@ contains
           END IF
 
           IF(print_tiles)THEN
-             WRITE(*,"('tile[',I4,'](',I4,':',I4,')(',I4,':',I4,'), interior:(',I4,':',I4,')(',I4,':',I4,') ')")                                       &
+             WRITE(*,"('tile[',I4,'](',I4,':',I4,')(',I4,':',I4,'), "// &
+                  &  "interior:(',I4,':',I4,')(',I4,':',I4,') ')") &
                   ith,                                                 &
                   tile(ith)%whole%xstart, tile(ith)%whole%xstop,       &
                   tile(ith)%whole%ystart, tile(ith)%whole%ystop,       &

### 2015-02-10 nemolite2d_psykal_omp/nemolite2d.f90
Change psykal nemolite2d to use tiling

@@ -119,7 +119,7 @@ subroutine step(grid, istp, &
   use kind_params_mod
   use grid_mod
   use field_mod
-  use time_step_mod, only: invoke_time_step
+  use time_step_mod, only: invoke_time_step_tiled
   use gocean2d_io_mod, only: model_write
   implicit none
   type(grid_type), intent(in) :: grid
@@ -129,7 +129,7 @@ subroutine step(grid, istp, &
   type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
   type(r2d_field), intent(in) :: hu, hv, ht
 
-  call invoke_time_step(istp, ssha, ssha_u, ssha_v, &
+  call invoke_time_step_tiled(istp, ssha, ssha_u, ssha_v, &
                         sshn, sshn_u, sshn_v, &
                         hu, hv, ht, ua, va, un, vn)
 

### 2015-02-10 nemolite2d_psykal_omp/time_step_mod.f90

@@ -3,7 +3,7 @@ module time_step_mod
 
   private
 
-  public invoke_time_step
+  public invoke_time_step, invoke_time_step_tiled
 
 contains
 
@@ -602,9 +602,9 @@ contains
     use boundary_conditions_mod
     implicit none
     integer,         intent(in)    :: istp
-    type(tiled_r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
-    type(tiled_r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
-    type(tiled_r2d_field), intent(in)    :: hu, hv, ht
+    type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
+    type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
+    type(r2d_field), intent(in)    :: hu, hv, ht
     ! Locals
     integer :: it, ji, jj, jiu, jiv
     integer :: M, N, idxt
@@ -673,7 +673,7 @@ contains
           do ji = ssha%tile(it)%internal%xstart, ssha%tile(it)%internal%xstop, 1
 
         call continuity_code(ji, jj,                             &
-                             ssha%tile(it)%data, sshn_t%data,             &
+                             ssha%data, sshn_t%data,             &
                              sshn_u%data, sshn_v%data,           &
                              hu%data, hv%data, un%data, vn%data, &
                              rdt, sshn_t%grid%area_t)
@@ -694,18 +694,12 @@ contains
 
 !    call timer_start('Momentum',idxt)
 
-!    do jj = ua%internal%ystart, ua%internal%ystop, 1
-!      do ji = ua%internal%xstart, ua%internal%xstop, 1
 !dir$ safe_address
 !$OMP DO SCHEDULE(RUNTIME)
-    do jj = 2, N, 1
-! SIMD
+    do it = 1, ua%ntiles, 1
+       do jj= ua%tile(it)%internal%ystart, ua%tile(it)%internal%ystop, 1
 !dir$ vector always
-      do ji = 2, M-1, 1
-!OMP DO SCHEDULE(RUNTIME)
-!    do it = 1, ntiles, 1
-!       do jj= tile(it)%internal%jstart, tile(it)%internal%jstop, 1
[…]

### 2015-02-11 api_v0.2/field_mod.f90
Fix whole/internal tile extents. Remove omp_tiling_mod module.

@@ -3,7 +3,6 @@ module field_mod
   use region_mod
   use halo_mod
   use grid_mod
-  use omp_tiling_mod
   use gocean_mod, only: gocean_stop
   implicit none
 
@@ -17,6 +16,13 @@ module field_mod
   integer, public, parameter :: F_POINTS   = 3
   !> A field that lives on all grid-points of the grid
   integer, public, parameter :: ALL_POINTS = 4
+
+  !> A field is sub-divided into tiles for coarse-grained OpenMP
+  type :: tile_type
+     type(region_type) :: internal
+     type(region_type) :: whole
+     !real(wp), dimension(:,:), allocatable :: data
+  end type tile_type
 
   type, public :: field_type
      !> Which mesh points the field is defined upon
@@ -42,6 +48,8 @@ module field_mod
 
   type, public, extends(field_type) :: r2d_field
      integer :: ntiles
+     !> The dimensions of the tiles into which the field
+     !! is sub-divided.
      type(tile_type), dimension(:), allocatable :: tile
      !> Array holding the actual field values
      real(wp), dimension(:,:), allocatable :: data
@@ -1068,14 +1076,13 @@ contains
     !> Optional specification of the dimensions of the tiling grid
     integer, intent(in), optional :: nx_arg, ny_arg
     integer :: nx, ny
-    INTEGER :: idx, idy, ival, jval ! For tile extent calculation
+    INTEGER :: ival, jval ! For tile extent calculation
     integer :: internal_width, internal_height
     INTEGER :: ierr, nwidth
     INTEGER :: ji,jj, ith
     INTEGER :: nthreads       ! No. of OpenMP threads being used
-    INTEGER :: jover, junder, idytmp, jover_per_row, junder_per_row
-    INTEGER :: iover, iunder, idxtmp, iover_per_col, iunder_per_col
-    LOGICAL :: nested_par     ! Whether OpenMP supports nested parallelism
+    INTEGER :: jover, junder, idytmp
+    INTEGER :: iover, iunder, idxtmp
     ! For doing stats on tile sizes
     INTEGER :: nvects, nvects_sum, nvects_min, nvects_max 
     LOGICAL, PARAMETER :: print_tiles = .TRUE.
@@ -1250,8 +1257,8 @@ contains
              fld%tile(ith)%whole%xstart    = fld%whole%xstart
              fld%tile(ith)%internal%xstart = ival
           else
-             fld%tile(ith)%whole%xstart    = ival - 1
[…]

### 2015-02-11 api_v0.2/omp_tiling_mod.f90

@@ -1,315 +1,0 @@
-module omp_tiling_mod
-  use kind_params_mod
-  use region_mod
-  private
-
-  ! For coarse-grained OpenMP tiling
-  TYPE :: tile_type
-     type(region_type) :: internal
-     type(region_type) :: whole
-     !real(wp), dimension(:,:), allocatable :: data
-  END TYPE tile_type
-
-  INTEGER, SAVE                                    :: ntiles
-  INTEGER, SAVE                                    :: ntilex, ntiley
-  !> Whether or not to automatically compute dimensions of tiling grid
-  LOGICAL, SAVE                                    :: auto_tile
-  ! Extent of this 1D array will be ntiles which will default to
-  ! the no. of threads
-  TYPE(tile_type), ALLOCATABLE, SAVE, DIMENSION(:) :: tile
-  INTEGER, SAVE                                    :: max_tile_width
-  INTEGER, SAVE                                    :: max_tile_height
-
-  public openmp_grid_init
-  public tile_type
-
-contains
-
-  !================================================
-
-  SUBROUTINE openmp_grid_init(xlen, ylen, nx_arg, ny_arg)
-    use omp_lib, only: omp_get_max_threads
-    implicit none
-    !> Dimensions of the model mesh
-    integer, intent(in) :: xlen, ylen
-    !> Optional specification of the dimensions of the tiling grid
-    integer, intent(in), optional :: nx_arg, ny_arg
-    integer :: nx, ny
-    INTEGER :: idx, idy, ival, jval ! For tile extent calculation
-    integer :: internal_width, internal_height
-    INTEGER :: ierr, nwidth
-    INTEGER :: ji,jj, ith
-    INTEGER :: nthreads       ! No. of OpenMP threads being used
-    INTEGER :: jover, junder, idytmp, jover_per_row, junder_per_row
-    INTEGER :: iover, iunder, idxtmp, iover_per_col, iunder_per_col
-    LOGICAL :: nested_par     ! Whether OpenMP supports nested parallelism
-    ! For doing stats on tile sizes
-    INTEGER :: nvects, nvects_sum, nvects_min, nvects_max 
-    LOGICAL, PARAMETER :: print_tiles = .TRUE.
-
-    ! Set-up regular grid of tiles
-    auto_tile = .TRUE.
-
-    ! Dimensions of the grid of tiles. 
-    if(get_grid_dims(nx,ny) )then
[…]

### 2015-02-11 nemolite2d_psykal_omp/time_step_mod.f90

@@ -12,7 +12,6 @@ contains
                               hu, hv, ht, ua, va, un, vn)
     use kind_params_mod
     use timing_mod
-    use omp_tiling_mod
     use field_mod
     use grid_mod
     use model_mod,       only: rdt, cbfr, visc
@@ -590,7 +589,6 @@ contains
                                     hu, hv, ht, ua, va, un, vn)
     use kind_params_mod
     use timing_mod
-    use omp_tiling_mod
     use field_mod
     use grid_mod
     use model_mod,       only: rdt, cbfr, visc

### 2015-02-11 api_v0.2/field_mod.f90
Remove old, unused code from field module and add some comments

@@ -19,11 +19,18 @@ module field_mod
 
   !> A field is sub-divided into tiles for coarse-grained OpenMP
   type :: tile_type
+     !> Tile region to use when only a field's 'internal'
+     !! points are required
      type(region_type) :: internal
+     !> Tile region to use when all of a field's points are required
      type(region_type) :: whole
+     ! Could potentially physically divide up an array into 
+     ! distinct tiles and store the data for each separately...
      !real(wp), dimension(:,:), allocatable :: data
   end type tile_type
 
+  !> The base field type. Intended to represent a global field
+  !! such as the x-component of velocity.
   type, public :: field_type
      !> Which mesh points the field is defined upon
      integer :: defined_on
@@ -46,6 +53,7 @@ module field_mod
      real(wp) :: data
   end TYPE scalar_field
 
+  !> A real, 2D field.
   type, public, extends(field_type) :: r2d_field
      integer :: ntiles
      !> The dimensions of the tiles into which the field
@@ -81,11 +89,6 @@ module field_mod
      module procedure r2d_field_constructor
   end interface r2d_field
 
-  ! User-defined constructor for tiled_r2d_field type objects
-!  interface tiled_r2d_field
-!     module procedure tiled_r2d_field_constructor
-!  end interface tiled_r2d_field
-
   !> Interface for the field checksum operation. Overloaded to take either
   !! a field object or a 2D, real(wp) array.
   interface field_checksum
@@ -93,8 +96,8 @@ module field_mod
   end interface field_checksum
 
   !> Info on the tile sizes
-  INTEGER, SAVE                                    :: max_tile_width
-  INTEGER, SAVE                                    :: max_tile_height
+  INTEGER, SAVE :: max_tile_width
+  INTEGER, SAVE :: max_tile_height
 
   public increment_field
   public copy_field
@@ -138,23 +141,6 @@ module field_mod
   logical, public, parameter :: TILED_FIELDS = .TRUE.
 
 contains
-
-  !===================================================
-
-!  function field_constructor(grid_ptr,    &
-!                             grid_points) result(self)
-!    implicit none
-!    ! Arguments
-!    !> Pointer to the grid on which this field lives
-!    type(grid_type),       intent(in), pointer :: grid_ptr
-!    !> Which grid-point type the field is defined on
-!    integer,               intent(in)          :: grid_points
-!    ! Local declarations
-!    type(field_type) :: self
-!
[…]

### 2015-02-11 api_v0.2/field_mod.f90
Add support for case when tiling switched off and parallelise the field_copy()

@@ -882,8 +882,17 @@ contains
     IMPLICIT none
     type(r2d_field), intent(in)    :: field_in
     type(r2d_field), intent(inout) :: field_out
-        
-    field_out%data(:,:) = field_in%data(:,:)
+    integer :: it, ji, jj
+
+!$OMP DO SCHEDULE(RUNTIME)
+    do it = 1, field_out%ntiles, 1
+       do jj= field_out%tile(it)%whole%ystart, field_out%tile(it)%whole%ystop
+          do ji = field_out%tile(it)%whole%xstart, field_out%tile(it)%whole%xstop
+             field_out%data(ji,jj) = field_in%data(ji,jj)
+          end do
+       end do
+    end do
+!$OMP END DO
         
   end subroutine copy_2dfield
 
@@ -1066,7 +1075,17 @@ contains
     integer :: ntilex, ntiley
 
     if(.not. TILED_FIELDS)then
-       fld%ntiles = 0
+       fld%ntiles = 1
+       allocate(fld%tile(fld%ntiles), Stat=ierr)
+       if(ierr /= 0 )then
+          call gocean_stop('Harness: ERROR: failed to allocate tiling structures')
+       end if
+
+       ! We only have one tile and so we can simply copy in the
+       ! regions already defined for the field as a whole
+       fld%tile(1)%whole    = fld%whole
+       fld%tile(1)%internal = fld%internal
+
        return
     end if
 

### 2015-02-11 nemolite2d_psykal_omp/time_step_mod.f90
Parallelise the field copies in the Next section of the time-step

@@ -1091,14 +1091,28 @@ end do
 !    call copy_field(ua, un)
 !    call copy_field(va, vn)
 !    call copy_field(ssha, sshn_t)
+
 !$OMP DO SCHEDULE(RUNTIME)
-     do jj = 1, N+1, 1
-       do ji = 1, M+1, 1
-          un%data(ji,jj) = ua%data(ji,jj)
-          vn%data(ji,jj) = va%data(ji,jj)
-          sshn_t%data(ji,jj) = ssha%data(ji,jj)
+    do it = 1, un%ntiles, 1 ! Each field has the same no. of tiles
+
+       do jj= un%tile(it)%whole%ystart, un%tile(it)%whole%ystop, 1
+          do ji = un%tile(it)%whole%xstart, un%tile(it)%whole%xstop, 1
+             un%data(ji,jj) = ua%data(ji,jj)
+          end do
        end do
-    end do
+
+       do jj= vn%tile(it)%whole%ystart, vn%tile(it)%whole%ystop, 1
+          do ji = vn%tile(it)%whole%xstart, vn%tile(it)%whole%xstop, 1
+             vn%data(ji,jj) = va%data(ji,jj)
+          end do
+       end do
+
+       do jj= sshn_t%tile(it)%whole%ystart, sshn_t%tile(it)%whole%ystop, 1
+          do ji = sshn_t%tile(it)%whole%xstart, sshn_t%tile(it)%whole%xstop, 1
+             sshn_t%data(ji,jj) = ssha%data(ji,jj)
+          end do
+       end do
+    end do ! End loop over tiles
 ! We have to block here since sshn_t is used in the following loop.
 ! We could avoid this by altering the following two loop nests to read from
 ! ssha%data instead of sshn_t%data.

### 2015-02-11 nemolite2d_psykal_omp/time_step_mod.f90
Put back in-lining of Continuity and remove tiling from u-Flather section as incorrect.

@@ -670,18 +670,18 @@ contains
        do jj= ssha%tile(it)%internal%ystart, ssha%tile(it)%internal%ystop, 1
           do ji = ssha%tile(it)%internal%xstart, ssha%tile(it)%internal%xstop, 1
 
-        call continuity_code(ji, jj,                             &
-                             ssha%data, sshn_t%data,             &
-                             sshn_u%data, sshn_v%data,           &
-                             hu%data, hv%data, un%data, vn%data, &
-                             rdt, sshn_t%grid%area_t)
-!!$         rtmp1 = (sshn_u%data(ji  ,jj ) + hu%data(ji  ,jj  ))*un%data(ji  ,jj)
-!!$         rtmp2 = (sshn_u%data(ji-1,jj ) + hu%data(ji-1,jj  ))*un%data(ji-1,jj)
-!!$         rtmp3 = (sshn_v%data(ji ,jj ) + hv%data(ji  ,jj  ))*vn%data(ji ,jj)
-!!$         rtmp4 = (sshn_v%data(ji ,jj-1) + hv%data(ji  ,jj-1))*vn%data(ji,jj-1)
-!!$
-!!$         ssha%data(ji,jj) = sshn_t%data(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
-!!$                       rdt / sshn_t%grid%area_t(ji,jj)
+!!$        call continuity_code(ji, jj,                             &
+!!$                             ssha%data, sshn_t%data,             &
+!!$                             sshn_u%data, sshn_v%data,           &
+!!$                             hu%data, hv%data, un%data, vn%data, &
+!!$                             rdt, sshn_t%grid%area_t)
+         rtmp1 = (sshn_u%data(ji  ,jj ) + hu%data(ji  ,jj  ))*un%data(ji  ,jj)
+         rtmp2 = (sshn_u%data(ji-1,jj ) + hu%data(ji-1,jj  ))*un%data(ji-1,jj)
+         rtmp3 = (sshn_v%data(ji ,jj ) + hv%data(ji  ,jj  ))*vn%data(ji ,jj)
+         rtmp4 = (sshn_v%data(ji ,jj-1) + hv%data(ji  ,jj-1))*vn%data(ji,jj-1)
+
+         ssha%data(ji,jj) = sshn_t%data(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+                       rdt / sshn_t%grid%area_t(ji,jj)
         end do
       end do
[…]

### 2015-02-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
The beginnings of a paper on nemolite2d results

--- 2014-11-21 shallow_perf.tex
@@
-%% shallow_perf.tex
+%% nemolite2d_perf.tex
 %%
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)}
+\title{Something about Performance, 2D Component of NEMO. (TBC)}
 %
@@
 
-\author{A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
+\author{H.~Lieu and J.~Holt, A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
+\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}
 \thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
@@
 
-\subsection{The `shallow' Program}
-For this work we have used a benchmarking program that solves the
-shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
-This software was originally written in 1984 by Paul Swarztrauber of
-the National Center for Atmospheric Research, US.  However, in common
-with many scientific codes, it has subsequently undergone some sort of
-evolutionary development with subsequent people making various changes
-and optimising it for previous generations of hardware.  In describing
-our work, we term the version of the Shallow program obtained at the
-beginning of this project the `original' version.
-
-Shallow is a very good test case for our purposes since the original
-version is short (some 600 lines) and contained within a single source
-file. This makes it relatively straightforward for a compiler to
-optimise. Its performance is thus quite a demanding target for our
-modified versions of the code to reproduce.
+\subsection{The `nemolite2d' Program}
+For this work we have used a  program
 
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of shallow, we made the kernels
+optimisation) of the \psykal version of nemolite2d, we made the kernels
 as fine-grained as possible. In this case, this resulted in eight
@@
 \centering
-\includegraphics[width=85mm]{psykal_shallow}
+%\includegraphics[width=85mm]{psykal_shallow}
 \caption{A schematic of the vanilla \psykal version of the shallow code.}
@@
 As with any full oceanographic model, boundary conditions must be
-applied at the edges of the model domain. In the case of Shallow we
-have periodic boundary conditions in both the $x$ and $y$ dimensions.
-These are simply implemented by having additional (`halo')
-rows/columns at the edges of the domain and ensuring the data in them
-is up-to-date before it is read. These rows/columns are updated by
-copying data from the corresponding row/column on the opposite side of
-the domain. In the current work, the lines of code to do these copies
-is manually inserted in the PSy-layer routine as required.
+applied at the edges of the model domain.
 
@@
 In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
-used with each compiler. These flags are particularly important for
-the \psykal versions of the code. The performance of the original
-version of the code is much less sensitive to the compiler
-options. This is because it consists of a single source file
-containing effectively a single routine (excluding those for I/O).
+used with each compiler. {\bf Explanation of the compiler flags?}
+%These flags are particularly important for
+%the \psykal versions of the code. The performance of the original
+%version of the code is much less sensitive to the compiler
+%options. This is because it consists of a single source file
+%containing effectively a single routine (excluding those for I/O).
 
@@
 Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
-Intel     & -O3 -fast -fno-inline-factor    \\
+Intel     & -O3 -fast -fno-inline-factor -xHost   \\
 Cray      & -O3 -O ipa5 -h wp               \\
@@
 version after it has been re-structured following the \psykal
-approach. These two versions of the code effectively provide upper and
-lower bounds, respectively, on the performance we expect to achieve.
+approach.
+
+{\bf Performance analysis as motivation for the code
+  tranformations. More detail on each transformation.}
 
@@
 
-\item Pass explicit array bounds to middle layer (rather than using
-  assumed-size array arguments)
-
-\item Move kernel routines into same module as middle layer
-
-\item Loop fusion
-
 \item In-lined field-copy operation
-
-\item Fused field-copy operation
 
@@
 
-\item Fully fuse first loop (only the outer loop of this loop nest was
-  fused in the first pass)
+\item Plus others\ldots
 
-\item Make the array/loop bounds explicitly constant for the
-      duration of the time-stepping loop.
 \end{enumerate}
@@
 In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
-original version of the shallow code for the range of compilers and
-hardware considered here. This summary demonstrates the effect of the
-larger last-level cache of the Intel Ivy Bridge system compared to
-that of the Haswell system; note the significant drop in performance
-in going from a problem size of $256^{2}$ to a problem size of
-$512^{2}$ for the black and green bars. The performance of the Ivy
-Bridge system with the Cray or Intel compiler (red and purple bars)
-only drops to this level when the domain size is increased to
-$1024^{2}$. At this point, the working set no longer fits within cache
-and performance is dominated by the bandwidth to main memory, making
-it relatively insensitive to the choice of compiler.
+original version of the nemolite2d code for the range of compilers and
+hardware considered here.
 
@@
 \centering
-\includegraphics[width=2.8in]{orig_summary}
-\caption{Summary of the performance of the original shallow code on 
+%\includegraphics[width=2.8in]{orig_summary}
+\caption{Summary of the performance of the original version of the 
+nemolite2d code on 
 the range of compilers and hardware under consideration.}
@@
 
-For this compiler-friendly form of the code, there is little
-performance difference between the executables produced by the Gnu and
-Intel compilers. However, the executable produced by the Cray compiler
-generally performs significantly better. {\bf Why is this; can we work
-out what it's doing that the others aren't?}
-
-Note that the results for the Gnu compiler on Ivy Bridge show the same
-drop-off (as problem size is increased) as those on the Haswell
-system. This may be because Archer's login nodes seem to be running on
-older hardware (E5-2650 with 20MB LLC) during the Archer upgrade. {\bf
-  Need to re-run these cases once Archer is back up.}
-
-Moving now the to the \psykal version of shallow,
+Moving now the to the \psykal version of nemolite2d,
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
@@
 (best) \psykal versions of shallow for each compiler/CPU combination.
-It is clear from this plot that the difference in the performance of
-the Gnu- and Intel-compiled excutables on the Haswell system is due to
-the Gnu compiler being unable to recover the performance achieved by
-the original code. In contrast there is no such problem for the
-Gnu-compiled executable on the Ivy Bridge system. We attribute this to
-the more recent version of the compiler on the latter system (4.9.1 as
-opposed to 4.8.2).
 
@@
 \centering
-\includegraphics[width=7.5cm]{best_psykal_summary}
+%\includegraphics[width=7.5cm]{best_psykal_summary}
 \caption{Summary of the best performance achieved by any \psykal 
@@
 \centering
-\includegraphics[width=2.8in]{slowdown_summary}
+%\includegraphics[width=2.8in]{slowdown_summary}
 \caption{Comparison of the performance of the best \psykal
@@
 \end{figure}
-
-In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
-\psykal version of shallow is never more than 2\% slower than the
-original and in some cases is faster.  This demonstrates that we can
-reliably recover the performance of the original version of the code,
-despite the significant restructuring required by the \psykal
-approach.
 
@@
 Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
-Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
-Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
-Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
-In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
-Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
-In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
-Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
-Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
 \hline
@@
 
-With the Gnu compiler (first and third clusters in
-Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
-shallow only achieves ~5-6\% of the performance of the
-original. Simply copying the code for each of the kernel subroutines
-into the module from which they are called has the most dramatic
-effect on the performance of the compiled code; it now achieves almost
-80\% of the performance of the original. The next most significant
-transformation is that of (manually) in-lining the kernel code itself
-into the bodies of the loops from which they are called. This gets the
-\psykal version to within 8\% of the performance of the original.  The
-dramatic performance improvement obtained from the first
-transformation demonstrates that the Gnu compiler is unable to
-optimise over separate files. Therefore, the fact that the
-time-stepping loop in the \psykal version still retains a call to an
-I/O routine and a call to the middle-layer, both of which are in
-different files, is likely to account for this remaining performance
-penalty of 8\%.
-
-With the Intel compiler, the key transformations to recover
-performance are different (second and fourth clusters in
-Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
-shows that even the vanilla \psykal version of shallow achieves some
-80\% of the performance of the original version. Only two code
-transformations were required to increase this to 100\%. The first of
-these was the fusion of the computational loops. However, we found
-that performance suffered (purple bar) when this was extended to
-include the loop required for field copies at the end of a time-step
-(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
-slow-down revealed that the compiler was unable to SIMD vectorise the
-newly-fused loop after it had in-lined the body of the kernel called
-from within it. Once this in-lining was done by hand (dark-blue bar),
-the compiler was able to vectorise the loop and the performance of the
-original version of shallow was recovered.
-
 \begin{figure}[!t]
@@
 \includegraphics[width=95mm]{opt_stages_256}
-\caption{The performance of the \psykal version of shallow for the
+\caption{The performance of the \psykal version of nemolite2d for the
   $256^{2}$ domain at each stage of optimisation. Results are given as
@@
 \end{figure}
-
-\begin{figure}
-\begin{verbatim}
-! Time smoothing
-DO J=1,N+1
-  DO I=1,M+1
-    CALL time_smooth_code(i,j,ufld,unew,uold)
-!    uold(i,j) = ufld(i,j) + alpha* &
-!     (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
-  END DO
-END DO
-
-! Copy new fields into current fields
-DO J=1,N+1
-  DO I=1,M+1
-    Ufld(I,J) = UNEW(I,J)
-  END DO
-END DO
-\end{verbatim}
-\caption{Example of the coding structure that required manual
-  in-lining of the kernel body (as indicated by commented-out lines)
-  to retrieve the performance of the original version of shallow with
-  the Intel compiler.}
-\label{FIG_time_smooth_code}
-\end{figure}
-
-The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
-evolution of the performance of the \psykal version of shallow with
-the Cray compiler. Again, this has significant differences from the
-behaviour seen with the Gnu and Intel compilers. As with the Intel
-compiler, the performance of the Vanilla \psykal version is fairly
-good at 74\% of that of the original. In contrast to the other
-compilers consisdered so far, the first significant transformation is
-simply to specify the bounds of the arrays being used within the
-middle layer using module variables (as opposed to specifying them as
-assumed size). This tells the compiler that all of the arrays used in
-the middle layer have the same extent and also that all of the
-computational loops are over almost all elements of these arrays. This
-simple change gives a \psykal version that achieves 88\% of the
-performance of the original.
-
-Subsequent transformations actually harm performance until the
-field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
-with the preceeding loop. The resulting \psykal version now achieves
-94\% of the original performance. Two further steps are required to
-match the latter. First, the first loop nest has to be fully fused
-(both inner and outer loops fused) and second, the array/loop bounds
-are passed as an argument to the middle layer. This is done in order
-to specify to the compiler that these bounds remain constant for the
-duration of the time-stepping loop.
 
@@
 % argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,shallow_perf}
+\bibliography{IEEEabrv,nemolite2d_perf}
 %

### 2015-02-11 docs/papers/nemolite2d_performance/bargraph.pl

@@ -1,0 +1,1 @@
+../shallow_performance/bargraph.pl

### 2015-02-12 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add data and plot of best psykal (serial) results
Add performance profile table

@@
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal version of shallow retains the Input/Output
+we ensure that any \psykal version of nemolite2d retains the Input/Output
 capability of the original. This aids in limiting the optimisations
@@
 the PSy layer) and a call to the I/O system
-(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
+(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single PSy-layer
 routine then consists of applying each of the kernels to all of the
@@
 doubly-nested loop around each kernel call, as indicated by the
-pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.
+pseudo-code in Figure~\ref{FIG_psykal_nemolite2d_structure}.
 
@@
 \centering
-%\includegraphics[width=85mm]{psykal_shallow}
-\caption{A schematic of the vanilla \psykal version of the shallow code.}
-\label{FIG_psykal_shallow_structure}
+%\includegraphics[width=85mm]{psykal_nemolite2d}
+\caption{A schematic of the vanilla \psykal version of the nemolite2d code.}
+\label{FIG_psykal_nemolite2d_structure}
 \end{figure}
@@
 Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
-Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
-IBM A2 (BGQ)     &       &             &         & 14.1     \\
+Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &          \\
+IBM A2 (BGQ)     &       &             &         &   --     \\
 IBM Power 8      &       &             &         &          \\
@@
 
+\begin{table}[!t]
+\caption{The performance profile of the original and \psykal 
+versions of nemolite2d ($128^{2}$ domain and Intel compiler).}
+\label{TABLE_profile}
+\centering
+\begin{tabular}{l|c|c|c|c}
+\hline
+Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{\psykal} \\
+        & Time (s) & \%  & Time (s) & \% \\
+\hline
+Continuity  & 0.096 & 2.7  & 0.33  & 6.6 \\
+Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
+BCs         & 0.16  & 4.6  & 0.29  & 5.7 \\
+Time-update & 0.29  & 8.1  & 0.41  & 8.1 \\
+\hline
+\end{tabular}
+\end{table}
+
 Beginning with the vanilla \psykal version, we then apply a series of
@@
 
-\item In-lined field-copy operation
+\item Addition of 'safe-loop' directives
 
-\item Manually in-line kernel bodies into middle-layer code
+\item In-line Momentum kernel bodies into middle-layer code
 
-\item Plus others\ldots
+\item Force SIMD vectorisation of the Momentum kernels using directives
+
+\item Work around limitations related to derived types
+
+\item In-line the Continuity kernel
+
+\item In-line remaining kernels (BCs and time-update)
+
+\item In-line field-copy operations
 
@@
 
-In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
-original version of the nemolite2d code for the range of compilers and
-hardware considered here.
+In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
+of the original version of the nemolite2d code for the range of
+compilers and hardware considered here.
 
@@
 \centering
-%\includegraphics[width=2.8in]{orig_summary}
+\includegraphics[width=2.8in]{orig_summary}
 \caption{Summary of the performance of the original version of the 
@@
 percentage difference between the performance of the original and the
-(best) \psykal versions of shallow for each compiler/CPU combination.
+(best) \psykal versions of nemolite2d for each compiler/CPU combination.
 
@@
 \centering
-%\includegraphics[width=7.5cm]{best_psykal_summary}
+\includegraphics[width=7.5cm]{best_psykal_summary}
 \caption{Summary of the best performance achieved by any \psykal 
-version of shallow with each of the compilers and CPUs under 
+version of nemolite2d with each of the compilers and CPUs under 
 consideration.}
@@
 \centering
-%\includegraphics[width=2.8in]{slowdown_summary}
+\includegraphics[width=2.8in]{slowdown_summary}
 \caption{Comparison of the performance of the best \psykal
@@
 Having show that, in general, we can recover and occasionally improve
-upon the performance of the original version of shallow, the next
+upon the performance of the original version of nemolite2d, the next
 logical step is to examine the necessary code transformations.  We do
@@
 \centering
-\includegraphics[width=95mm]{opt_stages_256}
+\includegraphics[width=90mm]{opt_stages_256}
 \caption{The performance of the \psykal version of nemolite2d for the

### 2015-02-12 nemolite2d_perf.tex
Add data on performance of orig. version of nemolite2d
Add data and plot of slowdown summary
WIP on safe-loop directives description

--- 2015-02-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
@@
 
+{\bf Description of the structure of the program at the Algorithm
+  level so that it links in to Profile data in results section.}
+
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal version of shallow retains the Input/Output
+we ensure that any \psykal version of nemolite2d retains the Input/Output
 capability of the original. This aids in limiting the optimisations
@@
 version after it has been re-structured following the \psykal
-approach.
+approach. In addition to this benchmarking, we profile these versions
+of the code at the algorithm level (using a high-level timing
+API). The resulting profiles are given in
+Table~\ref{TABLE_profile}. The Momentum section dominates the profile
+of both versions of the code, accounting for around 80\% of the
+wall-clock time spent doing time-stepping. In terms of performance
+loss when moving from the original to the \psykal version of the code
+however, the Momentum section is not too bad -- it slows down by some
+33\%. This is in contrast to the Continuity and BCs sections which
+slow down by factors of approximately three and two, respectively.
 
-{\bf Performance analysis as motivation for the code
-  tranformations. More detail on each transformation.}
+\begin{table}[!t]
+\caption{The performance profile of the original and \psykal versions
+  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
+  Intel compiler).}
+\label{TABLE_profile}
+\centering
+\begin{tabular}{l|c|c|c|c}
+\hline
+Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{Vanilla \psykal} \\
+        & Time (s) & \%  & Time (s) & \% \\
+\hline
+Continuity  & 0.10  & 2.7  & 0.33  & 6.6  \\
+Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
+BCs         & 0.16  & 4.6  & 0.29  & 5.7  \\
+Time-update & 0.29  & 8.1  & 0.41  & 8.1  \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 code transformations while obeying the \psykal separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
-the kernel and algorithm layers unchanged. The aim of these
-optimisations is to recover, as much as is possible, the structure and
-thus performance of the original version of the code. The
-transformations we have performed are as follows:
-\begin{enumerate}
+{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
+leaves the kernel and algorithm layers unchanged. The aim of these
+optimisations is to recover, as much as is possible, the performance
+of the original version of the code. The transformations we have
+performed and the reasons for them are described in the following sections.
 
-\item In-lined field-copy operation
+\subsection{Addition of `safe-loop' directives}
 
-\item Manually in-line kernel bodies into middle-layer code
+Much of the optimisations we have performed have been informed by the
+diagnostic output produced by either the Cray or Intel compilers. Many
+of the nemolite2d kernels contain conditional statements. These
+statements are there to check whether {\it e.g.} the current grid
+point is wet or neighbours a boundary point. A compiler is better able
+to optimise such a loop if it can be sure that all array accesses
+within the body of the loop are safe for every trip, irrespective of
+the conditional statements. In its diagnostic output the Cray compiler
+notes this with messages of the form:
+\begin{verbatim}
+ftn-6375 ftn: VECTOR File = nemolite2d.f90,
+ Line = 448 
+A loop starting at line 448 would benefit 
+from "!dir$ safe_address".
+\end{verbatim}
+In order that we could safely add such directives we altered the field
+module to allocate all arrays holding field data with extents greater
+than strictly required such that any array accesses within the kernels
+would remain within bounds.
 
-\item Plus others\ldots
+\subsection{In-line Momentum kernel bodies into middle-layer code}
 
-\end{enumerate}
+\subsection{Force SIMD vectorisation of the Momentum kernels using directives}
+
+\subsection{Work around limitations related to derived types}
+
+\subsection{In-line the Continuity kernel}
+
+\subsection{In-line remaining kernels (BCs and time-update)}
+
+\subsection{In-line field-copy operations}
 

### 2015-02-12 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
WIP on describing the in-lining of Mom. kernels

@@
 
+{\bf Description of the structure of the program at the Algorithm
+  level so that it links in to Profile data in results section.}
+
 Since any real oceanographic computational model must output results,
@@
 version after it has been re-structured following the \psykal
-approach.
-
-{\bf Performance analysis as motivation for the code
-  tranformations. More detail on each transformation.}
+approach. In addition to this benchmarking, we profile these versions
+of the code at the algorithm level (using a high-level timing
+API). The resulting profiles are given in
+Table~\ref{TABLE_profile}. The Momentum section dominates the profile
+of both versions of the code, accounting for around 80\% of the
+wall-clock time spent doing time-stepping. In terms of performance
+loss when moving from the original to the \psykal version of the code
+however, the Momentum section is not too bad -- it slows down by some
+33\%. This is in contrast to the Continuity and BCs sections which
+slow down by factors of approximately three and two, respectively.
 
 \begin{table}[!t]
-\caption{The performance profile of the original and \psykal 
-versions of nemolite2d ($128^{2}$ domain and Intel compiler).}
+\caption{The performance profile of the original and \psykal versions
+  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
+  Intel compiler).}
 \label{TABLE_profile}
@@
 \hline
-Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{\psykal} \\
+Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{Vanilla \psykal} \\
         & Time (s) & \%  & Time (s) & \% \\
 \hline
-Continuity  & 0.096 & 2.7  & 0.33  & 6.6 \\
+Continuity  & 0.10  & 2.7  & 0.33  & 6.6  \\
 Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
@@
 code transformations while obeying the \psykal separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
-the kernel and algorithm layers unchanged. The aim of these
-optimisations is to recover, as much as is possible, the structure and
-thus performance of the original version of the code. The
-transformations we have performed are as follows:
-\begin{enumerate}
+{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
+leaves the kernel and algorithm layers unchanged. The aim of these
+optimisations is to recover, as much as is possible, the performance
+of the original version of the code. The transformations we have
+performed and the reasons for them are described in the following sections.
 
-\item Addition of 'safe-loop' directives
+\subsection{Addition of `safe-address' directives}
 
-\item In-line Momentum kernel bodies into middle-layer code
+Much of the optimisations we have performed have been informed by the
+diagnostic output produced by either the Cray or Intel compilers. Many
+of the nemolite2d kernels contain conditional statements. These
+statements are there to check whether {\it e.g.} the current grid
+point is wet or neighbours a boundary point. A compiler is better able
+to optimise such a loop if it can be sure that all array accesses
+within the body of the loop are safe for every trip, irrespective of
+the conditional statements. In its diagnostic output the Cray compiler
+notes this with messages of the form:
+\begin{verbatim}
+ftn-6375 ftn: VECTOR File = nemolite2d.f90,
+ Line = 448 
+A loop starting at line 448 would benefit 
+from "!dir$ safe_address".
+\end{verbatim}
+In order that we could safely add such directives we altered the field
+module to allocate all field-data arrays with extents greater than
+strictly required.  We were then able to safely add the safe\_address
+before all of the loops where the Cray compiler indicated it might be
+useful (the Momentum loops and some of the BC loops).
 
-\item Force SIMD vectorisation of the Momentum kernels using directives
+\subsection{In-line Momentum kernel bodies into middle-layer code}
 
-\item Work around limitations related to derived types
+From our previous work~\cite{shallow_psykal} on a similar code we know
+that kernel in-lining is critical to obtaining performance with the
+Gnu and Intel compilers. For the Gnu compiler, this is because it
+cannot perform in-lining when routines are in separate source files
+{\bf do we have evidence for this?}. Although the Intel compiler does
+not have this limitation, we have found (both here and in our previous
+work~\cite{shallow_psykal}) that the number of optimisations it
+performs is reduced if it first has to in-line a routine.
 
-\item In-line the Continuity kernel
+The profiling data in Table~\ref{TABLE_profile} shows that it is the
+Momentum section that accounts for the bulk of the model run-time.  We
+therefore chose to in-line the two kernels (one for each of the $x$
+and $y$ components of momentum) called within this section.
 
-\item In-line remaining kernels (BCs and time-update)
+\subsection{Force SIMD vectorisation of the Momentum kernels using directives}
 
-\item In-line field-copy operations
+\subsection{Work around limitations related to derived types}
 
-\end{enumerate}
+\subsection{In-line the Continuity kernel}
+
+\subsection{In-line remaining kernels (BCs and time-update)}
+
+\subsection{In-line field-copy operations}
 

### 2015-02-12 nemolite2d_perf.tex
Text describing reasoning behind in-lining and SIMD directives

@@
 
-\subsection{Addition of `safe-loop' directives}
+\subsection{Addition of `safe-address' directives}
 
@@
 In order that we could safely add such directives we altered the field
-module to allocate all arrays holding field data with extents greater
-than strictly required such that any array accesses within the kernels
-would remain within bounds.
+module to allocate all field-data arrays with extents greater than
+strictly required.  We were then able to safely add the safe\_address
+before all of the loops where the Cray compiler indicated it might be
+useful (the Momentum loops and some of the BC loops).
 
@@
 
+From our previous work~\cite{shallow_psykal} on a similar code we know
+that kernel in-lining is critical to obtaining performance with both
+the Gnu and Intel compilers. For the Gnu compiler, this is because it
+cannot perform in-lining when routines are in separate source
+files. In our previous work we obtained an order-of-magnitude speed-up
+simply by moving subroutines into the module containing the middle
+layer (from which the kernels are called). A further performance
+improvement of roughly 30\% was obtained when the kernel code was
+manually inserted at the site of the subroutine call.
+
+Although the Intel compiler does do in-lining when routines are in
+separate source files, we have found (both here and in our previous
+work~\cite{shallow_psykal}) that the number of optimisations it
+performs is reduced if it first has to in-line a routine. For the
+Intel-compiled Shallow code, manually inserting kernel code at the
+site of the subroutine call increased performance by about 25\%.
+
+In fact, in-lining can have a significant effect on the Intel
+compiler's ability to vectorise a loop. Taking the loop that calls the
+kernel for the u-component of momentum as an example, before in-lining
+the compiler reports:
+\begin{verbatim}
+LOOP BEGIN at time_step_mod.f90(85,7) 
+  inlined into nemolite2d.f90(86,11)
+ remark #15335: loop was not vectorized: 
+  vectorization possible but seems 
+  inefficient. 
+ --- begin vector loop cost summary ---
+ scalar loop cost: 1307 
+ vector loop cost: 2391.000 
+ estimated potential speedup: 0.540 
+ lightweight vector operations: 1133 
+ medium-overhead vector operations: 1 
+ vectorized math library calls: 1 
+ --- end vector loop cost summary ---
+LOOP END
+\end{verbatim}
+After we have manually in-lined the kernel body, the compiler reports:
+\begin{verbatim}
+LOOP BEGIN at time_step_mod.f90(97,7) 
+  inlined into nemolite2d.f90(86,11)
+ LOOP WAS VECTORIZED
+ --- begin vector loop cost summary ---
+ scalar loop cost: 1253 
+ vector loop cost: 521.750 
+ estimated potential speedup: 2.350 
+ lightweight vector operations: 697 
+ medium-overhead vector operations: 1 
+ vectorized math library calls: 1 
+ --- end vector loop cost summary ---
+LOOP END
+\end{verbatim}
+Looking at the `estimated potential speedup' in the compiler outputs
+above, it is clear that the way in which the compiler vectorises the
+two versions must be very different. This conclusion is borne out by
+the fact that if one persuades the compiler to vectorise the first
+version (through the use of a directive) then the performance of the
+resulting binary is worse than that where the loop is left
+un-vectorised.
+
+The profiling data in Table~\ref{TABLE_profile} shows that it is the
+Momentum section that accounts for the bulk of the model run-time.  We
+therefore chose to in-line the two kernels (one for each of the $x$
+and $y$ components of momentum) called within this section. Given our
+findings with the shallow code, we have chosen to do full, manual
+inlining of the kernel code.
+
 \subsection{Force SIMD vectorisation of the Momentum kernels using directives}
 
+It turns out that the Cray-compiled binaries of both the original and
+\psykal versions of nemolite2d perform considerably less well than
+their Intel-compiled counterparts. Comparison of the diagnostic output
+from each of the compilers revealed that while the Intel compiler was
+happy to vectorise the Momentum loops, the Cray compiler was choosing
+not to:
+\begin{verbatim}
+   99.          do ji = 2, M-1, 1
+ A loop starting at line 99 was blocked 
+ with block size 256.
+
+ A loop starting at line 99 was not
+ vectorized because it contains 
+ conditional code which is more
+ efficient if executed in scalar mode.
+\end{verbatim}
+Inserting the Cray `vector always' directive persuaded the compiler
+to vectorise the loop:
+\begin{verbatim}
+  99.    !dir$ vector always
+  100.          do ji = 2, M-1, 1
+ A loop starting at line 100 was blocked
+ with block size 256.
+
+ A loop starting at line 100 requires an 
+ estimated 17 vector registers at 
+ line 151; 1 of these have been 
+ preemptively forced to memory.
+
+ A loop starting at line 100 was vectorized.
+\end{verbatim}
+which gave a significant performance improvement. This behaviour is in
+contrast to that obtained with the Intel compiler: its predictions
+about whether vectorising a loop would be beneficial were generally
+found to be reliable.
+
 \subsection{Work around limitations related to derived types}
+
+Further comparison of the diagnostic output from the Cray and Intel
+compilers revealed that the Cray compiler was vectorising the
+Continuity section while the Intel compiler reported that it was
+unable to do so due to dependencies. After some experimentation we
+found that this was due to the compiler's limited analysis of the way
+components of Fortran derived types were being used.
 

### 2015-02-12 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
WIP on describing Intels foibles with derived types

@@
 From our previous work~\cite{shallow_psykal} on a similar code we know
-that kernel in-lining is critical to obtaining performance with the
-Gnu and Intel compilers. For the Gnu compiler, this is because it
-cannot perform in-lining when routines are in separate source files
-{\bf do we have evidence for this?}. Although the Intel compiler does
-not have this limitation, we have found (both here and in our previous
+that kernel in-lining is critical to obtaining performance with both
+the Gnu and Intel compilers. For the Gnu compiler, this is because it
+cannot perform in-lining when routines are in separate source
+files. In our previous work we obtained an order-of-magnitude speed-up
+simply by moving subroutines into the module containing the middle
+layer (from which the kernels are called). A further performance
+improvement of roughly 30\% was obtained when the kernel code was
+manually inserted at the site of the subroutine call.
+
+Although the Intel compiler does do in-lining when routines are in
+separate source files, we have found (both here and in our previous
 work~\cite{shallow_psykal}) that the number of optimisations it
-performs is reduced if it first has to in-line a routine.
+performs is reduced if it first has to in-line a routine. For the
+Intel-compiled Shallow code, manually inserting kernel code at the
+site of the subroutine call increased performance by about 25\%.
+
+In fact, in-lining can have a significant effect on the Intel
+compiler's ability to vectorise a loop. Taking the loop that calls the
+kernel for the u-component of momentum as an example, before in-lining
+the compiler reports:
+\begin{verbatim}
+LOOP BEGIN at time_step_mod.f90(85,7) 
+  inlined into nemolite2d.f90(86,11)
+ remark #15335: loop was not vectorized: 
+  vectorization possible but seems 
+  inefficient. 
+ --- begin vector loop cost summary ---
+ scalar loop cost: 1307 
+ vector loop cost: 2391.000 
+ estimated potential speedup: 0.540 
+ lightweight vector operations: 1133 
+ medium-overhead vector operations: 1 
+ vectorized math library calls: 1 
+ --- end vector loop cost summary ---
+LOOP END
+\end{verbatim}
+After we have manually in-lined the kernel body, the compiler reports:
+\begin{verbatim}
+LOOP BEGIN at time_step_mod.f90(97,7) 
+  inlined into nemolite2d.f90(86,11)
+ LOOP WAS VECTORIZED
+ --- begin vector loop cost summary ---
+ scalar loop cost: 1253 
+ vector loop cost: 521.750 
+ estimated potential speedup: 2.350 
+ lightweight vector operations: 697 
+ medium-overhead vector operations: 1 
+ vectorized math library calls: 1 
+ --- end vector loop cost summary ---
+LOOP END
+\end{verbatim}
+Looking at the `estimated potential speedup' in the compiler outputs
+above, it is clear that the way in which the compiler vectorises the
+two versions must be very different. This conclusion is borne out by
+the fact that if one persuades the compiler to vectorise the first
+version (through the use of a directive) then the performance of the
+resulting binary is worse than that where the loop is left
+un-vectorised.
 
@@
 therefore chose to in-line the two kernels (one for each of the $x$
-and $y$ components of momentum) called within this section.
+and $y$ components of momentum) called within this section. Given our
+findings with the shallow code, we have chosen to do full, manual
+inlining of the kernel code.
 
@@
 
+It turns out that the Cray-compiled binaries of both the original and
+\psykal versions of nemolite2d perform considerably less well than
+their Intel-compiled counterparts. Comparison of the diagnostic output
+from each of the compilers revealed that while the Intel compiler was
+happy to vectorise the Momentum loops, the Cray compiler was choosing
+not to:
+\begin{verbatim}
+   99.          do ji = 2, M-1, 1
+ A loop starting at line 99 was blocked 
+ with block size 256.
+
+ A loop starting at line 99 was not
+ vectorized because it contains 
+ conditional code which is more
+ efficient if executed in scalar mode.
+\end{verbatim}
+Inserting the Cray `vector always' directive persuaded the compiler
+to vectorise the loop:
+\begin{verbatim}
+  99.    !dir$ vector always
+  100.          do ji = 2, M-1, 1
+ A loop starting at line 100 was blocked
+ with block size 256.
+
+ A loop starting at line 100 requires an 
+ estimated 17 vector registers at 
+ line 151; 1 of these have been 
+ preemptively forced to memory.
+
+ A loop starting at line 100 was vectorized.
+\end{verbatim}
+which gave a significant performance improvement. This behaviour is in
+contrast to that obtained with the Intel compiler: its predictions
+about whether vectorising a loop would be beneficial were generally
+found to be reliable.
+
 \subsection{Work around limitations related to derived types}
+
+Further comparison of the diagnostic output from the Cray and Intel
+compilers revealed that the Cray compiler was vectorising the
+Continuity section while the Intel compiler reported that it was
+unable to do so due to dependencies. After some experimentation we
+found that this was due to limitations in the compiler's analysis of
+the way components of Fortran derived types were being used. Each
+GOcean field object, in addition to the array holding the local
+section of the field, contains a pointer to a GOcean grid object. If a
+kernel requires grid-related quantities ({\it e.g.} the grid spacing)
+then these are obtained by passing it a reference to the appropriate
+array within the grid object. Although these grid-related quantities
+are read-only within a compute kernel, if they were obtained from the
+same object as that containing the array to which that kernel writes
+then the Intel compiler identified a dependency preventing
+vectorisation.  This limitation was simply removed by ensuring that
+all read-only quantities were accessed via field objects that were
+themselves read-only for the kernel at hand.
+\begin{verbatim}
+do jj = 2, N, 1
+ do ji = 2, M, 1
+
+   call continuity_code(ji, jj,      &
+                        ssha%data,   &
+                        sshn_t%data, &
+                        ...,         &
+                        sshn_t%grid%area_t)
+\end{verbatim}
+instead of
+\begin{verbatim}
+do jj = 2, N, 1
+ do ji = 2, M, 1
+
+   call continuity_code(ji, jj,      &
+                        ssha%data,   &
+                        sshn_t%data, &
+                        ...,         &
+                        ssha%grid%area_t)
+\end{verbatim}
 

### 2015-02-23 nemolite2d_perf.tex
More text describing opt. steps. Addition of potential column of profile data for optimised version

@@
 \centering
-\begin{tabular}{l|c|c|c|c}
+\begin{tabular}{l|c|c|c|c|c|c}
 \hline
-Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{Vanilla \psykal} \\
-        & Time (s) & \%  & Time (s) & \% \\
+Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
+\multicolumn{2}{c}{Optimised \psykal} \\
+        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
 \hline
-Continuity  & 0.10  & 2.7  & 0.33  & 6.6  \\
-Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
-BCs         & 0.16  & 4.6  & 0.29  & 5.7  \\
-Time-update & 0.29  & 8.1  & 0.41  & 8.1  \\
+Continuity  & 0.10  & 2.7  & 0.33  & 6.6  & 0.11 & \\
+Momentum    & 3.00  & 84.5 & 4.05  & 79.5 & 2.09 & \\
+BCs         & 0.16  & 4.6  & 0.29  & 5.7  & 0.27 & \\
+Time-update & 0.29  & 8.1  & 0.41  & 8.1  & 0.29 & \\
 \hline
@@
 \subsection{In-line Momentum kernel bodies into middle-layer code}
+
+The profiling data in Table~\ref{TABLE_profile} shows that it is the
+Momentum section that accounts for the bulk of the model run-time.  We
+therefore chose to attempt to optimise this section first. In-keeping
+with the \psykal approach, we are only permitted to optimise the middle
+(PSy) layer which, for this section comprises calls to two kernels,
+one for each of the $x$ and $y$ components of momentum. These kernels
+are relatively large; each comprises roughly 85 lines of Fortran
+(excluding variable declarations).
 
@@
 resulting binary is worse than that where the loop is left
-un-vectorised.
+un-vectorised. {\bf In principle we could investigate this by looking
+  at the assembler that the Intel compiler generates\ldots}
 
-The profiling data in Table~\ref{TABLE_profile} shows that it is the
-Momentum section that accounts for the bulk of the model run-time.  We
-therefore chose to in-line the two kernels (one for each of the $x$
-and $y$ components of momentum) called within this section. Given our
-findings with the shallow code, we have chosen to do full, manual
-inlining of the kernel code.
+For the best possible performance, we have chosen to do full, manual
+inlining for the two kernels making up the Momentum section.
 
@@
 
-Further comparison of the diagnostic output from the Cray and Intel
+Having optimised the Momentum section as much as permitted by the
+\psykal approach, we turn our attention to the three remaining
+sections of the code. The profile data in Table~\ref{TABLE_profile}
+shows that these regions are all comparable in terms of cost. What is
+striking however, is that the cost of the Continuity section increases
+by more than a factor of three in moving to the \psykal version of the
+code.
+
+Comparison of the diagnostic output from the Cray and Intel
 compilers revealed that the Cray compiler was vectorising the
@@
 unable to do so due to dependencies. After some experimentation we
-found that this was due to the compiler's limited analysis of the way
-components of Fortran derived types were being used.
+found that this was due to limitations in the compiler's analysis of
+the way components of Fortran derived types were being used. Each
+GOcean field object, in addition to the array holding the local
+section of the field, contains a pointer to a GOcean grid object. If a
+kernel requires grid-related quantities ({\it e.g.} the grid spacing)
+then these are obtained by passing it a reference to the appropriate
+array within the grid object. Although these grid-related quantities
+are read-only within a compute kernel, if they were obtained from the
+same object as that containing the array to which that kernel writes
+then the Intel compiler identified a dependency preventing
+vectorisation.  This limitation was simply removed by ensuring that
+all read-only quantities were accessed via field objects that were
+themselves read-only for the kernel at hand.  For instance, the call
+to the continuity kernel, which confused the Intel compiler, originally
+looked like this:
+\begin{verbatim}
+do jj = 2, N, 1
+ do ji = 2, M, 1
+
+   call continuity_code(ji, jj,      &
+                        ssha%data,   &
+                        sshn_t%data, &
+                        ...,         &
+                        ssha%grid%area_t)
+\end{verbatim}
+where {\it ssha} is the only field that is written to by the kernel.
+We remove any potential confusion by instead obtaining the
+grid-related (read-only) quantities from a field ({\it sshn\_t} in
+this case) that is only read by the kernel:
+\begin{verbatim}
+do jj = 2, N, 1
+ do ji = 2, M, 1
+
+   call continuity_code(ji, jj,      &
+                        ssha%data,   &
+                        sshn_t%data, &
+                        ...,         &
+                        sshn_t%grid%area_t)
+\end{verbatim}
 
@@
 
+As with the Momentum kernel, we know that obtaining optimal
+performance from both the Gnu and Intel compilers requires that a
+kernel be manually inlined at its call site. We do this for the
+Continuity kernel in this optimisation step.
+
 \subsection{In-line remaining kernels (BCs and time-update)}
+
+Having optimised the Continuity section we finally turn our attention
+to the Boundary Condition and Time-Update sections.
 

### 2015-02-24 nemolite2d_perf.tex
Correct profile timings of original version of code and finish text describing the optimisations

@@
 Table~\ref{TABLE_profile}. The Momentum section dominates the profile
-of both versions of the code, accounting for around 80\% of the
-wall-clock time spent doing time-stepping. In terms of performance
-loss when moving from the original to the \psykal version of the code
-however, the Momentum section is not too bad -- it slows down by some
-33\%. This is in contrast to the Continuity and BCs sections which
-slow down by factors of approximately three and two, respectively.
+of both versions of the code, accounting for around 70--80\% of the
+wall-clock time spent doing time-stepping. It is also the key section
+when we consider the performance loss when moving from the original to
+the \psykal version of the code; it slows down by a factor of
+two. Although less significant in terms of absolute time, the
+Continuity section is also dramatically slower in the \psykal version,
+this time by a factor of three. In contrast, the performance of the
+Time-update and Boundary Condition regions are not significantly
+affected by the move so the \psykal version.
 
@@
 Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
-\multicolumn{2}{c}{Optimised \psykal} \\
+\multicolumn{2}{c}{Final \psykal} \\
         & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
 \hline
-Continuity  & 0.10  & 2.7  & 0.33  & 6.6  & 0.11 & \\
-Momentum    & 3.00  & 84.5 & 4.05  & 79.5 & 2.09 & \\
-BCs         & 0.16  & 4.6  & 0.29  & 5.7  & 0.27 & \\
-Time-update & 0.29  & 8.1  & 0.41  & 8.1  & 0.29 & \\
+Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
+Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
+BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
+Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
 \hline
@@
 performance from both the Gnu and Intel compilers requires that a
-kernel be manually inlined at its call site. We do this for the
+kernel be manually in-lined at its call site. We do this for the
 Continuity kernel in this optimisation step.
@@
 Having optimised the Continuity section we finally turn our attention
-to the Boundary Condition and Time-Update sections.
+to the Boundary Condition and Time-update sections. The kernels in
+these sections are small and dominated by conditional statements.  We
+therefore limited our optimisation of them to manually in-lining each
+of the kernels into the PSy layer.
 
@@
 
-We shall see that these transformations do not always result in improved
-performance. Whether or not they do so depends both on the compiler
-and the problem size. We also emphasise that the aim of these
-optimisations is to recover, as far as is possible, the structure of
-the original version of the code. It may well be that transforming the
-code into some other structure would result in better performance on a
-particular architecture. However, exploring this optimisation space is
-beyond the scope of the present work.
+The Time-update section includes several array copies where fields for
+the current time-step become the fields at the previous time-step.
+Initially we implemented these copies as 'built-in' kernels (in the
+GOcean library) as they are specified in the algorithm layer. However,
+we obtained better performance by simply in-lining these array copies
+into the PSy layer.
+
+We shall see that the transformations we have just described do not
+always result in improved performance. Whether or not they do so
+depends both on the compiler used and the problem size. We also
+emphasise that the aim of these optimisations is to make the PSy layer
+as compiler-friendly as possible, following the lessons learned from
+our previous work with the Shallow code~\cite{shallow_psykal}. It may
+well be that transforming the code into some other structure would
+result in better performance on a particular architecture. However,
+exploring this optimisation space is beyond the scope of the present
+work.
 
 We explore the extent to which performance depends upon the problem
-size by using square domains of dimension 64, 128, 256, 512 and
-1024. This range allows us to investigate what happens when cache is
-exhausted as well as giving us some insight into the decisions that
-different compilers make when optimising the code.
+size by using square domains of dimension 64, 128, 256, 512 and 1024
+for the traditional, cache-based CPU systems. This range allows us to
+investigate what happens when cache is exhausted as well as giving us
+some insight into the decisions that different compilers make when
+optimising the code.
 

### 2015-02-24 nemolite2d_perf.tex
More text describing serial results

@@
 \section{Methodology(?)}
+\label{sec_methodology}
 
@@
 
+We first consider the performance of the code in serial and examine
+the effects of the transformations described in
+Section~\ref{sec_methodology}. Once we have arrived at an optimal form
+for the serial version of nemolite2d we then investigate its parallel
+performance on both CPU- and GPU-based systems.
+
+\subsection{Serial Performance}
+
 In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
 of the original version of the nemolite2d code for the range of
-compilers and hardware considered here.
+compilers and CPUs considered here. Unlike the Shallow code, the
+original version of nemolite2d has not been optimised. Although it is
+still a single source file it is, in common with NEMO itself,
+logically structured with separate subroutines performing different
+parts of the physics within each time step. This structuring and the
+heavy use of conditional statements favour the Intel compiler which
+significantly out-performs both the Gnu and Cray compilers. Only when
+the problem size falls out of cache does the performance gap begin to
+narrow. The reason for the performance deficit of the Cray-compiled
+binary comes down to SIMD vectorisation, an issue that we explore
+below.
 
@@
 nemolite2d code on 
-the range of compilers and hardware under consideration.}
+the range of compilers and CPU-based systems under consideration.}
 \label{FIG_orig_perf_summary}
@@
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest \psykal version for each of the compiler/CPU combinations. The
-only significant difference between this and the plot of the
-performance of the original version of the code in
-Figure~\ref{FIG_orig_perf_summary} is that the difference between the
-Gnu- and Intel-compiled executables is slightly more marked. This is
-clarified in Figure~\ref{FIG_slowdown_summary} which plots the
-percentage difference between the performance of the original and the
-(best) \psykal versions of nemolite2d for each compiler/CPU combination.
+fastest \psykal version for each of the compiler/CPU combinations.
+While the Intel compiler still produces the best-performing binary,
+the Cray-compiled binary is now a very close second. In fact, the
+performance of both the Gnu- and Cray-compiled \psykal versions is
+generally significantly greater than that of their respective original
+versions. We also note that best absolute performance (in terms of
+grid points processed per second) with any compiler is obtained with
+the $256^2$ domain.
 
@@
 
+Figure~\ref{FIG_slowdown_summary} plots the percentage difference
+between the performance of the original and the best \psykal versions
+of nemolite2d for each compiler/CPU combination. This shows that it is
+only the Intel-compiled binary running the $64^2$ domain that is
+slower with the \psykal version of the code (and then only by some
+3\%). For all other points in the space, the optimised \psykal version
+of the code performs better. The results for the Cray compiler are
+however somewhat sqewed by the fact that it did not SIMD vectorise key
+parts of the original version (see below).
+
 \begin{figure}[!t]
@@
 
-Having show that, in general, we can recover and occasionally improve
-upon the performance of the original version of nemolite2d, the next
-logical step is to examine the necessary code transformations.  We do
-this for the $256^{2}$ case since this fits within cache on both the
-Haswell and Ivy Bridge CPUs we are using here.
-Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
-after each transformation has been applied to the code. The same data
-is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
-of the \psykal version for a given compiler/CPU is given relative to
-the performance of the original with the same compiler/CPU.
+Having show that we can recover and often improve upon the performance
+of the original version of nemolite2d, the next logical step is to
+examine the necessary code transformations in detail.  We do this for
+the $256^{2}$ case since this fits within cache on the Ivy Bridge CPUs
+we are using here.  Table~\ref{TABLE_opt_breakdown} shows detailed
+timings for this case after each transformation has been applied to
+the code. The same data is visualised in
+Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
+version for a given compiler/CPU is given relative to the performance
+of the original with the same compiler/CPU.
 
@@
 
+\subsection{Parallel Performance}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}

### 2015-02-24 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
More text on serial perf. and data added to table summarising the optimisation steps

@@
 \centering
-\begin{tabular}{l|c|c|c|c}
+\begin{tabular}{l|c|c|c|c|c|c}
 \hline
-Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{Vanilla \psykal} \\
-        & Time (s) & \%  & Time (s) & \% \\
+Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
+\multicolumn{2}{c}{Final \psykal} \\
+        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
 \hline
-Continuity  & 0.10  & 2.7  & 0.33  & 6.6  \\
-Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
-BCs         & 0.16  & 4.6  & 0.29  & 5.7  \\
-Time-update & 0.29  & 8.1  & 0.41  & 8.1  \\
+Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
+Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
+BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
+Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
 \hline
@@
 all read-only quantities were accessed via field objects that were
-themselves read-only for the kernel at hand.
+themselves read-only for the kernel at hand.  For instance, the call
+to the continuity kernel, which confused the Intel compiler, originally
+looked like this:
+\begin{verbatim}
+do jj = 2, N, 1
+ do ji = 2, M, 1
+
+   call continuity_code(ji, jj,      &
+                        ssha%data,   &
+                        sshn_t%data, &
+                        ...,         &
+                        ssha%grid%area_t)
+\end{verbatim}
+where {\it ssha} is the only field that is written to by the kernel.
+We remove any potential confusion by instead obtaining the
+grid-related (read-only) quantities from a field ({\it sshn\_t} in
+this case) that is only read by the kernel:
 \begin{verbatim}
@@
 \end{verbatim}
-instead of
-\begin{verbatim}
-do jj = 2, N, 1
- do ji = 2, M, 1
-
-   call continuity_code(ji, jj,      &
-                        ssha%data,   &
-                        sshn_t%data, &
-                        ...,         &
-                        ssha%grid%area_t)
-\end{verbatim}
 
@@
 
+As with the Momentum kernel, we know that obtaining optimal
+performance from both the Gnu and Intel compilers requires that a
+kernel be manually in-lined at its call site. We do this for the
+Continuity kernel in this optimisation step.
+
 \subsection{In-line remaining kernels (BCs and time-update)}
+
+Having optimised the Continuity section we finally turn our attention
+to the Boundary Condition and Time-update sections. The kernels in
+these sections are small and dominated by conditional statements.  We
+therefore limited our optimisation of them to manually in-lining each
+of the kernels into the PSy layer.
 
@@
 
-Having show that, in general, we can recover and occasionally improve
-upon the performance of the original version of nemolite2d, the next
-logical step is to examine the necessary code transformations.  We do
-this for the $256^{2}$ case since this fits within cache on both the
-Haswell and Ivy Bridge CPUs we are using here.
-Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
-after each transformation has been applied to the code. The same data
-is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
-of the \psykal version for a given compiler/CPU is given relative to
-the performance of the original with the same compiler/CPU.
+Having show that we can recover, and often improve upon, the
+performance of the original version of nemolite2d, the next logical
+step is to examine the necessary code transformations in detail.  We
+do this for the $256^{2}$ case since this fits within cache on the Ivy
+Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
+detailed performance figures for this case after each transformation has been
+applied to the code. The same data is visualised in
+Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
+version for a given compiler/CPU is given relative to the performance
+of the original with the same compiler/CPU.
 
-\begin{table*}[!t]
+\begin{table}[!t]
 % increase table row spacing, adjust to taste
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
-code transformation.}
+\caption{Performance (millions of points updated per second) for
+  the $256^2$ case after each code transformation.}
 \label{TABLE_opt_breakdown}
 \centering
-\begin{tabular}{l|c|c|c|c|c|c}
+\begin{tabular}{l|c|c|c}
 \hline
-Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
+Compiler:           & Gnu & Intel & Cray  \\
 \hline
-CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
+CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
 \hline
-Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
-Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
+Original         & 3.87 & 12.2 & 6.83 \\
+Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
+Safe-loop        & 3.07 & 6.55 & 6.95 \\
+In-line Momentum & 4.31 & 10.7 & 6.93 \\
+SIMD Momentum    & 4.31 & 10.7 & 11.8 \\
+\parbox{2.5cm}{\raggedright Arrays from read-only objects} & 4.31 & 11.3 & 11.8 \\
+In-line Continuity         & 4.83 & 11.8 & 11.6 \\
+In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
+In-line field copies       & 5.92 & 12.5 & 11.4 \\
 \hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
+\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & -34.62 & -2.39 & -42.22 \\
 \hline
 \end{tabular}
-\end{table*}
+\end{table}
 
@@
 \includegraphics[width=90mm]{opt_stages_256}
-\caption{The performance of the \psykal version of nemolite2d for the
+\caption{Performance of the \psykal version of nemolite2d for the
   $256^{2}$ domain at each stage of optimisation. Results are given as

### 2015-02-25 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Tweak table containing perf. data for opt. stages and add text describing key features for each compiler.

@@
 \subsection{Addition of `safe-address' directives}
+\label{sec_safe_address}
 
@@
 \subsection{In-line Momentum kernel bodies into middle-layer code}
+\label{sec_inline_mom}
 
@@
 \subsection{In-line the Continuity kernel}
+\label{sec_cont_inline}
 
@@
 \subsection{In-line remaining kernels (BCs and time-update)}
+\label{sec_inline_all_kernels}
 
@@
 \subsection{In-line field-copy operations}
+\label{sec_inline_fldcopy}
 
@@
 
+Looking at the results for the Gnu compiler (and the Ivy Bridge CPU)
+first, all of the steps-up in performance correspond to kernel
+in-lining. None of the other transformations had any effect on the
+performance of the compiled code. In fact, simply in-lining the two
+kernels associated with the Momentum section was enough to improve
+upon the performance of the original code.
+
+With the Intel compiler, the single largest performance increase is
+again due to kernel in-lining (of the Momentum kernels). This is
+because the compiler does a much better job of SIMD vectorising the
+loops involved than it does when it first has to in-line the kernel
+itself (as evidenced by its own diagnostic output - see
+Section~\ref{sec_inline_mom}). However, although this gives a
+significant performance increase it is not sufficient to match the
+performance of the original version. This is only achieved by
+in-lining every kernel and making the lack of data dependencies
+between arrays accessed from different field objects more explicit.
+
+The Cray compiler is distinct from the other two in that kernel
+in-lining does not give any performance benefit and in fact, for the
+smaller kernels, it can actually hurt performance. Thus the key
+transformation is to encourage the compiler to SIMD vectorise the
+Momentum section via a compiler-specific directive (without this it
+concludes that such vectorisation would be inefficient). Of the other
+transformations, only the addition of the compiler-specific
+safe\_address directive (see Section~\ref{sec_safe_address}) was found
+to give any benefit.
+
 \begin{table}[!t]
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{Performance (millions of points updated per second) for
-  the $256^2$ case after each code transformation.}
+\caption{Performance (millions of points updated per second) for the
+  $256^2$ case after each code transformation. Where an optimisation
+  utilises compiler-specific directives then performance figures for
+  the other compilers are omitted.}
 \label{TABLE_opt_breakdown}
@@
 Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
-Safe-loop        & 3.07 & 6.55 & 6.95 \\
+Safe-address     & --   & --   & 6.95 \\
 In-line Momentum & 4.31 & 10.7 & 6.93 \\
-SIMD Momentum    & 4.31 & 10.7 & 11.8 \\
-\parbox{2.5cm}{\raggedright Arrays from read-only objects} & 4.31 & 11.3 & 11.8 \\
+SIMD Momentum    & --   & --   & 11.8 \\
+\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
 In-line Continuity         & 4.83 & 11.8 & 11.6 \\
@@
 \hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & -34.62 & -2.39 & -42.22 \\
+\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.62 & 2.39 & 42.22 \\
 \hline
@@
 \caption{Performance of the \psykal version of nemolite2d for the
-  $256^{2}$ domain at each stage of optimisation. Results are given as
-  a percentage of the performance of the original code for each
-  compiler/CPU combination. A figure greater than 100\% indicates that
-  the \psykal version performs better than the original.}
+  $256^{2}$ domain at each stage of optimisation. The first (black)
+  bar of each cluster gives the performance of the original version of
+  nemolite2d for that compiler/CPU combination.}
 \label{FIG_opt_stages_256}

### 2015-03-03 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Beginning to add OpenMP results and description

@@
 system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB. The CPU in the IBM BG/Q system is built around the
-PowerPC A2 core with a clock speed of 1.6~GHz. The chip has 32~MB of
-last-level cache.
+cache of 30~MB.
+
+We first describe the code transformations performed for the serial
+version of nemolite2d.  We then move on to the construction of
+parallel versions of the code using OpenMP and OpenACC. Again, we
+describe the key steps we have taken in this process in order to
+maximise the performance of the code.
+
+\subsection{Transformations of Serial Nemolite2D}
 
@@
 \hline
-                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
+                 & Gnu   & Intel       & Cray    & IBM      \\
 \hline
@@
 Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &          \\
-IBM A2 (BGQ)     &       &             &         &   --     \\
 IBM Power 8      &       &             &         &          \\
@@
 
-Before applying any code transformations, we benchmark the original
-version of the code. We also benchmark the vanilla, unoptimised
-version after it has been re-structured following the \psykal
-approach. In addition to this benchmarking, we profile these versions
-of the code at the algorithm level (using a high-level timing
+Before applying any code transformations, we first benchmark the
+original, serial version of the code. We also benchmark the vanilla,
+unoptimised version after it has been re-structured following the
+\psykal approach. In addition to this benchmarking, we profile these
+versions of the code at the algorithm level (using a high-level timing
 API). The resulting profiles are given in
@@
 Time-update and Boundary Condition regions are not significantly
-affected by the move so the \psykal version.
+affected by the move to the \psykal version.
 
@@
 
-\subsection{Addition of `safe-address' directives}
+\subsubsection{Addition of `safe-address' directives}
 \label{sec_safe_address}
@@
 
-\subsection{In-line Momentum kernel bodies into middle-layer code}
+\subsubsection{In-line Momentum kernel bodies into middle-layer code}
 \label{sec_inline_mom}
@@
 
-\subsection{Force SIMD vectorisation of the Momentum kernels using directives}
+\subsubsection{Force SIMD vectorisation of the Momentum kernels using directives}
 
@@
 
-\subsection{Work around limitations related to derived types}
+\subsubsection{Work around limitations related to derived types}
 
@@
 
-\subsection{In-line the Continuity kernel}
+\subsubsection{In-line the Continuity kernel}
 \label{sec_cont_inline}
@@
 
-\subsection{In-line remaining kernels (BCs and time-update)}
+\subsubsection{In-line remaining kernels (BCs and time-update)}
 \label{sec_inline_all_kernels}
@@
 
-\subsection{In-line field-copy operations}
+\subsubsection{In-line field-copy operations}
 \label{sec_inline_fldcopy}
@@
 optimising the code.
+
+\subsection{Construction of OpenMP-Parallel Nemolite2D}
+
+For this part of the work we began with the optimal \psykal version of
+the code, as obtained after applying the various transformations
+described in the previous section. As with the transformations of the
+serial code, our purpose here is to determine the functionality
+required of the PSyclone code generation system.
 
@@
 
+\subsubsection{OpenMP Version}
+
+For the OpenMP-parallel version of nemolite2d we began with the
+optimised form of the serial version, as described in the previous
+section.
+
+\begin{figure}
+\centering
+\includegraphics[width=2.8in]{omp_scaling_256_gnu}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
+  socket. The 24-thread runs employed hyperthreading.}
+\label{FIG_gnu_omp_256}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=2.8in]{omp_scaling_256_intel}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
+  socket. The 24-thread runs employed hyperthreading.}
+\label{FIG_intel_omp_256}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=2.8in]{omp_scaling_256_cray}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
+  socket. The 24-thread runs employed hyperthreading.}
+\label{FIG_cray_omp_256}
+\end{figure}
+
+\subsubsection{OpenACC Version}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-03-05 nemolite2d_perf.tex
Add text describing OpenMP optimisation stages

@@
 
-\subsection{Addition of `safe-address' directives}
+\subsubsection{Addition of `safe-address' directives}
+\label{sec_safe_address}
 
@@
 
-\subsection{In-line Momentum kernel bodies into middle-layer code}
+\subsubsection{In-line Momentum kernel bodies into middle-layer code}
+\label{sec_inline_mom}
 
@@
 
-\subsection{Force SIMD vectorisation of the Momentum kernels using directives}
+\subsubsection{Force SIMD vectorisation of the Momentum kernels 
+ using directives}
 
@@
 
-\subsection{In-line the Continuity kernel}
+\subsubsection{In-line the Continuity kernel}
+\label{sec_cont_inline}
 
@@
 
-\subsection{In-line remaining kernels (BCs and time-update)}
+\subsubsection{In-line remaining kernels (BCs and time-update)}
+\label{sec_inline_all_kernels}
 
@@
 
-\subsection{In-line field-copy operations}
+\subsubsection{In-line field-copy operations}
+\label{sec_inline_fldcopy}
 
@@
 optimising the code.
+
+\subsection{Construction of OpenMP-Parallel Nemolite2D}
+\label{sec_omp_steps}
+
+For this part of the work we began with the optimal \psykal version of
+the code, as obtained after applying the various transformations
+described in the previous section. As with the transformations of the
+serial code, our purpose here is to determine the functionality
+required of the PSyclone code-generation system.
+
+\subsubsection{Separate PARALLEL DOs}
+\label{sec_sep_dos}
+
+The simplest possible OpenMP-parallel implementation consists of
+parallelising each loop nest in the PSy layer. This was done by
+inserting an OpenMP PARALLEL DO directive before each loop nest so
+that the iterations of the outermost or $j$ loop (over the y-dimension
+of the model domain) are shared amongst the OpenMP threads. This
+leaves the innermost ($i$) loop available for SIMD vectorisation by
+the compiler.
+
+The loop nest dealing with the application of the Flather boundary
+condition to the y-component of velocity has a loop-carried dependency
+in $j$ which prevented its being executed in parallel. This was
+therefore left unchanged and executed on thread 0 only.
+
+\subsubsection{Single PARALLEL region}
+\label{sec_single_parallel}
+
+Although very simple to implement, the use of separate PARALLEL DO
+directives causes the team of OpenMP threads to be repeatedly created
+and destroyed. This may be avoided by keeping the thread team in
+existence for as long as possible using an OpenMP PARALLEL region.  We
+therefore enclosed the whole of the PSy layer (in this code, a single
+subroutine) within a single PARALLEL region. The directive preceeding
+each loop nest to be parallelised was then changed to an OpenMP DO. We
+ensured that the v-Flather loop nest was executed in serial (by the
+first thread to encounter it) by enclosing it within an OpenMP SINGLE
+section.
+
+\subsubsection{First-touch policy}
+\label{sec_1sttouch}
+
+When executing an OpenMP-parallel program on a Non-Uniform Memory
+Access (NUMA) compute node it becomes important to ensure that the
+memory locations accessed by each thread are local to the hardware
+core upon which it is executing. One way of doing this is to implement
+a so-called 'first-touch policy' whereby memory addresses that will
+generally be accessed by a given thread during program execution are
+first initialised by that thread. This is simply achieved by using an
+OpenMP-parallel loop to initialise newly-allocated arrays to some
+value, {\it e.g.} zero.
+
+Since data arrays are managed within the GOcean Library {\bf (is this
+  a good/appropriate name?)} this optimisation can again be
+implemented without changing user code (the Application and Kernel
+layers).
+
+\subsubsection{Minimise thread synchronisation}
+\label{sec_synch}
+
+By default, the OpenMP END DO directive includes an implicit barrier,
+thus causing all threads to wait until the slowest has completed the
+preceeding loop. Such synchronisation limits performance at larger
+thread counts and, for the nemolite2d code, is frequently unecessary.
+{\it E.g.} if a kernel does not make use of the results of a
+preceeding kernel call then there is clearly no need for threads to
+wait between the two kernels.
+
+We analysed the inter-dependencies of each of the code sections within
+the PSy layer and removed all unecessary barriers by adding the NOWAIT
+qualifier to the relevant OpenMP END DO or END SINGLE directives. This
+reduced the number of barriers from eleven down to four.
+
+\subsubsection{Amortize serial region}
+\label{sec_amortize}
+
+As previously mentioned, the v-Flather section has to be executed in
+serial becuase of a loop-carried dependence in $j$. (In principle we
+could choose to parallelise the inner, $i$ loop but that would inhibit
+its SIMD vectorisation.) This introduces a load-imbalance between the
+threads. We attempt to mitigate this by moving this serial section to
+before the (parallel) u-Flather section. Since these two sections are
+independent, the aim is that the thread that performs the serial,
+v-Flather computation then performs a smaller share of the following
+u-Flather loop. In practice, this requires that some form of dynamic
+thread scheduling is used.
+
+\subsubsection{Thread scheduling}
+\label{sec_scheduling}
+
+In order to investigate how thread scheduling affects performance we
+used the `runtime' argument to the OpenMP SCHEDULE qualifier for all
+of our OpenMP parallel loops. The actual schedule to use can then be
+set at run-time using the OMP\_SCHEDULE environment variable. We
+experimented with using the static, dynamic and guided (with varying
+chunk size) OpenMP schedules.
 
@@
 
-Having show that we can recover and often improve upon the performance
-of the original version of nemolite2d, the next logical step is to
-examine the necessary code transformations in detail.  We do this for
-the $256^{2}$ case since this fits within cache on the Ivy Bridge CPUs
-we are using here.  Table~\ref{TABLE_opt_breakdown} shows detailed
-timings for this case after each transformation has been applied to
-the code. The same data is visualised in
+Having show that we can recover, and often improve upon, the
+performance of the original version of nemolite2d, the next logical
+step is to examine the necessary code transformations in detail.  We
+do this for the $256^{2}$ case since this fits within cache on the Ivy
+Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
+detailed performance figures for this case after each transformation has been
+applied to the code. The same data is visualised in
 Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
@@
 
-\begin{table*}[!t]
+Looking at the results for the Gnu compiler (and the Ivy Bridge CPU)
+first, all of the steps-up in performance correspond to kernel
+in-lining. None of the other transformations had any effect on the
+performance of the compiled code. In fact, simply in-lining the two
+kernels associated with the Momentum section was enough to improve
+upon the performance of the original code.
+
+With the Intel compiler, the single largest performance increase is
+again due to kernel in-lining (of the Momentum kernels). This is
+because the compiler does a much better job of SIMD vectorising the
+loops involved than it does when it first has to in-line the kernel
+itself (as evidenced by its own diagnostic output - see
+Section~\ref{sec_inline_mom}). However, although this gives a
+significant performance increase it is not sufficient to match the
+performance of the original version. This is only achieved by
+in-lining every kernel and making the lack of data dependencies
+between arrays accessed from different field objects more explicit.
+
+The Cray compiler is distinct from the other two in that kernel
+in-lining does not give any performance benefit and in fact, for the
+smaller kernels, it can actually hurt performance. Thus the key
+transformation is to encourage the compiler to SIMD vectorise the
+Momentum section via a compiler-specific directive (without this it
+concludes that such vectorisation would be inefficient). Of the other
+transformations, only the addition of the compiler-specific
+safe\_address directive (see Section~\ref{sec_safe_address}) was found
+to give any benefit.
+
+\begin{table}[!t]
 % increase table row spacing, adjust to taste
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
-code transformation.}
+\caption{Performance (millions of points updated per second) for the
+  $256^2$ case after each code transformation. Where an optimisation
+  utilises compiler-specific directives then performance figures for
+  the other compilers are omitted.}
 \label{TABLE_opt_breakdown}
 \centering
-\begin{tabular}{l|c|c|c|c|c|c}
-\hline
-Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
-\hline
-CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
-\hline
-Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
-Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
-\hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
+\begin{tabular}{l|c|c|c}
+\hline
+Compiler:           & Gnu & Intel & Cray  \\
+\hline
+CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
+\hline
+Original         & 3.87 & 12.2 & 6.83 \\
+Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
+Safe-address     & --   & --   & 6.95 \\
+In-line Momentum & 4.31 & 10.7 & 6.93 \\
+SIMD Momentum    & --   & --   & 11.8 \\
+\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
+In-line Continuity         & 4.83 & 11.8 & 11.6 \\
+In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
+In-line field copies       & 5.92 & 12.5 & 11.4 \\
+\hline
+\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.62 & 2.39 & 42.22 \\
 \hline
 \end{tabular}
-\end{table*}
+\end{table}
 
@@
 \includegraphics[width=90mm]{opt_stages_256}
-\caption{The performance of the \psykal version of nemolite2d for the
-  $256^{2}$ domain at each stage of optimisation. Results are given as
-  a percentage of the performance of the original code for each
-  compiler/CPU combination. A figure greater than 100\% indicates that
-  the \psykal version performs better than the original.}
+\caption{Performance of the \psykal version of nemolite2d for the
+  $256^{2}$ domain at each stage of optimisation. The first (black)
+  bar of each cluster gives the performance of the original version of
+  nemolite2d for that compiler/CPU combination.}
 \label{FIG_opt_stages_256}
@@
 
+\subsubsection{OpenMP Version}
+
+As with the serial optimisations, we consider the effect of each of
+the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
+for the $256^2$
+domain. Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256} show the performance of each of the
+versions of the code for the Gnu, Intel and Cray compilers,
+respectively.
+
+\begin{figure}
+\centering
+\includegraphics[width=2.8in]{omp_scaling_256_gnu}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
+  socket. The 24-thread runs employed hyperthreading.}
+\label{FIG_gnu_omp_256}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=2.8in]{omp_scaling_256_intel}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
+  socket. The 24-thread runs employed hyperthreading.}
+\label{FIG_intel_omp_256}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=2.8in]{omp_scaling_256_cray}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
+  socket. The 24-thread runs employed hyperthreading.}
+\label{FIG_cray_omp_256}
+\end{figure}
+
+\subsubsection{OpenACC Version}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-03-09 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add graphs of GPU performance

@@
 \hyphenation{op-tical net-works semi-conduc-tor}
+
+\newlength{\picwidth}
+\setlength{\picwidth}{85mm}
 
@@
 
-\subsubsection{Force SIMD vectorisation of the Momentum kernels using directives}
+\subsubsection{Force SIMD vectorisation of the Momentum kernels 
+ using directives}
 
@@
 \subsection{Construction of OpenMP-Parallel Nemolite2D}
+\label{sec_omp_steps}
 
@@
 serial code, our purpose here is to determine the functionality
-required of the PSyclone code generation system.
+required of the PSyclone code-generation system.
+
+\subsubsection{Separate PARALLEL DOs}
+\label{sec_sep_dos}
+
+The simplest possible OpenMP-parallel implementation consists of
+parallelising each loop nest in the PSy layer. This was done by
+inserting an OpenMP PARALLEL DO directive before each loop nest so
+that the iterations of the outermost or $j$ loop (over the y-dimension
+of the model domain) are shared amongst the OpenMP threads. This
+leaves the innermost ($i$) loop available for SIMD vectorisation by
+the compiler.
+
+The loop nest dealing with the application of the Flather boundary
+condition to the y-component of velocity has a loop-carried dependency
+in $j$ which prevented its being executed in parallel. This was
+therefore left unchanged and executed on thread 0 only.
+
+\subsubsection{Single PARALLEL region}
+\label{sec_single_parallel}
+
+Although very simple to implement, the use of separate PARALLEL DO
+directives causes the team of OpenMP threads to be repeatedly created
+and destroyed. This may be avoided by keeping the thread team in
+existence for as long as possible using an OpenMP PARALLEL region.  We
+therefore enclosed the whole of the PSy layer (in this code, a single
+subroutine) within a single PARALLEL region. The directive preceeding
+each loop nest to be parallelised was then changed to an OpenMP DO. We
+ensured that the v-Flather loop nest was executed in serial (by the
+first thread to encounter it) by enclosing it within an OpenMP SINGLE
+section.
+
+\subsubsection{First-touch policy}
+\label{sec_1sttouch}
+
+When executing an OpenMP-parallel program on a Non-Uniform Memory
+Access (NUMA) compute node it becomes important to ensure that the
+memory locations accessed by each thread are local to the hardware
+core upon which it is executing. One way of doing this is to implement
+a so-called 'first-touch policy' whereby memory addresses that will
+generally be accessed by a given thread during program execution are
+first initialised by that thread. This is simply achieved by using an
+OpenMP-parallel loop to initialise newly-allocated arrays to some
+value, {\it e.g.} zero.
+
+Since data arrays are managed within the GOcean Library {\bf (is this
+  a good/appropriate name?)} this optimisation can again be
+implemented without changing user code (the Application and Kernel
+layers).
+
+\subsubsection{Minimise thread synchronisation}
+\label{sec_synch}
+
+By default, the OpenMP END DO directive includes an implicit barrier,
+thus causing all threads to wait until the slowest has completed the
+preceeding loop. Such synchronisation limits performance at larger
+thread counts and, for the nemolite2d code, is frequently unecessary.
+{\it E.g.} if a kernel does not make use of the results of a
+preceeding kernel call then there is clearly no need for threads to
+wait between the two kernels.
+
+We analysed the inter-dependencies of each of the code sections within
+the PSy layer and removed all unecessary barriers by adding the NOWAIT
+qualifier to the relevant OpenMP END DO or END SINGLE directives. This
+reduced the number of barriers from eleven down to four.
+
+\subsubsection{Amortize serial region}
+\label{sec_amortize}
+
+As previously mentioned, the v-Flather section has to be executed in
+serial becuase of a loop-carried dependence in $j$. (In principle we
+could choose to parallelise the inner, $i$ loop but that would inhibit
+its SIMD vectorisation.) This introduces a load-imbalance between the
+threads. We attempt to mitigate this by moving this serial section to
+before the (parallel) u-Flather section. Since these two sections are
+independent, the aim is that the thread that performs the serial,
+v-Flather computation then performs a smaller share of the following
+u-Flather loop. In practice, this requires that some form of dynamic
+thread scheduling is used.
+
+\subsubsection{Thread scheduling}
+\label{sec_scheduling}
+
+In order to investigate how thread scheduling affects performance we
+used the `runtime' argument to the OpenMP SCHEDULE qualifier for all
+of our OpenMP parallel loops. The actual schedule to use can then be
+set at run-time using the OMP\_SCHEDULE environment variable. We
+experimented with using the standard static, dynamic and guided (with varying
+chunk size) OpenMP schedules.
 
@@
 
-For the OpenMP-parallel version of nemolite2d we began with the
-optimised form of the serial version, as described in the previous
-section.
+As with the serial optimisations, we consider the effect of each of
+the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
+for the $256^2$
+domain. Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256} show the performance of each of the
+versions of the code for the Gnu, Intel and Cray compilers,
+respectively.
 
@@
 \centering
-\includegraphics[width=2.8in]{omp_scaling_256_gnu}
+\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
 \caption{Performance of the OpenMP-parallel version of \psykal
@@
 \centering
-\includegraphics[width=2.8in]{omp_scaling_256_intel}
+\includegraphics[width=\picwidth]{omp_scaling_256_intel}
 \caption{Performance of the OpenMP-parallel version of \psykal
@@
 \centering
-\includegraphics[width=2.8in]{omp_scaling_256_cray}
+\includegraphics[width=\picwidth]{omp_scaling_256_cray}
 \caption{Performance of the OpenMP-parallel version of \psykal
@@
 \subsubsection{OpenACC Version}
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{gpu_opt_stages}
+\caption{Performance of the various GPU implementations of nemolite2d.}
+\label{FIG_gpu_opt_stages}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{cpu_cf_gpu}
+\caption{Performance of the best OpenMP-parallel version of \psykal
+  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation.}
+\label{FIG_cpu_cf_gpu}
+\end{figure}
 

### 2015-03-09 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Added Jeremy as author

@@
 
-\author{H.~Lieu and J.~Holt, A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
-\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}
+\author{J.~Appleyard, H.~Lieu, J.~Holt, R.~W.~Ford, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
+\thanks{J.~Appleyard is with NVIDIA.}%
+\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
 \thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}).
+(\url{http://www.archer.ac.uk}) and some {\bf NVIDIA system}.
 

### 2015-03-10 nemolite2d_perf.tex
Add discussion of OpenMP and OpenACC results

@@
 
+\newlength{\picwidth}
+\setlength{\picwidth}{85mm}
+
 \newcommand{\psykal}{{PS}y{KA}l\ }
@@
 
-\author{H.~Lieu and J.~Holt, A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
-\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}
+\author{J.~Appleyard, R.~W.~Ford, J.~Holt, H.~Lieu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
+\thanks{J.~Appleyard is with NVIDIA.}%
+\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
 \thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
@@
 \renewcommand{\arraystretch}{1.3}
-\caption{The matrix of compilers and CPUs used in this work. The use
-  of a compiler on a given CPU is indicated by the specification of
-  the version of the compiler in the relevent element. No entry
-  indicates that a compiler was not available/used on that CPU.}
+\caption{The matrix of compilers and hardware used in this work. The
+  use of a compiler on a given system is indicated by the specification
+  of the version of the compiler in the relevent element. No entry
+  indicates that a compiler was not available/used on that system.}
 \label{TABLE_compilers}
 \centering
-\begin{tabular}{|l|c|c|c|c|}
-\hline
-                 & \multicolumn{4}{c|}{Compiler}            \\
-\hline
-                 & Gnu   & Intel       & Cray    & IBM      \\
-\hline
-Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
-Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &          \\
-IBM Power 8      &       &             &         &          \\
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+                 & \multicolumn{5}{c|}{Compiler}            \\
+\hline
+                 & Gnu   & Intel       & Cray    & IBM    &  PGI \\
+\hline
+Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
+Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
+IBM Power 8      &       &             &         &        &  \\
+NVIDIA Tesla K40 &       &             &         &        &  {\bf Jeremy} \\
 \hline
@@
 Cray      & -O3 -O ipa5 -h wp               \\
-IBM bgxlf & -O5 -qinline=auto:level=10      \\
+IBM       & -O5 -qinline=auto:level=10      \\
+PGI       & {\bf Jeremy to supply} \\
 \hline
@@
 set at run-time using the OMP\_SCHEDULE environment variable. We
-experimented with using the static, dynamic and guided (with varying
+experimented with using the standard static, dynamic and guided (with varying
 chunk size) OpenMP schedules.
@@
 
+We now turn to transformations related to parallelisation of the
+nemolite2d code; the introduction of OpenMP and OpenACC directives. In
+keeping with the \psykal approach, these transformations are
+restricted to the PSy (middle) layer of the code.
+
 \subsubsection{OpenMP Version}
@@
 the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
-for the $256^2$
-domain. Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+for the $256^2$ domain. For this we principally use a single Intel,
+Ivy Bridge socket which has 12 hardware cores and support for up to 24
+threads with hyperthreading.
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
 and~\ref{FIG_cray_omp_256} show the performance of each of the
-versions of the code for the Gnu, Intel and Cray compilers,
-respectively.
+versions of the code on this system for the Gnu, Intel and Cray
+compilers, respectively.
 
@@
 \centering
-\includegraphics[width=2.8in]{omp_scaling_256_gnu}
+\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
 \caption{Performance of the OpenMP-parallel version of \psykal
   nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading.}
+  socket. The 24-thread runs employed hyperthreading and the optimal
+  OpenMP schedule is given in parentheses.}
 \label{FIG_gnu_omp_256}
@@
 
+The simplest OpenMP implementation (black lines) fails to scale beyond
+two threads for any of the compilers. However, as soon as a single
+PARALLEL region is introduced, the situation is greatly improved with
+all three executables now scaling out to at least 12 threads (red
+lines). {\bf Add parallel efficiency lines to the OMP scaling plots to
+  highlight that things are still not brilliant.}
+
+In restricting ourselves to a single socket, we are keeping all
+threads within a single NUMA region. It is therefore surprising that
+implementing a `first-touch' policy has any effect and yet, for the
+Gnu- and Intel-compiled binaries, it appears to improve performance
+when hyperthreading is employed to run on 24 threads (green lines in
+Figures~\ref{FIG_gnu_omp_256} and ~\ref{FIG_intel_omp_256}). {\bf Can
+  we explain this?}
+
 \begin{figure}
 \centering
-\includegraphics[width=2.8in]{omp_scaling_256_intel}
+\includegraphics[width=\picwidth]{omp_scaling_256_intel}
 \caption{Performance of the OpenMP-parallel version of \psykal
   nemolite2d for the Intel compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading.}
+  socket. The 24-thread runs employed hyperthreading and the optimal
+  OpenMP schedule is given in parentheses.}
 \label{FIG_intel_omp_256}
@@
 \centering
-\includegraphics[width=2.8in]{omp_scaling_256_cray}
+\includegraphics[width=\picwidth]{omp_scaling_256_cray}
 \caption{Performance of the OpenMP-parallel version of \psykal
   nemolite2d for the Cray compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading.}
+  socket. The 24-thread runs employed hyperthreading and the optimal
+  OpenMP schedule is given in parentheses.}
 \label{FIG_cray_omp_256}
@@
 
+The final optimisation step found to have any significant effect is to
+minimise the amount of thread synchronisation by introducing the
+NOWAIT qualifier wherever possible (blue lines). For the Gnu compiler,
+this improves the performance of the executable on eight or more
+threads while, for the Intel compiler, it only gives an improvement
+for the 24-thread case. Moving the SINGLE region before a parallel
+loop is marginally beneficial for the Gnu- and Cray-compiled binaries
+and yet reduces the performance of the Intel binary (purple lines).
+
+We have used different scales for the $y$-axes in each of the plots in
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256} in order to highlight the performance
+differences between the code versions with a given compiler. However,
+the best performance obtained on twelve threads ({\it i.e.}  without
+hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
+Intel and Cray compilers, respectively. This serves to emphasise the
+democratisation that the introduction of OpenMP has had; for the
+serial case the Cray- and Intel-compiled exectables were a factor of
+two faster than the Gnu-compiled binary
+(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
+Gnu-compiled binary is only 6\% slower than that produced by the Intel
+compiler and is 13\% {\em faster} that that of the Cray compiler. In
+part this situation comes about because of the effect that adding the
+OpenMP compiler flag has on the optimisations performed by the
+compiler.  In particular, the Cray compiler no-longer vectorises the
+key Momentum section, despite the directive added during the serial
+optimisation work.
+
+A further complication is the choice of scheduling of the OpenMP
+threads.  We have investigated the performance of each of the
+executables (and thus the associated OpenMP run-time library) with the
+standard OpenMP {\it static}, {\it dynamic} and {\it guided}
+scheduling policies~\cite{openmp_standard}. For the Intel
+compiler/run-time, static loop scheduling was found to be best for all
+versions apart from that where we have attempted to amortize the cost
+of the SINGLE section. This is to be expected since that strategy
+requires some form of dynamic loop scheduling in order to reduce the
+load imbalance introduced by the SINGLE section.
+
+In contrast, some form of dynamic scheduling gave a performance
+improvement with the Gnu compiler/run-time even for the `first-touch'
+version of the code. This is despite the fact that this version
+contains (implicit) thread synchronisation after every parallel loop.
+For the Cray compiler/run-time, some form of dynamic scheduling became
+optimal once inter-thread synchronisation was reduced using the NOWAIT
+qualifiers.
+
 \subsubsection{OpenACC Version}
+
+The introduction of OpenACC directives proved to be simpler than
+producing the OpenMP version of nemolite2d {\bf - Jeremy to clarify
+  whether this is true}.  Consequently we only show the performance of
+{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{gpu_opt_stages}
+\caption{Performance of the various GPU implementations of nemolite2d.}
+\label{FIG_gpu_opt_stages}
+\end{figure}
+
+In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
+the OpenMP and OpenACC implementations of nemolite2d across the range
+of problem sizes considered. The OpenMP figures are the maximum
+performance obtained from a whole Intel Ivy Bridge socket by any
+version of the code on any number of threads for a given
+compiler/run-time. For the smallest domain size ($64^2$) the OpenMP
+version significantly outperforms the GPU because there is
+insufficient parallelism to fully utilise the GPU and one time-step
+takes only 80~$\mu$s. The execution of a single time-step is then
+dominated by the time taken to launch a kernel on the GPU rather than
+the execution of the kernel itself.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{cpu_cf_gpu}
+\caption{Performance of the best OpenMP-parallel version of \psykal
+  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
+\label{FIG_cpu_cf_gpu}
+\end{figure}
+
+Once the problem size is increased to $128^2$, a single time-step
+takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
+comparable in performance to the OpenACC version. For all of the
+larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
+version is considerably faster than the CPU.  For problem sizes of
+$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
+exhausted and performance becomes limited by the bandwidth to main
+memory. At this stage the OpenACC version of the code on the GPU is
+some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
+than the OpenMP version on the CPU. {\bf Can we relate this to the
+  difference in peak memory b/w between the GPU and CPU?}
 

### 2015-03-10 nemolite2d_perf.tex
Update OMP scaling plots to include parallel efficiency. Add grace files to svn.

@@
   nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading and the optimal
-  OpenMP schedule is given in parentheses.}
+  socket. The corresponding parallel efficiencies are shown using open
+  symbols and dashed lines. The 24-thread runs employed hyperthreading
+  and the optimal OpenMP schedule is given in parentheses.}
 \label{FIG_gnu_omp_256}
@@
   nemolite2d for the Intel compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading and the optimal
-  OpenMP schedule is given in parentheses.}
+  socket.  The corresponding parallel efficiencies are shown using
+  open symbols and dashed lines. The 24-thread runs employed
+  hyperthreading and the optimal OpenMP schedule is given in
+  parentheses.}
 \label{FIG_intel_omp_256}
@@
   nemolite2d for the Cray compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading and the optimal
-  OpenMP schedule is given in parentheses.}
+  socket.  The corresponding parallel efficiencies are shown using
+  open symbols and dashed lines. The 24-thread runs employed
+  hyperthreading and the optimal OpenMP schedule is given in
+  parentheses.}
 \label{FIG_cray_omp_256}
@@
 
-The final optimisation step found to have any significant effect is to
-minimise the amount of thread synchronisation by introducing the
-NOWAIT qualifier wherever possible (blue lines). For the Gnu compiler,
-this improves the performance of the executable on eight or more
-threads while, for the Intel compiler, it only gives an improvement
-for the 24-thread case. Moving the SINGLE region before a parallel
-loop is marginally beneficial for the Gnu- and Cray-compiled binaries
-and yet reduces the performance of the Intel binary (purple lines).
+The final optimisation step that we found to have any significant
+effect is to minimise the amount of thread synchronisation by
+introducing the NOWAIT qualifier wherever possible (blue lines). For
+the Gnu compiler, this improves the performance of the executable on
+eight or more threads while, for the Intel compiler, it only gives an
+improvement for the 24-thread case. Moving the SINGLE region before a
+parallel loop is marginally beneficial for the Gnu- and Cray-compiled
+binaries and yet reduces the performance of the Intel binary (purple
+lines).
 

### 2015-03-10 nemolite2d_perf.tex
Start new para on parallel efficiency

@@
 all three executables now scaling out to at least 12 threads (red
-lines). {\bf Add parallel efficiency lines to the OMP scaling plots to
-  highlight that things are still not brilliant.}
+lines, square symbols). {\bf Add parallel efficiency lines to the OMP
+  scaling plots to highlight that things are still not brilliant.}
 
@@
 Gnu- and Intel-compiled binaries, it appears to improve performance
-when hyperthreading is employed to run on 24 threads (green lines in
-Figures~\ref{FIG_gnu_omp_256} and ~\ref{FIG_intel_omp_256}). {\bf Can
-  we explain this?}
+when hyperthreading is employed to run on 24 threads (green lines and
+diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
+~\ref{FIG_intel_omp_256}). {\bf Can we explain this?}
 
@@
 effect is to minimise the amount of thread synchronisation by
-introducing the NOWAIT qualifier wherever possible (blue lines). For
-the Gnu compiler, this improves the performance of the executable on
-eight or more threads while, for the Intel compiler, it only gives an
-improvement for the 24-thread case. Moving the SINGLE region before a
-parallel loop is marginally beneficial for the Gnu- and Cray-compiled
-binaries and yet reduces the performance of the Intel binary (purple
-lines).
+introducing the NOWAIT qualifier wherever possible (blue lines and
+upwards-triangle symbols). For the Gnu compiler, this improves the
+performance of the executable on eight or more threads while, for the
+Intel compiler, it only gives an improvement for the 24-thread
+case. Moving the SINGLE region before a parallel loop is marginally
+beneficial for the Gnu- and Cray-compiled binaries and yet reduces the
+performance of the Intel binary (purple lines and right-pointing
+triangles).
+
+In order to quantify the scaling behaviour of the different versions
+of nemolite2d with the different compilers/run-time environments, we
+also plot the parallel efficiency in
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256}. We define parallel efficiency, $E(n)$, on
+a number of threads, $n$, as:
+\begin{equation}
+E(n) = 100 * \frac{P(n)}{ n*P(1)}
+\end{equation}
+where $P(n)$ is the performance of the code on $n$ threads.
 

### 2015-03-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add more text using parallel efficiency metric

@@
 
-\author{J.~Appleyard, H.~Lieu, J.~Holt, R.~W.~Ford, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
+\author{J.~Appleyard, R.~W.~Ford, J.~Holt, H.~Lieu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
 \thanks{J.~Appleyard is with NVIDIA.}%
@@
 the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
-for the $256^2$
-domain. Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+for the $256^2$ domain. For this we principally use a single Intel,
+Ivy Bridge socket which has 12 hardware cores and support for up to 24
+threads with hyperthreading.
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
 and~\ref{FIG_cray_omp_256} show the performance of each of the
-versions of the code for the Gnu, Intel and Cray compilers,
-respectively.
+versions of the code on this system for the Gnu, Intel and Cray
+compilers, respectively.
+
+In order to quantify the scaling behaviour of the different versions
+of nemolite2d with the different compilers/run-time environments, we
+also plot the parallel efficiency in
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
+parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
+\[
+E(n) = 100  \frac{P(n)}{ n P(1)}
+\]
+where $P(m)$ is the performance of the code on $m$ threads. For a
+perfect, linearly-scaling code, $E(n)$ will be 100\%.
 
@@
   nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading.}
+  socket. The corresponding parallel efficiencies are shown using open
+  symbols and dashed lines. The 24-thread runs employed hyperthreading
+  and the optimal OpenMP schedule is given in parentheses.}
 \label{FIG_gnu_omp_256}
 \end{figure}
+
+The simplest OpenMP implementation (black lines, circle symbols) fails
+to scale well for any of the compilers. For the Intel and Gnu
+versions, parallel efficiency is already less than 50\% on just four
+threads. The Cray version however does better and is about 45\%
+efficient on eight threads.
+
+With the move to a single PARALLEL region, the situation
+is greatly improved with all three executables now scaling out to at
+least 12 threads with $\sim70$\% efficiency (red lines, square symbols).
+
+In restricting ourselves to a single socket, we are keeping all
+threads within a single NUMA region. It is therefore surprising that
+implementing a `first-touch' policy has any effect and yet, for the
+Gnu- and Intel-compiled binaries, it appears to improve performance
+when hyperthreading is employed to run on 24 threads (green lines and
+diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
+~\ref{FIG_intel_omp_256}). {\bf Can we explain this?}
 
@@
   nemolite2d for the Intel compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading.}
+  socket.  The corresponding parallel efficiencies are shown using
+  open symbols and dashed lines. The 24-thread runs employed
+  hyperthreading and the optimal OpenMP schedule is given in
+  parentheses.}
 \label{FIG_intel_omp_256}
@@
   nemolite2d for the Cray compiler on a single Intel Ivy Bridge
-  socket. The 24-thread runs employed hyperthreading.}
+  socket.  The corresponding parallel efficiencies are shown using
+  open symbols and dashed lines. The 24-thread runs employed
+  hyperthreading and the optimal OpenMP schedule is given in
+  parentheses.}
 \label{FIG_cray_omp_256}
@@
 
+The final optimisation step that we found to have any significant
+effect is to minimise the amount of thread synchronisation by
+introducing the NOWAIT qualifier wherever possible (blue lines and
+upwards-triangle symbols). For the Gnu compiler, this improves the
+performance of the executable on eight or more threads while, for the
+Intel compiler, it only gives an improvement for the 24-thread
+case. Moving the SINGLE region before a parallel loop is marginally
+beneficial for the Gnu- and Cray-compiled binaries and yet reduces the
+performance of the Intel binary (purple lines and right-pointing
+triangles).
+
+We have used different scales for the $y$-axes in each of the plots in
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256} in order to highlight the performance
+differences between the code versions with a given compiler. However,
+the best performance obtained on twelve threads ({\it i.e.}  without
+hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
+Intel and Cray compilers, respectively. This serves to emphasise the
+democratisation that the introduction of OpenMP has had; for the
+serial case the Cray- and Intel-compiled exectables were a factor of
+two faster than the Gnu-compiled binary
+(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
+Gnu-compiled binary is only 6\% slower than that produced by the Intel
+compiler and is 13\% {\em faster} that that of the Cray compiler. In
+part this situation comes about because of the effect that adding the
+OpenMP compiler flag has on the optimisations performed by the
+compiler.  In particular, the Cray compiler no-longer vectorises the
+key Momentum section, despite the directive added during the serial
+optimisation work.
+
+A further complication is the choice of scheduling of the OpenMP
+threads.  We have investigated the performance of each of the
+executables (and thus the associated OpenMP run-time library) with the
+standard OpenMP {\it static}, {\it dynamic} and {\it guided}
+scheduling policies~\cite{openmp_standard}. For the Intel
+compiler/run-time, static loop scheduling was found to be best for all
+versions apart from that where we have attempted to amortize the cost
+of the SINGLE section. This is to be expected since that strategy
+requires some form of dynamic loop scheduling in order to reduce the
+load imbalance introduced by the SINGLE section.
+
+In contrast, some form of dynamic scheduling gave a performance
+improvement with the Gnu compiler/run-time even for the `first-touch'
+version of the code. This is despite the fact that this version
+contains (implicit) thread synchronisation after every parallel loop.
+For the Cray compiler/run-time, some form of dynamic scheduling became
+optimal once inter-thread synchronisation was reduced using the NOWAIT
+qualifiers.
+
 \subsubsection{OpenACC Version}
+
+The introduction of OpenACC directives proved to be simpler than
+producing the OpenMP version of nemolite2d {\bf - Jeremy to clarify
+  whether this is true}.  Consequently we only show the performance of
+{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.
 
@@
 
+In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
+the OpenMP and OpenACC implementations of nemolite2d across the range
+of problem sizes considered. The OpenMP figures are the maximum
+performance obtained from a whole Intel Ivy Bridge socket by any
+version of the code on any number of threads for a given
+compiler/run-time. For the smallest domain size ($64^2$) the OpenMP
+version significantly outperforms the GPU because there is
+insufficient parallelism to fully utilise the GPU and one time-step
+takes only 80~$\mu$s. The execution of a single time-step is then
+dominated by the time taken to launch a kernel on the GPU rather than
+the execution of the kernel itself.
+
 \begin{figure}
@@
 \caption{Performance of the best OpenMP-parallel version of \psykal
-  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation.}
+  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
 \label{FIG_cpu_cf_gpu}
 \end{figure}
+
+Once the problem size is increased to $128^2$, a single time-step
+takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
+comparable in performance to the OpenACC version. For all of the
+larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
+version is considerably faster than the CPU.  For problem sizes of
+$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
+exhausted and performance becomes limited by the bandwidth to main
+memory. At this stage the OpenACC version of the code on the GPU is
+some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
+than the OpenMP version on the CPU. {\bf Can we relate this to the
+  difference in peak memory b/w between the GPU and CPU?}
 

### 2015-03-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add plot and text for comparison of 1- and 2-socket performance results

@@
 
+Since NEMO and similar finite-difference codes tend to be
+memory-bandwidth bound, we checked the sensitivity of our performance
+results to this quantity by benchmarking using two sockets of Intel
+Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
+configuration we ensured that threads were evenly shared over the two
+sockets. The performance obtained for the $256^2$ case with the `early
+SINGLE' version of the code is compared with the single-socket
+performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
+available memory bandwidth in this way has little effect on
+performance - the two-socket performance figures track those from a
+single socket very closely. The only realy difference in performance
+is at 24 threads where, in addition to the difference in available
+memory bandwidth, the single-socket configuration is using
+hyperthreading while the two-socket case is not.  {\bf Have query in
+  with Archer helpdesk on behaviour seen with Cray compiler here. Also
+  would like to run 512 case on 2 sockets to see how it goes out to 48
+  threads.}
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{1and2sockets}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d on one and two sockets of Intel Ivy Bridge. The 24-thread
+  runs on a single socket used hyperthreading and the two-socket runs
+  had the threads shared equally between the sockets.}
+\label{FIG_omp_2socks}
+\end{figure}
+
 A further complication is the choice of scheduling of the OpenMP

### 2015-03-13 docs/papers/shallow_performance/shallow_perf.tex
Add Power8 results

@@
 
-\author{A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
+\author{R.~W.~Ford, M.~Modani, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
 \thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
 \thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
+\thanks{M.~Modani is with IBM.}
 }
@@
 scientific software by the drive towards Exascale computing is
-considerable.
+considerable. With power consumption becoming the overriding design
+constraint, CPU clock speeds are falling and the complex,
+multi-purpose compute core is being replaced by multiple, simpler
+cores. This philosophy can be seen at work in the rise of so-called
+accelerator based machines in the Top 500 List~\cite{top500} of
+supercomputers: five of the top-ten machines in the November 2014 list
+make use of Intel Xeon Phi's or NVIDIA GPUs. Four of the remaining
+five machines in the top ten are IBM BlueGene/Qs, the CPU of which has
+hardware support for running 64 threads.
+
+Achieving good performance on large numbers of light-weight cores
+requires exploiting as much parallelism in an application as possible
+and this results in increased complexity in the programming models
+that must be used. This in turn increases the burden of code
+maintenance and code development, in part because two specialisms are
+required: that of the scientific domain which a code is modelling
+({\it e.g.} oceanography) and that of computational science. The
+situation is currently complicated still further by the existence of
+competing hardware technology; if one was to begin writing a major
+scientific application today it is unclear whether one would target
+GPU, Xeon Phi, traditional CPU, FPGA or something else entirely. This
+is a problem because, generally speaking, these different technologies
+require different programming approaches.
 
@@
 finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
-Originally written by Paul Swarztrauber of the National Center for
-Atmospheric Research, US.  Organic development - optimised for
-previous hardware. In describing our work, we term the version of the
-shallow code obtained at the beginning of this work the `original'
-version.
+This software was originally written in 1984 by Paul Swarztrauber of
+the National Center for Atmospheric Research, US.  However, in common
+with many scientific codes, it has subsequently undergone some sort of
+evolutionary development with subsequent people making various changes
+and optimising it for previous generations of hardware.  In describing
+our work, we term the version of the Shallow program obtained at the
+beginning of this project the `original' version.
 
-In what follows we consider square domains of dimension 64, 128, 256,
-512 and 1024.
+Shallow is a very good test case for our purposes since the original
+version is short (some 600 lines) and contained within a single source
+file. This makes it relatively straightforward for a compiler to
+optimise. Its performance is thus quite a demanding target for our
+modified versions of the code to reproduce.
 
@@
 system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB. The CPU in the IBM BG/Q system is built around the
-PowerPC A2 core with a clock speed of 1.6~GHz. The chip has 32~MB of
+cache of 30~MB. The CPU in the IBM Power-8 system is built around the
+PowerPC xxx core with a clock speed of 4.1~GHz. The chip has xxx~MB of
 last-level cache.
@@
 \hline
-                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
+                 & Gnu   & Intel       & Cray    & IBM     \\
 \hline
@@
 Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
-IBM A2 (BGQ)     &       &             &         & 14.1     \\
-IBM Power 8      &       &             &         &          \\
+IBM Power 8      &       &             &         & 14.1     \\
 \hline
@@
 the \psykal versions of the code. The performance of the original
-version of the code is very much less sensitive to the compiler
+version of the code is much less sensitive to the compiler
 options. This is because it consists of a single source file
@@
 Cray      & -O3 -O ipa5 -h wp               \\
-IBM bgxlf & -O5 -qinline=auto:level=10    \\
+IBM       & \bf{CHECK} -O5 -qinline=auto:level=10      \\
 \hline
@@
 
-\begin{itemize}
+Before applying any code transformations, we benchmark the original
+version of the code. We also benchmark the vanilla, unoptimised
+version after it has been re-structured following the \psykal
+approach. These two versions of the code effectively provide upper and
+lower bounds, respectively, on the performance we expect to achieve.
 
-\item Performance of the 'original' version of shallow; i.e. the version we
-were supplied with at the start of this work.
+{\bf Performance analysis as motivation for the code
+  tranformations. More detail on each transformation.}
 
-\item Performance of the vanilla, unoptimised version after it has been
-re-structured following the \psykal approach.
-
-\end{itemize}
-
-Beginning with the vanilla \psykal version, we apply a series of
-optimisations while obeying the \psykal separation of concerns, {\it
-  i.e.} optimisation is restricted to the `middle layer' and leaves
+Beginning with the vanilla \psykal version, we then apply a series of
+code transformations while obeying the \psykal separation of concerns,
+{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
 the kernel and algorithm layers unchanged. The aim of these
@@
 thus performance of the original version of the code. The
-optimisations we have performed are as follows:
+transformations we have performed are as follows:
 \begin{enumerate}
@@
 beyond the scope of the present work.
+
+We explore the extent to which performance depends upon the problem
+size by using square domains of dimension 64, 128, 256, 512 and
+1024. This range allows us to investigate what happens when cache is
+exhausted as well as giving us some insight into the decisions that
+different compilers make when optimising the code.
 
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
+\caption{The mean time-per-step (s) for the $256^2$ case after each 
 code transformation.}
@@
 \hline
-Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
+Compiler:          &  Gnu  & Intel   & Gnu & Intel & Cray & IBM \\
 \hline
-CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
+CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & Power8 \\
 \hline
-Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
-Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
-Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
-Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
-Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
-In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
-Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
-In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
-Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
-Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
+Original                               &  7.47E-04 & 7.06E-04 &	7.70E-04 & 7.77E-04 & 6.13E-04 &  1.17E-03  \\
+Vanilla \psykal                        &  1.26E-02 & 8.64E-04 &	1.38E-02 & 9.82E-04 & 8.33E-04 &  7.87E-03  \\
+Explicitly-sized arrays                &  1.26E-02 & 8.62E-04 &	1.38E-02 & 9.65E-04 & 6.96E-04 &  6.74E-03  \\
+Kernels copied into module             &  1.10E-03 & 8.67E-04 &	9.42E-04 & 9.65E-04 & 7.10E-04 &  3.13E-03  \\
+Loops fused                            &  1.09E-03 & 7.89E-04 &	9.07E-04 & 8.80E-04 & 7.35E-04 &  3.60E-03  \\
+In-lined field-copy                    &  1.09E-03 & 7.90E-04 &	9.07E-04 & 8.73E-04 & 7.31E-04 &  3.62E-03  \\
+Fused field-copy                       &  1.09E-03 & 9.25E-04 &	1.02E-03 & 1.04E-03 & 6.50E-04 &  4.11E-03  \\
+In-line all kernels                    &  8.20E-04 & 7.14E-04 &	9.32E-04 & 8.52E-04 & 6.50E-04 &  1.07E-03  \\
+Fully fuse 1st loop nest               &  8.10E-04 & 7.18E-04 &	7.76E-04 & 7.92E-04 & 6.23E-04 &  1.07E-03  \\
+Explicit, constant array/loop bounds   &  8.05E-04 & 7.12E-04 &	7.81E-04 & 7.91E-04 & 6.13E-04 &  1.07E-03  \\
 \hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
+\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.8 & 1.8 & 0.03 & -9.0    \\
 \hline

### 2015-05-05 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add Mike to author list. Add plots of OMP performance for 32x32 domain.

@@
 
-\author{J.~Appleyard, R.~W.~Ford, J.~Holt, H.~Lieu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
+\author{J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Lieu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
 \thanks{J.~Appleyard is with NVIDIA.}%
 \thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
-\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
+\thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
 \thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
+  socket. The corresponding parallel efficiencies are shown using open
+  symbols and dashed lines. The 24-thread runs employed hyperthreading
+  and the optimal OpenMP schedule is given in parentheses.}
+\label{FIG_gnu_omp_32}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{omp_scaling_32_intel}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
+  socket.  The corresponding parallel efficiencies are shown using
+  open symbols and dashed lines. The 24-thread runs employed
+  hyperthreading and the optimal OpenMP schedule is given in
+  parentheses.}
+\label{FIG_intel_omp_32}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{omp_scaling_32_cray}
+\caption{Performance of the OpenMP-parallel version of \psykal
+  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
+  socket.  The corresponding parallel efficiencies are shown using
+  open symbols and dashed lines. The 24-thread runs employed
+  hyperthreading and the optimal OpenMP schedule is given in
+  parentheses.}
+\label{FIG_cray_omp_32}
+\end{figure}
+
 \subsubsection{OpenACC Version}

### 2015-05-06 nemolite2d_perf.tex
Add plot of effect of domain size on OpenMP scaling.

@@
 nemolite2d code; the introduction of OpenMP and OpenACC directives. In
-keeping with the \psykal approach, these transformations are
-restricted to the PSy (middle) layer of the code.
+keeping with the \psykal approach we do not modify either the
+Algorithm- or Kernel-layer code.  Any code changes are restricted to
+either the PSy (middle) layer or the underlying library that manages
+e.g. the construction of field objects.
 
@@
 Ivy Bridge socket which has 12 hardware cores and support for up to 24
-threads with hyperthreading.
+threads with hyperthreading (i.e. two threads per core).
 Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
@@
 compilers, respectively.
+
+In order to quantify the scaling behaviour of the different versions
+of nemolite2d with the different compilers/run-time environments, we
+also plot the parallel efficiency in
+Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
+and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
+parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
+\[
+E(n) = 100  \frac{P(n)}{ n P(1)}
+\]
+where $P(m)$ is the performance of the code on $m$ threads. For a
+perfect, linearly-scaling code, $E(n)$ will be 100\%.
+
+Since the space to explore consists of three different compilers, six
+different domain sizes, five stages of optimisation and six different
+thread counts we can only consider a slice through it in what follows.
+In order to inform our choice of domain size,
+Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
+most performant, Intel-compiled version of nemolite2d. Although it is
+common for production runs of NEMO to use MPI sub-domains of $20
+\times 20$ ({\bf do we have/need a reference for that?}), it is clear
+from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
+parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
+to fully test the performance of our OpenMP implementations we have
+chosen to examine results for the $256 \times 256$ domain and these
+are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
+and \ref{FIG_cray_omp_256}.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{omp_scaling_problem_size}
+\caption{The scaling behaviour of the most performant OpenMP-parallel
+  version of \psykal nemolite2d for the full range of domain sizes
+  considered on the CPU. Results are for the Intel compiler on a
+  single Intel Ivy Bridge socket. The corresponding parallel
+  efficiencies are shown using open symbols and dashed lines. The
+  24-thread runs employed hyperthreading.}
+\label{FIG_omp_domain_size}
+\end{figure}
 
@@
 
-The simplest OpenMP implementation (black lines) fails to scale beyond
-two threads for any of the compilers. However, as soon as a single
-PARALLEL region is introduced, the situation is greatly improved with
-all three executables now scaling out to at least 12 threads (red
-lines, square symbols). {\bf Add parallel efficiency lines to the OMP
-  scaling plots to highlight that things are still not brilliant.}
+The simplest OpenMP implementation (black lines, circle symbols) fails
+to scale well for any of the compilers. For the Intel and Gnu
+versions, parallel efficiency is already less than 50\% on just four
+threads. The Cray version however does better and is about 45\%
+efficient on eight threads.
+
+With the move to a single PARALLEL region, the situation
+is greatly improved with all three executables now scaling out to at
+least 12 threads with $\sim70$\% efficiency (red lines, square symbols).
 
@@
 triangles).
-
-In order to quantify the scaling behaviour of the different versions
-of nemolite2d with the different compilers/run-time environments, we
-also plot the parallel efficiency in
-Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
-and~\ref{FIG_cray_omp_256}. We define parallel efficiency, $E(n)$, on
-a number of threads, $n$, as:
-\begin{equation}
-E(n) = 100 * \frac{P(n)}{ n*P(1)}
-\end{equation}
-where $P(n)$ is the performance of the code on $n$ threads.
 
@@
 qualifiers.
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
+%\caption{Performance of the OpenMP-parallel version of \psykal
+%  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
+%  socket. The corresponding parallel efficiencies are shown using open
+%  symbols and dashed lines. The 24-thread runs employed hyperthreading
+%  and the optimal OpenMP schedule is given in parentheses.}
+%\label{FIG_gnu_omp_32}
+%\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=\picwidth]{omp_scaling_32_intel}
+%\caption{Performance of the OpenMP-parallel version of \psykal
+%  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
+%  socket.  The corresponding parallel efficiencies are shown using
+%  open symbols and dashed lines. The 24-thread runs employed
+%  hyperthreading and the optimal OpenMP schedule is given in
+%  parentheses.}
+%\label{FIG_intel_omp_32}
+%\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=\picwidth]{omp_scaling_32_cray}
+%\caption{Performance of the OpenMP-parallel version of \psykal
+%  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
+%  socket.  The corresponding parallel efficiencies are shown using
+%  open symbols and dashed lines. The 24-thread runs employed
+%  hyperthreading and the optimal OpenMP schedule is given in
+%  parentheses.}
+%\label{FIG_cray_omp_32}
+%\end{figure}
 

### 2015-05-06 shallow/Maxeler/CPUCode/GOceanCpuCode.c
Added input/output streams to LMEM for v and p

@@ -95,6 +95,8 @@ int main(void)
 	write_s.param_start = 0;
 	write_s.param_size  = lmem_size;
 	write_s.instream_cpu2lmem = unew;
+	write_s.instream_cpu2lmem_v = vnew;
+	write_s.instream_cpu2lmem_p = pnew;
 
 	GOcean_writeLMem_run(engine, &write_s);
 
@@ -105,7 +107,7 @@ int main(void)
 	// Read back from LMEM
 	printf("Reading %d bytes back from LMEM, sizeBytes\n", sizeBytes);
 	GOcean_readLMem_actions_t read_s;
-	GOcean_readLMem(lmem_size, 0, unew);
+	GOcean_readLMem(lmem_size, 0, unew, vnew, pnew);
 
 	max_unload(engine);
 	printf("Done.\n");

### 2015-06-08 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add text describing nemolite2d provided by Hedong

@@
 
-\author{J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Lieu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
+\author{J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Liu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
 \thanks{J.~Appleyard is with NVIDIA.}%
-\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
+\thanks{H.~Liu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
 \thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
@@
 \subsection{The `nemolite2d' Program}
-For this work we have used a  program
-
-{\bf Description of the structure of the program at the Algorithm
-  level so that it links in to Profile data in results section.}
+
+For this work we have used a program "nemolite2d" developed by
+ourselves.  “nemolite2d” is a vertically-averaged version of NEMO
+(Nucleus for European Modelling of the Ocean), which only keeps its
+dynamical part. The whole model system is represented through one
+continuity equation (for the update of the sea-surface height) and two
+vertically-integrated momentum equations (for the two velocity
+components, respectively).
+
+(add above three equations in .tex file).
+
+The external forcing includes surface wind stress, bottom friction,
+and open boundary barotropic forcing. A lateral slip boundary
+condition is applied along the coast lines. The open boundary
+conditions can be set as a clipped or a Flather's radiation
+condition (Flather, 1976). The bottom friction takes a semi-implicit
+form for the sake of model stability. As done in the original version
+of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
+used for the horizontal viscosity term.
+
+The traditional Arakawa C structured grid is employed here for the
+discretisation of the computational domain. A two-dimensional integer
+array is used to identify the different parts of the computational
+domain; it has the value of one for ocean, zero for land and minus
+one for ocean cells outside of the computational domain. This array
+serves as a mark to identify ocean cells, land cells, solid boundaries
+and open boundaries.
+
+For the sake of simplicity, the explicit Eulerian forward time
+stepping method is implemented here, except that the bottom friction
+takes a semi-implicit form.  The Coriolis force can be set in explicit
+or implicit form. The advection term is computed with a first-order
+upwind scheme.
+
+The order of the model computation is as follows:
+\begin{enumerate}
+\item Set the initial conditions (water depth, sea surface height, velocity)
+\item integrate the continuity equation for the new sea surface height
+\item update the different terms in the right hand side of the momentum equations (advection, Coriolis forcing(if set in explicit form), pressure gradient, and, horizontal viscosity
+\item update the velocity vectors by summing up the values in 3), and implicitly with the bottom friction and Coriolis forcing (if set in implicit form)
+\item apply the boundary conditions on to the open boundary and solid boundary cells.
+\end{enumerate}
 
@@
 perfect, linearly-scaling code, $E(n)$ will be 100\%.
+
+Since the space to explore consists of three different compilers, six
+different domain sizes, five stages of optimisation and six different
+thread counts we can only consider a slice through it in what follows.
+In order to inform our choice of domain size,
+Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
+most performant, Intel-compiled version of nemolite2d. Although it is
+common for production runs of NEMO to use MPI sub-domains of $20
+\times 20$ ({\bf do we have/need a reference for that?}), it is clear
+from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
+parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
+to fully test the performance of our OpenMP implementations we have
+chosen to examine results for the $256 \times 256$ domain and these
+are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
+and \ref{FIG_cray_omp_256}.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{omp_scaling_problem_size}
+\caption{The scaling behaviour of the most performant OpenMP-parallel
+  version of \psykal nemolite2d for the full range of domain sizes
+  considered on the CPU. Results are for the Intel compiler on a
+  single Intel Ivy Bridge socket. The corresponding parallel
+  efficiencies are shown using open symbols and dashed lines. The
+  24-thread runs employed hyperthreading.}
+\label{FIG_omp_domain_size}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
-\caption{Performance of the OpenMP-parallel version of \psykal
-  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-  socket. The corresponding parallel efficiencies are shown using open
-  symbols and dashed lines. The 24-thread runs employed hyperthreading
-  and the optimal OpenMP schedule is given in parentheses.}
-\label{FIG_gnu_omp_32}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_32_intel}
-\caption{Performance of the OpenMP-parallel version of \psykal
-  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
-  socket.  The corresponding parallel efficiencies are shown using
-  open symbols and dashed lines. The 24-thread runs employed
-  hyperthreading and the optimal OpenMP schedule is given in
-  parentheses.}
-\label{FIG_intel_omp_32}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_32_cray}
-\caption{Performance of the OpenMP-parallel version of \psykal
-  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
-  socket.  The corresponding parallel efficiencies are shown using
-  open symbols and dashed lines. The 24-thread runs employed
-  hyperthreading and the optimal OpenMP schedule is given in
-  parentheses.}
-\label{FIG_cray_omp_32}
-\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
+%\caption{Performance of the OpenMP-parallel version of \psykal
+%  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
+%  socket. The corresponding parallel efficiencies are shown using open
+%  symbols and dashed lines. The 24-thread runs employed hyperthreading
+%  and the optimal OpenMP schedule is given in parentheses.}
+%\label{FIG_gnu_omp_32}
+%\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=\picwidth]{omp_scaling_32_intel}
+%\caption{Performance of the OpenMP-parallel version of \psykal
+%  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
+%  socket.  The corresponding parallel efficiencies are shown using
+%  open symbols and dashed lines. The 24-thread runs employed
+%  hyperthreading and the optimal OpenMP schedule is given in
+%  parentheses.}
+%\label{FIG_intel_omp_32}
+%\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=\picwidth]{omp_scaling_32_cray}
+%\caption{Performance of the OpenMP-parallel version of \psykal
+%  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
+%  socket.  The corresponding parallel efficiencies are shown using
+%  open symbols and dashed lines. The 24-thread runs employed
+%  hyperthreading and the optimal OpenMP schedule is given in
+%  parentheses.}
+%\label{FIG_cray_omp_32}
+%\end{figure}
 

### 2015-06-30 parco2015_abstract/psyclone_parco2015.tex
Add src for ParCo2015 extended abstract

--- 2014-11-21 shallow_perf.tex
@@
-%% shallow_perf.tex
-%%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
+\documentclass{IOS-Book-Article}
 
 \usepackage{cite}
+\usepackage{mathptmx}
 
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
 \usepackage[pdftex]{graphicx}
@@
 \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.eps}
-\fi
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Do not use the stfloats baselinefloat ability as IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-\usepackage{dblfloatfix}
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
-% Axel Sommerfeldt. This package may be useful when used in conjunction with 
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
 
@@
 
+\def\hb{\hbox to 10.7 cm{}}
+
 \begin{document}
-%
-% paper title
+
+\pagestyle{headings}
+\def\thepage{}
+
+\begin{frontmatter}              % The preamble begins here.
+
 % can use linebreaks \\ within to get better formatting as desired
-\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)}
-%
-% author names and IEEE memberships
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
+\title{Towards Performance Portability in Earth-System Modelling with GOcean.}
 
-\author{A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
-\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
-\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
-}
+\markboth{}{November 2015\hb}
+%\subtitle{Subtitle}
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names.
+\author[A]{\fnms{A. R.} \snm{Porter}%
+\thanks{Corresponding Author: Andrew Porter, STFC Daresbury Laboratory, Warrington, WA4 4AD, UK; E-mail:
+andrew.porter@stfc.ac.uk.}},
+\author[A]{\fnms{R. W.} \snm{Ford}},
+\author[A]{\fnms{M.} \snm{Ashworth}},
+\author[A]{\fnms{G. D.} \snm{Riley}}
+and
+\author[B]{\fnms{M.} \snm{Modani}}
+\runningauthor{A.~R.~Porter et al.}
+\address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
+\address[B]{IBM}
 
-% The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
-The abstract goes here.
+This paper presents a domain specific approach to performance
+portability and code complexity reduction in finite element and finite
+difference codes. This approach has been developed for the UK Met
+Office's next generation atmospheric model which uses finite elements
+on a quasi-unform grid, and has also been prototyped on two finite
+difference Ocean model benchmarks, one of which is based on the NEMO
+ocean model. We first outline our approach and then present results
+for a range of compilers. These demonstrate that it is possible to
+retain and potentially improve upon the performance of an existing
+shallow water model while simultaneously reducing its code complexity.
 \end{abstract}
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
+\begin{keyword}
 Performance, Code-generation, Finite-difference
-\end{IEEEkeywords}
+\end{keyword}
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+\end{frontmatter}
+\markboth{November 2015\hb}{November 2015\hb}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HE" in caps to complete the first word.
-\IEEEPARstart{T}{he} challenge presented to the developers of
-scientific software by the drive towards Exascale computing is
-considerable. With power consumption becoming the overriding design
-constraint, CPU clock speeds are falling and the complex,
-multi-purpose compute core is being replaced by multiple, simpler
-cores. This philosophy can be seen at work in the rise of so-called
-accelerator based machines in the Top 500 List~\cite{top500} of
-supercomputers: five of the top-ten machines in the November 2014 list
-make use of Intel Xeon Phi's or NVIDIA GPUs. Four of the remaining
-five machines in the top ten are IBM BlueGene/Qs, the CPU of which has
-hardware support for running 64 threads.
+\section*{Introduction}
+
+The challenge presented to the developers of scientific software by
+the drive towards Exascale computing is considerable. With power
+consumption becoming the overriding design constraint, CPU clock
+speeds are falling and the complex, multi-purpose compute core is
+being replaced by multiple, simpler cores. This philosophy can be seen
+at work in the rise of so-called accelerator based machines in the Top
+500 List~\cite{top500} of supercomputers: five of the top-ten machines
+in the November 2014 list make use of Intel Xeon Phi's or NVIDIA
+GPUs. Four of the remaining five machines in the top ten are IBM
+BlueGene/Qs, the CPU of which has hardware support for running 64
+threads.
 
@@
 
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
+\subsection{The \psykal Approach}
 
-\subsection{The \psykal Approach}
-Proliferation of lightweight cores is resulting in increased
-complexity in the programming models required to achieve performance.
-Combine with the complexity of e.g. a model simulating circulation in
-the global ocean. Significant challenge in software engineering.
-Propose an approach for meeting this challenge by separating the
-computational science demands from the {\it e.g.} oceanographic
-demands.
+The \psykal approach separates code into three layers, the Algorithm
+layer, the PSy layer and the Kernel layer. Whilst this approach is
+general we have applied it to Atmosphere and Ocean models written in
+Fortran where domain decomposition is typically performed in the
+latitude-longitude direction, leaving columns of elements on each
+domain-decomposed partition.
 
-\begin{figure}
-\centering
-\caption{The \psykal Approach.}
-\end{figure}
+The top layer, in terms of calling hierarchy, is the Algorithm
+layer. This layer specifies the algorithm that thescientist would like
+to perform (in terms of calls to kernel and infrastructure routines)
+and logically operates on full fields. We say logically here as the
+fields may be domain decomposed, however the algorithm layer is not
+aware of this. It is the scientists responsibility to write this
+algorithm layer.
 
-While encapsulation makes it easier to produce more robust software,
-it can often come with a performance penalty. In this work we
-investigate in detail the hardware and compiler dependencies of this
-penalty and the steps required to remove it.
+The bottom layer, in terms of calling hierarchy, is the Kernel
+layer. The Kernel layer implements the science that the Algorithm
+layer calls, as a set of subroutines. These kernels operate on local
+fields (a set of elements, a single column of elements, or a set of
+columns, depending on the kernel). Again the scientist is responsible
+for writing this layer and there is no parallelism specified here, but
+there is likely to be input from an HPC expert and/or some coding
+rules to help make sure the kernels compile into efficient code.
+
+The PSy layer sits in-between the Algorithm and Kernel layers and its
+functional role is to link the algorithm calls to the associated
+kernel subroutines. As the Algorithm layer works on logically global
+fields and Kernel layer works on local fields the PSy layer is
+responsible for iterating over columns. It is also responsible for
+including any required distributed memory operations, such as halo
+swaps and reductions.
+
+As the PSy layer iterates over columns, the potential parallelism
+within this iteration space can be optimised and parallelised. The PSy
+layer can therefore be optimised for a particular hardware
+architecture, such as multi-core, many-core, GPGPUs, or some
+combination thereof with no change to the algorithm or kernel layer
+code. This approach therefore offers the potential for portable
+performance.
+
+As an example, consider the following ``traditional'' code fragment.
+\begin{verbatim}
+!$OMP PARALLEL DO ...
+do i = 1, nlat
+  do j = 1, nlon
+      do k = 1, levels
+      a(k,j,i) = ...
+      b(k,j,i) = ...
+    end do
+end do
+end do
+!$END OMP PARALLEL DO ...
+\end{verbatim}
+In the PSyKAl approach this code would be split into algorithm layer\ldots
+\begin{verbatim}
+call psy_...(a,b,...)
+\end{verbatim}
+PSy layer\ldots
+\begin{verbatim}
+subroutine psy_...(a,b,...)
+  !$OMP PARALLEL DO ...
+  do i = 1, nlat
+    do j = 1, nlon
+      call kern1(a,...)
+      call kern2(b,...)
+    end do
+  end do
+  !$END OMP PARALLEL DO
+end subroutine psy_...
+\end{verbatim}
+and kernel layer\ldots
+\begin{verbatim}
+subroutine kern1(a,...)
+  do k = 1, levels
+    a(k) = ...
+  end do
+end subroutine kern1
+subroutine kern2(b,...)
+  do k = 1, levels
+    b(k) = ...
+  end do
+end subroutine kern2
+\end{verbatim}
+
+Therefore parallelism is encapsulated in the PSy layer and the
+latitude-longitude iteration space can be parallelised in different
+ways; the directives could be changed, the loops could be split, halo
+calls could be added etc. None of these changes would require
+modification to the Algorithm or Kernel layers.
+
+One difference can be observed here. As kernels contain whole columns,
+it is not possible to manually merge the ``k'' loops into one (which is
+what was implemented in the original example code fragment). Such a
+modification would have to be performed by the compiler, if the
+compiler considered it appropriate.
+
+Clearly the splitting of code into separate layers will have an effect
+on performance. This overhead and how to get back to the performance
+of the ``traditional'' code, and potentially improve on it, will be
+discussed in the talk.
 
@@
 \centering
-\includegraphics[width=85mm]{psykal_shallow}
+\includegraphics[width=85mm]{../psykal_shallow}
 \caption{A schematic of the vanilla \psykal version of the shallow code.}
@@
 
-As with any full oceanographic model, boundary conditions must be
-applied at the edges of the model domain. In the case of Shallow we
-have periodic boundary conditions in both the $x$ and $y$ dimensions.
-These are simply implemented by having additional (`halo')
-rows/columns at the edges of the domain and ensuring the data in them
-is up-to-date before it is read. These rows/columns are updated by
-copying data from the corresponding row/column on the opposite side of
-the domain. In the current work, the lines of code to do these copies
-is manually inserted in the PSy-layer routine as required.
-
-{\bf Auto-generated BC code by reasoning about meta-data. But then,
-  that brings us on to meta-data...}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methodology(?)}
+\section{Methodology}
 
@@
 piece of hardware, the version number used in this work is specified.
-The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
-3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
-system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB. The CPU in the IBM BG/Q system is built around the
-PowerPC A2 core with a clock speed of 1.6~GHz. The chip has 32~MB of
-last-level cache.
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
 \begin{table}[!t]
@@
 \hline
-                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
+       CPU       & Gnu   & Intel       & Cray    & IBM     \\
 \hline
-Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
-Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
-IBM A2 (BGQ)     &       &             &         & 14.1     \\
-IBM Power 8      &       &             &         &          \\
-\hline
-\end{tabular}
-\end{table}
-
-In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
-used with each compiler. These flags are particularly important for
-the \psykal versions of the code. The performance of the original
-version of the code is much less sensitive to the compiler
-options. This is because it consists of a single source file
-containing effectively a single routine (excluding those for I/O).
-
-\begin{table}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-\caption{The compiler flags used in this work.}
-\label{TABLE_compiler_flags}
-\centering
-\begin{tabular}{l|c}
-\hline
-Compiler  &  Flags \\
-\hline
-Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
-Intel     & -O3 -fast -fno-inline-factor    \\
-Cray      & -O3 -O ipa5 -h wp               \\
-IBM bgxlf & -O5 -qinline=auto:level=10      \\
+Intel Haswell (E5-1620 v2)   & 4.8.2 & 14.0.0      &         & \\
+Intel Ivy Bridge  (E5-2697 v2) & 4.9.1 & 14.0.1.106  & 8.3.3 & \\
+IBM Power 8      &       &             &             & 14.1    \\
 \hline
@@
 code transformations while obeying the \psykal separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
-the kernel and algorithm layers unchanged. The aim of these
+{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
+leaves the kernel and algorithm layers unchanged. The aim of these
 optimisations is to recover, as much as is possible, the structure and
-thus performance of the original version of the code. The
-transformations we have performed are as follows:
-\begin{enumerate}
+thus performance of the original version of the code.
 
-\item Pass explicit array bounds to middle layer (rather than using
-  assumed-size array arguments)
-
-\item Move kernel routines into same module as middle layer
-
-\item Loop fusion
-
-\item In-lined field-copy operation
-
-\item Fused field-copy operation
-
-\item Manually in-line kernel bodies into middle-layer code
-
-\item Fully fuse first loop (only the outer loop of this loop nest was
-  fused in the first pass)
-
-\item Make the array/loop bounds explicitly constant for the
-      duration of the time-stepping loop.
-\end{enumerate}
-
-We shall see that these transformations do not always result in improved
+We shall see that these optimisations do not always result in improved
 performance. Whether or not they do so depends both on the compiler
@@
 
-In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
-original version of the shallow code for the range of compilers and
-hardware considered here. This summary demonstrates the effect of the
-larger last-level cache of the Intel Ivy Bridge system compared to
-that of the Haswell system; note the significant drop in performance
-in going from a problem size of $256^{2}$ to a problem size of
-$512^{2}$ for the black and green bars. The performance of the Ivy
-Bridge system with the Cray or Intel compiler (red and purple bars)
-only drops to this level when the domain size is increased to
-$1024^{2}$. At this point, the working set no longer fits within cache
-and performance is dominated by the bandwidth to main memory, making
-it relatively insensitive to the choice of compiler.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=2.8in]{orig_summary}
-\caption{Summary of the performance of the original shallow code on 
-the range of compilers and hardware under consideration.}
-\label{FIG_orig_perf_summary}
-\end{figure}
-
-For this compiler-friendly form of the code, there is little
-performance difference between the executables produced by the Gnu and
-Intel compilers. However, the executable produced by the Cray compiler
-generally performs significantly better. {\bf Why is this; can we work
-out what it's doing that the others aren't?}
-
-Note that the results for the Gnu compiler on Ivy Bridge show the same
-drop-off (as problem size is increased) as those on the Haswell
-system. This may be because Archer's login nodes seem to be running on
-older hardware (E5-2650 with 20MB LLC) during the Archer upgrade. {\bf
-  Need to re-run these cases once Archer is back up.}
-
-Moving now the to the \psykal version of shallow,
-Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest \psykal version for each of the compiler/CPU combinations. The
-only significant difference between this and the plot of the
-performance of the original version of the code in
-Figure~\ref{FIG_orig_perf_summary} is that the difference between the
-Gnu- and Intel-compiled executables is slightly more marked. This is
-clarified in Figure~\ref{FIG_slowdown_summary} which plots the
-percentage difference between the performance of the original and the
-(best) \psykal versions of shallow for each compiler/CPU combination.
-It is clear from this plot that the difference in the performance of
-the Gnu- and Intel-compiled excutables on the Haswell system is due to
-the Gnu compiler being unable to recover the performance achieved by
-the original code. In contrast there is no such problem for the
+In Figure~\ref{FIG_slowdown_summary} we plot the percentage difference
+between the performance of the original and the (best) \psykal
+versions of Shallow for each compiler/CPU combination.  It is clear
+from this plot that the difference in the performance of the Gnu- and
+Intel-compiled excutables on the Haswell system is due to the Gnu
+compiler being unable to recover the performance achieved by the
+original code. In contrast there is no such problem for the
 Gnu-compiled executable on the Ivy Bridge system. We attribute this to
@@
 \centering
-\includegraphics[width=7.5cm]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any \psykal 
-version of shallow with each of the compilers and CPUs under 
-consideration.}
-\label{FIG_best_psykal_perf_summary}
-\end{figure}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=2.8in]{slowdown_summary}
+\includegraphics[width=2.8in]{../slowdown_summary}
 \caption{Comparison of the performance of the best \psykal
@@
 
-Having show that, in general, we can recover and occasionally improve
-upon the performance of the original version of shallow, the next
-logical step is to examine the necessary code transformations.  We do
-this for the $256^{2}$ case since this fits within cache on both the
-Haswell and Ivy Bridge CPUs we are using here.
-Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
-after each transformation has been applied to the code. The same data
-is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
-of the \psykal version for a given compiler/CPU is given relative to
-the performance of the original with the same compiler/CPU.
+Having show that, in general, we can recover the performance of the
+original code we next examine the code transformations that are required
+(and must therefore be supported by the code-generation system).
 
-\begin{table*}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
-code transformation.}
-\label{TABLE_opt_breakdown}
-\centering
-\begin{tabular}{l|c|c|c|c|c|c}
-\hline
-Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
-\hline
-CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
-\hline
-Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
-Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
-Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
-Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
-Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
-In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
-Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
-In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
-Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
-Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
-\hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
-\hline
-\end{tabular}
-\end{table*}
-
-With the Gnu compiler (first and third clusters in
-Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
-shallow only achieves ~5-6\% of the performance of the
-original. Simply copying the code for each of the kernel subroutines
-into the module from which they are called has the most dramatic
-effect on the performance of the compiled code; it now achieves almost
-80\% of the performance of the original. The next most significant
-transformation is that of (manually) in-lining the kernel code itself
-into the bodies of the loops from which they are called. This gets the
-\psykal version to within 8\% of the performance of the original.  The
-dramatic performance improvement obtained from the first
-transformation demonstrates that the Gnu compiler is unable to
-optimise over separate files. Therefore, the fact that the
-time-stepping loop in the \psykal version still retains a call to an
-I/O routine and a call to the middle-layer, both of which are in
-different files, is likely to account for this remaining performance
-penalty of 8\%.
-
-With the Intel compiler, the key transformations to recover
-performance are different (second and fourth clusters in
-Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
-shows that even the vanilla \psykal version of shallow achieves some
-80\% of the performance of the original version. Only two code
-transformations were required to increase this to 100\%. The first of
-these was the fusion of the computational loops. However, we found
-that performance suffered (purple bar) when this was extended to
-include the loop required for field copies at the end of a time-step
-(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
-slow-down revealed that the compiler was unable to SIMD vectorise the
-newly-fused loop after it had in-lined the body of the kernel called
-from within it. Once this in-lining was done by hand (dark-blue bar),
-the compiler was able to vectorise the loop and the performance of the
-original version of shallow was recovered.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=95mm]{opt_stages_256}
-\caption{The performance of the \psykal version of shallow for the
-  $256^{2}$ domain at each stage of optimisation. Results are given as
-  a percentage of the performance of the original code for each
-  compiler/CPU combination. A figure greater than 100\% indicates that
-  the \psykal version performs better than the original.}
-\label{FIG_opt_stages_256}
-\end{figure}
-
-\begin{figure}
-\begin{verbatim}
-! Time smoothing
-DO J=1,N+1
-  DO I=1,M+1
-    CALL time_smooth_code(i,j,ufld,unew,uold)
-!    uold(i,j) = ufld(i,j) + alpha* &
-!     (unew(i,j) - 2.*ufld(i,j) + uold(i,j))
-  END DO
-END DO
-
-! Copy new fields into current fields
-DO J=1,N+1
-  DO I=1,M+1
-    Ufld(I,J) = UNEW(I,J)
-  END DO
-END DO
-\end{verbatim}
-\caption{Example of the coding structure that required manual
-  in-lining of the kernel body (as indicated by commented-out lines)
-  to retrieve the performance of the original version of shallow with
-  the Intel compiler.}
-\label{FIG_time_smooth_code}
-\end{figure}
-
-The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
-evolution of the performance of the \psykal version of shallow with
-the Cray compiler. Again, this has significant differences from the
-behaviour seen with the Gnu and Intel compilers. As with the Intel
-compiler, the performance of the Vanilla \psykal version is fairly
-good at 74\% of that of the original. In contrast to the other
-compilers consisdered so far, the first significant transformation is
-simply to specify the bounds of the arrays being used within the
-middle layer using module variables (as opposed to specifying them as
-assumed size). This tells the compiler that all of the arrays used in
-the middle layer have the same extent and also that all of the
-computational loops are over almost all elements of these arrays. This
-simple change gives a \psykal version that achieves 88\% of the
-performance of the original.
-
-Subsequent transformations actually harm performance until the
-field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
-with the preceeding loop. The resulting \psykal version now achieves
-94\% of the original performance. Two further steps are required to
-match the latter. First, the first loop nest has to be fully fused
-(both inner and outer loops fused) and second, the array/loop bounds
-are passed as an argument to the middle layer. This is done in order
-to specify to the compiler that these bounds remain constant for the
-duration of the time-stepping loop.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
-
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}
@@
 natural (oceanographic) science aspects. In this work we have shown
-that the \psykal approach has the potential to improve performance and
-to improve code structure/maintainability.
+that the \psykal approach has the potential to at least maintain (if
+not improve) performance while improving code
+structure/maintainability.
 
 The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal version of a code.
+to the performance of the \psykal version of a code. In our approach
+this problem is tackled using the PSyclone code-generation system
+which will be described in a future publication.
 
-% use section* for acknowledgement
-\section*{Acknowledgments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Acknowledgments}
 
@@
 
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,shallow_perf}
-%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliography{../shallow_perf}{}
+\bibliographystyle{unsrt}
 
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiographynophoto}{A.~R.~Porter}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{R.~Ford}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiographynophoto}{G.~Riley}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}

### 2015-08-04 docs/papers/shallow_performance/shallow_perf.tex
Add updated P8 results and some more discussion of them

@@
 \thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
-\thanks{M.~Modani is with IBM.}
+\thanks{M.~Modani is with IBM India.}
 }
@@
 
-\subsection{The `shallow' Program}
+\subsection{The `Shallow' Program}
 For this work we have used a benchmarking program that solves the
@@
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal version of shallow retains the Input/Output
+we ensure that any \psykal version of Shallow retains the Input/Output
 capability of the original. This aids in limiting the optimisations
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of shallow, we made the kernels
+optimisation) of the \psykal version of Shallow, we made the kernels
 as fine-grained as possible. In this case, this resulted in eight
@@
 \includegraphics[width=85mm]{psykal_shallow}
-\caption{A schematic of the vanilla \psykal version of the shallow code.}
+\caption{A schematic of the vanilla \psykal version of the Shallow code.}
 \label{FIG_psykal_shallow_structure}
@@
 system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB. The CPU in the IBM Power-8 system is built around the
-PowerPC xxx core with a clock speed of 4.1~GHz. The chip has xxx~MB of
-last-level cache.
+cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
+'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
+512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
+is shared with all other chiplets so that there is a total of 96~MB of
+last-level cache. The cores on the system we used had a clock speed of
+2~GHz.
 
@@
 Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
-IBM Power 8      &       &             &         & 14.1     \\
+IBM Power 8      &       &             &         & 15.1.2     \\
 \hline
@@
 \centering
-\begin{tabular}{l|c}
+\begin{tabular}{l|l}
 \hline
@@
 Cray      & -O3 -O ipa5 -h wp               \\
-IBM       & \bf{CHECK} -O5 -qinline=auto:level=10      \\
+IBM       & \parbox{5cm}{-O5 -qinline=auto:level=10\\ -qinline=auto:level=10 -qaltivec=be\\ -qdebug=NSIMDUBTHRESHOLD -qprefetch=dscr=0x1D7} \\
 \hline
@@
 In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
-original version of the shallow code for the range of compilers and
+original version of the Shallow code for the range of compilers and
 hardware considered here. This summary demonstrates the effect of the
@@
 and performance is dominated by the bandwidth to main memory, making
-it relatively insensitive to the choice of compiler.
+it relatively insensitive to the choice of compiler. The IBM Fortran
+compiler for Linux on Power 8 is relatively immature and failed to
+SIMD-vectorise two of the three major loop nests in the time-stepping
+loop. Figure~\ref{FIG_orig_perf_summary} therefore also includes the
+performance results obtained after the source had been modified to
+work-around this limitation (we fissioned one loop nest and manually
+unrolled another). Work is currently underway on updating the compiler
+so that these modifications will be unecessary with a future release.
 
@@
 \includegraphics[width=2.8in]{orig_summary}
-\caption{Summary of the performance of the original shallow code on 
-the range of compilers and hardware under consideration.}
+\caption{Summary of the performance of the original Shallow code on
+  the range of compilers and hardware under consideration. *After code
+  modifications to encourage SIMD vectorisation (see text).}
 \label{FIG_orig_perf_summary}
@@
 Intel compilers. However, the executable produced by the Cray compiler
-generally performs significantly better. {\bf Why is this; can we work
-out what it's doing that the others aren't?}
+generally performs significantly better for all but the largest domain
+size.
 
-Note that the results for the Gnu compiler on Ivy Bridge show the same
-drop-off (as problem size is increased) as those on the Haswell
-system. This may be because Archer's login nodes seem to be running on
-older hardware (E5-2650 with 20MB LLC) during the Archer upgrade. {\bf
-  Need to re-run these cases once Archer is back up.}
-
-Moving now the to the \psykal version of shallow,
+Moving now the to the \psykal version of Shallow,
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
@@
 percentage difference between the performance of the original and the
-(best) \psykal versions of shallow for each compiler/CPU combination.
+(best) \psykal versions of Shallow for each compiler/CPU combination.
 It is clear from this plot that the difference in the performance of
@@
 \caption{Summary of the best performance achieved by any \psykal 
-version of shallow with each of the compilers and CPUs under 
+version of Shallow with each of the compilers and CPUs under 
 consideration.}
@@
 In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
-\psykal version of shallow is never more than 2\% slower than the
+\psykal version of Shallow is never more than 2\% slower than the
 original and in some cases is faster.  This demonstrates that we can
@@
 
-Having show that, in general, we can recover and occasionally improve
-upon the performance of the original version of shallow, the next
+The most striking feature of Figure~\ref{FIG_slowdown_summary} is the
+fact that the best \psykal version of Shallow is consistently faster
+than the original on the Power 8. This confirms that this early
+version of the IBM compiler struggled with the original form of the
+Shallow source code.
+
+Having shown that, in general, we can recover and occasionally improve
+upon the performance of the original version of Shallow, the next
 logical step is to examine the necessary code transformations.  We do
-this for the $256^{2}$ case since this fits within cache on both the
-Haswell and Ivy Bridge CPUs we are using here.
-Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
-after each transformation has been applied to the code. The same data
-is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
-of the \psykal version for a given compiler/CPU is given relative to
-the performance of the original with the same compiler/CPU.
+this for the $256^{2}$ case since this fits within last-level cache on
+all of the CPUs we are using here.  Table~\ref{TABLE_opt_breakdown}
+shows detailed timings for this case after each transformation has
+been applied to the code. The same data is visualised in
+Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
+version for a given compiler/CPU is given relative to the performance
+of the original with the same compiler/CPU.
 
@@
 Explicit, constant array/loop bounds   &  8.05E-04 & 7.12E-04 &	7.81E-04 & 7.91E-04 & 6.13E-04 &  1.07E-03  \\
+Fission new-field update loops         &           &          &          &          &          &            \\
+Un-roll time-smoothing                 &           &          &          &          &          &            \\
 \hline
@@
 Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
-shallow only achieves ~5-6\% of the performance of the
+Shallow only achieves ~5-6\% of the performance of the
 original. Simply copying the code for each of the kernel subroutines
-into the module from which they are called has the most dramatic
-effect on the performance of the compiled code; it now achieves almost
-80\% of the performance of the original. The next most significant
-transformation is that of (manually) in-lining the kernel code itself
-into the bodies of the loops from which they are called. This gets the
-\psykal version to within 8\% of the performance of the original.  The
-dramatic performance improvement obtained from the first
-transformation demonstrates that the Gnu compiler is unable to
-optimise over separate files. Therefore, the fact that the
-time-stepping loop in the \psykal version still retains a call to an
-I/O routine and a call to the middle-layer, both of which are in
-different files, is likely to account for this remaining performance
-penalty of 8\%.
+into the module from which they are called ('module in-lining') has
+the most dramatic effect on the performance of the compiled code; it
+now achieves almost 80\% of the performance of the original. The next
+most significant transformation is that of manually in-lining the
+kernel code itself into the bodies of the loops from which they are
+called ('kernel in-lining'). This gets the \psykal version to within
+8\% of the performance of the original.  The dramatic performance
+improvement obtained from the first transformation demonstrates that
+the Gnu compiler is unable to optimise over separate files. Therefore,
+the fact that the time-stepping loop in the \psykal version still
+retains a call to an I/O routine and a call to the middle-layer, both
+of which are in different files, is likely to account for this
+remaining performance penalty of 8\%.
 
@@
 Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
-shows that even the vanilla \psykal version of shallow achieves some
+shows that even the vanilla \psykal version of Shallow achieves some
 80\% of the performance of the original version. Only two code
@@
 the compiler was able to vectorise the loop and the performance of the
-original version of shallow was recovered.
+original version of Shallow was recovered.
 
@@
 \includegraphics[width=95mm]{opt_stages_256}
-\caption{The performance of the \psykal version of shallow for the
+\caption{The performance of the \psykal version of Shallow for the
   $256^{2}$ domain at each stage of optimisation. Results are given as
@@
   in-lining of the kernel body (as indicated by commented-out lines)
-  to retrieve the performance of the original version of shallow with
+  to retrieve the performance of the original version of Shallow with
   the Intel compiler.}
@@
 The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
-evolution of the performance of the \psykal version of shallow with
+evolution of the performance of the \psykal version of Shallow with
 the Cray compiler. Again, this has significant differences from the
@@
 
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
+As with the Gnu compiler, it is module in-lining and kernel in-lining
+that are important for the IBM compiler (sixth cluster in
+Figure~\ref{FIG_opt_stages_256}). Fusing loop nests actually harms
+performance, just as it did for the Cray compiler.
 
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
+The take-home message of all this is that while we can employ a
+separation of concerns and recover performance, doing so is not
+straightforward. Determining how best to optimise even a simple code
+such as Shallow is highly compiler- and CPU-dependent. Decisions that
+are good on one platform may be bad for another and if these are
+written into the source code then they will accumulate over time and
+will almost certainly result in a code that does not perform optimally
+on any system.
 
@@
 to the performance of the \psykal version of a code.
+
+Some of our findings have been sent to the IBM compiler team and will
+serve to improve future releases of the compiler.
 

### 2015-08-05 SEQ/time_step_mod.f90
Fix errors in in-lined unew and vnew calc. Manually un-roll outer loop around time-smoothing for P8

@@ -1,6 +1,5 @@
 module time_step_mod
   use kind_params_mod
-!  use topology_mod, only: M, N
   use timing_mod
   implicit none
 
@@ -93,16 +92,16 @@ contains
 
           !CALL compute_unew_code(i+1, j, unew, uold, &
           !                       zfld, cvfld, hfld, tdt)
-          UNEW(I,J) = UOLD(I,J) +                                 &
-                      TDTS8*(Zfld(I,J+1)+Zfld(I,J)) *                   &
-                      (CVfld(I,J+1)+CVfld(I-1,J+1)+CVfld(I-1,J)+CVfld(I,J)) - &
-                       TDTSDX*(Hfld(I,J)-Hfld(I-1,J))
+          UNEW(I+1,J) = UOLD(I+1,J) +                                         &
+                      TDTS8*(Zfld(I+1,J+1)+Zfld(I+1,J)) *                     &
+                      (CVfld(I+1,J+1)+CVfld(I,J+1)+CVfld(I,J)+CVfld(I+1,J)) - &
+                       TDTSDX*(Hfld(I+1,J)-Hfld(I,J))
 
           !CALL compute_vnew_code(i, j+1, vnew, vold, &
           !                       zfld, cufld, hfld, tdt)
-          VNEW(I,J) = VOLD(I,J)-TDTS8*(Zfld(I+1,J)+Zfld(I,J))           &
-                      *(CUfld(I+1,J)+CUfld(I,J)+CUfld(I,J-1)+CUfld(I+1,J-1)) &
-                      -TDTSDY*(Hfld(I,J)-Hfld(I,J-1))
+          VNEW(I,J+1) = VOLD(I,J+1)-TDTS8*(Zfld(I+1,J+1)+Zfld(I,J+1))        &
+                      *(CUfld(I+1,J+1)+CUfld(I,J+1)+CUfld(I,J)+CUfld(I+1,J)) &
+                      -TDTSDY*(Hfld(I,J+1)-Hfld(I,J))
 
           !CALL compute_pnew_code(i, j, pnew, pold, &
           !                       cufld, cvfld, tdt)
@@ -135,8 +134,62 @@ contains
[…]

### 2015-08-05 shallow_perf.tex
Add a todo re performance analysis
Added an abstract and changed the author list appropriately

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)}
+\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes}
 %
@@
 
-\author{A.~R.~Porter, R.~Ford and G.~Riley% <-this % stops a space
-\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
-\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
+\author{ A.~R.~Porter, M.~Ashworth, R.~W.~Ford, M.~Modani, G.~D.~Riley% <-this % stops a space
+\thanks{A.~R.~Porter, M.~Ashworth, R.~W.~Ford and M.~Modani work in the Science and Technology Facilities Council's Daresbury Laboratory, UK.}% <-this % stops a space
+\thanks{G.~D.~Riley works in the School of Computer Science at the University of Manchester, Manchester, UK.}
 }
@@
 \begin{abstract}
-The abstract goes here.
+In this paper we evaluate the performance implications of applying a
+technique called \psykal to the finite difference domain. 
+%
+In \psykal, the code related to the underlying science is formally
+separated from code related to parallelisation and single core
+optimisations. This separation of concerns allows scientists to code
+their science independently of the underlying hardware architecture
+(thereby keeping a single code base) and for optimisation specialists
+to be able to tailor the code for a particular machine independently
+of the science code.
+%
+A finite difference shallow water benchmark optimised for cache based
+architectures is taken as the starting point. A vanilla \psykal version
+is written and the performance of the two compared. The optimisations
+that were applied to the original benchmark (loop fusion etc) are then
+applied to the \psykal version as a set of code modifications to the
+optimisation layer. Performance results are presented for the Cray,
+Intel and GNU compilers on Intel Ivybridge and Haswell processors and
+for the IBM compiler on Power8.
+%
+Results show that the combined set of code modifications obtain
+performance that is within a few percent of the original code for all
+compiler and architecture combinations on all tested problem sizes
+apart from GNU on Haswell and IBM on Power8.
+%
+As the GNU compiler performs well on Ivybridge the reduction in
+performance for the \psykal version was not analysed further.
+%
+In the case of the IBM compiler, the \psykal performance is
+significantly better than the original hand optimised code. Analysing
+the performance of the IBM compiler shows that a different structure
+for the original optimised code produces much improved performance and
+this performance is also reproduced using the \psykal version using
+appropriate transformations.
+%
+Therefore the \psykal approach can be used with negligable performance
+loss compared to the original optimised code, there is not a single
+best implementation of the code for all compilers tested, and the
+flexibility in applying code modifications to the optimisation layer
+allows for a significant performance improvement for the IBM Power8
+compared with the original code.
 \end{abstract}
@@
 system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB. The CPU in the IBM BG/Q system is built around the
-PowerPC A2 core with a clock speed of 1.6~GHz. The chip has 32~MB of
-last-level cache.
+cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
+'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
+512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
+is shared with all other chiplets so that there is a total of 96~MB of
+last-level cache. The cores on the system we used had a clock speed of
+3.3~GHz.
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
 \begin{table}[!t]
@@
 Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
-IBM A2 (BGQ)     &       &             &         & 14.1     \\
-IBM Power 8      &       &             &         &          \\
+IBM Power 8      &       &             &         & 15.1.2     \\
 \hline
@@
 Cray      & -O3 -O ipa5 -h wp               \\
-IBM bgxlf & -O5 -qinline=auto:level=10      \\
+IBM       & \parbox{5cm}{-O5 -qinline=auto:level=10\\ -qinline=auto:level=10 -qaltivec=be\\ -qdebug=NSIMDUBTHRESHOLD -qprefetch=dscr=0x1D7} \\
 \hline
@@
 lower bounds, respectively, on the performance we expect to achieve.
+
+{\bf Performance analysis as motivation for the code
+  tranformations. More detail on each transformation.}
 
@@
 \hline
-Compiler:          &  Gnu    & Intel   & Gnu & Intel & Cray & IBM bgxlf \\
+Compiler:          &  Gnu  & Intel   & Gnu & Intel & Cray & IBM \\
 \hline
-CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
+CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & Power8 \\
 \hline
-Original                             & 1.49 & 1.41 & 1.54 & 1.57 & 1.23 &        \\
-Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.96 & 1.67 &        \\
-Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.93 & 1.39 & 113.74 \\
-Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.93 & 1.42 &  68.28 \\
-Loops fused                          & 2.18 & 1.58 & 1.79 & 1.76 & 1.47 &  71.91 \\
-In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.75 & 1.46 &  83.36 \\
-Fused field-copy                     & 2.18 & 1.85 & 1.99 & 2.07 & 1.30 &  72.62 \\
-In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.59 & 1.30 &        \\
-Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.55 & 1.25 &        \\
-Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.54 & 1.23 &        \\
+Original                               &  7.47E-04 & 7.06E-04 &	7.70E-04 & 7.77E-04 & 6.13E-04 &  1.17E-03  \\
+Vanilla \psykal                        &  1.26E-02 & 8.64E-04 &	1.38E-02 & 9.82E-04 & 8.33E-04 &  7.87E-03  \\
+Explicitly-sized arrays                &  1.26E-02 & 8.62E-04 &	1.38E-02 & 9.65E-04 & 6.96E-04 &  6.74E-03  \\
+Kernels copied into module             &  1.10E-03 & 8.67E-04 &	9.42E-04 & 9.65E-04 & 7.10E-04 &  3.13E-03  \\
+Loops fused                            &  1.09E-03 & 7.89E-04 &	9.07E-04 & 8.80E-04 & 7.35E-04 &  3.60E-03  \\
+In-lined field-copy                    &  1.09E-03 & 7.90E-04 &	9.07E-04 & 8.73E-04 & 7.31E-04 &  3.62E-03  \\
+Fused field-copy                       &  1.09E-03 & 9.25E-04 &	1.02E-03 & 1.04E-03 & 6.50E-04 &  4.11E-03  \\
+In-line all kernels                    &  8.20E-04 & 7.14E-04 &	9.32E-04 & 8.52E-04 & 6.50E-04 &  1.07E-03  \\
+Fully fuse 1st loop nest               &  8.10E-04 & 7.18E-04 &	7.76E-04 & 7.92E-04 & 6.23E-04 &  1.07E-03  \\
+Explicit, constant array/loop bounds   &  8.05E-04 & 7.12E-04 &	7.81E-04 & 7.91E-04 & 6.13E-04 &  1.07E-03  \\
+Fission new-field update loops         &           &          &          &          &          &            \\
+Un-roll time-smoothing                 &           &          &          &          &          &            \\
 \hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
+\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.8 & 1.8 & 0.03 & -9.0    \\
 \hline

### 2015-08-06 docs/papers/shallow_performance/shallow_perf.tex
shallow paper: Correct the P8 clock speed in the text
fleshed out the psykal section and the conclusions

@@
 
-\author{R.~W.~Ford, M.~Modani, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
-\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
-\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
-\thanks{M.~Modani is with IBM India.}
+\author{ A.~R.~Porter, M.~Ashworth, R.~W.~Ford, M.~Modani, G.~D.~Riley% <-this % stops a space
+\thanks{A.~R.~Porter, M.~Ashworth, R.~W.~Ford and M.~Modani work in the Science and Technology Facilities Council's Daresbury Laboratory, UK.}% <-this % stops a space
+\thanks{G.~D.~Riley works in the School of Computer Science at the University of Manchester, Manchester, UK.}
 }
@@
 \begin{abstract}
-The abstract goes here.
+In this paper we evaluate the performance implications of applying a
+technique called \psykal to finite difference Ocean models. 
+%
+In \psykal, the code related to the underlying science is formally
+separated from code related to parallelisation and single core
+optimisations. This separation of concerns allows scientists to code
+their science independently of the underlying hardware architecture
+(thereby keeping a single code base) and for optimisation specialists
+to be able to tailor the code for a particular machine independently
+of the science code.
+%
+A finite difference shallow water benchmark optimised for cache based
+architectures is taken as the starting point. A vanilla \psykal version
+is written and the performance of the two compared. The optimisations
+that were applied to the original benchmark (loop fusion etc) are then
+applied to the \psykal version as a set of code modifications to the
+optimisation layer. Performance results are presented for the Cray,
+Intel and GNU compilers on Intel Ivybridge and Haswell processors and
+for the IBM compiler on Power8.
+%
+Results show that the combined set of code modifications obtain
+performance that is within a few percent of the original code for all
+compiler and architecture combinations on all tested problem sizes
+apart from GNU on Haswell, one problem size for Intel on Haswell and
+IBM on Power8.
+%
+As the GNU compiler performs well on Ivybridge the reduction in
+performance for the \psykal version was not analysed further. As the
+Intel performance on Haswell was better for the \psykal version than
+the original version, it was not analysed further.
+%
+In the case of the IBM compiler, the \psykal performance is
+significantly better than the original hand optimised code on all
+problem sizes. Analysing the performance of the IBM compiler shows
+that a different structure for the original optimised code produces
+much improved performance. Replicating this different structure using
+appropriate transformations in the \psykal version provides the
+same performance improvements.
+%
+Therefore, the \psykal approach can be used with negligable
+performance loss and sometimes small performance gains compared to the
+original optimised code, there is no single best hand optimised
+implementation of the code for all the compilers tested, and the
+flexibility in applying code modifications to the optimisation layer
+in the PSyKAl version allows for a significant performance improvement
+on the IBM Power8 compared with the original code.
+
 \end{abstract}
@@
 \subsection{The \psykal Approach}
-Proliferation of lightweight cores is resulting in increased
-complexity in the programming models required to achieve performance.
-Combine with the complexity of e.g. a model simulating circulation in
-the global ocean. Significant challenge in software engineering.
-Propose an approach for meeting this challenge by separating the
-computational science demands from the {\it e.g.} oceanographic
-demands.
 
-\begin{figure}
-\centering
-\caption{The \psykal Approach.}
-\end{figure}
+%Proliferation of lightweight cores is resulting in increased
+%complexity in the programming models required to achieve performance.
+%Combine with the complexity of e.g. a model simulating circulation in
+%the global ocean. Significant challenge in software engineering.
+%Propose an approach for meeting this challenge by separating the
+%computational science demands from the {\it e.g.} oceanographic
+%demands.
 
-While encapsulation makes it easier to produce more robust software,
-it can often come with a performance penalty. In this work we
-investigate in detail the hardware and compiler dependencies of this
-penalty and the steps required to remove it.
+%\begin{figure}
+%\centering
+%\caption{The \psykal Approach.}
+%\end{figure}
+
+The \psykal approach separates code into three layers; the Algorithm
+layer, the PSy layer and the Kernel layer. While this approach is
+general, we are currently applying it to Atmosphere and Ocean models
+written in Fortran where domain decomposition is typically performed
+in the latitude-longitude direction, leaving columns of elements on
+each domain-decomposed partition.
+
+The top layer, in terms of calling hierarchy, is the Algorithm
+layer. This layer specifies the algorithm that the scientist would like
+to perform (in terms of calls to kernel and infrastructure routines)
+and logically operates on full fields. We say logically here as the
+fields may be domain decomposed, however the algorithm layer is not
+aware of this. It is the scientist's responsibility to write this
+algorithm layer.
+
+The bottom layer, in terms of calling hierarchy, is the Kernel
+layer. The Kernel layer implements the science that the Algorithm
+layer calls, as a set of subroutines. These kernels operate on local
+fields (a set of elements, a single column of elements, or a set of
+columns, depending on the kernel). Again the scientist is responsible
+for writing this layer and there is no parallelism specified here,
+but, depending on the complexity of the Kernels, there may be input
+from an HPC expert and/or some coding rules to help ensure the kernels
+compile into efficient code.
+
+The PSy layer sits in-between the Algorithm and Kernel layers and its
+functional role is to link the algorithm calls to the associated
+kernel subroutines. As the Algorithm layer works on logically global
+fields and Kernel layer works on local fields, the PSy layer is
+responsible for iterating over columns. It is also responsible for
+including any required distributed-memory operations, such as halo
+swaps and reductions.
+
+As the PSy layer iterates over columns, the single core performance
+can be optimised by applying transformations such as loop fusion and
+loop blocking. Additionally, the potential parallelism within this
+iteration space can also be optimised and parallelised. The PSy layer
+can therefore be tailored for a particular hardware (such as
+multi-core, many-core, GPGPUs, or some combination thereof) and
+software (such as compiler, operating system, MPI implementation)
+configuration with no change to the algorithm or kernel layer
+code. This approach therefore offers the potential for portable
+performance.
+
+Clearly the separation of code into distinct layers may have an effect
+on performance. This overhead, and how to get back to the performance
+of a hand optimised code, and potentially improve on it, will be
+discussed in the remainder of this paper.
 
 \subsection{The `Shallow' Program}
-For this work we have used a benchmarking program that solves the
+
+For this work we use a benchmark called Shallow which solves the
 shallow-water equations on a bi-periodic plane following the
@@
 
-{\bf Auto-generated BC code by reasoning about meta-data. But then,
-  that brings us on to meta-data...}
+%{\bf Auto-generated BC code by reasoning about meta-data. But then,
+%  that brings us on to meta-data...}
 
@@
 
-Our aim in this work is to achieve portable performance. Consequently,
-we have performed tests with the hardware and compilers listed in
-Table~\ref{TABLE_compilers}. Where a compiler is available on a given
-piece of hardware, the version number used in this work is specified.
+The primary aim of this work is to determine whether any performance
+loss due to the \psykal separation of concerns can be negated by
+optimising the PSy layer. To evaluate this we perform tests on a range
+of hardware and a range of compilers. The hardware and compilers are
+listed in Table~\ref{TABLE_compilers}. Where a compiler is available
+on a given piece of hardware, the version number used in this work is
+specified.
+%
 The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
@@
 last-level cache. The cores on the system we used had a clock speed of
-2~GHz.
+3.3~GHz.
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
 \begin{table}[!t]
@@
 lower bounds, respectively, on the performance we expect to achieve.
+
+{\bf RF: Explain that we gradually apply transformations which restructure the psykal code to end up with the same optimised structure as the original code.
+Explain the reason behind the ordering and what each does - I think you said it was the order that you thought would make the biggest difference.
+If you can base it on performance analysis then all the better but I don't think that is required.}
 
@@
 
-The take-home message of all this is that while we can employ a
+The main conclusion of all this analysis is that while we can employ a
 separation of concerns and recover performance, doing so is not
@@
 
+{\bf RF: IBM compiler an example of the above point}
+
 \section{Conclusions}
@@
 related aspects of a computer model to be kept separate from the
-natural (oceanographic) science aspects. In this work we have shown
-that the \psykal approach has the potential to improve performance and
-to improve code structure/maintainability.
+natural (oceanographic) science aspects.
 
-The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal version of a code.
+As expected, applying this separation of concerns does reduce
+performance significantly when comparing with an existing optimised
+code. However, the application of code transformations to the
+performance/PSy layer and the use of appropriate compiler flags can
+recover any performance losses to within a few percent and in some
+cases, despite limiting ourselves to transformations which replicate
+the structure of the optimised code, result in slightly improved
+performance.
 
-Some of our findings have been sent to the IBM compiler team and will
-serve to improve future releases of the compiler.
+The IBM results demonstrate two key points.
+%
+Firstly, even for a relatively simple
+benchmark code, the code structure required to obtain good performance
+between different architecture and/or compiler combinations may
+differ. For hand optimised codes this implies the need to support
+multiple code bases in order to achieve portable performance.
+%
+Secondly, the \psykal approach directly addresses the problem of
+requiring different optimisations for different architectures and/or
+compilers by limiting these differences to a separate performance
+layer, thereby allowing the natural science code base to remain
+unchanged.
+
+In future work we will extend our performance portability analysis of
+the \psykal approach to shared memory parallelisation optimisations on
+different architectures using different parallalisation techniques (in
+particular OpenMP on multi/many-core and OpenACC on GPU's).
+%
+We will then analyse the performance of a domain specific compiler
+that is currently under development. This compiler will generate
+optimised PSy-layer code using a user-provided recipe of
+transformations, thereby removing the need for optimisation experts to
+manually write the PSy layer.
+
+% \bf{Perhaps we don't want to give too much away here so I've removed the optimisation space search plans even though it is mentioned in the text}
+% Finally we plan to use this compiler to allow us to search the optimisation sp%ace of transformations to automatically optimise a code for a particular compil%er/architecture combination or to determine whether we can find improved perfor%mance over existing hand optimised solutions and to automatically
 
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}).
+(\url{http://www.archer.ac.uk}). This work was funded by NERC project {\bf xxx}
 
@@
 
+\begin{IEEEbiographynophoto}{M.~Ashworth}
+Biography text here.
+\end{IEEEbiographynophoto}
+
 % if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{R.~Ford}
+\begin{IEEEbiographynophoto}{R.~W~Ford}
 Biography text here.
@@
 
-\begin{IEEEbiographynophoto}{G.~Riley}
+\begin{IEEEbiographynophoto}{M.~Modani}
+Biography text here.
+\end{IEEEbiographynophoto}
+
+\begin{IEEEbiographynophoto}{G.~D.~Riley}
 Biography text here.

### 2015-08-06 shallow_perf.tex
Change to use IOS formatting instead of IEEE

@@
 
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+\def\hb{\hbox to 10.7 cm{}}
 
@@
 \begin{document}
+
+\pagestyle{headings}
+\def\thepage{}
+
+\begin{frontmatter}              % The preamble begins here.
 %
@@
 \title{Towards Compiler-Agnostic Performance in Finite-Difference Codes}
-%
-% author names and IEEE memberships
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
 
-\author{ A.~R.~Porter, M.~Ashworth, R.~W.~Ford, M.~Modani, G.~D.~Riley% <-this % stops a space
-\thanks{A.~R.~Porter, M.~Ashworth, R.~W.~Ford and M.~Modani work in the Science and Technology Facilities Council's Daresbury Laboratory, UK.}% <-this % stops a space
-\thanks{G.~D.~Riley works in the School of Computer Science at the University of Manchester, Manchester, UK.}
-}
+\author[A]{\fnms{A. R.} \snm{Porter}%
+\thanks{Corresponding Author: Andrew Porter, STFC Daresbury Laboratory, Warrington, WA4 4AD, UK; E-mail:
+andrew.porter@stfc.ac.uk.}},
+\author[A]{\fnms{R. W.} \snm{Ford}},
+\author[A]{\fnms{M.} \snm{Ashworth}},
+\author[A]{\fnms{G. D.} \snm{Riley}}
+and
+\author[B]{\fnms{M.} \snm{Modani}}
+\runningauthor{A.~R.~Porter et al.}
+\address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
+\address[B]{IBM India}
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names.
-
-% The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
 In this paper we evaluate the performance implications of applying a
-technique called \psykal to the finite difference domain. 
+technique called \psykal to finite difference Ocean models. 
 %
@@
 compiler and architecture combinations on all tested problem sizes
-apart from GNU on Haswell and IBM on Power8.
+apart from GNU on Haswell, one problem size for Intel on Haswell and
+IBM on Power8.
 %
 As the GNU compiler performs well on Ivybridge the reduction in
-performance for the \psykal version was not analysed further.
+performance for the \psykal version was not analysed further. As the
+Intel performance on Haswell was better for the \psykal version than
+the original version, it was not analysed further.
 %
 In the case of the IBM compiler, the \psykal performance is
-significantly better than the original hand optimised code. Analysing
-the performance of the IBM compiler shows that a different structure
-for the original optimised code produces much improved performance and
-this performance is also reproduced using the \psykal version using
-appropriate transformations.
+significantly better than the original hand optimised code on all
+problem sizes. Analysing the performance of the IBM compiler shows
+that a different structure for the original optimised code produces
+much improved performance. Replicating this different structure using
+appropriate transformations in the \psykal version provides the
+same performance improvements.
 %
-Therefore the \psykal approach can be used with negligable performance
-loss compared to the original optimised code, there is not a single
-best implementation of the code for all compilers tested, and the
+Therefore, the \psykal approach can be used with negligable
+performance loss and sometimes small performance gains compared to the
+original optimised code, there is no single best hand optimised
+implementation of the code for all the compilers tested, and the
 flexibility in applying code modifications to the optimisation layer
-allows for a significant performance improvement for the IBM Power8
-compared with the original code.
+in the PSyKAl version allows for a significant performance improvement
+on the IBM Power8 compared with the original code.
+
 \end{abstract}
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
+\begin{keyword}
 Performance, Code-generation, Finite-difference
-\end{IEEEkeywords}
+\end{keyword}
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+\end{frontmatter}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HE" in caps to complete the first word.
-\IEEEPARstart{T}{he} challenge presented to the developers of
+% Introduction is not to be numbered so use \section*{}
+\section*{Introduction}
+The challenge presented to the developers of
 scientific software by the drive towards Exascale computing is
@@
 
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
+\subsection{The \psykal Approach}
 
-\subsection{The \psykal Approach}
-Proliferation of lightweight cores is resulting in increased
-complexity in the programming models required to achieve performance.
-Combine with the complexity of e.g. a model simulating circulation in
-the global ocean. Significant challenge in software engineering.
-Propose an approach for meeting this challenge by separating the
-computational science demands from the {\it e.g.} oceanographic
-demands.
+%Proliferation of lightweight cores is resulting in increased
+%complexity in the programming models required to achieve performance.
+%Combine with the complexity of e.g. a model simulating circulation in
+%the global ocean. Significant challenge in software engineering.
+%Propose an approach for meeting this challenge by separating the
+%computational science demands from the {\it e.g.} oceanographic
+%demands.
 
-\begin{figure}
-\centering
-\caption{The \psykal Approach.}
-\end{figure}
+The \psykal approach separates code into three layers; the Algorithm
+layer, the PSy layer and the Kernel layer. While this approach is
+general, we are currently applying it to Atmosphere and Ocean models
+written in Fortran where domain decomposition is typically performed
+in the latitude-longitude direction, leaving columns of elements on
+each domain-decomposed partition.
 
-While encapsulation makes it easier to produce more robust software,
-it can often come with a performance penalty. In this work we
-investigate in detail the hardware and compiler dependencies of this
-penalty and the steps required to remove it.
+The top layer, in terms of calling hierarchy, is the Algorithm
+layer. This layer specifies the algorithm that the scientist would like
+to perform (in terms of calls to kernel and infrastructure routines)
+and logically operates on full fields. We say logically here as the
+fields may be domain decomposed, however the algorithm layer is not
+aware of this. It is the scientist's responsibility to write this
+algorithm layer.
+
+The bottom layer, in terms of calling hierarchy, is the Kernel
+layer. The Kernel layer implements the science that the Algorithm
+layer calls, as a set of subroutines. These kernels operate on local
+fields (a set of elements, a single column of elements, or a set of
+columns, depending on the kernel). Again the scientist is responsible
+for writing this layer and there is no parallelism specified here,
+but, depending on the complexity of the Kernels, there may be input
+from an HPC expert and/or some coding rules to help ensure the kernels
+compile into efficient code.
+
+The PSy layer sits in-between the Algorithm and Kernel layers and its
+functional role is to link the algorithm calls to the associated
+kernel subroutines. As the Algorithm layer works on logically global
+fields and Kernel layer works on local fields, the PSy layer is
+responsible for iterating over columns. It is also responsible for
+including any required distributed-memory operations, such as halo
+swaps and reductions.
+
+As the PSy layer iterates over columns, the single core performance
+can be optimised by applying transformations such as loop fusion and
+loop blocking. Additionally, the potential parallelism within this
+iteration space can also be optimised and parallelised. The PSy layer
+can therefore be tailored for a particular hardware (such as
+multi-core, many-core, GPGPUs, or some combination thereof) and
+software (such as compiler, operating system, MPI implementation)
+configuration with no change to the algorithm or kernel layer
+code. This approach therefore offers the potential for portable
+performance.
+
+Clearly the separation of code into distinct layers may have an effect
+on performance. This overhead, and how to get back to the performance
+of a hand optimised code, and potentially improve on it, will be
+discussed in the remainder of this paper.
 
 \subsection{The `Shallow' Program}
-For this work we have used a benchmarking program that solves the
+
+For this work we use a benchmark called Shallow which solves the
 shallow-water equations on a bi-periodic plane following the
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}).
+(\url{http://www.archer.ac.uk}). This work was funded by NERC project {\bf xxx}
 
 % http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
+\bibliographystyle{unsrt}
 % argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,shallow_perf}
+\bibliography{shallow_perf}
 %
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiographynophoto}{A.~R.~Porter}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{R.~Ford}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiographynophoto}{G.~Riley}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}

### 2015-08-06 SEQ/time_step_mod.f90
Fission loop nest updating new versions of fields

@@ -96,13 +96,21 @@ contains
                       TDTS8*(Zfld(I+1,J+1)+Zfld(I+1,J)) *                     &
                       (CVfld(I+1,J+1)+CVfld(I,J+1)+CVfld(I,J)+CVfld(I+1,J)) - &
                        TDTSDX*(Hfld(I+1,J)-Hfld(I,J))
+       END DO
+    END DO
 
+    DO J=1, N, 1
+       DO I= 1, M, 1
           !CALL compute_vnew_code(i, j+1, vnew, vold, &
           !                       zfld, cufld, hfld, tdt)
           VNEW(I,J+1) = VOLD(I,J+1)-TDTS8*(Zfld(I+1,J+1)+Zfld(I,J+1))        &
                       *(CUfld(I+1,J+1)+CUfld(I,J+1)+CUfld(I,J)+CUfld(I+1,J)) &
                       -TDTSDY*(Hfld(I,J+1)-Hfld(I,J))
+       END DO
+    END DO
 
+    DO J=1, N, 1
+       DO I= 1, M, 1
           !CALL compute_pnew_code(i, j, pnew, pold, &
           !                       cufld, cvfld, tdt)
           PNEW(I,J) = POLD(I,J)-TDTSDX*(CUfld(I+1,J)-CUfld(I,J))   & 

### 2015-08-07 shallow_perf.tex
Remove table containing detailed timings and update text. Update IBM compiler flags and Gnu compiler version. Increase size of plots.

@@
 \hline
-Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
+Intel Haswell    & 4.9.3 & 14.0.0      &         &          \\
 Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
@@
 Cray      & -O3 -O ipa5 -h wp               \\
-IBM       & \parbox{5cm}{-O5 -qinline=auto:level=10\\ -qinline=auto:level=10 -qaltivec=be\\ -qdebug=NSIMDUBTHRESHOLD -qprefetch=dscr=0x1D7} \\
+IBM       & \parbox{5cm}{-O5 -qinline=auto:level=10\\ -qprefetch=dscr=0x1D7} \\
 \hline
@@
 and performance is dominated by the bandwidth to main memory, making
-it relatively insensitive to the choice of compiler. The IBM Fortran
-compiler for Linux on Power 8 is relatively immature and failed to
-SIMD-vectorise two of the three major loop nests in the time-stepping
-loop. Figure~\ref{FIG_orig_perf_summary} therefore also includes the
-performance results obtained after the source had been modified to
-work-around this limitation (we fissioned one loop nest and manually
-unrolled another). Work is currently underway on updating the compiler
-so that these modifications will be unecessary with a future release.
+it relatively insensitive to the choice of compiler.
+
+% The IBM Fortran
+%compiler for Linux on Power 8 is relatively immature and failed to
+%SIMD-vectorise two of the three major loop nests in the time-stepping
+%loop. Figure~\ref{FIG_orig_perf_summary} therefore also includes the
+%performance results obtained after the source had been modified to
+%work-around this limitation (we fissioned one loop nest and manually
+%unrolled another). Work is currently underway on updating the compiler
+%so that these modifications will be unecessary with a future release.
 
@@
 \centering
-\includegraphics[width=2.8in]{orig_summary}
+\includegraphics[width=120mm]{orig_summary}
 \caption{Summary of the performance of the original Shallow code on
-  the range of compilers and hardware under consideration. *After code
-  modifications to encourage SIMD vectorisation (see text).}
+  the range of compilers and hardware under consideration.}
 \label{FIG_orig_perf_summary}
@@
 \centering
-\includegraphics[width=7.5cm]{best_psykal_summary}
+\includegraphics[width=120mm]{best_psykal_summary}
 \caption{Summary of the best performance achieved by any \psykal 
@@
 \centering
-\includegraphics[width=2.8in]{slowdown_summary}
+\includegraphics[width=120mm]{slowdown_summary}
 \caption{Comparison of the performance of the best \psykal
@@
 this for the $256^{2}$ case since this fits within last-level cache on
-all of the CPUs we are using here.  Table~\ref{TABLE_opt_breakdown}
-shows detailed timings for this case after each transformation has
-been applied to the code. The same data is visualised in
-Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
-version for a given compiler/CPU is given relative to the performance
-of the original with the same compiler/CPU.
+all of the CPUs we are using here.  The performance of the \psykal
+version after each transformation has been applied to the code is
+plotted in Figure~\ref{FIG_opt_stages_256}. Results for a given
+compiler/CPU are given relative to the performance of the original
+version of Shallow with the same compiler/CPU.
 
-\begin{table*}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-\caption{The mean time-per-step (s) for the $256^2$ case after each 
-code transformation.}
-\label{TABLE_opt_breakdown}
+\begin{figure}[!t]
 \centering
-\begin{tabular}{l|c|c|c|c|c|c}
-\hline
-Compiler:          &  Gnu  & Intel   & Gnu & Intel & Cray & IBM \\
-\hline
-CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & Power8 \\
-\hline
-Original                               &  7.47E-04 & 7.06E-04 &	7.70E-04 & 7.77E-04 & 6.13E-04 &  1.17E-03  \\
-Vanilla \psykal                        &  1.26E-02 & 8.64E-04 &	1.38E-02 & 9.82E-04 & 8.33E-04 &  7.87E-03  \\
-Explicitly-sized arrays                &  1.26E-02 & 8.62E-04 &	1.38E-02 & 9.65E-04 & 6.96E-04 &  6.74E-03  \\
-Kernels copied into module             &  1.10E-03 & 8.67E-04 &	9.42E-04 & 9.65E-04 & 7.10E-04 &  3.13E-03  \\
-Loops fused                            &  1.09E-03 & 7.89E-04 &	9.07E-04 & 8.80E-04 & 7.35E-04 &  3.60E-03  \\
-In-lined field-copy                    &  1.09E-03 & 7.90E-04 &	9.07E-04 & 8.73E-04 & 7.31E-04 &  3.62E-03  \\
-Fused field-copy                       &  1.09E-03 & 9.25E-04 &	1.02E-03 & 1.04E-03 & 6.50E-04 &  4.11E-03  \\
-In-line all kernels                    &  8.20E-04 & 7.14E-04 &	9.32E-04 & 8.52E-04 & 6.50E-04 &  1.07E-03  \\
-Fully fuse 1st loop nest               &  8.10E-04 & 7.18E-04 &	7.76E-04 & 7.92E-04 & 6.23E-04 &  1.07E-03  \\
-Explicit, constant array/loop bounds   &  8.05E-04 & 7.12E-04 &	7.81E-04 & 7.91E-04 & 6.13E-04 &  1.07E-03  \\
-Fission new-field update loops         &           &          &          &          &          &            \\
-Un-roll time-smoothing                 &           &          &          &          &          &            \\
-\hline
-\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.8 & 1.8 & 0.03 & -9.0    \\
-\hline
-\end{tabular}
-\end{table*}
+\includegraphics[width=120mm]{opt_stages_256}
+\caption{The performance of the \psykal version of Shallow for the
+  $256^{2}$ domain at each stage of optimisation. Results are given as
+  a percentage of the performance of the original code for each
+  compiler/CPU combination.}
+\label{FIG_opt_stages_256}
+\end{figure}
 
@@
 original. Simply copying the code for each of the kernel subroutines
-into the module from which they are called ('module in-lining') has
+into the module from which they are called (`module in-lining') has
 the most dramatic effect on the performance of the compiled code; it
@@
 kernel code itself into the bodies of the loops from which they are
-called ('kernel in-lining'). This gets the \psykal version to within
+called (`kernel in-lining'). This gets the \psykal version to within
 8\% of the performance of the original.  The dramatic performance
@@
 performance are different (second and fourth clusters in
-Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
-shows that even the vanilla \psykal version of Shallow achieves some
-80\% of the performance of the original version. Only two code
-transformations were required to increase this to 100\%. The first of
-these was the fusion of the computational loops. However, we found
-that performance suffered (purple bar) when this was extended to
-include the loop required for field copies at the end of a time-step
-(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
-slow-down revealed that the compiler was unable to SIMD vectorise the
-newly-fused loop after it had in-lined the body of the kernel called
-from within it. Once this in-lining was done by hand (dark-blue bar),
-the compiler was able to vectorise the loop and the performance of the
-original version of Shallow was recovered.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=95mm]{opt_stages_256}
-\caption{The performance of the \psykal version of Shallow for the
-  $256^{2}$ domain at each stage of optimisation. Results are given as
-  a percentage of the performance of the original code for each
-  compiler/CPU combination. A figure greater than 100\% indicates that
-  the \psykal version performs better than the original.}
-\label{FIG_opt_stages_256}
-\end{figure}
+Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal version of
+Shallow achieves some 80\% of the performance of the original
+version. Only two code transformations were required to increase this
+to 100\%. The first of these was the fusion of the computational
+loops. However, we found that performance suffered (purple bar) when
+this was extended to include the loop required for field copies at the
+end of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
+investigation of this slow-down revealed that the compiler was unable
+to SIMD vectorise the newly-fused loop after it had in-lined the body
+of the kernel called from within it. Once this in-lining was done by
+hand (dark-blue bar), the compiler was able to vectorise the loop and
+the performance of the original version of Shallow was recovered.
 

### 2015-08-07 shallow_perf.tex
Change Manish affiliation to Hartree and acknowledge use of Hartree P8 system

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
@@
 \author[A]{\fnms{M.} \snm{Ashworth}},
-\author[A]{\fnms{G. D.} \snm{Riley}}
+\author[B]{\fnms{G. D.} \snm{Riley}}
 and
-\author[B]{\fnms{M.} \snm{Modani}}
+\author[C]{\fnms{M.} \snm{Modani}}
 \runningauthor{A.~R.~Porter et al.}
 \address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
-\address[B]{IBM India}
+\address[B]{The University of Manchester, UK}
+\address[C]{STFC Hartree Centre}
 
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}). This work was funded by NERC project {\bf xxx}
+(\url{http://www.archer.ac.uk}). We acknowledge use of the Hartree
+Centre IBM Power 8 system in this work. The STFC Hartree Centre is a
+research collaboratory in association with IBM providing High
+Performance Computing platforms funded by the UK's investment in
+e-Infrastructure. The Centre aims to develop and demonstrate next
+generation software, optimised to take advantage of the move towards
+exa-scale computing. This work was funded by NERC project {\bf xxx}
 

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Add NERC grant no. to acks

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
+
+\pagestyle{headings}
+\def\thepage{}
+
+\begin{frontmatter}              % The preamble begins here.
 %
@@
 \title{Towards Compiler-Agnostic Performance in Finite-Difference Codes}
-%
-% author names and IEEE memberships
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
 
-\author{ A.~R.~Porter, M.~Ashworth, R.~W.~Ford, M.~Modani, G.~D.~Riley% <-this % stops a space
-\thanks{A.~R.~Porter, M.~Ashworth, R.~W.~Ford and M.~Modani work in the Science and Technology Facilities Council's Daresbury Laboratory, UK.}% <-this % stops a space
-\thanks{G.~D.~Riley works in the School of Computer Science at the University of Manchester, Manchester, UK.}
-}
+\author[A]{\fnms{A. R.} \snm{Porter}%
+\thanks{Corresponding Author: Andrew Porter, STFC Daresbury Laboratory, Warrington, WA4 4AD, UK; E-mail:
+andrew.porter@stfc.ac.uk.}},
+\author[A]{\fnms{R. W.} \snm{Ford}},
+\author[A]{\fnms{M.} \snm{Ashworth}},
+\author[B]{\fnms{G. D.} \snm{Riley}}
+and
+\author[C]{\fnms{M.} \snm{Modani}}
+\runningauthor{A.~R.~Porter et al.}
+\address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
+\address[B]{The University of Manchester, UK}
+\address[C]{STFC Hartree Centre}
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names.
-
-% The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
@@
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
+\begin{keyword}
 Performance, Code-generation, Finite-difference
-\end{IEEEkeywords}
+\end{keyword}
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+\end{frontmatter}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HE" in caps to complete the first word.
-\IEEEPARstart{T}{he} challenge presented to the developers of
+% Introduction is not to be numbered so use \section*{}
+\section*{Introduction}
+The challenge presented to the developers of
 scientific software by the drive towards Exascale computing is
@@
 
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
 \subsection{The \psykal Approach}
@@
 %demands.
-
-%\begin{figure}
-%\centering
-%\caption{The \psykal Approach.}
-%\end{figure}
 
@@
 
-% \bf{Perhaps we don't want to give too much away here so I've removed the optimisation space search plans even though it is mentioned in the text}
-% Finally we plan to use this compiler to allow us to search the optimisation sp%ace of transformations to automatically optimise a code for a particular compil%er/architecture combination or to determine whether we can find improved perfor%mance over existing hand optimised solutions and to automatically
-
 % use section* for acknowledgement
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}). This work was funded by NERC project {\bf xxx}
+(\url{http://www.archer.ac.uk}). We acknowledge use of the Hartree
+Centre IBM Power 8 system in this work. The STFC Hartree Centre is a
+research collaboratory in association with IBM providing High
+Performance Computing platforms funded by the UK's investment in
+e-Infrastructure. The Centre aims to develop and demonstrate next
+generation software, optimised to take advantage of the move towards
+exa-scale computing. This work was funded by the NERC GOcean project,
+grant number NE/L012111/1.
 
 % http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
+\bibliographystyle{unsrt}
 % argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,shallow_perf}
+\bibliography{shallow_perf}
 %
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiographynophoto}{A.~R.~Porter}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-\begin{IEEEbiographynophoto}{M.~Ashworth}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{R.~W~Ford}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiographynophoto}{M.~Modani}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-\begin{IEEEbiographynophoto}{G.~D.~Riley}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Add description of compiler flags

@@
 
-%Proliferation of lightweight cores is resulting in increased
-%complexity in the programming models required to achieve performance.
-%Combine with the complexity of e.g. a model simulating circulation in
-%the global ocean. Significant challenge in software engineering.
-%Propose an approach for meeting this challenge by separating the
-%computational science demands from the {\it e.g.} oceanographic
-%demands.
-
 The \psykal approach separates code into three layers; the Algorithm
@@
 Clearly the separation of code into distinct layers may have an effect
-on performance. This overhead, and how to get back to the performance
-of a hand optimised code, and potentially improve on it, will be
+on performance. This overhead, how to get back to the performance
+of a hand-optimised code, and potentially improve on it, will be
 discussed in the remainder of this paper.
@@
 the domain. In the current work, the lines of code to do these copies
-is manually inserted in the PSy-layer routine as required.
+are manually inserted in the PSy-layer routine as required.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methodology(?)}
+\section{Methodology}
 
@@
 containing effectively a single routine (excluding those for I/O).
+For all compilers apart from Cray it was important to use flags that
+encourage in-lining. With the Cray compiler some benefit was obtained
+by increasing the level of Inter-Procedural Analysis and using
+'whole-program optimisation'. In order to get performant code from the
+IBM compiler it was found essential to make the pre-fetching
+considerably more aggressive than the default.
 
@@
 generation software, optimised to take advantage of the move towards
-exa-scale computing. This work was funded by the NERC GOcean project,
+exa-scale computing. This work was funded by the NERC `GOcean' project,
 grant number NE/L012111/1.

### 2015-08-07 shallow_perf.tex
small update for psykal approach

@@
 
-%Proliferation of lightweight cores is resulting in increased
-%complexity in the programming models required to achieve performance.
-%Combine with the complexity of e.g. a model simulating circulation in
-%the global ocean. Significant challenge in software engineering.
-%Propose an approach for meeting this challenge by separating the
-%computational science demands from the {\it e.g.} oceanographic
-%demands.
+The \psykal approach attempts to address the problems described in the
+previous section. It separates code into three layers; the Algorithm
+layer, the PSy layer and the Kernel layer. The approach has been
+developed in the GungHo project~\cite{GungHo}, which is creating a new
+Dynamical core for the Met Office, and its design has been influenced
+by earlier work on OP2~\cite{OP2,PYOP2}.
 
-The \psykal approach separates code into three layers; the Algorithm
-layer, the PSy layer and the Kernel layer. While this approach is
-general, we are currently applying it to Atmosphere and Ocean models
-written in Fortran where domain decomposition is typically performed
-in the latitude-longitude direction, leaving columns of elements on
-each domain-decomposed partition.
+While this approach is general, we are currently applying it to
+Atmosphere and Ocean models written in Fortran where domain
+decomposition is typically performed in the latitude-longitude
+direction, leaving columns of elements on each domain-decomposed
+partition.
 
@@
 generation software, optimised to take advantage of the move towards
-exa-scale computing. This work was funded by NERC project {\bf xxx}
+exa-scale computing. This work was funded by the NERC `GOcean' project,
+grant number NE/L012111/1.
 

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Add fuller description of transformations and way in which they were chosen

@@
 
-The \psykal approach separates code into three layers; the Algorithm
-layer, the PSy layer and the Kernel layer. While this approach is
-general, we are currently applying it to Atmosphere and Ocean models
-written in Fortran where domain decomposition is typically performed
-in the latitude-longitude direction, leaving columns of elements on
-each domain-decomposed partition.
+The \psykal approach attempts to address the problems described in the
+previous section. It separates code into three layers; the Algorithm
+layer, the PSy layer and the Kernel layer. The approach has been
+developed in the GungHo project~\cite{GungHo}, which is creating a new
+Dynamical core for the Met Office, and its design has been influenced
+by earlier work on OP2~\cite{OP2,PYOP2}.
+
+While this approach is general, we are currently applying it to
+Atmosphere and Ocean models written in Fortran where domain
+decomposition is typically performed in the latitude-longitude
+direction, leaving columns of elements on each domain-decomposed
+partition.
 
@@
 
-%{\bf Auto-generated BC code by reasoning about meta-data. But then,
-%  that brings us on to meta-data...}
+Excluding the application of boundary conditions, the code has three
+main sections: computation of mass fluxes, vorticity and sea-surface
+height; computation of the pressure and velocity components; and
+finally, smoothing of the fields in time and copying of the field
+values at the next time-step to be those at the current
+time-step. Profiling of the original code showed that these three
+sections each account for approximately one third of the run-time.
 
@@
 
-{\bf RF: Explain that we gradually apply transformations which restructure the psykal code to end up with the same optimised structure as the original code.
-Explain the reason behind the ordering and what each does - I think you said it was the order that you thought would make the biggest difference.
-If you can base it on performance analysis then all the better but I don't think that is required.}
+Beginning with the vanilla \psykal version, we then apply a series of
+(cumulative) code transformations while obeying the \psykal separation
+of concerns, {\it i.e.} optimisation is restricted to the middle,
+{PS}y layer and leaves the kernel and algorithm layers unchanged. The
+aim of these optimisations is to recover, as much as is possible, the
+structure and thus performance of the original version of the code. We
+began with two transformations that affect all of the kernels. After
+this, we determined the transformations to perform by comparing the
+timing profiles of the \psykal and original versions of the code. We
+then tackled the sections resposible for the largest slowdowns and
+iterated this process. While doing this we also made heavy use of the
+optimisation reports produced by the compilers and how these changed
+when moving from the original to the \psykal version of the code.
 
-{\bf Performance analysis as motivation for the code
-  tranformations. More detail on each transformation.}
-
-Beginning with the vanilla \psykal version, we then apply a series of
-code transformations while obeying the \psykal separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
-the kernel and algorithm layers unchanged. The aim of these
-optimisations is to recover, as much as is possible, the structure and
-thus performance of the original version of the code. The
-transformations we have performed are as follows:
+The transformations we have performed are as follows:
 \begin{enumerate}
 
-\item Pass explicit array bounds to middle layer (rather than using
-  assumed-size array arguments)
+\item Specify array bounds in the middle layer using module variables
+  (rather than using assumed-size array arguments);
 
-\item Move kernel routines into same module as middle layer
+\item Move all kernel subroutines into the same module as as that
+  containing the middle/PSy layer (module inlining);
 
-\item Loop fusion
+\item Loop fusion: all neighbouring loop nests were fully fused (where
+  possible) with the exception of the first loop nest where it was
+  found that fusing only the outer loop was more performant.
 
-\item In-lined field-copy operation
+\item Manually in-line (kernel inline) the three field-copy operations
+  performed as part of the time-update section;
 
-\item Fused field-copy operation
+\item Loop fuse the three field-copy operations with the
+  time-smoothing loop nest;
 
-\item Manually in-line kernel bodies into middle-layer code
+\item Manually in-line all kernel bodies into the middle-layer
+  subroutine (kernel inlining);
 
-\item Fully fuse first loop (only the outer loop of this loop nest was
-  fused in the first pass)
+\item Fully fuse the first loop nest (only the outer loop of this loop
+  nest was fused in the first pass);
 
-\item Make the array/loop bounds explicitly constant for the
-      duration of the time-stepping loop.
+\item Make the array/loop bounds explicitly constant for the duration
+  of the time-stepping loop (pass as arguments to the middle layer
+  rather than using module variables).
+
 \end{enumerate}
+
+In addition to these transformations we also applied two
+transformations that were suggested by the IBM compiler team. The
+first of these manually blocked the outer loop of the time-smoothing
+loop nest using a factor of four. The second effectively undoes our 3rd
+transformation by fissioning the first loop nest back into three
+separate loop nests.
 

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Re-written text about overall slowdowns/speedups. Removed one plot as essentially duplicated the plot of perf. of original code

@@
 larger last-level cache of the Intel Ivy Bridge system compared to
-that of the Haswell system; note the significant drop in performance
+that of the Haswell system; note the drop in performance
 in going from a problem size of $256^{2}$ to a problem size of
-$512^{2}$ for the black and green bars. The performance of the Ivy
-Bridge system with the Cray or Intel compiler (red and purple bars)
+$512^{2}$ for the first two bars. The performance of the Ivy
+Bridge system with the Cray or Intel compiler (penultimate two bars)
 only drops to this level when the domain size is increased to
@@
 
-For this compiler-friendly form of the code, there is little
+For this compiler-friendly form of the code, there is generally little
 performance difference between the executables produced by the Gnu and
-Intel compilers. However, the executable produced by the Cray compiler
-generally performs significantly better for all but the largest domain
-size.
+Intel compilers on both the Haswell and Ivy Bridge CPUs. However, the
+executable produced by the Cray compiler generally performs
+significantly better for all but the largest domain size. Analysis of
+the optimisation reports produced by the Cray and Intel compilers did
+not give any clues as to the cause of this difference and an in-depth
+investigation is outside the scope of this paper.
 
-Moving now the to the \psykal version of Shallow,
-Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest \psykal version for each of the compiler/CPU combinations. The
-only significant difference between this and the plot of the
-performance of the original version of the code in
-Figure~\ref{FIG_orig_perf_summary} is that the difference between the
-Gnu- and Intel-compiled executables is slightly more marked. This is
-clarified in Figure~\ref{FIG_slowdown_summary} which plots the
-percentage difference between the performance of the original and the
-(best) \psykal versions of Shallow for each compiler/CPU combination.
-It is clear from this plot that the difference in the performance of
-the Gnu- and Intel-compiled excutables on the Haswell system is due to
-the Gnu compiler being unable to recover the performance achieved by
-the original code. In contrast there is no such problem for the
-Gnu-compiled executable on the Ivy Bridge system. We attribute this to
-the more recent version of the compiler on the latter system (4.9.1 as
-opposed to 4.8.2).
+Moving now the to the \psykal version of Shallow, it turns out that a
+plot of the performance of the best performing \psykal version for
+each of the compiler/CPU combinations is very similar to
+Figure~\ref{FIG_orig_perf_summary}. Therefore, in
+Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
+  difference} between the performance of the original and the (best)
+\psykal versions of Shallow for each compiler/CPU combination. The
+most important feature of this plot is that the performance penalty
+incurred by the \psykal version is less than four percent with only
+one exception where it is 6.6\%.  In fact, for the Intel and Cray
+compilers, the \psykal version of Shallow is never more than 2\%
+slower than the original and in some cases is faster.  This
+demonstrates that we can reliably recover the performance of the
+original version of the code, despite the significant restructuring
+required by the \psykal approach.
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=120mm]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any \psykal 
-version of Shallow with each of the compilers and CPUs under 
-consideration.}
-\label{FIG_best_psykal_perf_summary}
-\end{figure}
+The exceptional case is that of the Gnu-compiled binary on the Haswell
+CPU for a problem size of $256^2$.  This is strange since the
+performance difference for the same case on the Ivy Bridge CPU is just
+0.8\%. {\bf installing 4.9.1 on desktop so can try running this case
+  with it. In fact there are considerable difference between
+  slow-downs seen with 4.8.2 and 4.9.3 on desktop -things are much
+  better with more recent compiler.}
 
@@
 
-In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
-\psykal version of Shallow is never more than 2\% slower than the
-original and in some cases is faster.  This demonstrates that we can
-reliably recover the performance of the original version of the code,
-despite the significant restructuring required by the \psykal
-approach.
+Turning to the cases where the \psykal version is significantly
+{\em faster} than the original, we see from
+Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
+this only occurs for the smallest domain size. This is no doubt due in
+part to the fact that the problem executes very quickly and therefore
+a small reduction in run-time will show as a large precentage. Since
+this problem is well contained in L2 cache, its performance will also
+be more sensitive to small differences, {\it e.g.} in the way that CPU
+registers are used.
 
-The most striking feature of Figure~\ref{FIG_slowdown_summary} is the
-fact that the best \psykal version of Shallow is consistently faster
-than the original on the Power 8. This confirms that this early
-version of the IBM compiler struggled with the original form of the
-Shallow source code.
+The performance of the Power 8 system with the $512^2$ domain size is
+somewhat anomolous: the \psykal version of the code performs
+significantly better than the original and in addition, this problem
+size is also the only one where the Power 8 out-performs the Haswell
+system (see Figure~\ref{FIG_orig_perf_summary}) for the original
+version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
+that the $512^2$ case marks a transition from the working set being in
+cache (sizes $\leq 256$) to spilling out of cache (size 1024). We
+therefore suggest that cache is being used more efficiently on the
+Power 8 system and this is further enhanced by the IBM-supplied
+transformations in the \psykal version.
 

### 2015-08-07 shallow_perf.tex
Update text describing transformations with gnu compiler

@@
 
-{\bf RF: Explain that we gradually apply transformations which restructure the psykal code to end up with the same optimised structure as the original code.
-Explain the reason behind the ordering and what each does - I think you said it was the order that you thought would make the biggest difference.
-If you can base it on performance analysis then all the better but I don't think that is required.}
+Beginning with the vanilla \psykal version, we then apply a series of
+(cumulative) code transformations while obeying the \psykal separation
+of concerns, {\it i.e.} optimisation is restricted to the middle,
+{PS}y layer and leaves the kernel and algorithm layers unchanged. The
+aim of these optimisations is to recover, as much as is possible, the
+structure and thus performance of the original version of the code. We
+began with two transformations that affect all of the kernels. After
+this, we determined the transformations to perform by comparing the
+timing profiles of the \psykal and original versions of the code. We
+then tackled the sections resposible for the largest slowdowns and
+iterated this process. While doing this we also made heavy use of the
+optimisation reports produced by the compilers and how these changed
+when moving from the original to the \psykal version of the code.
 
-{\bf Performance analysis as motivation for the code
-  tranformations. More detail on each transformation.}
-
-Beginning with the vanilla \psykal version, we then apply a series of
-code transformations while obeying the \psykal separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
-the kernel and algorithm layers unchanged. The aim of these
-optimisations is to recover, as much as is possible, the structure and
-thus performance of the original version of the code. The
-transformations we have performed are as follows:
+The transformations we have performed are as follows:
 \begin{enumerate}
 
-\item Pass explicit array bounds to middle layer (rather than using
-  assumed-size array arguments)
+\item Specify array bounds in the middle layer using module variables
+  (rather than using assumed-size array arguments);
 
-\item Move kernel routines into same module as middle layer
+\item Move all kernel subroutines into the same module as as that
+  containing the middle/PSy layer (`module inlining');
 
-\item Loop fusion
+\item Loop fusion: all neighbouring loop nests were fully fused (where
+  possible) with the exception of the first loop nest where it was
+  found that fusing only the outer loop was more performant.
 
-\item In-lined field-copy operation
+\item Manually in-line (`kernel inline') the three field-copy operations
+  performed as part of the time-update section;
 
-\item Fused field-copy operation
+\item Loop fuse the three field-copy operations with the
+  time-smoothing loop nest;
 
-\item Manually in-line kernel bodies into middle-layer code
+\item Manually in-line all kernel bodies into the middle-layer
+  subroutine (kernel inlining);
 
-\item Fully fuse first loop (only the outer loop of this loop nest was
-  fused in the first pass)
+\item Fully fuse the first loop nest (only the outer loop of this loop
+  nest was fused in the first pass);
 
-\item Make the array/loop bounds explicitly constant for the
-      duration of the time-stepping loop.
+\item Make the array/loop bounds explicitly constant for the duration
+  of the time-stepping loop (pass as arguments to the middle layer
+  rather than using module variables).
+
 \end{enumerate}
+
+In addition to these transformations we also applied two
+transformations that were suggested by the IBM compiler team. The
+first of these manually blocked the outer loop of the time-smoothing
+loop nest using a factor of four. The second effectively undoes our 3rd
+transformation by fissioning the first loop nest back into three
+separate loop nests.
 
@@
 
-For this compiler-friendly form of the code, there is little
+For this compiler-friendly form of the code, there is generally little
 performance difference between the executables produced by the Gnu and
-Intel compilers. However, the executable produced by the Cray compiler
-generally performs significantly better for all but the largest domain
-size.
+Intel compilers on both the Haswell and Ivy Bridge CPUs. However, the
+executable produced by the Cray compiler generally performs
+significantly better for all but the largest domain size. Analysis of
+the optimisation reports produced by the Cray and Intel compilers did
+not give any clues as to the cause of this difference and an in-depth
+investigation is outside the scope of this paper.
 
-Moving now the to the \psykal version of Shallow,
-Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest \psykal version for each of the compiler/CPU combinations. The
-only significant difference between this and the plot of the
-performance of the original version of the code in
-Figure~\ref{FIG_orig_perf_summary} is that the difference between the
-Gnu- and Intel-compiled executables is slightly more marked. This is
-clarified in Figure~\ref{FIG_slowdown_summary} which plots the
-percentage difference between the performance of the original and the
-(best) \psykal versions of Shallow for each compiler/CPU combination.
-It is clear from this plot that the difference in the performance of
-the Gnu- and Intel-compiled excutables on the Haswell system is due to
-the Gnu compiler being unable to recover the performance achieved by
-the original code. In contrast there is no such problem for the
-Gnu-compiled executable on the Ivy Bridge system. We attribute this to
-the more recent version of the compiler on the latter system (4.9.1 as
-opposed to 4.8.2).
+Moving now the to the \psykal version of Shallow, it turns out that a
+plot of the performance of the best performing \psykal version for
+each of the compiler/CPU combinations is very similar to
+Figure~\ref{FIG_orig_perf_summary}. Therefore, in
+Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
+  difference} between the performance of the original and the (best)
+\psykal versions of Shallow for each compiler/CPU combination. The
+most important feature of this plot is that the performance penalty
+incurred by the \psykal version is less than four percent with only
+one exception where it is 6.6\%.  In fact, for the Intel and Cray
+compilers, the \psykal version of Shallow is never more than 2\%
+slower than the original and in some cases is faster.  This
+demonstrates that we can reliably recover the performance of the
+original version of the code, despite the significant restructuring
+required by the \psykal approach.
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=120mm]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any \psykal 
-version of Shallow with each of the compilers and CPUs under 
-consideration.}
-\label{FIG_best_psykal_perf_summary}
-\end{figure}
+The exceptional case is that of the Gnu-compiled binary on the Haswell
+CPU for a problem size of $256^2$.  This is strange since the
+performance difference for the same case on the Ivy Bridge CPU is just
+0.8\%. We investigated this further by compiling and running this case
+with versions 4.8.3, 4.9.1 and 4.9.3 of the Gnu compiler on the
+Haswell system. These gave slow-downs of 7.8\%, 7.3\% and 6.6\%,
+respectively. From these results we conclude that the Gnu compiler's
+support for/ability to exploit the newer Haswell architecture is still
+maturing.
 
@@
 \centering
-\includegraphics[width=120mm]{opt_stages_256}
+\includegraphics[width=123mm]{opt_stages_256}
 \caption{The performance of the \psykal version of Shallow for the
@@
 original. Simply copying the code for each of the kernel subroutines
-into the module from which they are called (`module in-lining') has
-the most dramatic effect on the performance of the compiled code; it
-now achieves almost 80\% of the performance of the original. The next
-most significant transformation is that of manually in-lining the
-kernel code itself into the bodies of the loops from which they are
-called (`kernel in-lining'). This gets the \psykal version to within
-8\% of the performance of the original.  The dramatic performance
-improvement obtained from the first transformation demonstrates that
-the Gnu compiler is unable to optimise over separate files. Therefore,
-the fact that the time-stepping loop in the \psykal version still
-retains a call to an I/O routine and a call to the middle-layer, both
-of which are in different files, is likely to account for this
-remaining performance penalty of 8\%.
+into the module from which they are called (module in-lining) has the
+most dramatic effect on the performance of the compiled code; it now
+achieves more than 80\% of the performance of the original. The
+transformations required to get beyond this level then become
+architecture dependent: on Haswell loop fusion achieves more than 90\%
+while on Ivy Bridge it is fully-fusing the first loop nest. The
+dramatic performance improvement obtained from the module-inlining
+transformation demonstrates that the Gnu compiler is unable to
+optimise over separate files.
 

### 2015-08-07 shallow_perf.tex
Changes to abstract and conclusions to reflect updated P8 results

@@
 %
-% paper title
 % can use linebreaks \\ within to get better formatting as desired
@@
 In this paper we evaluate the performance implications of applying a
-technique called \psykal to finite difference Ocean models. 
+technique which we call \psykal to finite difference Ocean models. 
 %
@@
 is written and the performance of the two compared. The optimisations
-that were applied to the original benchmark (loop fusion etc) are then
+that were applied to the original benchmark (loop fusion {\it etc.}) are then
 applied to the \psykal version as a set of code modifications to the
@@
 performance that is within a few percent of the original code for all
-compiler and architecture combinations on all tested problem sizes
-apart from GNU on Haswell, one problem size for Intel on Haswell and
-IBM on Power8.
-%
-As the GNU compiler performs well on Ivybridge the reduction in
-performance for the \psykal version was not analysed further. As the
-Intel performance on Haswell was better for the \psykal version than
-the original version, it was not analysed further.
-%
-In the case of the IBM compiler, the \psykal performance is
-significantly better than the original hand optimised code on all
-problem sizes. Analysing the performance of the IBM compiler shows
-that a different structure for the original optimised code produces
-much improved performance. Replicating this different structure using
-appropriate transformations in the \psykal version provides the
-same performance improvements.
-%
+compiler and architecture combinations on all tested problem
+sizes. The only exception to this (other than where we see significant
+speed-ups) is the Gnu compiler on Haswell for one problem size. Our
+tests indicate that this may be due to immature support for that
+architecture in the Gnu compiler -- no such problem is seen on the Ivy
+Bridge system.
+%%
 Therefore, the \psykal approach can be used with negligable
@@
 that the $512^2$ case marks a transition from the working set being in
-cache (sizes $\leq 256$) to spilling out of cache (size 1024). We
+cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
 therefore suggest that cache is being used more efficiently on the
@@
 version. Only two code transformations were required to increase this
-to 100\%. The first of these was the fusion of the computational
-loops. However, we found that performance suffered (purple bar) when
-this was extended to include the loop required for field copies at the
-end of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
+to $\sim$100\%. The first of these was the fusion of the computational
+loops. However, we found that performance suffered (bar 5) when this
+was extended to include the loop required for field copies at the end
+of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
 investigation of this slow-down revealed that the compiler was unable
@@
 of the kernel called from within it. Once this in-lining was done by
-hand (dark-blue bar), the compiler was able to vectorise the loop and
-the performance of the original version of Shallow was recovered.
+hand (bar 6), the compiler was able to vectorise the loop and the
+performance of the original version of Shallow was recovered on the
+Haswell system. On Ivy Bridge the first loop nest also had to be fully
+fused (bar 7) to recover $\sim100\%$ of the performance.
 
@@
 are passed as an argument to the middle layer. This is done in order
-to specify to the compiler that these bounds remain constant for the
-duration of the time-stepping loop.
+to indicate to the compiler that these bounds remain constant for the
+duration of each trip of the time-stepping loop.
 
-As with the Gnu compiler, it is module in-lining and kernel in-lining
-that are important for the IBM compiler (sixth cluster in
-Figure~\ref{FIG_opt_stages_256}). Fusing loop nests actually harms
-performance, just as it did for the Cray compiler.
+Moving to the IBM compiler on the Power 8 system (sixth cluster in
+Figure~\ref{FIG_opt_stages_256}) we see that performance of the
+vanilla \psykal version is only 5\% of the original and several
+transformations are required to achieve comparable performance. As
+with the Cray compiler, specifying array bounds explicitly improves
+performance. Module-inlining also has a significant effect; doubling
+the performance from 10 to 20\% of the original. However, fusing loop
+nests actually harms performance, just as it did for the Cray
+compiler. Kernel inlining takes performance to over 50\% and it is the
+IBM-supplied transformations of loop unrolling and fissioning the
+first loop ({\it i.e.} undoing some of transformations three and
+seven) that finally recovers 96\% of the performance of the
+original. We note that this final pair of transformations significantly
+reduce the performance obtained with the Intel and Gnu compilers
+although they do not much perturb that obtained by the Cray
+compiler.
 
@@
 will almost certainly result in a code that does not perform optimally
-on any system.
-
-{\bf RF: IBM compiler an example of the above point}
+on any system. For instance, loop fusion benefits performance for the
+Gnu and Intel compilers and yet hurts performance with the Cray and
+IBM compilers.
 
@@
 As expected, applying this separation of concerns does reduce
-performance significantly when comparing with an existing optimised
+performance significantly when compared with an existing, optimised
 code. However, the application of code transformations to the
@@
 
-The IBM results demonstrate two key points.
+The IBM results in particular demonstrate two key points.
 %
-Firstly, even for a relatively simple
+First, even for a relatively simple
 benchmark code, the code structure required to obtain good performance
@@
 %
-Secondly, the \psykal approach directly addresses the problem of
+Second, the \psykal approach directly addresses the problem of
 requiring different optimisations for different architectures and/or
@@
 the \psykal approach to shared memory parallelisation optimisations on
-different architectures using different parallalisation techniques (in
+different architectures using different parallelisation techniques (in
 particular OpenMP on multi/many-core and OpenACC on GPU's).
 %
-We will then analyse the performance of a domain specific compiler
+We will then analyse the performance of a domain-specific compiler
 that is currently under development. This compiler will generate

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Minor tidying

@@
 
-\subsection{The \psykal Approach}
+\subsection*{The \psykal Approach}
 
@@
 
-\subsection{The `Shallow' Program}
+\subsection*{The `Shallow' Program}
 
@@
 \centering
-\includegraphics[width=85mm]{psykal_shallow}
+\includegraphics[width=100mm]{psykal_shallow}
 \caption{A schematic of the vanilla \psykal version of the Shallow code.}
@@
 cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
-'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
+`chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
 512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
@@
 \item Move all kernel subroutines into the same module as as that
-  containing the middle/PSy layer (module inlining);
+  containing the middle/PSy layer (`module inlining');
 
@@
 
-\item Manually in-line (kernel inline) the three field-copy operations
+\item Manually in-line (`kernel inline') the three field-copy operations
   performed as part of the time-update section;
@@
   of the time-stepping loop (pass as arguments to the middle layer
-  rather than using module variables).
+  rather than using module variables);
+
+\item Un-roll the outer loop of the time-smoothing loop nest using a
+blocking factor of four (IBM supplied);
+
+\item Fission the first loop nest back into three separate loop nests
+  (IBM supplied).
 
@@
 
-In addition to these transformations we also applied two
-transformations that were suggested by the IBM compiler team. The
-first of these manually blocked the outer loop of the time-smoothing
-loop nest using a factor of four. The second effectively undoes our 3rd
-transformation by fissioning the first loop nest back into three
-separate loop nests.
+As indicated above, the last two of these transformations were
+suggested by the IBM compiler team. The second of these effectively
+undoes our 3rd transformation by fissioning the first loop nest back
+into three separate loop nests.
 
@@
 and the problem size. We also emphasise that the aim of these
-optimisations is to recover, as far as is possible, the structure of
+optimisations is typically to recover, as far as is possible, the structure of
 the original version of the code. It may well be that transforming the
@@
 it relatively insensitive to the choice of compiler.
-
-% The IBM Fortran
-%compiler for Linux on Power 8 is relatively immature and failed to
-%SIMD-vectorise two of the three major loop nests in the time-stepping
-%loop. Figure~\ref{FIG_orig_perf_summary} therefore also includes the
-%performance results obtained after the source had been modified to
-%work-around this limitation (we fissioned one loop nest and manually
-%unrolled another). Work is currently underway on updating the compiler
-%so that these modifications will be unecessary with a future release.
 
@@
 performance difference for the same case on the Ivy Bridge CPU is just
-0.8\%. {\bf installing 4.9.1 on desktop so can try running this case
-  with it. In fact there are considerable difference between
-  slow-downs seen with 4.8.2 and 4.9.3 on desktop -things are much
-  better with more recent compiler.}
+0.8\%. We investigated this further by compiling and running this case
+with versions 4.8.3, 4.9.1 and 4.9.3 of the Gnu compiler on the
+Haswell system. These gave slow-downs of 7.8\%, 7.3\% and 6.6\%,
+respectively. From these results we conclude that the Gnu compiler's
+support for/ability to exploit the newer Haswell architecture is still
+maturing.
 

### 2015-09-07 docs/papers/shallow_performance/shallow_perf.tex
Re-word sentence to avoid <IBM results> to please IBM reviewer
PARCO paper: Re-ordering of results section following suggestion of reviewer

@@
 layer. The Kernel layer implements the science that the Algorithm
-layer calls, as a set of subroutines. These kernels operate on local
-fields (a set of elements, a single column of elements, or a set of
-columns, depending on the kernel). Again the scientist is responsible
+layer calls, as a set of subroutines. These kernels operate on fields
+that are local to the process doing the computation. (Depending on the
+type of kernel, these may be a set of elements, a single column of
+elements, or a set of columns.). Again the scientist is responsible
 for writing this layer and there is no parallelism specified here,
@@
 responsible for iterating over columns. It is also responsible for
-including any required distributed-memory operations, such as halo
-swaps and reductions.
+including any distributed-memory operations resulting from the
+decomposition of the simulation domain, such as halo swaps and
+reductions.
 
@@
 evolutionary development with subsequent people making various changes
-and optimising it for previous generations of hardware.  In describing
-our work, we term the version of the Shallow program obtained at the
-beginning of this project the `original' version.
+and optimising it for previous generations of hardware.  There is no
+complete record of the people involved and the precise hardware they
+were targetting. In describing our work, we term the version of the
+Shallow program obtained at the beginning of this project the
+`original' version.
 
@@
 we ensure that any \psykal version of Shallow retains the Input/Output
-capability of the original. This aids in limiting the optimisations
-that can be performed on the \psykal version to those that should also
-be applicable to full oceanographic models. Note that although we
-retain the I/O functionality, all of the results presented in this work
-carefully exclude the effects of I/O since it is compute performance
-that interests us here.
+capability of the original (it reads the problem specification from a
+namelist and outputs the solution fields every $n$ timesteps). This
+aids in limiting the optimisations that can be performed on the
+\psykal version to those that should also be applicable to full
+oceanographic models. Note that although we retain the I/O
+functionality, all of the results presented in this work carefully
+exclude the effects of I/O since it is compute performance that
+interests us here.
 
@@
 loss due to the \psykal separation of concerns can be negated by
-optimising the PSy layer. To evaluate this we perform tests on a range
-of hardware and a range of compilers. The hardware and compilers are
+optimising the PSy layer and if so, what the key transformations are
+(for a given system). To evaluate this we perform tests on a range of
+hardware and a range of compilers. The hardware and compilers are
 listed in Table~\ref{TABLE_compilers}. Where a compiler is available
@@
 
-Beginning with the vanilla \psykal version, we then apply a series of
-(cumulative) code transformations while obeying the \psykal separation
-of concerns, {\it i.e.} optimisation is restricted to the middle,
-{PS}y layer and leaves the kernel and algorithm layers unchanged. The
-aim of these optimisations is to recover, as much as is possible, the
-structure and thus performance of the original version of the code. We
-began with two transformations that affect all of the kernels. After
-this, we determined the transformations to perform by comparing the
-timing profiles of the \psykal and original versions of the code. We
-then tackled the sections resposible for the largest slowdowns and
-iterated this process. While doing this we also made heavy use of the
-optimisation reports produced by the compilers and how these changed
-when moving from the original to the \psykal version of the code.
+Beginning with the vanilla \psykal version, we then manually apply a
+series of (cumulative) code transformations while obeying the \psykal
+separation of concerns, {\it i.e.} optimisation is restricted to the
+middle, {PS}y layer and leaves the kernel and algorithm layers
+unchanged. The aim of these optimisations is to recover, as much as is
+possible, the structure and thus performance of the original version
+of the code. We began with two transformations that affect all of the
+kernels. After this, we determined the transformations to perform by
+comparing the timing profiles of the \psykal and original versions of
+the code. We then tackled the sections resposible for the largest
+slowdowns and iterated this process. While doing this we also made
+heavy use of the optimisation reports produced by the compilers and
+how these changed when moving from the original to the \psykal version
+of the code.
 
@@
 
+In this section we first discuss the performance of the original
+version of the code with the various compiler/CPU combinations. We
+then investigate the different code transformations that are required
+to recover good performance from the \psykal version of Shallow. Finally,
+we compare the most performant \psykal version  with the original
+version of Shallow for each compiler/CPU combination.
+
+\subsection{Performance of the Original Version of Shallow}
+
 In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
@@
 
-Moving now the to the \psykal version of Shallow, it turns out that a
-plot of the performance of the best performing \psykal version for
-each of the compiler/CPU combinations is very similar to
-Figure~\ref{FIG_orig_perf_summary}. Therefore, in
-Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
-  difference} between the performance of the original and the (best)
-\psykal versions of Shallow for each compiler/CPU combination. The
-most important feature of this plot is that the performance penalty
-incurred by the \psykal version is less than four percent with only
-one exception where it is 6.6\%.  In fact, for the Intel and Cray
-compilers, the \psykal version of Shallow is never more than 2\%
-slower than the original and in some cases is faster.  This
-demonstrates that we can reliably recover the performance of the
-original version of the code, despite the significant restructuring
-required by the \psykal approach.
+\subsection{Optimisations of the \psykal Version to Recover Performance}
 
-The exceptional case is that of the Gnu-compiled binary on the Haswell
-CPU for a problem size of $256^2$.  This is strange since the
-performance difference for the same case on the Ivy Bridge CPU is just
-0.8\%. We investigated this further by compiling and running this case
-with versions 4.8.3, 4.9.1 and 4.9.3 of the Gnu compiler on the
-Haswell system. These gave slow-downs of 7.8\%, 7.3\% and 6.6\%,
-respectively. From these results we conclude that the Gnu compiler's
-support for/ability to exploit the newer Haswell architecture is still
-maturing.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=120mm]{slowdown_summary}
-\caption{Comparison of the performance of the best \psykal
-version with that of the original version of the code. A negative value 
-indicates that the \psykal version is faster than the original.}
-\label{FIG_slowdown_summary}
-\end{figure}
-
-Turning to the cases where the \psykal version is significantly
-{\em faster} than the original, we see from
-Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
-this only occurs for the smallest domain size. This is no doubt due in
-part to the fact that the problem executes very quickly and therefore
-a small reduction in run-time will show as a large precentage. Since
-this problem is well contained in L2 cache, its performance will also
-be more sensitive to small differences, {\it e.g.} in the way that CPU
-registers are used.
-
-The performance of the Power 8 system with the $512^2$ domain size is
-somewhat anomolous: the \psykal version of the code performs
-significantly better than the original and in addition, this problem
-size is also the only one where the Power 8 out-performs the Haswell
-system (see Figure~\ref{FIG_orig_perf_summary}) for the original
-version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
-that the $512^2$ case marks a transition from the working set being in
-cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
-therefore suggest that cache is being used more efficiently on the
-Power 8 system and this is further enhanced by the IBM-supplied
-transformations in the \psykal version.
-
-Having shown that, in general, we can recover and occasionally improve
-upon the performance of the original version of Shallow, the next
-logical step is to examine the necessary code transformations.  We do
-this for the $256^{2}$ case since this fits within last-level cache on
-all of the CPUs we are using here.  The performance of the \psykal
-version after each transformation has been applied to the code is
-plotted in Figure~\ref{FIG_opt_stages_256}. Results for a given
-compiler/CPU are given relative to the performance of the original
-version of Shallow with the same compiler/CPU.
+We now examine the performance of the \psykal version of Shallow and
+the code transformations that are required to reach the levels obtained
+by the original.  We do this for the $256^{2}$ problem size since this
+fits within last-level cache on all of the CPUs we are using here.
+The performance of the \psykal version after each transformation has
+been applied to the code is plotted in
+Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
+given relative to the performance of the original version of Shallow
+with the same compiler/CPU.
 
@@
 
+\subsection{Performance Comparison of the Original and \psykal Versions}
+
+Finally, we compare the performance of the best \psykal version of
+Shallow with that of the original. In
+Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
+  difference} between the performance of the original and the (best)
+\psykal versions of Shallow for each compiler/CPU combination. The
+most important feature of this plot is that the performance penalty
+incurred by the \psykal version is less than four percent with only
+one exception where it is 6.6\%.  In fact, for the Intel and Cray
+compilers, the \psykal version of Shallow is never more than 2\%
+slower than the original and in some cases is faster.  This
+demonstrates that we can reliably recover the performance of the
+original version of the code, despite the significant restructuring
+required by the \psykal approach.
+
+The exceptional case is that of the Gnu-compiled binary on the Haswell
+CPU for a problem size of $256^2$.  This is strange since the
+performance difference for the same case on the Ivy Bridge CPU is just
+0.8\%. We investigated this further by compiling and running this case
+with versions 4.8.3, 4.9.1 and 4.9.3 of the Gnu compiler on the
+Haswell system. These gave slow-downs of 7.8\%, 7.3\% and 6.6\%,
+respectively. From these results we conclude that the Gnu compiler's
+support for/ability to exploit the newer Haswell architecture is still
+maturing.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=120mm]{slowdown_summary}
+\caption{Comparison of the performance of the best \psykal
+version with that of the original version of the code. A negative value 
+indicates that the \psykal version is faster than the original.}
+\label{FIG_slowdown_summary}
+\end{figure}
+
+Turning to the cases where the \psykal version is significantly
+{\em faster} than the original, we see from
+Figure~\ref{FIG_slowdown_summary} that, with the exception of Power 8,
+this only occurs for the smallest domain size. This is no doubt due in
+part to the fact that the problem executes very quickly and therefore
+a small reduction in run-time will show as a large precentage. Since
+this problem is well contained in L2 cache, its performance will also
+be more sensitive to small differences, {\it e.g.} in the way that CPU
+registers are used.
+
+The performance of the Power 8 system with the $512^2$ domain size is
+somewhat anomolous: the \psykal version of the code performs
+significantly better than the original and in addition, this problem
+size is also the only one where the Power 8 out-performs the Haswell
+system (see Figure~\ref{FIG_orig_perf_summary}) for the original
+version.  The results in Figure~\ref{FIG_orig_perf_summary} indicate
+that the $512^2$ case marks a transition from the working set being in
+cache (sizes $\leq 256^2$) to spilling out of cache (size $1024^2$). We
+therefore suggest that cache is being used more efficiently on the
+Power 8 system and this is further enhanced by the IBM-supplied
+transformations in the \psykal version.
+
 The main conclusion of all this analysis is that while we can employ a
@@
 
-The IBM results in particular demonstrate two key points.
+In particular, the results on the IBM system demonstrate two key points.
 %

### 2015-09-08 docs/papers/shallow_performance/shallow_perf.tex
parco paper: correct my psykal macro to not insert trailing blank space

@@
 
-\newcommand{\psykal}{{PS}y{KA}l\ }
+\newcommand{\psykal}{{PS}y{KA}l}
 
@@
 In this paper we evaluate the performance implications of applying a
-technique which we call \psykal to finite difference Ocean models. 
+technique which we call \psykal\ to finite difference Ocean models. 
 %
@@
 A finite difference shallow water benchmark optimised for cache based
-architectures is taken as the starting point. A vanilla \psykal version
+architectures is taken as the starting point. A vanilla \psykal\ version
 is written and the performance of the two compared. The optimisations
 that were applied to the original benchmark (loop fusion {\it etc.}) are then
-applied to the \psykal version as a set of code modifications to the
+applied to the \psykal\ version as a set of code modifications to the
 optimisation layer. Performance results are presented for the Cray,
@@
 %%
-Therefore, the \psykal approach can be used with negligable
+Therefore, the \psykal\ approach can be used with negligable
 performance loss and sometimes small performance gains compared to the
@@
 
-\subsection*{The \psykal Approach}
+\subsection*{The \psykal\ Approach}
 
-The \psykal approach attempts to address the problems described in the
+The \psykal\ approach attempts to address the problems described in the
 previous section. It separates code into three layers; the Algorithm
@@
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal version of Shallow retains the Input/Output
+we ensure that any \psykal\ version of Shallow retains the Input/Output
 capability of the original (it reads the problem specification from a
@@
 aids in limiting the optimisations that can be performed on the
-\psykal version to those that should also be applicable to full
+\psykal\ version to those that should also be applicable to full
 oceanographic models. Note that although we retain the I/O
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of Shallow, we made the kernels
+optimisation) of the \psykal\ version of Shallow, we made the kernels
 as fine-grained as possible. In this case, this resulted in eight
@@
 \includegraphics[width=100mm]{psykal_shallow}
-\caption{A schematic of the vanilla \psykal version of the Shallow code.}
+\caption{A schematic of the vanilla \psykal\ version of the Shallow code.}
 \label{FIG_psykal_shallow_structure}
@@
 applied at the edges of the model domain. In the case of Shallow we
-have periodic boundary conditions in both the $x$ and $y$ dimensions.
+have periodic boundary conditions in both the $x$- and $y$-directions.
 These are simply implemented by having additional (`halo')
@@
 values at the next time-step to be those at the current
-time-step. Profiling of the original code showed that these three
+time-step. Profiling of the original code showed that these 
 sections each account for approximately one third of the run-time.
@@
 The primary aim of this work is to determine whether any performance
-loss due to the \psykal separation of concerns can be negated by
+loss due to the \psykal\ separation of concerns can be negated by
 optimising the PSy layer and if so, what the key transformations are
@@
 used with each compiler. These flags are particularly important for
-the \psykal versions of the code. The performance of the original
+the \psykal\ versions of the code. The performance of the original
 version of the code is much less sensitive to the compiler
@@
 
-Beginning with the vanilla \psykal version, we then manually apply a
+Beginning with the vanilla \psykal\ version, we then manually apply a
 series of (cumulative) code transformations while obeying the \psykal
@@
 kernels. After this, we determined the transformations to perform by
-comparing the timing profiles of the \psykal and original versions of
+comparing the timing profiles of the \psykal\ and original versions of
 the code. We then tackled the sections resposible for the largest
@@
 heavy use of the optimisation reports produced by the compilers and
-how these changed when moving from the original to the \psykal version
+how these changed when moving from the original to the \psykal\ version
 of the code.
@@
 
-\item Move all kernel subroutines into the same module as as that
+\item Move all kernel subroutines into the same module as that
   containing the middle/PSy layer (`module inlining');
@@
 
-We shall see that these transformations do not always result in improved
-performance. Whether or not they do so depends both on the compiler
-and the problem size. We also emphasise that the aim of these
-optimisations is typically to recover, as far as is possible, the structure of
-the original version of the code. It may well be that transforming the
-code into some other structure would result in better performance on a
-particular architecture. However, exploring this optimisation space is
-beyond the scope of the present work.
+We shall see that these transformations do not always result in
+improved performance. Whether or not they do so depends both on the
+compiler and the problem size. We also emphasise that the aim of these
+optimisations is typically to recover, as far as is possible, the
+structure of the original version of the code. It may well be that
+transforming the code into some other structure would result in better
+performance on a particular architecture. However, exploring this
+optimisation space is beyond the scope of the present work.
 
 We explore the extent to which performance depends upon the problem
-size by using square domains of dimension 64, 128, 256, 512 and
+size by using square domains with side lengths of 64, 128, 256, 512 and
 1024. This range allows us to investigate what happens when cache is
@@
 then investigate the different code transformations that are required
-to recover good performance from the \psykal version of Shallow. Finally,
-we compare the most performant \psykal version  with the original
+to recover good performance from the \psykal\ version of Shallow. Finally,
+we compare the most performant \psykal\ version  with the original
 version of Shallow for each compiler/CPU combination.
@@
 
+Strangely, the third bar (Gnu compiler) in
+Figure~\ref{FIG_orig_perf_summary} shows the same drop in performance
+in moving from the $256^2$ case to the $512^2$ case, despite the
+larger last-level cache of the Ivy Bridge system. In order to
+investigate this further, we profiled the application to obtain
+hardware performance counter data. This revealed that the drop in
+performance is due to a XXX
+
+ With the
+Intel compiler this drop in efficiency only occurs when going to
+$512^2$. We therefore conclude that the drop in performance is due to
+the Gnu compiler generating code that YYY
+compared to that produced by the Intel and Cray
+compilers.
+
 For this compiler-friendly form of the code, there is generally little
 performance difference between the executables produced by the Gnu and
-Intel compilers on both the Haswell and Ivy Bridge CPUs. However, the
-executable produced by the Cray compiler generally performs
-significantly better for all but the largest domain size. Analysis of
-the optimisation reports produced by the Cray and Intel compilers did
-not give any clues as to the cause of this difference and an in-depth
-investigation is outside the scope of this paper.
+Intel compilers on both the Haswell and Ivy Bridge CPUs (with the
+exception of the $512^2$ case noted above). However, the executable
+produced by the Cray compiler generally performs significantly better
+for all but the largest domain size. Analysis of the optimisation
+reports produced by the Cray and Intel compilers did not give any
+clues as to the cause of this difference and an in-depth investigation
+is outside the scope of this paper.
 
-\subsection{Optimisations of the \psykal Version to Recover Performance}
+\subsection{Optimisations of the \psykal\ Version to Recover Performance}
 
-We now examine the performance of the \psykal version of Shallow and
+We now examine the performance of the \psykal\ version of Shallow and
 the code transformations that are required to reach the levels obtained
@@
 fits within last-level cache on all of the CPUs we are using here.
-The performance of the \psykal version after each transformation has
+The performance of the \psykal\ version after each transformation has
 been applied to the code is plotted in
@@
 \includegraphics[width=123mm]{opt_stages_256}
-\caption{The performance of the \psykal version of Shallow for the
+\caption{The performance of the \psykal\ version of Shallow for the
   $256^{2}$ domain at each stage of optimisation. Results are given as
@@
 With the Gnu compiler (first and third clusters in
-Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
+Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal\ version of
 Shallow only achieves ~5-6\% of the performance of the
@@
 performance are different (second and fourth clusters in
-Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal version of
+Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal\ version of
 Shallow achieves some 80\% of the performance of the original
@@
 The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
-evolution of the performance of the \psykal version of Shallow with
+evolution of the performance of the \psykal\ version of Shallow with
 the Cray compiler. Again, this has significant differences from the
 behaviour seen with the Gnu and Intel compilers. As with the Intel
-compiler, the performance of the Vanilla \psykal version is fairly
+compiler, the performance of the Vanilla \psykal\ version is fairly
 good at 74\% of that of the original. In contrast to the other
@@
 computational loops are over almost all elements of these arrays. This
-simple change gives a \psykal version that achieves 88\% of the
+simple change gives a \psykal\ version that achieves 88\% of the
 performance of the original.
@@
 field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
-with the preceeding loop. The resulting \psykal version now achieves
+with the preceeding loop. The resulting \psykal\ version now achieves
 94\% of the original performance. Two further steps are required to
@@
 Figure~\ref{FIG_opt_stages_256}) we see that performance of the
-vanilla \psykal version is only 5\% of the original and several
+vanilla \psykal\ version is only 5\% of the original and several
 transformations are required to achieve comparable performance. As
@@
 
-\subsection{Performance Comparison of the Original and \psykal Versions}
+\subsection{Performance Comparison of the Original and \psykal\ Versions}
 
-Finally, we compare the performance of the best \psykal version of
+Finally, we compare the performance of the best \psykal\ version of
 Shallow with that of the original. In
@@
   difference} between the performance of the original and the (best)
-\psykal versions of Shallow for each compiler/CPU combination. The
+\psykal\ versions of Shallow for each compiler/CPU combination. The
 most important feature of this plot is that the performance penalty
-incurred by the \psykal version is less than four percent with only
+incurred by the \psykal\ version is less than four percent with only
 one exception where it is 6.6\%.  In fact, for the Intel and Cray
-compilers, the \psykal version of Shallow is never more than 2\%
+compilers, the \psykal\ version of Shallow is never more than 2\%
 slower than the original and in some cases is faster.  This
@@
 original version of the code, despite the significant restructuring
-required by the \psykal approach.
+required by the \psykal\ approach.
 
@@
 version with that of the original version of the code. A negative value 
-indicates that the \psykal version is faster than the original.}
+indicates that the \psykal\ version is faster than the original.}
 \label{FIG_slowdown_summary}
@@
 
-Turning to the cases where the \psykal version is significantly
+Turning to the cases where the \psykal\ version is significantly
 {\em faster} than the original, we see from
@@
 The performance of the Power 8 system with the $512^2$ domain size is
-somewhat anomolous: the \psykal version of the code performs
+somewhat anomolous: the \psykal\ version of the code performs
 significantly better than the original and in addition, this problem
@@
 Power 8 system and this is further enhanced by the IBM-supplied
-transformations in the \psykal version.
+transformations in the \psykal\ version.
 
@@
 
-We have investigated the application of the \psykal separation of
+We have investigated the application of the \psykal\ separation of
 concerns approach to the domain of finite-difference ocean
@@
 %
-Second, the \psykal approach directly addresses the problem of
+Second, the \psykal\ approach directly addresses the problem of
 requiring different optimisations for different architectures and/or
@@
 In future work we will extend our performance portability analysis of
-the \psykal approach to shared memory parallelisation optimisations on
+the \psykal\ approach to shared memory parallelisation optimisations on
 different architectures using different parallelisation techniques (in

### 2015-09-08 docs/papers/shallow_performance/PSEA-2015_review_response.txt
parco paper: add doc giving response to reviewers

> Main:

> 1. The results can be broadly separated into i) comparison with the
> original 'shallow' benchmark and ii) Optimisation of the Psy
> layer. They are presented in this order but to my mind this seems like
> the wrong way round. I don't think that there is technically a 'right
> way round' as they are very separate, but I did expect to read about
> the optimisation first. Also while reading the document, it mentioned
> 'original and the (best) PsyKAl ...' on page 6, it left me wondering
> what 'best' is because I hadn't reached the bit about optimisation at
> that stage. Suggest to reorder.

We have split the Results section into three subsections and
re-ordered, as suggested.

> Other

> 2. space after 'PsyKal' and before ',' on second line of abstract.
Fixed.

> 3. para 2, page 2: should there be 'of' on line 3? in the
  programming of models?

We're talking here about programming models such as OpenMP, CUDA etc. rather
than about constructing scientific models.

> 4. para 4, page 3: is there a version number of shallow used in the tests?

GRAHAM.

> 5. last full para, page 3: x- and y-directions?
Fixed.

> 6. para 1, page 4: is it one third per section?
Re-worded.

> 7. page 5, second transformation: remove second 'as'
Fixed.

> 8. para 3, page 6: 64, 128 ... should these be squared also on the figure too.
Fixed.

> 9. para 4, page 6: in the statement 'for the first two bars...' I
  would say that this is also true for the third bar also. This is gnu
  on the IvyBridge. Is that understood?

> 10. para6, page 6: remove the before to.
> 11. para6, page 6: difference seems to be in italics.

> 12. para 2, page 8: could you reword this; "on Haswell loop fusion
 achieves more than 90% while on Ivy Bridge it is fully-fusing the
 first loop nest", I found it difficult to follow.

> 13. Figure 4 and labeling the transformations: The fact that you
  have labeled the run instances 0-10 means its difficult to compare
  the labels with the bars (for me at least). Rather than reading the
  shading, I found my self counting them and I kept getting mixed up
  because the sixth bar corresponds to transformation 5. Further, in
  the text on page 9, there is reference to bar 6 which I assume is
  transformation 6 but bar 7... I understand why its been laid out
  this way but I think the transformations/vanilla could be labeled
  1-11?

> 14. para 3, page 9: potentially change "operation of Figure 5 is
  fused" to "operation (lower loop in Figure 5) is fused."

> 15. pare 1, page 10: in the sentence; "remain constant for the
> duration of each trip of the time-stepping loop" I've never heard
> that terminology before, is that standard? I would expect to hear
> something like "remain constant for the duration of each iteration
> of the time-stepping loop"

> 16. the introduction of Figure 5 help to explain a bit more the
> detail of the applied transformations. This was good and it would
> have been nice to have had a bit more detail on the transformations
> from the start rather than disbursed through the text.

----------------------- REVIEW 2 ---------------------

> The paper presents a very detailed and rigorous analysis of 6 such
> combinations across 3 platforms and 4 compilers with overall promising
> results. However the description of the PSyKAI architecture and the
> responsibilities and interplay between the different layers does not
> become entirely clear. The authors present very little background and
> rationale as to how PSyKAI compares to similar approaches in the
> field. It should also be made very clear in the introduction that the
> PSyKAI layer needs to presently be written *manually*, since this
> makes the current approach effectively "hand-optimised code".

We have altered the text to make it clearer that in this work we are
manually optimising the PSy layer in order to learn what
transformations must be supported in our code generation system (which
will be the subject of a future paper).

> Clarification questions
> -----------------------

> Introduction

> 1) p2: In which way has the PSyKAI design been influenced by OP2? How does it compare?

RUPERT.

> 2) p2: It is unclear what is meant by "local fields" in the kernel
> context: what is the granularity? Does a kernel (as written by the
> scientist) operate on the entire partition? A colum? A vertical
> slice of the partition? A single point? For the PSyKAI layer to be
> automatically generated, the kernel interface would need to be be
> well defined.

The text has been re-worked slightly to emphasise that a Kernel, like
the Algorithm layer, has no concept of parallelism. The nature of the
field it operates on depends on the application.

> 3) p2: If the algorithm layer operates on a full field and the kernels
> on "local fields", is the PSyKAI layer responsible for the domain
> decomposition?

Yes it is. The text has been re-worked to emphasise this.

> 4) p3: Who implemented the "original" Shallow version benchmarked
> against and for which platform has it been optimised?

Paul Swarztrauber originally wrote the code but the header text
records modifications by at least two other people. The hardware that
they were optimising for is not known. A line has been added to this effect.

> 5) p3: What does "retaining the I/O capability" imply? Is bitwise
> reproducability of results a requirement?

It simply implies that the solution fields can be written to disk at a
user-defined interval. Text has been added to this effect. Bit-wise
reproducibility is not a requirement.

> 6) p4: it is unclear from Figure 1 and the description on the previous
> page whether invoke_tstep is a routine written by the scientist and
> 'transformed' or entirely the responsibility of PSyKAI. It the latter
> is the case, how can PSyKAI determine which kernels to call, in which
> order, which parameters to pass etc.?

> Methodology

> 7) p5: Some of the transformations (3, 4, 5, 7, 9, 10) seem to be very
> specific to the Shallow application. In which way do those represent
> optimisations 'independent of the science code' as claimed?

Of necessity, the code transformations required to recover performance
for a specific code are determined by the structure of that code. We
are not claiming that a general set of optimisations can be applied to
any given code. Instead, the PSyKAl approach separates the code which
must be optimised from the code describing the science. The
optimisations are restricted to the PSy layer and therefore do not
affect the scientific components of the code.

> 8) p5: Could transformations be encoded as rules in a reusable form?
> Presently every version of the PSyKAI layer is presumably a separate,
> independent implementation?

In this paper we are concerned with whether (and how) it is possible
to recover performance after applying the considerable restructuring
required to achieve the PSyKAl separation of concerns. To do this we
have manually written the PSy layer. As the reviewer correctly states,
this means we would have a separate implementation for each target
architecture; precisely the situation we are working to avoid. We are
therefore developing the PSyclone code-generation system to generate
and manipulate the PSy layer. This will be the subject of a future
paper.

> Conclusions

> 9) p11: As pointed out in the comments above it seems the PSyKAI
> layer would require more tightly specified interface to both the
> algorithmic and kernel layers to be suitable for automatic
> generation. It would be interesting to hear the authors' thoughts on
> this.

The reviewer is correct, however automatic generation is not the
subject of this paper. Suffice to say we have such interfaces and a
python-based code-generation system (PSyclone) that can generate correct
code. This wil be the subject of a future paper.

> 10) p11: How would these user-provided transformation recipes look like?

These recipes are implemented as python scripts that apply PSyclone
Transformations.

General comments / suggestions
------------------------------

> 1) There seems to be no preprint available yet for reference [2]

> 2) Figure 2/3: A better visual grouping of the bars by platform and
> compiler e.g. by using colour to differentiate compilers and hatching
> to different platforms would help with interpretation of the plots

We would love to use colour but the formatting guidlines say we can't.

> 3) Figure 4: The bars per cluster are quite hard to tell apart visually

### 2015-09-10 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add Flather paper to bib and cite it

@@
 conditions can be set as a clipped or a Flather's radiation
-condition (Flather, 1976). The bottom friction takes a semi-implicit
+condition~\cite{flather76}. The bottom friction takes a semi-implicit
 form for the sake of model stability. As done in the original version
@@
 version with that of the original version of the code. A negative value 
-indicates that the \psykal version is faster than the original.}
+indicates that the \psykal version is slower than the original.}
 \label{FIG_slowdown_summary}

### 2015-09-10 docs/papers/parco2015_abstract/psyclone_parco2015.tex

--- 2015-06-30 parco2015_abstract/psyclone_parco2015.tex
@@
 \address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
-\address[B]{IBM}
+\address[B]{IBM India}
 
@@
 This paper presents a domain specific approach to performance
-portability and code complexity reduction in finite element and finite
-difference codes. This approach has been developed for the UK Met
-Office's next generation atmospheric model which uses finite elements
-on a quasi-unform grid, and has also been prototyped on two finite
-difference Ocean model benchmarks, one of which is based on the NEMO
-ocean model. We first outline our approach and then present results
-for a range of compilers. These demonstrate that it is possible to
-retain and potentially improve upon the performance of an existing
-shallow water model while simultaneously reducing its code complexity.
+portability and the reduction of code complexity in finite-element and
+finite-difference codes. This approach has been developed for the UK
+Met Office's next generation atmospheric model which uses finite
+elements on a quasi-unform grid, and has also been prototyped on two
+finite difference Ocean model benchmarks, one of which is based on the
+NEMO ocean model. We first outline our approach and then present
+results for a range of compilers. These demonstrate that it is
+possible to retain and potentially improve upon the performance of an
+existing shallow water model while simultaneously reducing its code
+complexity.
 \end{abstract}
@@
 
-\subsection{The \psykal Approach}
+\section{The \psykal Approach}
 
-The \psykal approach separates code into three layers, the Algorithm
-layer, the PSy layer and the Kernel layer. Whilst this approach is
-general we have applied it to Atmosphere and Ocean models written in
+The \psykal approach separates code into three layers; the Algorithm
+layer, the PSy layer and the Kernel layer. While this approach is
+general, we have applied it to Atmosphere and Ocean models written in
 Fortran where domain decomposition is typically performed in the
@@
 The top layer, in terms of calling hierarchy, is the Algorithm
-layer. This layer specifies the algorithm that thescientist would like
+layer. This layer specifies the algorithm that the scientist would like
 to perform (in terms of calls to kernel and infrastructure routines)
@@
 fields may be domain decomposed, however the algorithm layer is not
-aware of this. It is the scientists responsibility to write this
+aware of this. It is the scientist's responsibility to write this
 algorithm layer.
@@
 kernel subroutines. As the Algorithm layer works on logically global
-fields and Kernel layer works on local fields the PSy layer is
+fields and Kernel layer works on local fields, the PSy layer is
 responsible for iterating over columns. It is also responsible for
-including any required distributed memory operations, such as halo
+including any required distributed-memory operations, such as halo
 swaps and reductions.
@@
 
-As an example, consider the following ``traditional'' code fragment.
+As an example, consider the following ``traditional'' code fragment:
 \begin{verbatim}
@@
 \end{verbatim}
-In the PSyKAl approach this code would be split into algorithm layer\ldots
+In the \psykal approach this code would be split into the {A}lgorithm
+layer\ldots
 \begin{verbatim}
@@
 \end{verbatim}
-and kernel layer\ldots
+and {K}ernel layer\ldots
 \begin{verbatim}
@@
 of the ``traditional'' code, and potentially improve on it, will be
-discussed in the talk.
+discussed in the remainder of this paper.
 
-\subsection{The `shallow' Program}
-For this work we have used a benchmarking program that solves the
-shallow-water equations on a bi-periodic plane following the
-finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
+\section{The `shallow' Program}
+
+For the work described here we have used a benchmarking program that
+solves the shallow-water equations on a bi-periodic plane, following
+the finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
 This software was originally written in 1984 by Paul Swarztrauber of
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of shallow, we made the kernels
-as fine-grained as possible. In this case, this resulted in eight
+optimisation) of the \psykal version of Shallow, we made the kernels
+as fine-grained as possible. This resulted in eight
 distinct kernels, each of which operated on a single field at a single
@@
   of a compiler on a given CPU is indicated by the specification of
-  the version of the compiler in the relevent element. No entry
+  the version of the compiler in the relevent element. The absence of an entry
   indicates that a compiler was not available/used on that CPU.}
@@
 version after it has been re-structured following the \psykal
-approach. These two versions of the code effectively provide upper and
-lower bounds, respectively, on the performance we expect to achieve.
+approach. These two versions of the code effectively provide
+approximate upper and lower bounds, respectively, on the performance
+we expect to achieve.
 
@@
 \centering
-\includegraphics[width=2.8in]{../slowdown_summary}
+\includegraphics[width=90mm]{slowdown_summary}
 \caption{Comparison of the performance of the best \psykal
@@
 that the \psykal approach has the potential to at least maintain (if
-not improve) performance while improving code
+not improve) performance while significantly improving code
 structure/maintainability.
 
-The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal version of a code. In our approach
-this problem is tackled using the PSyclone code-generation system
-which will be described in a future publication.
+The application of code transformations/optimisations to the
+middle/PSy layer is key to the performance of the \psykal version of a
+code. In our approach this problem is tackled using the PSyclone
+code-generation system which will be described in a future
+publication.
 

### 2015-09-10 shallow/Maxeler/CPUCode/GOceanCpuCode.c
Removed unecessary files from new branch

@@ -1,115 +1,0 @@
-#include <math.h>
-#include <stdio.h>
-#include <stdlib.h>
-
-#include "Maxfiles.h"
-#include "MaxSLiCInterface.h"
-
-int main(void)
-{
-	char size_str[] = "DOMAIN_SIZE";
-
-	max_file_t *maxfile = GOcean_init();
-	max_engine_t *engine = max_load(maxfile, "*");
-
-    long size = max_get_constant_uint64t(maxfile, size_str);
-    size *= size;
-	//int size = 256*256;
-
-	// lmem has to be aligned to the burst size
-	const int lmem_align = 48; // *8 = 384 bytes
-	int lmem_size = size;
-	if(size%lmem_align != 0){
-		lmem_size = lmem_align*(size/lmem_align + 1);
-	}
-
-	int sizeBytes = lmem_size * sizeof(double);
-	double *u = malloc(sizeBytes);
-	double *uold = malloc(sizeBytes);
-	double *v = malloc(sizeBytes);
-	double *vold = malloc(sizeBytes);
-	double *p = malloc(sizeBytes);
-	double *pold = malloc(sizeBytes);
-	// Output buffers
-	double *unew = malloc(sizeBytes);
-	double *vnew = malloc(sizeBytes);
-    double *pnew = malloc(sizeBytes);
-	double *uold_new = malloc(sizeBytes);
-	double *vold_new = malloc(sizeBytes);
-    double *pold_new = malloc(sizeBytes);
-
-	// TODO Generate input data using stream function as in original shallow
-	for(int i = 0; i < lmem_size; ++i) {
-		u[i] = (double)(random())/(double)RAND_MAX;
-		uold[i] = u[i];
-		v[i] = (double)(random())/(double)RAND_MAX;
-		vold[i] = v[i];
-		p[i] = (double)(random())/(double)RAND_MAX;
-		pold[i] = p[i];
-
-		unew[i] = (double)i;
-	}
-
-
-	GOcean_actions_t run_scalar;
-
-	run_scalar.param_N = size;
-    run_scalar.param_NLmem = lmem_size;
-
-	// Input buffers - stream in field{,_old}
-
-	// Data we stream straight into kernels
-	run_scalar.instream_p = p;
-	run_scalar.instream_pold = pold;
-	run_scalar.instream_u = u;
-	run_scalar.instream_uold = uold;
-	run_scalar.instream_v = v;
[…]

### 2015-09-10 shallow/OMP/wtime.c

@@ -1,20 +1,0 @@
-#include <sys/time.h>
-#ifdef _OPENMP
-#include <omp.h>
-#endif
-
-double wtime()
-{
-#if defined(_OPENMP) && (_OPENMP > 200010)
-   /* Use omp_get_wtime() if we can */
-   return omp_get_wtime();
-#else
-   /* Use a generic timer */
-   static int sec = -1;
-   struct timeval tv;
-   gettimeofday(&tv, (void *)0);
-   if (sec < 0) sec = tv.tv_sec;
-   return (tv.tv_sec - sec) + 1.0e-6*tv.tv_usec;
-#endif
-}
-    

### 2015-09-10 shallow/SEQ/apply_bcs_ct_mod.f90

@@ -1,91 +1,0 @@
-!> \brief Apply boundary conditions for field on CT
-!! \detail Applies cyclic boundary conditions for a 
-!! field defined on CT.
-module apply_bcs_ct_mod
-  use kind_params_mod
-  use kernel_mod
-  use argument_mod
-  implicit none
-
-  private
-
-  public invoke_apply_bcs_ct
-  public apply_bcs_ct_type, apply_bcs_ct_code
-
-  type, extends(kernel_type) :: apply_bcs_ct_type
-     type(arg), dimension(1) :: meta_args =    &
-          (/ arg(READWRITE, CT, POINTWISE)     & ! field
-           /)
-     !> This kernel writes only to internal points of the
-     !! simulation domain.
-     integer :: ITERATES_OVER = INTERNAL_PTS
-  contains
-    procedure, nopass :: code => apply_bcs_ct_code
-  end type apply_bcs_ct_type
-
-contains
-
-  !===================================================
-
-  !> Manual implementation of the code needed to invoke
-  !! apply_bcs_ct_code().
-  subroutine invoke_apply_bcs_ct(fld)
-    use field_mod
-    implicit none
-    type(r2d_field), intent(inout) :: fld
-    ! Locals
-    integer :: ihalo
-
-    ! Note that we do not loop over the full extent of the field.
-    ! Fields are allocated with extents (M+1,N+1).
-    ! Presumably the extra row and column are needed for periodic BCs.
-    ! We are updating the last row and the last column of a quantity on CT.
-    ! This looks like (using x to indicate a location that is written first
-    ! and y a location that is written second):
-    !
-    !  i=1   i=M
-    ! _ y  y  y  y 
-    ! /|o  o  o  x   j=N 
-    ! | o  o  o  x
-    ! \ o  o  o  x
-    !  \o  o  o _x   j=1
-    !    \______/|
-
-    !   vi-1j+1--fij+1---vij+1---fi+1j+1-vi+1j+1
-    !   |        |       |       |        |    
-    !   |        |       |       |        |    
-    !   Ti-1j----uij-----Tij-----ui+1j --Ti+1j--
-    !   |        |       |       |        |    
-    !   |        |       |       |        |    
-    !   vi-1j----fij-----vij-----fi+1j --vi+1j--
-    !   |        |       |       |        |    
[…]

### 2015-09-10 shallow/SEQ/apply_bcs_cu_mod.f90

@@ -1,95 +1,0 @@
-!> \brief Apply periodic boundary conditions for field defined on CU
-!! points.
-!! \detail Applies cyclic boundary conditions for a 
-!! field defined on CU.
-module apply_bcs_cu_mod
-  use kind_params_mod
-  use kernel_mod
-  use argument_mod
-  implicit none
-
-  private
-
-  public invoke_apply_bcs_cu
-  public apply_bcs_cu_type, apply_bcs_cu_code
-
-  type, extends(kernel_type) :: apply_bcs_cu_type
-     type(arg), dimension(1) :: meta_args =    &
-          (/ arg(READWRITE, CU, POINTWISE)     & ! field
-           /)
-     !> This kernel writes only to internal points of the
-     !! simulation domain.
-     integer :: ITERATES_OVER = INTERNAL_PTS
-  contains
-    procedure, nopass :: code => apply_bcs_cu_code
-  end type apply_bcs_cu_type
-
-contains
-
-  !===================================================
-
-  !> Manual implementation of the code needed to invoke
-  !! apply_bcs_cu_code().
-  subroutine invoke_apply_bcs_cu(fld)
-    use field_mod,    only: copy_field
-    use topology_mod, only: cu_grid
-    implicit none
-    real(wp), intent(inout), dimension(:,:) :: fld
-    ! Locals
-    integer :: ihalo
-
-    ! Note that we do not loop over the full extent of the field.
-    ! Arrays are allocated with extents (M+1,N+1) which gives the
-    ! necessary freedom to have the actual fields staggered.
-    ! The extra row and column are then also available for appling
-    ! periodic BCs.
-    ! We are updating the first column and last row of a quantity on CU.
-    ! This looks like (using x to indicate a location that is written first
-    ! and y a location that is written second):
-    !
-    !  i=1   i=M
-    ! _ y  y  y  y 
-    ! /|x  o  o  o   j=N 
-    ! | x  o  o  o
-    ! \ x  o  o  o
-    !  \x_ o  o  o   j=1
-    !   |\______/ 
-
-    !   vi-1j+1--fij+1---vij+1---fi+1j+1-vi+1j+1
-    !   |        |       |       |        |    
-    !   |        |       |       |        |    
-    !   Ti-1j----uij-----Tij-----ui+1j --Ti+1j--
-    !   |        |       |       |        |    
-    !   |        |       |       |        |    
[…]

### 2015-09-10 shallow/SEQ/apply_bcs_cv_mod.f90

@@ -1,91 +1,0 @@
-!> \brief Apply boundary conditions for field on CV
-!! \detail Applies cyclic boundary conditions for a 
-!! field defined on CV.
-module apply_bcs_cv_mod
-  use kind_params_mod
-  use kernel_mod
-  use argument_mod
-  implicit none
-
-  private
-
-  public invoke_apply_bcs_cv
-  public apply_bcs_cv_type, apply_bcs_cv_code
-
-  type, extends(kernel_type) :: apply_bcs_cv_type
-     type(arg), dimension(1) :: meta_args =    &
-          (/ arg(READWRITE, CV, POINTWISE)     & ! field
-           /)
-     !> This kernel writes only to external points
-     integer :: ITERATES_OVER = EXTERNAL_PTS
-  contains
-    procedure, nopass :: code => apply_bcs_cv_code
-  end type apply_bcs_cv_type
-
-contains
-
-  !===================================================
-
-  !> Manual implementation of the code needed to invoke
-  !! apply_bcs_cv_code().
-  subroutine invoke_apply_bcs_cv(fld)
-    use field_mod,    only: copy_field
-    use topology_mod, only: cv_grid
-    implicit none
-    real(wp), intent(inout), dimension(:,:) :: fld
-    ! Locals
-    integer :: ihalo
-
-    ! Note that we do not loop over the full extent of the field.
-    ! Fields are allocated with extents (M+1,N+1).
-    ! Presumably the extra row and column are needed for periodic BCs.
-    ! We are updating the first row and last col of a quantity on CV.
-    ! This looks like (using x to indicate a location that is written first
-    ! and y a location that is written second):
-    !
-    !  i=1   i=M
-    !  -o  o  o  y 
-    ! / o  o  o  y   j=N 
-    ! | o  o  o  y
-    ! \ o  o  o  y
-    !  \x  x  x _y   j=1
-    !    \______/|
-
-    !   vi-1j+1--fij+1---vij+1---fi+1j+1-vi+1j+1
-    !   |        |       |       |        |    
-    !   |        |       |       |        |    
-    !   Ti-1j----uij-----Tij-----ui+1j --Ti+1j--
-    !   |        |       |       |        |    
-    !   |        |       |       |        |    
-    !   vi-1j----fij-----vij-----fi+1j --vi+1j--
-    !   |        |       |       |        |    
[…]

### 2015-09-10 shallow/SEQ/model_mod.f90

@@ -1,129 +1,0 @@
-MODULE model_mod
-  use kind_params_mod
-  use field_mod
-  use shallow_io_mod
-  use timing_mod, ONLY: timer_init, timer_report
-  implicit none
-
-  integer :: itmax   !< number of timesteps
-
-  real(wp) :: dt  !< model timestep (seconds)
-  real(wp) :: tdt !< 2xdt apart from first step when is just dt
-
-  ! solution arrays
-  ! Fields are allocated with extents (M+1,N+1).
-  ! Presumably the extra row and column are needed for periodic BCs.
-
-  ! The finite-difference variables are defined on the mesh as follows:
-  !            u              u          
-  !    -------->-----P,H------>-----P,H
-  !    |              |              |   
-  !    |        rot   |        rot   |   
-  !v  /|\      o     /|\ v    o     /|\ v
-  !    |              |              |   
-  !    |              |              |   
-  !    -------->-----P,H------>-----P,H
-  !    |              |              |   
-  !    |        rot   |        rot   |   
-  !v  /|\      o     /|\ v    o     /|\ v
-  !    |              |              |   
-  !    |              |              |   
-  !    |------->------|------->------|   
-  !            u              u          
-
-  ! Calling the point at which P,H etc. are defined 'T' and the
-  ! point at which rotation/vorticity defined 'f' then:
-  !
-  !          f(i,j+1) v(i,j+1)  f(i+1,j+1)  v(i+1,j+1)  f(i+2,j+1)
-  !
-  !          u(i,j)   T(i,j)    u(i+1,j)    T(i+1,j)    u(i+2,j)
-  !
-  !          f(i,j)   v(i,j)    f(i+1,j)    v(i+1,j)    f...
-  !
-  !          u(i,j-1) T(i,j-1)  u(i+1,j-1)  T...        u...
-  !
-  !  So, T is dual of f.
-  !
-  ! This is consistent with the form of the original code:
-  !     DO J=1,N
-  !        DO I=1,M
-  !           CU(I+1,J) = .5*(P(I+1,J)+P(I,J))*U(I+1,J)
-  !           CV(I,J+1) = .5*(P(I,J+1)+P(I,J))*V(I,J+1)
-  !           Z(I+1,J+1) = &
-  !              (FSDX*(V(I+1,J+1)-V(I,J+1))-FSDY*(U(I+1,J+1)-U(I+1,J))) &
[…]

### 2015-09-10 shallow/SEQ/original/shallow_base.f90

@@ -1,595 +1,0 @@
-      PROGRAM shallow
-
-!     BENCHMARK WEATHER PREDICTION PROGRAM FOR COMPARING THE
-!     PREFORMANCE OF CURRENT SUPERCOMPUTERS. THE MODEL IS
-!     BASED OF THE PAPER - THE DYNAMICS OF FINITE-DIFFERENCE
-!     MODELS OF THE SHALLOW-WATER EQUATIONS, BY ROBERT SADOURNY
-!     J. ATM. SCIENCES, VOL 32, NO 4, APRIL 1975.
-!     
-!     CODE BY PAUL N. SWARZTRAUBER, NATIONAL CENTER FOR
-!     ATMOSPHERIC RESEARCH, BOULDER, CO,  OCTOBER 1984.
-!     Modified by Juliana Rew, NCAR, January 2006
-!
-!     In this version, shallow4.f, initial and calculated values
-!     of U, V, and P are written to a netCDF file
-!     for later use in visualizing the results. The netCDF data
-!     management library is freely available from
-!     http://www.unidata.ucar.edu/software/netcdf
-!     This code is still serial but has been brought up to modern
-!     Fortran constructs and uses portable intrinsic Fortran 90 timing routines
-!     This can be compiled on the IBM SP using:
-!     xlf90 -qmaxmem=-1 -g -o shallow4 -qfixed=132 -qsclk=micro \
-!     -I/usr/local/include shallow4.f -L/usr/local/lib32/r4i4 -l netcdf
-!     where the -L and -I point to local installation of netCDF
-!     
-!     Changes from shallow4.f (Annette Osprey, January 2010):
-!     - Converted to free-form fortran 90.  
-!     - Some tidying up of old commented-out code.   
-!     - Explicit type declarations.
-!     - Variables n, m, itmax and mprint read in from namelist. 
-!     - Dynamic array allocation.
-!     - Only write to netcdf at mprint timesteps.
-!     - Don't write wrap-around points to NetCDF file.
-!     - Use 8-byte reals. 
-
-
-      USE timing_mod
-      use shallow_io_mod, only: ascii_write
[…]

### 2015-09-10 shallow/SEQ/runme_loop_fuse.py

@@ -1,25 +1,0 @@
-from parse import parse,ParseError
-from psyGen import PSyFactory,GenerationError
-#from algGen import Alg
-api="gocean"
-filename="shallow_gocean.f90"
-ast,invokeInfo=parse(filename,api=api,invoke_name="invoke")
-psy=PSyFactory(api).create(invokeInfo)
-print psy.gen
-#alg=Alg(ast,psy)
-
-print psy.invokes.names
-schedule=psy.invokes.get('invoke_0').schedule
-schedule.view()
-
-from psyGen import TransInfo
-t=TransInfo()
-print t.list
-#lf=t.get_trans_name('DoubleLoopFuse')
-lf=t.get_trans_name('LoopFuse')
-
-newschedule,memento=lf.apply(schedule.children[0],schedule.children[1])
-#newschedule,memento=lf.apply(schedule.children[0].children[0].children[0],schedule.children[1].children[0].children[0])
-newschedule.view()
-#psy.invokes.get('invoke_0')._schedule=newschedule
-#print psy.gen

### 2015-09-10 shallow/SEQ/shallow_base.f90

@@ -1,222 +1,0 @@
-program shallow
-
-!> @mainpage BENCHMARK WEATHER PREDICTION PROGRAM FOR COMPARING THE
-!! PREFORMANCE OF CURRENT SUPERCOMPUTERS. THE MODEL IS
-!! BASED OF THE PAPER - THE DYNAMICS OF FINITE-DIFFERENCE
-!! MODELS OF THE SHALLOW-WATER EQUATIONS, BY ROBERT SADOURNY
-!! J. ATM. SCIENCES, VOL 32, NO 4, APRIL 1975.
-!!     
-!! CODE BY PAUL N. SWARZTRAUBER, NATIONAL CENTER FOR
-!! ATMOSPHERIC RESEARCH, BOULDER, CO,  OCTOBER 1984.
-!! Modified by Juliana Rew, NCAR, January 2006
-!
-!     In this version, shallow4.f, initial and calculated values
-!     of U, V, and P are written to a netCDF file
-!     for later use in visualizing the results. The netCDF data
-!     management library is freely available from
-!     http://www.unidata.ucar.edu/software/netcdf
-!     This code is still serial but has been brought up to modern
-!     Fortran constructs and uses portable intrinsic Fortran 90 timing routines
-!     This can be compiled on the IBM SP using:
-!     xlf90 -qmaxmem=-1 -g -o shallow4 -qfixed=132 -qsclk=micro \
-!     -I/usr/local/include shallow4.f -L/usr/local/lib32/r4i4 -l netcdf
-!     where the -L and -I point to local installation of netCDF
-!     
-!     Changes from shallow4.f (Annette Osprey, January 2010):
-!     - Converted to free-form fortran 90.  
-!     - Some tidying up of old commented-out code.   
-!     - Explicit type declarations.
-!     - Variables n, m, itmax and mprint read in from namelist. 
-!     - Dynamic array allocation.
-!     - Only write to netcdf at mprint timesteps.
-!     - Don't write wrap-around points to NetCDF file.
-!     - Use 8-byte reals. 
-!
-!     This version heavily modified as part of the GOcean-2D project
-!     with the mantra "all computation must occur in a kernel."
-!     Andrew Porter, April 2014
-
-  use shallow_io_mod
-  use timing_mod
-  use gocean_mod, only: model_write_log
[…]

### 2015-09-10 shallow/SEQ/time_step_mod.f90

@@ -1,171 +1,0 @@
-module time_step_mod
-  use kind_params_mod
-  use field_mod
-  implicit none
-
-contains
-
-  subroutine invoke_time_step(cufld, cvfld, ufld, unew, uold, &
-                              vfld, vnew, vold, &
-                              pfld, pnew, pold, &
-                              hfld, zfld, tdt)
-    use compute_cu_mod, only: compute_cu_code
-    use compute_cv_mod, only: compute_cv_code
-    use compute_z_mod,  only: compute_z_code
-    use compute_h_mod,  only: compute_h_code
-    use compute_unew_mod, only: compute_unew_code
-    use compute_vnew_mod, only: compute_vnew_code
-    use compute_pnew_mod, only: compute_pnew_code
-    use time_smooth_mod, only: time_smooth_code
-    use apply_bcs_mod, only: invoke_apply_bcs
-    implicit none
-    type(r2d_field), intent(inout) :: cufld, cvfld
-    type(r2d_field), intent(inout) :: unew, vnew, pnew
-    type(r2d_field), intent(inout) :: hfld, zfld, pfld, &
-                                               ufld, vfld
-    type(r2d_field), intent(inout) :: uold, vold, pold
-    real(wp),        intent(in)    :: tdt
-
-    ! Locals
-    integer :: I, J, idim1, idim2
-    real(wp) :: dx, dy
-
-    ! COMPUTE CAPITAL U, CAPITAL V, Z AND H
-
-    do J=cufld%internal%ystart, cufld%internal%ystop
-       do I=cufld%internal%xstart, cufld%internal%xstop
-
-          call compute_cu_code(i, j, cufld%data, pfld%data, ufld%data)
-       end do
-    end do
-    do J=cvfld%internal%ystart, cvfld%internal%ystop
-       do I=cvfld%internal%xstart, cvfld%internal%xstop
-
-          call compute_cv_code(i, j, cvfld%data, pfld%data, vfld%data)
-       end do
-    end do
-    dx = zfld%grid%dx
-    dy = zfld%grid%dy
-
-    do J=zfld%internal%ystart, zfld%internal%ystop, 1
-       do I=zfld%internal%xstart, zfld%internal%xstop, 1
-
-          call compute_z_code(i, j,      &
-                              zfld%data, &
-                              pfld%data, &
-                              ufld%data, &
-                              vfld%data, &
-                              dx, dy)
[…]

### 2015-09-10 nemolite2d_psykal/nemolite2d.f90
#2: undo loop-limits and in-lining in serial nemolite2d

@@ -30,7 +30,6 @@ program gocean2d
 
   ! time stepping index
   integer :: istp  
-  real(wp) :: rstp 
   integer :: itimer0
 
   ! Create the model grid. We use a NE offset (i.e. the U, V and F
@@ -83,9 +82,8 @@ program gocean2d
   do istp = nit000, nitend, 1
 
      !call model_write_log("('istp == ',I6)",istp)
-     rstp = real(istp, wp)
 
-     call step(rstp,                               &
+     call step(istp,                               &
                ua_fld, va_fld, un_fld, vn_fld,     &
                sshn_t_fld, sshn_u_fld, sshn_v_fld, &
                ssha_t_fld, ssha_u_fld, ssha_v_fld, &
@@ -125,7 +123,7 @@ subroutine step(istp,           &
   use gocean2d_io_mod, only: model_write
   implicit none
   !> The current time step
-  real(wp),           intent(in) :: istp
+  integer,            intent(in) :: istp
   type(r2d_field), intent(inout) :: un, vn, sshn, sshn_u, sshn_v
   type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
   type(r2d_field), intent(in) :: hu, hv, ht
@@ -133,28 +131,6 @@ subroutine step(istp,           &
   call invoke_time_step(istp, ssha, ssha_u, ssha_v, &
                         sshn, sshn_u, sshn_v, &
                         hu, hv, ht, ua, va, un, vn)
-
-!  call invoke(                                               &
-!              continuity(istp, ssha, sshn_t, sshn_u, sshn_v, &
-!                         hu, hv, un, vn),                    &
-!              momentum_u(ua, un, vn,                         &
-!                         ssha_u, sshn_t, sshn_u, sshn_v),    &
-!              momentum_v(va, un, vn, hu, hv, ht,             &
-!                         ssha_v, sshn_t, sshn_u, sshn_v),    &
-!              bc_ssh(istp, ssha),                            &
-!              bc_solid_u(ua),                                &
-!              bc_solid_v(va),                                &
-!              bc_flather_u(ua, hu, sshn_u),                  &
-!              bc_flather_v(va, hv, sshn_v),                  &
-!              copy_field(ua, un),                            &
-!              copy_field(va, vn),                            &
-!              copy_field(ssha, sshn_t),                      &
-!              next_sshu(sshn_u, sshn_t),                     &
[…]

### 2015-09-10 nemolite2d_psykal/time_step_mod.f90

@@ -16,19 +16,19 @@ contains
     use grid_mod
     use model_mod,       only: rdt, cbfr, visc
     use physical_params_mod, only: g, omega, d2r
-!    use momentum_mod,    only: momentum_v_code
-!    use momentum_mod,    only: momentum_u_code
-!    use continuity_mod,  only: continuity_code
-!    use time_update_mod, only: next_sshu_code, next_sshv_code
+    use momentum_mod,    only: momentum_v_code
+    use momentum_mod,    only: momentum_u_code
+    use continuity_mod,  only: continuity_code
+    use time_update_mod, only: next_sshu_code, next_sshv_code
     use boundary_conditions_mod
     implicit none
-    real(wp),        intent(in)    :: istp
+    integer,         intent(in)    :: istp
     type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
     type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
     type(r2d_field), intent(in)    :: hu, hv, ht
     ! Locals
     integer :: ji, jj, jiu, jiv
-    integer :: M, N, idxt
+    integer :: idxt
     ! Locals for momentum
     REAL(wp) :: u_e, u_w, v_n, v_s
     real(wp) :: v_nc, v_sc
@@ -43,468 +43,190 @@ contains
     ! Locals for BCs
     real(wp) :: amp_tide, omega_tide, rtime
 
-    M  = ssha%grid%simulation_domain%xstop
-    N  = ssha%grid%simulation_domain%ystop
+    !M  = ssha%grid%simulation_domain%xstop
+    !N  = ssha%grid%simulation_domain%ystop
 
-    ! In the general case we have to reason about whether or not the
-    ! domain has PBCs and what sort of offset convention the kernels
-    ! use. However, this is a middle layer specific to NEMOLite2D and
-    ! therefore we know that we have no periodic BCs and are using a
-    ! NE stagger
-    !txstart = 2 ! grid%simulation_domain%xstart
-    !tystart = 2 ! grid%simulation_domain%ystart
[…]

### 2015-09-10 api_v1.0/field_mod.f90
#2: Add OpenMP to field-copy routine

@@ -859,6 +859,25 @@ contains
     IMPLICIT none
     type(r2d_field), intent(in)    :: field_in
     type(r2d_field), intent(inout) :: field_out
+    integer :: ji, jj
+
+!$OMP PARALLEL DO SCHEDULE(RUNTIME) default(none) &
+!$OMP private(ji,jj) shared(field_out)
+    do jj= field_out%whole%ystart, field_out%whole%ystop
+       do ji = field_out%whole%xstart, field_out%whole%xstop
+          field_out%data(ji,jj) = field_in%data(ji,jj)
+       end do
+    end do
+!$OMP END PARALLEL DO
+        
+  end subroutine copy_2dfield
+
+  !===================================================
+
+  SUBROUTINE copy_2dfield_tiled(field_in, field_out)
+    IMPLICIT none
+    type(r2d_field), intent(in)    :: field_in
+    type(r2d_field), intent(inout) :: field_out
     integer :: it, ji, jj
 
 !$OMP DO SCHEDULE(RUNTIME)
@@ -871,7 +890,7 @@ contains
     end do
 !$OMP END DO
         
-  end subroutine copy_2dfield
+  end subroutine copy_2dfield_tiled
 
   !===================================================
 

### 2015-09-10 nemolite2d_psykal/time_step_mod.f90
#2: Removed un-used local vars

@@ -14,7 +14,7 @@ contains
     use timing_mod
     use field_mod
     use grid_mod
-    use model_mod,       only: rdt, cbfr, visc
+    use model_mod,       only: rdt
     use physical_params_mod, only: g, omega, d2r
     use momentum_mod,    only: momentum_v_code
     use momentum_mod,    only: momentum_u_code
@@ -27,21 +27,7 @@ contains
     type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
     type(r2d_field), intent(in)    :: hu, hv, ht
     ! Locals
-    integer :: ji, jj, jiu, jiv
-    integer :: idxt
-    ! Locals for momentum
-    REAL(wp) :: u_e, u_w, v_n, v_s
-    real(wp) :: v_nc, v_sc
-    real(wp) :: depe, depw, deps, depn
-    real(wp) :: hpg, adv, cor, vis
-    real(wp) :: dudx_e, dudx_w, dudy_s, dudy_n
-    real(wp) :: uu_e, uu_n, uu_s, uu_w
-    real(wp) :: u_ec, u_wc, vv_e, vv_n, vv_s, vv_w
-    real(wp) :: dvdx_e, dvdx_w, dvdy_n, dvdy_s
-    real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
-    ! end locals for momentum
-    ! Locals for BCs
-    real(wp) :: amp_tide, omega_tide, rtime
+    integer :: ji, jj
 
     !M  = ssha%grid%simulation_domain%xstop
     !N  = ssha%grid%simulation_domain%ystop

### 2015-09-10 nemolite2d_psykal/nemolite2d_omp_transform.py
#2: Change psyclone-omp script so that it puts an OMP PARALLEL DO around every loop in the invoke

@@ -10,8 +10,7 @@ psy=PSyFactory(api).create(invokeInfo)
 
 from psyGen import TransInfo
 t=TransInfo()
-ltrans = t.get_trans_name('GOceanOMPLoopTrans')
-rtrans = t.get_trans_name('OMPParallelTrans')
+ltrans = t.get_trans_name('GOceanOMPParallelLoopTrans')
 
 schedule=psy.invokes.get('invoke_0').schedule
 #schedule.view()
@@ -22,10 +21,6 @@ new_schedule=schedule
 for child in schedule.children:
     newschedule,memento=ltrans.apply(child)
     schedule = newschedule
-
-# Enclose all of these loops within a single OpenMP
-# PARALLEL region
-newschedule,memento = rtrans.apply(schedule.children)
 
 psy.invokes.get('invoke_0')._schedule=schedule
 print psy.gen

### 2015-09-10 api_v1.0/field_mod.f90
#2: remove use of default(none) in OMP directive in field_copy routine

@@ -862,7 +862,7 @@ contains
     integer :: ji, jj
 
 !$OMP PARALLEL DO SCHEDULE(RUNTIME) default(none) &
-!$OMP private(ji,jj) shared(field_out)
+!$OMP private(ji ,jj) shared(field_out, field_in)
     do jj= field_out%whole%ystart, field_out%whole%ystop
        do ji = field_out%whole%xstart, field_out%whole%xstop
           field_out%data(ji,jj) = field_in%data(ji,jj)

### 2015-09-10 nemolite2d_acc/nemolite2d.f90
#2: remove OpenACC versions since not applicable to this branch

@@ -1,1289 +1,0 @@
-    module sizes_mod
-
-         INTEGER,  PARAMETER :: sp = SELECTED_REAL_KIND(6, 37)
-         INTEGER,  PARAMETER :: dp = SELECTED_REAL_KIND(12, 307)
-         INTEGER,  PARAMETER :: wp = dp         
-                    
-   end module sizes_mod          
-                    
-    module cuda_mod
-         use cudafor
-         use sizes_mod
-         IMPLICIT NONE
-
-     contains    
-
-              
-        ! Assumes padded allocations
-        SUBROUTINE momentum_acc(un, vn, sshn, sshn_u, sshn_v, ssha_u, ssha_v, &
-                                                   hu, hv, ht, e1v, e2v, e1u, e2u, e1t, e2t, &
-                                                   gphiu, gphiv, e12u, e12v, pt, ua, va ,&
-                                                   ssha, e12t, &
-                                                   visc, omega, d2r, g, rdt, cbfr, jpi, jpj)
-                                               
-          REAL(wp), intent(in), device, dimension(-31:jpi,1:jpj) :: un, hu, gphiu, e1u, e2u, sshn_u, ssha_u, e12u
-          REAL(wp), intent(in), device, dimension(1:jpi,0:jpj)   :: vn, hv, gphiv, e1v, e2v, sshn_v, ssha_v, e12v
-          REAL(wp), intent(in), device, dimension(1:jpi,1:jpj)   :: sshn, ht, e1t, e2t, e12t
-          REAL(wp), intent(out), device, dimension(1:jpi,1:jpj)  :: ssha
-          
-          
-          INTEGER, intent(in), device, dimension(0:jpi+1+30,0:jpj+1) :: pt
-          INTEGER,  value, intent(in) :: jpi, jpj
-          REAL(wp), value, intent(in) :: visc, omega, d2r, g, rdt, cbfr
-          
-          REAL(wp), intent(out), device, dimension(-31:jpi,1:jpj) :: ua
[…]

### 2015-09-10 nemolite2d_psykal_acc/model_mod.f90

@@ -1,154 +1,0 @@
-MODULE model_mod
-  use kind_params_mod
-  use grid_mod
-  use field_mod
-  use timing_mod, ONLY: timer_init, timer_report
-  implicit none
-
-  public
-
-  !> start-end and record time steps
-  integer, save  :: nit000, nitend, irecord 
-  !> type (source?) of grid
-  integer, save  :: jphgr_msh
-
-  real(wp), save :: rdt             !< time step
-  REAL(wp), save :: cbfr            !< bottom friction coefficient
-  REAL(wp), save :: visc            !< backgroud/constant viscosity 
-
-  REAL(wp), save :: dep_const       !< constant depth
-
-  
-CONTAINS
-
-  !================================================
-
-  subroutine model_init(grid)
-    use gocean2d_io_mod
-    use gocean_mod, only: gocean_init
-    implicit none
-    type(grid_type), intent(inout) :: grid
-
-    !> Problem size, read from namelist
-    integer :: jpiglo, jpjglo
-    real(wp) :: dx, dy
-    integer  :: ierr
-    integer, dimension(:,:), allocatable :: tmask
-
-    ! Initialise timing system
-    call timer_init()
-
-    ! Read model configuration from namelist
-    call read_namelist(jpiglo, jpjglo, dx, dy, &
-                       nit000, nitend, irecord, &
-                       jphgr_msh, dep_const, rdt, cbfr, visc)
-
-    ! Set-up the T mask. This defines the model domain.
-    allocate(tmask(jpiglo,jpjglo), stat=ierr)
-    if(ierr /= 0)then
-       stop 'model_init: failed to allocate T mask'
-    end if
-
-    call setup_tpoints_mask(jpiglo, jpjglo, tmask)
-
-    ! Having specified the T points mask, we can set up mesh parameters
-    call grid_init(grid, jpiglo, jpjglo, dx, dy, tmask)
-
-    ! Initialise model IO 'system'
-    call model_write_init(jpiglo, jpjglo, irecord)
-
-    ! Clean-up. T-mask is now a part of the grid object.
-    deallocate(tmask)
-
-    ! Any other GOcean initialisation
-    call gocean_init()
-
-  end subroutine model_init
-
-  !================================================
-
-  SUBROUTINE model_finalise()
-    use gocean2d_io_mod, only: model_write_finalise
-    IMPLICIT none
-
[…]

### 2015-09-10 nemolite2d_psykal_acc/nemolite2d.f90

@@ -1,158 +1,0 @@
-program gocean2d
-  use timing_mod
-  use grid_mod
-  use field_mod
-  use initialisation_mod, only: initialisation
-  use model_mod
-  use gocean2d_io_mod, only: model_write
-  use gocean_mod,      only: model_write_log
-
-  !> A Horizontal 2D hydrodynamic ocean model which
-  !!   1) using structured grid
-  !!   2) using direct data addressing structures
-
-  implicit none
-
-  !> The grid on which our fields are defined
-  type(grid_type), target :: model_grid
-  !> Current ('now') sea-surface height at different grid points
-  type(r2d_field) :: sshn_u_fld, sshn_v_fld, sshn_t_fld
-  !> 'After' sea-surface height at different grid points
-  type(r2d_field) :: ssha_u_fld, ssha_v_fld, ssha_t_fld
-  !> Distance from sea-bed to mean sea level at the different grid points.
-  !! This is not time varying.
-  type(r2d_field) :: ht_fld, hu_fld, hv_fld
-  !> Current ('now') velocity components
-  type(r2d_field) :: un_fld, vn_fld
-  !> 'After' velocity components
-  type(r2d_field) :: ua_fld, va_fld
-
-  ! time stepping index
-  integer :: istp   
-  integer :: itimer0
-
-  ! Create the model grid. We use a NE offset (i.e. the U, V and F
-  ! points immediately to the North and East of a T point all have the
-  ! same i,j index).  This is the same offset scheme as used by NEMO.
-  model_grid = grid_type(ARAKAWA_C, &
-  !  BC_PERIODIC, BC_NON_PERIODIC ??
-                         (/BC_EXTERNAL,BC_EXTERNAL,BC_NONE/), &
-                         OFFSET_NE)
-
-  !! read in model parameters and configure the model grid 
-  CALL model_init(model_grid)
-
-  ! Create fields on this grid
-
-  ! Sea-surface height now (current time step)
-  sshn_u_fld = r2d_field(model_grid, U_POINTS)
-  sshn_v_fld = r2d_field(model_grid, V_POINTS)
-  sshn_t_fld = r2d_field(model_grid, T_POINTS)
-
-  ! Sea-surface height 'after' (next time step)
-  ssha_u_fld = r2d_field(model_grid, U_POINTS)
-  ssha_v_fld = r2d_field(model_grid, V_POINTS)
-  ssha_t_fld = r2d_field(model_grid, T_POINTS)
-
-  ! Distance from sea-bed to mean sea level
[…]

### 2015-09-10 nemolite2d_psykal_acc/time_step_mod.f90

@@ -1,621 +1,0 @@
-module time_step_mod
-  use kind_params_mod
-  implicit none
-
-  private
-
-  public invoke_time_step
-
-contains
-
-  subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
-                              sshn_t, sshn_u, sshn_v, &
-                              hu, hv, ht, ua, va, un, vn)
-    !> This routine is purely a wrapper that converts references to
-    !! data arrays within derived types to straightforward data
-    !! arrays in the called routine. This is to stop compilers
-    !! from panicking at the sight of a '%' symbol!
-    use field_mod
-    use grid_mod
-    implicit none
-    integer,         intent(in)    :: istp
-    type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
-    type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
-    type(r2d_field), intent(in)    :: hu, hv, ht
-                     
-    call invoke_time_step_arrays(istp,                                  &
-                                 ua%grid%nx, ua%grid%ny,                &
-                                 ua%grid%simulation_domain%xstop,       &
-                                 ua%grid%simulation_domain%ystop,       &
-                                 ua%grid%area_t,                        &
-                                 ua%grid%area_u,                        &
-                                 ua%grid%area_v,                        &
-                                 ua%grid%dx_u,                          &
-                                 ua%grid%dx_v,                          &
-                                 ua%grid%dx_t,                          &
-                                 ua%grid%dy_u,                          &
-                                 ua%grid%dy_v,                          &
-                                 ua%grid%dy_t,                          &
[…]

### 2015-09-14 api_v1.0/field_mod.f90
#2: change from multiple omp parallel dos to single omp parallel region

@@ -861,15 +861,14 @@ contains
     type(r2d_field), intent(inout) :: field_out
     integer :: ji, jj
 
-!$OMP PARALLEL DO SCHEDULE(RUNTIME) default(none) &
-!$OMP private(ji ,jj) shared(field_out, field_in)
+!$OMP DO SCHEDULE(RUNTIME)
     do jj= field_out%whole%ystart, field_out%whole%ystop
        do ji = field_out%whole%xstart, field_out%whole%xstop
           field_out%data(ji,jj) = field_in%data(ji,jj)
        end do
     end do
-!$OMP END PARALLEL DO
-        
+!$OMP END DO
+
   end subroutine copy_2dfield
 
   !===================================================

### 2015-09-14 nemolite2d_psykal/nemolite2d_omp_transform.py

@@ -10,7 +10,8 @@ psy=PSyFactory(api).create(invokeInfo)
 
 from psyGen import TransInfo
 t=TransInfo()
-ltrans = t.get_trans_name('GOceanOMPParallelLoopTrans')
+ltrans = t.get_trans_name('GOceanOMPLoopTrans')
+rtrans = t.get_trans_name('OMPParallelTrans')
 
 schedule=psy.invokes.get('invoke_0').schedule
 #schedule.view()
@@ -21,6 +22,10 @@ new_schedule=schedule
 for child in schedule.children:
     newschedule,memento=ltrans.apply(child)
     schedule = newschedule
+
+# Enclose all of these loops within a single OpenMP
+# PARALLEL region
+newschedule,memento = rtrans.apply(schedule.children)
 
 psy.invokes.get('invoke_0')._schedule=schedule
 print psy.gen

### 2015-09-14 nemolite2d_psykal/time_step_mod.f90

@@ -34,8 +34,9 @@ contains
 
     !call timer_start('Continuity',idxt)
 
-!$OMP PARALLEL DO SCHEDULE(RUNTIME), default(none), &
-!$OMP private(ji,jj) shared(ssha, sshn_t, sshn_u, sshn_v, hu, hv, un, vn, rdt)
+!$OMP PARALLEL default(shared), private(ji,jj)
+
+!$OMP DO SCHEDULE(RUNTIME)
     do jj = ssha%internal%ystart, ssha%internal%ystop, 1
       do ji = ssha%internal%xstart, ssha%internal%xstop, 1
 !    do jj = 2, N, 1
@@ -48,14 +49,13 @@ contains
                              rdt, sshn_t%grid%area_t)
       end do
     end do
-!$OMP END PARALLEL DO
+!$OMP END DO
 
     !call timer_stop(idxt)
 
     !call timer_start('Momentum',idxt)
 
-!$OMP PARALLEL DO SCHEDULE(RUNTIME), default(none) private(ji,jj) &
-!$OMP shared(ua, un, vn, hu, hv, ht, ssha_u, sshn_t, sshn_u, sshn_v)
+!$OMP DO SCHEDULE(RUNTIME)
     do jj = ua%internal%ystart, ua%internal%ystop, 1
       do ji = ua%internal%xstart, ua%internal%xstop, 1
 !    do jj = 2, N, 1
@@ -77,12 +77,11 @@ contains
 
       end do
     end do
-!$OMP END PARALLEL DO
+!$OMP END DO
 
 !    do jj = 2, N-1, 1
 !      do ji = 2, M, 1
-!$OMP PARALLEL DO SCHEDULE(RUNTIME), default(none) private(ji,jj) &
-!$OMP shared(va, un, vn, hu, hv, ht, ssha_v, sshn_t, sshn_u, sshn_v)
+!$OMP DO SCHEDULE(RUNTIME)
     do jj = va%internal%ystart, va%internal%ystop, 1
       do ji = va%internal%xstart, va%internal%xstop, 1
 
@@ -102,7 +101,7 @@ contains
 
       end do
     end do
-!$OMP END PARALLEL DO
+!$OMP END DO
 
     !call timer_stop(idxt)
 
@@ -110,8 +109,7 @@ contains
 
     !call timer_start('BCs', idxt)
 
-!$OMP PARALLEL DO SCHEDULE(RUNTIME), default(none) private(ji,jj) &
[…]

### 2015-09-14 nemolite2d_psykal/nemolite2d_omp_transform.py
#2: add fusion of field-copy loops to code-gen script

@@ -12,16 +12,30 @@ from psyGen import TransInfo
 t=TransInfo()
 ltrans = t.get_trans_name('GOceanOMPLoopTrans')
 rtrans = t.get_trans_name('OMPParallelTrans')
+ftrans = t.get_trans_name('GOceanLoopFuse')
 
 schedule=psy.invokes.get('invoke_0').schedule
 #schedule.view()
 new_schedule=schedule
+
+# Fuse the three field-copy operations
+# First the outer loops...
+new_schedule, memento = ftrans.apply(schedule.children[8],
+                                     schedule.children[9])
+new_schedule, memento = ftrans.apply(new_schedule.children[8],
+                                     new_schedule.children[9])
+# Then the inner loops...
+new_schedule, memento = ftrans.apply(schedule.children[8].children[0],
+                                     schedule.children[8].children[1])
+new_schedule, memento = ftrans.apply(schedule.children[8].children[0],
+                                     schedule.children[8].children[1])
 
 # Apply the OpenMP Loop transformation to *every* loop 
 # in the schedule
 for child in schedule.children:
     newschedule,memento=ltrans.apply(child)
     schedule = newschedule
+
 
 # Enclose all of these loops within a single OpenMP
 # PARALLEL region

### 2015-09-14 docs/papers/shallow_performance/shallow_perf.tex
parco paper: finish new para describing Gnu slowdown in going from 256 to 512.

@@
 The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
-3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
-system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
+3.7~GHz and 10~MB of last-level cache (LLC). The Intel Ivy
+Bridge-based system (Xeon E5-2697) has a clock speed of 2.7~GHz and a
+LLC of 30~MB. The CPU in the IBM Power-8 system is built around 12
 `chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
@@
 is shared with all other chiplets so that there is a total of 96~MB of
-last-level cache. The cores on the system we used had a clock speed of
-3.3~GHz.
+LLC. The cores on the system we used had a clock speed of 3.3~GHz.
 
@@
 hardware considered here. This summary demonstrates the effect of the
-larger last-level cache of the Intel Ivy Bridge system compared to
+larger LLC of the Intel Ivy Bridge system compared to
 that of the Haswell system; note the drop in performance
@@
 in moving from the $256^2$ case to the $512^2$ case, despite the
-larger last-level cache of the Ivy Bridge system. In order to
-investigate this further, we profiled the application to obtain
-hardware performance counter data. This revealed that the drop in
-performance is due to a XXX
-
- With the
-Intel compiler this drop in efficiency only occurs when going to
-$512^2$. We therefore conclude that the drop in performance is due to
-the Gnu compiler generating code that YYY
-compared to that produced by the Intel and Cray
-compilers.
+larger LLC of the Ivy Bridge system. In order to investigate this
+further, we profiled the application to obtain hardware performance
+counter data for the Gnu- and Intel-compiled executables. This
+revealed that the Gnu-compiled executable made more references to
+memory and had a higher LLC miss-rate compared with the Intel binary.
+We conclude that the Gnu compiler is not utilising registers as
+efficiently as the Intel compiler and that this results in greater
+memory traffic.
 
@@
 Shallow with that of the original. In
-Figure~\ref{FIG_slowdown_summary} we plot the percentage {\em
-  difference} between the performance of the original and the (best)
+Figure~\ref{FIG_slowdown_summary} we plot the percentage
+  difference between the performance of the original and the (best)
 \psykal\ versions of Shallow for each compiler/CPU combination. The

### 2015-09-14 docs/papers/shallow_performance/shallow_perf.tex
parco paper: re-worded paragraph describing transormations required by Gnu compiler

@@
 achieves more than 80\% of the performance of the original. The
-transformations required to get beyond this level then become
-architecture dependent: on Haswell loop fusion achieves more than 90\%
-while on Ivy Bridge it is fully-fusing the first loop nest. The
-dramatic performance improvement obtained from the module-inlining
-transformation demonstrates that the Gnu compiler is unable to
-optimise over separate files.
+transformation(s) required to get more than 90\% then become
+architecture dependent: on Haswell it is just loop fusion while on Ivy
+Bridge we do not exceed 90\% until we have done all of the steps up to
+and including fully-fusing (fusing the inner loops of) the first loop
+nest. The dramatic performance improvement obtained from the
+module-inlining transformation demonstrates that the Gnu compiler is
+unable to optimise over separate files.
 

### 2015-09-15 docs/papers/shallow_performance/shallow_perf.tex
parco paper: added labels to optimisation plot. Changed numbering of transformations so indexed from 1 instead of 0.

@@
 \centering
-\includegraphics[width=123mm]{opt_stages_256}
+\includegraphics[width=\textwidth]{opt_stages_256}
 \caption{The performance of the \psykal\ version of Shallow for the
@@
   a percentage of the performance of the original code for each
-  compiler/CPU combination.}
+  compiler/CPU combination. As an aid to understanding, significant
+  performance jumps are labelled with the corresponding transformation
+  number.}
 \label{FIG_opt_stages_256}

### 2015-09-15 docs/papers/shallow_performance/shallow_perf.tex
parco paper: more work on explicitly using numbering of transformations

@@
 Beginning with the vanilla \psykal\ version, we then manually apply a
-series of (cumulative) code transformations while obeying the \psykal
+series of (cumulative) code transformations while obeying the \psykal\ 
 separation of concerns, {\it i.e.} optimisation is restricted to the
@@
 \begin{enumerate}
+
+\item No optimisation: the vanilla \psykal\ version of the code;
 
@@
 
-As indicated above, the last two of these transformations were
+As indicated above, transformations 10 and 11 were
 suggested by the IBM compiler team. The second of these effectively
-undoes our 3rd transformation by fissioning the first loop nest back
+undoes transformation four by fissioning the first loop nest back
 into three separate loop nests.

### 2015-09-15 docs/papers/shallow_performance/PSEA-2015_review_response.txt
parco paper: updates to reviewer response doc
parco paper: further tweaks for first reviewer

@@
-> Main:
-
 > 1. The results can be broadly separated into i) comparison with the
@@
   on the IvyBridge. Is that understood?
+Text has been added to explain this.
 
-> 10. para6, page 6: remove the before to.
 > 11. para6, page 6: difference seems to be in italics.
+Fixed.
 
@@
  first loop nest", I found it difficult to follow.
+Re-worded.
 
@@
   1-11?
+We agree and have re-numbered the transformations from 1-11. We have
+also added labels to the key optimisation steps in the Figure and
+refer to the transformations by number in the text.
 
@@
   fused" to "operation (lower loop in Figure 5) is fused."
+Re-worded.
 
@@
 > of the time-stepping loop"
+Re-worded.
 
@@
 > from the start rather than disbursed through the text.
+
+DOES SPACE PERMIT ANY MORE DETAIL?
 

### 2015-09-15 docs/papers/shallow_performance/shallow_perf.tex
parco paper: further tweaks for first reviewer

@@
 flexibility in applying code modifications to the optimisation layer
-in the PSyKAl version allows for a significant performance improvement
+in the \psykal\ version allows for a significant performance improvement
 on the IBM Power8 compared with the original code.
@@
 version of the code. We also benchmark the vanilla, unoptimised
-version after it has been re-structured following the \psykal
+version after it has been re-structured following the \psykal\
 approach. These two versions of the code effectively provide upper and
@@
 how these changed when moving from the original to the \psykal\ version
-of the code.
+of the code. The transformations we have performed are as follows:
 
-The transformations we have performed are as follows:
-\begin{enumerate}
+\newcounter{tnum}
+
+\begin{list}{T\arabic{tnum}.}{\usecounter{tnum}\setlength{\rightmargin}{\leftmargin}}
 
@@
 
-\end{enumerate}
+\end{list}
 
-As indicated above, transformations 10 and 11 were
+As indicated above, transformations T10 and T11 were
 suggested by the IBM compiler team. The second of these effectively
@@
 We now examine the performance of the \psykal\ version of Shallow and
-the code transformations that are required to reach the levels obtained
-by the original.  We do this for the $256^{2}$ problem size since this
-fits within last-level cache on all of the CPUs we are using here.
-The performance of the \psykal\ version after each transformation has
-been applied to the code is plotted in
+the code transformations that are required to reach the levels
+obtained by the original.  We do this for the $256^{2}$ problem size
+since this fits within LLC on all of the CPUs we are using here.  The
+performance of the \psykal\ version after each transformation has been
+manually applied to the code is plotted in
 Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
@@
 original. Simply copying the code for each of the kernel subroutines
-into the module from which they are called (module in-lining) has the
+into the module from which they are called (module in-lining, T3) has the
 most dramatic effect on the performance of the compiled code; it now
@@
 transformation(s) required to get more than 90\% then become
-architecture dependent: on Haswell it is just loop fusion while on Ivy
+architecture dependent: on Haswell it is just loop fusion (T4) while on Ivy
 Bridge we do not exceed 90\% until we have done all of the steps up to
-and including fully-fusing (fusing the inner loops of) the first loop
-nest. The dramatic performance improvement obtained from the
-module-inlining transformation demonstrates that the Gnu compiler is
-unable to optimise over separate files.
+and including T8; fully fusing (fusing the inner loops of)
+the first loop nest. The dramatic performance improvement obtained
+from the module-inlining (T3) demonstrates that the Gnu
+compiler is unable to optimise over separate source files.
 
@@
 version. Only two code transformations were required to increase this
-to $\sim$100\%. The first of these was the fusion of the computational
-loops. However, we found that performance suffered (bar 5) when this
+to $\sim$100\%. The first of these was T4, the fusion of the computational
+loops. However, we found that performance suffered (T6) when this
 was extended to include the loop required for field copies at the end
@@
 of the kernel called from within it. Once this in-lining was done by
-hand (bar 6), the compiler was able to vectorise the loop and the
+hand (T7), the compiler was able to vectorise the loop and the
 performance of the original version of Shallow was recovered on the
 Haswell system. On Ivy Bridge the first loop nest also had to be fully
-fused (bar 7) to recover $\sim100\%$ of the performance.
+fused (T8) to recover $\sim$100\% of the performance.
 
@@
 behaviour seen with the Gnu and Intel compilers. As with the Intel
-compiler, the performance of the Vanilla \psykal\ version is fairly
+compiler, the performance of the Vanilla \psykal\ version, T1,  is fairly
 good at 74\% of that of the original. In contrast to the other
@@
 
-Subsequent transformations actually harm performance until the
-field-copy operation of Figure~\ref{FIG_time_smooth_code} is fused
-with the preceeding loop. The resulting \psykal\ version now achieves
-94\% of the original performance. Two further steps are required to
-match the latter. First, the first loop nest has to be fully fused
-(both inner and outer loops fused) and second, the array/loop bounds
-are passed as an argument to the middle layer. This is done in order
-to indicate to the compiler that these bounds remain constant for the
-duration of each trip of the time-stepping loop.
+Subsequent transformations actually harm performance until T6 where
+the field-copy operation (lower loop in
+Figure~\ref{FIG_time_smooth_code}) is fused with the preceeding
+loop. The resulting \psykal\ version now achieves 94\% of the original
+performance. Two further steps are required to match the latter: the
+first loop nest is fully fused (T8 - both inner and outer loops fused)
+and second, the array/loop bounds are passed as an argument to the
+middle layer (T9). This is done in order to indicate to the compiler
+that these bounds remain constant for the duration of each iteration
+of the time-stepping loop.
 
@@
 transformations are required to achieve comparable performance. As
-with the Cray compiler, specifying array bounds explicitly improves
-performance. Module-inlining also has a significant effect; doubling
+with the Cray compiler, specifying array bounds explicitly, T2, improves
+performance. Module-inlining, T3, also has a significant effect; doubling
 the performance from 10 to 20\% of the original. However, fusing loop
 nests actually harms performance, just as it did for the Cray
-compiler. Kernel inlining takes performance to over 50\% and it is the
-IBM-supplied transformations of loop unrolling and fissioning the
-first loop ({\it i.e.} undoing some of transformations three and
-seven) that finally recovers 96\% of the performance of the
-original. We note that this final pair of transformations significantly
+compiler. Kernel inlining (T7) takes performance to over 50\% and it is the
+IBM-supplied transformations of T10 (loop unrolling) and T11 (fissioning of the
+first loop) that finally recovers 96\% of the performance of the
+original. (\textit{i.e.} we must undo some of transformations three and
+seven). We note that this final pair of transformations significantly
 reduce the performance obtained with the Intel and Gnu compilers
@@
 \includegraphics[width=120mm]{slowdown_summary}
-\caption{Comparison of the performance of the best \psykal
+\caption{Comparison of the performance of the best \psykal\
 version with that of the original version of the code. A negative value 

### 2015-09-15 docs/papers/shallow_performance/PSEA-2015_review_response.txt
parco paper: updates to response to second reviewer

@@
 
+PSyKAl is a design approach, not a tool. Figure 1 illustrates the
+structure of the Shallow code after it has been re-structured
+following PSyKAl principles. How the code in the different layers
+is produced is not the subject of this paper.
+
 > Methodology
@@
 
+RUPERT/GRAHAM
+
 > 2) Figure 2/3: A better visual grouping of the bars by platform and
@@
 > to different platforms would help with interpretation of the plots
+> 3) Figure 4: The bars per cluster are quite hard to tell apart visually
 
-We would love to use colour but the formatting guidlines say we can't.
-
-> 3) Figure 4: The bars per cluster are quite hard to tell apart visually
+We would love to use colour but the formatting guidlines say we
+can't. However, we have added additional labelling to what was Figure
+3 in order to aid interpretation.

### 2015-10-01 docs/papers/shallow_performance/shallow_perf.tex
Modifications to the paper to provide a little information about how PSyKAl relates to OP2 and to fix various typos and inconsistencies.

@@
 compiler and architecture combinations on all tested problem
-sizes. The only exception to this (other than where we see significant
-speed-ups) is the Gnu compiler on Haswell for one problem size. Our
+sizes. The only exception to this (other than where we see performance
+improvement) is the Gnu compiler on Haswell for one problem size. Our
 tests indicate that this may be due to immature support for that
 architecture in the Gnu compiler -- no such problem is seen on the Ivy
-Bridge system.
+Bridge system. Further, the original code performed poorly using the
+IBM compiler on Power8 and needed to be modified to obtain performant
+code.
 %%
-Therefore, the \psykal\ approach can be used with negligable
+Therefore, the \psykal\ approach can be used with negligible
 performance loss and sometimes small performance gains compared to the
@@
 
-While this approach is general, we are currently applying it to
+In common with OP2, the \psykal approach separates out the science
+code and the performance related code and does this via a separate
+layer. The calls that specify parallelism in both approaches are
+similar in terms of where they are placed in the code and in their
+semantics but the \psykal approach supports the specification of more
+than one Kernel in a parallel region of code, compared with one for
+OP2, giving more scope for optimisation. Another difference is that
+the description metadata is provided with the Kernel in the \psykal
+approach whereas it is provided as part of the Kernel call in OP2.
+
+While the \psykal approach is general, we are currently applying it to
 Atmosphere and Ocean models written in Fortran where domain
@@
 and logically operates on full fields. We say logically here as the
-fields may be domain decomposed, however the algorithm layer is not
+fields may be domain decomposed, however the Algorithm layer is not
 aware of this. It is the scientist's responsibility to write this
-algorithm layer.
+Algorithm layer.
 
@@
 software (such as compiler, operating system, MPI implementation)
-configuration with no change to the algorithm or kernel layer
+configuration with no change to the Algorithm or Kernel layer
 code. This approach therefore offers the potential for portable
@@
 complete record of the people involved and the precise hardware they
-were targetting. In describing our work, we term the version of the
+were targeting. In describing our work, we term the version of the
 Shallow program obtained at the beginning of this project the
@@
 the PSy layer) and a call to the I/O system
-(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
+(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy layer
 routine then consists of applying each of the kernels to all of the
 points on the model mesh requiring an update. In its basic,
-unoptimised (`vanilla') form, this PSy-layer routine then contains a
+unoptimised (`vanilla') form, this PSy layer routine then contains a
 doubly-nested loop around each kernel call, as indicated by the
@@
 the domain. In the current work, the lines of code to do these copies
-are manually inserted in the PSy-layer routine as required.
+are manually inserted in the PSy layer routine as required.
 
@@
   of a compiler on a given CPU is indicated by the specification of
-  the version of the compiler in the relevent element. No entry
+  the version of the compiler in the relevant element. No entry
   indicates that a compiler was not available/used on that CPU.}
@@
 separation of concerns, {\it i.e.} optimisation is restricted to the
-middle, {PS}y layer and leaves the kernel and algorithm layers
+middle, {PS}y layer and leaves the Kernel and Algorithm layers
 unchanged. The aim of these optimisations is to recover, as much as is
@@
 comparing the timing profiles of the \psykal\ and original versions of
-the code. We then tackled the sections resposible for the largest
+the code. We then tackled the sections responsible for the largest
 slowdowns and iterated this process. While doing this we also made
@@
 
-\item Manually in-line all kernel bodies into the middle-layer
+\item Manually in-line all kernel bodies into the middle layer
   subroutine (kernel inlining);
@@
 Figure~\ref{FIG_opt_stages_256}. Results for a given compiler/CPU are
-given relative to the performance of the original version of Shallow
+presented relative to the performance of the original version of Shallow
 with the same compiler/CPU.
@@
 architecture dependent: on Haswell it is just loop fusion (T4) while on Ivy
-Bridge we do not exceed 90\% until we have done all of the steps up to
+Bridge we do not exceed 90\% until we have applied all of the steps up to
 and including T8; fully fusing (fusing the inner loops of)
@@
 good at 74\% of that of the original. In contrast to the other
-compilers consisdered so far, the first significant transformation is
+compilers considered so far, the first significant transformation is
 simply to specify the bounds of the arrays being used within the
@@
 the field-copy operation (lower loop in
-Figure~\ref{FIG_time_smooth_code}) is fused with the preceeding
+Figure~\ref{FIG_time_smooth_code}) is fused with the preceding
 loop. The resulting \psykal\ version now achieves 94\% of the original
@@
 Finally, we compare the performance of the best \psykal\ version of
-Shallow with that of the original. In
+Shallow with that of the original for different problem sizes. In
 Figure~\ref{FIG_slowdown_summary} we plot the percentage
@@
 part to the fact that the problem executes very quickly and therefore
-a small reduction in run-time will show as a large precentage. Since
+a small reduction in run-time will show as a large percentage. Since
 this problem is well contained in L2 cache, its performance will also
@@
 The performance of the Power 8 system with the $512^2$ domain size is
-somewhat anomolous: the \psykal\ version of the code performs
+somewhat anomalous: the \psykal\ version of the code performs
 significantly better than the original and in addition, this problem
@@
 
-In particular, the results on the IBM system demonstrate two key points.
+More unpredictably, the code structure required to obtain good
+performance for different architecture and/or compiler combinations
+differs, even for this relatively simple benchmark code. For hand
+optimised codes this implies the need to support multiple code bases
+in order to achieve portable performance.
 %
-First, even for a relatively simple
-benchmark code, the code structure required to obtain good performance
-between different architecture and/or compiler combinations may
-differ. For hand optimised codes this implies the need to support
-multiple code bases in order to achieve portable performance.
-%
-Second, the \psykal\ approach directly addresses the problem of
-requiring different optimisations for different architectures and/or
-compilers by limiting these differences to a separate performance
-layer, thereby allowing the natural science code base to remain
-unchanged.
+The \psykal\ approach directly addresses the problem of requiring
+different optimisations for different architectures and/or compilers
+by limiting these differences to a separate performance layer, thereby
+allowing the natural science code base to remain unchanged.
 
@@
 that is currently under development. This compiler will generate
-optimised PSy-layer code using a user-provided recipe of
+optimised PSy layer code using a user-provided recipe of
 transformations, thereby removing the need for optimisation experts to
@@
 % use section* for acknowledgement
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 

### 2015-10-01 docs/papers/shallow_performance/shallow_perf.tex
forgot to add a \ to the end of \psykal in the paper so any spaces were removed.

@@
 %
-In \psykal, the code related to the underlying science is formally
+In \psykal\, the code related to the underlying science is formally
 separated from code related to parallelisation and single core
@@
 
-In common with OP2, the \psykal approach separates out the science
+In common with OP2, the \psykal\ approach separates out the science
 code and the performance related code and does this via a separate
@@
 similar in terms of where they are placed in the code and in their
-semantics but the \psykal approach supports the specification of more
+semantics but the \psykal\ approach supports the specification of more
 than one Kernel in a parallel region of code, compared with one for
 OP2, giving more scope for optimisation. Another difference is that
-the description metadata is provided with the Kernel in the \psykal
+the description metadata is provided with the Kernel in the \psykal\
 approach whereas it is provided as part of the Kernel call in OP2.
 
-While the \psykal approach is general, we are currently applying it to
+While the \psykal\ approach is general, we are currently applying it to
 Atmosphere and Ocean models written in Fortran where domain

### 2015-10-01 docs/papers/shallow_performance/shallow_perf.tex
Final minor changes to paper text.

@@
 In common with OP2, the \psykal\ approach separates out the science
-code and the performance related code and does this via a separate
-layer. The calls that specify parallelism in both approaches are
-similar in terms of where they are placed in the code and in their
-semantics but the \psykal\ approach supports the specification of more
-than one Kernel in a parallel region of code, compared with one for
-OP2, giving more scope for optimisation. Another difference is that
-the description metadata is provided with the Kernel in the \psykal\
-approach whereas it is provided as part of the Kernel call in OP2.
+code and the performance related code via separate layers. The calls
+that specify parallelism in both approaches are similar in terms of
+where they are placed in the code and in their semantics but the
+\psykal\ approach supports the specification of more than one kernel
+in a parallel region of code, compared with one for OP2, giving more
+scope for optimisation. Another difference is that the description
+metadata is provided with a kernel in the \psykal\ approach whereas it
+is provided as part of the kernel call in OP2.
 

### 2015-10-02 docs/papers/shallow_performance/PSEA-2015_review_response.txt
parco paper: clarify that gungho paper (ref 2) has been accepted for publication and we are awaiting a date
parco paper: final tweaks to text and response to reviewers

@@
 
-GRAHAM.
+The version used in these tests has been released in version 1.0 of
+the GOcean benchmark. The text has been updated accordingly.
 
@@
 
-DOES SPACE PERMIT ANY MORE DETAIL?
+Unfortunately space does not permit us to describe the transformations
+in more detail.
 
@@
 
-RUPERT.
+Text has been added that describes how OP2 and PSyKAl compare.
 
@@
 are not claiming that a general set of optimisations can be applied to
-any given code. Instead, the PSyKAl approach separates the code which
-must be optimised from the code describing the science. The
-optimisations are restricted to the PSy layer and therefore do not
-affect the scientific components of the code.
+any given code (quite the opposite in fact). Instead, the PSyKAl
+approach separates the code which must be optimised from the code
+describing the science. The optimisations are restricted to the PSy
+layer and therefore do not affect the scientific components of the
+code. This is now emphasised in the text.
 
@@
 
-RUPERT/GRAHAM
+This is correct. The paper referred to has been accepted for publication
+but we are awaiting notification from the publisher as to when
+it will actually appear.
 

### 2015-10-02 docs/papers/shallow_performance/shallow_perf.tex
parco paper: final tweaks to text and response to reviewers

@@
 %
-A finite difference shallow water benchmark optimised for cache based
+A finite difference shallow water benchmark optimised for cache-based
 architectures is taken as the starting point. A vanilla \psykal\ version
@@
 performance loss and sometimes small performance gains compared to the
-original optimised code, there is no single best hand optimised
-implementation of the code for all the compilers tested, and the
-flexibility in applying code modifications to the optimisation layer
-in the \psykal\ version allows for a significant performance improvement
-on the IBM Power8 compared with the original code.
+original optimised code. We also find that there is no single best
+hand-optimised implementation of the code for all of the compilers
+tested.
 
@@
 situation is currently complicated still further by the existence of
-competing hardware technology; if one was to begin writing a major
+competing hardware technology; if one were to begin writing a major
 scientific application today it is unclear whether one would target
@@
 In common with OP2, the \psykal\ approach separates out the science
-code and the performance related code via separate layers. The calls
+code and the performance-related code into distinct layers. The calls
 that specify parallelism in both approaches are similar in terms of
-where they are placed in the code and in their semantics but the
+where they are placed in the code and in their semantics. However, the
 \psykal\ approach supports the specification of more than one kernel
@@
 loop blocking. Additionally, the potential parallelism within this
-iteration space can also be optimised and parallelised. The PSy layer
+iteration space can also be exploited and optimised. The PSy layer
 can therefore be tailored for a particular hardware (such as
@@
 Shallow program obtained at the beginning of this project the
-`original' version.
+`original' version. This code is available as part of version 1.0 of
+the GOcean benchmark (\url{https://puma.nerc.ac.uk/trac/GOcean}).
 
@@
 that is currently under development. This compiler will generate
-optimised PSy layer code using a user-provided recipe of
+optimised PSy layer code by following a user-provided recipe of
 transformations, thereby removing the need for optimisation experts to
@@
 
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
 \bibliographystyle{unsrt}
-% argument is your BibTeX string definitions and bibliography database(s)
 \bibliography{shallow_perf}

### 2015-10-02 nemolite2d_psykal/nemolite2d_acc_transform.py
OpenACC: change nemolite script to use new OpenACCDataTrans transformation

@@ -2,8 +2,10 @@
 # before generating the PSy layer Fortran code.
 
 def trans(psy):
-    from transformations import OpenACCParallelTrans
+    from transformations import OpenACCParallelTrans, \
+        OpenACCDataTrans
     atrans = OpenACCParallelTrans()
+    dtrans = OpenACCDataTrans()
 
     invoke = psy.invokes.get('invoke_0')
     schedule = invoke.schedule
@@ -12,9 +14,13 @@ def trans(psy):
 
     # Apply the OpenMP Loop transformation to *every* loop 
     # in the schedule
+    from psyGen import Loop
     for child in schedule.children:
-        newschedule,memento=atrans.apply(child)
-        schedule = newschedule
+        if isinstance(child, Loop):
+            newschedule, memento = atrans.apply(child)
+            schedule = newschedule
+
+    newschedule, memento = dtrans.apply(schedule)
 
     invoke.schedule = newschedule
     newschedule.view()

### 2015-10-15 docs/papers/shallow_performance/shallow_perf.tex
parco paper: changes following comments by Eike

@@
 A finite difference shallow water benchmark optimised for cache-based
-architectures is taken as the starting point. A vanilla \psykal\ version
-is written and the performance of the two compared. The optimisations
-that were applied to the original benchmark (loop fusion {\it etc.}) are then
-applied to the \psykal\ version as a set of code modifications to the
-optimisation layer. Performance results are presented for the Cray,
-Intel and GNU compilers on Intel Ivybridge and Haswell processors and
-for the IBM compiler on Power8.
+architectures is taken as the starting point. A vanilla
+\psykal\ version is written and the performance of the two
+compared. The optimisations that were applied to the original
+benchmark (loop fusion {\it etc.}) are then manually applied to the
+\psykal\ version as a set of code modifications to the optimisation
+layer. Performance results are presented for the Cray, Intel and GNU
+compilers on Intel Ivybridge and Haswell processors and for the IBM
+compiler on Power8.
 %
@@
 from an HPC expert and/or some coding rules to help ensure the kernels
-compile into efficient code.
+compile into efficient code.  In an alternative approach, kernels are
+generated from a high-level specification~\cite{firedrake,fenics},
+potentially allowing them to be optimised automatically~\cite{coffee}.
 
@@
 loop blocking. Additionally, the potential parallelism within this
-iteration space can also be exploited and optimised. The PSy layer
-can therefore be tailored for a particular hardware (such as
-multi-core, many-core, GPGPUs, or some combination thereof) and
-software (such as compiler, operating system, MPI implementation)
-configuration with no change to the Algorithm or Kernel layer
-code. This approach therefore offers the potential for portable
-performance.
+iteration space can also be exploited and optimised. The PSy layer can
+therefore be tailored for a particular hardware (such as multi-core,
+many-core, GPGPUs, or some combination thereof) and software (such as
+compiler, operating system, MPI implementation) configuration with no
+change to the Algorithm or Kernel layer code. This approach therefore
+offers the potential for portable performance. In this work we apply
+optimisations to the PSy layer manually. The development of a tool to
+automate this process will be the subject of a future paper.
 
@@
 values at the next time-step to be those at the current
-time-step. Profiling of the original code showed that these 
-sections each account for approximately one third of the run-time.
+time-step. Profiling the original code showed that these 
+sections each account for around one third of the run-time.
 
@@
 
-As indicated above, transformations T10 and T11 were
-suggested by the IBM compiler team. The second of these effectively
-undoes transformation four by fissioning the first loop nest back
-into three separate loop nests.
+As indicated above, transformations T10 and T11 were suggested by the
+IBM compiler team. The second of these effectively undoes
+transformation four by fissioning the first loop nest back into three
+separate loop nests. Note that this sequence of transformations is
+specific to Shallow: we are currently working upon a code-generation
+system that will allow a user to choose from a menu of transformations
+when optimising a PSy layer and this will be the subject of a future
+paper.
 
-We shall see that these transformations do not always result in
+We shall see that the transformations T1--T11 do not always result in
 improved performance. Whether or not they do so depends both on the
@@
 optimisations is typically to recover, as far as is possible, the
-structure of the original version of the code. It may well be that
-transforming the code into some other structure would result in better
-performance on a particular architecture. However, exploring this
-optimisation space is beyond the scope of the present work.
+structure of the original version of the code.  Therefore this
+sequence of transformations is specific to Shallow. It may well be
+that transforming the code into some other structure would result in
+better performance on a particular architecture. However, exploring
+this optimisation space is beyond the scope of the present work.  We
+are currently developing a code-generation system that will allow a
+user to choose from a menu of transformations when optimising a PSy
+layer and this will be the subject of a future paper.
 
@@
 in moving from the $256^2$ case to the $512^2$ case, despite the
-larger LLC of the Ivy Bridge system. In order to investigate this
-further, we profiled the application to obtain hardware performance
-counter data for the Gnu- and Intel-compiled executables. This
-revealed that the Gnu-compiled executable made more references to
-memory and had a higher LLC miss-rate compared with the Intel binary.
-We conclude that the Gnu compiler is not utilising registers as
-efficiently as the Intel compiler and that this results in greater
-memory traffic.
+larger LLC of the Ivy Bridge system. In order to investigate this, we
+profiled the application to obtain hardware performance counter data
+for the Gnu- and Intel-compiled executables. This revealed that the
+Gnu-compiled executable made more references to memory and had a
+higher LLC miss-rate compared with the Intel binary.  We conclude that
+the Gnu compiler is not utilising registers as efficiently as the
+Intel compiler and that this results in greater memory traffic.
 
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}). We acknowledge use of the Hartree
-Centre IBM Power 8 system in this work. The STFC Hartree Centre is a
-research collaboratory in association with IBM providing High
-Performance Computing platforms funded by the UK's investment in
-e-Infrastructure. The Centre aims to develop and demonstrate next
-generation software, optimised to take advantage of the move towards
-exa-scale computing. This work was funded by the NERC `GOcean' project,
-grant number NE/L012111/1.
+(\url{http://www.archer.ac.uk}) and the Hartree Centre IBM Power 8
+system. The STFC Hartree Centre is a research collaboratory in
+association with IBM providing HPC platforms funded by the UK's
+investment in e-Infrastructure. The Centre aims to develop and
+demonstrate next generation software, optimised for
+the move towards exa-scale computing. This work was funded by the NERC
+`GOcean' project, grant number NE/L012111/1.
 

### 2015-11-16 docs/NEMO/psyclone_in_nemo.tex
Add draft report/work plan from meeting at Met Office on 12-13/11/2015

subroutine ldf_lsp_triad(...)

  call invoke( horiz_density_grad(expan_coeff, temp, salin, density_grad) )
  if (partial_steps) then
    call invoke( bottom_grad_corr(density_grad, expan_coeff) )
  end if
  call invoke( vertical_density_grad(temp, salin, ...),       &
               recip_depth_w_mix(...),                        &
               set(..., 0),                                   &
               triad_slope_at_mixed_layer(density_grad, ...), &
               calc_triad(...) )

end subroutine ldf_lsp_triad

### 2015-11-16 docs/NEMO/psyclone_and_nemo.tex
Add initial draft of discussion of use of PSyclone in NEMO.

\documentclass{article}

\newcommand{\psykal}{{PS}y{KA}l}

\begin{document}

\title{Would NEMO benefit from using the PSyclone system?}

\author{A. R. Porter and R. W. Ford and T. Graham and M. Bell}

\maketitle

\section{Introduction}

This report describes the outcomes of an initial meeting looking at
the possible use of the PSyclone code-generation system with the NEMO
ocean model.

Fundamental to the use of PSyclone is the \psykal\ (Parallel System,
Kernel, Algorithm) separation of concerns. In this approach a program
is re-structured into three layers; the Algorithm, PSy and Kernel layers.
The Algorithm and Kernel layers are the responsibility of the natural
 scientist while all code related to parallelism is contained within the
PSy layer.

PSyclone is currently being developed as part of the GungHo project
and forms a component of the Dynamo (the prototype dynamical core)
build system. Dynamo uses Finite Elements on a grid that is formally
unstructured in the horizontal but structured in the vertical.  In
contrast, NEMO uses Finite Difference on a (tri-polar) regular,
latitude-longitude grid.

In the GOcean2D project we investigated the use of the \psykal\ and
PSyclone approach for models using a finite-difference scheme on a
regular lat-lon grid. For this work we took two different
shallow-water models, one of which was the 2D, free-surface component
of NEMO. After restructuring these programs according to the \psykal\
separation of concerns we were able to show that the performance of
the original code could be recovered with optimisations restricted to
the PSy layer.

The question now is, would this approach be valuable for the full NEMO
code?  Key to this is the fact that the full NEMO code is a
combination of both 3D and 2D components and thus we have to decide
how the loop over levels (the ``k-loop'') might fit within the PSyclone
framework. In Dynamo this loop over levels is not managed by PSyclone
but instead is performed within each kernel as a way of amortizing the
cost of the indirection associated with the unstructured mesh.

We believe that there are three possible places in which to locate the
k loop:

\begin{enumerate}

\item the Algorithm Layer;

\item the PSy Layer (i.e. make it part of the iteration space managed
  by PSyclone);

\item the Kernel Layer.

\end{enumerate}

NEMO currently uses an $(i,j,k)$ array-index ordering and on current
(CPU-based) architectures reaches a strong-scaling limit when the MPI
sub-domains are about $40\times 40$ grid points. In contrast, the
number of vertical levels in current configurations is typically
$\sim75$. In addition it is likely that the majority of loops over
levels in NEMO are parallelisable. It is therefore highly desirable to
be able to exploit this and this lends weight to option (2). In order
for the PSy layer to manage the iteration space in k it must be
possible to formulate a complete set of rules for the different types
of k loops in the whole of NEMO. For instance, it must be possible to
infer the value of the lower bound of the k-loop around a given kernel
from a knowledge of the grid-point type of the field being updated by
a kernel (plus other meta-data describing the kernel). The number of
exceptions to this must be small if an effective Domain Specific
Language is to be developed for PSyclone's application to NEMO.

However, NEMO has had several decades of development and optimisation
and has thus evolved to make optimal use of the data layout associated
with the $i,j,k$ index ordering. As a consequence there are many
examples where a loop over levels itself contains several, individual
$i,j$ loop nests, e.g.:

\begin{verbatim}
DO jk = 2, jpkm1
   ! Kernel 1
   DO jj = 2, jpjm1
      DO ji = 2, jpim1
         fldb(ji,jj,jk) = ...
      END DO
   END DO
   ! Kernel 2
   DO jj = 2, jpjm1
      DO ji = 2, jpim1
         fldc(ji,jj,jk) = ...
      END DO
   END DO
END DO
\end{verbatim}

This structure favours option (1) since it naturally follows that the
resulting algorithm code would look something like:

\begin{verbatim}
DO jk = 2, jpkm1
   call invoke( kernel1(flda, fldb), &
                kernel2(fldb, fldc) )
END DO
\end{verbatim}

However, a key part of \psykal\ is that the Algorithm layer operates on
logically global fields. In other words it has no concept that the
fields may be domain decomposed (as would be done in an MPI
implementation). This is handled in the PSy layer where any necessary
halo swaps are performed. Unfortunately, the implication of the above
code structure is then that any halo swaps required in the invoke
would be performed separately for \emph{every level} of the
model. This would introduce an unacceptable overhead and we are
therefore forced to rule out option (1).

Finally, there remains option (3) in which the loop over levels is the
responsibility of a kernel. Although this is the approach taken in
Dynamo, the fact that NEMO uses an $i,j,k$ index ordering precludes it
here. This is because, by design, the PSy layer already manages the
loops over i and j and therefore a loop over k within the kernel would
become the innermost loop. The resulting array accesses would have a
stride of $jpi*jpj$ resulting in extremely poor cache use and no SIMD
vectorisation.

We believe therefore that the least intrusive and most performant way
to introduce PSyclone to NEMO would be option (2). This is dependent
on it being possible to construct a formal description of the majority
of k loops (i.e. the values to use for the loop bounds) within NEMO.

\section{Incremental introduction of PSyclone to NEMO}

One of the key advantages of the PSyclone system is that it is
possible for it to co-exist with existing code structures including
infrastructure that makes use of MPI.

As a first step, a section of code must be re-worked to have the
\psykal\ structure. In the context of NEMO, a routine such as
ldf\_slp\_triad could be tackled. The rules that must be followed when
re-structuring the code are:
\begin{itemize}
\item All computation on field values must be performed in kernels;
\item It must be possible to execute any given kernel in parallel but
  a kernel itself must contain no code relating to parallelism
  (e.g. no MPI or OpenMP);
\item All loops over $i$, $j$ and $k$ must be performed in the PSy
  layer;
\item The Algorithm layer must also contain no code relating to
  parallelism.
\end{itemize}
Initially the PSy layer may be written manually. This will then
subsequently be used to inform the development of PSyclone to support
NEMO.

\subsection{An example: ldf\_slp\_triad}

Rupert mocking this up.

\section{Risks}

Although initial discussions and a code walk through indicate that the
\psykal/PSyclone approach could be both beneficial and effective for
NEMO there remain risks. We discuss those here.

\subsection{Tri-diagonal Solves}

If PSyclone is to control the iteration space in $k$ as well as in $i,
j$ then this assumes that that space is parallelisable. A clear
exception to this is those routines that perform an implicit solve in
the vertical dimension (e.g. tra\_zdf\_imp). This results in the need
to solve a problem with tri-diagonal form. Kernel meta-data will be
required to identify such cases. Once PSyclone is able to identify
such kernels we can envisage it generating calls to optimised solver
routines.

\subsection{OBS}

The OBS (Observation) component of NEMO allows prognostic fields to be
extracted from the model at user-specified geographical locations.  It
is possible that the looping structure within these routines might be
difficult to classify in a way that is suitable for treatment with
PSyclone.

\subsection{AGRIF}

AGRIF (Adaptive Grid Refinement In Fortran) gives NEMO the ability to
run with two-way nesting of higher-resolution meshes. This is achieved
through a pre-processing step in which the tool changes the NEMO
source. In the resulting code field objects (Fortran derived types)
are passed down from the AGRIF infrastructure into wrappers for the
original NEMO routines.

Since PSyclone too works by reading and modifying/generating Fortran
source files it is obviously questionable whether it could safely
inter-operate with AGRIF. However, once PSyclone has generated the PSy
layer and re-written the Algorithm layer the resulting source code is
standard Fortran. Since AGRIF directives
(e.g. !\$AGRIF\_DO\_NOT\_TREAT) are comments they will remain unchanged
in both the Algorithm and the Kernels.

The most significant issue is then probably the underlying
infrastructure.  Both AGRIF and PSyclone normally work with their own
definitions of field objects rather than raw Fortran arrays. Although
this is not essential for PSyclone per se (the GOcean 0.1 API for
instance does not use these field objects) it does present problems
for its support of MPI. This is because PSyclone requires some way of
keeping track of whether a field's halos are `clean' or `dirty' and it
is natural to encapsulate this within a field object. It is however
possible to envisage other ways of doing this, e.g. by using the ISO
Fortran-C interface to get the address of a Fortran array in memory
and then using this to index into some internal data structure.

\subsection{CPP Keys}

NEMO currently makes use of CPP keys to configure the code at compile
time. Options include e.g. whether to include MPI support and which
ice model to use. It is likely that PSyclone can safely be applied to
the code after this pre-processing stage but this must be tested.

\subsection{Tracers}

Although the core of NEMO contains only two active tracers
(Temperature and Salinity) it can be used with biogeochemical
models. In those configurations there can be of order $\sim 50$ tracer
species, all of which may be treated independently. This has
significant implications for both data structures and our ability to
exploit the additional parallelism that is available.

\section{Suggested workplan - 1 Week}

In this section we consider what might be accomplished by one person
working intensively with NEMO and PSyclone over the course of one
week. The aim here would be to address some of the risks identified in
the previous section. We emphasise that there is no need for PSyclone
to be able to generate code for NEMO at this stage; all that is
required is a manual implementation of the \psykal\ separation of
concerns for at least one routine. For the purposes of testing with
AGRIF this routine need not be complex.

The tasks that might be tackled are:
\begin{enumerate}

\item Begin with a working AGRIF configuration. Manually 
 re-structure one routine following \psykal\ rules (1 day).

\item Attempt to process the resulting code with AGRIF and 
 document issues (1+ days).

\item Dependent on time remaining after the first two tasks, identify
  those routines that have complex structure in the k loop and attempt
  to classify/manually re-structure them (1 day).

\item If time allows, examine the use of tracers within NEMO and how
  this might be supported by PSyclone infrastructure.

\end{enumerate}
The first two tasks require a person with both experience of NEMO and
AGRIF with perhaps some additional support on \psykal\ concepts.  The
latter tasks call for someone familar with \psykal\ and PSyclone
concepts, with input from a NEMO expert.

\section{Suggested workplan - 1 year}

NEMO is a relatively large and complex code with an extensive user
community. Introducing \psykal/PSyclone to such a code base is not to
be undertaken lightly and therefore we suggest an initial project of
one year duration that would enable a thorough investigation of all
the issues. The primary aims of such a project would be: to establish
that the use of PSyclone with NEMO (including AGRIF) is technically
feasible; that the natural scientists are happy with the resulting
code and that it performs at least as well as the original.

In this section we present a possible work plan that could be followed
should funding be available for one FTE for one year.

\subsection{WP1: Identify Target NEMO Configurations}
\label{wp1_target_configs}

NEMO supports a wide variety of model configurations and these
influence which source code is physically executed. It is therefore
very important to choose representative configurations that will be
used as the basis for the work. At least one of these configurations
must use AGRIF and another must make significant use of tracers.  Any
other key technologies (e.g. coupling to sea ice) must be covered by
the chosen set of configurations.

Ideally there should be a test suite that covers these configurations
so that the physical correctness of the output of any developed code
can be verified. It is possible that SETTE will fulfil this
requirement.

\subsection{WP2: Classify the Types of Iteration Space}
\label{wp2_classify}

As described earlier, for PSyclone to be useful it must be possible to
classify the types of loop structure in the vast majority of the NEMO
code base. The routines involved in the selected configurations must
therefore be identified (e.g. by performing tracing experiments with a
tool such as Tau). The loop structures in these routines must then be
examined. A strategy for supporting those routines that cannot be put
into \psykal\ form must be developed.

As a part of this package we will develop the extensions to the GOcean
1.0 Kernel meta-data that are required to support the
three-dimensional finite-difference scheme used in NEMO.

\subsection{WP3: Infrastructure support}
\label{wp3_infrastructure}

Field objects or arrays?

The existing NEMO code base is already MPI parallel and therefore any
introduction of PSyclone must work within this framework.

The NEMO source code is under almost constant scientific
development. This must be allowed for in any strategy to introduce
PSyclone.

\subsection{WP4: Tracer Support}
\label{wp4_tracers}

As biogeochemical models are of increasing importance we must consider
how best to support them in any \psykal\ version of NEMO. This
includes data structures, infrastructure support and Kernel meta-data.

\subsection{WP5: Performance}
\label{wp5_perf}

Performance is a key issue for NEMO and is one of the reasons for
considering PSyclone -- the resulting code will no longer be
restricted to one particular architecture. However, a key first step
is to ensure that the introduction of the \psykal\ separation of
concerns does not harm performance on current machines. Or, if it
does, then that performance can be recovered through code
transformations. We have shown that this is the case for two
shallow-water models~\cite{XXX} but this work must be extended for the
3D parts of NEMO. Benchmarking of the configurations chosen in WP1
will inform the key routines used to examine this issue. Note that
this benchmarking must be performed on processor counts typical of
those used in production runs.

\subsection{WP6: PSyclone support}
\label{wp6_autogen}

PSyclone already contains some support for finite difference in two
dimensions.  (One of the codes developed as part of the GOcean project
is based upon the free-surface component of NEMO.) It is therefore
natural to first introduce PSyclone code-generation to those parts of NEMO
that are two dimensional.

While initial experiments may be performed by manually running
PSyclone on individual NEMO routines, this process must ultimately be
made a part of the standard build system. Part of this build system
involves running the C pre-processor (CPP) over the source since some
NEMO functionality (e.g. which ice model to use) is chosen at compile
time through CPP keys. We plan to introduce the PSyclone
code-generation step after this pre-processing but before AGRIF
transforms the source.

Once PSyclone is incorporated into the NEMO build system, the next
stage is to extend the PSyclone API to support 3D parts of NEMO.  This
will require the Kernel meta-data and rules developed under Work
Package 2.

\end{document}

### 2015-11-19 nemolite2d_psykal/nemolite2d_omp_transform.py
Convert nemolite2d-omp script into form that can be passed to generate using -s flag. Change Makefile to do so.

@@ -1,31 +1,25 @@
-# Python script which uses PSyclone to apply transformations
-# before generating the PSy layer Fortran code.
-from parse import parse,ParseError
-from psyGen import PSyFactory,GenerationError
-api="gocean1.0"
-filename="nemolite2d_alg.f90"
-ast,invokeInfo=parse(filename,api=api,invoke_name="invoke")
-psy=PSyFactory(api).create(invokeInfo)
-#print psy.invokes.names
+def trans(psy):
+    ''' Python script intended to be passed to PSyclone's generate()
+    function via the -s option. Applies OpenMP to every loop before
+    enclosing them all within a single OpenMP PARALLEL region. '''
 
-from psyGen import TransInfo
-t=TransInfo()
-ltrans = t.get_trans_name('GOceanOMPLoopTrans')
-rtrans = t.get_trans_name('OMPParallelTrans')
+    from psyGen import TransInfo
+    tinfo = TransInfo()
+    ltrans = tinfo.get_trans_name('GOceanOMPLoopTrans')
+    rtrans = tinfo.get_trans_name('OMPParallelTrans')
 
-schedule=psy.invokes.get('invoke_0').schedule
-#schedule.view()
-new_schedule=schedule
+    schedule = psy.invokes.get('invoke_0').schedule
+    # schedule.view()
 
-# Apply the OpenMP Loop transformation to *every* loop 
-# in the schedule
-for child in schedule.children:
-    newschedule,memento=ltrans.apply(child)
-    schedule = newschedule
+    # Apply the OpenMP Loop transformation to *every* loop
+    # in the schedule
+    for child in schedule.children:
+        newschedule, _ = ltrans.apply(child)
+        schedule = newschedule
 
-# Enclose all of these loops within a single OpenMP
-# PARALLEL region
-newschedule,memento = rtrans.apply(schedule.children)
+    # Enclose all of these loops within a single OpenMP
+    # PARALLEL region
+    newschedule, _ = rtrans.apply(schedule.children)
 
-psy.invokes.get('invoke_0')._schedule=schedule
-print psy.gen
+    psy.invokes.get('invoke_0').schedule = newschedule
+    return psy

### 2015-11-19 nemolite2d_psykal/nemolite2d_acc_transform.py
OpenACC: allow nemolite2d_acc_transform.py to be run as a script OR passed to generate(). Tidy for pep8 and pylint.

@@ -1,7 +1,11 @@
-# Python script which uses PSyclone to apply transformations
-# before generating the PSy layer Fortran code.
+
+'''Python script intended to be passed to PSyclone's generate()
+funcation via the -s option. Performs OpenACC transformations. '''
+
 
 def trans(psy):
+    ''' Take the supplied psy object, apply OpenACC transformations
+    to the schedule of invoke_0 and return the new psy object '''
     from transformations import OpenACCParallelTrans, \
         OpenACCDataTrans
     atrans = OpenACCParallelTrans()
@@ -9,31 +13,34 @@ def trans(psy):
 
     invoke = psy.invokes.get('invoke_0')
     schedule = invoke.schedule
-    #schedule.view()
-    new_schedule=schedule
+    # schedule.view()
 
-    # Apply the OpenMP Loop transformation to *every* loop 
+    # Apply the OpenMP Loop transformation to *every* loop
     # in the schedule
     from psyGen import Loop
     for child in schedule.children:
         if isinstance(child, Loop):
-            newschedule, memento = atrans.apply(child)
+            newschedule, _ = atrans.apply(child)
             schedule = newschedule
 
-    newschedule, memento = dtrans.apply(schedule)
+    newschedule, _ = dtrans.apply(schedule)
 
     invoke.schedule = newschedule
     newschedule.view()
     return psy
 
-from parse import parse,ParseError
-from psyGen import PSyFactory,GenerationError
-api="gocean1.0"
-filename="nemolite2d_alg.f90"
-ast,invokeInfo=parse(filename,api=api,invoke_name="invoke")
-psy=PSyFactory(api).create(invokeInfo)
-#print psy.invokes.names
 
-new_psy = trans(psy)
+if __name__ == "__main__":
+    from parse import parse
+    from psyGen import PSyFactory
+    API = "gocean1.0"
+    FILENAME = "nemolite2d_alg.f90"
+    _, INVOKEINFO = parse(FILENAME,
+                          api=API,
+                          invoke_name="invoke")
+    PSY = PSyFactory(API).create(INVOKEINFO)
+    # print PSY.invokes.names
 
-print new_psy.gen
+    NEW_PSY = trans(PSY)
+
+    print NEW_PSY.gen

### 2015-11-19 docs/NEMO/psyclone_and_nemo.tex
updated version of psyclone and nemo document

@@
 \newcommand{\psykal}{{PS}y{KA}l}
+\newcommand{\psyclone}{{PS}yclone}
 
@@
 
-\title{Would NEMO benefit from using the PSyclone system?}
+\title{Would NEMO benefit from using the \psyclone\ system?}
 
@@
 This report describes the outcomes of an initial meeting looking at
-the possible use of the PSyclone code-generation system with the NEMO
+the possible use of the \psyclone\ code-generation system with the NEMO
 ocean model.
 
-Fundamental to the use of PSyclone is the \psykal\ (Parallel System,
+Fundamental to the use of \psyclone\ is the \psykal\ (Parallel System,
 Kernel, Algorithm) separation of concerns. In this approach a program
-is re-structured into three layers; the Algorithm, PSy and Kernel layers.
-The Algorithm and Kernel layers are the responsibility of the natural
- scientist while all code related to parallelism is contained within the
-PSy layer.
+is re-structured into three layers; the Algorithm, PSy and Kernel
+layers.  The Algorithm and Kernel layers are the responsibility of the
+natural scientist while all code related to parallelism is contained
+within the PSy layer. Thus the scientist does not need to concern
+themselves with optimisation issues, such as loop fusion, redundant
+computation in halos, etc. and parallelism issues, such as placement
+of directives and halo swap calls.
 
-PSyclone is currently being developed as part of the GungHo project
+The role of \psyclone\ is to generate the PSy layer (and modify the
+Algorithm layer). This role is split into two parts: correctness and
+performance. The correctness aspect of \psyclone\ is often underplayed
+but there are potentially significant maintenance benefits in this
+approach. For example, the horizontal iteration space, ji and jj (and
+potentially the vertical iteration space), is generated in the PSy
+layer which can help avoid indexing errors.
+
+\psyclone\ has been designed from the ground up with performance and
+performance portability in mind. Rather than automatically optimising
+code, as compilers do, \psyclone\ allows an optimisation expert to
+specify a set of optimisations in a script, which it then applies to
+the code. The \psykal separation of concerns can be considered a
+stripping back of many of the optimisations that have been hard coded
+into a program and \psyclone\ then allows such optimisations to be
+specified in a script via a set of transformations (a recipe of
+transformations if you like) to apply to the code. This hopefully
+means that there is no (or at least minimal) loss in performance
+compared to hand optimised code when using \psyclone.
+
+The advantage of a recipe is that it can be changed, {\bf with no
+  change to the science code}. Thus, for example, one recipe could
+target MPI and OpenMP using the Intel compiler on a Cray with a
+particular configuration of NEMO. Another recipe could target MPI and
+OpenACC using the PGI compiler on GPU's for a different NEMO
+configuration. One of the problems with performance portability is
+that optimisations for one architecture are not necessarily beneficial
+for another. It is difficult enough to get different versions of the
+same compiler on the same architecture to work well on a complex code
+without changes to source code. If such optimisations are hard coded
+then it is very difficult to change the code to work well on another
+architecture and almost impossible to have performance portability
+with a single code base. \psyclone\ aims to address this problem.
+
+\psyclone\ is currently being developed as part of the GungHo project
 and forms a component of the Dynamo (the prototype dynamical core)
-build system. Dynamo uses Finite Elements on a grid that is formally
+build system. \psyclone\ can currently generate correct sequential code
+for Dynamo and supports transformations such as loop colouring, loop
+fusion, OpenMP and module inlining. We are currently working
+on OpenACC transformations and are about to start work on MPI support.
+
+Dynamo uses Finite Elements on a grid that is formally
 unstructured in the horizontal but structured in the vertical.  In
@@
 In the GOcean2D project we investigated the use of the \psykal\ and
-PSyclone approach for models using a finite-difference scheme on a
+\psyclone\ approach for models using a finite-difference scheme on a
 regular lat-lon grid. For this work we took two different
 shallow-water models, one of which was the 2D, free-surface component
-of NEMO. After restructuring these programs according to the \psykal\
-separation of concerns we were able to show that the performance of
-the original code could be recovered with optimisations restricted to
-the PSy layer.
+of NEMO. After restructuring these programs according to the
+\psykal\ separation of concerns we were able to show that the
+performance of the original code could be recovered with optimisations
+restricted to the PSy layer. We also demonstrated that this code could
+run efficiently on multi-core architectures using OpenMP directives
+and GPU architectures using OpenACC, again with optimisations
+restricted to the PSy layer.
 
-The question now is, would this approach be valuable for the full NEMO
-code?  Key to this is the fact that the full NEMO code is a
-combination of both 3D and 2D components and thus we have to decide
-how the loop over levels (the ``k-loop'') might fit within the PSyclone
-framework. In Dynamo this loop over levels is not managed by PSyclone
-but instead is performed within each kernel as a way of amortizing the
-cost of the indirection associated with the unstructured mesh.
+The question now is, would this approach be valuable and applicable
+for the full NEMO code?  Key to this is the fact that the full NEMO
+code is a combination of both 3D and 2D components and thus we have to
+decide how the loop over levels (the ``k-loop'') might fit within the
+\psyclone\ framework. In Dynamo this loop over levels is not managed by
+\psyclone\ but instead is performed within each kernel as a way of
+amortising the cost of the indirection associated with the
+unstructured mesh.
 
 We believe that there are three possible places in which to locate the
-k loop:
+k loop in NEMO:
 
@@
 \item the PSy Layer (i.e. make it part of the iteration space managed
-  by PSyclone);
+  by \psyclone);
 
@@
 exceptions to this must be small if an effective Domain Specific
-Language is to be developed for PSyclone's application to NEMO.
+Language is to be developed for \psyclone\'s application to NEMO.
 
@@
 We believe therefore that the least intrusive and most performant way
-to introduce PSyclone to NEMO would be option (2). This is dependent
+to introduce \psyclone\ to NEMO would be option (2). This is dependent
 on it being possible to construct a formal description of the majority
@@
 
-\section{Incremental introduction of PSyclone to NEMO}
+\section{Incremental introduction of \psyclone\ to NEMO}
 
-One of the key advantages of the PSyclone system is that it is
+One of the key advantages of the \psyclone\ system is that it is
 possible for it to co-exist with existing code structures including
@@
 Initially the PSy layer may be written manually. This will then
-subsequently be used to inform the development of PSyclone to support
+subsequently be used to inform the development of \psyclone\ to support
 NEMO.
@@
 Although initial discussions and a code walk through indicate that the
-\psykal/PSyclone approach could be both beneficial and effective for
+\psykal/\psyclone\ approach could be both beneficial and effective for
 NEMO there remain risks. We discuss those here.
@@
 
-If PSyclone is to control the iteration space in $k$ as well as in $i,
+If \psyclone\ is to control the iteration space in $k$ as well as in $i,
 j$ then this assumes that that space is parallelisable. A clear
@@
 to solve a problem with tri-diagonal form. Kernel meta-data will be
-required to identify such cases. Once PSyclone is able to identify
+required to identify such cases. Once \psyclone\ is able to identify
 such kernels we can envisage it generating calls to optimised solver
@@
 difficult to classify in a way that is suitable for treatment with
-PSyclone.
+\psyclone.
 
@@
 
-Since PSyclone too works by reading and modifying/generating Fortran
+Since \psyclone\ too works by reading and modifying/generating Fortran
 source files it is obviously questionable whether it could safely
-inter-operate with AGRIF. However, once PSyclone has generated the PSy
+inter-operate with AGRIF. However, once \psyclone\ has generated the PSy
 layer and re-written the Algorithm layer the resulting source code is
@@
 The most significant issue is then probably the underlying
-infrastructure.  Both AGRIF and PSyclone normally work with their own
+infrastructure.  Both AGRIF and \psyclone\ normally work with their own
 definitions of field objects rather than raw Fortran arrays. Although
-this is not essential for PSyclone per se (the GOcean 0.1 API for
+this is not essential for \psyclone\ per se (the GOcean 0.1 API for
 instance does not use these field objects) it does present problems
-for its support of MPI. This is because PSyclone requires some way of
+for its support of MPI. This is because \psyclone\ requires some way of
 keeping track of whether a field's halos are `clean' or `dirty' and it
@@
 time. Options include e.g. whether to include MPI support and which
-ice model to use. It is likely that PSyclone can safely be applied to
+ice model to use. It is likely that \psyclone\ can safely be applied to
 the code after this pre-processing stage but this must be tested.
@@
 
+\section{Index re-reordering}
+
+\psyclone\ does not directly support the idea of re-ordering NEMO's data
+layout. This question has been asked of \psyclone\ before as it is
+thought that i,j,k may work best on some architectures and k,i,j on
+others. However, the separation of science code and optimisations and
+parallelism combined with the separation of the i,j,k looping
+structure from the Kernel science code should make it much more
+tractable to manually evaluate the benefits of data reordering. In
+future work we would be interested in exploring the possibility of
+supporting data re-ordering within \psyclone\ in cases where the PSy
+layer controls the full iteration space.
+
 \section{Suggested workplan - 1 Week}
 
-In this section we consider what might be accomplished by one person
-working intensively with NEMO and PSyclone over the course of one
-week. The aim here would be to address some of the risks identified in
-the previous section. We emphasise that there is no need for PSyclone
-to be able to generate code for NEMO at this stage; all that is
-required is a manual implementation of the \psykal\ separation of
-concerns for at least one routine. For the purposes of testing with
-AGRIF this routine need not be complex.
+In this section we consider what might be accomplished by two people
+with the required skills working intensively with NEMO and \psyclone\
+over the course of one week. The aim here would be to address some of
+the risks identified in the previous section. We emphasise that there
+is no need for \psyclone\ to be able to generate code for NEMO at this
+stage; all that is required is a manual implementation of the
+\psykal\ separation of concerns for at least one routine. For the
+purposes of testing with AGRIF this routine need not be complex.
 
@@
 \item If time allows, examine the use of tracers within NEMO and how
-  this might be supported by PSyclone infrastructure.
+  this might be supported by \psyclone\ infrastructure.
 
@@
 AGRIF with perhaps some additional support on \psykal\ concepts.  The
-latter tasks call for someone familar with \psykal\ and PSyclone
+latter tasks call for someone familiar with \psykal\ and \psyclone\
 concepts, with input from a NEMO expert.
@@
 NEMO is a relatively large and complex code with an extensive user
-community. Introducing \psykal/PSyclone to such a code base is not to
+community. Introducing \psykal/\psyclone\ to such a code base is not to
 be undertaken lightly and therefore we suggest an initial project of
@@
 the issues. The primary aims of such a project would be: to establish
-that the use of PSyclone with NEMO (including AGRIF) is technically
+that the use of \psyclone\ with NEMO (including AGRIF) is technically
 feasible; that the natural scientists are happy with the resulting
@@
 
-As described earlier, for PSyclone to be useful it must be possible to
+As described earlier, for \psyclone\ to be useful it must be possible to
 classify the types of loop structure in the vast majority of the NEMO
@@
 The existing NEMO code base is already MPI parallel and therefore any
-introduction of PSyclone must work within this framework.
+introduction of \psyclone\ must work within this framework.
 
@@
 development. This must be allowed for in any strategy to introduce
-PSyclone.
+\psyclone.
 
@@
 Performance is a key issue for NEMO and is one of the reasons for
-considering PSyclone -- the resulting code will no longer be
+considering \psyclone\ -- the resulting code will no longer be
 restricted to one particular architecture. However, a key first step
@@
 
-\subsection{WP6: PSyclone support}
+\subsection{WP6: \psyclone\ support}
 \label{wp6_autogen}
 
-PSyclone already contains some support for finite difference in two
+\psyclone\ already contains some support for finite difference in two
 dimensions.  (One of the codes developed as part of the GOcean project
 is based upon the free-surface component of NEMO.) It is therefore
-natural to first introduce PSyclone code-generation to those parts of NEMO
+natural to first introduce \psyclone\ code-generation to those parts of NEMO
 that are two dimensional.
@@
 While initial experiments may be performed by manually running
-PSyclone on individual NEMO routines, this process must ultimately be
+\psyclone\ on individual NEMO routines, this process must ultimately be
 made a part of the standard build system. Part of this build system
@@
 NEMO functionality (e.g. which ice model to use) is chosen at compile
-time through CPP keys. We plan to introduce the PSyclone
+time through CPP keys. We plan to introduce the \psyclone\
 code-generation step after this pre-processing but before AGRIF
@@
 
-Once PSyclone is incorporated into the NEMO build system, the next
-stage is to extend the PSyclone API to support 3D parts of NEMO.  This
+Once \psyclone\ is incorporated into the NEMO build system, the next
+stage is to extend the \psyclone\ API to support 3D parts of NEMO.  This
 will require the Kernel meta-data and rules developed under Work

### 2015-11-20 api_v1.0/field_mod.f90
OpenACC: modify field checksum routine so that it derefs and then passes down to array_checksum. This is necessary for Cray to get the correct pointer when doing update host(field)

@@ -910,26 +910,48 @@ contains
     type(r2d_field), intent(in) :: field
     real(wp) :: val
 
+    val = array_checksum(field%data, field%data_on_device,           &
+                         field%internal%xstart, field%internal%xstop, &
+                         field%internal%ystart, field%internal%ystop)
+    return
+
+! The code below fails with the Cray compiler - the update host(field%data)
+! seems to get the wrong pointer.
+
     ! If we're using OpenACC then make sure we get the data back from
     ! the GPU
-    if(field%data_on_device)then
-!$acc update host(field%data)
-    end if
-
-    !> \todo Could add an OpenMP implementation
-    val = SUM( ABS(field%data(field%internal%xstart:field%internal%xstop, &
-                              field%internal%ystart:field%internal%ystop)) )
+!    if(field%data_on_device)then
+!!acc update host(field%data)
+!    end if
+!
+!    !> \todo Could add an OpenMP implementation
+!    val = SUM( ABS(field%data(field%internal%xstart:field%internal%xstop, &
+!                              field%internal%ystart:field%internal%ystop)) )
   end function fld_checksum
 
   !===================================================
 
   !> Compute the checksum of ALL of the elements of supplied array
-  function array_checksum(field) result(val)
+  function array_checksum(field, update, &
+                          xstart, xstop, &
+                          ystart, ystop) result(val)
     implicit none
     real(wp), dimension(:,:), intent(in) :: field
+    logical, optional, intent(in) :: update
+    integer, optional, intent(in) :: xstart, xstop, ystart, ystop
     real(wp) :: val
 
-    val = SUM( ABS(field(:,:) ) )
+    if( present(update) )then
+       if(update)then
+          !$acc update host(field)
+       end if
+    end if
+
+    if( present(xstart) )then
+       val = SUM( ABS(field(xstart:xstop,ystart:ystop) ) )
+    else
+       val = SUM( ABS(field(:,:)) )
+    end if
 
   end function array_checksum
 

### 2015-11-20 nemolite2d_psykal_acc/nemolite2d.f90
Correct the intents of all fields to allow the <data on device> member to be set

@@ -126,7 +126,7 @@ subroutine step(grid, istp, &
   integer,         intent(in) :: istp
   type(r2d_field), intent(inout) :: un, vn, sshn, sshn_u, sshn_v
   type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
-  type(r2d_field), intent(in) :: hu, hv, ht
+  type(r2d_field), intent(inout) :: hu, hv, ht
 
   call invoke_time_step(istp, ssha, ssha_u, ssha_v, &
                         sshn, sshn_u, sshn_v, &

### 2015-11-20 nemolite2d_psykal_acc/time_step_mod.f90

@@ -21,7 +21,7 @@ contains
     integer,         intent(in)    :: istp
     type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
     type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
-    type(r2d_field), intent(in)    :: hu, hv, ht
+    type(r2d_field), intent(inout) :: hu, hv, ht
                      
     call invoke_time_step_arrays(istp,                                  &
                                  ua%grid%nx, ua%grid%ny,                &

### 2015-11-20 docs/NEMO/example/horizontal_density_gradient_u_mod.f90
Added an example of what NEMO would look like if it were to be PSyclone'd.

@@ -1,0 +1,33 @@
+module horizontal_density_gradient_u_mod
+
+  implicit none
+
+  private
+  public horizontal_density_gradient_u, horizontal_density_gradient_u_code
+
+  ! metadata
+  type, extends(kernel_type) :: horizontal_density_gradient_u
+     type(arg), dimension(3) :: meta_args = &
+          (/ arg(WRITE, CU),                & ! zgru
+             arg(READ,  CT, STENCIL(E,1)),  & ! prd
+             arg(READ,  GRID_MASK_U)        & ! umask
+          /)
+     integer :: iterates_over = 3D_LOCAL_POINTS
+     integer :: index_offset  = OFFSET_NE
+   contains
+     procedure, nopass :: code => horizontal_density_gradient_u_code
+  end type horizontal_density_gradient_u
+	
+contains
+
+  subroutine horizontal_density_gradient_u_code(ji,jj,jk,zgru,prd,umask)
+
+    integer,  intent(in)                    :: ji,jj,jk
+    real(wp), intent(in),  dimension(:,:,:) :: prd, umask
+    real(wp), intent(out), dimension(:,:,:) :: zgru
+
+    zgru(ji,jj,jk) = umask(ji,jj,jk) * ( prd(ji+1,jj,jk) - prd(ji,jj,jk) )
+
+  end subroutine horizontal_density_gradient_u_code
+
+end module horizontal_density_gradient_u_mod

### 2015-11-20 docs/NEMO/example/horizontal_density_gradient_v_mod.f90

@@ -1,0 +1,33 @@
+module horizontal_density_gradient_v_mod
+
+  implicit none
+
+  private
+  public horizontal_density_gradient_v, horizontal_density_gradient_v_code
+
+  ! metadata
+  type, extends(kernel_type) :: horizontal_density_gradient_v
+     type(arg), dimension(3) :: meta_args = &
+          (/ arg(WRITE, CV),                & ! zgrv
+             arg(READ,  CT, STENCIL(N,1)),  & ! prd
+             arg(READ,  GRID_MASK_V)        & ! vmask
+          /)
+     integer :: iterates_over = 2D_LOCAL_POINTS
+     integer :: index_offset  = OFFSET_NE
+   contains
+     procedure, nopass :: code => horizontal_density_gradient_v_code
+  end type horizontal_density_gradient_v
+	
+contains
+
+  subroutine horizontal_density_gradient_v_code(ji,jj,jk,zgrv,prd,vmask)
+
+    integer,  intent(in)                    :: ji,jj,jk
+    real(wp), intent(in),  dimension(:,:,:) :: prd, vmask
+    real(wp), intent(out), dimension(:,:,:) :: zgrv
+
+    zgrv(ji,jj,jk) = vmask(ji,jj,jk) * ( prd(ji,jj+1,jk) - prd(ji,jj,jk) )
+
+  end subroutine horizontal_density_gradient_v_code
+
+end module horizontal_density_gradient_v_mod

### 2015-11-20 docs/NEMO/example/ldf_slp_alg.f90

@@ -1,0 +1,44 @@
+module ldfslp_mod
+
+  IMPLICIT NONE
+
+contains
+
+  subroutine ldf_slp(uslp,wslpi,vslp,wslpj,...)
+    !! ...
+    !! ** Purpose :   Compute the slopes of neutral surface
+    !! ...
+    type(r3d_field), intent(in)  :: ...
+    type(r3d_field), intent(out) :: uslp,wslpi,vslp,wslpj
+
+    ! local fields
+    type(r3d_field) :: zgru,zgrv,zwz,zww,zdzr
+
+    zgru = r3d_field(model_grid, U_POINTS)
+    zgrv = r3d_field(model_grid, V_POINTS)
+    zwz  = r3d_field(model_grid, U_POINTS)
+    zww  = r3d_field(model_grid, V_POINTS)
+    zdzr = r3d_field(model_grid, T_POINTS)
+
+    call invoke( horizontal_density_gradient_u(zgru,prd),                   &
+                 horizontal_density_gradient_v(zgrv,prd) )
+    if (ln_zps) then
+      call invoke( partial_step_correction_u(zgru,gru),                     &
+                   partial_step_correction_v(zgrv,grv) )
+    end if
+    call invoke( vertical_density_gradient_at_t(zdzr,prd,pn2),              &
+                 slope_mixing_layer_u(uslpml,prd,pn2,zgru,zdzr),            &
+                 slope_mixing_layer_v(vslpml,prd,pn2,zgrv,zdzr),            &
+                 slope_mixing_layer_wi(wslpiml,prd,pn2,zgru),               &
+                 slope_mixing_layer_wj(wslpjml,prd,pn2,zgrv),               &
+                 vertical_density_gradient_at_u(zwz,zgru,zdzr,uslpml),      &
+                 vertical_density_gradient_at_v(zww,zgrv,zdzr,vslpml),      &
+                 horizontal_shapiro_filter(uslp,zwz),                       &
+                 horizontal_shapiro_filter(vslp,zww),                       &
+                 vertical_density_gradient_at_wi(zwz,zgru,prd,pn2,wslpiml), &
+                 vertical_density_gradient_at_wj(zww,zgrv,prd,pn2,wslpjml), &
+                 horizontal_shapiro_filter(wslpi,zwz),                      &
+                 horizontal_shapiro_filter(wslpj,zww) )
+  end subroutine ldf_slp
+
+end module ldfslp_mod

### 2015-11-20 docs/NEMO/example/ldf_slp_alg_processed.f90

@@ -1,0 +1,31 @@
+module ldfslp_mod
+
+  IMPLICIT NONE
+
+contains
+
+  subroutine ldf_slp(uslp,wslpi,vslp,wslpj,...)
+    !! ...
+    !! ** Purpose :   Compute the slopes of neutral surface
+    !! ...
+    type(r3d_field), intent(in)  :: ...
+    type(r3d_field), intent(out) :: uslp,wslpi,vslp,wslpj
+
+    ! local fields
+    type(r3d_field) :: zgru,zgrv,zwz,zww,zdzr
+
+    zgru = r3d_field(model_grid, U_POINTS)
+    zgrv = r3d_field(model_grid, V_POINTS)
+    zwz  = r3d_field(model_grid, U_POINTS)
+    zww  = r3d_field(model_grid, V_POINTS)
+    zdzr = r3d_field(model_grid, T_POINTS)
+
+    call invoke_0(zgru,prd,zgrv)
+    if (ln_zps) then
+      call invoke_1(zgru,gru,zgrv,grv)
+    end if
+    call invoke_2(zdzr,prd,pn2,uslpml,zgru,vslpml,zgrv,wslpiml,wslpjml,zwz,uslpml, &
+                  zww,uslp,vslp,zwz,wslpi,wslpj)
+  end subroutine ldf_slp
+
+end module ldfslp_mod

### 2015-11-20 docs/NEMO/example/ldfslp.F90

@@ -1,0 +1,798 @@
+MODULE ldfslp
+   !!======================================================================
+   !!                       ***  MODULE  ldfslp  ***
+   !! Ocean physics: slopes of neutral surfaces
+   !!======================================================================
+   !! History :  OPA  ! 1994-12  (G. Madec, M. Imbard)  Original code
+   !!            8.0  ! 1997-06  (G. Madec)  optimization, lbc
+   !!            8.1  ! 1999-10  (A. Jouzeau)  NEW profile in the mixed layer
+   !!   NEMO     1.0  ! 2002-10  (G. Madec)  Free form, F90
+   !!             -   ! 2005-10  (A. Beckmann)  correction for s-coordinates
+   !!            3.3  ! 2010-10  (G. Nurser, C. Harris, G. Madec)  add Griffies operator
+   !!             -   ! 2010-11  (F. Dupond, G. Madec)  bug correction in slopes just below the ML
+   !!            3.7  ! 2013-12  (F. Lemarie, G. Madec)  add limiter on triad slopes
+   !!----------------------------------------------------------------------
+
+   !!----------------------------------------------------------------------
+   !!   ldf_slp       : calculates the slopes of neutral surface   (Madec operator)
+   !!   ldf_slp_triad : calculates the triads of isoneutral slopes (Griffies operator)
+   !!   ldf_slp_mxl   : calculates the slopes at the base of the mixed layer (Madec operator)
+   !!   ldf_slp_init  : initialization of the slopes computation
+   !!----------------------------------------------------------------------
+   USE oce            ! ocean dynamics and tracers
+   USE dom_oce        ! ocean space and time domain
+   USE ldfdyn         ! lateral diffusion: eddy viscosity coef.
+   USE phycst         ! physical constants
+   USE zdfmxl         ! mixed layer depth
+   USE eosbn2         ! equation of states
+   !
+   USE in_out_manager ! I/O manager
+   USE prtctl         ! Print control
+   USE lbclnk         ! ocean lateral boundary conditions (or mpp link)
+   USE lib_mpp        ! distribued memory computing library
[…]

### 2015-11-20 docs/NEMO/example/psy.f90

@@ -1,0 +1,46 @@
+! Manually written as an example. Will be generated by PSyclone.
+module psy
+
+  implicit none
+
+contains
+
+  subroutine invoke_0(zgru,prd,zgrv)
+
+    type(r3d_field), intent(inout) :: zgru,prd,zgrv
+
+    use horiz_density_gradient_u_mod, only : horizontal_density_gradient_u_code
+    use horiz_density_gradient_v_mod, only : horizontal_density_gradient_v_code
+    use grid_mod, only : grid
+
+    jpk = grid%upper("K")
+    jpj = grid%upper("J")
+    jpi = grid%upper("I")
+
+    do jk = 1, jpk
+      do jj = 1, jpj-1
+        do ji = 1, jpi-1
+          call horizontal_density_gradient_u_code(ji,jj,jk,zgru%data,prd%data,grid%umask)
+       end do
+      end do
+    end do
+
+    do jk = 1, jpk
+      do jj = 1, jpj-1
+        do ji = 1, jpi-1
+          call horizontal_density_gradient_v_code(ji,jj,jk,zgrv%data,prd%data,grid%vmask)
+       end do
+      end do
+    end do
+
+  end subroutine invoke_0
+
+  subroutine invoke_1(...)
+    ...
+  end subroutine invoke_1
+
+  subroutine invoke_2(...)
+    ...
+  end subroutine invoke_2
+
+end module psy

### 2015-11-20 docs/NEMO/psyclone_and_nemo.tex
NEMO workplan: added content to WP on infrastructure and ref. for parco2015 paper

@@
 specify a set of optimisations in a script, which it then applies to
-the code. The \psykal separation of concerns can be considered a
+the code. The \psykal\ separation of concerns can be considered a
 stripping back of many of the optimisations that have been hard coded
@@
 exceptions to this must be small if an effective Domain Specific
-Language is to be developed for \psyclone\'s application to NEMO.
+Language is to be developed for \psyclone's application to NEMO.
 
@@
 
-\section{Index re-reordering}
+\section{Index Re-ordering}
 
@@
 layout. This question has been asked of \psyclone\ before as it is
-thought that i,j,k may work best on some architectures and k,i,j on
+thought that $i,j,k$ may work best on some architectures and $k,i,j$ on
 others. However, the separation of science code and optimisations and
-parallelism combined with the separation of the i,j,k looping
+parallelism combined with the separation of the $i,j,k$ looping
 structure from the Kernel science code should make it much more
@@
 
-\section{Suggested workplan - 1 Week}
+\section{Suggested Workplan - 1 Week}
 
@@
 
-\section{Suggested workplan - 1 year}
+\section{Suggested Workplan - 1 Year}
 
@@
 
-\subsection{WP3: Infrastructure support}
+\subsection{WP3: Infrastructure Support}
 \label{wp3_infrastructure}
 
-Field objects or arrays?
+\psyclone\ must interface to some sort of infrastructure that provides
+support for manipulating fields - e.g. creating them and performing
+halo swaps. Obviously NEMO already contains support for performing
+domain decomposition and MPI operations. Any incremental introduction
+of \psyclone\ must therefore work within this framework.
 
-The existing NEMO code base is already MPI parallel and therefore any
-introduction of \psyclone\ must work within this framework.
+In the latest versions of GOcean and Dynamo the infrastructure is
+built around the concept of field objects. While these have benefits,
+such an approach would be a significant change to NEMO and may also
+conflict with the use of AGRIF. In the initial stages of this work it
+will therefore probably be necessary for the infrastructure to work
+with fields as Fortran arrays rather than derived types.
 
@@
 transformations. We have shown that this is the case for two
-shallow-water models~\cite{XXX} but this work must be extended for the
-3D parts of NEMO. Benchmarking of the configurations chosen in WP1
-will inform the key routines used to examine this issue. Note that
-this benchmarking must be performed on processor counts typical of
-those used in production runs.
+shallow-water models~\cite{psykal_shallow} but this work must be
+extended for the 3D parts of NEMO. Benchmarking of the configurations
+chosen in WP1 will inform the key routines used to examine this
+issue. Note that this benchmarking must be performed on processor
+counts typical of those used in production runs.
 
@@
 
+\begin{thebibliography}{1}
+
+\bibitem{psykal_shallow} A.~R.~Porter, R.~W.~Ford, M.~Ashworth, G.~D.~Riley and M.~Modani, ``Towards Compiler-Agnostic Performance in Finite-Difference Codes'', to appear in the Proceedings of ParCo 2015.
+
+\end{thebibliography}
+
 \end{document}

### 2015-11-20 docs/NEMO/psyclone_and_nemo.tex
NEMO: add discussion of example code provided by Rupert

@@
 
-\subsection{An example: ldf\_slp\_triad}
+\subsection{An example: ldf\_slp}
 
-Rupert mocking this up.
+As an example, we have manually re-structured the {\it ldf\_slp}
+routine following the \psykal\ separation of concerns. The resulting
+code is too long to include here but may be obtained from the GOcean
+SVN repository on puma:
+\begin{verbatim}
+> svn list https://puma.nerc.ac.uk/svn/GOcean_svn/GOcean/trunk/\
+docs/NEMO/example
+horizontal_density_gradient_u_mod.f90
+horizontal_density_gradient_v_mod.f90
+ldf_slp_alg.f90
+ldf_slp_alg_processed.f90
+ldfslp.F90
+psy.f90
+\end{verbatim}
+
+Here, ldfslp.F90 is the original routine, ldf\_slp\_alg.f90 is the
+Algorithm layer, psy.f90 contains the (manually written) PSy layer and
+the horizontal\_density\_$\ast$.f90 files contain kernels.
+ldf\_slp\_alg\_processed.f90 contains the Algorithm layer as it would
+appear after processing by \psyclone.
 

### 2015-12-02 docs/NEMO/psyclone_and_nemo.tex
NEMO: tweak example text
NEMO: updates to psyclone_and_nemo discussion doc following feedback from Mike Bell

@@
 computation in halos, etc. and parallelism issues, such as placement
-of directives and halo swap calls.
+of directives and halo swap calls. This separation also improves the
+readability and modularity of code and thus greatly aids with
+maintenance.
 
@@
 architecture and almost impossible to have performance portability
-with a single code base. \psyclone\ aims to address this problem.
+with a single code base. The situation is even more complex if, as
+with NEMO, a single code base supports different model configurations
+that may themselves require different optimisations. \psyclone\ aims
+to address this problem by keeping the optimisation space fully
+separate from the science code.
 
@@
 
+Note that this example has only been produced to inform this
+document. We make no claim that the meta-data etc. is correct or
+sufficient for NEMO.
+
 \section{Risks}
@@
 
+\subsection{Run-time Switches}
+
+In common with most simulation packages, NEMO makes use of run-time
+flags in order to enable and disable certain functionality. This
+becomes an issue for PSyclone when this functionality relates to
+choices in the specific Algorithm that is run. As an example, the
+ldf\_slp routine contains code that is only required if partial steps
+are in use:
+
+\begin{verbatim}
+DO jk = 1, jpk             !==   i- & j-gradient of density   ==!
+   DO jj = 1, jpjm1
+      DO ji = 1, fs_jpim1   ! vector opt.
+         zgru(ji,jj,jk) = umask(ji,jj,jk) * ( ... )
+      END DO
+   END DO
+END DO
+IF( ln_zps ) THEN    ! partial steps correction at  bottom ocean level
+   DO jj = 1, jpjm1
+      DO ji = 1, jpim1
+         zgru(ji,jj,mbku(ji,jj)) = ...
+      END DO
+   END DO
+ENDIF
+!
+zdzr(:,:,1) = 0._wp  !== Local vertical density gradient at T-point == !
+DO jk = 2, jpkm1
+   zdzr(:,:,jk) = ...
+END DO
+\end{verbatim}
+
+As PSyclone currently has no support for control-flow statements
+within an Invoke, the Algorithm Layer for this example would have to
+contain three separate Invokes and might look something like:
+
+\begin{verbatim}
+CALL invoke( density_igradient(...), &
+             density_jgradient(...) )
+IF( ln_zps ) THEN
+   CALL invoke( partial_steps_correction(...) )
+END IF
+CALL invoke( local_density_kgradient(...) )
+\end{verbatim}
+
+Since optimisation can only be performed within an Invoke, this
+structure might result in less performant code. If this proved to be
+the case then one might consider re-writing the Algorithm as:
+
+\begin{verbatim}
+IF( ln_zps ) THEN
+   CALL invoke( density_igradient(...),        &
+                density_jgradient(...),        &
+                partial_steps_correction(...), &
+                local_density_kgradient(...) )
+ELSE
+   CALL invoke( density_igradient(...),        &
+                density_jgradient(...),        &
+                local_density_kgradient(...) )
+END IF
+\end{verbatim}
+
+This issue is not NEMO specific and in fact is currently under
+discussion within the GungHo project. Any solution developed there
+will be immediately applicable to NEMO.
+
 \subsection{Tracers}
@@
 models. In those configurations there can be of order $\sim 50$ tracer
-species, all of which may be treated independently. This has
-significant implications for both data structures and our ability to
-exploit the additional parallelism that is available.
+species. Although these can be treated independently within the
+advection routines there are interactions between them in the
+biological routines. This has significant implications for both data
+structures and our ability to exploit the additional parallelism that
+is available. For the latter we would need additional meta-data to
+describe this parallelism and, crucially, indicate when it is not
+available.
 

### 2016-01-07 nemolite2d_perf.tex
nemolite2d paper: update reference to top500 list in intro

@@
 cores. This philosophy can be seen at work in the rise of so-called
-accelerator based machines in the Top 500 List~\cite{top500} of
-supercomputers: five of the top-ten machines in the November 2014 list
-make use of Intel Xeon Phi's or NVIDIA GPUs. Four of the remaining
-five machines in the top ten are IBM BlueGene/Qs, the CPU of which has
-hardware support for running 64 threads.
+accelerator-based machines in the Top 500 List~\cite{top500} of
+supercomputers: four of the top-ten machines in the November 2015 list
+make use of Intel Xeon Phi's or NVIDIA GPUs. Two of the remaining six
+machines are IBM BlueGene/Qs, the CPU of which has hardware support
+for running 64 threads.
 
@@
 \subsection{The `nemolite2d' Program}
-For this work we have used a  program
-
-{\bf Description of the structure of the program at the Algorithm
-  level so that it links in to Profile data in results section.}
+
+For this work we have used a program "nemolite2d" developed by
+ourselves.  “nemolite2d” is a vertically-averaged version of NEMO
+(Nucleus for European Modelling of the Ocean), which only keeps its
+dynamical part. The whole model system is represented through one
+continuity equation (for the update of the sea-surface height) and two
+vertically-integrated momentum equations (for the two velocity
+components, respectively).
+
+(add above three equations in .tex file).
+
+The external forcing includes surface wind stress, bottom friction,
+and open boundary barotropic forcing. A lateral slip boundary
+condition is applied along the coast lines. The open boundary
+conditions can be set as a clipped or a Flather's radiation
+condition~\cite{flather76}. The bottom friction takes a semi-implicit
+form for the sake of model stability. As done in the original version
+of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
+used for the horizontal viscosity term.
+
+The traditional Arakawa C structured grid is employed here for the
+discretisation of the computational domain. A two-dimensional integer
+array is used to identify the different parts of the computational
+domain; it has the value of one for ocean, zero for land and minus
+one for ocean cells outside of the computational domain. This array
+serves as a mark to identify ocean cells, land cells, solid boundaries
+and open boundaries.
+
+For the sake of simplicity, the explicit Eulerian forward time
+stepping method is implemented here, except that the bottom friction
+takes a semi-implicit form.  The Coriolis force can be set in explicit
+or implicit form. The advection term is computed with a first-order
+upwind scheme.
+
+The order of the model computation is as follows:
+\begin{enumerate}
+\item Set the initial conditions (water depth, sea surface height, velocity)
+\item integrate the continuity equation for the new sea surface height
+\item update the different terms in the right hand side of the momentum equations (advection, Coriolis forcing(if set in explicit form), pressure gradient, and, horizontal viscosity
+\item update the velocity vectors by summing up the values in 3), and implicitly with the bottom friction and Coriolis forcing (if set in implicit form)
+\item apply the boundary conditions on to the open boundary and solid boundary cells.
+\end{enumerate}
 

### 2016-01-07 nemolite2d_perf.tex
nemolite2d paper: add notes column to compiler flags table

@@
 \centering
-\begin{tabular}{l|c}
+\begin{tabular}{l|l|l}
 \hline
-Compiler  &  Flags \\
+Compiler  &  Flags                                    & Notes \\
 \hline
-Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
-Intel     & -O3 -fast -fno-inline-factor -xHost   \\
-Cray      & -O3 -O ipa5 -h wp               \\
-IBM       & -O5 -qinline=auto:level=10      \\
-PGI       & {\bf Jeremy to supply} \\
+Gnu       & -Ofast -mtune=native -finline-limit=50000 &   \\
+Intel     & -O3 -fast -fno-inline-factor -xHost       &   \\
+Cray      & -O3 -O ipa5 -h wp                         &   \\
+IBM       & -O5 -qinline=auto:level=10                &   \\
+PGI       & {\bf Jeremy to supply}                    &   \\
 \hline

### 2016-01-08 nemolite2d_perf.tex
nemolite2d paper: add nemo reference

@@
   \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
   % and their extensions so you won't have to specify these with
@@
 
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
 % make the title area
@@
 \begin{IEEEkeywords}
-Performance, Code-generation, Finite-difference
+Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC
 \end{IEEEkeywords}
@@
 ourselves.  “nemolite2d” is a vertically-averaged version of NEMO
-(Nucleus for European Modelling of the Ocean), which only keeps its
+(Nucleus for European Modelling of the Ocean~\cite{nemo}), which only keeps its
 dynamical part. The whole model system is represented through one

### 2016-01-08 nemolite2d_perf.tex
nemolite2d paper: update description of code and b.c.s

@@
 \item integrate the continuity equation for the new sea surface height
-\item update the different terms in the right hand side of the momentum equations (advection, Coriolis forcing(if set in explicit form), pressure gradient, and, horizontal viscosity
+\item update the different terms in the right hand side of the momentum equations; advection, Coriolis forcing (if set in explicit form), pressure gradient, and horizontal viscosity
 \item update the velocity vectors by summing up the values in 3), and implicitly with the bottom friction and Coriolis forcing (if set in implicit form)
-\item apply the boundary conditions on to the open boundary and solid boundary cells.
+\item apply the boundary conditions on the open-  and solid-boundary cells.
 \end{enumerate}
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of nemolite2d, we made the kernels
-as fine-grained as possible. In this case, this resulted in eight
-distinct kernels, each of which operated on a single field at a single
-point (since we have chosen to use point-wise kernels). With a little
-bit of tidying/re-structuring, we found it was possible to express the
+optimisation) of the \psykal version of nemolite2d, we made the
+kernels as fine-grained as possible. This resulted in eight distinct
+kernels, each of which updates a single field at a single point (since
+we have chosen to use point-wise kernels). With a little bit of
+tidying/re-structuring, we found it was possible to express the
 contents of the main time-stepping loop as a single invoke (a call to
 the PSy layer) and a call to the I/O system
-(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single PSy-layer
-routine then consists of applying each of the kernels to all of the
-points on the model mesh requiring an update. In its basic,
-unoptimised (`vanilla') form, this PSy-layer routine then contains a
-doubly-nested loop around each kernel call, as indicated by the
-pseudo-code in Figure~\ref{FIG_psykal_nemolite2d_structure}.
+(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
+gives us a single PSy-layer routine which consists of applying each of
+the kernels to all of the points requiring an update on the model
+mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
+routine then contains a doubly-nested loop around each kernel call, as
+indicated by the pseudo-code in
+Figure~\ref{FIG_psykal_nemolite2d_structure}.
 
@@
 As with any full oceanographic model, boundary conditions must be
-applied at the edges of the model domain.
-
-{\bf Auto-generated BC code by reasoning about meta-data. But then,
-  that brings us on to meta-data...}
+applied at the edges of the model domain. Since nemolite2d applies
+external boundary conditions (e.g. barotropic forcing), this is done
+via user-supplied kernels.
 

### 2016-01-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d paper: fix typos and tweak text. Update details of shallow paper and correct citation of OpenMP

@@
 As with any full oceanographic model, boundary conditions must be
-applied at the edges of the model domain.
-
-{\bf Auto-generated BC code by reasoning about meta-data. But then,
-  that brings us on to meta-data...}
+applied at the edges of the model domain. Since nemolite2d applies
+external boundary conditions ({\it e.g.} barotropic forcing), this is
+done via user-supplied kernels.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methodology(?)}
+\section{Methodology}
 \label{sec_methodology}
@@
 piece of hardware, the version number used in this work is specified.
-The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
-3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
+%The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
+%3.7~GHz and 10~MB of last-level cache. 
+The Intel Ivy Bridge-based
 system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
@@
 % \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
 %
@@
 \hline
-Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
+%Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
 Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
@@
 \caption{The performance profile of the original and \psykal versions
-  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
-  Intel compiler).}
+  of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
+  $128^{2}$ domain and the Intel compiler).}
 \label{TABLE_profile}
@@
 
-\subsubsection{Addition of `safe-address' directives}
+\subsubsection{Addition of {\tt safe\_address} directives}
 \label{sec_safe_address}
@@
 module to allocate all field-data arrays with extents greater than
-strictly required.  We were then able to safely add the safe\_address
+strictly required.  We were then able to safely add the {\tt safe\_address}
 before all of the loops where the Cray compiler indicated it might be
@@
 \end{verbatim}
-Inserting the Cray `vector always' directive persuaded the compiler
+Inserting the Cray {\tt vector always} directive persuaded the compiler
 to vectorise the loop:
@@
 array within the grid object. Although these grid-related quantities
-are read-only within a compute kernel, if they were obtained from the
-same object as that containing the array to which that kernel writes
+are read-only within a compute kernel, if they were referenced from the
+same field object as that containing an array to which the kernel writes
 then the Intel compiler identified a dependency preventing
@@
 that the iterations of the outermost or $j$ loop (over the y-dimension
-of the model domain) are shared amongst the OpenMP threads. This
+of the model domain) are shared out amongst the OpenMP threads. This
 leaves the innermost ($i$) loop available for SIMD vectorisation by
@@
 condition to the y-component of velocity has a loop-carried dependency
-in $j$ which prevented its being executed in parallel. This was
+in $j$ which appears to prevent its being executed in parallel. This was
 therefore left unchanged and executed on thread 0 only.
@@
 Although very simple to implement, the use of separate PARALLEL DO
-directives causes the team of OpenMP threads to be repeatedly created
-and destroyed. This may be avoided by keeping the thread team in
-existence for as long as possible using an OpenMP PARALLEL region.  We
-therefore enclosed the whole of the PSy layer (in this code, a single
+directives results in a lot of thread synchronisation and can also
+cause the team of OpenMP threads to be repeatedly created and
+destroyed. This may be avoided by keeping the thread team in existence
+for as long as possible using an OpenMP PARALLEL region.  We therefore
+enclosed the whole of the PSy layer (in this code, a single
 subroutine) within a single PARALLEL region. The directive preceeding
@@
 
-Since data arrays are managed within the GOcean Library {\bf (is this
-  a good/appropriate name?)} this optimisation can again be
-implemented without changing user code (the Application and Kernel
-layers).
+Since data arrays are managed within the GOcean Library this
+optimisation can again be implemented without changing user code ({\it
+  i.e.} the Application and Kernel layers).
 
@@
 of the original version of the nemolite2d code for the range of
-compilers and CPUs considered here. Unlike the Shallow code, the
-original version of nemolite2d has not been optimised. Although it is
-still a single source file it is, in common with NEMO itself,
-logically structured with separate subroutines performing different
-parts of the physics within each time step. This structuring and the
-heavy use of conditional statements favour the Intel compiler which
-significantly out-performs both the Gnu and Cray compilers. Only when
-the problem size falls out of cache does the performance gap begin to
-narrow. The reason for the performance deficit of the Cray-compiled
-binary comes down to SIMD vectorisation, an issue that we explore
-below.
+compilers and CPUs considered here. Unlike the Shallow
+code~\cite{shallow_psykal}, the original version of nemolite2d has not
+been optimised. Although it is still a single source file it is, in
+common with NEMO itself, logically structured with separate
+subroutines performing different parts of the physics within each time
+step. This structuring and the heavy use of conditional statements
+favour the Intel compiler which significantly out-performs both the
+Gnu and Cray compilers. Only when the problem size falls out of cache
+does the performance gap begin to narrow. The reason for the
+performance deficit of the Cray-compiled binary comes down to (a lack
+of) SIMD vectorisation, an issue that we explore below.
 
@@
 \hline
-\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.62 & 2.39 & 42.22 \\
+\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.6 & 2.39 & 42.2 \\
 \hline
@@
 \centering
-\includegraphics[width=90mm]{opt_stages_256}
+\includegraphics[width=93mm]{opt_stages_256}
 \caption{Performance of the \psykal version of nemolite2d for the
@@
 either the PSy (middle) layer or the underlying library that manages
-e.g. the construction of field objects.
+{\it e.g.} the construction of field objects.
 
@@
 diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
-~\ref{FIG_intel_omp_256}). {\bf Can we explain this?}
+~\ref{FIG_intel_omp_256}). {\bf Can we explain this? Perhaps with HW
+  counters?}
 
@@
 performance - the two-socket performance figures track those from a
-single socket very closely. The only realy difference in performance
+single socket very closely. The only significant difference in performance
 is at 24 threads where, in addition to the difference in available
@@
   threads.}
+
+{\bf Add roofline model to aid understanding of this result.}
 

### 2016-01-11 nemolite2d_perf.tex
nemolite2d paper: correct \psykal definition and start work on Conclusions

@@
 
-\subsection{The \psykal Approach}
+\subsection{The \psykal\ Approach}
 Proliferation of lightweight cores is resulting in increased
@@
 \centering
-\caption{The \psykal Approach.}
+\caption{The \psykal\ Approach.}
 \end{figure}
@@
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal version of nemolite2d retains the Input/Output
+we ensure that any \psykal\ version of nemolite2d retains the Input/Output
 capability of the original. This aids in limiting the optimisations
-that can be performed on the \psykal version to those that should also
+that can be performed on the \psykal\ version to those that should also
 be applicable to full oceanographic models. Note that although we
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal version of nemolite2d, we made the
+optimisation) of the \psykal\ version of nemolite2d, we made the
 kernels as fine-grained as possible. This resulted in eight distinct
@@
 %\includegraphics[width=85mm]{psykal_nemolite2d}
-\caption{A schematic of the vanilla \psykal version of the nemolite2d code.}
+\caption{A schematic of the vanilla \psykal\ version of the nemolite2d code.}
 \label{FIG_psykal_nemolite2d_structure}
@@
 applied at the edges of the model domain. Since nemolite2d applies
-external boundary conditions (e.g. barotropic forcing), this is done
-via user-supplied kernels.
+external boundary conditions ({\it e.g.} barotropic forcing), this is
+done via user-supplied kernels.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methodology(?)}
+\section{Methodology}
 \label{sec_methodology}
@@
 %These flags are particularly important for
-%the \psykal versions of the code. The performance of the original
+%the \psykal\ versions of the code. The performance of the original
 %version of the code is much less sensitive to the compiler
@@
 unoptimised version after it has been re-structured following the
-\psykal approach. In addition to this benchmarking, we profile these
+\psykal\ approach. In addition to this benchmarking, we profile these
 versions of the code at the algorithm level (using a high-level timing
@@
 when we consider the performance loss when moving from the original to
-the \psykal version of the code; it slows down by a factor of
+the \psykal\ version of the code; it slows down by a factor of
 two. Although less significant in terms of absolute time, the
-Continuity section is also dramatically slower in the \psykal version,
+Continuity section is also dramatically slower in the \psykal\ version,
 this time by a factor of three. In contrast, the performance of the
 Time-update and Boundary Condition regions are not significantly
-affected by the move to the \psykal version.
+affected by the move to the \psykal\ version.
 
 \begin{table}[!t]
-\caption{The performance profile of the original and \psykal versions
-  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
-  Intel compiler).}
+\caption{The performance profile of the original and \psykal\ versions
+  of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
+  $128^{2}$ domain and the Intel compiler).}
 \label{TABLE_profile}
@@
 
-Beginning with the vanilla \psykal version, we then apply a series of
-code transformations while obeying the \psykal separation of concerns,
+Beginning with the vanilla \psykal\ version, we then apply a series of
+code transformations while obeying the \psykal\ separation of concerns,
 {\it i.e.} optimisation is restricted to the middle, {PS}y layer and
@@
 therefore chose to attempt to optimise this section first. In-keeping
-with the \psykal approach, we are only permitted to optimise the middle
+with the \psykal\ approach, we are only permitted to optimise the middle
 (PSy) layer which, for this section comprises calls to two kernels,
@@
 It turns out that the Cray-compiled binaries of both the original and
-\psykal versions of nemolite2d perform considerably less well than
+\psykal\ versions of nemolite2d perform considerably less well than
 their Intel-compiled counterparts. Comparison of the diagnostic output
@@
 Having optimised the Momentum section as much as permitted by the
-\psykal approach, we turn our attention to the three remaining
+\psykal\ approach, we turn our attention to the three remaining
 sections of the code. The profile data in Table~\ref{TABLE_profile}
@@
 striking however, is that the cost of the Continuity section increases
-by more than a factor of three in moving to the \psykal version of the
+by more than a factor of three in moving to the \psykal\ version of the
 code.
@@
 
-For this part of the work we began with the optimal \psykal version of
+For this part of the work we began with the optimal \psykal\ version of
 the code, as obtained after applying the various transformations
@@
 
-Moving now the to the \psykal version of nemolite2d,
+Moving now the to the \psykal\ version of nemolite2d,
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest \psykal version for each of the compiler/CPU combinations.
+fastest \psykal\ version for each of the compiler/CPU combinations.
 While the Intel compiler still produces the best-performing binary,
 the Cray-compiled binary is now a very close second. In fact, the
-performance of both the Gnu- and Cray-compiled \psykal versions is
+performance of both the Gnu- and Cray-compiled \psykal\ versions is
 generally significantly greater than that of their respective original
@@
 \includegraphics[width=7.5cm]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any \psykal 
+\caption{Summary of the best performance achieved by any \psykal\ 
 version of nemolite2d with each of the compilers and CPUs under 
@@
 Figure~\ref{FIG_slowdown_summary} plots the percentage difference
-between the performance of the original and the best \psykal versions
+between the performance of the original and the best \psykal\ versions
 of nemolite2d for each compiler/CPU combination. This shows that it is
 only the Intel-compiled binary running the $64^2$ domain that is
-slower with the \psykal version of the code (and then only by some
-3\%). For all other points in the space, the optimised \psykal version
+slower with the \psykal\ version of the code (and then only by some
+3\%). For all other points in the space, the optimised \psykal\ version
 of the code performs better. The results for the Cray compiler are
@@
 \includegraphics[width=2.8in]{slowdown_summary}
-\caption{Comparison of the performance of the best \psykal
+\caption{Comparison of the performance of the best \psykal\
 version with that of the original version of the code. A negative value 
-indicates that the \psykal version is slower than the original.}
+indicates that the \psykal\ version is slower than the original.}
 \label{FIG_slowdown_summary}
@@
 applied to the code. The same data is visualised in
-Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
+Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal\
 version for a given compiler/CPU is given relative to the performance
@@
 \centering
-\includegraphics[width=90mm]{opt_stages_256}
-\caption{Performance of the \psykal version of nemolite2d for the
+\includegraphics[width=93mm]{opt_stages_256}
+\caption{Performance of the \psykal\ version of nemolite2d for the
   $256^{2}$ domain at each stage of optimisation. The first (black)
@@
 nemolite2d code; the introduction of OpenMP and OpenACC directives. In
-keeping with the \psykal approach we do not modify either the
+keeping with the \psykal\ approach we do not modify either the
 Algorithm- or Kernel-layer code.  Any code changes are restricted to
 either the PSy (middle) layer or the underlying library that manages
-e.g. the construction of field objects.
+{\it e.g.} the construction of field objects.
 
@@
 \caption{The scaling behaviour of the most performant OpenMP-parallel
-  version of \psykal nemolite2d for the full range of domain sizes
+  version of \psykal\ nemolite2d for the full range of domain sizes
   considered on the CPU. Results are for the Intel compiler on a
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_gnu}
-\caption{Performance of the OpenMP-parallel version of \psykal
+\caption{Performance of the OpenMP-parallel version of \psykal\
   nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_intel}
-\caption{Performance of the OpenMP-parallel version of \psykal
+\caption{Performance of the OpenMP-parallel version of \psykal\
   nemolite2d for the Intel compiler on a single Intel Ivy Bridge
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_cray}
-\caption{Performance of the OpenMP-parallel version of \psykal
+\caption{Performance of the OpenMP-parallel version of \psykal\
   nemolite2d for the Cray compiler on a single Intel Ivy Bridge
@@
 
+{\bf Add roofline model to aid understanding of this result.}
+
 \begin{figure}
@@
 \includegraphics[width=\picwidth]{1and2sockets}
-\caption{Performance of the OpenMP-parallel version of \psykal
+\caption{Performance of the OpenMP-parallel version of \psykal\
   nemolite2d on one and two sockets of Intel Ivy Bridge. The 24-thread
@@
 %\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
-%\caption{Performance of the OpenMP-parallel version of \psykal
+%\caption{Performance of the OpenMP-parallel version of \psykal\
 %  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
@@
 %\includegraphics[width=\picwidth]{omp_scaling_32_intel}
-%\caption{Performance of the OpenMP-parallel version of \psykal
+%\caption{Performance of the OpenMP-parallel version of \psykal\
 %  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
@@
 %\includegraphics[width=\picwidth]{omp_scaling_32_cray}
-%\caption{Performance of the OpenMP-parallel version of \psykal
+%\caption{Performance of the OpenMP-parallel version of \psykal\
 %  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
@@
 \includegraphics[width=\picwidth]{cpu_cf_gpu}
-\caption{Performance of the best OpenMP-parallel version of \psykal
+\caption{Performance of the best OpenMP-parallel version of \psykal\
   nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
@@
 
-We have investigated the application of the \psykal separation of
-concerns approach to the domain of finite-difference ocean
-models. This approach enables the computational science (performance)
-related aspects of a computer model to be kept separate from the
-natural (oceanographic) science aspects. In this work we have shown
-that the \psykal approach has the potential to improve performance and
-to improve code structure/maintainability.
+We have investigated the application of the \psykal\ separation of
+concerns approach to the domain of shared-memory parallel,
+finite-difference ocean models. This approach enables the
+computational science (performance) related aspects of a computer
+model to be kept separate from the natural (oceanographic) science
+aspects.
+
+We have used a new and un-optimised, two-dimension model extracted
+from the NEMO ocean model for this work. As a consequence of this, the
+introduction of the \psykal\ separation of concerns followed by
+suitable transformations of the PSy Layer is found to improve
+performance. This is in contrast to our previous
+experience~\cite{shallow_psykal} with tackling the Shallow code which
+has had been optimised over the course of several years. In that case
+we were able to recover\ldots
 
 The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal version of a code.
-
+to the performance of the \psykal\ version of a code.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % use section* for acknowledgement
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}) and some {\bf NVIDIA system}.
+(\url{http://www.archer.ac.uk}) and some {\bf NVIDIA system - or is it
+  Emerald?}.
 

### 2016-01-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d paper: add roofline plot to makefile and document

@@
 when we consider the performance loss when moving from the original to
-the \psykal version of the code; it slows down by a factor of
+the \psykal\ version of the code; it slows down by a factor of
 two. Although less significant in terms of absolute time, the
-Continuity section is also dramatically slower in the \psykal version,
+Continuity section is also dramatically slower in the \psykal\ version,
 this time by a factor of three. In contrast, the performance of the
 Time-update and Boundary Condition regions are not significantly
-affected by the move to the \psykal version.
+affected by the move to the \psykal\ version.
 
 \begin{table}[!t]
-\caption{The performance profile of the original and \psykal versions
+\caption{The performance profile of the original and \psykal\ versions
   of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
@@
 \includegraphics[width=93mm]{opt_stages_256}
-\caption{Performance of the \psykal version of nemolite2d for the
+\caption{Performance of the \psykal\ version of nemolite2d for the
   $256^{2}$ domain at each stage of optimisation. The first (black)
@@
 nemolite2d code; the introduction of OpenMP and OpenACC directives. In
-keeping with the \psykal approach we do not modify either the
+keeping with the \psykal\ approach we do not modify either the
 Algorithm- or Kernel-layer code.  Any code changes are restricted to
@@
 
-{\bf Add roofline model to aid understanding of this result.}
-
 \begin{figure}
@@
 
+\subsection{Performance Analysis with the Roofline Model}
+
+Although we have investigated how the performance of the \psykal
+version of nemolite2d compares with that or the original, we have not
+addressed how efficient the original actually is.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{roofline}
+\caption{Comparison of the performance achieved by nemolite2d and Shallow}
+\label{FIG_roofline}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-01-11 nemolite2d_perf.tex
nemolite2d paper: explain calc for roofline model

@@
 
+\subsection{Performance Analysis with the Roofline Model}
+
+Although we have investigated how the performance of the \psykal
+version of nemolite2d compares with that or the original, we have not
+addressed how efficient the original actually is. In order to do so we
+use the Roofline Model~\cite{roofline} which provides a relatively
+simple way of characterising the performance of a code in terms of
+whether it is memory-bandwidth bound or compute bound. To do so we
+follow the approach suggested in~\cite{para_pearls} and use the
+well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
+in order to get the upper bounds on the memory bandwidth and FLOPS for
+the test hardware. Since we are using an Intel Ivybridge CPU we used
+the Intel Math Kernel Library implementation of LINPACK.
+
+A key component of the Roofline model is the Arithmetic Intensity of
+the code being executed:
+\begin{equation}
+AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
+\end{equation}
+We calculated this quantity manually by examining the
+source code and counting the number of memory references and
+arithmetic operations that it contained. In doing this counting we
+assume that any references to adjacent array elements are fetched in a
+single cache-line and thus only count once.
+
+To illustrate this, consider the following code fragment (which forms
+the body of a doubly-nested loop over {\tt i} and {\tt j} with {\tt i}
+innermost):
+\begin{verbatim}
+CU(I+1,J) = .5d0*(P(I+1,J)+P(I,J))*U(I+1,J)
+CV(I,J+1) = .5d0*(P(I,J+1)+P(I,J))*V(I,J+1)
+Z(I+1,J+1) =(FSDX*(V(I+1,J+1)-V(I,J+1))- &
+             FSDY*(U(I+1,J+1)-U(I+1,J)))/&
+            (P(I,J)+P(I+1,J)+            &
+             P(I+1,J+1)+P(I,J+1))
+H(I,J) = P(I,J)+0.25d0*                   &
+         (U(I+1,J)*U(I+1,J)+U(I,J)*U(I,J) & 
+         +V(I,J+1)*V(I,J+1)+V(I,J)*V(I,J))
+\end{verbatim}
+This code writes to four memory locations ({\tt CU}, {\tt CV}, {\tt Z}
+and {\tt H}) and reads from ten distinct locations.  However, if we
+assume caching of reads from adjacent locations ({\it e.g.} {\tt
+  P(i,j)} and {\tt P(i+1,j)}) then we have only six distinct reads.
+If we assume that each of these ten memory accesses will touch a
+separate cache line then this code fragment will result in the moving
+of ten cache lines, each of which is 64 bytes (eight double-precision
+words). However, as noted above, this code fragment is the body of a
+loop nest. Therefore, upon the next loop trip {\tt i} will have been
+incremented by one and the majority of the accessed values will still
+be in cache. In fact, a new cache line will only have to be fetched
+every seven trips of the (innermost) loop and thus the mean memory
+traffic per loop trip is approximately $640/7$ bytes.
+
+We can also count the number of floating-point arithmetic operations
+in this code fragment which gives us 12 multiplications and 12
+additions. $AI$ for this code fragment is then $24*7/640 = 0.26$.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{roofline}
+\caption{Comparison of the performance achieved by nemolite2d and Shallow}
+\label{FIG_roofline}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-01-13 nemolite2d_perf.tex
nemolite2d paper: move detailed discussion of roofline model into its own document

--- 2016-01-11 nemolite2d_perf.tex
@@
 
+\subsection{Performance Analysis with the Roofline Model}
+
+Although we have investigated how the performance of the \psykal
+version of nemolite2d compares with that or the original, we have not
+addressed how efficient the original actually is. In order to do so we
+use the Roofline Model~\cite{roofline} which provides a relatively
+simple way of characterising the performance of a code in terms of
+whether it is memory-bandwidth bound or compute bound. To do so we
+follow the approach suggested in~\cite{para_pearls} and use the
+well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
+in order to get the upper bounds on the memory bandwidth and FLOPS for
+the test hardware. Since we are using an Intel Ivybridge CPU we used
+the Intel Math Kernel Library implementation of LINPACK.
+
+A key component of the Roofline model is the Arithmetic Intensity of
+the code being executed:
+\begin{equation}
+AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
+\end{equation}
+We calculated this quantity manually by examining the
+source code and counting the number of memory references and
+arithmetic operations that it contained. In doing this counting we
+assume that any references to adjacent array elements are fetched in a
+single cache-line and thus only count once.
+
+\begin{figure}
+\centering
+\includegraphics[width=\picwidth]{roofline}
+\caption{Comparison of the performance achieved by nemolite2d and Shallow}
+\label{FIG_roofline}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-01-13 docs/papers/roofline_model/gocean_roofline.tex
nemolite2d paper: move detailed discussion of roofline model into its own document

\documentclass[12pt]{article}
\usepackage[pdftex]{graphicx}
% and their extensions so you won't have to specify these with
% every instance of \includegraphics
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\usepackage{url}

\newcommand{\psykal}{{PS}y{KA}l}

\begin{document}

\title{Analysing the Performance of Shallow-Water Models with the
 Roofline Model}

\author{A.~.R.~Porter and R.~W.~Ford}

\maketitle

\section{Performance Analysis with the Roofline Model}

Although we have investigated how the performance of the \psykal
version of nemolite2d compares with that or the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
in order to get the upper bounds on the memory bandwidth and FLOPS for
the test hardware. Since we are using an Intel CPUs we used
the Intel Math Kernel Library implementation of LINPACK.

A key component of the Roofline model is the Arithmetic Intensity of
the code being executed:
\begin{equation}
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\end{equation}
We calculated this quantity manually by examining the
source code and counting the number of memory references and
arithmetic operations that it contained. In doing this counting we
assume that any references to adjacent array elements are fetched in a
single cache-line and thus only count once.

To illustrate this, consider the following code fragment (which forms
the body of a doubly-nested loop over {\tt i} and {\tt j} with {\tt i}
innermost):
\begin{verbatim}
CU(I+1,J) = .5d0*(P(I+1,J)+P(I,J))*U(I+1,J)
CV(I,J+1) = .5d0*(P(I,J+1)+P(I,J))*V(I,J+1)
Z(I+1,J+1) =(FSDX*(V(I+1,J+1)-V(I,J+1))- &
             FSDY*(U(I+1,J+1)-U(I+1,J)))/&
            (P(I,J)+P(I+1,J)+            &
             P(I+1,J+1)+P(I,J+1))
H(I,J) = P(I,J)+0.25d0*                   &
         (U(I+1,J)*U(I+1,J)+U(I,J)*U(I,J) & 
         +V(I,J+1)*V(I,J+1)+V(I,J)*V(I,J))
\end{verbatim}
This code writes to four memory locations ({\tt CU}, {\tt CV}, {\tt Z}
and {\tt H}) and reads from ten distinct locations.  However, if we
assume caching of reads from adjacent locations ({\it e.g.} {\tt
  P(i,j)} and {\tt P(i+1,j)}) then we have only six distinct reads.
If we assume that each of these ten memory accesses will touch a
separate cache line then this code fragment will result in the moving
of ten cache lines, each of which is 64 bytes (eight double-precision
words). However, as noted above, this code fragment is the body of a
loop nest. Therefore, upon the next loop trip {\tt i} will have been
incremented by one and the majority of the accessed values will still
be in cache. In fact, a new cache line will only have to be fetched
every seven trips of the (innermost) loop and thus the mean memory
traffic per loop trip is approximately $640/7$ bytes.

We can also count the number of floating-point arithmetic operations
in this code fragment which gives us 12 multiplications and 12
additions. $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) x-Momentum
kernel of nemolite2d then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to
sin(). There are 31 distinct memory references. This gives an $AI$ of
approximately $108*7/(31*64)$ or 0.38 FLOPs/byte.

We measure the mean time to execute this region of code as 0.0020
seconds (Intel v.16 on Haswell E5-1620 v2) for the 256 domain. This
gives us $\frac{108*256^2}{0.002} = 3.5$ GFLOPS. In contrast,
instrumenting this section of code with the LIKWID marker API and
using likwid-perfctr we measure 4.7 GFLOPS.

For the 128 the measured time per kernel call is 0.570033E-03 s.
$108*128^2/0.570e-3 = 3.1$ GFLOPS which is 25\% less than we measure
directly with likwid.  We can use the value reported by likwid to get
the FLOP count: 4.269 GFLOPS $\times 1.677 s = 7.1615217$ GFLOPs. This is for
3000 steps of $128^2$ so one kernel call is: $7.1615217/(3000*128^2) =
145.7$ FLOPs. This means that $AI = 145.7*7/(31*64) = 0.514$ FLOP/byte

\begin{figure}
\centering
\includegraphics[width=120mm]{roofline_archer}
\caption{Comparison of the performance achieved by nemolite2d and Shallow on Archer (Intel Ivybridge)}
\label{FIG_roofline_archer}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=120mm]{roofline_haswell}
\caption{Comparison of the performance achieved by nemolite2d and Shallow on Intel Haswell}
\label{FIG_roofline_haswell}
\end{figure}

\bibliographystyle{unsrt}
\bibliography{gocean_roofline}

\end{document}

### 2016-01-13 docs/papers/roofline_model/gocean_roofline.tex
roofline: add table of basic data for two problem sizes of nemolite2d

@@
 
+\begin{table}
+\begin{tabular}{l|l|l}
+\hline
+\hline
+                                    &    128       &   256 \\
+\hline
+        Runtime (RDTSC) [s]        &   1.697142   &     6.446171  \\
+        Runtime unhalted [s]       & 1.750448e+00 &   6.734944e+00 \\
+            Clock [MHz]            & 3.818981e+03 &   3.875309e+03 \\
+                CPI                & 6.503796e-01 &   6.348858e-01 \\
+\hline
+        L1 DTLB load misses        &    1317912   &    6748712   \\
+       L1 DTLB load miss rate      & 1.326630e-04 & 1.723565e-04 \\
+  L1 DTLB load miss duration [Cyc] & 2.101039e+01 & 2.013584e+01 \\
+        L1 DTLB store misses       &     87257    &    401036    \\
+      L1 DTLB store miss rate      & 8.783423e-06 & 1.024213e-05 \\
+ L1 DTLB store miss duration [Cyc] & 2.151055e+01 & 2.669793e+01 \\
+\hline
+  L2D load bandwidth [MBytes/s] & 1.080936e+04 & 1.446214e+04 \\
+  L2D load data volume [GBytes] &  18.1215488  &   92.177536  \\
+ L2D evict bandwidth [MBytes/s] & 4.501487e+02 & 5.065637e+02 \\
+ L2D evict data volume [GBytes] &  0.75465984  &  3.22869248  \\
+     L2 bandwidth [MBytes/s]    & 1.130282e+04 & 1.498363e+04 \\
+     L2 data volume [GBytes]    & 18.948807808 & 95.501414592 \\
+\hline
+  L3 load bandwidth [MBytes/s] & 4.511686e+03 & 4.607617e+03 \\
+  L3 load data volume [GBytes] &   7.6168768  &   29.692832  \\
+ L3 evict bandwidth [MBytes/s] & 2.771148e+02 & 2.722823e+02 \\
+ L3 evict data volume [GBytes] &  0.467840512 &  1.75466688  \\
+    L3 bandwidth [MBytes/s]    & 4.788801e+03 & 4.879899e+03 \\
+    L3 data volume [GBytes]    &  8.084717312 &  31.44749888 \\
+\hline
+\hline
+\end{tabular}
+\end{table}
+
 \bibliographystyle{unsrt}

### 2016-01-19 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more results to doc

--- 2016-01-13 docs/papers/roofline_model/gocean_roofline.tex
@@
 \maketitle
+
+\section{Introduction}
+
+We want to use the Roofline Model to establish the efficiency (or
+otherwise) of two shallow-water benchmark codes that we have used in
+the GOcean project.
+
+\section{Hardware Details}
+
+CPU name:       Intel(R) Xeon(R) CPU E5-1620 v2 @ 3.70GHz
+CPU type:       Intel Xeon IvyBridge EN/EP/EX processor
+
+Level 1 32 kB
+Level 2 256 kB
+Level 3 10 MB
+
+\section{Hardware Performance Counters}
+
+The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
+access to Hardware Performance Counters (HWPCs) on Intel and AMD CPUs.
+However, there are issues with the accuracy of these counters,
+especially for FLOP-related measurements on Ivybridge architectures. A
+tool is included in the likwid distribution that tests the accuracy of
+HWPC-derived data by running micro-benchmarks for which the FLOP count
+and memory traffic are known:
+\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
+The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
+Figures~\ref{FIG_flops_dp_stream_test}
+and~\ref{FIG_flops_dp_triad_test}. It appears that once the memory
+footprint spills out of L1 cache (which all of our nemolite2d
+benchmark configurations do), the performance counter results
+significantly overestimate the FLOPs performed.
+
+\begin{figure}
+\centering
+\includegraphics[width=120mm]{FLOPS_DP_stream_ivybridgeEP.pdf}
+\caption{Results of the likwid accuracy test for the FLOPS\_DP group
+  and the stream benchmark on an Intel Xeon E5-1620 v2.}
+\label{FIG_flops_dp_stream_test}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=120mm]{FLOPS_DP_triad_ivybridgeEP.pdf}
+\caption{Results of the likwid accuracy test for the FLOPS\_DP group
+  and the triad benchmark on an Intel Xeon E5-1620 v2.}
+\label{FIG_flops_dp_triad_test}
+\end{figure}
+
+We therefore need some other way of measuring FLOPS. We could use
+likwid on an older Intel CPU (to which we have root access) or try
+another tool on the IvybridgeEP system. Other possible tools include
+Intel's VTune or OProfile.
 
@@
 
+\begin{table}
+\begin{tabular}{l|l|l|l|l}
+\hline
+\hline
+               & \multicolumn{4}{c}{Problem size} \\
+               & 32     &    64        &    128       &   256 \\
+\hline
+Footprint (MB) & 0.1 & 0.5          & 2.3          & 9.2 \\
+\hline
+Runtime (RDTSC) [s]   & 0.144868 &  0.4485108   &   1.697142   &   6.446171  \\
+Runtime unhalted [s]  & 1.199456e-01 & 4.475135e-01 & 1.750448e+00 &   6.734944e+00 \\
+CPI                   & 6.431867e-01 & 6.423220e-01 & 6.503796e-01 &   6.348858e-01 \\
+\hline
+MFLOPS           & 3.112780e+03 & 3.877720e+03  & 4.163209e+03 & 4.304728e+03 \\
+AVX MFLOPS       & 2.840625e+03 & 3.660086e+03  & 3.993397e+03 & 4.164647e+03 \\
+Packed MUOPS/s   & 7.423377e+02 & 9.584935e+02  & 1.046781e+03 & 1.092280e+03 \\
+Scalar MUOPS/s   & 2.077927e+02 & 1.306897e+02  & 7.294811e+01 & 3.784506e+01 \\
+\hline
+L1 DTLB load misses          &      2007     &      17015   &    1317912   &    6748712   \\
+L1 DTLB load miss rate       &  2.949686e-06 &  6.607264e-06& 1.326630e-04 & 1.723565e-04 \\
+L1 DTLB load miss durn [Cyc] &  3.970154e+01 &  3.080153e+01& 2.101039e+01 & 2.013584e+01 \\
+L1 DTLB store misses         &       77      &       16     &     87257    &    401036    \\
+L1 DTLB store miss rate      &  1.131668e-07 &  6.213119e-09& 8.783423e-06 & 1.024213e-05 \\
+L1 DTLB store miss durn [Cyc]&  5.633766e+01 &       29     & 2.151055e+01 & 2.669793e+01 \\
+\hline
+L2D load bandwidth [MB/s]  &  3.811099e+03 &  5.030691e+03 & 1.080936e+04 & 1.446214e+04 \\
+L2D load data volume [GB]  &   0.541220288 &   2.34655488  &  18.1215488  &   92.177536  \\
+L2D evict bandwidth [MB/s] &  3.131102e+02 &  3.094459e+02 & 4.501487e+02 & 5.065637e+02 \\
+L2D evict data volume [GB] &   0.04446528  &   0.144340352 &  0.75465984  &  3.22869248  \\
+L2 bandwidth [MB/s]        &  4.600834e+03 &  5.493404e+03 & 1.130282e+04 & 1.498363e+04 \\
+L2 data volume [GB]        &   0.653371776 &   2.562386304 & 18.948807808 & 95.501414592 \\
+\hline
+L3 load bandwidth [MB/s]  &  2.434255e+03 & 4.626177e+03 & 4.511686e+03 & 4.607617e+03 \\
+L3 load data volume [GB]  &   0.371968512 &  2.07489024  &   7.6168768  &   29.692832  \\
+L3 evict bandwidth [MB/s] &  5.652167e+02 & 4.828169e+02 & 2.771148e+02 & 2.722823e+02 \\
+L3 evict data volume [GB] &   0.086368448 &  0.216548608 &  0.467840512 &  1.75466688  \\
+L3 bandwidth [MB/s]       &  2.999472e+03 & 5.108994e+03 & 4.788801e+03 & 4.879899e+03 \\
+L3 data volume [GB]       &   0.45833696  &  2.291438848 &  8.084717312 &  31.44749888 \\
+\hline
+\hline
+\end{tabular}
+\caption{Hardware performance counter data for the u-Momentum kernel on the Intel E5-1620 v2 processor.}
+\label{TAB_HWPC_MOM_U}
+\end{table}
+
 \bibliographystyle{unsrt}

### 2016-01-20 docs/papers/roofline_model/gocean_roofline.tex
roofline: start to add description of kernels

@@
 \section{Performance Analysis with the Roofline Model}
+
+\subsection{Kernel Characteristics}
+
+Although the Continuity kernel is very simple, the u-Momentum kernel
+is not and in particular, includes a call to the {\tt sin()} function.
+\begin{verbatim}
+  my_sum = 0.0d0
+  do i = 1, 1000
+    my_sum = my_sum + sin(real(i))
+  end do
+  write (*,*) 'My sum = ', my_sum
+\end{verbatim}
+
+\begin{table}
+\begin{tabular}{l|c|c}
+                         & Continuity & u-Momentum \\
+FLOPs                    &    14      &   108      \\
+Distinct memory refs     &    10      &   31       \\
+Arithmetic intensity (FLOPs/byte) & 0.153  & 0.38  \\
+Working set size (bytes) &  $64n^2$   &            \\
+\end{tabular}
+\end{table}
 

### 2016-01-20 docs/papers/roofline_model/gocean_roofline.tex
roofline: first R version of plot and tweaks to doc

@@
 the GOcean project.
+
+Although we have investigated how the performance of the \psykal
+version of nemolite2d compares with that or the original, we have not
+addressed how efficient the original actually is. In order to do so we
+use the Roofline Model~\cite{roofline} which provides a relatively
+simple way of characterising the performance of a code in terms of
+whether it is memory-bandwidth bound or compute bound.
 
@@
 
-\subsection{Kernel Characteristics}
+\subsection{Calculating Kernel Characteristics}
 
@@
 is not and in particular, includes a call to the {\tt sin()} function.
+In order to estimate the cost of this function in FLOPs the following
+code was compiled (with the Intel compiler and -O1) and run with
+likwid-perfctr:
 \begin{verbatim}
@@
 \end{verbatim}
-
-\begin{table}
-\begin{tabular}{l|c|c}
-                         & Continuity & u-Momentum \\
-FLOPs                    &    14      &   108      \\
-Distinct memory refs     &    10      &   31       \\
-Arithmetic intensity (FLOPs/byte) & 0.153  & 0.38  \\
-Working set size (bytes) &  $64n^2$   &            \\
-\end{tabular}
-\end{table}
-
-Although we have investigated how the performance of the \psykal
-version of nemolite2d compares with that or the original, we have not
-addressed how efficient the original actually is. In order to do so we
-use the Roofline Model~\cite{roofline} which provides a relatively
-simple way of characterising the performance of a code in terms of
-whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\cite{para_pearls} and use the
-well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
-in order to get the upper bounds on the memory bandwidth and FLOPS for
-the test hardware. Since we are using an Intel CPUs we used
-the Intel Math Kernel Library implementation of LINPACK.
+This measurement revealed that the {\tt sin()} function requires
+approximately 13 FLOPs.
 
@@
 
-To illustrate this, consider the following code fragment (which forms
-the body of a doubly-nested loop over {\tt i} and {\tt j} with {\tt i}
-innermost):
+To illustrate this, we consider the first loop-nest in the
+time-stepping section of the Shallow code. This is a doubly-nested
+loop over {\tt i} and {\tt j} with {\tt i} innermost). The body of the
+loop is:
 \begin{verbatim}
@@
 additions. $AI$ for this code fragment is then $24*7/640 = 0.26$
-FLOPs/byte.  If we do the same analysis for the (longer) x-Momentum
+FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
 kernel of nemolite2d then we find that there are 51 addition
@@
 sin(). There are 31 distinct memory references. This gives an $AI$ of
-approximately $108*7/(31*64)$ or 0.38 FLOPs/byte.
+approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte.
+
+\begin{table}
+  \begin{tabular}{l|c|c|c}
+                           & Shallow & \multicolumn{2}{c}{nemolite2d} \\
+                           & 1st Loop nest & Continuity & u-Momentum \\
+\hline                                
+FLOPs                      & 24      &    14      &   107+13   \\
+Distinct memory refs       & 10      &    10      &   31       \\
+$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
+Working set size (bytes)   & $56n^2$ &  $64n^2$   &            \\
+\end{tabular}
+\caption{Details of the two nemolite2d kernels investigated with the
+  roofline model.}
+\label{TAB_kernel_details}
+\end{table}
+
+In applying the roofline model we follow the approach suggested
+in~\cite{para_pearls} and use the well-known STREAM~\cite{stream} and
+LINPACK~\cite{linpack} benchmarks in order to get the upper bounds on
+the memory bandwidth and FLOPS for the test hardware. Since we are
+using an Intel CPUs we used the Intel Math Kernel Library
+implementation of LINPACK.
 

### 2016-01-20 docs/papers/roofline_model/roofline_haswell.R

@@ -1,0 +1,57 @@
+# CPU CONSTANTS
+# For single core of Xeon E5-1620 v2 (my desktop), as measured with 
+# the Intel MKL version of linpack. This is therefore using
+# 256-bit AVX instructions (SIMD)
+PEAK_GFLOPS= 28.1
+# An AVX instruction can do 4 d.p. FLOPs in one go so if we were
+# to turn it off altogether we get a factor four reduction:
+NOSIMD_PEAK_GFLOPS <- PEAK_GFLOPS/4.0
+
+# MEM CONSTANTS
+# For single core of E5-1620 v2 as measured with STREAM
+# with arrays of 15M elements. Therefore, this is bandwidth to 
+# main memory, not cache. Units are GB/s.
+PEAK_MEM_BW= 14.5
+# Using arrays of 0.25M elements I think we get bandwidth to
+# L3 cache:
+PEAK_L3_BW= 39.6
+
+XMIN= 0.05
+XMAX= 4.0
+
+cpu_roof <- PEAK_GFLOPS
+PEAK_BW <- max(PEAK_MEM_BW,PEAK_L3_BW)
+
+mydata <- read.table("./roofline.dat", sep=",")
+xvals <- seq(from=XMIN, to=XMAX, length.out=100)
+yvals <- mydata$V2
+# Width of bars to draw
+barW <- 0.02
+
+mem_roof <- function(x, peak) {
+  result <- x*peak
+  return(result)
+}
+
+cpu_ceiling <- function(x, peak_flops) {
+    roof <- x*PEAK_BW
+    vals <- rep(peak_flops, length(x)) 
+    result <- pmin(roof, vals)
+    return(result)
+}
+
+peak_cpu_ceiling <- function(x) cpu_ceiling(x, PEAK_GFLOPS)
+nosimd_cpu_ceiling <- function(x) cpu_ceiling(x, NOSIMD_PEAK_GFLOPS)
+
+mem_ceiling <- function(x) {
+    mem_roof <- x*PEAK_MEM_BW
+    peaklist <- rep(PEAK_GFLOPS, length(x))
+    result <- pmin(mem_roof, peaklist)
+    return(result)
+}
+
+plot(mydata$V1, mydata$V2, log="xy", xlim=c(XMIN,XMAX), ylim=c(0.01,30.0), xlab="Operational intensity (FLOPs/byte)", ylab="GFLOPS")
[…]

### 2016-01-20 docs/papers/roofline_model/roofline_haswell.R
Roofline: add haswell results file and update name in R script

@@ -22,7 +22,7 @@ XMAX= 4.0
 cpu_roof <- PEAK_GFLOPS
 PEAK_BW <- max(PEAK_MEM_BW,PEAK_L3_BW)
 
-mydata <- read.table("./roofline.dat", sep=",")
+mydata <- read.table("./roofline_haswell.dat", sep=",")
 xvals <- seq(from=XMIN, to=XMAX, length.out=100)
 yvals <- mydata$V2
 # Width of bars to draw
@@ -54,4 +54,4 @@ plot(mydata$V1, mydata$V2, log="xy", xlim=c(XMIN,XMAX), ylim=c(0.01,30.0), xlab=
 rect(xleft=(1.0-barW)*mydata$V1, ybottom=0.01, xright=(1.0+barW)*mydata$V1, ytop=yvals)
 curve(mem_ceiling, xvals, add=TRUE, col="blue")
 curve(nosimd_cpu_ceiling, xvals, add=TRUE, col="blue")
-curve(peak_cpu_ceiling, xvals, add=TRUE, col="red")
+curve(peak_cpu_ceiling, xvals, add=TRUE, col="red", lty=1, lwd=2)

### 2016-01-20 docs/papers/roofline_model/roofline_haswell.R
roofline: Add label info to data file

@@ -18,6 +18,8 @@ PEAK_L3_BW= 39.6
 
 XMIN= 0.05
 XMAX= 4.0
+YMIN= 0.5
+YMAX= 32.0
 
 cpu_roof <- PEAK_GFLOPS
 PEAK_BW <- max(PEAK_MEM_BW,PEAK_L3_BW)
@@ -25,6 +27,8 @@ PEAK_BW <- max(PEAK_MEM_BW,PEAK_L3_BW)
 mydata <- read.table("./roofline_haswell.dat", sep=",")
 xvals <- seq(from=XMIN, to=XMAX, length.out=100)
 yvals <- mydata$V2
+labels <- mydata$V3
+
 # Width of bars to draw
 barW <- 0.02
 
@@ -50,8 +54,10 @@ mem_ceiling <- function(x) {
     return(result)
 }
 
-plot(mydata$V1, mydata$V2, log="xy", xlim=c(XMIN,XMAX), ylim=c(0.01,30.0), xlab="Operational intensity (FLOPs/byte)", ylab="GFLOPS")
+plot(mydata$V1, mydata$V2, log="xy", xlim=c(XMIN,XMAX), ylim=c(YMIN,YMAX), xlab="Operational intensity (FLOPs/byte)", ylab="GFLOPS")
+axis(2, at=c(0.5,1,2,4,8,16,32))
 rect(xleft=(1.0-barW)*mydata$V1, ybottom=0.01, xright=(1.0+barW)*mydata$V1, ytop=yvals)
 curve(mem_ceiling, xvals, add=TRUE, col="blue")
 curve(nosimd_cpu_ceiling, xvals, add=TRUE, col="blue")
 curve(peak_cpu_ceiling, xvals, add=TRUE, col="red", lty=1, lwd=2)
+grid(equilogs=FALSE)

### 2016-01-21 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
add model equations

@@
 %
-%\usepackage[cmex10]{amsmath}
+\usepackage[cmex10]{amsmath}
 % A popular package from the American Mathematical Society that provides
@@
 dynamical part. The whole model system is represented through one
-continuity equation (for the update of the sea-surface height) and two
-vertically-integrated momentum equations (for the two velocity
+continuity equation \eqref{continuity} (for the update of the sea-surface height) and two
+vertically-integrated momentum equations \eqref{momentum} (for the two velocity
 components, respectively).
 
-(add above three equations in .tex file).
+\begin{equation}\label{continuity}
+ \frac{\partial \zeta}{\partial t} + \vec{\nabla} \cdot (\vec{U}h) = 0
+\end{equation}
+\begin{equation}\label{momentum}
+ \frac{\partial \vec{U}h}{\partial t}
+   + \vec{U} \cdot \vec{\nabla} (\vec{U}h) = 
+   - gh \vec{\nabla} \zeta - 2h \vec{\Omega} \times \vec{u}
+   + \nu h \Delta \vec{U}
+\end{equation}
+
+where $\zeta$ and $\vec{U}$ represent the sea surface height and horizontal velocity vector
+respectively. $h$ is the total water depth, $\Omega$ is the Earth rotation velocity vector.
+$g$ is the gravity. $\nu$ is the kinematic viscosity coefficient.
 
@@
 memory. At this stage the OpenACC version of the code on the GPU is
-some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
+some {\bf 3.7 (this is the factor for the CUDA version!)}sssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssss times faster
 than the OpenMP version on the CPU. {\bf Can we relate this to the
@@
 version of nemolite2d compares with that or the original, we have not
-addressed how efficient the original actually is.
+addressed how efficient the original actually is. In order to do so we
+use the Roofline Model~\cite{roofline} which provides a relatively
+simple way of characterising the performance of a code in terms of
+whether it is memory-bandwidth bound or compute bound. To do so we
+follow the approach suggested in~\cite{para_pearls} and use the
+well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
+in order to get the upper bounds on the memory bandwidth and FLOPS for
+the test hardware. Since we are using an Intel Ivybridge CPU we used
+the Intel Math Kernel Library implementation of LINPACK.
+
+A key component of the Roofline model is the Arithmetic Intensity of
+the code being executed:
+\begin{equation}
+AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
+\end{equation}
+We calculated this quantity manually by examining the
+source code and counting the number of memory references and
+arithmetic operations that it contained. In doing this counting we
+assume that any references to adjacent array elements are fetched in a
+single cache-line and thus only count once.
 

### 2016-01-21 docs/papers/roofline_model/gocean_roofline.tex
roofline: updates with linpack/stream results on Archer and desktop

@@
 
-CPU name:       Intel(R) Xeon(R) CPU E5-1620 v2 @ 3.70GHz
-CPU type:       Intel Xeon IvyBridge EN/EP/EX processor
+In applying the roofline model we follow the approach suggested
+in~\cite{para_pearls} and use the well-known STREAM~\cite{stream} and
+LINPACK~\cite{linpack} benchmarks in order to get the upper bounds on
+the memory bandwidth and FLOPS for the test hardware. Since we are
+using Intel CPUs we used the Intel Math Kernel Library
+implementation of LINPACK.
 
-Level 1 32 kB
-Level 2 256 kB
-Level 3 10 MB
+\begin{table}
+\begin{tabular}{l|c|c|c|c}
+  & \multicolumn{2}{c|}{E5-1620 v.~2, 3.7GHz} & \multicolumn{2}{c}{E5-2697 v.~2, 2.7GHz}  \\
+  \hline
+  L2 Cache (KB)             & \multicolumn{2}{c|}{256} &  \multicolumn{2}{c}{256}  \\
+  L3 Cache (MB)             & \multicolumn{2}{c|}{10}  &  \multicolumn{2}{c}{30}  \\
+  \hline
+  Array size ($10^6$ words) &   0.25 &  5      &  0.25  & 0.5    & 15 \\
+  STREAM, Triad (GB/s)      &  35.49 & 17.64   & 29.271 & 13.8426 & 9.6218 \\
+  \hline
+ LINPACK (GFLOPS)      & \multicolumn{2}{c|}{28.32}  & \multicolumn{2}{c}{25.41} \\
+\hline
+\end{tabular}
+\caption{The STREAM and LINPACK benchmark results for a single thread
+  running on each of the two CPUs considered here.}
+\label{TAB_stream_linpack}
+\end{table}
 
@@
 
-In applying the roofline model we follow the approach suggested
-in~\cite{para_pearls} and use the well-known STREAM~\cite{stream} and
-LINPACK~\cite{linpack} benchmarks in order to get the upper bounds on
-the memory bandwidth and FLOPS for the test hardware. Since we are
-using an Intel CPUs we used the Intel Math Kernel Library
-implementation of LINPACK.
-
 We measure the mean time to execute this region of code as 0.0020
@@
 \label{FIG_roofline_haswell}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=120mm]{nemolite2d_kernel_perf}
+  \caption{Performance of the Continuity and u-Momentum kernels as a
+    function of SIMD vectorisation and problem size. Results are for
+    v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.}
+  \label{FIG_kernel_perf}
 \end{figure}

### 2016-01-21 docs/papers/roofline_model/gocean_roofline.tex
roofline: add plot of STREAM2 results on desktop and table containing vital statistics

@@
 the memory bandwidth and FLOPS for the test hardware. Since we are
-using Intel CPUs we used the Intel Math Kernel Library
-implementation of LINPACK.
+using Intel CPUs we used the Intel Math Kernel Library implementation
+of LINPACK.
+
+Interpreting the results of the STREAM benchmark in order to get
+bandwidth figures for all levels of the cache hierarchy is not
+straightforward. We therefore used the STREAM2 benchmark which is
+intended for this purpose. In Figure~\ref{FIG_stream2_desktop} we plot
+the results obtained from the DAXPY component of the STREAM2 benchmark
+on the E5-1620 CPU. The plateaux in bandwidth for L2, L3 and main
+memory are clear but that for L1 is considerably noisier. The final
+values we have used in the roofline plots are given in
+Table~\ref{TAB_stream_linpack}.
+
+\begin{figure}
+\includegraphics{stream2_desktop}
+\caption{Results from running the STREAM2 benchmark on the E5-1620 CPU.}
+\label{FIG_stream2_desktop}
+\end{figure}
 
 \begin{table}
-\begin{tabular}{l|c|c|c|c}
-  & \multicolumn{2}{c|}{E5-1620 v.~2, 3.7GHz} & \multicolumn{2}{c}{E5-2697 v.~2, 2.7GHz}  \\
+\begin{tabular}{|l|c|c|c|c|}
 \hline
-  L2 Cache (KB)             & \multicolumn{2}{c|}{256} &  \multicolumn{2}{c}{256}  \\
-  L3 Cache (MB)             & \multicolumn{2}{c|}{10}  &  \multicolumn{2}{c}{30}  \\
+  \multicolumn{5}{|c|}{E5-1620 v.~2, 3.7GHz} \\
   \hline
-  Array size ($10^6$ words) &   0.25 &  5      &  0.25  & 0.5    & 15 \\
-  STREAM, Triad (GB/s)      &  35.49 & 17.64   & 29.271 & 13.8426 & 9.6218 \\
+LINPACK (GFLOPS) & \multicolumn{4}{c|}{28.32} \\
 \hline
- LINPACK (GFLOPS)      & \multicolumn{2}{c|}{28.32}  & \multicolumn{2}{c}{25.41} \\
+ &  L1 Cache & L2 Cache & L3 Cache & Main memory \\
+ &   32 KB   & 256 KB   & 10 MB    &  --         \\
+\hline
+STREAM2 (GB/s)   &  117  &  65.8  & 46.7 &  20.5 \\
+  \hline
+\multicolumn{5}{|c|}{E5-2697 v.~2, 2.7GHz}  \\
+\hline
+LINPACK (GFLOPS) & \multicolumn{4}{c|}{25.41} \\
+\hline
+ & L1 Cache & L2 Cache & L3 Cache & Main memory \\
+ &   32 KB  & 256 KB   & 30 MB    &  --         \\
+\hline
+STREAM2 (GB/s) &   &   &  &  \\
 \hline

### 2016-01-21 docs/papers/roofline_model/gocean_roofline.tex
roofline: update desktop roofline plot with figures obtained from stream2. Update/restructure text incl. start of Tools discussion

@@
 
+Despite the unreliability of FLOP counts, we hope that the trends in
+HWPC data as the problem size is varied will be useful. This raw data
+is given in Table~\ref{TAB_hwpc_mom_u}.
+
+\begin{table}
+\begin{tabular}{l|l|l|l|l}
+\hline
+\hline
+               & \multicolumn{4}{c}{Problem size} \\
+               & 32     &    64        &    128       &   256 \\
+\hline
+Footprint (MB) & 0.1 & 0.5          & 2.3          & 9.2 \\
+\hline
+Runtime (RDTSC) [s]   & 0.144868 &  0.4485108   &   1.697142   &   6.446171  \\
+Runtime unhalted [s]  & 1.199456e-01 & 4.475135e-01 & 1.750448e+00 &   6.734944e+00 \\
+CPI                   & 6.431867e-01 & 6.423220e-01 & 6.503796e-01 &   6.348858e-01 \\
+\hline
+MFLOPS           & 3.112780e+03 & 3.877720e+03  & 4.163209e+03 & 4.304728e+03 \\
+AVX MFLOPS       & 2.840625e+03 & 3.660086e+03  & 3.993397e+03 & 4.164647e+03 \\
+Packed MUOPS/s   & 7.423377e+02 & 9.584935e+02  & 1.046781e+03 & 1.092280e+03 \\
+Scalar MUOPS/s   & 2.077927e+02 & 1.306897e+02  & 7.294811e+01 & 3.784506e+01 \\
+\hline
+L1 DTLB load misses          &      2007     &      17015   &    1317912   &    6748712   \\
+L1 DTLB load miss rate       &  2.949686e-06 &  6.607264e-06& 1.326630e-04 & 1.723565e-04 \\
+L1 DTLB load miss durn [Cyc] &  3.970154e+01 &  3.080153e+01& 2.101039e+01 & 2.013584e+01 \\
+L1 DTLB store misses         &       77      &       16     &     87257    &    401036    \\
+L1 DTLB store miss rate      &  1.131668e-07 &  6.213119e-09& 8.783423e-06 & 1.024213e-05 \\
+L1 DTLB store miss durn [Cyc]&  5.633766e+01 &       29     & 2.151055e+01 & 2.669793e+01 \\
+\hline
+L2D load bandwidth [MB/s]  &  3.811099e+03 &  5.030691e+03 & 1.080936e+04 & 1.446214e+04 \\
+L2D load data volume [GB]  &   0.541220288 &   2.34655488  &  18.1215488  &   92.177536  \\
+L2D evict bandwidth [MB/s] &  3.131102e+02 &  3.094459e+02 & 4.501487e+02 & 5.065637e+02 \\
+L2D evict data volume [GB] &   0.04446528  &   0.144340352 &  0.75465984  &  3.22869248  \\
+L2 bandwidth [MB/s]        &  4.600834e+03 &  5.493404e+03 & 1.130282e+04 & 1.498363e+04 \\
+L2 data volume [GB]        &   0.653371776 &   2.562386304 & 18.948807808 & 95.501414592 \\
+\hline
+L3 load bandwidth [MB/s]  &  2.434255e+03 & 4.626177e+03 & 4.511686e+03 & 4.607617e+03 \\
+L3 load data volume [GB]  &   0.371968512 &  2.07489024  &   7.6168768  &   29.692832  \\
+L3 evict bandwidth [MB/s] &  5.652167e+02 & 4.828169e+02 & 2.771148e+02 & 2.722823e+02 \\
+L3 evict data volume [GB] &   0.086368448 &  0.216548608 &  0.467840512 &  1.75466688  \\
+L3 bandwidth [MB/s]       &  2.999472e+03 & 5.108994e+03 & 4.788801e+03 & 4.879899e+03 \\
+L3 data volume [GB]       &   0.45833696  &  2.291438848 &  8.084717312 &  31.44749888 \\
+\hline
+\hline
+\end{tabular}
+\caption{Hardware performance counter data obtained using the likwid
+  tool for the u-Momentum kernel on the Intel E5-1620 v2 processor.}
+\label{TAB_hwpc_mom_u}
+\end{table}
+
 \section{Performance Analysis with the Roofline Model}
@@
 \caption{Details of the two nemolite2d kernels investigated with the
-  roofline model.}
+  roofline model. $n$ is the problem size -- the length of one side of
+  the square domain.}
 \label{TAB_kernel_details}
@@
 
-\begin{table}
-\begin{tabular}{l|l|l|l|l}
-\hline
-\hline
-               & \multicolumn{4}{c}{Problem size} \\
-               & 32     &    64        &    128       &   256 \\
-\hline
-Footprint (MB) & 0.1 & 0.5          & 2.3          & 9.2 \\
-\hline
-Runtime (RDTSC) [s]   & 0.144868 &  0.4485108   &   1.697142   &   6.446171  \\
-Runtime unhalted [s]  & 1.199456e-01 & 4.475135e-01 & 1.750448e+00 &   6.734944e+00 \\
-CPI                   & 6.431867e-01 & 6.423220e-01 & 6.503796e-01 &   6.348858e-01 \\
-\hline
-MFLOPS           & 3.112780e+03 & 3.877720e+03  & 4.163209e+03 & 4.304728e+03 \\
-AVX MFLOPS       & 2.840625e+03 & 3.660086e+03  & 3.993397e+03 & 4.164647e+03 \\
-Packed MUOPS/s   & 7.423377e+02 & 9.584935e+02  & 1.046781e+03 & 1.092280e+03 \\
-Scalar MUOPS/s   & 2.077927e+02 & 1.306897e+02  & 7.294811e+01 & 3.784506e+01 \\
-\hline
-L1 DTLB load misses          &      2007     &      17015   &    1317912   &    6748712   \\
-L1 DTLB load miss rate       &  2.949686e-06 &  6.607264e-06& 1.326630e-04 & 1.723565e-04 \\
-L1 DTLB load miss durn [Cyc] &  3.970154e+01 &  3.080153e+01& 2.101039e+01 & 2.013584e+01 \\
-L1 DTLB store misses         &       77      &       16     &     87257    &    401036    \\
-L1 DTLB store miss rate      &  1.131668e-07 &  6.213119e-09& 8.783423e-06 & 1.024213e-05 \\
-L1 DTLB store miss durn [Cyc]&  5.633766e+01 &       29     & 2.151055e+01 & 2.669793e+01 \\
-\hline
-L2D load bandwidth [MB/s]  &  3.811099e+03 &  5.030691e+03 & 1.080936e+04 & 1.446214e+04 \\
-L2D load data volume [GB]  &   0.541220288 &   2.34655488  &  18.1215488  &   92.177536  \\
-L2D evict bandwidth [MB/s] &  3.131102e+02 &  3.094459e+02 & 4.501487e+02 & 5.065637e+02 \\
-L2D evict data volume [GB] &   0.04446528  &   0.144340352 &  0.75465984  &  3.22869248  \\
-L2 bandwidth [MB/s]        &  4.600834e+03 &  5.493404e+03 & 1.130282e+04 & 1.498363e+04 \\
-L2 data volume [GB]        &   0.653371776 &   2.562386304 & 18.948807808 & 95.501414592 \\
-\hline
-L3 load bandwidth [MB/s]  &  2.434255e+03 & 4.626177e+03 & 4.511686e+03 & 4.607617e+03 \\
-L3 load data volume [GB]  &   0.371968512 &  2.07489024  &   7.6168768  &   29.692832  \\
-L3 evict bandwidth [MB/s] &  5.652167e+02 & 4.828169e+02 & 2.771148e+02 & 2.722823e+02 \\
-L3 evict data volume [GB] &   0.086368448 &  0.216548608 &  0.467840512 &  1.75466688  \\
-L3 bandwidth [MB/s]       &  2.999472e+03 & 5.108994e+03 & 4.788801e+03 & 4.879899e+03 \\
-L3 data volume [GB]       &   0.45833696  &  2.291438848 &  8.084717312 &  31.44749888 \\
-\hline
-\hline
-\end{tabular}
-\caption{Hardware performance counter data for the u-Momentum kernel on the Intel E5-1620 v2 processor.}
-\label{TAB_HWPC_MOM_U}
-\end{table}
+\section{Tools}
+
+This work has deliberately focused on the application of a relatively
+simple performance model to two very simple, stencil based simulation
+codes.  Despite this, getting the data necessary for the construction
+of the roofline model has not been straightforward. Once the roofline
+model had been constructed, it was also difficult to establish why the
+performance of a section of code was apparently poor.
+
+The reasons for this include:
+\begin{itemize}
+\item HWPC data on its own doesn't give much insight into performance issues;
+\item FLOPS data produced by likwid is unreliable on Ivy Bridge processors;
+\item Persuading any tool to give HWPC data can be difficult.
+likwid is relatively straightforward but root access was required on a Redhat-based linux distribution.
+Thus far we haven't succeeded in getting this data from Intel's VTune.
+\item The Intel compiler tends to report ``Loop was vectorised'' but
+  it is hard to judge how efficient this vectorisation is. For
+  instance we have found occasions where backing-off from AVX to SSE
+  instructions actually gave better performing code. Ways of establishing the quality of the code produced by the compiler are missing;
+\end{itemize}
+
+It feels as though there is a need for a tool that addresses some of
+these issues.
+
+\section{Conclusions}
 

### 2016-01-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: added stream2 results for Archer

@@
 straightforward. We therefore used the STREAM2 benchmark which is
-intended for this purpose. In Figure~\ref{FIG_stream2_desktop} we plot
+intended for this purpose. In Figure~\ref{FIG_stream2} we plot
 the results obtained from the DAXPY component of the STREAM2 benchmark
@@
 \includegraphics{stream2_desktop}
-\caption{Results from running the STREAM2 benchmark on the E5-1620 CPU.}
-\label{FIG_stream2_desktop}
+\includegraphics{stream2_archer}
+\caption{Results from running the STREAM2 benchmark on the E5-1620
+  (top) and E5-2697 (bottom) CPUs.}
+\label{FIG_stream2}
 \end{figure}
@@
 \hline
-STREAM2 (GB/s) &   &   &  &  \\
+STREAM2 (GB/s) & 108 & 54.0 & 37.7  &  17.6 \\
 \hline
@@
 The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
-Figures~\ref{FIG_flops_dp_stream_test}
-and~\ref{FIG_flops_dp_triad_test}. It appears that once the memory
+Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
 footprint spills out of L1 cache (which all of our nemolite2d
@@
 \includegraphics[width=120mm]{FLOPS_DP_stream_ivybridgeEP.pdf}
-\caption{Results of the likwid accuracy test for the FLOPS\_DP group
-  and the stream benchmark on an Intel Xeon E5-1620 v2.}
-\label{FIG_flops_dp_stream_test}
-\end{figure}
-
-\begin{figure}
-\centering
 \includegraphics[width=120mm]{FLOPS_DP_triad_ivybridgeEP.pdf}
 \caption{Results of the likwid accuracy test for the FLOPS\_DP group
-  and the triad benchmark on an Intel Xeon E5-1620 v2.}
-\label{FIG_flops_dp_triad_test}
+  and the stream (top) and triad (bottom) benchmark on an Intel Xeon
+  E5-1620 v2.}
+\label{FIG_flops_dp_test}
 \end{figure}
@@
 We can also count the number of floating-point arithmetic operations
-in this code fragment which gives us 12 multiplications and 12
-additions. $AI$ for this code fragment is then $24*7/640 = 0.26$
+(FLOPs) in this code fragment which gives us 12 multiplications and 12
+additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
 FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
@@
 sin(). There are 31 distinct memory references. This gives an $AI$ of
-approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte.
+approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we assume
+the call to {\tt sin()} costs 13 FLOPs). The results of this analysis
+are summarised in Table~\ref{TAB_kernel_details}.
 
 \begin{table}
-  \begin{tabular}{l|c|c|c}
-                           & Shallow & \multicolumn{2}{c}{nemolite2d} \\
-                           & 1st Loop nest & Continuity & u-Momentum \\
+\begin{tabular}{|l|c|c|c|}
+\hline
+Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
+\hline
+Kernel                     & 1st loop nest & Continuity & u-Momentum \\
 \hline                                
@@
 $AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
-Working set size (bytes)   & $56n^2$ &  $64n^2$   &            \\
+Working set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
+\hline
 \end{tabular}
-\caption{Details of the two nemolite2d kernels investigated with the
-  roofline model. $n$ is the problem size -- the length of one side of
-  the square domain.}
+\caption{Details of the kernels investigated with the roofline
+  model. $n$ is the problem size -- the length of one side of the
+  square domain.}
 \label{TAB_kernel_details}
@@
 \includegraphics[width=120mm]{roofline_archer}
-\caption{Comparison of the performance achieved by nemolite2d and Shallow on Archer (Intel Ivybridge)}
+\caption{Comparison of the performance achieved by nemolite2d and
+  Shallow on Archer (Intel Ivybridge)}
 \label{FIG_roofline_archer}
@@
 \includegraphics[width=120mm]{roofline_haswell}
-\caption{Comparison of the performance achieved by nemolite2d and Shallow on Intel Haswell}
+\caption{Comparison of the performance achieved by nemolite2d and
+  Shallow on Intel Haswell}
 \label{FIG_roofline_haswell}
@@
 \item FLOPS data produced by likwid is unreliable on Ivy Bridge processors;
-\item Persuading any tool to give HWPC data can be difficult.
-likwid is relatively straightforward but root access was required on a Redhat-based linux distribution.
-Thus far we haven't succeeded in getting this data from Intel's VTune.
+\item Persuading any tool to give HWPC data can be difficult. likwid
+  is relatively straightforward but root access was required on a
+  Redhat-based linux distribution. Thus far we haven't succeeded in
+  getting this data from Intel's VTune.
 \item The Intel compiler tends to report ``Loop was vectorised'' but
@@
   instance we have found occasions where backing-off from AVX to SSE
-  instructions actually gave better performing code. Ways of establishing the quality of the code produced by the compiler are missing;
+  instructions actually gave better performing code. Ways of
+  establishing the quality of the code produced by the compiler are
+  missing;
 \end{itemize}
@@
 It feels as though there is a need for a tool that addresses some of
-these issues.
+these issues. It could be that Intel's Vectorization Advisor will be
+useful and this requires investigation.
 

### 2016-01-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: fleshing out text now we have results

@@
 otherwise) of two shallow-water benchmark codes that we have used in
-the GOcean project.
+the GOcean project. The first of these codes, Shallow, was originally
+written by Paul Swarztrauber over 30 years ago and has been optimised
+by various people over that time. In contrast, nemolite2d is a new,
+un-optimised code that has been extracted from NEMO.
 
 Although we have investigated how the performance of the \psykal
-version of nemolite2d compares with that or the original, we have not
+version of nemolite2d compares with that of the original, we have not
 addressed how efficient the original actually is. In order to do so we
@@
 
+In this report we investigate the application of the Roofline Model to
+these two, serial, shallow-water codes.
+
 \section{Hardware Details}
 
-In applying the roofline model we follow the approach suggested
-in~\cite{para_pearls} and use the well-known STREAM~\cite{stream} and
-LINPACK~\cite{linpack} benchmarks in order to get the upper bounds on
-the memory bandwidth and FLOPS for the test hardware. Since we are
-using Intel CPUs we used the Intel Math Kernel Library implementation
-of LINPACK.
-
-Interpreting the results of the STREAM benchmark in order to get
-bandwidth figures for all levels of the cache hierarchy is not
-straightforward. We therefore used the STREAM2 benchmark which is
-intended for this purpose. In Figure~\ref{FIG_stream2} we plot
-the results obtained from the DAXPY component of the STREAM2 benchmark
-on the E5-1620 CPU. The plateaux in bandwidth for L2, L3 and main
-memory are clear but that for L1 is considerably noisier. The final
-values we have used in the roofline plots are given in
-Table~\ref{TAB_stream_linpack}.
-
-\begin{figure}
-\includegraphics{stream2_desktop}
-\includegraphics{stream2_archer}
-\caption{Results from running the STREAM2 benchmark on the E5-1620
-  (top) and E5-2697 (bottom) CPUs.}
-\label{FIG_stream2}
-\end{figure}
+We have performed this investigation on two different Intel CPUs; an
+E5-1620 v.2 and an E5-2697 v.2. Both of these use the Intel Ivy Bridge
+microarchitecture and have Ivy Bridge-EP cores. More details are given
+in Table~\ref{TAB_stream_linpack}.
 
@@
 \end{tabular}
-\caption{The STREAM and LINPACK benchmark results for a single thread
-  running on each of the two CPUs considered here.}
+\caption{The STREAM2 (DAXPY) and LINPACK benchmark results for a
+  single thread running on each of the two CPUs considered here.}
 \label{TAB_stream_linpack}
@@
 
+In applying the roofline model we follow the approach suggested
+in~\cite{para_pearls} and use the well-known STREAM~\cite{stream} and
+LINPACK~\cite{linpack} benchmarks in order to get the upper bounds on
+the memory bandwidth and FLOPS for the test hardware. Since we are
+using Intel CPUs we used the Intel Math Kernel Library implementation
+of LINPACK.
+
+Interpreting the results of the STREAM benchmark in order to get
+bandwidth figures for all levels of the cache hierarchy is not
+straightforward. We therefore used the STREAM2 benchmark which is
+intended for this purpose. In Figure~\ref{FIG_stream2} we plot the
+results obtained from the DAXPY component of the STREAM2 benchmark on
+both of the CPUs examined here. The plateaux in bandwidth for L2, L3
+and main memory are clear but that for L1 is considerably noisier. The
+final values we have used in the roofline plots are given in
+Table~\ref{TAB_stream_linpack}.
+
+\begin{figure}
+\includegraphics{stream2_desktop}
+\includegraphics{stream2_archer}
+\caption{Results from running the STREAM2 benchmark on the E5-1620
+  (top) and E5-2697 (bottom) CPUs.}
+\label{FIG_stream2}
+\end{figure}
+
 \section{Hardware Performance Counters}
 
+Initially, we hoped to use Hardware Performance Counters (HWPCs) in
+order to get an accurate measure of the FLOPS and memory traffic
+associated with specific kernels in each of the models.
+
 The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
-access to Hardware Performance Counters (HWPCs) on Intel and AMD CPUs.
+access to HWPCs on Intel and AMD CPUs.
 However, there are issues with the accuracy of these counters,
-especially for FLOP-related measurements on Ivybridge architectures. A
+especially for FLOP-related measurements on Ivy Bridge architectures. A
 tool is included in the likwid distribution that tests the accuracy of
@@
 is not and in particular, includes a call to the {\tt sin()} function.
-In order to estimate the cost of this function in FLOPs the following
+In order to estimate the cost of this function in FLOPs, the following
 code was compiled (with the Intel compiler and -O1) and run with
@@
 \end{equation}
-We calculated this quantity manually by examining the
-source code and counting the number of memory references and
-arithmetic operations that it contained. In doing this counting we
-assume that any references to adjacent array elements are fetched in a
-single cache-line and thus only count once.
+Since HWPC data proved to be unreliable, we calculated this quantity
+manually by examining the source code and counting the number of
+memory references and arithmetic operations that it contained. In
+doing this counting we assume that any references to adjacent array
+elements are fetched in a single cache-line and thus only count once.
 
@@
 time-stepping section of the Shallow code. This is a doubly-nested
-loop over {\tt i} and {\tt j} with {\tt i} innermost). The body of the
+loop over {\tt i} and {\tt j} with {\tt i} innermost. The body of the
 loop is:
@@
 and {\tt H}) and reads from ten distinct locations.  However, if we
-assume caching of reads from adjacent locations ({\it e.g.} {\tt
+assume caching of reads from locations adjacent in memory ({\it e.g.} {\tt
   P(i,j)} and {\tt P(i+1,j)}) then we have only six distinct reads.
@@
 kernel of nemolite2d then we find that there are 51 addition
-operations, 56 multiplication/division operations and one call to
-sin(). There are 31 distinct memory references. This gives an $AI$ of
-approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we assume
-the call to {\tt sin()} costs 13 FLOPs). The results of this analysis
-are summarised in Table~\ref{TAB_kernel_details}.
+operations, 56 multiplication/division operations and one call to {\tt
+  sin()}. There are 31 distinct memory references. This gives an $AI$
+of approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we
+assume the call to {\tt sin()} costs 13 FLOPs). The results of this
+analysis are summarised in Table~\ref{TAB_kernel_details}.
 
@@
 $AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
-Working set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
+Working-set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
 \hline
@@
 
-We measure the mean time to execute this region of code as 0.0020
-seconds (Intel v.16 on Haswell E5-1620 v2) for the 256 domain. This
-gives us $\frac{108*256^2}{0.002} = 3.5$ GFLOPS. In contrast,
-instrumenting this section of code with the LIKWID marker API and
-using likwid-perfctr we measure 4.7 GFLOPS.
+We measure the mean time to execute the u-Momentum kernel over the
+whole domain as 0.0020 seconds (Intel v.16 on the E5-1620) for the
+$256^2$ problem size. This gives us a performance of
+$\frac{108*256^2}{0.002} = 3.5$ GFLOPS. In contrast, instrumenting
+this section of code with the LIKWID marker API and using
+likwid-perfctr we measure 4.7 GFLOPS.
 
-For the 128 the measured time per kernel call is 0.570033E-03 s.
-$108*128^2/0.570e-3 = 3.1$ GFLOPS which is 25\% less than we measure
-directly with likwid.  We can use the value reported by likwid to get
-the FLOP count: 4.269 GFLOPS $\times 1.677 s = 7.1615217$ GFLOPs. This is for
-3000 steps of $128^2$ so one kernel call is: $7.1615217/(3000*128^2) =
-145.7$ FLOPs. This means that $AI = 145.7*7/(31*64) = 0.514$ FLOP/byte
+For the $128^2$ domain, the measured time is 0.570033E-03 s. This
+gives a performance of $108*128^2/0.570e-3 = 3.1$ GFLOPS which is 25\%
+less than we measure directly with likwid.
+
+Using these figures we can finally construct the Roofline model for
+the two CPUs and the various kernels. In
+Figure~\ref{FIG_roofline_archer} we plot the model for the E5-2697 CPU
+and show the performance of the 1st loop-nest of Shallow and the
+u-Momentum kernel of nemolite2d.
 
@@
 \caption{Comparison of the performance achieved by nemolite2d and
-  Shallow on Archer (Intel Ivybridge)}
+  Shallow on the E5-2697}
 \label{FIG_roofline_archer}
@@
 \includegraphics[width=120mm]{roofline_haswell}
-\caption{Comparison of the performance achieved by nemolite2d and
-  Shallow on Intel Haswell}
+\caption{Comparison of the performance achieved by different nemolite2d and
+kernels on the E5-1620.}
 \label{FIG_roofline_haswell}

### 2016-01-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: fix archer roofline plot to have nice bars. Add discussion of Archer results

@@
 words). However, as noted above, this code fragment is the body of a
-loop nest. Therefore, upon the next loop trip {\tt i} will have been
+loop nest. Therefore, upon the next loop trip, {\tt i} will have been
 incremented by one and the majority of the accessed values will still
@@
   sin()}. There are 31 distinct memory references. This gives an $AI$
-of approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we
+of approximately $(107+13)*7/(31*64) = 0.42$ FLOPs/byte (where we
 assume the call to {\tt sin()} costs 13 FLOPs). The results of this
@@
 We measure the mean time to execute the u-Momentum kernel over the
-whole domain as 0.0020 seconds (Intel v.16 on the E5-1620) for the
+whole domain as 0.0021 seconds (Intel v.16 on the E5-1620) for the
 $256^2$ problem size. This gives us a performance of
-$\frac{108*256^2}{0.002} = 3.5$ GFLOPS. In contrast, instrumenting
+$\frac{120\times 256^2}{0.0021} = 3.7$ GFLOPS. In contrast, instrumenting
 this section of code with the LIKWID marker API and using
 likwid-perfctr we measure 4.7 GFLOPS.
-
-For the $128^2$ domain, the measured time is 0.570033E-03 s. This
-gives a performance of $108*128^2/0.570e-3 = 3.1$ GFLOPS which is 25\%
-less than we measure directly with likwid.
+For the $128^2$ domain, the measured time is 0.556E-03 s. This gives a
+performance of $\frac{120\times 128^2}{0.556\times10^{-3}} = 3.5$
+GFLOPS which is some 25\% less than we measure directly with likwid.
 
@@
 and show the performance of the 1st loop-nest of Shallow and the
-u-Momentum kernel of nemolite2d.
+u-Momentum kernel of nemolite2d. Two things are immediately clear from
+this plot: first, the performance of the first loop nest of Shallow is
+clearly memory-bandwidth limited and second, the u-Momentum kernel of
+NEMOLite2D is no-where near approaching this limit.
 
@@
 
+ Two performance points are plotted for the Shallow kernel
+ corresponding to domains of $512^2$ and $1024^2$. Using the equation
+ for working-set size in Table~\ref{TAB_kernel_details}, these
+ correspond to 14.7 and 58.7~MB, respectively. Since the E5-2697 has
+ an L3 cache of 30~MB, the $512^2$ case should be contained in cache
+ while the $1024^2$ case will spill to main memory. This is borne out
+ by the plot; the $1024^2$ case is achieving a performance in-line
+ with saturating the bandwidth to main memory. The $512^2$ case is
+ approaching the L3 bandwidth limit. We attribute the shortfall to the
+ fact that the kernel is not run in isolation but as part of a
+ time-stepping loop -- other kernels requiring different arrays are
+ run on either side of it. {\bf BUT, the working-set for the
+   time-stepping loop as a whole has 13 arrays giving a size of 27~MB
+   so it should fit in L3 cache?}
+
 \begin{figure}
@@
 \includegraphics[width=120mm]{roofline_haswell}
-\caption{Comparison of the performance achieved by different nemolite2d and
-kernels on the E5-1620.}
+\caption{Comparison of the performance achieved by different
+  NEMOLite2D kernels on the E5-1620 CPU.}
 \label{FIG_roofline_haswell}
 \end{figure}
+
+In contrast to those results, the performance of the NEMOLite2D
+u-Momentum kernel shows little sensitivity to working-set size -- the
+$128^2$, $256^2$ and $512^2$ performance points in
+Figure~\ref{FIG_roofline_archer} are very close together.
 

### 2016-01-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: add beginnings of discussion of roofline results

@@
 \includegraphics[width=120mm]{roofline_archer}
-\caption{Comparison of the performance achieved by nemolite2d and
-  Shallow on the E5-2697}
+\caption{Comparison of the performance achieved by NEMOLite2D and
+  Shallow on the E5-2697 CPU.}
 \label{FIG_roofline_archer}
@@
    so it should fit in L3 cache?}
+
+In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
+obtained for the E5-1620 CPU. In order to investigate the poor
+performance of the u-Momentum kernel of NEMOLite2D, we also show the
+performance of the (simpler) Continuity kernel from that code. As we
+would expect, the performance of the u-Momentum kernel remains poor on
+this CPU. However, the Continuity kernel performs well. The
+working-set sizes for this kernel for the $128^2$ and $256^2$ domains
+are approximately 1~MB and 4.2~MB, respectively. Both of these should
+therefore fit comfortably in L3 cache. However, again, this kernel is
+one of several within the time-stepping loop of NEMOLite2D. For the
+$128^2$ domain this loop as a whole has a working-set size of
+$18\times128^2\times 8$ bytes or 2.4~MB. {\bf Again, this should all
+  fit comfortably in l3 cache so why are we nowhere near the L3
+  bandwidth limit?}
 

### 2016-01-25 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more discussion of roofline results
roofline: add note to paper re next stage

@@
 
+In contrast to those results, the performance of the NEMOLite2D
+u-Momentum kernel shows little sensitivity to working-set size -- the
+$128^2$, $256^2$ and $512^2$ performance points in
+Figure~\ref{FIG_roofline_archer} are very close together.
+
 In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
@@
   fit comfortably in l3 cache so why are we nowhere near the L3
-  bandwidth limit?}
+  bandwidth limit? Need to look at performance variation with problem size and maybe HWPCs at the same time.}
 
@@
 \end{figure}
-
-In contrast to those results, the performance of the NEMOLite2D
-u-Momentum kernel shows little sensitivity to working-set size -- the
-$128^2$, $256^2$ and $512^2$ performance points in
-Figure~\ref{FIG_roofline_archer} are very close together.
 

### 2016-01-26 docs/papers/roofline_model/gocean_roofline.tex
roofline: added more points for investigation
roofline: add hwpc plot for cont. kernel

@@
   fit comfortably in l3 cache so why are we nowhere near the L3
-  bandwidth limit? Need to look at performance variation with problem size and maybe HWPCs at the same time.}
+  bandwidth limit?
+Need to look at performance variation with problem size and maybe HWPCs at the same time.
+Is my figure for l3 bandwidth correct?
+What does loop1 of Shallow achieve on smaller problem sizes?}
 
@@
 
+In order to investigate why the performance of the Continuity kernel
+is falling well short of the maximum predicted by the roofline model
+we have experimented with the level of SIMD
+vectorisation. Figure~\ref{FIG_kernel_perf} shows the performance of
+the Continuity and u-Momentum kernels as a function of problem size.
+
 \begin{figure}
@@
   \label{FIG_kernel_perf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=120mm]{nemolite2d_cont_perf}
+  \caption{Performance of the Continuity kernel (compiled with AVX
+    instructions) and HWPC data as a function of problem size. Results
+    are for v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.}
+  \label{FIG_cont_perf}
 \end{figure}

### 2016-01-26 docs/papers/roofline_model/gocean_roofline.tex
roofline: update of stream2 results on desktop

@@
 final values we have used in the roofline plots are given in
-Table~\ref{TAB_stream_linpack}.
+Table~\ref{TAB_stream_linpack}. On the E5-1620 the benchmark was
+compiled with the Intel v.16 compiler with the flags ``-O3
+-fast''. Strangely, we found that dropping ``-fast'' from the link
+stage resulted in better bandwidth to main memory. Removing ``-fast''
+altogether and just using ``-O3 -xHost'' gave identical results.
 
@@
 \caption{Results from running the STREAM2 benchmark on the E5-1620
-  (top) and E5-2697 (bottom) CPUs.}
+  (top) and E5-2697 (bottom) CPUs. Error bars indicate the maximum and
+  minimum values obtained from 200 runs performed for each array
+  size. }
 \label{FIG_stream2}
 \end{figure}
+
+The bandwidths obtained on the E5-1620 are significantly greater than
+those on the E5-2697.  {\bf need to check by re-running on Archer with
+  200 repeats and new min/max results}.
 

### 2016-01-27 docs/papers/roofline_model/gocean_roofline.tex
roofline: add data and plots for investigation of daxpy kernels
roofline: update text and labelling of daxpy roofline

--- 2016-01-26 docs/papers/roofline_model/gocean_roofline.tex
@@
 results obtained from the DAXPY component of the STREAM2 benchmark on
-both of the CPUs examined here. The plateaux in bandwidth for L2, L3
-and main memory are clear but that for L1 is considerably noisier. The
-final values we have used in the roofline plots are given in
+the E5-2697 CPU (using v.~14 {\bf (check this)} of the Intel
+compiler. The plateaux in bandwidth for L2, L3 and main memory are
+clear but that for L1 is considerably noisier. The final values we
+have used in the roofline plots are given in
 Table~\ref{TAB_stream_linpack}.
@@
 \begin{figure}
-\includegraphics{stream2_desktop}
-\includegraphics{stream2_archer}
-\caption{Results from running the STREAM2 benchmark on the E5-1620
-  (top) and E5-2697 (bottom) CPUs.}
+\includegraphics[width=120mm]{stream2_archer}
+\caption{Results from running the STREAM2 benchmark on the E5-2697
+  CPUs. Error bars indicate the maximum and minimum values obtained
+  from 200 repeats of each case.  size. }
 \label{FIG_stream2}
+\end{figure}
+
+ The equivalent results for the E5-1620
+CPU are shown in Figure~\ref{FIG_daxpy_bw}. This plot has the same
+shape as that for the E5-2697 but is shifted upwards indicating a
+considerable increase in bandwidth accross all levels of the cache
+hierarchy. {\bf need to check by re-running on Archer with 200 repeats
+  and new min/max results}. In addition, the width of the plateau
+corresponding to the L3 cache is narrower because of the smaller size
+of this cache on the E5-1620.
+
+\begin{figure}
+\includegraphics[width=120mm]{daxpy_plot}
+\caption{Bandwidth obtained on the E5-1620 CPU by the original and
+  modified versions of the daxpy kernel from the STREAM2 benchmark.}
+\label{FIG_daxpy_bw}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=120mm]{roofline_daxpy}
+\caption{Roofline plot of daxpy-based kernels on the E5-1620 CPU. Each
+  cluster of bars represents a different kernel. Bars within a cluster
+  are for problem sizes chosen to fit within the L3, L2 and L1 cache
+  (arrays of 94,427, 8271 and 1533 elements, respectively). }
+\label{FIG_daxpy_roofline}
 \end{figure}
@@
   fit comfortably in l3 cache so why are we nowhere near the L3
-  bandwidth limit?
-Need to look at performance variation with problem size and maybe HWPCs at the same time.
-Is my figure for l3 bandwidth correct?
-What does loop1 of Shallow achieve on smaller problem sizes?}
+  bandwidth limit? Need to look at performance variation with problem
+  size and maybe HWPCs at the same time. Is my figure for l3 bandwidth
+  correct? What does loop1 of Shallow achieve on smaller problem
+  sizes?}
 

### 2016-01-27 docs/papers/roofline_model/gocean_roofline.tex
roofline: add some description of the daxpy kernels.

@@
 \label{FIG_daxpy_bw}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=120mm]{roofline_daxpy}
-\caption{Roofline plot of daxpy-based kernels on the E5-1620 CPU. Each
-  cluster of bars represents a different kernel. Bars within a cluster
-  are for problem sizes chosen to fit within the L3, L2 and L1 cache
-  (arrays of 94,427, 8271 and 1533 elements, respectively). }
-\label{FIG_daxpy_roofline}
 \end{figure}
@@
 
+As well as examining the performance of the Continuity kernel, we have
+also experimented with altering the DAXPY kernel of the STREAM2
+benchmark so as to vary its $AI$. We did this by adding further
+arithmetic operations but making use of the same array
+elements. We used four different kernels:
+\begin{enumerate}
+\item {\tt b(i) = b(i) + scalar*a(i)}
+\item {\tt b(i) = b(i) + scalar*a(i)+ a(i)*b(i)}
+\item {\tt b(i) = b(i) + scalar*a(i)+ a(i)*b(i)*b(i)}
+\item {\tt b(i) = b(i) + scalar*a(i)+ a(i)*b(i)*b(i)*b(i)}
+\end{enumerate}
+Figure~\ref{FIG_daxpy_roofline} shows the performance of these kernels
+on a roofline plot for the E5-1620 CPU. They generally reach L2 and L3
+bandwidth apart from at the lowest AI. Continuity kernel has AI of
+{\bf xxx more discussion here}
+
+\begin{figure}
+\includegraphics[width=120mm]{roofline_daxpy}
+\caption{Roofline plot of daxpy-based kernels on the E5-1620 CPU. Each
+  cluster of bars represents a different kernel. Bars within a cluster
+  are for problem sizes chosen to fit within the L3, L2 and L1 cache
+  (arrays of 94,427, 8271 and 1533 elements, respectively). }
+\label{FIG_daxpy_roofline}
+\end{figure}
+
 \section{Tools}

### 2016-01-28 docs/papers/roofline_model/gocean_roofline.tex
roofline: update stream2 results from Archer. Update bandwidth values used in roofline plots to match those in table in text

@@
 \hline
-STREAM2 (GB/s)   &  117  &  65.8  & 46.7 &  20.5 \\
+STREAM2 (GB/s)   &  160  &  61.0  & 46.0 &  20.5 \\
   \hline
@@
 \hline
-STREAM2 (GB/s) & 108 & 54.0 & 37.7  &  17.6 \\
+STREAM2 (GB/s) & 132      & 59.0     & 40.0     &  20.2 \\
 \hline
@@
 
- The equivalent results for the E5-1620
-CPU are shown in Figure~\ref{FIG_daxpy_bw}. This plot has the same
-shape as that for the E5-2697 but is shifted upwards indicating a
-considerable increase in bandwidth accross all levels of the cache
-hierarchy. {\bf need to check by re-running on Archer with 200 repeats
-  and new min/max results}. In addition, the width of the plateau
-corresponding to the L3 cache is narrower because of the smaller size
-of this cache on the E5-1620.
+ The equivalent results for the E5-1620 CPU are shown in
+ Figure~\ref{FIG_daxpy_bw}. This plot has the same shape as that for
+ the E5-2697 but is shifted upwards indicating an increase in
+ bandwidth accross all levels of the cache hierarchy, but particularly
+ for L3 and L1. In addition, the width of the plateau corresponding to
+ the L3 cache is narrower because of the smaller size of this cache on
+ the E5-1620.
 
@@
 
-Despite the unreliability of FLOP counts, we hope that the trends in
-HWPC data as the problem size is varied will be useful. This raw data
-is given in Table~\ref{TAB_hwpc_mom_u}.
+Despite the unreliability of the absolute values of the FLOP counts,
+we hope that the trends in HWPC data as the problem size is varied
+will be useful. This raw data is given in Table~\ref{TAB_hwpc_mom_u}.
 
@@
 whole domain as 0.0021 seconds (Intel v.16 on the E5-1620) for the
-$256^2$ problem size. This gives us a performance of
-$\frac{120\times 256^2}{0.0021} = 3.7$ GFLOPS. In contrast, instrumenting
-this section of code with the LIKWID marker API and using
-likwid-perfctr we measure 4.7 GFLOPS.
-For the $128^2$ domain, the measured time is 0.556E-03 s. This gives a
-performance of $\frac{120\times 128^2}{0.556\times10^{-3}} = 3.5$
-GFLOPS which is some 25\% less than we measure directly with likwid.
+$256^2$ problem size. This gives us a performance of $\frac{120\times
+  256^2}{0.0021} = 3.7$ GFLOPS. (In contrast, instrumenting this
+section of code with the LIKWID marker API and using likwid-perfctr we
+measure 4.7 GFLOPS.)  For the $128^2$ domain, the measured time is
+0.556E-03 s. This gives a performance of $\frac{120\times
+  128^2}{0.556\times10^{-3}} = 3.5$ GFLOPS which again is some 25\% less
+than we measure directly with likwid.
 
@@
 with saturating the bandwidth to main memory. The $512^2$ case is
- approaching the L3 bandwidth limit. We attribute the shortfall to the
- fact that the kernel is not run in isolation but as part of a
- time-stepping loop -- other kernels requiring different arrays are
- run on either side of it. {\bf BUT, the working-set for the
-   time-stepping loop as a whole has 13 arrays giving a size of 27~MB
-   so it should fit in L3 cache?}
+approaching the L3 bandwidth limit. We investigate the reasons for
+the fact that it does not achieve the L3 bandwidth limit below.
 
-In contrast to those results, the performance of the NEMOLite2D
-u-Momentum kernel shows little sensitivity to working-set size -- the
-$128^2$, $256^2$ and $512^2$ performance points in
-Figure~\ref{FIG_roofline_archer} are very close together.
+In contrast to the results obtained with the Shallow kernel, the
+performance of the NEMOLite2D u-Momentum kernel shows little
+sensitivity to working-set size -- the $128^2$, $256^2$ and $512^2$
+performance points in Figure~\ref{FIG_roofline_archer} are very close
+together.
 
@@
 would expect, the performance of the u-Momentum kernel remains poor on
-this CPU. However, the Continuity kernel performs well. The
-working-set sizes for this kernel for the $128^2$ and $256^2$ domains
-are approximately 1~MB and 4.2~MB, respectively. Both of these should
-therefore fit comfortably in L3 cache. However, again, this kernel is
-one of several within the time-stepping loop of NEMOLite2D. For the
-$128^2$ domain this loop as a whole has a working-set size of
-$18\times128^2\times 8$ bytes or 2.4~MB. {\bf Again, this should all
-  fit comfortably in l3 cache so why are we nowhere near the L3
-  bandwidth limit? Need to look at performance variation with problem
-  size and maybe HWPCs at the same time. Is my figure for l3 bandwidth
-  correct? What does loop1 of Shallow achieve on smaller problem
-  sizes?}
+this CPU. For the $512^2$ domain the Continuity kernel performs well,
+reaching the limit imposed by the bandwidth to main memory. However,
+the working-set sizes for this kernel for the $128^2$ and $256^2$
+domains are approximately 1~MB and 4.2~MB, respectively and therefore
+both of these should fit comfortably within L3 cache. Despite this,
+the performance achieved with these domain sizes falls short of the L3
+bandwidth limit.  Although this kernel is one of several within the
+time-stepping loop of NEMOLite2D, the working-set size of this loop
+for the $128^2$ domain is just 2.4~MB which again is well within the
+capacity of L3.  . Need to look at performance variation with problem
+size and maybe HWPCs at the same time.  What does loop1 of Shallow
+achieve on smaller problem sizes?
 

### 2016-01-28 docs/papers/roofline_model/gocean_roofline.tex
roofline: add set-associativity data to table giving cache specs

@@
 &  L1 Cache & L2 Cache & L3 Cache & Main memory \\
-                 &   32 KB   & 256 KB   & 10 MB    &  --         \\
+$n$-way set associative & 8    & 8   & 10 & -- \\
+Size                  &   32 KB/core   & 256 KB/core   & 10 MB    &  --         \\
 \hline
@@
                & L1 Cache & L2 Cache & L3 Cache & Main memory \\
-               &   32 KB  & 256 KB   & 30 MB    &  --         \\
+$n$-way set associative & 8    & 8   & 20 & -- \\
+Size           &   32 KB/core  & 256 KB/core   & 30 MB    &  --         \\
 \hline

### 2016-01-29 docs/papers/roofline_model/gocean_roofline.tex
roofline: update kernel-vect-perf plot and add discussion

@@
 for the $128^2$ domain is just 2.4~MB which again is well within the
-capacity of L3.  . Need to look at performance variation with problem
-size and maybe HWPCs at the same time.  What does loop1 of Shallow
-achieve on smaller problem sizes?
+capacity of L3.
 
@@
 vectorisation. Figure~\ref{FIG_kernel_perf} shows the performance of
-the Continuity and u-Momentum kernels as a function of problem size.
+the Continuity and u-Momentum kernels when compiled without
+vectorisation (``-no-vec''), with SSE4 (Streaming SIMD Extensions)
+vectorisation (``-axSSE4.2'') and with AVX (``-xHost''). SSE supports
+vector lengths of 128 bits (two double-precision words) and AVX 512
+bits (four double-precision words). For the 64, 128 and 256 problem
+sizes, changing from no vectorisation to SSE increases the performance
+of the Continuity kernel by up to roughly 60\%; some way short of the
+ideal speed-up of 100\% that would be obtained with perfect
+SSE vectorisation of a compute-bound kernel. Unsurprisingly then, moving
+from SSE to AVX vectors does not give any performance improvement and
+in fact, is slightly detrimental for most of the problem sizes.
 
@@
 \end{figure}
+
+Peak absolute performance is obtained for the 256 problem size with
+SSE instructions. The 512 problem size spills from L3 cache and
+becomes bound by the bandwidth to main memory: SIMD vectorisation
+gives no performance benefit. The fact that best performance is
+obtained when the problem size fits within L3 indicates that there are
+other overheads that dominate when the problem size is small enough to
+fit within L2. This is possibly because the inner, SIMD-vectorised
+loop then only has a trip-count of 35. (This is not 32 because of the way
+the simulation domain is set-up; it is the {\em internal} region of
+the domain that has dimension $32^2$.)
+
+Turning now to the u-Momentum kernel (lower plot in
+Figure~\ref{FIG_kernel_perf}), it is striking that the kernel
+performance is almost independent of problem size. The reason for this
+becomes clear when we edit the kernel to remove the many IF statements
+(present because of the application of boundary conditions). The blue
+symbols mark the resulting performance which is significantly
+increased and again reaches a maximum at the 256 problem size.  It
+should be noted that the compiler reported that this loop had been
+vectorised, even before the IF statements were removed. Clearly the
+vectorisation is more efficient once the code is simplified. However,
+even with the up-lift in performance given by removing all the IF
+statements, the performance of this kernel is still far short of even
+the main-memory bandwidth limit, despite the fact that the $256^2$
+problem fits within L3 cache.
 

### 2016-01-29 docs/papers/roofline_model/gocean_roofline.tex
roofline: update continuity perf data for SSE case. Update discussion

@@
 the simulation domain is set-up; it is the {\em internal} region of
-the domain that has dimension $32^2$.)
+the domain that has dimension $32^2$.) {\bf Remainder and xxx loops.}
 
@@
 
+In order to help further understand the performance of the Continuity
+kernel in Figure~\ref{FIG_cont_perf} we plot the performance of the
+Continuity kernel alongside HWPC-derived metrics (obtained using the
+likwid tool). We use the version compiled with SSE instructions since
+that gave the best performance. This plot confirms that the $32^2$
+problem size is dominated by bandwidth to L2. Sizes $64^2$, $128^2$
+and $256^2$ are dominated by bandwidth to L2 and L3 and $512^2$ makes
+heavy use of main memory.
+
 \begin{figure}
@@
   \includegraphics[width=120mm]{nemolite2d_cont_perf}
-  \caption{Performance of the Continuity kernel (compiled with AVX
-    instructions) and HWPC data as a function of problem size. Results
-    are for v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.}
+  \caption{Performance of the Continuity kernel (bars -- right y-axis)
+    and HWPC data (open circles -- left y-axis) as a function of
+    problem size. Results are for v.16 of the Intel compiler on a Xeon
+    E5-1620 v.2 CPU.}
   \label{FIG_cont_perf}

### 2016-02-03 docs/papers/roofline_model/gocean_roofline.tex
roofline: change to grace plot of perf and bw of daxpy kernels. Add non-aligned daxpy result to daxpy roofline.
roofline: add daxpy perf. for main-memory problem size. Add discussion of daxpy roofline plot

@@
 \begin{figure}
-\includegraphics[width=120mm]{daxpy_plot}
+\includegraphics[width=120mm]{daxpy_bw_perf}
 \caption{Bandwidth obtained on the E5-1620 CPU by the original and
@@
 the simulation domain is set-up; it is the {\em internal} region of
-the domain that has dimension $32^2$.) {\bf Remainder and xxx loops.}
+the domain that has dimension $32^2$.) {\bf Remainder and peel loops.}
 
@@
 arithmetic operations but making use of the same array
-elements. We used four different kernels:
+elements. We used five different kernels:
 \begin{enumerate}
-\item {\tt b(i) = b(i) + scalar*a(i)}
-\item {\tt b(i) = b(i) + scalar*a(i)+ a(i)*b(i)}
-\item {\tt b(i) = b(i) + scalar*a(i)+ a(i)*b(i)*b(i)}
-\item {\tt b(i) = b(i) + scalar*a(i)+ a(i)*b(i)*b(i)*b(i)}
+\item {\tt y(i) = y(i) + a*x(i)}, AI = 0.125
+\item {\tt y(i) = y(i) + a*x(i)+ x(i)*y(i)}, AI = 0.167
+\item {\tt y(i) = y(i) + a*x(i)+ x(i)*y(i)*y(i)}, AI = 0.208
+\item {\tt y(i) = y(i) + a*x(i)+ x(i)*y(i)*y(i)*y(i)}, AI = 0.25
+\item {\tt y(i) = y(i) + a*x(i)+ x(i-1)*y(i)*y(i)*y(i)}, AI = 0.25
 \end{enumerate}
-Figure~\ref{FIG_daxpy_roofline} shows the performance of these kernels
-on a roofline plot for the E5-1620 CPU. They generally reach L2 and L3
-bandwidth apart from at the lowest AI. Continuity kernel has AI of
-{\bf xxx more discussion here}
+The fifth of these introduces an un-aligned array access ({\tt
+  x(i-1)}) so as to better represent real-life finite difference
+kernels.  Figure~\ref{FIG_daxpy_roofline} shows the performance of
+these kernels on a roofline plot for the E5-1620 CPU. The performance
+of each kernel is plotted for four problem sizes, chosen so as to fit
+within L1, L2, L3 cache and main memory. The performance of kernels 2,
+3 and 4 is entirely consistent with the roofline model for problem
+sizes that fit within the L3 and L3 caches (red and black bars). Their
+performance for the smallest problem size doesn't quite reach the L1
+bandwidth limit but does appear to track the roofline.  Kernel 1 has
+the lowest AI of the five kernels and this appears to limit its
+performance; at no point does the performance reach the limit imposed
+by the bandwidth to the relevant level of the memory hierarchy. {\bf need to expore the reason for this}
+
+The performance of kernel 5 is plotted using patterned bars in
+Figure~\ref{FIG_daxpy_roofline}. This kernel has the same AI as kernel
+4 but has an un-aligned array access. Comparing the performance of
+these two kernels we see that the resulting reduction in the
+efficiency with which the kernel can be SIMD vectorised (because of
+the need for gather/scatter operations) harms performance,
+particularly when the cost of accessing memory is small (i.e. when the
+problem is contained in L1 or L2). Since all of the NEMOLite2D kernels
+contain un-aligned array accesses, this goes some way to accounting
+for the fact that they do not reach the theoretical limits of the
+roofline plot in Figure~\ref{FIG_roofline_haswell}.
 
@@
 \caption{Roofline plot of daxpy-based kernels on the E5-1620 CPU. Each
-  cluster of bars represents a different kernel. Bars within a cluster
-  are for problem sizes chosen to fit within the L3, L2 and L1 cache
-  (arrays of 94,427, 8271 and 1533 elements, respectively). }
+  cluster of bars represents a different kernel. The green, black, red and
+  blue bars are for problem sizes chosen to fit within main memory, L3, L2 and
+  L1 caches (arrays of 3.3M, 94K, 8.3K and 1.5K elements),
+  respectively. Bars with patterned fill are for kernel 5 (unaligned
+  access).}
 \label{FIG_daxpy_roofline}

### 2016-02-04 api_v1.0/grid_mod.f90
API: add get_tmask method

@@ -97,6 +97,8 @@ module grid_mod
 
      !> Coordinates of grid (T) points in horizontal plane
      real(wp), allocatable :: xt(:,:), yt(:,:)
+   contains
+     procedure :: get_tmask
 
   end type grid_type
 
@@ -109,6 +111,15 @@ module grid_mod
 contains
 
   !============================================
+  function get_tmask(self) result(tmask)
+    implicit none
+    class (grid_type), target, intent(in) :: self
+    integer, pointer :: tmask(:,:)
+
+    tmask => self%tmask
+
+    return
+  end function get_tmask
 
   !> Basic constructor for the grid type. Full details
   !! are fleshed-out by the grid_init() routine.

### 2016-02-04 api_v1.0/timing_mod.f90
nemolite2d: update timing to use OpenMP

@@ -54,7 +54,15 @@ MODULE timing_mod
    ! Counter register). If false then the Fortran intrinsic SYSTEM_CLOCK 
    ! is used.
    LOGICAL, PARAMETER :: use_rdtsc_timer = .FALSE.
-
+   ! Intel-specific rdtsc timer (reads the Time Stamp Counter register)
+   INTEGER, PARAMETER :: RDTSC_TIMER = 0
+   ! Use the OpenMP wtime routine (must link against OpenMP)
+   INTEGER, PARAMETER :: OMP_TIMER=1
+   ! Use the Fortran intrinsic timer (precision may be limited)
+   INTEGER, PARAMETER :: INTRINSIC_TIMER=2
+   ! Which of these timers to use by default
+   INTEGER :: base_timer = OMP_TIMER
+   
    !-------------------------------------------------------------------
    ! Publicly-accessible routines
 
@@ -80,20 +88,27 @@ MODULE timing_mod
 !$      END IF
 
       ! Initialise the timer structures
+      select case(base_timer)
 
-      IF(use_rdtsc_timer)THEN
-         iclk_rate =1
+      case(OMP_TIMER)
+         iclk_rate = 1
          iclk_max = 1
-      ELSE
-         CALL SYSTEM_CLOCK(COUNT_RATE=iclk_rate, COUNT_MAX=iclk_max)
-         WRITE (*,"('System clock, cycles/sec =',I7,', max count = ',I11)") &
-                iclk_rate, iclk_max
-      END IF
+         write (*,"('TIMING: using OpenMP omp_get_wtime()')")
+      case(RDTSC_TIMER)
+         iclk_rate = 1
+         iclk_max = 1
+         write (*,"('TIMING: using Intel Time Stamp Counter register')")
+      case(INTRINSIC_TIMER)
+         call SYSTEM_CLOCK(COUNT_RATE=iclk_rate, COUNT_MAX=iclk_max)
+         write (*,"('TIMING: using Fortran intrinsic system clock, cycles/sec =',I7, &
+              &   ', max count = ',I11)") iclk_rate, iclk_max
+      end select
 
       nThreads = 1
 !$    nThreads = omp_get_max_threads()
 
-      WRITE (*,"('Allocating timer structures for ',I3,' threads.')") nThreads
+      WRITE (*,"('TIMING: Allocating timer structures for ',I3,' threads.')") &
+           nThreads
 
[…]

### 2016-02-04 nemolite2d_psykal/time_step_mod_align.f90
nemolite2d: add align version of time_step_mod and add nemolite2d_align target to Makefile

@@ -1,0 +1,744 @@
+module time_step_mod
+  implicit none
+
+  private
+
+  public invoke_time_step
+
+contains
+
+  subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
+                              sshn_t, sshn_u, sshn_v, &
+                              hu, hv, ht, ua, va, un, vn)
+    use kind_params_mod
+    use timing_mod
+    use field_mod
+    use grid_mod
+    use model_mod,       only: rdt, cbfr, visc
+    use physical_params_mod, only: g, omega, d2r
+!    use momentum_mod,    only: momentum_v_code
+!    use momentum_mod,    only: momentum_u_code
+!    use continuity_mod,  only: continuity_code
+!    use time_update_mod, only: next_sshu_code, next_sshv_code
+    use boundary_conditions_mod
+    !use likwid
+    implicit none
+    real(wp),        intent(in)    :: istp
+    type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
+    type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
+    type(r2d_field), intent(in)    :: hu, hv, ht
+    ! Locals
+    integer :: ji, jj, jiu, jiv
+    integer :: M, N, idxt
+    ! Locals for momentum
+    REAL(wp) :: u_e, u_w, v_n, v_s
+    real(wp) :: v_nc, v_sc
+    real(wp) :: depe, depw, deps, depn
+    real(wp) :: hpg, adv, cor, vis
+    real(wp) :: dudx_e, dudx_w, dudy_s, dudy_n
+    real(wp) :: uu_e, uu_n, uu_s, uu_w
+    real(wp) :: u_ec, u_wc, vv_e, vv_n, vv_s, vv_w
+    real(wp) :: dvdx_e, dvdx_w, dvdy_n, dvdy_s
+    real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
+    ! end locals for momentum
+    ! Locals for BCs
+    real(wp) :: amp_tide, omega_tide, rtime
+
+    M  = ssha%grid%simulation_domain%xstop
+    if(mod(M, 2) .ne. 0)then
+       M = M + 1
+    end if
+!DIR$ ASSUME (MOD(M,2) .EQ. 0)
[…]

### 2016-02-04 nemolite2d_psykal/time_step_mod_align.f90
nemolite2d: un-comment alignment directives and fix makefile

@@ -85,26 +85,6 @@ contains
                                   hu%data, hv%data, &
                                   un%data, vn%data, sshn_t%grid%area_t)
 
-    !call timer_start('Continuity',idxt)
-    !call likwid_markerStartRegion('Continuity')
-
-!    do jj = ssha%internal%ystart, ssha%internal%ystop, 1
-!      do ji = ssha%internal%xstart, ssha%internal%xstop, 1
-!!$    do jj = 2, N, 1
-!!$      do ji = 2, M, 1
-!!$
-!!$         rtmp1 = (sshn_u%data(ji  ,jj ) + hu%data(ji  ,jj  ))*un%data(ji  ,jj)
-!!$         rtmp2 = (sshn_u%data(ji-1,jj ) + hu%data(ji-1,jj  ))*un%data(ji-1,jj)
-!!$         rtmp3 = (sshn_v%data(ji ,jj ) + hv%data(ji  ,jj  ))*vn%data(ji ,jj)
-!!$         rtmp4 = (sshn_v%data(ji ,jj-1) + hv%data(ji  ,jj-1))*vn%data(ji,jj-1)
-!!$
-!!$         ssha%data(ji,jj) = sshn_t%data(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
-!!$                       rdt / sshn_t%grid%area_t(ji,jj)
-!!$      end do
-!!$    end do
-
-    !call likwid_markerStopRegion('Continuity')
-    !call timer_stop(idxt)
 
 !!$    call invoke_momentum_u_arrays(sshn_t%grid%nx, sshn_t%grid%ny, M, N, &
 !!$                                  ua%data, un%data, vn%data, &
@@ -555,10 +535,10 @@ contains
     real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
     integer :: nrepeat, ic
     
-!!!DIRx$ ASSUME (MOD(NX,4) .EQ. 0)
-!!!DIRx$ ASSUME (MOD(M,2) .EQ. 0)
-!!!DIRx$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
-!!!DIRx$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
+!DIR$ ASSUME (MOD(NX,4) .EQ. 0)
+!DIR$ ASSUME (MOD(M,2) .EQ. 0)
[…]

### 2016-02-16 api_v1.0/global_parameters_mod.F90
Update Continuity kernel for AVX alignment

@@ -12,7 +12,7 @@ module global_parameters_mod
   integer, parameter, public :: NAME_LEN = 1024
 
   ! What boundary to align arrays on
-  ! 256 bit = 4 d.p. words
+  ! AVX is 256 bit = 4 d.p. words
   integer, parameter, public :: ALIGNMENT = 4
 
   ! Iteration spaces for kernels.

### 2016-02-16 nemolite2d_psykal/time_step_mod_align.f90

@@ -10,6 +10,7 @@ contains
   subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
                               sshn_t, sshn_u, sshn_v, &
                               hu, hv, ht, ua, va, un, vn)
+    use global_parameters_mod, only: ALIGNMENT
     use kind_params_mod
     use timing_mod
     use field_mod
@@ -45,10 +46,10 @@ contains
     real(wp) :: amp_tide, omega_tide, rtime
 
     M  = ssha%grid%simulation_domain%xstop
-    if(mod(M, 2) .ne. 0)then
-       M = M + 1
+    if(mod(M, ALIGNMENT) .ne. 0)then
+       M = (M/ALIGNMENT + 1)*ALIGNMENT
     end if
-!DIR$ ASSUME (MOD(M,2) .EQ. 0)
+!DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
     N  = ssha%grid%simulation_domain%ystop
 
     ! In the general case we have to reason about whether or not the
@@ -521,6 +522,7 @@ contains
   subroutine invoke_continuity_arrays(nx, ny, M, N, rdt, ssha, &
                                       sshn_t, sshn_u, sshn_v, &
                                       hu, hv, un, vn, area_t)
+    use global_parameters_mod, only: ALIGNMENT
     use kind_params_mod
     use timing_mod
     implicit none
@@ -535,24 +537,24 @@ contains
     real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
     integer :: nrepeat, ic
     
-!DIR$ ASSUME (MOD(NX,4) .EQ. 0)
-!DIR$ ASSUME (MOD(M,2) .EQ. 0)
+!DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
+!DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
 !DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
[…]

### 2016-02-17 docs/papers/roofline_model/gocean_roofline.tex
Roofline: add perf-plot of Continuity kernel for various SIMD optimisations

@@
 
-Peak absolute performance is obtained for the 256 problem size with
-SSE instructions. The 512 problem size spills from L3 cache and
+Peak absolute performance is obtained for the $256^2$ problem size
+with SSE instructions. The 512 problem size spills from L3 cache and
 becomes bound by the bandwidth to main memory: SIMD vectorisation
@@
 other overheads that dominate when the problem size is small enough to
-fit within L2. This is possibly because the inner, SIMD-vectorised
-loop then only has a trip-count of 35. (This is not 32 because of the way
-the simulation domain is set-up; it is the {\em internal} region of
-the domain that has dimension $32^2$.) {\bf Remainder and peel loops.}
+fit within L2. We checked the efficiency of the SIMD vectorised code
+by looking at the compiler diagnostic output. On the basis of this we
+added Intel-specific directives to the code in order to specify that
+the arrays were aligned on 64-byte boundaries. We also manually coded
+the 'peel loop' (iterations required to reach a 64-byte
+boundary). Having done this the compiler reported that no peel or
+remainder loops were generated and the resulting (SSE) vectorised loop
+had an estimated speed-up of 1.9. (This estimated speed-up was
+previously 1.6; in agreement with the speed-up we measured.) After
+making these changes the performance of the Continuity kernel with the
+$32^2$ domain increased from 4.22 to 4.57 GFLOPs.
 
@@
   \label{FIG_cont_perf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=120mm]{continuity_opt}
+  \caption{Performance of the Continuity kernel with optimisations
+    intended to aid SIMD vectorisation. Results are for v.16 of the
+    Intel compiler on a Xeon E5-1620 v.2 CPU.}
+  \label{FIG_cont_vec}
 \end{figure}

### 2016-02-17 api_v1.0/timing_mod.f90
Timing: add calculation of error estimate to timing module

@@ -22,6 +22,7 @@ MODULE timing_mod
 
    INTEGER :: iclk_rate ! Ticks per second of Fortran timer
    INTEGER :: iclk_max  ! Max value that Fortran timer can return
+   REAL(wp) :: clock_tick_s ! Time in seconds between clock ticks
 
    INTEGER, PARAMETER :: LABEL_LEN  = 128
    INTEGER, PARAMETER :: MAX_TIMERS = 60
@@ -93,15 +94,20 @@ MODULE timing_mod
       case(OMP_TIMER)
          iclk_rate = 1
          iclk_max = 1
+!$       clock_tick_s = omp_get_wtick()
          write (*,"('TIMING: using OpenMP omp_get_wtime()')")
+         write (*,"('TIMING: time between clock ticks =',1E13.5,' (s)')") &
+               clock_tick_s
       case(RDTSC_TIMER)
          iclk_rate = 1
          iclk_max = 1
+         clock_tick_s = 1.0d0 ! TODO work out how to get this quantity
          write (*,"('TIMING: using Intel Time Stamp Counter register')")
       case(INTRINSIC_TIMER)
          call SYSTEM_CLOCK(COUNT_RATE=iclk_rate, COUNT_MAX=iclk_max)
          write (*,"('TIMING: using Fortran intrinsic system clock, cycles/sec =',I7, &
               &   ', max count = ',I11)") iclk_rate, iclk_max
+         clock_tick_s = 1.0d0/REAL(iclk_rate)
       end select
 
       nThreads = 1
@@ -309,7 +315,7 @@ MODULE timing_mod
          WRITE(*," (' Timed using Fortran SYSTEM_CLOCK intrinsic. Units are seconds.')")
       END IF
       WRITE(*," ('-------------------------------------------------------------------')")
-      WRITE(*," ('Region',26x,'Counts      Total         Average')")
+      WRITE(*," ('Region',26x,'Counts      Total         Average      Error')")
       WRITE(*," ('-------------------------------------------------------------------')")
       DO jt = 1, nThreads, 1
          IF(jt > 1)THEN
@@ -327,9 +333,10 @@ MODULE timing_mod
             END IF
 
             ! Truncate the label to 32 chars for table-formatting purposes
-            WRITE(*,"((A),1x,I5,1x,E13.6,2x,E13.6)") &
+            WRITE(*,"((A),1x,I5,1x,E13.6,2x,E13.6,1x,E13.6)") &
[…]

### 2016-02-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: correction to Continuity AI and updates to bar plot of simd performance

@@
 FLOPs                      & 24      &    14      &   107+13   \\
-Distinct memory refs       & 10      &    10      &   31       \\
-$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
-Working-set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
+Distinct memory refs       & 10      &    12      &   31       \\
+$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
+Working-set size (bytes)   & $56n^2$ &  $72n^2$   &   $136n^2$ \\
 \hline
@@
 the working-set sizes for this kernel for the $128^2$ and $256^2$
-domains are approximately 1~MB and 4.2~MB, respectively and therefore
+domains are approximately 1.1~MB and 4.7~MB, respectively and therefore
 both of these should fit comfortably within L3 cache. Despite this,
@@
 bandwidth limit.  Although this kernel is one of several within the
-time-stepping loop of NEMOLite2D, the working-set size of this loop
+time-stepping loop of NEMOLite2D, the working-set size of the latter
 for the $128^2$ domain is just 2.4~MB which again is well within the
@@
 we have experimented with the level of SIMD
-vectorisation. Figure~\ref{FIG_kernel_perf} shows the performance of
+vectorisation. Figure~\ref{FIG_cont_vec} shows the performance of the
+Continuity kernel compiled with differing levels of SIMD vectorisation
+and after code modifications intended to improve vectorisation.
+Compiling without auto-vectorisation (``-no-vec''), the kernel
+achieves a performance that is largely limited by the bandwidth to
+main memory. Note that there is very little opportunity for temporal
+cache re-use in the Continuity kernel -- only three of the twelve
+distinct memory references are to previous rows of the array. The size
+of this effect can be seen in the small drop in performance (black
+bars) in going from the $256^2$ to the $512^2$ domain as at this point
+the working-set size spills out of L3 cache.
+
+Figure~\ref{FIG_kernel_perf} shows the performance of
 the Continuity and u-Momentum kernels when compiled without
@@
     function of SIMD vectorisation and problem size. Results are for
-    v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.}
+    v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.{\bf This plot
+requires updating or dropping in favour of the clustered-bar form.}}
   \label{FIG_kernel_perf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=130mm]{continuity_opt}
+  \caption{Performance of the Continuity kernel with optimisations
+    intended to aid SIMD vectorisation. Results are for v.16 of the
+    Intel compiler on a Xeon E5-1620 v.2 CPU.}
+  \label{FIG_cont_vec}
 \end{figure}
@@
   \includegraphics[width=120mm]{nemolite2d_cont_perf}
-  \caption{Performance of the Continuity kernel (bars -- right y-axis)
-    and HWPC data (open circles -- left y-axis) as a function of
+  \caption{Performance of the Continuity kernel (open circles -- right y-axis)
+    and associated HWPC data (bars -- left y-axis) as a function of
     problem size. Results are for v.16 of the Intel compiler on a Xeon
@@
   \label{FIG_cont_perf}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=120mm]{continuity_opt}
-  \caption{Performance of the Continuity kernel with optimisations
-    intended to aid SIMD vectorisation. Results are for v.16 of the
-    Intel compiler on a Xeon E5-1620 v.2 CPU.}
-  \label{FIG_cont_vec}
 \end{figure}

### 2016-02-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more text describing clustered bar chart results

@@
 \hline
-Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
+                             & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
+WSS of time-stepping (bytes) & $104n^2$ & \multicolumn{2}{c|}{$180n^2$} \\ 
 \hline
@@
 $AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
-Working-set size (bytes)   & $56n^2$ &  $72n^2$   &   $136n^2$ \\
+WSS (bytes)                & $56n^2$ &  $72n^2$   &   $136n^2$ \\
 \hline
@@
 \caption{Details of the kernels investigated with the roofline
-  model. $n$ is the problem size -- the length of one side of the
-  square domain.}
+  model. WSS is the Working Set Size and $n$ is the problem size --
+  the length of one side of the square domain.}
 \label{TAB_kernel_details}
@@
 the working-set sizes for this kernel for the $128^2$ and $256^2$
-domains are approximately 1.1~MB and 4.7~MB, respectively and therefore
-both of these should fit comfortably within L3 cache. Despite this,
-the performance achieved with these domain sizes falls short of the L3
-bandwidth limit.  Although this kernel is one of several within the
-time-stepping loop of NEMOLite2D, the working-set size of the latter
-for the $128^2$ domain is just 2.4~MB which again is well within the
-capacity of L3.
+domains are approximately 1.1~MB and 4.7~MB, respectively and
+therefore both of these should fit comfortably within L3
+cache. Despite this, the performance achieved with these domain sizes
+falls short of the L3 bandwidth limit.  However, the Continuity kernel is just
+one of several within the time-stepping loop of NEMOLite2D; the loop
+as a whole has a working-set size of $180\times n^{2}$ bytes. For the
+$32^2$ domain this is 180~KB, for $52^2$ it is 475~KB and for the
+$128^2$ domain it is 2.9~MB which is well within the capacity of L3.
 
@@
 and after code modifications intended to improve vectorisation.
-Compiling without auto-vectorisation (``-no-vec''), the kernel
-achieves a performance that is largely limited by the bandwidth to
-main memory. Note that there is very little opportunity for temporal
-cache re-use in the Continuity kernel -- only three of the twelve
-distinct memory references are to previous rows of the array. The size
-of this effect can be seen in the small drop in performance (black
-bars) in going from the $256^2$ to the $512^2$ domain as at this point
-the working-set size spills out of L3 cache.
+Compiling without auto-vectorisation (compiler flag ``-no-vec''), the
+kernel achieves a performance that appears to be largely limited by
+the bandwidth to main memory. Note that there is very little
+opportunity for temporal cache re-use in the Continuity kernel -- only
+three of the twelve distinct memory references are to previous rows of
+the array. The size of this effect can be seen in the small drop in
+performance (black bars) in going from the $256^2$ to the $512^2$
+domain as at this point the working-set size of this kernel spills out
+of L3 cache.
 
-Figure~\ref{FIG_kernel_perf} shows the performance of
-the Continuity and u-Momentum kernels when compiled without
-vectorisation (``-no-vec''), with SSE4 (Streaming SIMD Extensions)
-vectorisation (``-axSSE4.2'') and with AVX (``-xHost''). SSE supports
-vector lengths of 128 bits (two double-precision words) and AVX 512
-bits (four double-precision words). For the 64, 128 and 256 problem
-sizes, changing from no vectorisation to SSE increases the performance
-of the Continuity kernel by up to roughly 60\%; some way short of the
-ideal speed-up of 100\% that would be obtained with perfect
-SSE vectorisation of a compute-bound kernel. Unsurprisingly then, moving
-from SSE to AVX vectors does not give any performance improvement and
-in fact, is slightly detrimental for most of the problem sizes.
+\begin{figure}
+  \centering
+  \includegraphics[width=130mm]{continuity_opt}
+  \caption{Performance of the Continuity kernel with optimisations
+    intended to aid SIMD vectorisation. Results are for v.16 of the
+    Intel compiler on a Xeon E5-1620 v.2 CPU. Optimisations are
+    cumulative for results with a given SIMD instruction set.}
+  \label{FIG_cont_vec}
+\end{figure}
+
+The E5-1620 CPU supports two different SIMD instruction sets;
+Streaming SIMD Extensions (SSE; flag ``-axSSE4.2'') and Advanced
+Vector Extensions (AVX; flag ``-xHost''). SSE supports vector lengths
+of 128 bits (two double-precision words) and AVX 256 bits (four
+double-precision words). For the problem sizes that largely fit within
+L3 or L2 cache, switching on the use of SSE instructions provides a
+speed-up of between 1.5 and 1.8 - the factor increasing with problem
+size until the WSS of the time-stepping loop begins to fall out of L3
+cache. It is therefore likely that the reason this factor is below the
+ideal value of 2.0 is the overhead associated with constructing the
+loop and handling any ``peel'' or ``remainder'' loops. As the problem
+size increases, this overhead (as a fraction of the vectorisable
+workload) decreases.
+
+Figure~\ref{FIG_kernel_perf} shows the performance of the Continuity
+and u-Momentum kernels when compiled without vectorisation
+(``-no-vec''), with SSE vectorisation (``-axSSE4.2'') and with AVX
+(``-xHost''). SSE supports vector lengths of 128 bits (two
+double-precision words) and AVX 512 bits (four double-precision
+words). For the 64, 128 and 256 problem sizes, changing from no
+vectorisation to SSE increases the performance of the Continuity
+kernel by up to roughly 60\%; some way short of the ideal speed-up of
+100\% that would be obtained with perfect SSE vectorisation of a
+compute-bound kernel. Unsurprisingly then, moving from SSE to AVX
+vectors does not give any performance improvement and in fact, is
+slightly detrimental for most of the problem sizes.
 
@@
   \label{FIG_kernel_perf}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=130mm]{continuity_opt}
-  \caption{Performance of the Continuity kernel with optimisations
-    intended to aid SIMD vectorisation. Results are for v.16 of the
-    Intel compiler on a Xeon E5-1620 v.2 CPU.}
-  \label{FIG_cont_vec}
 \end{figure}

### 2016-02-23 nemolite2d_psykal/time_step_mod_align_preload.f90
nemolite2d: create separate versions of the continuity kernel for each optimisation

@@ -535,11 +535,27 @@ contains
     ! Locals
     integer :: jj, ji, idxt
     real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
-    integer :: nrepeat, ic
+    integer :: nrepeat
 !DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
 !DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
+
+    ! Hack to ensure all our arrays are loaded into cache - we
+    ! simply execute the whole loop here first before entering
+    ! the timed region
+    do jj = 2, N, 1
+      do ji = 2, M, 1
+
+         rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
+         rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
+
+         ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+                       rdt / area_t(ji,jj)
+      end do
+    end do
 
     nrepeat = 1
 

### 2016-02-23 nemolite2d_psykal/time_step_mod_alignall_preload.f90

@@ -535,16 +535,33 @@ contains
     ! Locals
     integer :: jj, ji, idxt
     real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
-    integer :: nrepeat, ic
+    integer :: nrepeat
 !DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
 !DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
 
+    ! Hack to ensure all our arrays are loaded into cache - we
+    ! simply execute the whole loop here first before entering
+    ! the timed region
+    do jj = 2, N, 1
+      do ji = 2, M, 1
+
+         rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
+         rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
+
+         ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+                       rdt / area_t(ji,jj)
+      end do
+    end do
+
     nrepeat = 1
 
     call timer_start('Continuity',idxt,nrepeat)
     !call likwid_markerStartRegion('Continuity')
+    !do ic=1, nrepeat
 !DIR$ VECTOR ALIGNED
     do jj = 2, N, 1
 
@@ -561,14 +578,13 @@ contains
       do ji = ALIGNMENT+1, M, 1
 
          rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
-         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj))*un(ji-1,jj)
+         rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))*un(ji-ALIGNMENT,jj)
[…]

### 2016-02-23 nemolite2d_psykal/time_step_mod_alignall_preload_nopeel.f90

@@ -535,40 +535,48 @@ contains
     ! Locals
     integer :: jj, ji, idxt
     real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
-    integer :: nrepeat, ic
+    integer :: nrepeat
 !DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
 !DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
 
-    nrepeat = 1
-
-    call timer_start('Continuity',idxt,nrepeat)
-    !call likwid_markerStartRegion('Continuity')
-!DIR$ VECTOR ALIGNED
+    ! Hack to ensure all our arrays are loaded into cache - we
+    ! simply execute the whole loop here first before entering
+    ! the timed region
     do jj = 2, N, 1
+      do ji = 2, M, 1
 
-      ! Explicit peel loop
-      do ji = 2, ALIGNMENT
          rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
          rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
          rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
          rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
-         ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
-               rdt / area_t(ji,jj)
-      end do
-      
-      do ji = ALIGNMENT+1, M, 1
-
-         rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
-         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj))*un(ji-1,jj)
-         rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
-         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji ,jj-1))*vn(ji,jj-1)
 
[…]

### 2016-02-23 nemolite2d_psykal/time_step_mod_alignall_preload_nopeel_8arrays.f90

@@ -535,11 +535,27 @@ contains
     ! Locals
     integer :: jj, ji, idxt
     real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
-    integer :: nrepeat, ic
+    integer :: nrepeat
 !DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
 !DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
 !DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
+
+    ! Hack to ensure all our arrays are loaded into cache - we
+    ! simply execute the whole loop here first before entering
+    ! the timed region
+    do jj = 2, N, 1
+      do ji = 2, M, 1
+
+         rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
+         rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
+
+         ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+                       rdt / area_t(ji,jj)
+      end do
+    end do
 
     nrepeat = 1
 
@@ -548,27 +564,16 @@ contains
 !DIR$ VECTOR ALIGNED
     do jj = 2, N, 1
 
-      ! Explicit peel loop
-      do ji = 2, ALIGNMENT
-         rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
-         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
-         rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
-         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
[…]

### 2016-02-23 docs/papers/roofline_model/gocean_roofline.tex
roofline: text describing SIMD opts and update of AVX performance figures

@@
 
+Examining the optimisation report produced by the Intel compiler
+revealed that the majority of array accesses within the loop were not
+aligned:
+\begin{verbatim}
+Example of unaligned access report here
+\end{verbatim}
+and that this inhibited efficient vectorisation. Tackling this issue
+requires two steps; first, ensure that the arrays are allocated to
+align with 64-byte boundaries and second, ensure that the compiler is
+aware of this where those arrays are used. The first requirement is
+easily met by passing the ``-align array64byte'' flag to
+the compiler when building the code. This ensures that any array that
+is constructed using a call to ALLOCATE is aligned on 64-byte
+boundaries in memory.
+
+The second requirement is complicated by the fact that the NEMOLite2D
+code makes extensive use of Fortran derived types --- the data arrays
+used for computation are members of derived types. The information
+about the alignment of these arrays cannot therefore be propagated
+through the code by module {\tt use} statements. In fact, even the
+Intel compiler directive which can be used to specify the alignment
+of a particular array at its point of use will not accept references
+to members of a derived type. It was therefore necessary to wrap the
+kernel call (and associated loops) within a routine and pass in
+references to the arrays. Within the wrapper routine the corresponding
+dummy arguments are declared as simple arrays and their alignment can
+then be specified using the {\tt ASSUME\_ALIGNED} directive:
+\begin{verbatim}
+subroutine invoke_continuity_arrays(nx, ny, ssha, ...)
+  integer, intent(in)   :: nx, ny
+  real(wp), intent(out) :: ssha(nx,ny)
+!DIR$ ASSUME_ALIGNED ssha:64
+  ...
+\end{verbatim}
+This approach also enables the extent of the arrays ({\tt nx} and {\tt
+  ny} in this example) to be specified explicitly.
+
+The continuity kernel includes array accesses of the form {\tt
+  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
+then this specifies a location in memory {\tt Nx} words before that
+occupied by {\tt a(i,j)}. In order for such an access to be aligned,
+{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
+AVX) and this must be specified to the compiler. We therefore padded Nx
+as required and added a directive within the wrapper routine:
+\begin{verbatim}
+!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
+\end{verbatim}
+
+Once all of these steps were complete, the only un-aligned array
+accesses reported by the compiler were those of the form {\tt
+  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
+unaligned accesses are unavoidable in a stencil-based
+finite-difference code.
+
+The final SIMD-related optimisations are concerned with what happens
+at the beginning and end of a SIMD-vectorisable loop. If the loop
+updates e.g. {\tt a(i,j)} and begins on a value of {\tt i} that
+  results in unaligned memory accesses (e.g. {\tt i = 2}) then the
+  compiler must insert a ``peel'' loop to handle those iterations
+  required to reach a point where memory is accessed on a 64-byte
+  boundary.  Similarly, if the trip-count of the resulting
+  vectorisable loop is not an integer multiple of the SIMD vector
+  width then the compiler must insert a ``remainder'' loop to handle
+  those iterations left over.  We removed the remainder loop by
+  ensuring that the length of the inner loop was a multiple of the
+  vector width and adding an {\tt ASSUME} compiler directive
+  (c.f. that for {\tt Nx} above). This was straightforward since we
+  had already padded the array extents themselves and NEMOLite2D uses
+  a mask to define the active computational domain - the additional
+  grid points resulting from the padding were simply marked as
+  external to the domain.  Despite doing this we found that the
+  remainder loop was not removed until we manually introduced the peel
+  loop (rather than relying on the compiler).
+
+The inner loop of the kernel where all array accesses are aligned and
+the number of distinct streams has been reduced to eight (with a
+concommitant reduction in the number of FLOPs):
+\begin{verbatim}
+do ji = ALIGNMENT+1, M, 1
+   rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+   rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))* &
+           un(ji-ALIGNMENT,jj)
+   rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+   ssha(ji,jj) =  (rtmp2 - rtmp1 - rtmp3) / sshn_t(ji,jj)
+end do
+\end{verbatim}
+
 Figure~\ref{FIG_kernel_perf} shows the performance of the Continuity

### 2016-02-24 nemolite2d_psykal/time_step_mod_alignall_preload.f90
nemolite2d: improve the alignall code so only refs array(i,j-1) instead of array(i,j-ALIGNMENT)

@@ -568,7 +568,7 @@ contains
       ! Explicit peel loop
       do ji = 2, ALIGNMENT
          rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
-         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
+         rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))*un(ji-ALIGNMENT,jj)
          rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
          rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
          ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
@@ -580,7 +580,7 @@ contains
          rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
          rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))*un(ji-ALIGNMENT,jj)
          rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
-         rtmp4 = (sshn_v(ji ,jj-ALIGNMENT) + hv(ji ,jj-ALIGNMENT))*vn(ji,jj-ALIGNMENT)
+         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji ,jj-1))*vn(ji,jj-1)
          ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
                        rdt / area_t(ji,jj)
       end do

### 2016-02-24 nemolite2d_psykal/time_step_mod_alignall_preload_nopeel.f90

@@ -572,7 +572,7 @@ contains
          rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
          rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))*un(ji-ALIGNMENT,jj)
          rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
-         rtmp4 = (sshn_v(ji ,jj-ALIGNMENT) + hv(ji ,jj-ALIGNMENT))*vn(ji,jj-ALIGNMENT)
+         rtmp4 = (sshn_v(ji ,jj-1) + hv(ji ,jj-1))*vn(ji,jj-1)
          ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
                        rdt / area_t(ji,jj)
       end do

### 2016-03-07 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more discussion of Continuity SIMD results.

@@
 performance points in Figure~\ref{FIG_roofline_archer} are very close
-together.
+together and well below any of the roofline limits.
 
@@
 cache. Despite this, the performance achieved with these domain sizes
-falls short of the L3 bandwidth limit.  However, the Continuity kernel is just
-one of several within the time-stepping loop of NEMOLite2D; the loop
-as a whole has a working-set size of $180\times n^{2}$ bytes. For the
-$32^2$ domain this is 180~KB, for $52^2$ it is 475~KB and for the
-$128^2$ domain it is 2.9~MB which is well within the capacity of L3.
+falls short of the L3 bandwidth limit.  Even when we consider that the
+Continuity kernel is just one of several within the time-stepping loop
+of NEMOLite2D we see that the L3 bandwidth limit is still applicable;
+the loop as a whole has a working-set size of $180\times n^{2}$
+bytes. For the $32^2$ domain this is 180~KB ({\it i.e.}\ within L2),
+while the $52^2$ (475~KB) through to the $128^2$ domain (2.9~MB) fall
+well within the capacity of L3.
 
@@
 \end{verbatim}
+where {\tt ALIGNMENT} is an integer parameter declared to have the value two
+when compiling with SSE and four with AVX.
 
-Once all of these steps were complete, the only un-aligned array
+We can see in Figure~\ref{FIG_cont_vec} that these changes (``SSE
+align'') significantly improve the performance of the smaller
+domains. Again, as the domain size increases this improvement reduces
+until, for $512^2$ and $1024^2$ (where we are limited by the bandwidth
+to main memory), this optimisation has no effect.
+
+Since the Continuity kernel is one of several in the time-stepping
+loop, we experimented with pre-loading the fields into cache simply by
+executing the Continuity kernel immediately prior to executing it for
+a second time with timing. Figure~\ref{FIG_cont_vec} shows that this
+``pre-loading'' slightly improved the kernel performance for those
+problem sizes that fit within cache. {\bf Discuss this}
+
+Once all of the above steps were complete, the only un-aligned array
 accesses reported by the compiler were those of the form {\tt
@@
 unaligned accesses are unavoidable in a stencil-based
-finite-difference code.
+finite-difference code. However, in order to quantify how much these
+accesses limit the performance of the kernel we experimented with
+replacing e.g. {\tt a(i-1,j)} with {\tt a(i-ALIGNMENT,j)}. The
+resulting kernel had no un-aligned accesses and performed
+significantly better; actually reaching or exceeding the L3-bandwidth
+roofline for domains $64^2$ -- $256^2$.
 

### 2016-03-09 docs/papers/roofline_model/gocean_roofline.tex
More updates to roofline doc

@@
 written by Paul Swarztrauber over 30 years ago and has been optimised
-by various people over that time. In contrast, nemolite2d is a new,
+by various people over that time. In contrast, NEMOLite2D is a new,
 un-optimised code that has been extracted from NEMO.
@@
 Although we have investigated how the performance of the \psykal
-version of nemolite2d compares with that of the original, we have not
+version of NEMOLite2D compares with that of the original, we have not
 addressed how efficient the original actually is. In order to do so we
@@
 \caption{Results from running the STREAM2 benchmark on the E5-2697
-  CPUs. Error bars indicate the maximum and minimum values obtained
-  from 200 repeats of each case.  size. }
+  CPU.}
 \label{FIG_stream2}
@@
 Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
-footprint spills out of L1 cache (which all of our nemolite2d
+footprint spills out of L1 cache (which all of our NEMOLite2D
 benchmark configurations do), the performance counter results
@@
 FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
-kernel of nemolite2d then we find that there are 51 addition
+kernel of NEMOLite2D then we find that there are 51 addition
 operations, 56 multiplication/division operations and one call to {\tt
@@
 \hline
-                             & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
+                             & Shallow & \multicolumn{2}{c|}{NEMOLite2D} \\
 WSS of time-stepping (bytes) & $104n^2$ & \multicolumn{2}{c|}{$180n^2$} \\ 
@@
 and show the performance of the 1st loop-nest of Shallow and the
-u-Momentum kernel of nemolite2d. Two things are immediately clear from
+u-Momentum kernel of NEMOLite2D. Two things are immediately clear from
 this plot: first, the performance of the first loop nest of Shallow is
@@
 for working-set size in Table~\ref{TAB_kernel_details}, these
-correspond to 14.7 and 58.7~MB, respectively. Since the E5-2697 has
-an L3 cache of 30~MB, the $512^2$ case should be contained in cache
-while the $1024^2$ case will spill to main memory. This is borne out
-by the plot; the $1024^2$ case is achieving a performance in-line
+correspond to 14.7 and 58.7~MB, respectively. Since the E5-2697 has an
+L3 cache of 30~MB, the $512^2$ case should be contained in cache while
+the $1024^2$ case will spill to main memory. This is borne out by the
+plot; the $1024^2$ case is achieving a performance roughly in-line
 with saturating the bandwidth to main memory. The $512^2$ case is
-approaching the L3 bandwidth limit. We investigate the reasons for
-the fact that it does not achieve the L3 bandwidth limit below.
+approaching the L3 bandwidth limit. We investigate the reasons for the
+fact that it does not achieve the L3 bandwidth limit below.
 
@@
 {\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
-AVX) and this must be specified to the compiler. We therefore padded Nx
+AVX) and this must be specified to the compiler. We therefore padded {\tt Nx}
 as required and added a directive within the wrapper routine:
@@
 significantly better; actually reaching or exceeding the L3-bandwidth
-roofline for domains $64^2$ -- $256^2$.
+roofline for domains of $52^2$ -- $256^2$ (Figure~\ref{FIG_cont_vec} -
+``align all'').
 
@@
 at the beginning and end of a SIMD-vectorisable loop. If the loop
-updates e.g. {\tt a(i,j)} and begins on a value of {\tt i} that
-  results in unaligned memory accesses (e.g. {\tt i = 2}) then the
-  compiler must insert a ``peel'' loop to handle those iterations
+updates {\it e.g.}\ {\tt a(i,j)} and begins on a value of {\tt i} that
+results in unaligned memory accesses ({\it e.g.}\ {\tt i = 2}) then
+the compiler must insert a ``peel'' loop to handle those iterations
 required to reach a point where memory is accessed on a 64-byte
-  boundary.  Similarly, if the trip-count of the resulting
-  vectorisable loop is not an integer multiple of the SIMD vector
-  width then the compiler must insert a ``remainder'' loop to handle
-  those iterations left over.  We removed the remainder loop by
-  ensuring that the length of the inner loop was a multiple of the
-  vector width and adding an {\tt ASSUME} compiler directive
-  (c.f. that for {\tt Nx} above). This was straightforward since we
-  had already padded the array extents themselves and NEMOLite2D uses
-  a mask to define the active computational domain - the additional
-  grid points resulting from the padding were simply marked as
-  external to the domain.  Despite doing this we found that the
-  remainder loop was not removed until we manually introduced the peel
-  loop (rather than relying on the compiler).
+boundary.  Similarly, if the trip-count of the resulting vectorisable
+loop is not an integer multiple of the SIMD vector width then the
+compiler must insert a ``remainder'' loop to handle those iterations
+left over.
+
+We removed the remainder loop by ensuring that the length of the inner
+loop was a multiple of the vector width and adding an {\tt ASSUME}
+compiler directive ({\it c.f.}\ that for {\tt Nx} above). This was
+straightforward since we had already padded the array extents
+themselves and NEMOLite2D uses a mask to define the active
+computational domain - the additional grid points resulting from the
+padding were simply marked as external to the domain.  Despite doing
+this we found that the remainder loop was not removed until we
+manually introduced the peel loop (rather than relying on the
+compiler). Figure~\ref{FIG_cont_vec} shows that the resulting binary
+was the most performant of all those compiled with SSE on all problem
+sizes that are contained within L3 cache with the exception of $52^2$.
+{\bf Why is $52^2$ an exception? Something to do with tuning it to fit in L2?}
+
+Since the caches of both of the CPUs used are eight-way associative,
+we also considered whether we might be experiencing ``cache
+thrashing'' whereby different locations in memory are mapped to the
+same location in memory.
 

### 2016-03-09 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more discussion of SIMD results and correct perf results for 52x52 domain as I actually ran 50x50

@@
 size increases, this overhead (as a fraction of the vectorisable
-workload) decreases.
+workload) decreases. {\bf if this is the case then why is my 128x8 domain
+not faster than the 32x32?}
+
+When compiling with AVX instructions rather than SSE, the performance
+of the Continuity kernel was typically just below that of the latter
+for all of the problem sizes fitting within cache. We attribute this
+to the higher overheads associated with managing the longer AVX
+vectors.
 
@@
 
+In what follows, the code modifications were no longer restricted to
+those that maintained correctness since we were interested in what
+characteristics of the code were limiting its performance.
+
 Since the Continuity kernel is one of several in the time-stepping
@@
 thrashing'' whereby different locations in memory are mapped to the
-same location in memory.
-
-The inner loop of the kernel where all array accesses are aligned and
-the number of distinct streams has been reduced to eight (with a
-concommitant reduction in the number of FLOPs):
+same location in cache. We therefore constructed a version of the
+kernel with the number of distinct streams reduced from twelve to
+eight. In order to maintain the same Arithmetic Intensity as the
+original kernel we also reduced the number of FLOPs. The inner loop of
+the resulting kernel was:
 \begin{verbatim}
@@
 \end{verbatim}
+
+Figure~\ref{FIG_cont_vec} shows that the absolute performance of this
+modified kernel (``8 arrays'') is mostly lower than that of the kernel
+with twelve streams and certainly never exceeds it. We therefore
+conclude that cache-thrashing is not a problem for this kernel.
 

### 2016-03-10 docs/papers/roofline_model/gocean_roofline.tex
roofline: correct some typos

@@
 written by Paul Swarztrauber over 30 years ago and has been optimised
-by various people over that time. In contrast, NEMOLite2D is a new,
+by various people since that time. In contrast, NEMOLite2D is a new,
 un-optimised code that has been extracted from NEMO.
 
-Although we have investigated how the performance of the \psykal
+Although we have investigated how the performance of the \psykal\ 
 version of NEMOLite2D compares with that of the original, we have not
@@
 \hline
-  \multicolumn{5}{|c|}{E5-1620 v.~2, 3.7GHz} \\
+  \multicolumn{5}{|c|}{E5-1620 v.2, 3.7GHz} \\
   \hline
@@
   \hline
-\multicolumn{5}{|c|}{E5-2697 v.~2, 2.7GHz}  \\
+\multicolumn{5}{|c|}{E5-2697 v.2, 2.7GHz}  \\
 \hline

### 2016-03-30 docs/papers/roofline_model/gocean_roofline.tex
roofline: correct daxpy AI and update continuity-opt plot to only show sse results

@@
 
+\begin{verbatim}
+!DIR$ VECTOR ALIGNED
+do ic = 1, nrepeat, 1
+  do jj = 2, N, 1
+
+    ! Explicit peel loop
+    do ji = 2, ALIGNMENT
+       rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+       rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
+       rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+       rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
+       ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+             rdt / area_t(ji,jj)
+    end do
+
+    do ji = ALIGNMENT+1, M, 1
+
+       rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+       rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj))*un(ji-1,jj)
+       rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+       rtmp4 = (sshn_v(ji ,jj-1) + hv(ji ,jj-1))*vn(ji,jj-1)
+
+       ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+                     rdt / area_t(ji,jj)
+    end do
+
+  end do
+end do
+\end{verbatim}
+
 Since the caches of both of the CPUs used are eight-way associative,
@@
 In order to help further understand the performance of the Continuity
-kernel in Figure~\ref{FIG_cont_perf} we plot the performance of the
+kernel, in Figure~\ref{FIG_cont_perf} we plot the performance of the
 Continuity kernel alongside HWPC-derived metrics (obtained using the
@@
     problem size. Results are for v.16 of the Intel compiler on a Xeon
-    E5-1620 v.2 CPU.}
+    E5-1620 v.2 CPU using only the SSE SIMD instructions.}
   \label{FIG_cont_perf}
@@
 \begin{enumerate}
-\item {\tt y(i) = y(i) + a*x(i)}, AI = 0.125
+\item {\tt y(i) = y(i) + a*x(i)}, AI = 0.083
 \item {\tt y(i) = y(i) + a*x(i)+ x(i)*y(i)}, AI = 0.167
@@
 of each kernel is plotted for four problem sizes, chosen so as to fit
-within L1, L2, L3 cache and main memory. The performance of kernels 2,
-3 and 4 is entirely consistent with the roofline model for problem
-sizes that fit within the L3 and L3 caches (red and black bars). Their
+within L1, L2, L3 cache and main memory. The performance of kernels
+1--4 is entirely consistent with the roofline model for problem sizes
+that fit within the L3 and L3 caches (red and black bars). Their
 performance for the smallest problem size doesn't quite reach the L1
-bandwidth limit but does appear to track the roofline.  Kernel 1 has
-the lowest AI of the five kernels and this appears to limit its
-performance; at no point does the performance reach the limit imposed
-by the bandwidth to the relevant level of the memory hierarchy. {\bf need to expore the reason for this}
+bandwidth limit but does appear to track the roofline.
 

### 2016-03-30 nemolite2d_perf.tex
nemolite2d paper: tidying

@@
 
-\author{J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Liu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
+\author{A.~R.~Porter, J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Liu and G.~D.~Riley% <-this % stops a space
+\thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
 \thanks{J.~Appleyard is with NVIDIA.}%
 \thanks{H.~Liu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
-\thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
 \thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space

### 2016-05-09 docs/papers/roofline_model/gocean_roofline.tex
roofline: update stream2 results for desktop with effect of turbo/stepping

@@
 \hline
-STREAM2 (GB/s)   &  160  &  61.0  & 46.0 &  20.5 \\
+STREAM2 (GB/s)   &  140  &  65.0  & 46.7 &  20.5 \\
   \hline
@@
 straightforward. We therefore used the STREAM2 benchmark which is
-intended for this purpose. In Figure~\ref{FIG_stream2} we plot the
+intended for this purpose. In Figure~\ref{FIG_stream2_archer} we plot the
 results obtained from the DAXPY component of the STREAM2 benchmark on
@@
 \includegraphics[width=120mm]{stream2_archer}
-\caption{Results from running the STREAM2 benchmark on the E5-2697
-  CPU.}
-\label{FIG_stream2}
+\caption{Bandwidth results from running the DAXPY component of the
+  STREAM2 benchmark on the E5-2697 CPU.}
+\label{FIG_stream2_archer}
 \end{figure}
@@
  The equivalent results for the E5-1620 CPU are shown in
- Figure~\ref{FIG_daxpy_bw}. This plot has the same shape as that for
- the E5-2697 but is shifted upwards indicating an increase in
+ Figure~\ref{FIG_stream2_desktop}. This plot has the same shape as
+ that for the E5-2697 but is shifted upwards indicating an increase in
  bandwidth accross all levels of the cache hierarchy, but particularly
@@
  the L3 cache is narrower because of the smaller size of this cache on
- the E5-1620.
+ the E5-1620. Figure~\ref{FIG_stream2_desktop} also shows the effect
+ on performance of disabling the turbo and frequency-stepping modes of
+ the CPU. (Both sets of results are with hyper-threading disabled in
+ order to reduce noise due to resource contention.) This reduces the
+ performance obtained from the various levels of the cache hierarchy
+ by a few percent but increases the bandwidth obtained to main-memory
+ by a similar amount.
+
+\begin{figure}
+\includegraphics[width=120mm]{stream2_desktop}
+\caption{Bandwidth results from running the DAXPY component of the
+  STREAM2 benchmark on the E5-1620 CPU. Three data sets are plotted
+  corresponding to three consecutive runs of the benchmark.}
+\label{FIG_stream2_desktop}
+\end{figure}
 
@@
   modified versions of the daxpy kernel from the STREAM2 benchmark.}
-\label{FIG_daxpy_bw}
+\label{FIG_stream2_daxpy}
 \end{figure}

### 2016-05-27 nemolite2d_psykal/continuity_mod.f90
nemolite2d_align: move invoke_continuity_arrays() back into continuity_mod.f90 to enable better control over in-lining

@@ -62,6 +62,69 @@ contains
 
   end subroutine invoke_continuity
 
+  !=================================================================
+
+  subroutine invoke_continuity_arrays(nx, ny, M, N, rdt, ssha, &
+                                      sshn_t, sshn_u, sshn_v, &
+                                      hu, hv, un, vn, area_t)
+    use global_parameters_mod, only: ALIGNMENT
+    use kind_params_mod
+    use dl_timer, only: timer_start, timer_stop
+    implicit none
+    integer, intent(in) :: nx, ny, M, N
+    real(wp), intent(in) :: rdt
+    real(wp), intent(out) :: ssha(nx,ny)
+    real(wp), intent(in)  :: sshn_u(nx,ny), sshn_v(nx,ny), sshn_t(nx,ny)
+    real(wp), intent(in)  :: un(nx,ny), vn(nx,ny)
+    real(wp), intent(in)  :: hu(nx,ny), hv(nx,ny), area_t(nx,ny)
+    ! Locals
+    integer :: jj, ji
+    real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
+    !> For timing
+    integer, save :: idxt
+    integer :: nrepeat, ic
+!xxxDIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
+!xxxDIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
+!xxxDIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
+!xxxDIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
+
+    !nrepeat = 100 
+    nrepeat = 1
+   
+    call timer_start(idxt, label='Continuity', num_repeats=nrepeat)
+    !call likwid_markerStartRegion('Continuity')
+!xxxxxDIR$ VECTOR ALIGNED
+    !do ic = 1, nrepeat, 1
+    do jj = 2, N, 1
+
+      ! Explicit peel loop
+      do ji = 2, ALIGNMENT
+         rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+         rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
[…]

### 2016-05-27 nemolite2d_psykal/time_step_mod_align.f90

@@ -20,7 +20,7 @@ contains
     use physical_params_mod, only: g, omega, d2r
 !    use momentum_mod,    only: momentum_v_code
 !    use momentum_mod,    only: momentum_u_code
-!    use continuity_mod,  only: continuity_code
+    use continuity_mod,  only: invoke_continuity_arrays
 !    use time_update_mod, only: next_sshu_code, next_sshv_code
     use boundary_conditions_mod
     implicit none
@@ -521,69 +521,6 @@ contains
     call timer_stop(next_timer)
 
   end subroutine invoke_time_step
-
-  !=================================================================
-
-  subroutine invoke_continuity_arrays(nx, ny, M, N, rdt, ssha, &
-                                      sshn_t, sshn_u, sshn_v, &
-                                      hu, hv, un, vn, area_t)
-    use global_parameters_mod, only: ALIGNMENT
-    use kind_params_mod
-    use dl_timer, only: timer_start, timer_stop
-    implicit none
-    integer, intent(in) :: nx, ny, M, N
-    real(wp), intent(in) :: rdt
-    real(wp), intent(out) :: ssha(nx,ny)
-    real(wp), intent(in)  :: sshn_u(nx,ny), sshn_v(nx,ny), sshn_t(nx,ny)
-    real(wp), intent(in)  :: un(nx,ny), vn(nx,ny)
-    real(wp), intent(in)  :: hu(nx,ny), hv(nx,ny), area_t(nx,ny)
-    ! Locals
-    integer :: jj, ji, idxt
-    real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
-    !> For timing
-    real(wp) :: t1, t2
-    integer :: nrepeat, ic
-!DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
-!DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
-!DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
-!DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
-
-    nrepeat = 100 
-    !nrepeat = 1
-   
[…]

### 2016-06-01 shallow/SEQ/original/shallow_base.f90
shallow: update to use new dl_timer API

@@ -78,19 +78,25 @@
       INTEGER :: ierr
 
       !> Integer tags for timers
-      INTEGER :: idxt0, idxt1
-
+      INTEGER :: idxt0, cu_timer, pbcs_timer, unew_timer, tsmooth_timer
 
 !  ** Initialisations ** 
-      CALL timer_init()
-
+      call timer_init()
+      
 !     Read in namelist 
       OPEN(unit=input_unit, file=nml_name, status='old',iostat=ierr)
-        CALL check(ierr, "open "//nml_name)
+      CALL check(ierr, "open "//nml_name)
       READ(unit=input_unit, nml=global_domain, iostat=ierr)
-        CALL check(ierr, "read "//nml_name)
+      CALL check(ierr, "read "//nml_name)
       READ(unit=input_unit, nml=io_control, iostat=ierr)
-        CALL check(ierr, "read "//nml_name)
+      CALL check(ierr, "read "//nml_name)
+
+      call timer_register(idxt0,     label='Time-stepping', &
+                          num_repeats=itmax)
+      call timer_register(cu_timer,  label='Compute CU,CV,CZ,H')
+      call timer_register(pbcs_timer,label='PBCs')
+      call timer_register(unew_timer,label='Compute {U,V,P}NEW')
+      call timer_register(tsmooth_timer,label='Time smooth')
 
 !     Set up arrays
       m_len = m+1
@@ -210,7 +216,7 @@
       TIME = 0.
 
       !     Start timer
-      CALL timer_start('Time-stepping',idxt0)
+      CALL timer_start(idxt0)
 
 !  ** Start of time loop ** 
       DO ncycle=1,itmax
@@ -221,7 +227,7 @@
 
          !call system_clock(count=c1, count_rate=r,count_max=max)
          !T100 = c1
-         CALL timer_start('Compute CU,CV,CZ,H',idxt1)
+         CALL timer_start(cu_timer)
 
          DO J=1,N
             DO I=1,M
@@ -234,11 +240,11 @@
             END DO
          END DO
 
-         CALL timer_stop(idxt1)
+         CALL timer_stop(cu_timer)
          !call system_clock(count=c2,count_rate=r,count_max=max)
          !T100 = dble(c2-T100)/dble(r)
 
-         CALL timer_start('PBCs',idxt1)
+         CALL timer_start(pbcs_timer)
 !        PERIODIC CONTINUATION
[…]

### 2016-06-01 shallow/SEQ/shallow_base.f90

@@ -157,7 +157,7 @@ program shallow
   CALL model_write(0, p_fld, u_fld, v_fld)
 
   !     Start timer
-  CALL timer_start('Time-stepping',idxt0)
+  CALL timer_start(idxt0, label='Time-stepping')
 
   !====================================
   ! Perform the first time step

### 2016-07-08 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolited2d paper: merge in updates. Roofline plot is now for E5-1620 rather than Archer.

@@
 memory. At this stage the OpenACC version of the code on the GPU is
-some {\bf 3.7 (this is the factor for the CUDA version!)}sssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssssss times faster
+some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
 than the OpenMP version on the CPU. {\bf Can we relate this to the
@@
 follow the approach suggested in~\cite{para_pearls} and use the
-well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
+ STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
 in order to get the upper bounds on the memory bandwidth and FLOPS for
@@
 \includegraphics[width=\picwidth]{roofline}
-\caption{Comparison of the performance achieved by nemolite2d and Shallow}
+\caption{Comparison of the performance achieved by NEMOLite2D and
+  Shallow on the E5-1620 CPU. Results for the former are for the
+  $256^2$ domain since that gave the best performance.}
 \label{FIG_roofline}
 \end{figure}
+
+In Figure~\ref{FIG_roofline} we show the performance of kernels from
+both Shallow and NEMOLite2D on the roofline model for the E5-2697 CPU.
+This demonstrates that the Shallow kernel is achieving a performance
+roughly consistent with saturating the available memory bandwidth. In
+contrast, the kernel taken from NEMOLite2D is struggling to reach a
+performance consistent with saturating the bandwidth to main
+memory. The fact that reducing the problem size does not improve the
+performance points to more fundamental issues with the way that the
+kernel is implemented.
+
+We investigated this using the simpler Continuity kernel since it too
+performs poorly while being much shorter.
+{\bf Plot of Cont. GFLOPS as function of optimisations here.}
+\begin{itemize}
+\item AVX versus SSE versus no SIMD vectorisation.
+\item Telling the compiler about alignment
+\item The division operation
+\end{itemize}
+Move back to Momentum kernel. Discuss further issues:
+\begin{itemize}
+\item IFs within loop (masking)
+\end{itemize}
 

### 2016-07-08 nemolite2d_perf.tex
nemolite2d paper: add CPU ceilings based on DAG analysis to the roofline plot.
nemolite2d paper: update roofline plot with Shallow perf results for desktop rather than Archer
nemolite2d paper: add description of performance model and refs to Agner Fog

@@
 follow the approach suggested in~\cite{para_pearls} and use the
-well-known STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
-in order to get the upper bounds on the memory bandwidth and FLOPS for
-the test hardware. Since we are using an Intel Ivybridge CPU we used
-the Intel Math Kernel Library implementation of LINPACK.
+STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
+get the upper bounds on the memory bandwidth and floating-point
+operations per second (FLOPS) for the test hardware. Since we are
+using an Intel Ivybridge CPU we used the Intel Math Kernel Library
+implementation of LINPACK.
 
@@
 \includegraphics[width=\picwidth]{roofline}
-\caption{Comparison of the performance achieved by nemolite2d and Shallow}
+\caption{Comparison of the performance achieved by NEMOLite2D and
+  Shallow on the E5-1620 CPU. Results for the former are for the
+  $256^2$ domain since that gave the best performance.}
 \label{FIG_roofline}
 \end{figure}
+
+In Figure~\ref{FIG_roofline} we show the performance of kernels from
+both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
+This demonstrates that the Shallow kernel is achieving a performance
+roughly consistent with saturating the available memory bandwidth. In
+contrast, the kernel taken from NEMOLite2D is struggling to reach a
+performance consistent with saturating the bandwidth to main
+memory. We experimented with reducing the problem size (so as to
+ensure it fitted within cache) but that did not significantly improve
+the performance of the kernel. This then points to more fundamental issues
+with the way that the kernel is implemented which are not captured in the
+simple roofline model.
+
+In order to aid our understanding of kernel performance we have
+developed a tool capable of parsing Fortran code and generating a
+Directed Acyclic Graph (DAG) of the data flow. This tool eases the
+laborious and error-prone process of counting memory accesses and
+FLOPs as well as providing information on those operations that are
+rate-limiting or on the critical path. Using the details of the Intel
+Ivy Bridge microarchitecture published by Fog~\cite{fog_microarch,
+  fog_instructions} we have constructed performance estimates of the
+NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
+(i.e. assuming that all nodes in the DAG are executed in serial) we
+get a lower-bound performance estimate of $0.6391 \times CLOCK\_SPEED$
+FLOPS which gives 2.46~GFLOPS at a clock speed of
+3.85~GHz.
+
+Alternatively, we may construct an upper bound by assuming the
+out-of-order execution engine of the Ivy Bridge core is able to
+perfectly schedule and pipeline all operations such that those that go
+to different execution ports are always run in parallel. In the Ivy
+Bridge core floating-point multiplication and division operations go
+to port 0 while addition and subtraction go to port
+1~\cite{fog_instructions}. Therefore we sum the cost of all
+multiplication and division operations in the DAG and compare that
+with the sum of all addition and subtraction operations. The greater
+of these two quantities is then taken to be the cost of executing the
+kernel; all of the operations on the other port are assumed to be done
+in parallel. This gives a performance estimate of $1.029 \times
+CLOCK\_SPEED$ FLOPS or 3.96 GFLOPS at 3.85~GHz.
+
+These performance estimates are plotted as CPU ceilings in
+Figure~\ref{FIG_roofline}. The performance of the Momentum kernel is
+seen to fall between these two bounds which demonstrates that its
+performance is not memory-bandwidth limited, as might have been
+assumed by its low AI. Therefore, although the performance of this
+kernel is clearly sub-optimal, this is due to the balance of
+floating-point operations that it contains. Improving this is outside
+the scope of the current work since we are focused on the \psykal
+separation of concerns and the introduction of parallelism in the PSy
+layer.
 
@@
 
-% that's all folks
 \end{document}

### 2016-07-08 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d paper: remove some of my TODO comments and add SIMD to discussion of kernel performance

@@
 In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
-used with each compiler. {\bf Explanation of the compiler flags?}
-%These flags are particularly important for
-%the \psykal\ versions of the code. The performance of the original
-%version of the code is much less sensitive to the compiler
-%options. This is because it consists of a single source file
-%containing effectively a single routine (excluding those for I/O).
+used with each compiler. For the IBM, Gnu and Intel compilers, we
+include flags to encourage in-lining of kernel bodies. The Intel flag
+``-xHost'' enables the highest-level of SIMD vectorisation supported
+by the host CPU (AVX in this case). The Intel flag ``-fast'' and Cray
+flags ``-O ipa5'' and ``-h wp'' enable inter-procedural optimisation.
 
@@
 \end{verbatim}
-Looking at the `estimated potential speedup' in the compiler outputs
+Looking at the `estimated potential speedup' in the compiler output
 above, it is clear that the way in which the compiler vectorises the
@@
 resulting binary is worse than that where the loop is left
-un-vectorised. {\bf In principle we could investigate this by looking
-  at the assembler that the Intel compiler generates\ldots}
-
-For the best possible performance, we have chosen to do full, manual
+un-vectorised. In principle this could be investigated further by
+looking at the assembler that the Intel compiler generates but that is
+outside the scope of this work.
+
+For the best possible performance, we have therefore chosen to do full, manual
 inlining for the two kernels making up the Momentum section.
@@
 diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
-~\ref{FIG_intel_omp_256}). {\bf Can we explain this? Perhaps with HW
-  counters?}
+~\ref{FIG_intel_omp_256}).
 
@@
 performance - the two-socket performance figures track those from a
-single socket very closely. The only significant difference in performance
-is at 24 threads where, in addition to the difference in available
-memory bandwidth, the single-socket configuration is using
-hyperthreading while the two-socket case is not.  {\bf Have query in
-  with Archer helpdesk on behaviour seen with Cray compiler here. Also
-  would like to run 512 case on 2 sockets to see how it goes out to 48
-  threads.}
+single socket very closely. The only significant difference in
+performance is at 24 threads where, in addition to the difference in
+available memory bandwidth, the single-socket configuration is using
+hyperthreading while the two-socket case is not.  The discrepancy in
+the performance of the Cray- and Intel-compiled binaries at this
+thread count is under investigation by Cray.
 
@@
 follow the approach suggested in~\cite{para_pearls} and use the
- STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks
-in order to get the upper bounds on the memory bandwidth and FLOPS for
-the test hardware. Since we are using an Intel Ivybridge CPU we used
-the Intel Math Kernel Library implementation of LINPACK.
+STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
+get the upper bounds on the memory bandwidth and floating-point
+operations per second (FLOPS) for the test hardware. Since we are
+using an Intel Ivybridge CPU we used the Intel Math Kernel Library
+implementation of LINPACK.
 
@@
 In Figure~\ref{FIG_roofline} we show the performance of kernels from
-both Shallow and NEMOLite2D on the roofline model for the E5-2697 CPU.
+both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
 This demonstrates that the Shallow kernel is achieving a performance
@@
 performance consistent with saturating the bandwidth to main
-memory. The fact that reducing the problem size does not improve the
-performance points to more fundamental issues with the way that the
-kernel is implemented.
-
-We investigated this using the simpler Continuity kernel since it too
-performs poorly while being much shorter.
-{\bf Plot of Cont. GFLOPS as function of optimisations here.}
-\begin{itemize}
-\item AVX versus SSE versus no SIMD vectorisation.
-\item Telling the compiler about alignment
-\item The division operation
-\end{itemize}
-Move back to Momentum kernel. Discuss further issues:
-\begin{itemize}
-\item IFs within loop (masking)
-\end{itemize}
+memory. We experimented with reducing the problem size (so as to
+ensure it fitted within cache) but that did not significantly improve
+the performance of the kernel. This then points to more fundamental issues
+with the way that the kernel is implemented which are not captured in the
+simple roofline model.
+
+In order to aid our understanding of kernel performance we have
+developed a tool capable of parsing Fortran code and generating a
+Directed Acyclic Graph (DAG) of the data flow. This tool eases the
+laborious and error-prone process of counting memory accesses and
+FLOPs as well as providing information on those operations that are
+rate-limiting or on the critical path. Using the details of the Intel
+Ivy Bridge microarchitecture published by Fog~\cite{fog_microarch,
+  fog_instructions} we have constructed performance estimates of the
+NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
+(i.e. assuming that all nodes in the DAG are executed in serial) we
+get a lower-bound performance estimate of $0.6391 \times CLOCK\_SPEED$
+FLOPS which gives 2.46~GFLOPS at a clock speed of
+3.85~GHz.
+
+Alternatively, we may construct an upper bound by assuming the
+out-of-order execution engine of the Ivy Bridge core is able to
+perfectly schedule and pipeline all operations such that those that go
+to different execution ports are always run in parallel. In the Ivy
+Bridge core floating-point multiplication and division operations go
+to port 0 while addition and subtraction go to port
+1~\cite{fog_instructions}. Therefore we sum the cost of all
+multiplication and division operations in the DAG and compare that
+with the sum of all addition and subtraction operations. The greater
+of these two quantities is then taken to be the cost of executing the
+kernel; all of the operations on the other port are assumed to be done
+in parallel. This gives a performance estimate of $1.029 \times
+CLOCK\_SPEED$ FLOPS or 3.96 GFLOPS at 3.85~GHz.
+
+These performance estimates are plotted as CPU ceilings (horizontal
+dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
+Momentum kernel is seen to fall between these two bounds which
+demonstrates that its performance is not memory-bandwidth limited, as
+might have been assumed by its low AI. Therefore, although the
+performance of this kernel is clearly sub-optimal, this is due to the
+balance of floating-point operations that it contains. Improving this
+is outside the scope of the current work since here we are focused on
+the \psykal separation of concerns and the introduction of parallelism
+in the PSy layer.
+
+Enabling SIMD vectorisation for this kernel does not significantly
+improve its performance (Figure~\ref{FIG_roofline}) and in fact,
+limiting it to SSE instructions (vector width of two double-precision
+floating point numbers) rather than AVX (vector width of four) was
+found to produce a slightly more performant version. The poor
+efficiency of the SIMD version is highlighted by the fact that the
+corresponding kernel performance no longer falls within the bounds of
+our performance estimate. This is because we have incorporated the
+effect of SSE by assuming perfect vectorisation which gives a
+performance increase of a factor of two.
 

### 2016-07-11 docs/papers/roofline_model/gocean_roofline.tex
roofline: rename roofline_haswell to roofline_desktop

@@
 
-In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
+In Figure~\ref{FIG_roofline_desktop} we plot the Roofline model
 obtained for the E5-1620 CPU. In order to investigate the poor
@@
 \centering
-\includegraphics[width=120mm]{roofline_haswell}
+\includegraphics[width=120mm]{roofline_desktop}
 \caption{Comparison of the performance achieved by different
   NEMOLite2D kernels on the E5-1620 CPU.}
-\label{FIG_roofline_haswell}
+\label{FIG_roofline_desktop}
 \end{figure}
@@
 for the fact that they do not reach the theoretical limits of the
-roofline plot in Figure~\ref{FIG_roofline_haswell}.
+roofline plot in Figure~\ref{FIG_roofline_desktop}.
 

### 2016-07-12 nemolite2d_psykal/time_step_mod_cont_only.f90
nemolite2d: add time-stepping loop containing only Continuity kernle

@@ -1,0 +1,100 @@
+module time_step_mod
+  implicit none
+
+  private
+
+  public invoke_time_step
+
+contains
+
+  subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
+                              sshn_t, sshn_u, sshn_v, &
+                              hu, hv, ht, ua, va, un, vn)
+    use kind_params_mod
+    use dl_timer
+    use field_mod
+    use grid_mod
+    use model_mod,       only: rdt, cbfr, visc
+    use physical_params_mod, only: g, omega, d2r
+    use boundary_conditions_mod
+    implicit none
+    real(wp),        intent(in)    :: istp
+    type(r2d_field), intent(inout) :: un, vn, sshn_t, sshn_u, sshn_v
+    type(r2d_field), intent(inout) :: ua, va, ssha, ssha_u, ssha_v
+    type(r2d_field), intent(in)    :: hu, hv, ht
+    ! Locals
+    integer :: ji, jj, jiu, jiv
+    integer :: M, N
+    integer :: cont_timer, mom_timer, bc_timer, next_timer
+    ! Locals for momentum
+    REAL(wp) :: u_e, u_w, v_n, v_s
+    real(wp) :: v_nc, v_sc
+    real(wp) :: depe, depw, deps, depn
+    real(wp) :: hpg, adv, cor, vis
+    real(wp) :: dudx_e, dudx_w, dudy_s, dudy_n
+    real(wp) :: uu_e, uu_n, uu_s, uu_w
+    real(wp) :: u_ec, u_wc, vv_e, vv_n, vv_s, vv_w
+    real(wp) :: dvdx_e, dvdx_w, dvdy_n, dvdy_s
+    real(wp) :: rtmp1, rtmp2, rtmp3, rtmp4
+    ! end locals for momentum
+    ! Locals for BCs
+    real(wp) :: amp_tide, omega_tide, rtime
+
+    M  = ssha%grid%simulation_domain%xstop
+    N  = ssha%grid%simulation_domain%ystop
+
+    ! In the general case we have to reason about whether or not the
+    ! domain has PBCs and what sort of offset convention the kernels
+    ! use. However, this is a middle layer specific to NEMOLite2D and
+    ! therefore we know that we have no periodic BCs and are using a
+    ! NE stagger
[…]

### 2016-07-12 nemolite2d_psykal/continuity_mod.f90
nemolite2d: add basic, array version of Continuity kernel to Cont module

@@ -83,17 +83,17 @@ contains
     !> For timing
     integer, save :: idxt
     integer :: nrepeat, ic
-!xxxDIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
-!xxxDIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
-!xxxDIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
-!xxxDIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
+!DIR$ ASSUME (MOD(NX,ALIGNMENT) .EQ. 0)
+!DIR$ ASSUME (MOD(M,ALIGNMENT) .EQ. 0)
+!DIR$ ASSUME_ALIGNED ssha:64, sshn_u:64, sshn_v:64, sshn_t:64
+!DIR$ ASSUME_ALIGNED un:64, vn:64, hu:64, hv:64, area_t:64
 
     !nrepeat = 100 
     nrepeat = 1
    
     call timer_start(idxt, label='Continuity', num_repeats=nrepeat)
     !call likwid_markerStartRegion('Continuity')
-!xxxxxDIR$ VECTOR ALIGNED
+!DIR$ VECTOR ALIGNED
     !do ic = 1, nrepeat, 1
     do jj = 2, N, 1
 
@@ -124,6 +124,56 @@ contains
     call timer_stop(idxt)
 
   end subroutine invoke_continuity_arrays
+
+  !=================================================================
+
+  subroutine invoke_continuity_arrays_basic(nx, ny, M, N, rdt, ssha, &
+                                            sshn_t, sshn_u, sshn_v, &
+                                            hu, hv, un, vn, area_t)
+    use kind_params_mod
+    use dl_timer, only: timer_start, timer_stop
+    implicit none
+    integer, intent(in) :: nx, ny, M, N
+    real(wp), intent(in) :: rdt
+    real(wp), intent(out) :: ssha(nx,ny)
+    real(wp), intent(in)  :: sshn_u(nx,ny), sshn_v(nx,ny), sshn_t(nx,ny)
+    real(wp), intent(in)  :: un(nx,ny), vn(nx,ny)
[…]

### 2016-07-12 nemolite2d_psykal/time_step_mod_align_nodiv.f90
nemolite2d: bring no-div version up-to-date with dl-timer API

@@ -6,6 +6,9 @@ module time_step_mod
 
   public invoke_time_step
 
+  integer, save :: timer_momu, timer_momv, timer_cont, &
+                   timer_bcs, timer_next
+  
 contains
 
   subroutine invoke_time_step(istp, ssha, ssha_u, ssha_v, &
@@ -53,7 +56,12 @@ contains
     N  = ssha%grid%simulation_domain%ystop
 
     if(istp .eq. 1)then
-      write (*,"('Loop bounds (inner,outer) = ',I4,1x I4)") M, N
+       write (*,"('Loop bounds (inner,outer) = ',I4,1x I4)") M, N
+       call timer_register(timer_momu,label='Momentum-u')
+       call timer_register(timer_momv,label='Momentum-v')
+       call timer_register(timer_cont,label='Continuity')
+       call timer_register(timer_next,label='Next'      )
+       call timer_register(timer_bcs, label='BCs'       )
     end if
 
     ! In the general case we have to reason about whether or not the
@@ -105,7 +113,7 @@ contains
 !!$                                  un%grid%area_u, &
 !!$                                  un%grid%gphiu)
 
-    call timer_start('Momentum-u',idxt)
+    call timer_start(handle=timer_momu)
     !call likwid_markerStartRegion('Momentum-u')
 
 !    do jj = ua%internal%ystart, ua%internal%ystop, 1
@@ -221,9 +229,9 @@ contains
     end do
  
     !call likwid_markerStopRegion('Momentum-u')
-    call timer_stop(idxt)
+    call timer_stop(timer_momu)
 
-    call timer_start('Momentum-v',idxt)
+    call timer_start(handle=timer_momv)
 
 !dir$ safe_address
     do jj = 2, N-1, 1
@@ -342,11 +350,11 @@ contains
       end do
     end do
 
-    call timer_stop(idxt)
+    call timer_stop(timer_momv)
 
     ! Apply open and solid boundary conditions
 
-    call timer_start('BCs', idxt)
+    call timer_start(handle=timer_bcs)
 
 !    DO jj = ssha%internal%ystart, ssha%internal%ystop 
 !       DO ji = ssha%internal%xstart, ssha%internal%xstop 
@@ -455,11 +463,11 @@ contains
        END DO
     END DO
 
[…]

### 2016-08-01 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-perf: copy intro from shallow paper and add references

@@
 accelerator-based machines in the Top 500 List~\cite{top500} of
-supercomputers: four of the top-ten machines in the November 2015 list
-make use of Intel Xeon Phi's or NVIDIA GPUs. Two of the remaining six
-machines are IBM BlueGene/Qs, the CPU of which has hardware support
-for running 64 threads.
+supercomputers: four of the top-ten machines in the June 2016 list
+make use of many-core processors (Intel Xeon Phi, NVIDIA GPU or NRCPC
+SW26010). Two of the remaining six machines are IBM BlueGene/Qs, the
+CPU of which has hardware support for running 64 threads.
 
@@
 \subsection{The \psykal\ Approach}
+
+The \psykal\ approach attempts to address the problems described in the
+previous section. It separates code into three layers; the Algorithm
+layer, the PSy layer and the Kernel layer. The approach has been
+developed in the GungHo project~\cite{GungHo}, which is creating a new
+Dynamical core for the Met Office, and its design has been influenced
+by earlier work on OP2~\cite{OP2,PYOP2}.
+
+In common with OP2, the \psykal\ approach separates out the science
+code and the performance-related code into distinct layers. The calls
+that specify parallelism in both approaches are similar in terms of
+where they are placed in the code and in their semantics. However, the
+\psykal\ approach supports the specification of more than one kernel
+in a parallel region of code, compared with one for OP2, giving more
+scope for optimisation. Another difference is that the description
+metadata is provided with a kernel in the \psykal\ approach whereas it
+is provided as part of the kernel call in OP2.
+
+While the \psykal\ approach is general, we are currently applying it to
+Atmosphere and Ocean models written in Fortran where domain
+decomposition is typically performed in the latitude-longitude
+direction, leaving columns of elements on each domain-decomposed
+partition.
+
+The top layer, in terms of calling hierarchy, is the Algorithm
+layer. This layer specifies the algorithm that the scientist would like
+to perform (in terms of calls to kernel and infrastructure routines)
+and logically operates on full fields. We say logically here as the
+fields may be domain decomposed, however the Algorithm layer is not
+aware of this. It is the scientist's responsibility to write this
+Algorithm layer.
+
+The bottom layer, in terms of calling hierarchy, is the Kernel
+layer. The Kernel layer implements the science that the Algorithm
+layer calls, as a set of subroutines. These kernels operate on fields
+that are local to the process doing the computation. (Depending on the
+type of kernel, these may be a set of elements, a single column of
+elements, or a set of columns.). Again the scientist is responsible
+for writing this layer and there is no parallelism specified here,
+but, depending on the complexity of the Kernels, there may be input
+from an HPC expert and/or some coding rules to help ensure the kernels
+compile into efficient code.  In an alternative approach, kernels are
+generated from a high-level specification~\cite{firedrake,fenics},
+potentially allowing them to be optimised automatically~\cite{coffee}.
+
+The PSy layer sits in-between the Algorithm and Kernel layers and its
+functional role is to link the algorithm calls to the associated
+kernel subroutines. As the Algorithm layer works on logically global
+fields and Kernel layer works on local fields, the PSy layer is
+responsible for iterating over columns. It is also responsible for
+including any distributed-memory operations resulting from the
+decomposition of the simulation domain, such as halo swaps and
+reductions.
+
+As the PSy layer iterates over columns, the single core performance
+can be optimised by applying transformations such as loop fusion and
+loop blocking. Additionally, the potential parallelism within this
+iteration space can also be exploited and optimised. The PSy layer can
+therefore be tailored for a particular hardware (such as multi-core,
+many-core, GPGPUs, or some combination thereof) and software (such as
+compiler, operating system, MPI implementation) configuration with no
+change to the Algorithm or Kernel layer code. This approach therefore
+offers the potential for portable performance. In this work we apply
+optimisations to the PSy layer manually. The development of a tool to
+automate this process will be the subject of a future paper.
+
+Clearly the separation of code into distinct layers may have an effect
+on performance. This overhead, how to get back to the performance
+of a hand-optimised code, and potentially improve on it, will be
+discussed in the remainder of this paper.
+
+%xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
+
 Proliferation of lightweight cores is resulting in increased

### 2016-08-01 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d: update to intro

@@
 
+In a previous paper~\cite{shallow_psykal} we introduced a possible
+approach to tackling this problem which we term \psykal\ (discussed
+below). In that work we considered the implications for serial
+performance of the extensive code re-structuring required by the
+approach when applied to a shallow-water model. We found that although
+the re-structuring did initially incur a sizeable performance penalty,
+it was possible to transform the resulting code to recover performance
+while obeying the \psykal\ separation of concerns. In this work we
+move to looking at parallel performance within the \psykal\ approach.
+
 % needed in second column of first page if using \IEEEpubid
@@
 
-The \psykal\ approach attempts to address the problems described in the
-previous section. It separates code into three layers; the Algorithm
-layer, the PSy layer and the Kernel layer. The approach has been
-developed in the GungHo project~\cite{GungHo}, which is creating a new
-Dynamical core for the Met Office, and its design has been influenced
-by earlier work on OP2~\cite{OP2,PYOP2}.
+The \psykal\ approach attempts to address the problems described in
+the previous section. It separates code into three layers; the
+Algorithm layer, the PSy layer and the Kernel layer. The approach has
+been developed in the GungHo project~\cite{GungHo}, which is creating
+a new Dynamical core for the UK Met Office, and its design has been
+influenced by earlier work on OP2~\cite{OP2, PYOP2}.
 
@@
 decomposition is typically performed in the latitude-longitude
-direction, leaving columns of elements on each domain-decomposed
+dimension, leaving columns of elements on each domain-decomposed
 partition.
@@
 
-%xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
-
-Proliferation of lightweight cores is resulting in increased
-complexity in the programming models required to achieve performance.
-Combine with the complexity of e.g. a model simulating circulation in
-the global ocean. Significant challenge in software engineering.
-Propose an approach for meeting this challenge by separating the
-computational science demands from the {\it e.g.} oceanographic
-demands.
-
-\begin{figure}
-\centering
-\caption{The \psykal\ Approach.}
-\end{figure}
-
-While encapsulation makes it easier to produce more robust software,
-it can often come with a performance penalty. In this work we
-investigate in detail the hardware and compiler dependencies of this
-penalty and the steps required to remove it.
-
-\subsection{The `nemolite2d' Program}
-
-For this work we have used a program "nemolite2d" developed by
-ourselves.  “nemolite2d” is a vertically-averaged version of NEMO
-(Nucleus for European Modelling of the Ocean~\cite{nemo}), which only keeps its
-dynamical part. The whole model system is represented through one
-continuity equation \eqref{continuity} (for the update of the sea-surface height) and two
-vertically-integrated momentum equations \eqref{momentum} (for the two velocity
-components, respectively).
+\subsection{The `NEMOLite2D' Program}
+
+For this work we have used a program, `NEMOLite2D,' developed by
+ourselves.  NEMOLite2D is a vertically-averaged version of NEMO
+(Nucleus for European Modelling of the Ocean~\cite{nemo}), retaining
+only its dynamical part. The whole model system is represented through
+one continuity equation \eqref{continuity} (for the update of the
+sea-surface height) and two vertically-integrated momentum equations
+\eqref{momentum} (for the two velocity components, respectively).
 
@@
 
-where $\zeta$ and $\vec{U}$ represent the sea surface height and horizontal velocity vector
-respectively. $h$ is the total water depth, $\Omega$ is the Earth rotation velocity vector.
-$g$ is the gravity. $\nu$ is the kinematic viscosity coefficient.
+where $\zeta$ and $\vec{U}$ represent the sea-surface height and
+horizontal velocity vector, respectively. $h$ is the total water
+depth, $\Omega$ is the Earth rotation velocity vector.  $g$ is the
+acceleration due to gravity. $\nu$ is the kinematic viscosity coefficient.
 
 The external forcing includes surface wind stress, bottom friction,
-and open boundary barotropic forcing. A lateral slip boundary
+and open-boundary barotropic forcing. A lateral-slip boundary
 condition is applied along the coast lines. The open boundary
-conditions can be set as a clipped or a Flather's radiation
+condition can be set as a clipped or Flather's radiation
 condition~\cite{flather76}. The bottom friction takes a semi-implicit
@@
 one for ocean cells outside of the computational domain. This array
-serves as a mark to identify ocean cells, land cells, solid boundaries
+enables the identification of ocean cells, land cells, solid boundaries
 and open boundaries.
@@
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal\ version of nemolite2d retains the Input/Output
+we ensure that any \psykal\ version of NEMOLite2D retains the Input/Output
 capability of the original. This aids in limiting the optimisations
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal\ version of nemolite2d, we made the
+optimisation) of the \psykal\ version of NEMOLite2D, we made the
 kernels as fine-grained as possible. This resulted in eight distinct
@@
 mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
-routine then contains a doubly-nested loop around each kernel call, as
-indicated by the pseudo-code in
-Figure~\ref{FIG_psykal_nemolite2d_structure}.
+routine then contains a doubly-nested loop around each kernel call.
 
@@
 \centering
-%\includegraphics[width=85mm]{psykal_nemolite2d}
-\caption{A schematic of the vanilla \psykal\ version of the nemolite2d code.}
+\begin{verbatim}
+DO istep = 1, nsteps
+  call invoke( continuity(ssha_t, ...), &
+               momentum_u(ua, un, ...), &
+               ...,                     &
+               next_sshu(sshn_u, ...),  &
+               next_sshv(sshn_v, ...) )
+  call model_write(istep, sshn_t, un, vn)
+END DO
+\end{verbatim}
+\caption{A schematic of the top-level of the \psykal\ version of the
+ NEMOLite2D code.}
 \label{FIG_psykal_nemolite2d_structure}
@@
 As with any full oceanographic model, boundary conditions must be
-applied at the edges of the model domain. Since nemolite2d applies
+applied at the edges of the model domain. Since NEMOLite2D applies
 external boundary conditions ({\it e.g.} barotropic forcing), this is
@@
 We first describe the code transformations performed for the serial
-version of nemolite2d.  We then move on to the construction of
+version of NEMOLite2D.  We then move on to the construction of
 parallel versions of the code using OpenMP and OpenACC. Again, we
 describe the key steps we have taken in this process in order to
-maximise the performance of the code.
-
-\subsection{Transformations of Serial Nemolite2D}
+maximise the performance of the code. In both cases our aim is to
+identify those transformations which must be supported by any tool
+which seeks to auto-generate a performant PSy layer.
+
+\subsection{Transformations of Serial NEMOLite2D}
 
@@
 \caption{The performance profile of the original and \psykal\ versions
-  of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
+  of NEMOLite2D on the Intel Ivy Bridge CPU (for 2000 time-steps of the
   $128^{2}$ domain and the Intel compiler).}
@@
 diagnostic output produced by either the Cray or Intel compilers. Many
-of the nemolite2d kernels contain conditional statements. These
+of the NEMOLite2D kernels contain conditional statements. These
 statements are there to check whether {\it e.g.} the current grid
@@
 It turns out that the Cray-compiled binaries of both the original and
-\psykal\ versions of nemolite2d perform considerably less well than
+\psykal\ versions of NEMOLite2D perform considerably less well than
 their Intel-compiled counterparts. Comparison of the diagnostic output
@@
 
-\subsection{Construction of OpenMP-Parallel Nemolite2D}
+\subsection{Construction of OpenMP-Parallel NEMOLite2D}
 \label{sec_omp_steps}
@@
 serial code, our purpose here is to determine the functionality
-required of the PSyclone code-generation system.
+required of a tool that seeks to generate the PSy layer.
 
@@
 inserting an OpenMP PARALLEL DO directive before each loop nest so
-that the iterations of the outermost or $j$ loop (over the y-dimension
-of the model domain) are shared out amongst the OpenMP threads. This
-leaves the innermost ($i$) loop available for SIMD vectorisation by
-the compiler.
+that the iterations of the outermost or $j$ loop (over the latitude
+dimension of the model domain) are shared out amongst the OpenMP
+threads. This leaves the innermost ($i$) loop available for SIMD
+vectorisation by the compiler.
 
@@
 condition to the y-component of velocity has a loop-carried dependency
-in $j$ which appears to prevent its being executed in parallel. This was
-therefore left unchanged and executed on thread 0 only.
+in $j$ which appears to prevent its being executed in
+parallel.\footnote{Only once this work was complete did we establish
+  that boundary conditions are enforced such that it can be executed
+  in parallel.}  This was therefore left unchanged and executed on
+thread 0 only.
 
@@
 preceeding loop. Such synchronisation limits performance at larger
-thread counts and, for the nemolite2d code, is frequently unecessary.
+thread counts and, for the NEMOLite2D code, is frequently unecessary.
 {\it E.g.} if a kernel does not make use of the results of a
@@
 Section~\ref{sec_methodology}. Once we have arrived at an optimal form
-for the serial version of nemolite2d we then investigate its parallel
+for the serial version of NEMOLite2D we then investigate its parallel
 performance on both CPU- and GPU-based systems.
@@
 In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
-of the original version of the nemolite2d code for the range of
+of the original version of the NEMOLite2D code for the range of
 compilers and CPUs considered here. Unlike the Shallow
-code~\cite{shallow_psykal}, the original version of nemolite2d has not
+code~\cite{shallow_psykal}, the original version of NEMOLite2D has not
 been optimised. Although it is still a single source file it is, in
@@
 \caption{Summary of the performance of the original version of the 
-nemolite2d code on 
+NEMOLite2D code on 
 the range of compilers and CPU-based systems under consideration.}
@@
 
-Moving now the to the \psykal\ version of nemolite2d,
+Moving now the to the \psykal\ version of NEMOLite2D,
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
@@
 \caption{Summary of the best performance achieved by any \psykal\ 
-version of nemolite2d with each of the compilers and CPUs under 
+version of NEMOLite2D with each of the compilers and CPUs under 
 consideration.}
@@
 between the performance of the original and the best \psykal\ versions
-of nemolite2d for each compiler/CPU combination. This shows that it is
+of NEMOLite2D for each compiler/CPU combination. This shows that it is
 only the Intel-compiled binary running the $64^2$ domain that is
@@
 Having show that we can recover, and often improve upon, the
-performance of the original version of nemolite2d, the next logical
+performance of the original version of NEMOLite2D, the next logical
 step is to examine the necessary code transformations in detail.  We
@@
 \includegraphics[width=93mm]{opt_stages_256}
-\caption{Performance of the \psykal\ version of nemolite2d for the
+\caption{Performance of the \psykal\ version of NEMOLite2D for the
   $256^{2}$ domain at each stage of optimisation. The first (black)
   bar of each cluster gives the performance of the original version of
-  nemolite2d for that compiler/CPU combination.}
+  NEMOLite2D for that compiler/CPU combination.}
 \label{FIG_opt_stages_256}
@@
 We now turn to transformations related to parallelisation of the
-nemolite2d code; the introduction of OpenMP and OpenACC directives. In
+NEMOLite2D code; the introduction of OpenMP and OpenACC directives. In
 keeping with the \psykal\ approach we do not modify either the
@@
 In order to quantify the scaling behaviour of the different versions
-of nemolite2d with the different compilers/run-time environments, we
+of NEMOLite2D with the different compilers/run-time environments, we
 also plot the parallel efficiency in
@@
 Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
-most performant, Intel-compiled version of nemolite2d. Although it is
+most performant, Intel-compiled version of NEMOLite2D. Although it is
 common for production runs of NEMO to use MPI sub-domains of $20
@@
 \caption{The scaling behaviour of the most performant OpenMP-parallel
-  version of \psykal\ nemolite2d for the full range of domain sizes
+  version of \psykal\ NEMOLite2D for the full range of domain sizes
   considered on the CPU. Results are for the Intel compiler on a
@@
 \caption{Performance of the OpenMP-parallel version of \psykal\
-  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
+  NEMOLite2D for the Gnu compiler on a single Intel Ivy Bridge
   socket. The corresponding parallel efficiencies are shown using open
@@
 \caption{Performance of the OpenMP-parallel version of \psykal\
-  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
+  NEMOLite2D for the Intel compiler on a single Intel Ivy Bridge
   socket.  The corresponding parallel efficiencies are shown using
@@
 \caption{Performance of the OpenMP-parallel version of \psykal\
-  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
+  NEMOLite2D for the Cray compiler on a single Intel Ivy Bridge
   socket.  The corresponding parallel efficiencies are shown using
@@
 \caption{Performance of the OpenMP-parallel version of \psykal\
-  nemolite2d on one and two sockets of Intel Ivy Bridge. The 24-thread
+  NEMOLite2D on one and two sockets of Intel Ivy Bridge. The 24-thread
   runs on a single socket used hyperthreading and the two-socket runs
@@
 The introduction of OpenACC directives proved to be simpler than
-producing the OpenMP version of nemolite2d {\bf - Jeremy to clarify
+producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
   whether this is true}.  Consequently we only show the performance of
@@
 \includegraphics[width=\picwidth]{gpu_opt_stages}
-\caption{Performance of the various GPU implementations of nemolite2d.}
+\caption{Performance of the various GPU implementations of NEMOLite2D.}
 \label{FIG_gpu_opt_stages}
@@
 In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
-the OpenMP and OpenACC implementations of nemolite2d across the range
+the OpenMP and OpenACC implementations of NEMOLite2D across the range
 of problem sizes considered. The OpenMP figures are the maximum
@@
 \caption{Performance of the best OpenMP-parallel version of \psykal\
-  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
+  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
 \label{FIG_cpu_cf_gpu}
@@
 Although we have investigated how the performance of the \psykal
-version of nemolite2d compares with that or the original, we have not
+version of NEMOLite2D compares with that or the original, we have not
 addressed how efficient the original actually is. In order to do so we

### 2016-08-19 nemolite2d_perf.tex
shallow-perf: tidying of existing text

@@
 
+In a previous paper~\cite{shallow_psykal} we introduced a possible
+approach to tackling this problem which we term \psykal\ (discussed
+below). In that work we considered the implications for serial
+performance of the extensive code re-structuring required by the
+approach when applied to the `Shallow' shallow-water model. We found
+that although the re-structuring did initially incur a sizeable
+performance penalty, it was possible to transform the resulting code
+to recover performance while obeying the \psykal\ separation of
+concerns. In this work we move to looking at parallel performance
+within the \psykal\ approach.
+
 % needed in second column of first page if using \IEEEpubid
@@
 \subsection{The \psykal\ Approach}
-Proliferation of lightweight cores is resulting in increased
-complexity in the programming models required to achieve performance.
-Combine with the complexity of e.g. a model simulating circulation in
-the global ocean. Significant challenge in software engineering.
-Propose an approach for meeting this challenge by separating the
-computational science demands from the {\it e.g.} oceanographic
-demands.
-
-\begin{figure}
-\centering
-\caption{The \psykal\ Approach.}
-\end{figure}
-
-While encapsulation makes it easier to produce more robust software,
-it can often come with a performance penalty. In this work we
-investigate in detail the hardware and compiler dependencies of this
-penalty and the steps required to remove it.
-
-\subsection{The `nemolite2d' Program}
-
-For this work we have used a program "nemolite2d" developed by
-ourselves.  “nemolite2d” is a vertically-averaged version of NEMO
-(Nucleus for European Modelling of the Ocean~\cite{nemo}), which only keeps its
-dynamical part. The whole model system is represented through one
-continuity equation \eqref{continuity} (for the update of the sea-surface height) and two
-vertically-integrated momentum equations \eqref{momentum} (for the two velocity
-components, respectively).
+
+The \psykal\ approach attempts to address the problems described in
+the previous section. It separates code into three layers; the
+Algorithm layer, the PSy layer and the Kernel layer. The approach has
+been developed in the GungHo project~\cite{GungHo}, which is creating
+a new Dynamical core for the UK Met Office, and its design has been
+influenced by earlier work on OP2~\cite{OP2, PYOP2}.
+
+In common with OP2, the \psykal\ approach separates out the science
+code and the performance-related code into distinct layers. The calls
+that specify parallelism in both approaches are similar in terms of
+where they are placed in the code and in their semantics. However, the
+\psykal\ approach supports the specification of more than one kernel
+in a parallel region of code, compared with one for OP2, giving more
+scope for optimisation. In addition, the metadata describing a kernel
+is included with the kernel code in the \psykal\ approach whereas it
+is provided as a part of the kernel call in OP2.
+
+While the \psykal\ approach is general, we are currently applying it to
+Atmosphere and Ocean models written in Fortran where domain
+decomposition is typically performed in the latitude-longitude
+dimension, leaving columns of elements on each domain-decomposed
+partition.
+
+The top layer, in terms of calling hierarchy, is the Algorithm
+layer. This layer specifies the algorithm that the scientist would like
+to perform (in terms of calls to kernel and infrastructure routines)
+and logically operates on full fields. We say logically here as the
+fields may be domain decomposed, however the Algorithm layer is not
+aware of this. It is the scientist's responsibility to write this
+Algorithm layer.
+
+The bottom layer, in terms of calling hierarchy, is the Kernel
+layer. The Kernel layer implements the science that the Algorithm
+layer calls, as a set of subroutines. These kernels operate on fields
+that are local to the process doing the computation. (Depending on the
+type of kernel, these may be a set of elements, a single column of
+elements, or a set of columns.) Again the scientist is responsible for
+writing this layer and there is no parallelism specified here, but,
+depending on the complexity of the Kernels, there may be input from an
+HPC expert and/or some coding rules to help ensure that the kernels
+compile into efficient code.  In an alternative
+approach~\cite{firedrake,fenics}, kernels are generated from a
+high-level specification, potentially allowing them to be optimised
+automatically~\cite{coffee}.
+
+The PSy layer sits in-between the Algorithm and Kernel layers and its
+functional role is to link the algorithm calls to the associated
+kernel subroutines. As the Algorithm layer works on logically global
+fields and Kernel layer works on local fields, the PSy layer is
+responsible for iterating over columns. It is also responsible for
+including any distributed-memory operations resulting from the
+decomposition of the simulation domain, such as halo swaps and
+reductions.
+
+As the PSy layer iterates over columns, the single-core performance
+can be optimised by applying transformations such as loop fusion and
+loop blocking. Additionally, the potential parallelism within this
+iteration space can also be exploited and optimised. The PSy layer can
+therefore be tailored for a particular hardware (such as multi-core,
+many-core, GPGPUs, or some combination thereof) and software (such as
+compiler, operating system, MPI implementation) configuration with no
+change to the Algorithm or Kernel layer code. This approach therefore
+offers the potential for portable performance. In this work we apply
+optimisations to the PSy layer manually. The development of a tool to
+automate this process will be the subject of a future paper.
+
+Clearly the separation of code into distinct layers may have an effect
+on performance. This overhead, how to get back to the performance of a
+parallel, hand-optimised code, and potentially improve on it, will be
+discussed in the remainder of this paper.
+
+\subsection{The `NEMOLite2D' Program}
+
+For this work we have used a program, `NEMOLite2D,' developed by
+ourselves.  NEMOLite2D is a vertically-averaged version of NEMO
+(Nucleus for European Modelling of the Ocean~\cite{nemo}), retaining
+only its dynamical part. The whole model system is represented through
+one continuity equation \eqref{continuity} (for the update of the
+sea-surface height) and two vertically-integrated momentum equations
+\eqref{momentum} (for the two velocity components, respectively).
 
@@
 
-The order of the model computation is as follows:
+The sequence of the model computation is as follows:
 \begin{enumerate}
@@
 mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
-routine then contains a doubly-nested loop around each kernel call, as
-indicated by the pseudo-code in
-Figure~\ref{FIG_psykal_nemolite2d_structure}.
+routine then contains a doubly-nested loop (over the two dimensions of
+the model grid) around each kernel call.
 
@@
 are relatively large; each comprises roughly 85 lines of Fortran
-(excluding variable declarations).
+executable statements.
 
@@
 inserting an OpenMP PARALLEL DO directive before each loop nest so
-that the iterations of the outermost or $j$ loop (over the y-dimension
-of the model domain) are shared out amongst the OpenMP threads. This
-leaves the innermost ($i$) loop available for SIMD vectorisation by
-the compiler.
+that the iterations of the outermost or $j$ loop (over the latitude
+dimension of the model domain) are shared out amongst the OpenMP
+threads. This leaves the innermost ($i$) loop available for SIMD
+vectorisation by the compiler.
 
 The loop nest dealing with the application of the Flather boundary
-condition to the y-component of velocity has a loop-carried dependency
-in $j$ which appears to prevent its being executed in parallel. This was
-therefore left unchanged and executed on thread 0 only.
+condition to the y-component of velocity ($v$) has a loop-carried
+dependency in $j$ which appears to prevent its being executed in
+parallel.\footnote{Only once this work was complete did we establish
+  that boundary conditions are enforced such that it can safely be
+  executed in parallel.}  This was therefore left unchanged and
+executed on thread 0 only.
 
@@
 each loop nest to be parallelised was then changed to an OpenMP DO. We
-ensured that the v-Flather loop nest was executed in serial (by the
+ensured that the $v$-Flather loop nest was executed in serial (by the
 first thread to encounter it) by enclosing it within an OpenMP SINGLE
@@
 Since data arrays are managed within the GOcean Library this
-optimisation can again be implemented without changing user code ({\it
-  i.e.} the Application and Kernel layers).
+optimisation can again be implemented without changing the
+natural-science code ({\it i.e.} the Application and Kernel layers).
 
@@
 
-As previously mentioned, the v-Flather section has to be executed in
+As previously mentioned, the $v$-Flather section was executed in
 serial becuase of a loop-carried dependence in $j$. (In principle we
@@
 threads. We attempt to mitigate this by moving this serial section to
-before the (parallel) u-Flather section. Since these two sections are
+before the (parallel) $u$-Flather section. Since these two sections are
 independent, the aim is that the thread that performs the serial,
-v-Flather computation then performs a smaller share of the following
-u-Flather loop. In practice, this requires that some form of dynamic
+$v$-Flather computation then performs a smaller share of the following
+$u$-Flather loop. In practice, this requires that some form of dynamic
 thread scheduling is used.
@@
 set at run-time using the OMP\_SCHEDULE environment variable. We
-experimented with using the standard static, dynamic and guided (with varying
-chunk size) OpenMP schedules.
+experimented with using the standard static, dynamic and guided (with
+varying chunk size) OpenMP schedules.
 
@@
 Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
-detailed performance figures for this case after each transformation has been
-applied to the code. The same data is visualised in
-Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal\
-version for a given compiler/CPU is given relative to the performance
-of the original with the same compiler/CPU.
+detailed performance figures for this case after each transformation
+has been applied to the code. The same data is visualised in
+Figure~\ref{FIG_opt_stages_256}.
 
@@
 Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
-most performant, Intel-compiled version of nemolite2d. Although it is
+most performant, Intel-compiled version of NEMOLite2D. Although it is
 common for production runs of NEMO to use MPI sub-domains of $20
-\times 20$ ({\bf do we have/need a reference for that?}), it is clear
-from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
-parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
-to fully test the performance of our OpenMP implementations we have
-chosen to examine results for the $256 \times 256$ domain and these
-are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
-and \ref{FIG_cray_omp_256}.
+\times 20$, it is clear from Figure~\ref{FIG_omp_domain_size} that the
+limited quantity of parallelism in the $32 \times 32$ domain inhibits
+scaling. Therefore, to fully test the performance of our OpenMP
+implementations we have chosen to examine results for the $256 \times
+256$ domain and these are shown in Figures~\ref{FIG_gnu_omp_256},
+\ref{FIG_intel_omp_256} and \ref{FIG_cray_omp_256}.
 
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_gnu}
-\caption{Performance of the OpenMP-parallel version of \psykal\
-  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-  socket. The corresponding parallel efficiencies are shown using open
-  symbols and dashed lines. The 24-thread runs employed hyperthreading
-  and the optimal OpenMP schedule is given in parentheses.}
+\caption{Performance of the OpenMP-parallel version of
+  \psykal\ NEMOLite2D for the $256^2$ domain with the Gnu compiler on
+  a single Intel Ivy Bridge socket. The corresponding parallel
+  efficiencies are shown using open symbols and dashed lines. The
+  24-thread runs employed hyperthreading and the optimal OpenMP
+  schedule is given in parentheses.}
 \label{FIG_gnu_omp_256}
@@
 threads. The Cray version however does better and is about 45\%
-efficient on eight threads.
+efficient on eight threads (Figure~\ref{FIG_cray_omp_256}).
 
@@
 OpenMP compiler flag has on the optimisations performed by the
-compiler.  In particular, the Cray compiler no-longer vectorises the
+compiler.  In particular, the Cray compiler no longer vectorises the
 key Momentum section, despite the directive added during the serial
-optimisation work.
+optimisation work. This defficiency has been reported to Cray.
 
@@
 results to this quantity by benchmarking using two sockets of Intel
-Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
-configuration we ensured that threads were evenly shared over the two
-sockets. The performance obtained for the $256^2$ case with the `early
-SINGLE' version of the code is compared with the single-socket
-performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
-available memory bandwidth in this way has little effect on
-performance - the two-socket performance figures track those from a
-single socket very closely. The only significant difference in performance
-is at 24 threads where, in addition to the difference in available
-memory bandwidth, the single-socket configuration is using
-hyperthreading while the two-socket case is not.  {\bf Have query in
-  with Archer helpdesk on behaviour seen with Cray compiler here. Also
-  would like to run 512 case on 2 sockets to see how it goes out to 48
-  threads.}
+Ivy Bridge ({\it i.e.} using a complete node of ARCHER, a Cray
+XC30). For this configuration we ensured that threads were evenly
+shared over the two sockets. The performance obtained for the $256^2$
+case with the `early SINGLE' version of the code is compared with the
+single-socket performance in
+Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the available
+memory bandwidth in this way has little effect on performance - the
+two-socket performance figures track those from a single socket very
+closely. The only significant difference in performance is at 24
+threads where, in addition to the difference in available memory
+bandwidth, the single-socket configuration is using hyperthreading
+while the two-socket case is not.  The discrepancy in the performance
+of the Cray- and Intel-compiled binaries at this thread count is under
+investigation by Cray.
 
@@
 
-The introduction of OpenACC directives proved to be simpler than
-producing the OpenMP version of nemolite2d {\bf - Jeremy to clarify
+The advantage of the \psykal\ re-structuring becomes apparent if we
+wish to run NEMOLite2D on different hardware, \textit{e.g.} a
+GPU. This is because the necessary code modifications are, by design,
+limited to the middle PSy layer. In order to port the Fortran
+NEMOLite2D to GPU we used OpenACC directives as opposed to the OpenMP
+directives used for the multi-core CPU.
+
+The introduction of these directives proved to be simpler than
+producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
   whether this is true}.  Consequently we only show the performance of
@@
 
-Although we have investigated how the performance of the \psykal
-version of nemolite2d compares with that or the original, we have not
+Although we have investigated how the performance of the \psykal\
+version of NEMOLite2D compares with that or the original, we have not
 addressed how efficient the original actually is. In order to do so we
@@
 \includegraphics[width=\picwidth]{roofline}
-\caption{Comparison of the performance achieved by NEMOLite2D and
-  Shallow on the E5-1620 CPU. Results for the former are for the
-  $256^2$ domain since that gave the best performance.}
+\caption{Comparison of the performance achieved by kernels from
+  NEMOLite2D and Shallow on a roof-line plot for the E5-1620
+  CPU. Results for the former are for the $256^2$ domain since that
+  gave the best performance. See the text for a discussion of the
+  different CPU ceilings (dashed lines).}
 \label{FIG_roofline}
@@
 In Figure~\ref{FIG_roofline} we show the performance of kernels from
-both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
+both Shallow and NEMOLite2D on the roofline model XSfor the E5-1620 CPU.
 This demonstrates that the Shallow kernel is achieving a performance
@@
 NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
-(i.e. assuming that all nodes in the DAG are executed in serial) we
-get a lower-bound performance estimate of $0.6391 \times CLOCK\_SPEED$
-FLOPS which gives 2.46~GFLOPS at a clock speed of
+(ILP), \textit{i.e.} assuming that all nodes in the DAG are executed
+in serial, we get a lower-bound performance estimate of $0.6391 \times
+CLOCK\_SPEED$ FLOPS which gives 2.46~GFLOPS at a clock speed of
 3.85~GHz.
@@
 to different execution ports are always run in parallel. In the Ivy
-Bridge core floating-point multiplication and division operations go
+Bridge core, floating-point multiplication and division operations go
 to port 0 while addition and subtraction go to port
@@
 
-These performance estimates are plotted as CPU ceilings in
-Figure~\ref{FIG_roofline}. The performance of the Momentum kernel is
-seen to fall between these two bounds which demonstrates that its
-performance is not memory-bandwidth limited, as might have been
-assumed by its low AI. Therefore, although the performance of this
-kernel is clearly sub-optimal, this is due to the balance of
-floating-point operations that it contains. Improving this is outside
-the scope of the current work since we are focused on the \psykal
-separation of concerns and the introduction of parallelism in the PSy
-layer.
+These performance estimates are plotted as CPU ceilings (horizontal
+dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
+Momentum kernel is seen to fall between these two bounds which
+demonstrates that its performance is not memory-bandwidth limited, as
+might have been assumed by its low AI. Therefore, although the
+performance of this kernel is clearly sub-optimal, this is due to the
+balance of floating-point operations that it contains. Improving this
+is outside the scope of the current work since here we are focused on
+the \psykal\ separation of concerns and the introduction of parallelism
+in the PSy layer.
+
+Enabling SIMD vectorisation for this kernel does not significantly
+improve its performance (Figure~\ref{FIG_roofline}) and in fact,
+limiting it to SSE instructions (vector width of two double-precision
+floating point numbers) rather than AVX (vector width of four) was
+found to produce a slightly more performant version. The poor
+efficiency of the SIMD version is highlighted by the fact that the
+corresponding kernel performance no longer falls within the bounds of
+our performance estimate. This is because we have incorporated the
+effect of SSE into that estimate by simply assuming perfect
+vectorisation which gives a performance increase of a factor of two.
 

### 2016-08-19 nemolite2d_perf.tex
nemolite-perf: added some conclusions!

@@
   whether this is true}.  Consequently we only show the performance of
-{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.
+{\bf xxx} versions of the code in
+Figure~\ref{FIG_gpu_opt_stages}. {\bf discuss performance drop in
+  OpenACC version c.f. CUDA.}
 
@@
 concerns approach to the domain of shared-memory parallel,
-finite-difference ocean models. This approach enables the
+finite-difference shallow-water models. This approach enables the
 computational science (performance) related aspects of a computer
@@
 introduction of the \psykal\ separation of concerns followed by
-suitable transformations of the PSy Layer is found to improve
+suitable transformations of the PSy Layer is actually found to improve
 performance. This is in contrast to our previous
@@
 has had been optimised over the course of several years. In that case
-we were able to recover\ldots
+we were able to recover more than 90\% of the performance of the
+original for the Gnu, Intel, Cray and IBM compilers on both Intel and
+IBM Power CPUs.
 
 The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal\ version of a code.
+to the performance of the \psykal\ version of a code. For both
+NEMLite2D and Shallow we have found that for serial performance, the
+most important transformation is that of in-lining the kernel source
+at the call site. However, for the more complex NEMOLite2D code the
+Cray compiler also had to be coerced into performing SIMD
+vectorisation through the use of souce-code directives.
+
+In this work we have also demonstrated the introduction of parallelism
+into the PSy layer with both OpenMP and OpenACC directives. In both
+cases we were able to leave the natural-science parts of the code
+unchanged. For OpenMP we achieved a parallel efficiency of $\sim70\%$
+on 12 threads by enclosing the body of the (single) PSy layer routine
+within a single PARALLEL region. Removal of unecessary synchronisation
+points through use of the NOWAIT clause boosted 12-thread performance
+by approximately 10\% with the Gnu and Cray compilers. The
+\psykal\ re-structuring of this (admittedly small) code was not found
+to pose any problems for the introduction of performant OpenMP.
+
+{\bf Some words about how good the OpenACC, \psykal\ version is compared
+to the CUDA version.}
+
+Having demonstrated that the \psykal\ separation of concerns may be
+applied to finite-difference codes with minimal performance loss, our
+next step will be to consider the automatic generation of the PSy
+layer. In future work we will analyse the performance of a
+domain-specific compiler that performs this task. This compiler
+(which we have named `PSyclone') is currently under development.
 

### 2016-08-19 nemolite2d_perf.tex
nemolite2d-perf: added an abstract

@@
 
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
-The abstract goes here.
+
+We present an approach which we call \psykal\ that aims to achieve
+portable performance for parallel, finite-difference Ocean models.  In
+\psykal\ the code related to the underlying science is formally
+separated from code related to parallelisation and single-core
+optimisations. This separation of concerns allows scientists to code
+their science independently of the underlying hardware architecture
+and for optimisation specialists to be able to tailor the code for a
+particular machine independently of the science code. We have taken
+the free-surface part of the NEMO ocean model and created a new,
+shallow-water model named NEMOLite2D. In doing this we have a code
+which is of a manageable size and yet which incorporates elements of
+full ocean models (input/output, boundary conditions, \textit{etc.}).
+We have then manually constructed a \psykal\ version of this code and
+investigated the transformations that must be applied to the
+middle/PSy layer in order to achieve good performance, both serial and
+parallel. We have produced versions of the PSy layer parallelised with
+both OpenMP and OpenACC; in both cases we were able to leave the
+natural-science parts of the code unchanged while achieving good
+performance on both multi-core CPUs and GPUs. In quantifying whether
+or not the obtained performance is `good' we also consider the
+limitations of the basic roofline model and attempt to improve on it
+by generating kernel-specific CPU ceilings.
+
 \end{abstract}

### 2016-08-31 nemolite2d_perf.tex
nemolite2d-perf: remove hardware table and put details into text

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Something about Performance, 2D Component of NEMO. (TBC)}
+\title{Portable Multi- and Many-Core Performance for Finite Difference Codes;
+  Application to the Free-Surface Component of NEMO.}
 %
@@
 
-Our aim in this work is to achieve portable performance. Consequently,
-we have performed tests with the hardware and compilers listed in
-Table~\ref{TABLE_compilers}. Where a compiler is available on a given
-piece of hardware, the version number used in this work is specified.
-%The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
-%3.7~GHz and 10~MB of last-level cache. 
-The Intel Ivy Bridge-based
-system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB.
+Our aim in this work is to achieve portable performance, especially
+between multi-core CPU and many-core GPU systems. Consequently, we
+have performed tests on both an Intel Ivy Bridge CPU (E5-2697 at
+2.7~GHz) and on an NVIDIA Tesla K40 GPU.  On the Intel-based system we
+have used the Gnu, Intel and Cray Fortran compilers (versions 4.9.1,
+15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
+was compiled using version 15.10 of the PGI compiler.
 
@@
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-%
-\begin{table}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-\caption{The matrix of compilers and hardware used in this work. The
-  use of a compiler on a given system is indicated by the specification
-  of the version of the compiler in the relevent element. No entry
-  indicates that a compiler was not available/used on that system.}
-\label{TABLE_compilers}
-\centering
-\begin{tabular}{|l|c|c|c|c|c|}
-\hline
-                 & \multicolumn{5}{c|}{Compiler}            \\
-\hline
-                 & Gnu   & Intel       & Cray    & IBM    &  PGI \\
-\hline
-%Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
-Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
-IBM Power 8      &       &             &         &        &  \\
-NVIDIA Tesla K40 &       &             &         &        &  {\bf Jeremy} \\
-\hline
-\end{tabular}
-\end{table}
-
 In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
-used with each compiler. For the IBM, Gnu and Intel compilers, we
+used with each compiler. For the Gnu and Intel compilers, we
 include flags to encourage in-lining of kernel bodies. The Intel flag
@@
 Cray      & -O3 -O ipa5 -h wp                         &   \\
-IBM       & -O5 -qinline=auto:level=10                &   \\
-PGI       & {\bf Jeremy to supply}                    &   \\
+PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 &   \\
 \hline

### 2016-08-31 nemolite2d_perf.tex
nemolite2d-perf: add para on habakkuk and OpenMP to conclusions

@@
 performance penalty, it was possible to transform the resulting code
-to recover performance while obeying the \psykal\ separation of
-concerns. In this work we move to looking at parallel performance
-within the \psykal\ approach.
+to recover performance (for a variety of CPU/compiler combinations)
+while obeying the \psykal\ separation of concerns. In this work we
+move to looking at portable parallel performance within the
+\psykal\ approach.
 
@@
 NEMOLite2D to GPU we used OpenACC directives as opposed to the OpenMP
-directives used for the multi-core CPU.
+directives used for the multi-core CPU. We also experimented with
+using CUDA directly in order to quantify any performance penalty
+incurred.
 
@@
 In Figure~\ref{FIG_roofline} we show the performance of kernels from
-both Shallow and NEMOLite2D on the roofline model XSfor the E5-1620 CPU.
+both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
 This demonstrates that the Shallow kernel is achieving a performance
@@
 
+Investigation of the absolute serial performance of the NEMLite2D code
+using the roofline model revealed that it was still significantly
+below any of the traditional roofline ceilings. We have developed a
+code-analysis tool~\cite{habakkuk} that is capable of providing more
+accurate ceilings by analysing the nature of the floating point
+computations performed by a kernel. In future work we aim to extend
+this tool such that it can be applied to code parallelised using
+OpenMP.
+
 The application of code transformations to the middle/PSy layer is key

### 2016-09-01 nemolite2d_perf.tex
nemolite-perf: update GPU-perf plots and captions so that we use psykal GPU numbers where appropriate

@@
 15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
-was compiled using version 15.10 of the PGI compiler.
+was compiled using version 15.5 of the PGI compiler.
 
@@
 \includegraphics[width=\picwidth]{gpu_opt_stages}
-\caption{Performance of the various GPU implementations of NEMOLite2D.}
+\caption{Performance of the \psykal\ (OpenACC) and CUDA GPU
+  implementations of NEMOLite2D. The performance for the OpenACC
+  version of the original code is not shown since it is identical to
+  that of the \psykal\ version.}
 \label{FIG_gpu_opt_stages}
@@
 \caption{Performance of the best OpenMP-parallel version of \psykal\
-  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
+  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the \psykal\ GPU implementation (using OpenACC).}
 \label{FIG_cpu_cf_gpu}
@@
 Having demonstrated that the \psykal\ separation of concerns may be
-applied to finite-difference codes with minimal performance loss, our
-next step will be to consider the automatic generation of the PSy
-layer. In future work we will analyse the performance of a
-domain-specific compiler that performs this task. This compiler
-(which we have named `PSyclone') is currently under development.
+applied to 2D finite-difference codes with minimal performance loss,
+our next steps will be; 1) to consider the automatic generation of the
+PSy layer and 2) to look at extending the approach to the full NEMO
+model (i.e. three dimensions). In future work we will analyse the
+performance of a domain-specific compiler that performs the automatic
+generation of the PSy layer. This compiler (which we have named
+`PSyclone') is currently under development.
 

### 2016-09-01 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-perf: tidy-up discussion of GPU results.

@@
 performance of the extensive code re-structuring required by the
-approach when applied to a shallow-water model. We found that although
-the re-structuring did initially incur a sizeable performance penalty,
-it was possible to transform the resulting code to recover performance
+approach when applied to the `Shallow' shallow-water model. We found
+that although the re-structuring did initially incur a sizeable
+performance penalty, it was possible to transform the resulting code
+to recover performance (for a variety of CPU/compiler combinations)
 while obeying the \psykal\ separation of concerns. In this work we
-move to looking at parallel performance within the \psykal\ approach.
+move to looking at portable parallel performance within the
+\psykal\ approach.
 
@@
 in a parallel region of code, compared with one for OP2, giving more
-scope for optimisation. Another difference is that the description
-metadata is provided with a kernel in the \psykal\ approach whereas it
-is provided as part of the kernel call in OP2.
+scope for optimisation. In addition, the metadata describing a kernel
+is included with the kernel code in the \psykal\ approach whereas it
+is provided as a part of the kernel call in OP2.
 
@@
 type of kernel, these may be a set of elements, a single column of
-elements, or a set of columns.). Again the scientist is responsible
-for writing this layer and there is no parallelism specified here,
-but, depending on the complexity of the Kernels, there may be input
-from an HPC expert and/or some coding rules to help ensure the kernels
-compile into efficient code.  In an alternative approach, kernels are
-generated from a high-level specification~\cite{firedrake,fenics},
-potentially allowing them to be optimised automatically~\cite{coffee}.
+elements, or a set of columns.) Again the scientist is responsible for
+writing this layer and there is no parallelism specified here, but,
+depending on the complexity of the Kernels, there may be input from an
+HPC expert and/or some coding rules to help ensure that the kernels
+compile into efficient code.  In an alternative
+approach~\cite{firedrake,fenics}, kernels are generated from a
+high-level specification, potentially allowing them to be optimised
+automatically~\cite{coffee}.
 
@@
 
-As the PSy layer iterates over columns, the single core performance
+As the PSy layer iterates over columns, the single-core performance
 can be optimised by applying transformations such as loop fusion and
@@
 Clearly the separation of code into distinct layers may have an effect
-on performance. This overhead, how to get back to the performance
-of a hand-optimised code, and potentially improve on it, will be
+on performance. This overhead, how to get back to the performance of a
+parallel, hand-optimised code, and potentially improve on it, will be
 discussed in the remainder of this paper.
@@
 mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
-routine then contains a doubly-nested loop around each kernel call.
+routine then contains a doubly-nested loop (over the two dimensions of
+the model grid) around each kernel call.
 
@@
 
-Our aim in this work is to achieve portable performance. Consequently,
-we have performed tests with the hardware and compilers listed in
-Table~\ref{TABLE_compilers}. Where a compiler is available on a given
-piece of hardware, the version number used in this work is specified.
-%The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
-%3.7~GHz and 10~MB of last-level cache. 
-The Intel Ivy Bridge-based
-system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
-cache of 30~MB.
+Our aim in this work is to achieve portable performance, especially
+between multi-core CPU and many-core GPU systems. Consequently, we
+have performed tests on both an Intel Ivy Bridge CPU (E5-2697 at
+2.7~GHz) and on an NVIDIA Tesla K40 GPU.  On the Intel-based system we
+have used the Gnu, Intel and Cray Fortran compilers (versions 4.9.1,
+15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
+was compiled using version 15.5 of the PGI compiler.
 
@@
 Cray      & -O3 -O ipa5 -h wp                         &   \\
-IBM       & -O5 -qinline=auto:level=10                &   \\
-PGI       & {\bf Jeremy to supply}                    &   \\
+PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 & \\
 \hline
@@
 The loop nest dealing with the application of the Flather boundary
-condition to the y-component of velocity has a loop-carried dependency
-in $j$ which appears to prevent its being executed in
+condition to the y-component of velocity ($v$) has a loop-carried
+dependency in $j$ which appears to prevent its being executed in
 parallel.\footnote{Only once this work was complete did we establish
-  that boundary conditions are enforced such that it can be executed
-  in parallel.}  This was therefore left unchanged and executed on
-thread 0 only.
+  that boundary conditions are enforced such that it can safely be
+  executed in parallel.}  This was therefore left unchanged and
+executed on thread 0 only.
 
@@
 common for production runs of NEMO to use MPI sub-domains of $20
-\times 20$ ({\bf do we have/need a reference for that?}), it is clear
-from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
-parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
-to fully test the performance of our OpenMP implementations we have
-chosen to examine results for the $256 \times 256$ domain and these
-are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
-and \ref{FIG_cray_omp_256}.
+\times 20$, it is clear from Figure~\ref{FIG_omp_domain_size} that the
+limited quantity of parallelism in the $32 \times 32$ domain inhibits
+scaling. Therefore, to fully test the performance of our OpenMP
+implementations we have chosen to examine results for the $256 \times
+256$ domain and these are shown in Figures~\ref{FIG_gnu_omp_256},
+\ref{FIG_intel_omp_256} and \ref{FIG_cray_omp_256}.
 
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_gnu}
-\caption{Performance of the OpenMP-parallel version of \psykal\
-  NEMOLite2D for the Gnu compiler on a single Intel Ivy Bridge
-  socket. The corresponding parallel efficiencies are shown using open
-  symbols and dashed lines. The 24-thread runs employed hyperthreading
-  and the optimal OpenMP schedule is given in parentheses.}
+\caption{Performance of the OpenMP-parallel version of
+  \psykal\ NEMOLite2D for the $256^2$ domain with the Gnu compiler on
+  a single Intel Ivy Bridge socket. The corresponding parallel
+  efficiencies are shown using open symbols and dashed lines. The
+  24-thread runs employed hyperthreading and the optimal OpenMP
+  schedule is given in parentheses.}
 \label{FIG_gnu_omp_256}
@@
 results to this quantity by benchmarking using two sockets of Intel
-Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
-configuration we ensured that threads were evenly shared over the two
-sockets. The performance obtained for the $256^2$ case with the `early
-SINGLE' version of the code is compared with the single-socket
-performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
-available memory bandwidth in this way has little effect on
-performance - the two-socket performance figures track those from a
-single socket very closely. The only significant difference in
-performance is at 24 threads where, in addition to the difference in
-available memory bandwidth, the single-socket configuration is using
-hyperthreading while the two-socket case is not.  The discrepancy in
-the performance of the Cray- and Intel-compiled binaries at this
-thread count is under investigation by Cray.
+Ivy Bridge ({\it i.e.} using a complete node of ARCHER, a Cray
+XC30). For this configuration we ensured that threads were evenly
+shared over the two sockets. The performance obtained for the $256^2$
+case with the `early SINGLE' version of the code is compared with the
+single-socket performance in
+Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the available
+memory bandwidth in this way has little effect on performance - the
+two-socket performance figures track those from a single socket very
+closely. The only significant difference in performance is at 24
+threads where, in addition to the difference in available memory
+bandwidth, the single-socket configuration is using hyperthreading
+while the two-socket case is not.  The discrepancy in the performance
+of the Cray- and Intel-compiled binaries at this thread count is under
+investigation by Cray.
 
@@
 
-The introduction of OpenACC directives proved to be simpler than
-producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
-  whether this is true}.  Consequently we only show the performance of
-{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.
+The advantage of the \psykal\ re-structuring becomes apparent if we
+wish to run NEMOLite2D on different hardware, \textit{e.g.} a
+GPU. This is because the necessary code modifications are, by design,
+limited to the middle PSy layer. In order to demonstrate this and to
+check for any limitations imposed by the \psykal\ re-structuring, we
+had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
+used OpenACC directives for this porting as that approach is similar
+to the use of OpenMP directives and works well within the \psykal
+apprach. He also experimented with using CUDA directly within the
+original form of NEMOLite2D in order to quantify any performance
+penalty incurred by taking the \psykal/OpenACC approach.
+
+The performance of these two versions of the code is plotted in
+Figure~\ref{FIG_gpu_opt_stages}.  Since the \psykal restructuring of
+NEMOLite2D did not prove to be a barrier to achieving a performant
+code when using OpenACC directives, we only report the performance of
+the \psykal\ OpenACC version. For problem sizes up to $256^2$ the
+difference in the performance of the two versions is less than 5\% and
+for the larger domains it is at most 11\%. This demonstrates that the
+performance cost of the \psykal\ approach in utilising a GPU for
+NEMOLite2D is minimal.
 
@@
 memory. At this stage the OpenACC version of the code on the GPU is
-some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
-than the OpenMP version on the CPU. {\bf Can we relate this to the
-  difference in peak memory b/w between the GPU and CPU?}
+some 3.4 times faster than the best OpenMP version on the CPU.
+
+%Peak memory bandwidth on the K40 GPU is 288 GB/s and we measured
+%(using STREAM2) bandwidths of 46.0 and 20.5 GB/s to L3 and main
+%memory, respectively on the CPU.
 
@@
 
-Although we have investigated how the performance of the \psykal
+Although we have investigated how the performance of the \psykal\
 version of NEMOLite2D compares with that or the original, we have not
@@
 In Figure~\ref{FIG_roofline} we show the performance of kernels from
-both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
+both Shallow and NEMOLite2D on the roofline model for an Intel E5-1620 CPU.
 This demonstrates that the Shallow kernel is achieving a performance
@@
 is outside the scope of the current work since here we are focused on
-the \psykal separation of concerns and the introduction of parallelism
+the \psykal\ separation of concerns and the introduction of parallelism
 in the PSy layer.
@@
 our performance estimate. This is because we have incorporated the
-effect of SSE by assuming perfect vectorisation which gives a
-performance increase of a factor of two.
+effect of SSE into that estimate by simply assuming perfect
+vectorisation which gives a performance increase of a factor of two.
 
@@
 has had been optimised over the course of several years. In that case
-we were able to recover\ldots
+we were able to recover more than 90\% of the performance of the
+original for the Gnu, Intel, Cray and IBM compilers on both Intel and
+IBM Power CPUs.
+
+Investigation of the absolute serial performance of the NEMLite2D code
+using the roofline model revealed that it was still significantly
+below any of the traditional roofline ceilings. We have developed a
+code-analysis tool~\cite{habakkuk} that is capable of providing more
+accurate ceilings by analysing the nature of the floating point
+computations performed by a kernel. In future work we aim to extend
+this tool such that it can be applied to code parallelised using
+OpenMP.
 
 The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal\ version of a code.
+to the performance of the \psykal\ version of a code. For both
+NEMLite2D and Shallow we have found that for serial performance, the
+most important transformation is that of in-lining the kernel source
+at the call site. However, for the more complex NEMOLite2D code the
+Cray compiler also had to be coerced into performing SIMD
+vectorisation through the use of souce-code directives.
+
+In this work we have also demonstrated the introduction of parallelism
+into the PSy layer with both OpenMP and OpenACC directives. In both
+cases we were able to leave the natural-science parts of the code
+unchanged. For OpenMP we achieved a parallel efficiency of $\sim70\%$
+on 12 threads by enclosing the body of the (single) PSy layer routine
+within a single PARALLEL region. Removal of unecessary synchronisation
+points through use of the NOWAIT clause boosted 12-thread performance
+by approximately 10\% with the Gnu and Cray compilers. The
+\psykal\ re-structuring of this (admittedly small) code was not found
+to pose any problems for the introduction of performant OpenMP.
+Similarly, we have also demonstrated good GPU performance using OpenACC
+in a \psykal\ version of the code.
+
+Having demonstrated that the \psykal\ separation of concerns may be
+applied to 2D finite-difference codes with minimal performance loss,
+our next steps will be; 1) to consider the automatic generation of the
+PSy layer and 2) to look at extending the approach to the full NEMO
+model (i.e. three dimensions). In future work we will analyse the
+performance of a domain-specific compiler that performs the automatic
+generation of the PSy layer. This compiler (which we have named
+`PSyclone') is currently under development.
 
@@
 This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}) and some {\bf NVIDIA system - or is it
-  Emerald?}.
+(\url{http://www.archer.ac.uk}) and Emerald, a GPU-accelerated High
+Performance Computer, made available by the Science \& Engineering
+South Consortium operated in partnership with the STFC
+Rutherford-Appleton Laboratory
+(\url{http://www.ses.ac.uk/high-performance-computing/emerald/}).
 

### 2016-11-15 docs/papers/roofline_habakkuk/habakkuk.tex
Beginnings of skeleton of Habakkuk paper

\documentclass[12pt]{article}
\usepackage[pdftex]{graphicx}
% and their extensions so you won't have to specify these with
% every instance of \includegraphics
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\usepackage{url}

\newcommand{\psykal}{{PS}y{KA}l}

\begin{document}

\title{Analysing the Performance of Shallow-Water Models with the
 Roofline Model}

\author{A.~.R.~Porter and R.~W.~Ford}

\maketitle

\section{Introduction}

Shallow implementation of Sadourny~\cite{sadourny75}.

\section{The Roofline Model and its Limitations}

\subsection{What the Roofline Model (RFM) is}

Arithmetic intensity and MAX(memory bandwidth, FLOPS)

\subsection{Approaches to constructing a RFM}

Theoretical limits versus experimentally-measured values (STREAM and
LINPACK).

Measuring Arithmetic Intensity. Hardware Performance Counters (access,
reliability, difficulty of interpretation) or eye-balling the code
(error prone and time consuming).

\subsection{Basic Assumptions}
\subsection{When the assumptions break down}

All FLOPs are not equal. For instance, on Ivy Bridge a division costs
approximately eight times as much as a multiplication.
Instruction-level parallelism - operations despatched to different
execution ports can be executed in parallel.

\section{How to remove some of the assumptions}

Parse the (Fortran) source code.

\section{Conclusions}

\bibliographystyle{unsrt}
\bibliography{habakkuk}

\end{document}

### 2016-11-16 docs/papers/roofline_habakkuk/habakkuk.tex
Habakkuk paper - Begin to discuss multi-core performance

@@
 Shallow implementation of Sadourny~\cite{sadourny75}.
+NEMOLite2D.
 
@@
 
-Parse the (Fortran) source code.
+Parse the (Fortran) source code.  Construct a Directed Acyclic Graph
+to represent the source.  Use the properties of the DAG to
+characterise the processed source.  At its simplest, this can just be
+a count of FLOPs and memory (array) accesses.  Can then be extended to
+allow for caching of consecutive array accesses as in stencil
+operations in a direct-addressed model.
+
+\subsubsection{Scheduling}
+
+Obtaining a performance estimate by simply counting all of the FLOPs
+that must be performed is equivalent to assuming that they are all
+executed sequentially. However, a core a current CPU is likely to have
+support for Instruction Level Parallelism.
+
+\subsection{Single Instruction Multiple Data}
+
+\subsection{Multi-core Performance}
+
+So far we have considered only single-core performance. In reality,
+any application seeking to achieve good performance will run in
+parallel on multiple CPU cores.  If we restrict ourselves to running a
+single process/thread per core then clearly that thread will have
+exclusive use of that core's registers and execution units. Resources
+shared between cores such as L2 cache or bandwidth to main memory will
+however be shared by the executing threads.
+
+\section{Results}
+
+\subsection{Shallow}
+
+\subsection{NEMOLite2D}
+
+\subsection{Tracer Advection in NEMO}
 

### 2017-01-10 docs/NEMO/psyclone_and_nemo.tex
Add section on key NEMO issues that PSyclone can help with.

@@
 of k loops (i.e. the values to use for the loop bounds) within NEMO.
+
+\section{NEMO Performance/Maintenance Issues that \psyclone\ Could Help With}
+
+This section contains a list of those issues that have been
+highlighted at various NEMO HPC and HPC subgroup meetings that
+\psyclone\ and the PSyKAl separation of concerns will help to address:
+
+\begin{enumerate}
+
+\item Removal of unecessary halo swaps - PSyclone only generates calls to
+  perform halo swaps when they are necessary for correctness.
+
+\item Ability to experiment with collecting of multiple halo swaps
+
+\item Aiding loop vectorisation through compiler directives
+
+\item Cache-blocking of loops to better utilise memory bandwidth
+
+\item Support for OpenACC so that NEMO can run on GPU-based systems
+
+\item Support for OpenMP so that NEMO can run in hybrid OpenMP/MPI mode
+
+\end{enumerate}
+
+For invasive and repetitive code changes (such as re-structuring loops
+for cache blocking) the PSyclone approach is particularly
+valuable. This is because such changes are then implemented as a
+PSyclone transformation which only affects the generated code - the
+oceanographer does not have to worry about the resulting code
+complexity (potentially a nested loop with depth six). Tuning of the
+cache blocking itself can also be easily done by an HPC expert since
+the PSyclone tool would apply any changes to the whole code base.
 

### 2017-01-10 docs/NEMO/psyclone_and_nemo.tex
Update NEMO doc to reflect PSyclones current functionality.

@@
 
-\psyclone\ is currently being developed as part of the GungHo project
-and forms a component of the Dynamo (the prototype dynamical core)
-build system. \psyclone\ can currently generate correct sequential code
-for Dynamo and supports transformations such as loop colouring, loop
-fusion, OpenMP and module inlining. We are currently working
-on OpenACC transformations and are about to start work on MPI support.
+\psyclone\ is being developed as part of the GungHo project
+and forms a component of the LFRic (the prototype Met Office
+atmosphere model implementing GungHo science) build
+system. \psyclone\ can currently generate correct sequential and
+MPI-parallel code for Dynamo and supports transformations such as loop
+colouring, loop fusion, OpenMP and module inlining. We plan to
+introduce support for OpenACC along similar lines to the OpenMP
+transformations.
 
@@
 We believe that there are three possible places in which to locate the
-k loop in NEMO:
+$k$ loop in NEMO:
 
@@
 on it being possible to construct a formal description of the majority
-of k loops (i.e. the values to use for the loop bounds) within NEMO.
+of $k$ loops (i.e. the values to use for the loop bounds) within NEMO.
 
@@
 One of the key advantages of the \psyclone\ system is that it is
-possible for it to co-exist with existing code structures including
+possible for it to co-exist with existing code structures, including
 infrastructure that makes use of MPI.

### 2017-01-11 docs/NEMO/psyclone_and_nemo.tex
Break up introduction into two sections and update the reference to the Shallow paper

@@
 performance of the original code could be recovered with optimisations
-restricted to the PSy layer. We also demonstrated that this code could
-run efficiently on multi-core architectures using OpenMP directives
-and GPU architectures using OpenACC, again with optimisations
-restricted to the PSy layer.
+restricted to the PSy layer.\cite{psykal_shallow} We also demonstrated
+that this code could run efficiently on multi-core architectures using
+OpenMP directives and GPU architectures using OpenACC, again with
+optimisations restricted to the PSy layer.
 
 The question now is, would this approach be valuable and applicable
-for the full NEMO code?  Key to this is the fact that the full NEMO
-code is a combination of both 3D and 2D components and thus we have to
-decide how the loop over levels (the ``k-loop'') might fit within the
-\psyclone\ framework. In Dynamo this loop over levels is not managed by
-\psyclone\ but instead is performed within each kernel as a way of
-amortising the cost of the indirection associated with the
+for the full NEMO code? We expand on this in the rest of this document.
+
+\section{Managing the Iteration Space}
+
+In contrast to the codes tackled in GOcean, the full NEMO code
+contains a combination of both 3D and 2D components and thus we have
+to decide how the loop over levels (the ``k-loop'') might fit within
+the \psyclone\ framework. In Dynamo this loop over levels is not
+managed by \psyclone\ but instead is performed within each kernel as a
+way of amortising the cost of the indirection associated with the
 unstructured mesh.
@@
 
-\bibitem{psykal_shallow} A.~R.~Porter, R.~W.~Ford, M.~Ashworth, G.~D.~Riley and M.~Modani, ``Towards Compiler-Agnostic Performance in Finite-Difference Codes'', to appear in the Proceedings of ParCo 2015.
+\bibitem{psykal_shallow} A.~R.~Porter, R.~W.~Ford, M.~Ashworth, G.~D.~Riley and M.~Modani, ``Towards Compiler-Agnostic Performance in Finite-Difference Codes'', in ``Parallel Computing: On the Road to Exascale'', editors Gerhard R.
+Joubert and Hugh Leather and Mark Parsons and Frans Peters and Mark Sawyer, 27, 647-658 (2016) IOS Press, (Amsterdam, New York, Tokyo)
 

### 2017-01-24 docs/papers/roofline_habakkuk/habakkuk.tex
habakkuk paper: tweak text and add mulsd/divsd overlap data file

@@
 that must be performed is equivalent to assuming that they are all
-executed sequentially. However, a core a current CPU is likely to have
-support for Instruction Level Parallelism.
+executed sequentially. However, a core on any recent generation of CPU
+is likely to have support for Instruction Level Parallelism.
 
@@
 exclusive use of that core's registers and execution units. Resources
-shared between cores such as L2 cache or bandwidth to main memory will
+shared between cores (such as L2 cache or bandwidth to main memory) will
 however be shared by the executing threads.

### 2017-01-24 docs/papers/roofline_habakkuk/habakkuk.tex
habakkuk paper: add figure and update data for mulsd/divsd overlap

@@
 executed sequentially. However, a core on any recent generation of CPU
-is likely to have support for Instruction Level Parallelism.
+is likely to have support for Instruction Level Parallelism. For
+instance, in Intel Ivybridge micro-architecture a core has six
+execution ports to which micro-operations are
+despatched. Floating-point multiplication and division is handled by
+port 0 while addition is handled by port 1.
+
+\begin{figure}
+\includegraphics[width=130mm]{mulsd_divsd}
+\caption{Cost in cycles on an Intel E5-1620 v.2 CPU of the inner-loop
+  body as a function of number of independent MULSD and DIVSD
+  instructions.}
+\label{FIG_mul_div_overlap}
+\end{figure}
 

### 2017-05-24 docs/papers/roofline_habakkuk/habakkuk.tex
Move nemolite2d/shallow results in from general roofline report.

@@
 
+Although the Continuity kernel is very simple, the u-Momentum kernel
+is not and in particular, includes a call to the {\tt sin()} function.
+In order to estimate the cost of this function in FLOPs, the following
+code was compiled (with the Intel compiler and -O1) and run with
+likwid-perfctr:
+\begin{verbatim}
+  my_sum = 0.0d0
+  do i = 1, 1000
+    my_sum = my_sum + sin(real(i))
+  end do
+  write (*,*) 'My sum = ', my_sum
+\end{verbatim}
+This measurement revealed that the {\tt sin()} function requires
+approximately 13 FLOPs.
+
+A key component of the Roofline model is the Arithmetic Intensity of
+the code being executed:
+\begin{equation}
+AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
+\end{equation}
+Since HWPC data proved to be unreliable, we calculated this quantity
+manually by examining the source code and counting the number of
+memory references and arithmetic operations that it contained. In
+doing this counting we assume that any references to adjacent array
+elements are fetched in a single cache-line and thus only count once.
+
+To illustrate this, we consider the first loop-nest in the
+time-stepping section of the Shallow code. This is a doubly-nested
+loop over {\tt i} and {\tt j} with {\tt i} innermost. The body of the
+loop is:
+\begin{verbatim}
+CU(I+1,J) = .5d0*(P(I+1,J)+P(I,J))*U(I+1,J)
+CV(I,J+1) = .5d0*(P(I,J+1)+P(I,J))*V(I,J+1)
+Z(I+1,J+1) =(FSDX*(V(I+1,J+1)-V(I,J+1))- &
+             FSDY*(U(I+1,J+1)-U(I+1,J)))/&
+            (P(I,J)+P(I+1,J)+            &
+             P(I+1,J+1)+P(I,J+1))
+H(I,J) = P(I,J)+0.25d0*                   &
+         (U(I+1,J)*U(I+1,J)+U(I,J)*U(I,J) & 
+         +V(I,J+1)*V(I,J+1)+V(I,J)*V(I,J))
+\end{verbatim}
+This code writes to four memory locations ({\tt CU}, {\tt CV}, {\tt Z}
+and {\tt H}) and reads from ten distinct locations.  However, if we
+assume caching of reads from locations adjacent in memory ({\it e.g.} {\tt
+  P(i,j)} and {\tt P(i+1,j)}) then we have only six distinct reads.
+If we assume that each of these ten memory accesses will touch a
+separate cache line then this code fragment will result in the moving
+of ten cache lines, each of which is 64 bytes (eight double-precision
+words). However, as noted above, this code fragment is the body of a
+loop nest. Therefore, upon the next loop trip, {\tt i} will have been
+incremented by one and the majority of the accessed values will still
+be in cache. In fact, a new cache line will only have to be fetched
+every seven trips of the (innermost) loop and thus the mean memory
+traffic per loop trip is approximately $640/7$ bytes.
+
+We can also count the number of floating-point arithmetic operations
+(FLOPs) in this code fragment which gives us 12 multiplications and 12
+additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
+FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
+kernel of NEMOLite2D then we find that there are 51 addition
+operations, 56 multiplication/division operations and one call to {\tt
+  sin()}. There are 31 distinct memory references. This gives an $AI$
+of approximately $(107+13)*7/(31*64) = 0.42$ FLOPs/byte (where we
+assume the call to {\tt sin()} costs 13 FLOPs). The results of this
+analysis are summarised in Table~\ref{TAB_kernel_details}.
+
+\begin{table}
+\begin{tabular}{|l|c|c|c|}
+\hline
+                             & Shallow & \multicolumn{2}{c|}{NEMOLite2D} \\
+WSS of time-stepping (bytes) & $104n^2$ & \multicolumn{2}{c|}{$180n^2$} \\ 
+\hline
+Kernel                     & 1st loop nest & Continuity & u-Momentum \\
+\hline                                
+FLOPs                      & 24      &    14      &   107+13   \\
+Distinct memory refs       & 10      &    12      &   31       \\
+$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
+WSS (bytes)                & $56n^2$ &  $72n^2$   &   $136n^2$ \\
+\hline
+\end{tabular}
+\caption{Details of the kernels investigated with the roofline
+  model. WSS is the Working Set Size and $n$ is the problem size --
+  the length of one side of the square domain.}
+\label{TAB_kernel_details}
+\end{table}
+
+We measure the mean time to execute the u-Momentum kernel over the
+whole domain as 0.0021 seconds (Intel v.16 on the E5-1620) for the
+$256^2$ problem size. This gives us a performance of $\frac{120\times
+  256^2}{0.0021} = 3.7$ GFLOPS. (In contrast, instrumenting this
+section of code with the LIKWID marker API and using likwid-perfctr we
+measure 4.7 GFLOPS.)  For the $128^2$ domain, the measured time is
+0.556E-03 s. This gives a performance of $\frac{120\times
+  128^2}{0.556\times10^{-3}} = 3.5$ GFLOPS which again is some 25\% less
+than we measure directly with likwid.
+
+Using these figures we can finally construct the Roofline model for
+the two CPUs and the various kernels. In
+Figure~\ref{FIG_roofline_archer} we plot the model for the E5-2697 CPU
+and show the performance of the 1st loop-nest of Shallow and the
+u-Momentum kernel of NEMOLite2D. Two things are immediately clear from
+this plot: first, the performance of the first loop nest of Shallow is
+clearly memory-bandwidth limited and second, the u-Momentum kernel of
+NEMOLite2D is no-where near approaching this limit.
+
+\begin{figure}
+\centering
+\includegraphics[width=120mm]{roofline_archer}
+\caption{Comparison of the performance achieved by NEMOLite2D and
+  Shallow on the E5-2697 CPU.}
+\label{FIG_roofline_archer}
+\end{figure}
+
+Two performance points are plotted for the Shallow kernel
+corresponding to domains of $512^2$ and $1024^2$. Using the equation
+for working-set size in Table~\ref{TAB_kernel_details}, these
+correspond to 14.7 and 58.7~MB, respectively. Since the E5-2697 has an
+L3 cache of 30~MB, the $512^2$ case should be contained in cache while
+the $1024^2$ case will spill to main memory. This is borne out by the
+plot; the $1024^2$ case is achieving a performance roughly in-line
+with saturating the bandwidth to main memory. The $512^2$ case is
+approaching the L3 bandwidth limit. We investigate the reasons for the
+fact that it does not achieve the L3 bandwidth limit below.
+
+In contrast to the results obtained with the Shallow kernel, the
+performance of the NEMOLite2D u-Momentum kernel shows little
+sensitivity to working-set size -- the $128^2$, $256^2$ and $512^2$
+performance points in Figure~\ref{FIG_roofline_archer} are very close
+together and well below any of the roofline limits.
+
+In Figure~\ref{FIG_roofline_desktop} we plot the Roofline model
+obtained for the E5-1620 CPU. In order to investigate the poor
+performance of the u-Momentum kernel of NEMOLite2D, we also show the
+performance of the (simpler) Continuity kernel from that code. As we
+would expect, the performance of the u-Momentum kernel remains poor on
+this CPU. For the $512^2$ domain the Continuity kernel performs well,
+reaching the limit imposed by the bandwidth to main memory. However,
+the working-set sizes for this kernel for the $128^2$ and $256^2$
+domains are approximately 1.1~MB and 4.7~MB, respectively and
+therefore both of these should fit comfortably within L3
+cache. Despite this, the performance achieved with these domain sizes
+falls short of the L3 bandwidth limit.  Even when we consider that the
+Continuity kernel is just one of several within the time-stepping loop
+of NEMOLite2D we see that the L3 bandwidth limit is still applicable;
+the loop as a whole has a working-set size of $180\times n^{2}$
+bytes. For the $32^2$ domain this is 180~KB ({\it i.e.}\ within L2),
+while the $52^2$ (475~KB) through to the $128^2$ domain (2.9~MB) fall
+well within the capacity of L3.
+
+\begin{figure}
+\centering
+\includegraphics[width=120mm]{roofline_desktop}
+\caption{Comparison of the performance achieved by different
+  NEMOLite2D kernels on the E5-1620 CPU.}
+\label{FIG_roofline_desktop}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=120mm]{roofline_desktop_cont_limits}
+  \includegraphics[width=120mm]{roofline_desktop_mom_limits}
+  \caption{Roofline with computed ceilings for the Continuity and
+    Momentum kernels.
+    Kernel performance shown
+    is for the $256^2$ domain on the E5-1620 CPU.}
+\end{figure}
+
+In order to investigate why the performance of the Continuity kernel
+is falling well short of the maximum predicted by the roofline model
+we have experimented with the level of SIMD
+vectorisation. Figure~\ref{FIG_cont_vec} shows the performance of the
+Continuity kernel compiled with differing levels of SIMD vectorisation
+and after code modifications intended to improve vectorisation.
+Compiling without auto-vectorisation (compiler flag ``-no-vec''), the
+kernel achieves a performance that appears to be largely limited by
+the bandwidth to main memory. Note that there is very little
+opportunity for temporal cache re-use in the Continuity kernel -- only
+three of the twelve distinct memory references are to previous rows of
+the array. The size of this effect can be seen in the small drop in
+performance (black bars) in going from the $256^2$ to the $512^2$
+domain as at this point the working-set size of this kernel spills out
+of L3 cache.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=130mm]{continuity_opt}
+  \caption{Performance of the Continuity kernel with optimisations
+    intended to aid SIMD vectorisation. Results are for v.16 of the
+    Intel compiler on a Xeon E5-1620 v.2 CPU. Optimisations are
+    cumulative for results with a given SIMD instruction set.}
+  \label{FIG_cont_vec}
+\end{figure}
+
+The E5-1620 CPU supports two different SIMD instruction sets;
+Streaming SIMD Extensions (SSE; flag ``-axSSE4.2'') and Advanced
+Vector Extensions (AVX; flag ``-xHost''). SSE supports vector lengths
+of 128 bits (two double-precision words) and AVX 256 bits (four
+double-precision words). For the problem sizes that largely fit within
+L3 or L2 cache, switching on the use of SSE instructions provides a
+speed-up of between 1.5 and 1.8 - the factor increasing with problem
+size until the WSS of the time-stepping loop begins to fall out of L3
+cache. It is therefore likely that the reason this factor is below the
+ideal value of 2.0 is the overhead associated with constructing the
+loop and handling any ``peel'' or ``remainder'' loops. As the problem
+size increases, this overhead (as a fraction of the vectorisable
+workload) decreases. {\bf if this is the case then why is my 128x8 domain
+not faster than the 32x32?}
+
+When compiling with AVX instructions rather than SSE, the performance
+of the Continuity kernel was typically just below that of the latter
+for all of the problem sizes fitting within cache. We attribute this
+to the higher overheads associated with managing the longer AVX
+vectors.
+
+Examining the optimisation report produced by the Intel compiler
+revealed that the majority of array accesses within the loop were not
+aligned:
+\begin{verbatim}
+Example of unaligned access report here
+\end{verbatim}
+and that this inhibited efficient vectorisation. Tackling this issue
+requires two steps; first, ensure that the arrays are allocated to
+align with 64-byte boundaries and second, ensure that the compiler is
+aware of this where those arrays are used. The first requirement is
+easily met by passing the ``-align array64byte'' flag to
+the compiler when building the code. This ensures that any array that
+is constructed using a call to ALLOCATE is aligned on 64-byte
+boundaries in memory.
+
+The second requirement is complicated by the fact that the NEMOLite2D
+code makes extensive use of Fortran derived types --- the data arrays
+used for computation are members of derived types. The information
+about the alignment of these arrays cannot therefore be propagated
+through the code by module {\tt use} statements. In fact, even the
+Intel compiler directive which can be used to specify the alignment
+of a particular array at its point of use will not accept references
+to members of a derived type. It was therefore necessary to wrap the
+kernel call (and associated loops) within a routine and pass in
+references to the arrays. Within the wrapper routine the corresponding
+dummy arguments are declared as simple arrays and their alignment can
+then be specified using the {\tt ASSUME\_ALIGNED} directive:
+\begin{verbatim}
+subroutine invoke_continuity_arrays(nx, ny, ssha, ...)
+  integer, intent(in)   :: nx, ny
+  real(wp), intent(out) :: ssha(nx,ny)
+!DIR$ ASSUME_ALIGNED ssha:64
+  ...
+\end{verbatim}
+This approach also enables the extent of the arrays ({\tt nx} and {\tt
+  ny} in this example) to be specified explicitly.
+
+The continuity kernel includes array accesses of the form {\tt
+  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
+then this specifies a location in memory {\tt Nx} words before that
+occupied by {\tt a(i,j)}. In order for such an access to be aligned,
+{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
+AVX) and this must be specified to the compiler. We therefore padded {\tt Nx}
+as required and added a directive within the wrapper routine:
+\begin{verbatim}
+!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
+\end{verbatim}
+where {\tt ALIGNMENT} is an integer parameter declared to have the value two
+when compiling with SSE and four with AVX.
+
+We can see in Figure~\ref{FIG_cont_vec} that these changes (``SSE
+align'') significantly improve the performance of the smaller
+domains. Again, as the domain size increases this improvement reduces
+until, for $512^2$ and $1024^2$ (where we are limited by the bandwidth
+to main memory), this optimisation has no effect.
+
+In what follows, the code modifications were no longer restricted to
+those that maintained correctness since we were interested in what
+characteristics of the code were limiting its performance.
+
+Since the Continuity kernel is one of several in the time-stepping
+loop, we experimented with pre-loading the fields into cache simply by
+executing the Continuity kernel immediately prior to executing it for
+a second time with timing. Figure~\ref{FIG_cont_vec} shows that this
+``pre-loading'' slightly improved the kernel performance for those
+problem sizes that fit within cache. {\bf Discuss this}
+
+Once all of the above steps were complete, the only un-aligned array
+accesses reported by the compiler were those of the form {\tt
+  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
+unaligned accesses are unavoidable in a stencil-based
+finite-difference code. However, in order to quantify how much these
+accesses limit the performance of the kernel we experimented with
+replacing e.g. {\tt a(i-1,j)} with {\tt a(i-ALIGNMENT,j)}. The
+resulting kernel had no un-aligned accesses and performed
+significantly better; actually reaching or exceeding the L3-bandwidth
+roofline for domains of $52^2$ -- $256^2$ (Figure~\ref{FIG_cont_vec} -
+``align all'').
+
+The final SIMD-related optimisations are concerned with what happens
+at the beginning and end of a SIMD-vectorisable loop. If the loop
+updates {\it e.g.}\ {\tt a(i,j)} and begins on a value of {\tt i} that
+results in unaligned memory accesses ({\it e.g.}\ {\tt i = 2}) then
+the compiler must insert a ``peel'' loop to handle those iterations
+required to reach a point where memory is accessed on a 64-byte
+boundary.  Similarly, if the trip-count of the resulting vectorisable
+loop is not an integer multiple of the SIMD vector width then the
+compiler must insert a ``remainder'' loop to handle those iterations
+left over.
+
+We removed the remainder loop by ensuring that the length of the inner
+loop was a multiple of the vector width and adding an {\tt ASSUME}
+compiler directive ({\it c.f.}\ that for {\tt Nx} above). This was
+straightforward since we had already padded the array extents
+themselves and NEMOLite2D uses a mask to define the active
+computational domain - the additional grid points resulting from the
+padding were simply marked as external to the domain.  Despite doing
+this we found that the remainder loop was not removed until we
+manually introduced the peel loop (rather than relying on the
+compiler). Figure~\ref{FIG_cont_vec} shows that the resulting binary
+was the most performant of all those compiled with SSE on all problem
+sizes that are contained within L3 cache with the exception of $52^2$.
+{\bf Why is $52^2$ an exception? Something to do with tuning it to fit in L2?}
+
+\begin{verbatim}
+!DIR$ VECTOR ALIGNED
+do ic = 1, nrepeat, 1
+  do jj = 2, N, 1
+
+    ! Explicit peel loop
+    do ji = 2, ALIGNMENT
+       rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+       rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj  ))*un(ji-1,jj)
+       rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+       rtmp4 = (sshn_v(ji ,jj-1) + hv(ji  ,jj-1))*vn(ji,jj-1)
+       ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+             rdt / area_t(ji,jj)
+    end do
+
+    do ji = ALIGNMENT+1, M, 1
+
+       rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+       rtmp2 = (sshn_u(ji-1,jj ) + hu(ji-1,jj))*un(ji-1,jj)
+       rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+       rtmp4 = (sshn_v(ji ,jj-1) + hv(ji ,jj-1))*vn(ji,jj-1)
+
+       ssha(ji,jj) = sshn_t(ji,jj) + (rtmp2 - rtmp1 + rtmp4 - rtmp3) * &
+                     rdt / area_t(ji,jj)
+    end do
+
+  end do
+end do
+\end{verbatim}
+
+Since the caches of both of the CPUs used are eight-way associative,
+we also considered whether we might be experiencing ``cache
+thrashing'' whereby different locations in memory are mapped to the
+same location in cache. We therefore constructed a version of the
+kernel with the number of distinct streams reduced from twelve to
+eight. In order to maintain the same Arithmetic Intensity as the
+original kernel we also reduced the number of FLOPs. The inner loop of
+the resulting kernel was:
+\begin{verbatim}
+do ji = ALIGNMENT+1, M, 1
+   rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
+   rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))* &
+           un(ji-ALIGNMENT,jj)
+   rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
+   ssha(ji,jj) =  (rtmp2 - rtmp1 - rtmp3) / sshn_t(ji,jj)
+end do
+\end{verbatim}
+
+Figure~\ref{FIG_cont_vec} shows that the absolute performance of this
+modified kernel (``8 arrays'') is mostly lower than that of the kernel
+with twelve streams and certainly never exceeds it. We therefore
+conclude that cache-thrashing is not a problem for this kernel.
+
+Figure~\ref{FIG_kernel_perf} shows the performance of the Continuity
+and u-Momentum kernels when compiled without vectorisation
+(``-no-vec''), with SSE vectorisation (``-axSSE4.2'') and with AVX
+(``-xHost''). SSE supports vector lengths of 128 bits (two
+double-precision words) and AVX 512 bits (four double-precision
+words). For the 64, 128 and 256 problem sizes, changing from no
+vectorisation to SSE increases the performance of the Continuity
+kernel by up to roughly 60\%; some way short of the ideal speed-up of
+100\% that would be obtained with perfect SSE vectorisation of a
+compute-bound kernel. Unsurprisingly then, moving from SSE to AVX
+vectors does not give any performance improvement and in fact, is
+slightly detrimental for most of the problem sizes.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=120mm]{nemolite2d_kernel_perf}
+  \caption{Performance of the Continuity and u-Momentum kernels as a
+    function of SIMD vectorisation and problem size. Results are for
+    v.16 of the Intel compiler on a Xeon E5-1620 v.2 CPU.{\bf This plot
+requires updating or dropping in favour of the clustered-bar form.}}
+  \label{FIG_kernel_perf}
+\end{figure}
+
+Peak absolute performance is obtained for the $256^2$ problem size
+with SSE instructions. The 512 problem size spills from L3 cache and
+becomes bound by the bandwidth to main memory: SIMD vectorisation
+gives no performance benefit. The fact that best performance is
+obtained when the problem size fits within L3 indicates that there are
+other overheads that dominate when the problem size is small enough to
+fit within L2. We checked the efficiency of the SIMD vectorised code
+by looking at the compiler diagnostic output. On the basis of this we
+added Intel-specific directives to the code in order to specify that
+the arrays were aligned on 64-byte boundaries. We also manually coded
+the 'peel loop' (iterations required to reach a 64-byte
+boundary). Having done this the compiler reported that no peel or
+remainder loops were generated and the resulting (SSE) vectorised loop
+had an estimated speed-up of 1.9. (This estimated speed-up was
+previously 1.6; in agreement with the speed-up we measured.) After
+making these changes the performance of the Continuity kernel with the
+$32^2$ domain increased from 4.22 to 4.57 GFLOPs.
+
+Turning now to the u-Momentum kernel (lower plot in
+Figure~\ref{FIG_kernel_perf}), it is striking that the kernel
+performance is almost independent of problem size. The reason for this
+becomes clear when we edit the kernel to remove the many IF statements
+(present because of the application of boundary conditions). The blue
+symbols mark the resulting performance which is significantly
+increased and again reaches a maximum at the 256 problem size.  It
+should be noted that the compiler reported that this loop had been
+vectorised, even before the IF statements were removed. Clearly the
+vectorisation is more efficient once the code is simplified. However,
+even with the up-lift in performance given by removing all the IF
+statements, the performance of this kernel is still far short of even
+the main-memory bandwidth limit, despite the fact that the $256^2$
+problem fits within L3 cache.
+
+In order to help further understand the performance of the Continuity
+kernel, in Figure~\ref{FIG_cont_perf} we plot the performance of the
+Continuity kernel alongside HWPC-derived metrics (obtained using the
+likwid tool). We use the version compiled with SSE instructions since
+that gave the best performance. This plot confirms that the $32^2$
+problem size is dominated by bandwidth to L2. Sizes $64^2$, $128^2$
+and $256^2$ are dominated by bandwidth to L2 and L3 and $512^2$ makes
+heavy use of main memory.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=120mm]{nemolite2d_cont_perf}
+  \caption{Performance of the Continuity kernel (open circles -- right y-axis)
+    and associated HWPC data (bars -- left y-axis) as a function of
+    problem size. Results are for v.16 of the Intel compiler on a Xeon
+    E5-1620 v.2 CPU using only the SSE SIMD instructions.}
+  \label{FIG_cont_perf}
+\end{figure}
+
 \subsection{Tracer Advection in NEMO}

### 2017-05-25 nemolite2d_perf.tex
nemolite2d-paper: updates to abstract and conclusions

@@
 
-We present an approach which we call \psykal\ that aims to achieve
+We present an approach which we call \psykal\ that is designed to achieve
 portable performance for parallel, finite-difference Ocean models.  In
@@
 or not the obtained performance is `good' we also consider the
-limitations of the basic roofline model and attempt to improve on it
+limitations of the basic roofline model and improve on it
 by generating kernel-specific CPU ceilings.
@@
 \begin{IEEEkeywords}
-Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC
+Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC, roofline
 \end{IEEEkeywords}
@@
 performance of the extensive code re-structuring required by the
-approach when applied to the `Shallow' shallow-water model. We found
-that although the re-structuring did initially incur a sizeable
-performance penalty, it was possible to transform the resulting code
-to recover performance (for a variety of CPU/compiler combinations)
-while obeying the \psykal\ separation of concerns. In this work we
-move to looking at portable parallel performance within the
-\psykal\ approach.
+approach when applied to the `Shallow' shallow-water model
+(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
+re-structuring did initially incur a sizeable performance penalty, it
+was possible to transform the resulting code to recover performance
+(for a variety of CPU/compiler combinations) while obeying the
+\psykal\ separation of concerns. In this work we move to looking at
+portable parallel performance within the \psykal\ approach.
 
@@
 For this work we have used a program, `NEMOLite2D,' developed by
-ourselves.  NEMOLite2D is a vertically-averaged version of NEMO
-(Nucleus for European Modelling of the Ocean~\cite{nemo}), retaining
-only its dynamical part. The whole model system is represented through
-one continuity equation \eqref{continuity} (for the update of the
-sea-surface height) and two vertically-integrated momentum equations
-\eqref{momentum} (for the two velocity components, respectively).
+ourselves (https://puma.nerc.ac.uk/trac/GOcean).  NEMOLite2D is a
+vertically-averaged version of NEMO (Nucleus for European Modelling of
+the Ocean~\cite{nemo}), retaining only its dynamical part. The whole
+model system is represented through one continuity equation
+\eqref{continuity} (for the update of the sea-surface height) and two
+vertically-integrated momentum equations \eqref{momentum} (for the two
+velocity components, respectively).
 
@@
 GPU. This is because the necessary code modifications are, by design,
-limited to the middle PSy layer. In order to port the Fortran
-NEMOLite2D to GPU we used OpenACC directives as opposed to the OpenMP
-directives used for the multi-core CPU. We also experimented with
-using CUDA directly in order to quantify any performance penalty
-incurred.
-
-The introduction of these directives proved to be simpler than
-producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
-  whether this is true}.  Consequently we only show the performance of
-{\bf xxx} versions of the code in
-Figure~\ref{FIG_gpu_opt_stages}. {\bf discuss performance drop in
-  OpenACC version c.f. CUDA.}
+limited to the middle PSy layer. In order to demonstrate this and to
+check for any limitations imposed by the \psykal\ re-structuring, we
+had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
+used OpenACC directives for this porting as that approach is similar
+to the use of OpenMP directives and works well within the \psykal
+apprach. He also experimented with using CUDA directly within the
+original form of NEMOLite2D in order to quantify any performance
+penalty incurred by taking the \psykal/OpenACC approach.
+
+The performance of these two versions of the code is plotted in
+Figure~\ref{FIG_gpu_opt_stages}.  Since the \psykal restructuring of
+NEMOLite2D did not prove to be a barrier to achieving a performant
+code when using OpenACC directives, we only report the performance of
+the \psykal\ OpenACC version. For problem sizes up to $256^2$ the
+difference in the performance of the two versions is less than 5\% and
+for the larger domains it is at most 11\%. This demonstrates that the
+performance cost of the \psykal\ approach in utilising a GPU for
+NEMOLite2D is minimal.
 
@@
 
-We have used a new and un-optimised, two-dimension model extracted
+We have used a new and un-optimised, two-dimensional model extracted
 from the NEMO ocean model for this work. As a consequence of this, the
@@
 experience~\cite{shallow_psykal} with tackling the Shallow code which
-has had been optimised over the course of several years. In that case
-we were able to recover more than 90\% of the performance of the
-original for the Gnu, Intel, Cray and IBM compilers on both Intel and
-IBM Power CPUs.
-
-Investigation of the absolute serial performance of the NEMLite2D code
-using the roofline model revealed that it was still significantly
-below any of the traditional roofline ceilings. We have developed a
-code-analysis tool~\cite{habakkuk} that is capable of providing more
-accurate ceilings by analysing the nature of the floating point
-computations performed by a kernel. In future work we aim to extend
-this tool such that it can be applied to code parallelised using
-OpenMP.
+has been optimised over many years. In that case we were able to
+recover (to within a few percent) the performance of the original and
+in some cases exceed it, in spite of limiting ourselves to
+transformations which replicated the structure of the original,
+optimised code.
+
+Investigation of the absolute serial performance of the NEMOLite2D
+code using the roofline model revealed that it was still significantly
+below any of the traditional roofline ceilings. We have developed
+Habakkuk (https://github.com/arporter/habakkuk), a code-analysis tool
+that is capable of providing more realistic ceilings by analysing the
+nature of the floating point computations performed by a kernel. The
+bounds produced by Habakkuk are in good agreement with the measured
+performance of the principal (Momentum) kernel in NEMOLite2D. In
+future work we aim to extend this tool to account for SIMD operations
+and make it applicable to code parallelised using OpenMP.
 
@@
 to the performance of the \psykal\ version of a code. For both
-NEMLite2D and Shallow we have found that for serial performance, the
+NEMOLite2D and Shallow we have found that for serial performance, the
 most important transformation is that of in-lining the kernel source
-at the call site. However, for the more complex NEMOLite2D code the
-Cray compiler also had to be coerced into performing SIMD
-vectorisation through the use of souce-code directives.
+at the call site, i.e. within the PSy layer. (Although we have done
+this in-lining manually for this work, our aim is that in future, such
+transformations will be performed automatically at compile-time and
+therefore do not affect the code that a scientist writes.)
+For the more complex NEMOLite2D code the Cray compiler also
+had to be coerced into performing SIMD vectorisation through the use
+of souce-code directives.
 
@@
 to pose any problems for the introduction of performant OpenMP.
-
-{\bf Some words about how good the OpenACC, \psykal\ version is compared
-to the CUDA version.}
-
-Having demonstrated that the \psykal\ separation of concerns may be
-applied to 2D finite-difference codes with minimal performance loss,
-our next steps will be; 1) to consider the automatic generation of the
-PSy layer and 2) to look at extending the approach to the full NEMO
+Similarly, we have also demonstrated good GPU performance using OpenACC
+in a \psykal\ version of the code.
+
+This paper demonstrates that the \psykal\ separation of concerns may
+be applied to 2D finite-difference codes without loss of
+performance. We have also shown that the resulting code is amenable to
+efficient parallelisation on both GPU and shared-memory CPU systems.
+This then means that it is possible to achieve performance portability
+while maintaining single-source science code.
+
+Our next steps will be first, to consider the automatic generation of the
+PSy layer and second, to look at extending the approach to the full NEMO
 model (i.e. three dimensions). In future work we will analyse the
@@
 generation of the PSy layer. This compiler (which we have named
-`PSyclone') is currently under development.
+`PSyclone', see https://github.com/stfc/psyclone) is currently under
+development.
 

### 2017-05-26 nemolite2d_perf.tex
nemolite2d-paper: change to copernicus style

@@
-%% nemolite2d_perf.tex
-%%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-\usepackage{cite}
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  \usepackage[pdftex]{graphicx}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.eps}
-\fi
-
-% *** MATH PACKAGES ***
-%
-\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
+%% 2-column papers and discussion papers
+\documentclass[gmd, manuscript]{copernicus}
+
+%% \usepackage commands included in the copernicus.cls:
+%\usepackage[german, english]{babel}
+%\usepackage{tabularx}
+%\usepackage{cancel}
+%\usepackage{multirow}
+%\usepackage{supertabular}
 %\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Do not use the stfloats baselinefloat ability as IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-\usepackage{dblfloatfix}
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
-% Axel Sommerfeldt. This package may be useful when used in conjunction with 
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-\usepackage{url}
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+%\usepackage{algorithm}
+%\usepackage{amsthm}
+%\usepackage{float}
+%\usepackage{subfig}
+%\usepackage{rotating}
+
+%\usepackage{cite}
+%\usepackage{url}
 
 \newlength{\picwidth}
-\setlength{\picwidth}{85mm}
-
-\newcommand{\psykal}{{PS}y{KA}l}
+\setlength{\picwidth}{83mm}
 
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
+
 \title{Portable Multi- and Many-Core Performance for Finite Difference Codes;
   Application to the Free-Surface Component of NEMO.}
-%
-% author names and IEEE memberships
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{A.~R.~Porter, J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Liu and G.~D.~Riley% <-this % stops a space
-\thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
-\thanks{J.~Appleyard is with NVIDIA.}%
-\thanks{H.~Liu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
-\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
-}
-
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names.
-
-% The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% make the title area
+
+\Author[1]{Andrew}{Porter}
+\Author[2]{Jeremy}{Appleyard}
+\Author[1]{Mike}{Ashworth}
+\Author[1]{Rupert}{Ford}
+\Author[3]{Jason}{Holt}
+\Author[3]{Hedong}{Liu}
+\Author[4]{Graham}{Riley}
+
+\affil[1]{Science and Technology Facilities Council, Daresbury Laboratory, UK}
+\affil[2]{NVIDIA}
+\affil[3]{National Oceanography Centre, Liverpool, UK}
+\affil[4]{University of Manchester, Manchester, UK}
+
+\runningtitle{Portable Multi- and Many-Core Performance for Finite Difference Codes}
+
+\runningauthor{A.~R.~Porter et al}
+
+\correspondence{Andrew Porter (andrew.porter@stfc.ac.uk)}
+
+\received{}
+\pubdiscuss{} %% only important for two-stage journals
+\revised{}
+\accepted{}
+\published{}
+
+\firstpage{1}
 \maketitle
@@
 
-We present an approach which we call \psykal\ that is designed to achieve
+We present an approach which we call {PS}y{KA}l that is designed to achieve
 portable performance for parallel, finite-difference Ocean models.  In
-\psykal\ the code related to the underlying science is formally
+{PS}y{KA}l the code related to the underlying science is formally
 separated from code related to parallelisation and single-core
@@
 full ocean models (input/output, boundary conditions, \textit{etc.}).
-We have then manually constructed a \psykal\ version of this code and
+We have then manually constructed a {PS}y{KA}l version of this code and
 investigated the transformations that must be applied to the
@@
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC, roofline
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+\copyrightstatement{Copyright Science and Technology Facilities Council, 2017}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HE" in caps to complete the first word.
-\IEEEPARstart{T}{he} challenge presented to the developers of
-scientific software by the drive towards Exascale computing is
-considerable. With power consumption becoming the overriding design
-constraint, CPU clock speeds are falling and the complex,
-multi-purpose compute core is being replaced by multiple, simpler
-cores. This philosophy can be seen at work in the rise of so-called
-accelerator-based machines in the Top 500 List~\cite{top500} of
-supercomputers: four of the top-ten machines in the June 2016 list
-make use of many-core processors (Intel Xeon Phi, NVIDIA GPU or NRCPC
-SW26010). Two of the remaining six machines are IBM BlueGene/Qs, the
-CPU of which has hardware support for running 64 threads.
+\introduction
+
+The challenge presented to the developers of scientific software by
+the drive towards Exascale computing is considerable. With power
+consumption becoming the overriding design constraint, CPU clock
+speeds are falling and the complex, multi-purpose compute core is
+being replaced by multiple, simpler cores. This philosophy can be seen
+at work in the rise of so-called accelerator-based machines in the Top
+500 List~\cite{top500} of supercomputers: four of the top-ten machines
+in the June 2016 list make use of many-core processors (Intel Xeon
+Phi, NVIDIA GPU or NRCPC SW26010). Two of the remaining six machines
+are IBM BlueGene/Qs, the CPU of which has hardware support for running
+64 threads.
 
@@
 In a previous paper~\cite{shallow_psykal} we introduced a possible
-approach to tackling this problem which we term \psykal\ (discussed
+approach to tackling this problem which we term {PS}y{KA}l (discussed
 below). In that work we considered the implications for serial
@@
 (for a variety of CPU/compiler combinations) while obeying the
-\psykal\ separation of concerns. In this work we move to looking at
-portable parallel performance within the \psykal\ approach.
-
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
-\subsection{The \psykal\ Approach}
-
-The \psykal\ approach attempts to address the problems described in
+{PS}y{KA}l separation of concerns. In this work we move to looking at
+portable parallel performance within the {PS}y{KA}l approach.
+
+\subsection{The {PS}y{KA}l Approach}
+
+The {PS}y{KA}l approach attempts to address the problems described in
 the previous section. It separates code into three layers; the
@@
 
-In common with OP2, the \psykal\ approach separates out the science
+In common with OP2, the {PS}y{KA}l approach separates out the science
 code and the performance-related code into distinct layers. The calls
@@
 where they are placed in the code and in their semantics. However, the
-\psykal\ approach supports the specification of more than one kernel
+{PS}y{KA}l approach supports the specification of more than one kernel
 in a parallel region of code, compared with one for OP2, giving more
 scope for optimisation. In addition, the metadata describing a kernel
-is included with the kernel code in the \psykal\ approach whereas it
+is included with the kernel code in the {PS}y{KA}l approach whereas it
 is provided as a part of the kernel call in OP2.
 
-While the \psykal\ approach is general, we are currently applying it to
+While the {PS}y{KA}l approach is general, we are currently applying it to
 Atmosphere and Ocean models written in Fortran where domain
@@
 Since any real oceanographic computational model must output results,
-we ensure that any \psykal\ version of NEMOLite2D retains the Input/Output
+we ensure that any {PS}y{KA}l version of NEMOLite2D retains the Input/Output
 capability of the original. This aids in limiting the optimisations
-that can be performed on the \psykal\ version to those that should also
+that can be performed on the {PS}y{KA}l version to those that should also
 be applicable to full oceanographic models. Note that although we
@@
 In order to maximise the flexibility (and thus potential for
-optimisation) of the \psykal\ version of NEMOLite2D, we made the
+optimisation) of the {PS}y{KA}l version of NEMOLite2D, we made the
 kernels as fine-grained as possible. This resulted in eight distinct
@@
 \end{verbatim}
-\caption{A schematic of the top-level of the \psykal\ version of the
+\caption{A schematic of the top-level of the {PS}y{KA}l version of the
  NEMOLite2D code.}
@@
 unoptimised version after it has been re-structured following the
-\psykal\ approach. In addition to this benchmarking, we profile these
+{PS}y{KA}l approach. In addition to this benchmarking, we profile these
 versions of the code at the algorithm level (using a high-level timing
@@
 when we consider the performance loss when moving from the original to
-the \psykal\ version of the code; it slows down by a factor of
+the {PS}y{KA}l version of the code; it slows down by a factor of
 two. Although less significant in terms of absolute time, the
-Continuity section is also dramatically slower in the \psykal\ version,
+Continuity section is also dramatically slower in the {PS}y{KA}l version,
 this time by a factor of three. In contrast, the performance of the
 Time-update and Boundary Condition regions are not significantly
-affected by the move to the \psykal\ version.
+affected by the move to the {PS}y{KA}l version.
 
 \begin{table}[!t]
-\caption{The performance profile of the original and \psykal\ versions
+\caption{The performance profile of the original and {PS}y{KA}l versions
   of NEMOLite2D on the Intel Ivy Bridge CPU (for 2000 time-steps of the
@@
 \hline
-Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
-\multicolumn{2}{c}{Final \psykal} \\
+Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla {PS}y{KA}l} &
+\multicolumn{2}{c}{Final {PS}y{KA}l} \\
         & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
@@
 
-Beginning with the vanilla \psykal\ version, we then apply a series of
-code transformations while obeying the \psykal\ separation of concerns,
+Beginning with the vanilla {PS}y{KA}l version, we then apply a series of
+code transformations while obeying the {PS}y{KA}l separation of concerns,
 {\it i.e.} optimisation is restricted to the middle, {PS}y layer and
@@
 therefore chose to attempt to optimise this section first. In-keeping
-with the \psykal\ approach, we are only permitted to optimise the middle
+with the {PS}y{KA}l approach, we are only permitted to optimise the middle
 (PSy) layer which, for this section comprises calls to two kernels,
@@
 It turns out that the Cray-compiled binaries of both the original and
-\psykal\ versions of NEMOLite2D perform considerably less well than
+{PS}y{KA}l versions of NEMOLite2D perform considerably less well than
 their Intel-compiled counterparts. Comparison of the diagnostic output
@@
 Having optimised the Momentum section as much as permitted by the
-\psykal\ approach, we turn our attention to the three remaining
+{PS}y{KA}l approach, we turn our attention to the three remaining
 sections of the code. The profile data in Table~\ref{TABLE_profile}
@@
 striking however, is that the cost of the Continuity section increases
-by more than a factor of three in moving to the \psykal\ version of the
+by more than a factor of three in moving to the {PS}y{KA}l version of the
 code.
@@
 
-For this part of the work we began with the optimal \psykal\ version of
+For this part of the work we began with the optimal {PS}y{KA}l version of
 the code, as obtained after applying the various transformations
@@
 
-Moving now the to the \psykal\ version of NEMOLite2D,
+Moving now the to the {PS}y{KA}l version of NEMOLite2D,
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest \psykal\ version for each of the compiler/CPU combinations.
+fastest {PS}y{KA}l version for each of the compiler/CPU combinations.
 While the Intel compiler still produces the best-performing binary,
 the Cray-compiled binary is now a very close second. In fact, the
-performance of both the Gnu- and Cray-compiled \psykal\ versions is
+performance of both the Gnu- and Cray-compiled {PS}y{KA}l versions is
 generally significantly greater than that of their respective original
@@
 \includegraphics[width=7.5cm]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any \psykal\ 
+\caption{Summary of the best performance achieved by any {PS}y{KA}l 
 version of NEMOLite2D with each of the compilers and CPUs under 
@@
 Figure~\ref{FIG_slowdown_summary} plots the percentage difference
-between the performance of the original and the best \psykal\ versions
+between the performance of the original and the best {PS}y{KA}l versions
 of NEMOLite2D for each compiler/CPU combination. This shows that it is
 only the Intel-compiled binary running the $64^2$ domain that is
-slower with the \psykal\ version of the code (and then only by some
-3\%). For all other points in the space, the optimised \psykal\ version
+slower with the {PS}y{KA}l version of the code (and then only by some
+3\%). For all other points in the space, the optimised {PS}y{KA}l version
 of the code performs better. The results for the Cray compiler are
@@
 \includegraphics[width=2.8in]{slowdown_summary}
-\caption{Comparison of the performance of the best \psykal\
+\caption{Comparison of the performance of the best {PS}y{KA}l
 version with that of the original version of the code. A negative value 
-indicates that the \psykal\ version is slower than the original.}
+indicates that the {PS}y{KA}l version is slower than the original.}
 \label{FIG_slowdown_summary}
@@
 Original         & 3.87 & 12.2 & 6.83 \\
-Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
+Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
 Safe-address     & --   & --   & 6.95 \\
@@
 \includegraphics[width=93mm]{opt_stages_256}
-\caption{Performance of the \psykal\ version of NEMOLite2D for the
+\caption{Performance of the {PS}y{KA}l version of NEMOLite2D for the
   $256^{2}$ domain at each stage of optimisation. The first (black)
@@
 NEMOLite2D code; the introduction of OpenMP and OpenACC directives. In
-keeping with the \psykal\ approach we do not modify either the
+keeping with the {PS}y{KA}l approach we do not modify either the
 Algorithm- or Kernel-layer code.  Any code changes are restricted to
@@
 \caption{The scaling behaviour of the most performant OpenMP-parallel
-  version of \psykal\ NEMOLite2D for the full range of domain sizes
+  version of {PS}y{KA}l NEMOLite2D for the full range of domain sizes
   considered on the CPU. Results are for the Intel compiler on a
@@
 \caption{Performance of the OpenMP-parallel version of
-  \psykal\ NEMOLite2D for the $256^2$ domain with the Gnu compiler on
+  {PS}y{KA}l NEMOLite2D for the $256^2$ domain with the Gnu compiler on
   a single Intel Ivy Bridge socket. The corresponding parallel
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_intel}
-\caption{Performance of the OpenMP-parallel version of \psykal\
+\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
   NEMOLite2D for the Intel compiler on a single Intel Ivy Bridge
@@
 \includegraphics[width=\picwidth]{omp_scaling_256_cray}
-\caption{Performance of the OpenMP-parallel version of \psykal\
+\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
   NEMOLite2D for the Cray compiler on a single Intel Ivy Bridge
@@
 \includegraphics[width=\picwidth]{1and2sockets}
-\caption{Performance of the OpenMP-parallel version of \psykal\
+\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
   NEMOLite2D on one and two sockets of Intel Ivy Bridge. The 24-thread
@@
 %\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
-%\caption{Performance of the OpenMP-parallel version of \psykal\
+%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
 %  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
@@
 %\includegraphics[width=\picwidth]{omp_scaling_32_intel}
-%\caption{Performance of the OpenMP-parallel version of \psykal\
+%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
 %  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
@@
 %\includegraphics[width=\picwidth]{omp_scaling_32_cray}
-%\caption{Performance of the OpenMP-parallel version of \psykal\
+%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
 %  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
@@
 
-The advantage of the \psykal\ re-structuring becomes apparent if we
+The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
 wish to run NEMOLite2D on different hardware, \textit{e.g.} a
@@
 limited to the middle PSy layer. In order to demonstrate this and to
-check for any limitations imposed by the \psykal\ re-structuring, we
+check for any limitations imposed by the {PS}y{KA}l re-structuring, we
 had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
 used OpenACC directives for this porting as that approach is similar
-to the use of OpenMP directives and works well within the \psykal
+to the use of OpenMP directives and works well within the {PS}y{KA}l
 apprach. He also experimented with using CUDA directly within the
 original form of NEMOLite2D in order to quantify any performance
-penalty incurred by taking the \psykal/OpenACC approach.
+penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
 
 The performance of these two versions of the code is plotted in
-Figure~\ref{FIG_gpu_opt_stages}.  Since the \psykal restructuring of
+Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring of
 NEMOLite2D did not prove to be a barrier to achieving a performant
 code when using OpenACC directives, we only report the performance of
-the \psykal\ OpenACC version. For problem sizes up to $256^2$ the
+the {PS}y{KA}l OpenACC version. For problem sizes up to $256^2$ the
 difference in the performance of the two versions is less than 5\% and
 for the larger domains it is at most 11\%. This demonstrates that the
-performance cost of the \psykal\ approach in utilising a GPU for
+performance cost of the {PS}y{KA}l approach in utilising a GPU for
 NEMOLite2D is minimal.
@@
 \includegraphics[width=\picwidth]{gpu_opt_stages}
-\caption{Performance of the \psykal\ (OpenACC) and CUDA GPU
+\caption{Performance of the {PS}y{KA}l (OpenACC) and CUDA GPU
   implementations of NEMOLite2D. The performance for the OpenACC
   version of the original code is not shown since it is identical to
-  that of the \psykal\ version.}
+  that of the {PS}y{KA}l version.}
 \label{FIG_gpu_opt_stages}
@@
 \includegraphics[width=\picwidth]{cpu_cf_gpu}
-\caption{Performance of the best OpenMP-parallel version of \psykal\
-  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the \psykal\ GPU implementation (using OpenACC).}
+\caption{Performance of the best OpenMP-parallel version of {PS}y{KA}l\
+  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the {PS}y{KA}l GPU implementation (using OpenACC).}
 \label{FIG_cpu_cf_gpu}
@@
 
-Although we have investigated how the performance of the \psykal\
+Although we have investigated how the performance of the {PS}y{KA}l
 version of NEMOLite2D compares with that or the original, we have not
@@
 is outside the scope of the current work since here we are focused on
-the \psykal\ separation of concerns and the introduction of parallelism
+the {PS}y{KA}l separation of concerns and the introduction of parallelism
 in the PSy layer.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-We have investigated the application of the \psykal\ separation of
+\conclusions
+
+We have investigated the application of the {PS}y{KA}l separation of
 concerns approach to the domain of shared-memory parallel,
@@
 from the NEMO ocean model for this work. As a consequence of this, the
-introduction of the \psykal\ separation of concerns followed by
+introduction of the {PS}y{KA}l separation of concerns followed by
 suitable transformations of the PSy Layer is actually found to improve
@@
 The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal\ version of a code. For both
+to the performance of the {PS}y{KA}l version of a code. For both
 NEMOLite2D and Shallow we have found that for serial performance, the
@@
 by approximately 10\% with the Gnu and Cray compilers. The
-\psykal\ re-structuring of this (admittedly small) code was not found
+{PS}y{KA}l re-structuring of this (admittedly small) code was not found
 to pose any problems for the introduction of performant OpenMP.
 Similarly, we have also demonstrated good GPU performance using OpenACC
-in a \psykal\ version of the code.
-
-This paper demonstrates that the \psykal\ separation of concerns may
+in a {PS}y{KA}l version of the code.
+
+This paper demonstrates that the {PS}y{KA}l separation of concerns may
 be applied to 2D finite-difference codes without loss of
@@
 
+\codeavailability{The NEMOLite2D package is available as the GOcean
+ 2.0 benchmark, available from https://puma.nerc.ac.uk/trac/GOcean.}
+
+\competinginterests{No competing interests are present.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% use section* for acknowledgement
-\section*{Acknowledgments}
+\begin{acknowledgements}
 
@@
 
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,nemolite2d_perf}
-%
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiographynophoto}{A.~R.~Porter}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{R.~Ford}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiographynophoto}{G.~Riley}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
+\end{acknowledgements}
+
+\bibliographystyle{copernicus}
+\bibliography{nemolite2d_perf}
 

### 2017-05-26 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: change e.g. \it to \textit for copernicus style.

@@
-%% nemolite2d_perf.tex
-%%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-\usepackage{cite}
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  \usepackage[pdftex]{graphicx}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.eps}
-\fi
-
-% *** MATH PACKAGES ***
-%
-\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
+%% 2-column papers and discussion papers
+\documentclass[gmd, manuscript]{copernicus}
+
+%% \usepackage commands included in the copernicus.cls:
+%\usepackage[german, english]{babel}
+%\usepackage{tabularx}
+%\usepackage{cancel}
+%\usepackage{multirow}
+%\usepackage{supertabular}
 %\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Do not use the stfloats baselinefloat ability as IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-\usepackage{dblfloatfix}
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
-% Axel Sommerfeldt. This package may be useful when used in conjunction with 
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-\usepackage{url}
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+%\usepackage{algorithm}
+%\usepackage{amsthm}
+%\usepackage{float}
+%\usepackage{subfig}
+%\usepackage{rotating}
 
 \newlength{\picwidth}
-\setlength{\picwidth}{85mm}
-
-\newcommand{\psykal}{{PS}y{KA}l}
+\setlength{\picwidth}{83mm}
 
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
+
 \title{Portable Multi- and Many-Core Performance for Finite Difference Codes;
   Application to the Free-Surface Component of NEMO.}
-%
-% author names and IEEE memberships
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{A.~R.~Porter, J.~Appleyard, M.~Ashworth, R.~W.~Ford, J.~Holt, H.~Liu and G.~D.~Riley% <-this % stops a space
-\thanks{M.~Ashworth, R.~W.~Ford and A.~R.~Porter are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
-\thanks{J.~Appleyard is with NVIDIA.}%
-\thanks{H.~Liu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
-\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
-}
-
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names.
-
-% The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% make the title area
+
+\Author[1]{Andrew}{Porter}
+\Author[2]{Jeremy}{Appleyard}
+\Author[1]{Mike}{Ashworth}
+\Author[1]{Rupert}{Ford}
+\Author[3]{Jason}{Holt}
+\Author[3]{Hedong}{Liu}
+\Author[4]{Graham}{Riley}
+
+\affil[1]{Science and Technology Facilities Council, Daresbury Laboratory, UK}
+\affil[2]{NVIDIA}
+\affil[3]{National Oceanography Centre, Liverpool, UK}
+\affil[4]{University of Manchester, Manchester, UK}
+
+\runningtitle{Portable Multi- and Many-Core Performance for Finite Difference Codes}
+
+\runningauthor{A.~R.~Porter et al}
+
+\correspondence{Andrew Porter (andrew.porter@stfc.ac.uk)}
+
+\received{}
+\pubdiscuss{} %% only important for two-stage journals
+\revised{}
+\accepted{}
+\published{}
+
+\firstpage{1}
 \maketitle
@@
 
-We present an approach which we call \psykal\ that aims to achieve
+We present an approach which we call {PS}y{KA}l that is designed to achieve
 portable performance for parallel, finite-difference Ocean models.  In
-\psykal\ the code related to the underlying science is formally
+{PS}y{KA}l the code related to the underlying science is formally
 separated from code related to parallelisation and single-core
@@
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+\copyrightstatement{Copyright Science and Technology Facilities Council, 2017}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HE" in caps to complete the first word.
-\IEEEPARstart{T}{he} challenge presented to the developers of
-scientific software by the drive towards Exascale computing is
-considerable. With power consumption becoming the overriding design
-constraint, CPU clock speeds are falling and the complex,
-multi-purpose compute core is being replaced by multiple, simpler
-cores. This philosophy can be seen at work in the rise of so-called
-accelerator-based machines in the Top 500 List~\cite{top500} of
-supercomputers: four of the top-ten machines in the June 2016 list
-make use of many-core processors (Intel Xeon Phi, NVIDIA GPU or NRCPC
-SW26010). Two of the remaining six machines are IBM BlueGene/Qs, the
-CPU of which has hardware support for running 64 threads.
+\introduction
+
+The challenge presented to the developers of scientific software by
+the drive towards Exascale computing is considerable. With power
+consumption becoming the overriding design constraint, CPU clock
+speeds are falling and the complex, multi-purpose compute core is
+being replaced by multiple, simpler cores. This philosophy can be seen
+at work in the rise of so-called accelerator-based machines in the Top
+500 List~\citep{top500} of supercomputers: four of the top-ten machines
+in the June 2016 list make use of many-core processors (Intel Xeon
+Phi, NVIDIA GPU or NRCPC SW26010). Two of the remaining six machines
+are IBM BlueGene/Qs, the CPU of which has hardware support for running
+64 threads.
 
@@
 required: that of the scientific domain which a code is modelling
-({\it e.g.} oceanography) and that of computational science. The
+(\textit{e.g.} oceanography) and that of computational science. The
 situation is currently complicated still further by the existence of
@@
 
-In a previous paper~\cite{shallow_psykal} we introduced a possible
-approach to tackling this problem which we term \psykal\ (discussed
+In a previous paper~\citep{shallow_psykal} we introduced a possible
+approach to tackling this problem which we term {PS}y{KA}l (discussed
 below). In that work we considered the implications for serial
 performance of the extensive code re-structuring required by the
-approach when applied to the `Shallow' shallow-water model. We found
-that although the re-structuring did initially incur a sizeable
-performance penalty, it was possible to transform the resulting code
-to recover performance (for a variety of CPU/compiler combinations)
-while obeying the \psykal\ separation of concerns. In this work we
-move to looking at portable parallel performance within the
-\psykal\ approach.
-
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
-\subsection{The \psykal\ Approach}
-
-The \psykal\ approach attempts to address the problems described in
+approach when applied to the `Shallow' shallow-water model
+(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
+re-structuring did initially incur a sizeable performance penalty, it
+was possible to transform the resulting code to recover performance
+(for a variety of CPU/compiler combinations) while obeying the
+{PS}y{KA}l separation of concerns. In this work we move to looking at
+portable parallel performance within the {PS}y{KA}l approach.
+
+\subsection{The {PS}y{KA}l Approach}
+
+The {PS}y{KA}l approach attempts to address the problems described in
 the previous section. It separates code into three layers; the
 Algorithm layer, the PSy layer and the Kernel layer. The approach has
-been developed in the GungHo project~\cite{GungHo}, which is creating
+been developed in the GungHo project~\citep{GungHo}, which is creating
 a new Dynamical core for the UK Met Office, and its design has been
-influenced by earlier work on OP2~\cite{OP2, PYOP2}.
-
-In common with OP2, the \psykal\ approach separates out the science
+influenced by earlier work on OP2~\citep{OP2, PYOP2}.
+
+In common with OP2, the {PS}y{KA}l approach separates out the science
 code and the performance-related code into distinct layers. The calls
@@
 compile into efficient code.  In an alternative
-approach~\cite{firedrake,fenics}, kernels are generated from a
+approach~\citep{firedrake,fenics}, kernels are generated from a
 high-level specification, potentially allowing them to be optimised
-automatically~\cite{coffee}.
+automatically~\citep{coffee}.
 
@@
 For this work we have used a program, `NEMOLite2D,' developed by
-ourselves.  NEMOLite2D is a vertically-averaged version of NEMO
-(Nucleus for European Modelling of the Ocean~\cite{nemo}), retaining
-only its dynamical part. The whole model system is represented through
-one continuity equation \eqref{continuity} (for the update of the
-sea-surface height) and two vertically-integrated momentum equations
-\eqref{momentum} (for the two velocity components, respectively).
+ourselves (https://puma.nerc.ac.uk/trac/GOcean).  NEMOLite2D is a
+vertically-averaged version of NEMO (Nucleus for European Modelling of
+the Ocean~\citep{nemo}), retaining only its dynamical part. The whole
+model system is represented through one continuity equation
+\eqref{continuity} (for the update of the sea-surface height) and two
+vertically-integrated momentum equations \eqref{momentum} (for the two
+velocity components, respectively).
 
@@
 condition can be set as a clipped or Flather's radiation
-condition~\cite{flather76}. The bottom friction takes a semi-implicit
+condition~\citep{flather76}. The bottom friction takes a semi-implicit
 form for the sake of model stability. As done in the original version
@@
 applied at the edges of the model domain. Since NEMOLite2D applies
-external boundary conditions ({\it e.g.} barotropic forcing), this is
+external boundary conditions (\textit{e.g.} barotropic forcing), this is
 done via user-supplied kernels.
@@
 
-Beginning with the vanilla \psykal\ version, we then apply a series of
-code transformations while obeying the \psykal\ separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
+Beginning with the vanilla {PS}y{KA}l version, we then apply a series of
+code transformations while obeying the {PS}y{KA}l separation of concerns,
+\textit{i.e.} optimisation is restricted to the middle, {PS}y layer and
 leaves the kernel and algorithm layers unchanged. The aim of these
@@
 
-\subsubsection{Addition of {\tt safe\_address} directives}
+\subsubsection{Addition of \texttt{safe\_address} directives}
 \label{sec_safe_address}
@@
 of the NEMOLite2D kernels contain conditional statements. These
-statements are there to check whether {\it e.g.} the current grid
+statements are there to check whether \textit{e.g.} the current grid
 point is wet or neighbours a boundary point. A compiler is better able
@@
 module to allocate all field-data arrays with extents greater than
-strictly required.  We were then able to safely add the {\tt safe\_address}
+strictly required.  We were then able to safely add the \texttt{safe\_address}
 before all of the loops where the Cray compiler indicated it might be
@@
 
-From our previous work~\cite{shallow_psykal} on a similar code we know
+From our previous work~\citep{shallow_psykal} on a similar code we know
 that kernel in-lining is critical to obtaining performance with both
@@
 separate source files, we have found (both here and in our previous
-work~\cite{shallow_psykal}) that the number of optimisations it
+work~\citep{shallow_psykal}) that the number of optimisations it
 performs is reduced if it first has to in-line a routine. For the
@@
 \end{verbatim}
-Inserting the Cray {\tt vector always} directive persuaded the compiler
+Inserting the Cray \texttt{vector always} directive persuaded the compiler
 to vectorise the loop:
@@
 section of the field, contains a pointer to a GOcean grid object. If a
-kernel requires grid-related quantities ({\it e.g.} the grid spacing)
+kernel requires grid-related quantities (\textit{e.g.} the grid spacing)
 then these are obtained by passing it a reference to the appropriate
@@
 \end{verbatim}
-where {\it ssha} is the only field that is written to by the kernel.
+where \textit{ssha} is the only field that is written to by the kernel.
 We remove any potential confusion by instead obtaining the
-grid-related (read-only) quantities from a field ({\it sshn\_t} in
+grid-related (read-only) quantities from a field (\textit{sshn\_t} in
 this case) that is only read by the kernel:
@@
 as compiler-friendly as possible, following the lessons learned from
-our previous work with the Shallow code~\cite{shallow_psykal}. It may
+our previous work with the Shallow code~\citep{shallow_psykal}. It may
 well be that transforming the code into some other structure would
@@
 OpenMP-parallel loop to initialise newly-allocated arrays to some
-value, {\it e.g.} zero.
+value, \textit{e.g.} zero.
 
@@
 optimisation can again be implemented without changing the
-natural-science code ({\it i.e.} the Application and Kernel layers).
+natural-science code (\textit{i.e.} the Application and Kernel layers).
 
@@
 thread counts and, for the NEMOLite2D code, is frequently unecessary.
-{\it E.g.} if a kernel does not make use of the results of a
+\textit{E.g.} if a kernel does not make use of the results of a
 preceeding kernel call then there is clearly no need for threads to
@@
 compilers and CPUs considered here. Unlike the Shallow
-code~\cite{shallow_psykal}, the original version of NEMOLite2D has not
+code~\citep{shallow_psykal}, the original version of NEMOLite2D has not
 been optimised. Although it is still a single source file it is, in
@@
 \centering
-\includegraphics[width=2.8in]{orig_summary}
+\includegraphics[width=\picwidth]{orig_summary}
 \caption{Summary of the performance of the original version of the 
@@
 \centering
-\includegraphics[width=7.5cm]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any \psykal\ 
+\includegraphics[width=\picwidth]{best_psykal_summary}
+\caption{Summary of the best performance achieved by any {PS}y{KA}l 
 version of NEMOLite2D with each of the compilers and CPUs under 
@@
 \centering
-\includegraphics[width=2.8in]{slowdown_summary}
-\caption{Comparison of the performance of the best \psykal\
+\includegraphics[width=\picwidth]{slowdown_summary}
+\caption{Comparison of the performance of the best {PS}y{KA}l
 version with that of the original version of the code. A negative value 
-indicates that the \psykal\ version is slower than the original.}
+indicates that the {PS}y{KA}l version is slower than the original.}
 \label{FIG_slowdown_summary}
@@
 \hline
-\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.6 & 2.39 & 42.2 \\
+\parbox{2.5cm}{\raggedright \%-speed-up of best \textit{c.f.} original} & 34.6 & 2.39 & 42.2 \\
 \hline
@@
 \centering
-\includegraphics[width=93mm]{opt_stages_256}
-\caption{Performance of the \psykal\ version of NEMOLite2D for the
+\includegraphics[width=\picwidth]{opt_stages_256}
+\caption{Performance of the {PS}y{KA}l version of NEMOLite2D for the
   $256^{2}$ domain at each stage of optimisation. The first (black)
@@
 
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
-
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
 \subsection{Parallel Performance}
@@
 NEMOLite2D code; the introduction of OpenMP and OpenACC directives. In
-keeping with the \psykal\ approach we do not modify either the
+keeping with the {PS}y{KA}l approach we do not modify either the
 Algorithm- or Kernel-layer code.  Any code changes are restricted to
 either the PSy (middle) layer or the underlying library that manages
-{\it e.g.} the construction of field objects.
+\textit{e.g.} the construction of field objects.
 
@@
 differences between the code versions with a given compiler. However,
-the best performance obtained on twelve threads ({\it i.e.}  without
+the best performance obtained on twelve threads (\textit{i.e.}  without
 hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
@@
 Gnu-compiled binary is only 6\% slower than that produced by the Intel
-compiler and is 13\% {\em faster} that that of the Cray compiler. In
+compiler and is 13\% \textit{faster} that that of the Cray compiler. In
 part this situation comes about because of the effect that adding the
@@
 results to this quantity by benchmarking using two sockets of Intel
-Ivy Bridge ({\it i.e.} using a complete node of ARCHER, a Cray
+Ivy Bridge (\textit{i.e.} using a complete node of ARCHER, a Cray
 XC30). For this configuration we ensured that threads were evenly
@@
 executables (and thus the associated OpenMP run-time library) with the
-standard OpenMP {\it static}, {\it dynamic} and {\it guided}
-scheduling policies~\cite{openmp_standard}. For the Intel
+standard OpenMP \textit{static}, \textit{dynamic} and \textit{guided}
+scheduling policies~\citep{openmp_standard}. For the Intel
 compiler/run-time, static loop scheduling was found to be best for all
@@
 
-Although we have investigated how the performance of the \psykal\
+Although we have investigated how the performance of the {PS}y{KA}l
 version of NEMOLite2D compares with that or the original, we have not
 addressed how efficient the original actually is. In order to do so we
-use the Roofline Model~\cite{roofline} which provides a relatively
+use the Roofline Model~\citep{roofline} which provides a relatively
 simple way of characterising the performance of a code in terms of
 whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\cite{para_pearls} and use the
-STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
+follow the approach suggested in~\citep{para_pearls} and use the
+STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
 get the upper bounds on the memory bandwidth and floating-point
@@
 rate-limiting or on the critical path. Using the details of the Intel
-Ivy Bridge microarchitecture published by Fog~\cite{fog_microarch,
+Ivy Bridge microarchitecture published by Fog~\citep{fog_microarch,
   fog_instructions} we have constructed performance estimates of the
@@
 to port 0 while addition and subtraction go to port
-1~\cite{fog_instructions}. Therefore we sum the cost of all
+1~\citep{fog_instructions}. Therefore we sum the cost of all
 multiplication and division operations in the DAG and compare that
@@
 
-We have used a new and un-optimised, two-dimension model extracted
+We have used a new and un-optimised, two-dimensional model extracted
 from the NEMO ocean model for this work. As a consequence of this, the
-introduction of the \psykal\ separation of concerns followed by
+introduction of the {PS}y{KA}l separation of concerns followed by
 suitable transformations of the PSy Layer is actually found to improve
 performance. This is in contrast to our previous
-experience~\cite{shallow_psykal} with tackling the Shallow code which
-has had been optimised over the course of several years. In that case
-we were able to recover more than 90\% of the performance of the
-original for the Gnu, Intel, Cray and IBM compilers on both Intel and
-IBM Power CPUs.
-
-Investigation of the absolute serial performance of the NEMLite2D code
-using the roofline model revealed that it was still significantly
-below any of the traditional roofline ceilings. We have developed a
-code-analysis tool~\cite{habakkuk} that is capable of providing more
-accurate ceilings by analysing the nature of the floating point
-computations performed by a kernel. In future work we aim to extend
-this tool such that it can be applied to code parallelised using
-OpenMP.
+experience~\citep{shallow_psykal} with tackling the Shallow code which
+has been optimised over many years. In that case we were able to
+recover (to within a few percent) the performance of the original and
+in some cases exceed it, in spite of limiting ourselves to
+transformations which replicated the structure of the original,
+optimised code.
+
+Investigation of the absolute serial performance of the NEMOLite2D
+code using the roofline model revealed that it was still significantly
+below any of the traditional roofline ceilings. We have developed
+Habakkuk (https://github.com/arporter/habakkuk), a code-analysis tool
+that is capable of providing more realistic ceilings by analysing the
+nature of the floating point computations performed by a kernel. The
+bounds produced by Habakkuk are in good agreement with the measured
+performance of the principal (Momentum) kernel in NEMOLite2D. In
+future work we aim to extend this tool to account for SIMD operations
+and make it applicable to code parallelised using OpenMP.
 
 The application of code transformations to the middle/PSy layer is key
-to the performance of the \psykal\ version of a code. For both
-NEMLite2D and Shallow we have found that for serial performance, the
+to the performance of the {PS}y{KA}l version of a code. For both
+NEMOLite2D and Shallow we have found that for serial performance, the
 most important transformation is that of in-lining the kernel source
-at the call site. However, for the more complex NEMOLite2D code the
-Cray compiler also had to be coerced into performing SIMD
-vectorisation through the use of souce-code directives.
+at the call site, i.e. within the PSy layer. (Although we have done
+this in-lining manually for this work, our aim is that in future, such
+transformations will be performed automatically at compile-time and
+therefore do not affect the code that a scientist writes.)
+For the more complex NEMOLite2D code the Cray compiler also
+had to be coerced into performing SIMD vectorisation through the use
+of souce-code directives.
 
@@
 by approximately 10\% with the Gnu and Cray compilers. The
-\psykal\ re-structuring of this (admittedly small) code was not found
+{PS}y{KA}l re-structuring of this (admittedly small) code was not found
 to pose any problems for the introduction of performant OpenMP.
 Similarly, we have also demonstrated good GPU performance using OpenACC
-in a \psykal\ version of the code.
-
-Having demonstrated that the \psykal\ separation of concerns may be
-applied to 2D finite-difference codes with minimal performance loss,
-our next steps will be; 1) to consider the automatic generation of the
-PSy layer and 2) to look at extending the approach to the full NEMO
+in a {PS}y{KA}l version of the code.
+
+This paper demonstrates that the {PS}y{KA}l separation of concerns may
+be applied to 2D finite-difference codes without loss of
+performance. We have also shown that the resulting code is amenable to
+efficient parallelisation on both GPU and shared-memory CPU systems.
+This then means that it is possible to achieve performance portability
+while maintaining single-source science code.
+
+Our next steps will be first, to consider the automatic generation of the
+PSy layer and second, to look at extending the approach to the full NEMO
 model (i.e. three dimensions). In future work we will analyse the
@@
 generation of the PSy layer. This compiler (which we have named
-`PSyclone') is currently under development.
+`PSyclone', see https://github.com/stfc/psyclone) is currently under
+development.
+
+\codeavailability{The NEMOLite2D package is available as the GOcean
+ 2.0 benchmark, available from https://puma.nerc.ac.uk/trac/GOcean.}
+
+\competinginterests{No competing interests are present.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% use section* for acknowledgement
-\section*{Acknowledgments}
+\begin{acknowledgements}
 

### 2017-05-26 nemolite2d_perf.tex
nemolite2d-paper: use \picwidth consistently
nemolite2d-paper: update top-500 stats in intro

@@
 %\usepackage{rotating}
-
-%\usepackage{cite}
-%\usepackage{url}
 
@@
 at work in the rise of so-called accelerator-based machines in the Top
-500 List~\cite{top500} of supercomputers: four of the top-ten machines
-in the June 2016 list make use of many-core processors (Intel Xeon
-Phi, NVIDIA GPU or NRCPC SW26010). Two of the remaining six machines
-are IBM BlueGene/Qs, the CPU of which has hardware support for running
-64 threads.
+500 List (http://www.top500.org/) of supercomputers: six of the
+top-ten machines in the November 2016 list make use of many-core
+processors (Intel Xeon Phi, NVIDIA GPU or NRCPC SW26010). Two of the
+remaining four machines are IBM BlueGene/Qs, the CPU of which has
+hardware support for running 64 threads.
 
@@
 
-In a previous paper~\cite{shallow_psykal} we introduced a possible
+In a previous paper~\citep{shallow_psykal} we introduced a possible
 approach to tackling this problem which we term {PS}y{KA}l (discussed
@@
 Algorithm layer, the PSy layer and the Kernel layer. The approach has
-been developed in the GungHo project~\cite{GungHo}, which is creating
+been developed in the GungHo project~\citep{GungHo}, which is creating
 a new Dynamical core for the UK Met Office, and its design has been
-influenced by earlier work on OP2~\cite{OP2, PYOP2}.
+influenced by earlier work on OP2~\citep{OP2, PYOP2}.
 
@@
 vertically-averaged version of NEMO (Nucleus for European Modelling of
-the Ocean~\cite{nemo}), retaining only its dynamical part. The whole
+the Ocean~\citep{nemo}), retaining only its dynamical part. The whole
 model system is represented through one continuity equation
@@
 code transformations while obeying the {PS}y{KA}l separation of concerns,
-{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
+\textit{i.e.} optimisation is restricted to the middle, {PS}y layer and
 leaves the kernel and algorithm layers unchanged. The aim of these
@@
 \centering
-\includegraphics[width=7.5cm]{best_psykal_summary}
+\includegraphics[width=\picwidth]{best_psykal_summary}
 \caption{Summary of the best performance achieved by any {PS}y{KA}l 
@@
 \centering
-\includegraphics[width=2.8in]{slowdown_summary}
+\includegraphics[width=\picwidth]{slowdown_summary}
 \caption{Comparison of the performance of the best {PS}y{KA}l
@@
 \centering
-\includegraphics[width=93mm]{opt_stages_256}
+\includegraphics[width=\picwidth]{opt_stages_256}
 \caption{Performance of the {PS}y{KA}l version of NEMOLite2D for the
@@
 either the PSy (middle) layer or the underlying library that manages
-{\it e.g.} the construction of field objects.
+\textit{e.g.} the construction of field objects.
 
@@
 addressed how efficient the original actually is. In order to do so we
-use the Roofline Model~\cite{roofline} which provides a relatively
+use the Roofline Model~\citep{roofline} which provides a relatively
 simple way of characterising the performance of a code in terms of
 whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\cite{para_pearls} and use the
-STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
+follow the approach suggested in~\citep{para_pearls} and use the
+STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
 get the upper bounds on the memory bandwidth and floating-point
@@
 performance. This is in contrast to our previous
-experience~\cite{shallow_psykal} with tackling the Shallow code which
+experience~\citep{shallow_psykal} with tackling the Shallow code which
 has been optimised over many years. In that case we were able to

### 2017-05-31 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: para on GOcean infrastructure

@@
 \affil[1]{Science and Technology Facilities Council, Daresbury Laboratory, UK}
-\affil[2]{NVIDIA}
+\affil[2]{NVIDIA Corporation}
 \affil[3]{National Oceanography Centre, Liverpool, UK}
@@
 at work in the rise of so-called accelerator-based machines in the Top
-500 List~\citep{top500} of supercomputers: four of the top-ten machines
-in the June 2016 list make use of many-core processors (Intel Xeon
-Phi, NVIDIA GPU or NRCPC SW26010). Two of the remaining six machines
-are IBM BlueGene/Qs, the CPU of which has hardware support for running
-64 threads.
+500 List of supercomputers (http://www.top500.org/): six of the
+top-ten machines in the November 2016 list make use of many-core
+processors (Intel Xeon Phi, NVIDIA GPU or NRCPC SW26010). Two of the
+remaining four machines are IBM BlueGene/Qs, the CPU of which has
+hardware support for running 64 threads.
 
@@
 {PS}y{KA}l separation of concerns. In this work we move to looking at
-portable parallel performance within the {PS}y{KA}l approach.
+portable \textit{parallel} performance within the {PS}y{KA}l approach.
 
@@
 kernel subroutines. As the Algorithm layer works on logically global
-fields and Kernel layer works on local fields, the PSy layer is
+fields and the Kernel layer works on local fields, the PSy layer is
 responsible for iterating over columns. It is also responsible for
@@
 therefore be tailored for a particular hardware (such as multi-core,
-many-core, GPGPUs, or some combination thereof) and software (such as
-compiler, operating system, MPI implementation) configuration with no
-change to the Algorithm or Kernel layer code. This approach therefore
-offers the potential for portable performance. In this work we apply
-optimisations to the PSy layer manually. The development of a tool to
-automate this process will be the subject of a future paper.
+many-core, GPUs, or some combination thereof) and software (such as
+compiler, operating system, MPI library, \textit{etc}.) configuration
+with no change to the Algorithm or Kernel layer code. This approach
+therefore offers the potential for portable performance. In this work
+we apply optimisations to the PSy layer manually. The development of a
+tool to automate this process will be the subject of a future paper.
 
@@
 depth, $\Omega$ is the Earth rotation velocity vector.  $g$ is the
-acceleration due to gravity. $\nu$ is the kinematic viscosity coefficient.
+acceleration due to gravity and $\nu$ is the kinematic viscosity coefficient.
 
@@
 that interests us here.
+
+In the Algorithm layer, fields (and grids) are treated as logically
+global objects. Therefore, as part of creating the {PS}y{KA}l version
+of NEMOLite2D, we represent fields with derived types instead of
+arrays in this layer. These types then hold information about the
+associated mesh and the extents of 'internal' and 'halo' regions as
+well as the data arrays themselves. This frees the natural scientist
+from having to consider these issues and allows for a certain degree
+of flexibility in the actual implementation (e.g. padding for
+alignment or increasing array extent to allow for other
+optimisations). The support for this is implemented as a library
+(which we term the GOcean Infrastructure) and is common to the
+{PS}y{KA}l versions of both NEMOLite2D and Shallow.
 
@@
 \caption{A schematic of the top-level of the {PS}y{KA}l version of the
- NEMOLite2D code.}
+  NEMOLite2D code. The kernels listed as arguments to the \texttt{invoke}
+call specify the operations to be performed.}
 \label{FIG_psykal_nemolite2d_structure}
@@
 \hline
-Compiler  &  Flags                                    & Notes \\
+Compiler  &  Flags                                                       \\
 \hline
-Gnu       & -Ofast -mtune=native -finline-limit=50000 &   \\
-Intel     & -O3 -fast -fno-inline-factor -xHost       &   \\
-Cray      & -O3 -O ipa5 -h wp                         &   \\
-PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 & \\
+Gnu       & -Ofast -mtune=native -finline-limit=50000                    \\
+Intel     & -O3 -fast -fno-inline-factor -xHost                          \\
+Cray      & -O3 -O ipa5 -h wp                                            \\
+PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 \\
 \hline
@@
 \begin{verbatim}
-ftn-6375 ftn: VECTOR File = nemolite2d.f90,
- Line = 448 
 A loop starting at line 448 would benefit 

### 2017-05-31 nemolite2d_perf.tex
nemolite-paper: add para describing constant loop bounds implementation

@@
 at work in the rise of so-called accelerator-based machines in the Top
-500 List (http://www.top500.org/) of supercomputers: six of the
+500 List of supercomputers (http://www.top500.org/): six of the
 top-ten machines in the November 2016 list make use of many-core
@@
 \begin{table}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
 \caption{The compiler flags used in this work.}
@@
 \centering
-\begin{tabular}{l|l|l}
+\begin{tabular}{l|l}
 \hline
-Compiler  &  Flags                                    & Notes \\
+Compiler  &  Flags                                                       \\
 \hline
-Gnu       & -Ofast -mtune=native -finline-limit=50000 &   \\
-Intel     & -O3 -fast -fno-inline-factor -xHost       &   \\
-Cray      & -O3 -O ipa5 -h wp                         &   \\
-PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 & \\
+Gnu       & -Ofast -mtune=native -finline-limit=50000                    \\
+Intel     & -O3 -fast -fno-inline-factor -xHost                          \\
+Cray      & -O3 -O ipa5 -h wp                                            \\
+PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 \\
 \hline
@@
 
+\subsubsection{Constant loop bounds}
+
+In the vanilla {PS}y layer, the lower and upper bounds for each loop
+over grid points are obtained from the relevant components of the
+derived type representing the field being updated by the kernel being
+called from within the loop. In our previous
+work~\citep{shallow_psykal} we found that the Cray compiler in
+particular produced more performant code if we changed the PSy layer
+such that the array extents are looked up once at the beginning of the
+PSy routine and then used to specify the loop bounds. We have therefore
+applied that transformation to the PSy layer of NEMOLite2D.
+
 \subsubsection{Addition of \texttt{safe\_address} directives}
@@
 \end{verbatim}
-In order that we could safely add such directives we altered the field
-module to allocate all field-data arrays with extents greater than
-strictly required.  We were then able to safely add the \texttt{safe\_address}
-before all of the loops where the Cray compiler indicated it might be
-useful (the Momentum loops and some of the BC loops).
-
-\subsubsection{In-line Momentum kernel bodies into middle-layer code}
+In order that we could safely add such directives we altered the
+GOcean infrastructure to allocate all field-data arrays with extents
+greater than strictly required.  We were then able to safely add the
+\texttt{safe\_address} before all of the loops where the Cray compiler
+indicated it might be useful (the Momentum loops and some of the BC
+loops).
+
+\subsubsection{In-line Momentum kernel bodies into middle-layer}
 \label{sec_inline_mom}
@@
 separate source files, we have found (both here and in our previous
-work~\citep{shallow_psykal}) that the number of optimisations it
+work~\citep{shallow_psykal}) that the extent of the optimisations it
 performs is reduced if it first has to in-line a routine. For the
@@
  estimated potential speedup: 0.540 
- lightweight vector operations: 1133 
- medium-overhead vector operations: 1 
- vectorized math library calls: 1 
+ ...
  --- end vector loop cost summary ---
@@
  estimated potential speedup: 2.350 
- lightweight vector operations: 697 
- medium-overhead vector operations: 1 
- vectorized math library calls: 1 
+ ...
  --- end vector loop cost summary ---
@@
 \begin{verbatim}
-do jj = 2, N, 1
- do ji = 2, M, 1
-
    call continuity_code(ji, jj,      &
@@
 of the original version of the NEMOLite2D code for the range of
-compilers and CPUs considered here. Unlike the Shallow
+compilers considered here. Unlike the Shallow
 code~\citep{shallow_psykal}, the original version of NEMOLite2D has not
@@
 favour the Intel compiler which significantly out-performs both the
-Gnu and Cray compilers. Only when the problem size falls out of cache
+Gnu and Cray compilers. Only when the problem size spills out of cache
 does the performance gap begin to narrow. The reason for the
@@
 \caption{Summary of the performance of the original version of the 
-NEMOLite2D code on 
-the range of compilers and CPU-based systems under consideration.}
+NEMOLite2D code on an Intel Ivy Bridge CPU for
+the range of compilers under consideration.}
 \label{FIG_orig_perf_summary}
@@
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest {PS}y{KA}l version for each of the compiler/CPU combinations.
+fastest {PS}y{KA}l version for each of the compiler/problem-size combinations.
 While the Intel compiler still produces the best-performing binary,
@@
 \caption{Summary of the best performance achieved by any {PS}y{KA}l 
-version of NEMOLite2D with each of the compilers and CPUs under 
-consideration.}
+version of NEMOLite2D for each of the compilers under consideration.}
 \label{FIG_best_psykal_perf_summary}
@@
 between the performance of the original and the best {PS}y{KA}l versions
-of NEMOLite2D for each compiler/CPU combination. This shows that it is
+of NEMOLite2D for each compiler/problem-size combination. This shows that it is
 only the Intel-compiled binary running the $64^2$ domain that is
@@
 performance of the compiled code. In fact, simply in-lining the two
-kernels associated with the Momentum section was enough to improve
-upon the performance of the original code.
+kernels associated with the Momentum section was sufficient to exceed
+the performance of the original code.
 
@@
 \begin{table}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-\caption{Performance (millions of points updated per second) for the
+  \caption{Performance (millions of points updated per second) on an Intel
+    Ivy Bridge CPU for the
   $256^2$ case after each code transformation. Where an optimisation
@@
 Compiler:           & Gnu & Intel & Cray  \\
-\hline
-CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
 \hline

### 2017-05-31 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-paper: final tweaks after review by ARP

@@
 \centering
-\begin{tabular}{l|l|l}
+\begin{tabular}{l|l}
 \hline
@@
 concludes that such vectorisation would be inefficient). Of the other
-transformations, only the addition of the compiler-specific
-safe\_address directive (see Section~\ref{sec_safe_address}) was found
-to give any benefit.
+transformations, only the change to constant loop bounds and the
+addition of the compiler-specific safe\_address directive (see
+Section~\ref{sec_safe_address}) were found to (slightly) improve
+performance.
 
 \begin{table}[!t]
-% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-\caption{Performance (millions of points updated per second) for the
+  \caption{Performance (millions of points updated per second) on an Intel
+    Ivy Bridge CPU for the
   $256^2$ case after each code transformation. Where an optimisation
@@
 \hline
-CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
-\hline
 Original             & 3.87 & 12.2 & 6.83 \\
-Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
+Vanilla {PS}y{KA}l   & 2.59 & 6.27 & 6.35 \\
+Constant loop bounds & 3.07 & 6.55 & 6.73 \\
 Safe-address         & --   & --   & 6.95 \\
@@
 
-\subsubsection{OpenMP Version}
+\subsubsection{OpenMP}
 
@@
 and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
-parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
+parallel efficiency (\%), $E(n)$, on $n$ threads, as:
 \[
@@
 \]
-where $P(m)$ is the performance of the code on $m$ threads. For a
+where $P(n)$ is the performance of the code on $n$ threads,
+\textit{e.g.} grid-points updated per second. For a
 perfect, linearly-scaling code, $E(n)$ will be 100\%.
@@
 threads. The Cray version however does better and is about 45\%
-efficient on eight threads (Figure~\ref{FIG_cray_omp_256}).
+efficient on eight threads (right axis, Figure~\ref{FIG_cray_omp_256}).
 
@@
 standard OpenMP \textit{static}, \textit{dynamic} and \textit{guided}
-scheduling policies~\citep{openmp_standard}. For the Intel
-compiler/run-time, static loop scheduling was found to be best for all
-versions apart from that where we have attempted to amortize the cost
-of the SINGLE section. This is to be expected since that strategy
-requires some form of dynamic loop scheduling in order to reduce the
-load imbalance introduced by the SINGLE section.
+scheduling policies. For the Intel compiler/run-time, static loop
+scheduling was found to be best for all versions apart from that where
+we have attempted to amortize the cost of the SINGLE section. This is
+to be expected since that strategy requires some form of dynamic loop
+scheduling in order to reduce the load imbalance introduced by the
+SINGLE section.
 
@@
 
-\subsubsection{OpenACC Version}
+\subsubsection{OpenACC}
 
@@
 Although we have investigated how the performance of the {PS}y{KA}l
-version of NEMOLite2D compares with that or the original, we have not
+version of NEMOLite2D compares with that of the original, we have not
 addressed how efficient the original actually is. In order to do so we
@@
 whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\citep{para_pearls} and use the
+follow the approach suggested in~\cite{para_pearls} and use the
 STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
@@
 
-A key component of the Roofline model is the Arithmetic Intensity of
-the code being executed:
-\begin{equation}
+A key component of the Roofline model is the Operational or Arithmetic
+Intensity of the code being executed:
+\[
 AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
-\end{equation}
+\]
 We calculated this quantity manually by examining the
@@
 arithmetic operations that it contained. In doing this counting we
-assume that any references to adjacent array elements are fetched in a
-single cache-line and thus only count once.
+assume that any references to adjacent array elements (\textit{e.g.}
+$u(i,j)$ and $u(i+1,j)$) are fetched in a single cache line and thus
+only count once.
 

### 2017-05-31 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: pluralize line about launching kernels on GPU

@@
 takes only 80~$\mu$s. The execution of a single time-step is then
-dominated by the time taken to launch a kernel on the GPU rather than
-the execution of the kernel itself.
+dominated by the time taken to launch the kernels on the GPU rather than
+the execution of the kernels themselves.
 

### 2017-06-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: WIP on description of OpenACC work

@@
 As the PSy layer iterates over columns, the single-core performance
-can be optimised by applying transformations such as loop fusion and
-loop blocking. Additionally, the potential parallelism within this
+can be optimised by applying transformations such as manipulation of
+loop bounds (e.g. padding for SIMD) and kernel
+in-lining. Additionally, the potential parallelism within this
 iteration space can also be exploited and optimised. The PSy layer can
@@
 arrays in this layer. These types then hold information about the
-associated mesh and the extents of 'internal' and 'halo' regions as
+associated mesh and the extents of `internal' and `halo' regions as
 well as the data arrays themselves. This frees the natural scientist
@@
 
-In order to maximise the flexibility (and thus potential for
-optimisation) of the {PS}y{KA}l version of NEMOLite2D, we made the
-kernels as fine-grained as possible. This resulted in eight distinct
+In re-structuring NEMOLite2D to conform to the {PS}y{KA}l separation
+of concerns we must break up the computation into multiple
+kernels. The more of these there are, the greater the potential for
+optimisation of the {PS}y layer. This re-structuring gave eight distinct
 kernels, each of which updates a single field at a single point (since
@@
 maximise the performance of the code. In both cases our aim is to
-identify those transformations which must be supported by any tool
+identify those transformations which must be supported by a tool
 which seeks to auto-generate a performant PSy layer.
@@
 Initially we implemented these copies as 'built-in' kernels (in the
-GOcean library) as they are specified in the algorithm layer. However,
-we obtained better performance by simply in-lining these array copies
-into the PSy layer.
+GOcean infrastructure) as they are specified in the algorithm
+layer. However, we obtained better performance by simply in-lining
+these array copies into the PSy layer.
 
@@
 
-For this part of the work we began with the optimal {PS}y{KA}l version of
-the code, as obtained after applying the various transformations
-described in the previous section. As with the transformations of the
-serial code, our purpose here is to determine the functionality
-required of a tool that seeks to generate the PSy layer.
+For this part of the work we began with the optimised {PS}y{KA}l
+version of the code, as obtained after applying the various
+transformations described in the previous section. As with the
+transformations of the serial code, our purpose here is to determine
+the functionality required of a tool that seeks to generate the PSy
+layer.
 
@@
 
-Since data arrays are managed within the GOcean Library this
+Since data arrays are managed within the GOcean infrastructure this
 optimisation can again be implemented without changing the
@@
 
+\subsection{Construction of OpenACC-Parallel NEMOLite2D}
+\label{sec_oacc_steps}
+
+The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
+wish to run NEMOLite2D on different hardware, \textit{e.g.} a
+GPU. This is because the necessary code modifications are, by design,
+limited to the middle PSy layer. In order to demonstrate this and to
+check for any limitations imposed by the {PS}y{KA}l re-structuring, we
+had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
+used OpenACC directives for this porting as that approach is similar
+to the use of OpenMP directives and works well within the {PS}y{KA}l
+approach.
+
+\subsubsection{Data Movement}
+
+Although the advent of technologies such as NVLink are alleviating the
+bottleneck presented by the connection of the GPU to the CPU, it
+remains critical to minimise data movement between the memory spaces
+of the two processing units. In NEMOLite2D this is achieved by
+performing all computation on the GPU. The whole time-stepping loop
+can then be enclosed inside a single OpenACC \texttt{data} region and it
+is only necessary to bring data back to the CPU for the purposes
+of I/O.
+
+\subsubsection{Kernel Acceleration}
+
+Moving the kernels to execute on the GPU was achieved by using the
+OpenACC \texttt{kernels} directive.
+
+He also experimented with using CUDA directly within the
+original form of NEMOLite2D in order to quantify any performance
+penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the effects of the transformations described in
-Section~\ref{sec_methodology}. Once we have arrived at an optimal form
-for the serial version of NEMOLite2D we then investigate its parallel
-performance on both CPU- and GPU-based systems.
+Section~\ref{sec_methodology}. Once we have arrived at an optimised
+form for the serial version of NEMOLite2D we then investigate its
+parallel performance on both CPU- and GPU-based systems.
 
@@
 check for any limitations imposed by the {PS}y{KA}l re-structuring, we
-had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
+had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
 used OpenACC directives for this porting as that approach is similar

### 2017-06-20 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: add methodology text and update GPU perf results

@@
 core upon which it is executing. One way of doing this is to implement
-a so-called 'first-touch policy' whereby memory addresses that will
+a so-called `first-touch policy' whereby memory addresses that will
 generally be accessed by a given thread during program execution are
@@
 Moving the kernels to execute on the GPU was achieved by using the
-OpenACC \texttt{kernels} directive.
-
-He also experimented with using CUDA directly within the
-original form of NEMOLite2D in order to quantify any performance
-penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
+OpenACC \texttt{kernels} directive in each Fortran routine containing
+loops over grid points. (In the {PS}y{KA}l version this is just the
+single {PS}y layer subroutine). This directive instructs the OpenACC
+compiler to automatically create a GPU kernel for each loop nest it
+encounters.
+
+\subsubsection{Force parallelisation of Flather kernels}
+
+The PGI compiler was unable to determine whether the loops applying
+the Flather boundary condition in the $x$ and $y$ directions were safe
+to parallelise. This information therefore had to be supplied by
+inserting a \texttt{loop independent} OpenACC directive before each of
+the two loops.
+
+\subsubsection{Loop collapse}
+
+When using the OpenACC \texttt{kernels} directive the compiler creates
+GPU kernels by parallelising only the outer loop of each loop nest.
+For the majority of loop nests in NEMOLite2D all of the iterations are
+independent. It is therefore beneficial to expose the maximum amount
+of parallelism to the GPU and we do this by instructing the compiler
+to parallelise both levels (i.e. \texttt{ji} and \texttt{jj}) of a
+loop nest. This is achieved using the OpenACC \texttt{loop collapse(2)}
+directive.
+
+\subsubsection{Shortloop}
+
+As well as accounting for the bulk of the execution time, the Momentum
+kernels ($u$ and $v$) are also relatively large.
+
+\subsubsection{Kernel fusion}
+
+\subsubsection{CUDA}
+
+We also experimented with using CUDA directly within the original form
+of NEMOLite2D in order to quantify any performance penalty incurred by
+taking the {PS}y{KA}l/OpenACC approach.
 

### 2017-06-20 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-paper: add text describing short loop and kernel fusin

@@
 For the majority of loop nests in NEMOLite2D all of the iterations are
-independent. It is therefore beneficial to expose the maximum amount
-of parallelism to the GPU and we do this by instructing the compiler
-to parallelise both levels (i.e. \texttt{ji} and \texttt{jj}) of a
-loop nest. This is achieved using the OpenACC \texttt{loop collapse(2)}
-directive.
+independent and the loop bodies consist of just a handful of
+executable statements. For small kernels the loop start-up cost can be
+significant and therefore it is therefore beneficial to generate as
+many threads as will fit on the GPU and then re-use them as
+required. For this approach to be efficient we must expose the maximum
+amount of parallelism to the GPU and we do this by instructing the
+compiler to parallelise both levels (i.e. \texttt{ji} and \texttt{jj})
+of a loop nest. This is achieved using the OpenACC \texttt{loop
+  collapse(2)} directive for all of the smaller kernels in NEMOLite2D.
 
@@
 
-As well as accounting for the bulk of the execution time, the Momentum
-kernels ($u$ and $v$) are also relatively large.
+In contrast to all of the other NEMOLite2D kernels, the Momentum
+kernels ($u$ and $v$) are relatively large in terms of the number of
+executable statements they contain. In turn, this means that the
+corresponding kernels will require more state and thus more registers
+on the GPU. Therefore, rather than generating as many threads as will
+fit on the GPU, it is more efficient only to generate as many as
+required by the problem. Once slots become free on the GPU new threads
+will launch. This minimises the overhead associated with thread
+creation and eases register pressure.  The compiler can be persuaded
+to adopt the above strategy through the use of the (PGI-specific)
+\texttt{loop shortloop} directive which tells it that there is not
+likely to be much thread re-use in the following loop. This directive
+was applied to both the \texttt{ji} and \texttt{jj} loops in both of
+the Momentum kernels.
 
 \subsubsection{Kernel fusion}
+
+Both of the Momentum kernels read from 20 double-precision arrays and
+thus require considerable memory bandwidth.  However, the majority of
+these arrays are used by both the $u$ and $v$ kernels. It is therefore
+possible to reduce the memory-bandwidth requirements by fusing the two
+kernels together. In practice, this means fusing the two,
+doubly-nested loops so that we have a single loop nest containing both
+the $u$ and $v$ kernel bodies/calls.
 

### 2017-06-20 nemolite2d_perf.tex
nemolite-paper: correct collapse and short-loop descriptions. Update gpu perf data

@@
 
+\subsection{Construction of OpenACC-Parallel NEMOLite2D}
+\label{sec_oacc_steps}
+
+The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
+wish to run NEMOLite2D on different hardware, \textit{e.g.} a
+GPU. This is because the necessary code modifications are, by design,
+limited to the middle PSy layer. In order to demonstrate this and to
+check for any limitations imposed by the {PS}y{KA}l re-structuring, we
+had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
+used OpenACC directives for this porting as that approach is similar
+to the use of OpenMP directives and works well within the {PS}y{KA}l
+approach.
+
+\subsubsection{Data Movement}
+
+Although the advent of technologies such as NVLink are alleviating the
+bottleneck presented by the connection of the GPU to the CPU, it
+remains critical to minimise data movement between the memory spaces
+of the two processing units. In NEMOLite2D this is achieved by
+performing all computation on the GPU. The whole time-stepping loop
+can then be enclosed inside a single OpenACC \texttt{data} region and it
+is only necessary to bring data back to the CPU for the purposes
+of I/O.
+
+\subsubsection{Kernel Acceleration}
+
+Moving the kernels to execute on the GPU was achieved by using the
+OpenACC \texttt{kernels} directive in each Fortran routine containing
+loops over grid points. (In the {PS}y{KA}l version this is just the
+single {PS}y layer subroutine). This directive instructs the OpenACC
+compiler to automatically create a GPU kernel for each loop nest it
+encounters.
+
+\subsubsection{Force parallelisation of Flather kernels}
+
+The PGI compiler was unable to determine whether the loops applying
+the Flather boundary condition in the $x$ and $y$ directions were safe
+to parallelise. This information therefore had to be supplied by
+inserting a \texttt{loop independent} OpenACC directive before each of
+the two loops.
+
+\subsubsection{Loop collapse}
+
+When using the OpenACC \texttt{kernels} directive the compiler creates
+GPU kernels by parallelising only the outer loop of each loop nest.
+For the majority of loop nests in NEMOLite2D all of the iterations are
+independent and the loop bodies consist of just a handful of
+executable statements. For small kernels the loop start-up cost can be
+significant and therefore it is beneficial to generate as many threads
+as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
+data size is greatre than the number of threads). For this approach to
+be efficient we must expose the maximum amount of parallelism to the
+GPU and we do this by instructing the compiler to parallelise both
+levels (i.e. \texttt{ji} and \texttt{jj}) of a loop nest. This is
+achieved using the OpenACC \texttt{loop collapse(2)} directive for all
+of the smaller kernels in NEMOLite2D.
+
+\subsubsection{Shortloop}
+
+In contrast to all of the other NEMOLite2D kernels, the Momentum
+kernels ($u$ and $v$) are relatively large in terms of the number of
+executable statements they contain. In turn, this means that the
+corresponding kernels will require more state and thus more registers
+on the GPU. Therefore, rather than generating as many threads as will
+fit on the GPU, it is more efficient only to generate as many as
+required by the problem so as to minimise reigster pressure. Once
+slots become free on the GPU, new threads will launch with the
+associated cost amortized by the longer execution time of the larger
+kernel. The compiler can be persuaded
+to adopt the above strategy through the use of the (PGI-specific)
+\texttt{loop shortloop} directive which tells it that there is not
+likely to be much thread re-use in the following loop. This directive
+was applied to both the \texttt{ji} and \texttt{jj} loops in both of
+the Momentum kernels.
+
+\subsubsection{Kernel fusion}
+
+Both of the Momentum kernels read from 20 double-precision arrays and
+thus require considerable memory bandwidth.  However, the majority of
+these arrays are used by both the $u$ and $v$ kernels. It is therefore
+possible to reduce the memory-bandwidth requirements by fusing the two
+kernels together. In practice, this means fusing the two,
+doubly-nested loops so that we have a single loop nest containing both
+the $u$ and $v$ kernel bodies/calls.
+
+\subsubsection{CUDA}
+
+We also experimented with using CUDA directly within the original form
+of NEMOLite2D in order to quantify any performance penalty incurred by
+taking the {PS}y{KA}l/OpenACC approach.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 concludes that such vectorisation would be inefficient). Of the other
-transformations, only the addition of the compiler-specific
-safe\_address directive (see Section~\ref{sec_safe_address}) was found
-to give any benefit.
+transformations, only the change to constant loop bounds and the
+addition of the compiler-specific safe\_address directive (see
+Section~\ref{sec_safe_address}) were found to (slightly) improve
+performance.
 
@@
 Original             & 3.87 & 12.2 & 6.83 \\
-Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
+Vanilla {PS}y{KA}l   & 2.59 & 6.27 & 6.35 \\
+Constant loop bounds & 3.07 & 6.55 & 6.73 \\
 Safe-address         & --   & --   & 6.95 \\

### 2017-06-20 nemolite2d_perf.tex
nemolite2d-paper: update cpu-gpu comparison data. Add discussion of GPU optimisation results.

@@
 check for any limitations imposed by the {PS}y{KA}l re-structuring, we
-had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
-used OpenACC directives for this porting as that approach is similar
-to the use of OpenMP directives and works well within the {PS}y{KA}l
-approach.
+had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. OpenACC
+directives were used as that approach is similar to the use of OpenMP
+directives and works well within the {PS}y{KA}l approach. In order to
+quantify any performance penalty incurred by taking the
+{PS}y{KA}l/OpenACC approach, we experimented with using CUDA directly
+within the original form of NEMOLite2D.
 
@@
 as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
-data size is greatre than the number of threads). For this approach to
+data size is greater than the number of threads). For this approach to
 be efficient we must expose the maximum amount of parallelism to the
@@
 fit on the GPU, it is more efficient only to generate as many as
-required by the problem so as to minimise reigster pressure. Once
+required by the problem so as to minimise register pressure. Once
 slots become free on the GPU, new threads will launch with the
@@
 
-Both of the Momentum kernels read from 20 double-precision arrays and
+Both of the Momentum kernels read from 16 double-precision arrays and
 thus require considerable memory bandwidth.  However, the majority of
@@
 of NEMOLite2D in order to quantify any performance penalty incurred by
-taking the {PS}y{KA}l/OpenACC approach.
+taking the {PS}y{KA}l/OpenACC approach. To do this we used PGI's
+support for CUDA Fortran to create a CUDA kernel for the (fused)
+Momentum kernel. The only significant code change with this approach
+is the explicit set-up of the grid- and block-sizes and the way in
+which the kernel is launched (i.e. use of the \texttt{call kernel $<<<$
+  gridDim1, blockDim1 $>>>$(args)} syntax).
 
@@
 \includegraphics[width=\picwidth]{opt_stages_256}
-\caption{Performance of the {PS}y{KA}l version of NEMOLite2D for the
+\caption{Serial performance of the {PS}y{KA}l version of NEMOLite2D for the
   $256^{2}$ domain at each stage of optimisation. The first (black)
@@
 
-The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
-wish to run NEMOLite2D on different hardware, \textit{e.g.} a
-GPU. This is because the necessary code modifications are, by design,
-limited to the middle PSy layer. In order to demonstrate this and to
-check for any limitations imposed by the {PS}y{KA}l re-structuring, we
-had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
-used OpenACC directives for this porting as that approach is similar
-to the use of OpenMP directives and works well within the {PS}y{KA}l
-apprach. He also experimented with using CUDA directly within the
-original form of NEMOLite2D in order to quantify any performance
-penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
-
-The performance of these two versions of the code is plotted in
-Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring of
-NEMOLite2D did not prove to be a barrier to achieving a performant
+In contrast to the CPU, we have just a single compiler to consider
+when looking at OpenACC performance. We are therefore able to
+investigate the effect of the various optimisations described in
+Section~\ref{sec_oacc_steps} for the full range of problem sizes.  All
+of the reported performance figures were obtained on an NVIDIA Tesla
+K40m GPU running in `boost' mode (enabled with the command
+\texttt{nvidia-smi -ac 3004,875}).
+
+The performance of the various versions of the code is plotted in
+Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring
+of NEMOLite2D did not prove to be a barrier to achieving a performant
 code when using OpenACC directives, we only report the performance of
-the {PS}y{KA}l OpenACC version. For problem sizes up to $256^2$ the
-difference in the performance of the two versions is less than 5\% and
-for the larger domains it is at most 11\%. This demonstrates that the
-performance cost of the {PS}y{KA}l approach in utilising a GPU for
-NEMOLite2D is minimal.
+the {PS}y{KA}l OpenACC versions.
 
@@
 \caption{Performance of the {PS}y{KA}l (OpenACC) and CUDA GPU
-  implementations of NEMOLite2D. The performance for the OpenACC
-  version of the original code is not shown since it is identical to
-  that of the {PS}y{KA}l version.}
+  implementations of NEMOLite2D. Optimisations are added to the code
+  in an incremental fashion. The performance for the OpenACC version
+  of the original code is not shown since it is identical to that of
+  the Vanilla {PS}y{KA}l version.}
 \label{FIG_gpu_opt_stages}
 \end{figure}
+
+The smallest domain ($64^2$) does not contain sufficient parallelism
+to fully utilise the GPU and only the parallelisation of the Flather
+kernels has much effect on performance. In fact, this is a significnt
+optimisation for all except the largest domain size where only changes
+to the Momentum kernel yield significant gains. Collapsing the 2D loop
+nests has a small but beneficial effect for the majority of problem
+sizes. Gains here are limited by the fact that this optimisation was
+only applied to the non-Momentum kernels which only account for som
+30\% of the run-time.
+
+Given the significance of the two Momentum kernels in the execution
+profile it is no surprise that their optimisation yields the most
+significant gains. Using the \texttt{shortloop} directive gives
+roughly a 25\% increase in performance for domains of $512^2$ and
+greater. Fusing the two momentum kernels gives a further boost of
+around 5\%.
+
+Note that all of the previous optimisations are restricted to the PSy
+layer and in most cases are simply a case of adding directives or
+clauses to directives. However, the question then arises as to the
+cost of restricting optimisations to the PSy layer. In particular, how
+does the performance of the OpenACC version of NEMOLite2D compare with
+a version where those restrictions are lifted?
+Figure~\ref{FIG_gpu_opt_stages} also shows the performance of the
+version of the code where the (fused) Momentum kernel has been
+replaced by a CUDA kernel. For domains up to $256^2$ the difference in
+the performance of the two versions is less than 5\% and for the
+larger domains it is at most 12\%. This demonstrates that the
+performance cost of the {PS}y{KA}l approach in utilising a GPU for
+NEMOLite2D is minimal. Discussions with NVIDIA regarding the
+discrepancy in performance between the CUDA and OpenACC versions are
+also on-going.
 

### 2017-06-22 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-paper: specify roofline is for Ivy Bridge and explain lack of SIMD performance benefit.

@@
 executable statements. For small kernels the loop start-up cost can be
-significant and therefore it is therefore beneficial to generate as
-many threads as will fit on the GPU and then re-use them as
-required. For this approach to be efficient we must expose the maximum
-amount of parallelism to the GPU and we do this by instructing the
-compiler to parallelise both levels (i.e. \texttt{ji} and \texttt{jj})
-of a loop nest. This is achieved using the OpenACC \texttt{loop
-  collapse(2)} directive for all of the smaller kernels in NEMOLite2D.
+significant and therefore it is beneficial to generate as many threads
+as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
+data size is greater than the number of threads). For this approach to
+be efficient we must expose the maximum amount of parallelism to the
+GPU and we do this by instructing the compiler to parallelise both
+levels (i.e. \texttt{ji} and \texttt{jj}) of a loop nest. This is
+achieved using the OpenACC \texttt{loop collapse(2)} directive for all
+of the smaller kernels in NEMOLite2D.
 
@@
 fit on the GPU, it is more efficient only to generate as many as
-required by the problem. Once slots become free on the GPU new threads
-will launch. This minimises the overhead associated with thread
-creation and eases register pressure.  The compiler can be persuaded
+required by the problem so as to minimise register pressure. Once
+slots become free on the GPU, new threads will launch with the
+associated cost amortized by the longer execution time of the larger
+kernel. The compiler can be persuaded
 to adopt the above strategy through the use of the (PGI-specific)
@@
 Gnu-compiled binary is only 6\% slower than that produced by the Intel
-compiler and is 13\% \textit{faster} that that of the Cray compiler. In
+compiler and is 13\% faster that that of the Cray compiler. In
 part this situation comes about because of the effect that adding the
@@
 addressed how efficient the original actually is. In order to do so we
+consider the performance of the code on the Intel Ivy Bridge CPU and
 use the Roofline Model~\citep{roofline} which provides a relatively
@@
 whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\cite{para_pearls} and use the
-STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
-get the upper bounds on the memory bandwidth and floating-point
-operations per second (FLOPS) for the test hardware. Since we are
-using an Intel Ivybridge CPU we used the Intel Math Kernel Library
-implementation of LINPACK.
+follow the approach suggested in~\cite{para_pearls} and construct a
+roofline model using the STREAM~\citep{stream} and
+LINPACK~\citep{linpack} benchmarks in order to get the appropriate
+upper bounds on the memory bandwidth and floating-point operations per
+second (FLOPS). Since we are using an Intel Ivybridge CPU we used the
+Intel Math Kernel Library implementation of LINPACK.
 
 A key component of the Roofline model is the Operational or Arithmetic
-Intensity of the code being executed:
+Intensity (AI) of the code being executed:
 \[
@@
 In order to aid our understanding of kernel performance we have
-developed a tool capable of parsing Fortran code and generating a
-Directed Acyclic Graph (DAG) of the data flow. This tool eases the
-laborious and error-prone process of counting memory accesses and
-FLOPs as well as providing information on those operations that are
-rate-limiting or on the critical path. Using the details of the Intel
-Ivy Bridge microarchitecture published by Fog~\citep{fog_microarch,
+developed a tool, ``Habakkuk'' (https://github.com/arporter/habakkuk),
+capable of parsing Fortran code and generating a Directed Acyclic
+Graph (DAG) of the data flow. Habakkuk eases the laborious and
+error-prone process of counting memory accesses and FLOPs as well as
+providing information on those operations that are rate-limiting or on
+the critical path. Using the details of the Intel Ivy Bridge
+microarchitecture published by Fog~\citep{fog_microarch,
   fog_instructions} we have constructed performance estimates of the
@@
 vectorisation which gives a performance increase of a factor of two.
+Further investigation of this issue revealed that several of the
+NEMOLite2D kernels make frequent use of floating-point division
+operations. Although SSE and AVX versions of the division operation
+are available, on Ivy Bridge they do not provide any performance
+benefit~\citep{fog_instructions}. A straightforward optimisation then
+would be to alter the kernels in order to reduce the number of
+division operations. However, kernel optimisation is outside the scope
+of this paper since it breaks the PSyKAl separation of concerns.
 

### 2017-06-22 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: final pre-submission tweaks.

@@
 15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
-was compiled using version 15.5 of the PGI compiler.
+was compiled using version 15.10 of the PGI compiler.
 
@@
 
-%\begin{figure}
-%\centering
-%\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
-%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-%  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-%  socket. The corresponding parallel efficiencies are shown using open
-%  symbols and dashed lines. The 24-thread runs employed hyperthreading
-%  and the optimal OpenMP schedule is given in parentheses.}
-%\label{FIG_gnu_omp_32}
-%\end{figure}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\picwidth]{omp_scaling_32_intel}
-%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-%  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
-%  socket.  The corresponding parallel efficiencies are shown using
-%  open symbols and dashed lines. The 24-thread runs employed
-%  hyperthreading and the optimal OpenMP schedule is given in
-%  parentheses.}
-%\label{FIG_intel_omp_32}
-%\end{figure}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\picwidth]{omp_scaling_32_cray}
-%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-%  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
-%  socket.  The corresponding parallel efficiencies are shown using
-%  open symbols and dashed lines. The 24-thread runs employed
-%  hyperthreading and the optimal OpenMP schedule is given in
-%  parentheses.}
-%\label{FIG_cray_omp_32}
-%\end{figure}
-
 \subsubsection{OpenACC}
 
-In contrast to the CPU, we have just a single compiler to consider
-when looking at OpenACC performance. We are therefore able to
-investigate the effect of the various optimisations described in
-Section~\ref{sec_oacc_steps} for the full range of problem sizes.  All
-of the reported performance figures were obtained on an NVIDIA Tesla
-K40m GPU running in `boost' mode (enabled with the command
-\texttt{nvidia-smi -ac 3004,875}).
+In contrast to the CPU, we only had access to the PGI compiler when
+looking at OpenACC performance. We therefore investigate the effect of
+the various optimisations described in Section~\ref{sec_oacc_steps}
+for the full range of problem sizes.  All of the reported performance
+figures were obtained on an NVIDIA Tesla K40m GPU running in `boost'
+mode (enabled with the command \texttt{nvidia-smi -ac 3004,875}).
 
 The performance of the various versions of the code is plotted in
-Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring
-of NEMOLite2D did not prove to be a barrier to achieving a performant
-code when using OpenACC directives, we only report the performance of
-the {PS}y{KA}l OpenACC versions.
+Figure~\ref{FIG_gpu_opt_stages}.  We do not show the performance of
+the OpenACC version of the original code as it was identical to that
+of the Vanilla {PS}y{KA}l version.
 
@@
 to fully utilise the GPU and only the parallelisation of the Flather
-kernels has much effect on performance. In fact, this is a significnt
+kernels has much effect on performance. In fact, this is a significant
 optimisation for all except the largest domain size where only changes
-to the Momentum kernel yield significant gains. Collapsing the 2D loop
+to the Momentum kernel yield sizeable gains. Collapsing the 2D loop
 nests has a small but beneficial effect for the majority of problem
 sizes. Gains here are limited by the fact that this optimisation was
-only applied to the non-Momentum kernels which only account for som
-30\% of the run-time.
+only beneficial for the non-Momentum kernels and these only account
+for some 30\% of the run-time.
 
@@
 performance cost of the {PS}y{KA}l approach in utilising a GPU for
-NEMOLite2D is minimal. Discussions with NVIDIA regarding the
-discrepancy in performance between the CUDA and OpenACC versions are
-also on-going.
+NEMOLite2D is minimal.  The discrepancy in performance between the
+CUDA and OpenACC versions has been fed back to the PGI compiler team.
 
@@
 
-%Peak memory bandwidth on the K40 GPU is 288 GB/s and we measured
-%(using STREAM2) bandwidths of 46.0 and 20.5 GB/s to L3 and main
-%memory, respectively on the CPU.
-
 \subsection{Performance Analysis with the Roofline Model}
@@
 version of NEMOLite2D compares with that of the original, we have not
-addressed how efficient the original actually is. In order to do so we
-consider the performance of the code on the Intel Ivy Bridge CPU and
-use the Roofline Model~\citep{roofline} which provides a relatively
-simple way of characterising the performance of a code in terms of
-whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\cite{para_pearls} and construct a
-roofline model using the STREAM~\citep{stream} and
-LINPACK~\citep{linpack} benchmarks in order to get the appropriate
-upper bounds on the memory bandwidth and floating-point operations per
-second (FLOPS). Since we are using an Intel Ivybridge CPU we used the
-Intel Math Kernel Library implementation of LINPACK.
+addressed how efficient the original actually is. Without this
+information we have no way of knowing whether further optimisations
+might yield worthwhile performance improvements. Therefore, we
+consider the performance of the original, serial NEMOLIte2D code on
+the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
+which provides a relatively simple way of characterising the
+performance of a code in terms of whether it is memory-bandwidth bound
+or compute bound. To do so we follow the approach suggested
+in~\cite{para_pearls} and construct a roofline model using the
+STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order
+to obtain appropriate upper bounds on the memory bandwidth and
+floating-point operations per second (FLOPS), respectively. Since we
+are using an Intel Ivy Bridge CPU we used the Intel Math Kernel Library
+implementation of LINPACK.
 
@@
 might have been assumed by its low AI. Therefore, although the
-performance of this kernel is clearly sub-optimal, this is due to the
-balance of floating-point operations that it contains. Improving this
-is outside the scope of the current work since here we are focused on
-the {PS}y{KA}l separation of concerns and the introduction of parallelism
-in the PSy layer.
+performance of this kernel is well below the peak performance of the
+CPU, this is due to the balance of floating-point operations that it
+contains. Improving this is outside the scope of the current work
+since here we are focused on the {PS}y{KA}l separation of concerns and
+the introduction of parallelism in the PSy layer.
 
@@
 corresponding kernel performance no longer falls within the bounds of
-our performance estimate. This is because we have incorporated the
-effect of SSE into that estimate by simply assuming perfect
-vectorisation which gives a performance increase of a factor of two.
-Further investigation of this issue revealed that several of the
-NEMOLite2D kernels make frequent use of floating-point division
-operations. Although SSE and AVX versions of the division operation
-are available, on Ivy Bridge they do not provide any performance
-benefit~\citep{fog_instructions}. A straightforward optimisation then
-would be to alter the kernels in order to reduce the number of
-division operations. However, kernel optimisation is outside the scope
-of this paper since it breaks the PSyKAl separation of concerns.
+the performance estimates produced by Habakkuk. This is because we
+have incorporated the effect of SSE into that estimate by simply
+assuming perfect vectorisation which gives a performance increase of a
+factor of two.  Further investigation of this issue revealed that
+several of the NEMOLite2D kernels make frequent use of floating-point
+division operations. Although SSE and AVX versions of the division
+operation are available, on Ivy Bridge they do not provide any
+performance benefit~\citep{fog_instructions}. A straightforward
+optimisation then would be to alter the kernels in order to reduce the
+number of division operations. However, again, kernel optimisation is
+outside the scope of this paper since it breaks the PSyKAl separation
+of concerns.
 
@@
 below any of the traditional roofline ceilings. We have developed
-Habakkuk (https://github.com/arporter/habakkuk), a code-analysis tool
+Habakkuk, a code-analysis tool
 that is capable of providing more realistic ceilings by analysing the
@@
 \codeavailability{The NEMOLite2D package is available as the GOcean
- 2.0 benchmark, available from https://puma.nerc.ac.uk/trac/GOcean.}
+  2.0 benchmark, available from
+  https://puma.nerc.ac.uk/trac/GOcean. Habakkuk is available from
+  https://github.com/arporter/habakkuk.}
 

### 2017-06-23 docs/papers/nemolite2d_performance/nemolite2d_perf_abstract.tex
nemolite-paper: add separate abstract tex file

--- 2017-05-31 nemolite2d_perf.tex
@@
 
-\copyrightstatement{Copyright Science and Technology Facilities Council, 2017}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\introduction
-
-The challenge presented to the developers of scientific software by
-the drive towards Exascale computing is considerable. With power
-consumption becoming the overriding design constraint, CPU clock
-speeds are falling and the complex, multi-purpose compute core is
-being replaced by multiple, simpler cores. This philosophy can be seen
-at work in the rise of so-called accelerator-based machines in the Top
-500 List of supercomputers (http://www.top500.org/): six of the
-top-ten machines in the November 2016 list make use of many-core
-processors (Intel Xeon Phi, NVIDIA GPU or NRCPC SW26010). Two of the
-remaining four machines are IBM BlueGene/Qs, the CPU of which has
-hardware support for running 64 threads.
-
-Achieving good performance on large numbers of light-weight cores
-requires exploiting as much parallelism in an application as possible
-and this results in increased complexity in the programming models
-that must be used. This in turn increases the burden of code
-maintenance and code development, in part because two specialisms are
-required: that of the scientific domain which a code is modelling
-(\textit{e.g.} oceanography) and that of computational science. The
-situation is currently complicated still further by the existence of
-competing hardware technology; if one was to begin writing a major
-scientific application today it is unclear whether one would target
-GPU, Xeon Phi, traditional CPU, FPGA or something else entirely. This
-is a problem because, generally speaking, these different technologies
-require different programming approaches.
-
-In a previous paper~\citep{shallow_psykal} we introduced a possible
-approach to tackling this problem which we term {PS}y{KA}l (discussed
-below). In that work we considered the implications for serial
-performance of the extensive code re-structuring required by the
-approach when applied to the `Shallow' shallow-water model
-(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
-re-structuring did initially incur a sizeable performance penalty, it
-was possible to transform the resulting code to recover performance
-(for a variety of CPU/compiler combinations) while obeying the
-{PS}y{KA}l separation of concerns. In this work we move to looking at
-portable \textit{parallel} performance within the {PS}y{KA}l approach.
-
-\subsection{The {PS}y{KA}l Approach}
-
-The {PS}y{KA}l approach attempts to address the problems described in
-the previous section. It separates code into three layers; the
-Algorithm layer, the PSy layer and the Kernel layer. The approach has
-been developed in the GungHo project~\citep{GungHo}, which is creating
-a new Dynamical core for the UK Met Office, and its design has been
-influenced by earlier work on OP2~\citep{OP2, PYOP2}.
-
-In common with OP2, the {PS}y{KA}l approach separates out the science
-code and the performance-related code into distinct layers. The calls
-that specify parallelism in both approaches are similar in terms of
-where they are placed in the code and in their semantics. However, the
-{PS}y{KA}l approach supports the specification of more than one kernel
-in a parallel region of code, compared with one for OP2, giving more
-scope for optimisation. In addition, the metadata describing a kernel
-is included with the kernel code in the {PS}y{KA}l approach whereas it
-is provided as a part of the kernel call in OP2.
-
-While the {PS}y{KA}l approach is general, we are currently applying it to
-Atmosphere and Ocean models written in Fortran where domain
-decomposition is typically performed in the latitude-longitude
-dimension, leaving columns of elements on each domain-decomposed
-partition.
-
-The top layer, in terms of calling hierarchy, is the Algorithm
-layer. This layer specifies the algorithm that the scientist would like
-to perform (in terms of calls to kernel and infrastructure routines)
-and logically operates on full fields. We say logically here as the
-fields may be domain decomposed, however the Algorithm layer is not
-aware of this. It is the scientist's responsibility to write this
-Algorithm layer.
-
-The bottom layer, in terms of calling hierarchy, is the Kernel
-layer. The Kernel layer implements the science that the Algorithm
-layer calls, as a set of subroutines. These kernels operate on fields
-that are local to the process doing the computation. (Depending on the
-type of kernel, these may be a set of elements, a single column of
-elements, or a set of columns.) Again the scientist is responsible for
-writing this layer and there is no parallelism specified here, but,
-depending on the complexity of the Kernels, there may be input from an
-HPC expert and/or some coding rules to help ensure that the kernels
-compile into efficient code.  In an alternative
-approach~\citep{firedrake,fenics}, kernels are generated from a
-high-level specification, potentially allowing them to be optimised
-automatically~\citep{coffee}.
-
-The PSy layer sits in-between the Algorithm and Kernel layers and its
-functional role is to link the algorithm calls to the associated
-kernel subroutines. As the Algorithm layer works on logically global
-fields and the Kernel layer works on local fields, the PSy layer is
-responsible for iterating over columns. It is also responsible for
-including any distributed-memory operations resulting from the
-decomposition of the simulation domain, such as halo swaps and
-reductions.
-
-As the PSy layer iterates over columns, the single-core performance
-can be optimised by applying transformations such as loop fusion and
-loop blocking. Additionally, the potential parallelism within this
-iteration space can also be exploited and optimised. The PSy layer can
-therefore be tailored for a particular hardware (such as multi-core,
-many-core, GPUs, or some combination thereof) and software (such as
-compiler, operating system, MPI library, \textit{etc}.) configuration
-with no change to the Algorithm or Kernel layer code. This approach
-therefore offers the potential for portable performance. In this work
-we apply optimisations to the PSy layer manually. The development of a
-tool to automate this process will be the subject of a future paper.
-
-Clearly the separation of code into distinct layers may have an effect
-on performance. This overhead, how to get back to the performance of a
-parallel, hand-optimised code, and potentially improve on it, will be
-discussed in the remainder of this paper.
-
-\subsection{The `NEMOLite2D' Program}
-
-For this work we have used a program, `NEMOLite2D,' developed by
-ourselves (https://puma.nerc.ac.uk/trac/GOcean).  NEMOLite2D is a
-vertically-averaged version of NEMO (Nucleus for European Modelling of
-the Ocean~\citep{nemo}), retaining only its dynamical part. The whole
-model system is represented through one continuity equation
-\eqref{continuity} (for the update of the sea-surface height) and two
-vertically-integrated momentum equations \eqref{momentum} (for the two
-velocity components, respectively).
-
-\begin{equation}\label{continuity}
- \frac{\partial \zeta}{\partial t} + \vec{\nabla} \cdot (\vec{U}h) = 0
-\end{equation}
-\begin{equation}\label{momentum}
- \frac{\partial \vec{U}h}{\partial t}
-   + \vec{U} \cdot \vec{\nabla} (\vec{U}h) = 
-   - gh \vec{\nabla} \zeta - 2h \vec{\Omega} \times \vec{u}
-   + \nu h \Delta \vec{U}
-\end{equation}
-
-where $\zeta$ and $\vec{U}$ represent the sea-surface height and
-horizontal velocity vector, respectively. $h$ is the total water
-depth, $\Omega$ is the Earth rotation velocity vector.  $g$ is the
-acceleration due to gravity and $\nu$ is the kinematic viscosity coefficient.
-
-The external forcing includes surface wind stress, bottom friction,
-and open-boundary barotropic forcing. A lateral-slip boundary
-condition is applied along the coast lines. The open boundary
-condition can be set as a clipped or Flather's radiation
-condition~\citep{flather76}. The bottom friction takes a semi-implicit
-form for the sake of model stability. As done in the original version
-of NEMO, a constant or Smagorinsky horizontal viscosity coefficient is
-used for the horizontal viscosity term.
-
-The traditional Arakawa C structured grid is employed here for the
-discretisation of the computational domain. A two-dimensional integer
-array is used to identify the different parts of the computational
-domain; it has the value of one for ocean, zero for land and minus
-one for ocean cells outside of the computational domain. This array
-enables the identification of ocean cells, land cells, solid boundaries
-and open boundaries.
-
-For the sake of simplicity, the explicit Eulerian forward time
-stepping method is implemented here, except that the bottom friction
-takes a semi-implicit form.  The Coriolis force can be set in explicit
-or implicit form. The advection term is computed with a first-order
-upwind scheme.
-
-The sequence of the model computation is as follows:
-\begin{enumerate}
-\item Set the initial conditions (water depth, sea surface height, velocity)
-\item integrate the continuity equation for the new sea surface height
-\item update the different terms in the right hand side of the momentum equations; advection, Coriolis forcing (if set in explicit form), pressure gradient, and horizontal viscosity
-\item update the velocity vectors by summing up the values in 3), and implicitly with the bottom friction and Coriolis forcing (if set in implicit form)
-\item apply the boundary conditions on the open- and solid-boundary cells.
-\end{enumerate}
-
-Since any real oceanographic computational model must output results,
-we ensure that any {PS}y{KA}l version of NEMOLite2D retains the Input/Output
-capability of the original. This aids in limiting the optimisations
-that can be performed on the {PS}y{KA}l version to those that should also
-be applicable to full oceanographic models. Note that although we
-retain the I/O functionality, all of the results presented in this work
-carefully exclude the effects of I/O since it is compute performance
-that interests us here.
-
-In the Algorithm layer, fields (and grids) are treated as logically
-global objects. Therefore, as part of creating the {PS}y{KA}l version
-of NEMOLite2D, we represent fields with derived types instead of
-arrays in this layer. These types then hold information about the
-associated mesh and the extents of 'internal' and 'halo' regions as
-well as the data arrays themselves. This frees the natural scientist
-from having to consider these issues and allows for a certain degree
-of flexibility in the actual implementation (e.g. padding for
-alignment or increasing array extent to allow for other
-optimisations). The support for this is implemented as a library
-(which we term the GOcean Infrastructure) and is common to the
-{PS}y{KA}l versions of both NEMOLite2D and Shallow.
-
-In order to maximise the flexibility (and thus potential for
-optimisation) of the {PS}y{KA}l version of NEMOLite2D, we made the
-kernels as fine-grained as possible. This resulted in eight distinct
-kernels, each of which updates a single field at a single point (since
-we have chosen to use point-wise kernels). With a little bit of
-tidying/re-structuring, we found it was possible to express the
-contents of the main time-stepping loop as a single invoke (a call to
-the PSy layer) and a call to the I/O system
-(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
-gives us a single PSy-layer routine which consists of applying each of
-the kernels to all of the points requiring an update on the model
-mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
-routine then contains a doubly-nested loop (over the two dimensions of
-the model grid) around each kernel call.
-
-\begin{figure}
-\centering
-\begin{verbatim}
-DO istep = 1, nsteps
-  call invoke( continuity(ssha_t, ...), &
-               momentum_u(ua, un, ...), &
-               ...,                     &
-               next_sshu(sshn_u, ...),  &
-               next_sshv(sshn_v, ...) )
-  call model_write(istep, sshn_t, un, vn)
-END DO
-\end{verbatim}
-\caption{A schematic of the top-level of the {PS}y{KA}l version of the
-  NEMOLite2D code. The kernels listed as arguments to the \texttt{invoke}
-call specify the operations to be performed.}
-\label{FIG_psykal_nemolite2d_structure}
-\end{figure}
-
-As with any full oceanographic model, boundary conditions must be
-applied at the edges of the model domain. Since NEMOLite2D applies
-external boundary conditions (\textit{e.g.} barotropic forcing), this is
-done via user-supplied kernels.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methodology}
-\label{sec_methodology}
-
-Our aim in this work is to achieve portable performance, especially
-between multi-core CPU and many-core GPU systems. Consequently, we
-have performed tests on both an Intel Ivy Bridge CPU (E5-2697 at
-2.7~GHz) and on an NVIDIA Tesla K40 GPU.  On the Intel-based system we
-have used the Gnu, Intel and Cray Fortran compilers (versions 4.9.1,
-15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
-was compiled using version 15.5 of the PGI compiler.
-
-We first describe the code transformations performed for the serial
-version of NEMOLite2D.  We then move on to the construction of
-parallel versions of the code using OpenMP and OpenACC. Again, we
-describe the key steps we have taken in this process in order to
-maximise the performance of the code. In both cases our aim is to
-identify those transformations which must be supported by any tool
-which seeks to auto-generate a performant PSy layer.
-
-\subsection{Transformations of Serial NEMOLite2D}
-
-In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
-used with each compiler. For the Gnu and Intel compilers, we
-include flags to encourage in-lining of kernel bodies. The Intel flag
-``-xHost'' enables the highest-level of SIMD vectorisation supported
-by the host CPU (AVX in this case). The Intel flag ``-fast'' and Cray
-flags ``-O ipa5'' and ``-h wp'' enable inter-procedural optimisation.
-
-\begin{table}[!t]
-\caption{The compiler flags used in this work.}
-\label{TABLE_compiler_flags}
-\centering
-\begin{tabular}{l|l}
-\hline
-Compiler  &  Flags                                                       \\
-\hline
-Gnu       & -Ofast -mtune=native -finline-limit=50000                    \\
-Intel     & -O3 -fast -fno-inline-factor -xHost                          \\
-Cray      & -O3 -O ipa5 -h wp                                            \\
-PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 \\
-\hline
-\end{tabular}
-\end{table}
-
-Before applying any code transformations, we first benchmark the
-original, serial version of the code. We also benchmark the vanilla,
-unoptimised version after it has been re-structured following the
-{PS}y{KA}l approach. In addition to this benchmarking, we profile these
-versions of the code at the algorithm level (using a high-level timing
-API). The resulting profiles are given in
-Table~\ref{TABLE_profile}. The Momentum section dominates the profile
-of both versions of the code, accounting for around 70--80\% of the
-wall-clock time spent doing time-stepping. It is also the key section
-when we consider the performance loss when moving from the original to
-the {PS}y{KA}l version of the code; it slows down by a factor of
-two. Although less significant in terms of absolute time, the
-Continuity section is also dramatically slower in the {PS}y{KA}l version,
-this time by a factor of three. In contrast, the performance of the
-Time-update and Boundary Condition regions are not significantly
-affected by the move to the {PS}y{KA}l version.
-
-\begin{table}[!t]
-\caption{The performance profile of the original and {PS}y{KA}l versions
-  of NEMOLite2D on the Intel Ivy Bridge CPU (for 2000 time-steps of the
-  $128^{2}$ domain and the Intel compiler).}
-\label{TABLE_profile}
-\centering
-\begin{tabular}{l|c|c|c|c|c|c}
-\hline
-Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla {PS}y{KA}l} &
-\multicolumn{2}{c}{Final {PS}y{KA}l} \\
-        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
-\hline
-Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
-Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
-BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
-Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
-\hline
-\end{tabular}
-\end{table}
-
-Beginning with the vanilla {PS}y{KA}l version, we then apply a series of
-code transformations while obeying the {PS}y{KA}l separation of concerns,
-\textit{i.e.} optimisation is restricted to the middle, {PS}y layer and
-leaves the kernel and algorithm layers unchanged. The aim of these
-optimisations is to recover, as much as is possible, the performance
-of the original version of the code. The transformations we have
-performed and the reasons for them are described in the following sections.
-
-\subsubsection{Constant loop bounds}
-
-In the vanilla {PS}y layer, the lower and upper bounds for each loop
-over grid points are obtained from the relevant components of the
-derived type representing the field being updated by the kernel being
-called from within the loop. In our previous
-work~\citep{shallow_psykal} we found that the Cray compiler in
-particular produced more performant code if we changed the PSy layer
-such that the array extents are looked up once at the beginning of the
-PSy routine and then used to specify the loop bounds. We have therefore
-applied that transformation to the PSy layer of NEMOLite2D.
-
-\subsubsection{Addition of \texttt{safe\_address} directives}
-\label{sec_safe_address}
-
-Much of the optimisations we have performed have been informed by the
-diagnostic output produced by either the Cray or Intel compilers. Many
-of the NEMOLite2D kernels contain conditional statements. These
-statements are there to check whether \textit{e.g.} the current grid
-point is wet or neighbours a boundary point. A compiler is better able
-to optimise such a loop if it can be sure that all array accesses
-within the body of the loop are safe for every trip, irrespective of
-the conditional statements. In its diagnostic output the Cray compiler
-notes this with messages of the form:
-\begin{verbatim}
-A loop starting at line 448 would benefit 
-from "!dir$ safe_address".
-\end{verbatim}
-In order that we could safely add such directives we altered the
-GOcean infrastructure to allocate all field-data arrays with extents
-greater than strictly required.  We were then able to safely add the
-\texttt{safe\_address} before all of the loops where the Cray compiler
-indicated it might be useful (the Momentum loops and some of the BC
-loops).
-
-\subsubsection{In-line Momentum kernel bodies into middle-layer}
-\label{sec_inline_mom}
-
-The profiling data in Table~\ref{TABLE_profile} shows that it is the
-Momentum section that accounts for the bulk of the model run-time.  We
-therefore chose to attempt to optimise this section first. In-keeping
-with the {PS}y{KA}l approach, we are only permitted to optimise the middle
-(PSy) layer which, for this section comprises calls to two kernels,
-one for each of the $x$ and $y$ components of momentum. These kernels
-are relatively large; each comprises roughly 85 lines of Fortran
-executable statements.
-
-From our previous work~\citep{shallow_psykal} on a similar code we know
-that kernel in-lining is critical to obtaining performance with both
-the Gnu and Intel compilers. For the Gnu compiler, this is because it
-cannot perform in-lining when routines are in separate source
-files. In our previous work we obtained an order-of-magnitude speed-up
-simply by moving subroutines into the module containing the middle
-layer (from which the kernels are called). A further performance
-improvement of roughly 30\% was obtained when the kernel code was
-manually inserted at the site of the subroutine call.
-
-Although the Intel compiler does do in-lining when routines are in
-separate source files, we have found (both here and in our previous
-work~\citep{shallow_psykal}) that the extent of the optimisations it
-performs is reduced if it first has to in-line a routine. For the
-Intel-compiled Shallow code, manually inserting kernel code at the
-site of the subroutine call increased performance by about 25\%.
-
-In fact, in-lining can have a significant effect on the Intel
-compiler's ability to vectorise a loop. Taking the loop that calls the
-kernel for the u-component of momentum as an example, before in-lining
-the compiler reports:
-\begin{verbatim}
-LOOP BEGIN at time_step_mod.f90(85,7) 
-  inlined into nemolite2d.f90(86,11)
- remark #15335: loop was not vectorized: 
-  vectorization possible but seems 
-  inefficient. 
- --- begin vector loop cost summary ---
- scalar loop cost: 1307 
- vector loop cost: 2391.000 
- estimated potential speedup: 0.540 
- ...
- --- end vector loop cost summary ---
-LOOP END
-\end{verbatim}
-After we have manually in-lined the kernel body, the compiler reports:
-\begin{verbatim}
-LOOP BEGIN at time_step_mod.f90(97,7) 
-  inlined into nemolite2d.f90(86,11)
- LOOP WAS VECTORIZED
- --- begin vector loop cost summary ---
- scalar loop cost: 1253 
- vector loop cost: 521.750 
- estimated potential speedup: 2.350 
- ...
- --- end vector loop cost summary ---
-LOOP END
-\end{verbatim}
-Looking at the `estimated potential speedup' in the compiler output
-above, it is clear that the way in which the compiler vectorises the
-two versions must be very different. This conclusion is borne out by
-the fact that if one persuades the compiler to vectorise the first
-version (through the use of a directive) then the performance of the
-resulting binary is worse than that where the loop is left
-un-vectorised. In principle this could be investigated further by
-looking at the assembler that the Intel compiler generates but that is
-outside the scope of this work.
-
-For the best possible performance, we have therefore chosen to do full, manual
-inlining for the two kernels making up the Momentum section.
-
-\subsubsection{Force SIMD vectorisation of the Momentum kernels 
- using directives}
-
-It turns out that the Cray-compiled binaries of both the original and
-{PS}y{KA}l versions of NEMOLite2D perform considerably less well than
-their Intel-compiled counterparts. Comparison of the diagnostic output
-from each of the compilers revealed that while the Intel compiler was
-happy to vectorise the Momentum loops, the Cray compiler was choosing
-not to:
-\begin{verbatim}
-  99. do ji = 2, M-1, 1
- A loop starting at line 99 was blocked 
- with block size 256.
-
- A loop starting at line 99 was not
- vectorized because it contains 
- conditional code which is more
- efficient if executed in scalar mode.
-\end{verbatim}
-Inserting the Cray \texttt{vector always} directive persuaded the compiler
-to vectorise the loop:
-\begin{verbatim}
- 99. !dir$ vector always
- 100. do ji = 2, M-1, 1
- A loop starting at line 100 was blocked
- with block size 256.
-
- A loop starting at line 100 requires an 
- estimated 17 vector registers at 
- line 151; 1 of these have been 
- preemptively forced to memory.
-
- A loop starting at line 100 was vectorized.
-\end{verbatim}
-which gave a significant performance improvement. This behaviour is in
-contrast to that obtained with the Intel compiler: its predictions
-about whether vectorising a loop would be beneficial were generally
-found to be reliable.
-
-\subsubsection{Work around limitations related to derived types}
-
-Having optimised the Momentum section as much as permitted by the
-{PS}y{KA}l approach, we turn our attention to the three remaining
-sections of the code. The profile data in Table~\ref{TABLE_profile}
-shows that these regions are all comparable in terms of cost. What is
-striking however, is that the cost of the Continuity section increases
-by more than a factor of three in moving to the {PS}y{KA}l version of the
-code.
-
-Comparison of the diagnostic output from the Cray and Intel
-compilers revealed that the Cray compiler was vectorising the
-Continuity section while the Intel compiler reported that it was
-unable to do so due to dependencies. After some experimentation we
-found that this was due to limitations in the compiler's analysis of
-the way components of Fortran derived types were being used. Each
-GOcean field object, in addition to the array holding the local
-section of the field, contains a pointer to a GOcean grid object. If a
-kernel requires grid-related quantities (\textit{e.g.} the grid spacing)
-then these are obtained by passing it a reference to the appropriate
-array within the grid object. Although these grid-related quantities
-are read-only within a compute kernel, if they were referenced from the
-same field object as that containing an array to which the kernel writes
-then the Intel compiler identified a dependency preventing
-vectorisation.  This limitation was simply removed by ensuring that
-all read-only quantities were accessed via field objects that were
-themselves read-only for the kernel at hand.  For instance, the call
-to the continuity kernel, which confused the Intel compiler, originally
-looked like this:
-\begin{verbatim}
-   call continuity_code(ji, jj,      &
-                        ssha%data,   &
-                        sshn_t%data, &
-                        ...,         &
-                        ssha%grid%area_t)
-\end{verbatim}
-where \textit{ssha} is the only field that is written to by the kernel.
-We remove any potential confusion by instead obtaining the
-grid-related (read-only) quantities from a field (\textit{sshn\_t} in
-this case) that is only read by the kernel:
-\begin{verbatim}
-   call continuity_code(ji, jj,      &
-                        ssha%data,   &
-                        sshn_t%data, &
-                        ...,         &
-                        sshn_t%grid%area_t)
-\end{verbatim}
-
-\subsubsection{In-line the Continuity kernel}
-\label{sec_cont_inline}
-
-As with the Momentum kernel, we know that obtaining optimal
-performance from both the Gnu and Intel compilers requires that a
-kernel be manually in-lined at its call site. We do this for the
-Continuity kernel in this optimisation step.
-
-\subsubsection{In-line remaining kernels (BCs and time-update)}
-\label{sec_inline_all_kernels}
-
-Having optimised the Continuity section we finally turn our attention
-to the Boundary Condition and Time-update sections. The kernels in
-these sections are small and dominated by conditional statements.  We
-therefore limited our optimisation of them to manually in-lining each
-of the kernels into the PSy layer.
-
-\subsubsection{In-line field-copy operations}
-\label{sec_inline_fldcopy}
-
-The Time-update section includes several array copies where fields for
-the current time-step become the fields at the previous time-step.
-Initially we implemented these copies as 'built-in' kernels (in the
-GOcean library) as they are specified in the algorithm layer. However,
-we obtained better performance by simply in-lining these array copies
-into the PSy layer.
-
-We shall see that the transformations we have just described do not
-always result in improved performance. Whether or not they do so
-depends both on the compiler used and the problem size. We also
-emphasise that the aim of these optimisations is to make the PSy layer
-as compiler-friendly as possible, following the lessons learned from
-our previous work with the Shallow code~\citep{shallow_psykal}. It may
-well be that transforming the code into some other structure would
-result in better performance on a particular architecture. However,
-exploring this optimisation space is beyond the scope of the present
-work.
-
-We explore the extent to which performance depends upon the problem
-size by using square domains of dimension 64, 128, 256, 512 and 1024
-for the traditional, cache-based CPU systems. This range allows us to
-investigate what happens when cache is exhausted as well as giving us
-some insight into the decisions that different compilers make when
-optimising the code.
-
-\subsection{Construction of OpenMP-Parallel NEMOLite2D}
-\label{sec_omp_steps}
-
-For this part of the work we began with the optimal {PS}y{KA}l version of
-the code, as obtained after applying the various transformations
-described in the previous section. As with the transformations of the
-serial code, our purpose here is to determine the functionality
-required of a tool that seeks to generate the PSy layer.
-
-\subsubsection{Separate PARALLEL DOs}
-\label{sec_sep_dos}
-
-The simplest possible OpenMP-parallel implementation consists of
-parallelising each loop nest in the PSy layer. This was done by
-inserting an OpenMP PARALLEL DO directive before each loop nest so
-that the iterations of the outermost or $j$ loop (over the latitude
-dimension of the model domain) are shared out amongst the OpenMP
-threads. This leaves the innermost ($i$) loop available for SIMD
-vectorisation by the compiler.
-
-The loop nest dealing with the application of the Flather boundary
-condition to the y-component of velocity ($v$) has a loop-carried
-dependency in $j$ which appears to prevent its being executed in
-parallel.\footnote{Only once this work was complete did we establish
-  that boundary conditions are enforced such that it can safely be
-  executed in parallel.}  This was therefore left unchanged and
-executed on thread 0 only.
-
-\subsubsection{Single PARALLEL region}
-\label{sec_single_parallel}
-
-Although very simple to implement, the use of separate PARALLEL DO
-directives results in a lot of thread synchronisation and can also
-cause the team of OpenMP threads to be repeatedly created and
-destroyed. This may be avoided by keeping the thread team in existence
-for as long as possible using an OpenMP PARALLEL region.  We therefore
-enclosed the whole of the PSy layer (in this code, a single
-subroutine) within a single PARALLEL region. The directive preceeding
-each loop nest to be parallelised was then changed to an OpenMP DO. We
-ensured that the $v$-Flather loop nest was executed in serial (by the
-first thread to encounter it) by enclosing it within an OpenMP SINGLE
-section.
-
-\subsubsection{First-touch policy}
-\label{sec_1sttouch}
-
-When executing an OpenMP-parallel program on a Non-Uniform Memory
-Access (NUMA) compute node it becomes important to ensure that the
-memory locations accessed by each thread are local to the hardware
-core upon which it is executing. One way of doing this is to implement
-a so-called 'first-touch policy' whereby memory addresses that will
-generally be accessed by a given thread during program execution are
-first initialised by that thread. This is simply achieved by using an
-OpenMP-parallel loop to initialise newly-allocated arrays to some
-value, \textit{e.g.} zero.
-
-Since data arrays are managed within the GOcean Library this
-optimisation can again be implemented without changing the
-natural-science code (\textit{i.e.} the Application and Kernel layers).
-
-\subsubsection{Minimise thread synchronisation}
-\label{sec_synch}
-
-By default, the OpenMP END DO directive includes an implicit barrier,
-thus causing all threads to wait until the slowest has completed the
-preceeding loop. Such synchronisation limits performance at larger
-thread counts and, for the NEMOLite2D code, is frequently unecessary.
-\textit{E.g.} if a kernel does not make use of the results of a
-preceeding kernel call then there is clearly no need for threads to
-wait between the two kernels.
-
-We analysed the inter-dependencies of each of the code sections within
-the PSy layer and removed all unecessary barriers by adding the NOWAIT
-qualifier to the relevant OpenMP END DO or END SINGLE directives. This
-reduced the number of barriers from eleven down to four.
-
-\subsubsection{Amortize serial region}
-\label{sec_amortize}
-
-As previously mentioned, the $v$-Flather section was executed in
-serial becuase of a loop-carried dependence in $j$. (In principle we
-could choose to parallelise the inner, $i$ loop but that would inhibit
-its SIMD vectorisation.) This introduces a load-imbalance between the
-threads. We attempt to mitigate this by moving this serial section to
-before the (parallel) $u$-Flather section. Since these two sections are
-independent, the aim is that the thread that performs the serial,
-$v$-Flather computation then performs a smaller share of the following
-$u$-Flather loop. In practice, this requires that some form of dynamic
-thread scheduling is used.
-
-\subsubsection{Thread scheduling}
-\label{sec_scheduling}
-
-In order to investigate how thread scheduling affects performance we
-used the `runtime' argument to the OpenMP SCHEDULE qualifier for all
-of our OpenMP parallel loops. The actual schedule to use can then be
-set at run-time using the OMP\_SCHEDULE environment variable. We
-experimented with using the standard static, dynamic and guided (with
-varying chunk size) OpenMP schedules.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-We first consider the performance of the code in serial and examine
-the effects of the transformations described in
-Section~\ref{sec_methodology}. Once we have arrived at an optimal form
-for the serial version of NEMOLite2D we then investigate its parallel
-performance on both CPU- and GPU-based systems.
-
-\subsection{Serial Performance}
-
-In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
-of the original version of the NEMOLite2D code for the range of
-compilers considered here. Unlike the Shallow
-code~\citep{shallow_psykal}, the original version of NEMOLite2D has not
-been optimised. Although it is still a single source file it is, in
-common with NEMO itself, logically structured with separate
-subroutines performing different parts of the physics within each time
-step. This structuring and the heavy use of conditional statements
-favour the Intel compiler which significantly out-performs both the
-Gnu and Cray compilers. Only when the problem size spills out of cache
-does the performance gap begin to narrow. The reason for the
-performance deficit of the Cray-compiled binary comes down to (a lack
-of) SIMD vectorisation, an issue that we explore below.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\picwidth]{orig_summary}
-\caption{Summary of the performance of the original version of the 
-NEMOLite2D code on an Intel Ivy Bridge CPU for
-the range of compilers under consideration.}
-\label{FIG_orig_perf_summary}
-\end{figure}
-
-Moving now the to the {PS}y{KA}l version of NEMOLite2D,
-Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest {PS}y{KA}l version for each of the compiler/problem-size combinations.
-While the Intel compiler still produces the best-performing binary,
-the Cray-compiled binary is now a very close second. In fact, the
-performance of both the Gnu- and Cray-compiled {PS}y{KA}l versions is
-generally significantly greater than that of their respective original
-versions. We also note that best absolute performance (in terms of
-grid points processed per second) with any compiler is obtained with
-the $256^2$ domain.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\picwidth]{best_psykal_summary}
-\caption{Summary of the best performance achieved by any {PS}y{KA}l 
-version of NEMOLite2D for each of the compilers under consideration.}
-\label{FIG_best_psykal_perf_summary}
-\end{figure}
-
-Figure~\ref{FIG_slowdown_summary} plots the percentage difference
-between the performance of the original and the best {PS}y{KA}l versions
-of NEMOLite2D for each compiler/problem-size combination. This shows that it is
-only the Intel-compiled binary running the $64^2$ domain that is
-slower with the {PS}y{KA}l version of the code (and then only by some
-3\%). For all other points in the space, the optimised {PS}y{KA}l version
-of the code performs better. The results for the Cray compiler are
-however somewhat sqewed by the fact that it did not SIMD vectorise key
-parts of the original version (see below).
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\picwidth]{slowdown_summary}
-\caption{Comparison of the performance of the best {PS}y{KA}l
-version with that of the original version of the code. A negative value 
-indicates that the {PS}y{KA}l version is slower than the original.}
-\label{FIG_slowdown_summary}
-\end{figure}
-
-Having show that we can recover, and often improve upon, the
-performance of the original version of NEMOLite2D, the next logical
-step is to examine the necessary code transformations in detail.  We
-do this for the $256^{2}$ case since this fits within cache on the Ivy
-Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
-detailed performance figures for this case after each transformation
-has been applied to the code. The same data is visualised in
-Figure~\ref{FIG_opt_stages_256}.
-
-Looking at the results for the Gnu compiler (and the Ivy Bridge CPU)
-first, all of the steps-up in performance correspond to kernel
-in-lining. None of the other transformations had any effect on the
-performance of the compiled code. In fact, simply in-lining the two
-kernels associated with the Momentum section was sufficient to exceed
-the performance of the original code.
-
-With the Intel compiler, the single largest performance increase is
-again due to kernel in-lining (of the Momentum kernels). This is
-because the compiler does a much better job of SIMD vectorising the
-loops involved than it does when it first has to in-line the kernel
-itself (as evidenced by its own diagnostic output - see
-Section~\ref{sec_inline_mom}). However, although this gives a
-significant performance increase it is not sufficient to match the
-performance of the original version. This is only achieved by
-in-lining every kernel and making the lack of data dependencies
-between arrays accessed from different field objects more explicit.
-
-The Cray compiler is distinct from the other two in that kernel
-in-lining does not give any performance benefit and in fact, for the
-smaller kernels, it can actually hurt performance. Thus the key
-transformation is to encourage the compiler to SIMD vectorise the
-Momentum section via a compiler-specific directive (without this it
-concludes that such vectorisation would be inefficient). Of the other
-transformations, only the addition of the compiler-specific
-safe\_address directive (see Section~\ref{sec_safe_address}) was found
-to give any benefit.
-
-\begin{table}[!t]
-  \caption{Performance (millions of points updated per second) on an Intel
-    Ivy Bridge CPU for the
-  $256^2$ case after each code transformation. Where an optimisation
-  utilises compiler-specific directives then performance figures for
-  the other compilers are omitted.}
-\label{TABLE_opt_breakdown}
-\centering
-\begin{tabular}{l|c|c|c}
-\hline
-Compiler:           & Gnu & Intel & Cray  \\
-\hline
-Original         & 3.87 & 12.2 & 6.83 \\
-Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
-Safe-address     & --   & --   & 6.95 \\
-In-line Momentum & 4.31 & 10.7 & 6.93 \\
-SIMD Momentum    & --   & --   & 11.8 \\
-\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
-In-line Continuity         & 4.83 & 11.8 & 11.6 \\
-In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
-In-line field copies       & 5.92 & 12.5 & 11.4 \\
-\hline
-\parbox{2.5cm}{\raggedright \%-speed-up of best \textit{c.f.} original} & 34.6 & 2.39 & 42.2 \\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\picwidth]{opt_stages_256}
-\caption{Performance of the {PS}y{KA}l version of NEMOLite2D for the
-  $256^{2}$ domain at each stage of optimisation. The first (black)
-  bar of each cluster gives the performance of the original version of
-  NEMOLite2D for that compiler/CPU combination.}
-\label{FIG_opt_stages_256}
-\end{figure}
-
-\subsection{Parallel Performance}
-
-We now turn to transformations related to parallelisation of the
-NEMOLite2D code; the introduction of OpenMP and OpenACC directives. In
-keeping with the {PS}y{KA}l approach we do not modify either the
-Algorithm- or Kernel-layer code.  Any code changes are restricted to
-either the PSy (middle) layer or the underlying library that manages
-\textit{e.g.} the construction of field objects.
-
-\subsubsection{OpenMP Version}
-
-As with the serial optimisations, we consider the effect of each of
-the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
-for the $256^2$ domain. For this we principally use a single Intel,
-Ivy Bridge socket which has 12 hardware cores and support for up to 24
-threads with hyperthreading (i.e. two threads per core).
-Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
-and~\ref{FIG_cray_omp_256} show the performance of each of the
-versions of the code on this system for the Gnu, Intel and Cray
-compilers, respectively.
-
-In order to quantify the scaling behaviour of the different versions
-of NEMOLite2D with the different compilers/run-time environments, we
-also plot the parallel efficiency in
-Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
-and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
-parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
-\[
-E(n) = 100  \frac{P(n)}{ n P(1)}
-\]
-where $P(m)$ is the performance of the code on $m$ threads. For a
-perfect, linearly-scaling code, $E(n)$ will be 100\%.
-
-Since the space to explore consists of three different compilers, six
-different domain sizes, five stages of optimisation and six different
-thread counts we can only consider a slice through it in what follows.
-In order to inform our choice of domain size,
-Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
-most performant, Intel-compiled version of NEMOLite2D. Although it is
-common for production runs of NEMO to use MPI sub-domains of $20
-\times 20$, it is clear from Figure~\ref{FIG_omp_domain_size} that the
-limited quantity of parallelism in the $32 \times 32$ domain inhibits
-scaling. Therefore, to fully test the performance of our OpenMP
-implementations we have chosen to examine results for the $256 \times
-256$ domain and these are shown in Figures~\ref{FIG_gnu_omp_256},
-\ref{FIG_intel_omp_256} and \ref{FIG_cray_omp_256}.
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_problem_size}
-\caption{The scaling behaviour of the most performant OpenMP-parallel
-  version of {PS}y{KA}l NEMOLite2D for the full range of domain sizes
-  considered on the CPU. Results are for the Intel compiler on a
-  single Intel Ivy Bridge socket. The corresponding parallel
-  efficiencies are shown using open symbols and dashed lines. The
-  24-thread runs employed hyperthreading.}
-\label{FIG_omp_domain_size}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
-\caption{Performance of the OpenMP-parallel version of
-  {PS}y{KA}l NEMOLite2D for the $256^2$ domain with the Gnu compiler on
-  a single Intel Ivy Bridge socket. The corresponding parallel
-  efficiencies are shown using open symbols and dashed lines. The
-  24-thread runs employed hyperthreading and the optimal OpenMP
-  schedule is given in parentheses.}
-\label{FIG_gnu_omp_256}
-\end{figure}
-
-The simplest OpenMP implementation (black lines, circle symbols) fails
-to scale well for any of the compilers. For the Intel and Gnu
-versions, parallel efficiency is already less than 50\% on just four
-threads. The Cray version however does better and is about 45\%
-efficient on eight threads (Figure~\ref{FIG_cray_omp_256}).
-
-With the move to a single PARALLEL region, the situation
-is greatly improved with all three executables now scaling out to at
-least 12 threads with $\sim70$\% efficiency (red lines, square symbols).
-
-In restricting ourselves to a single socket, we are keeping all
-threads within a single NUMA region. It is therefore surprising that
-implementing a `first-touch' policy has any effect and yet, for the
-Gnu- and Intel-compiled binaries, it appears to improve performance
-when hyperthreading is employed to run on 24 threads (green lines and
-diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
-~\ref{FIG_intel_omp_256}).
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_256_intel}
-\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-  NEMOLite2D for the Intel compiler on a single Intel Ivy Bridge
-  socket.  The corresponding parallel efficiencies are shown using
-  open symbols and dashed lines. The 24-thread runs employed
-  hyperthreading and the optimal OpenMP schedule is given in
-  parentheses.}
-\label{FIG_intel_omp_256}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{omp_scaling_256_cray}
-\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-  NEMOLite2D for the Cray compiler on a single Intel Ivy Bridge
-  socket.  The corresponding parallel efficiencies are shown using
-  open symbols and dashed lines. The 24-thread runs employed
-  hyperthreading and the optimal OpenMP schedule is given in
-  parentheses.}
-\label{FIG_cray_omp_256}
-\end{figure}
-
-The final optimisation step that we found to have any significant
-effect is to minimise the amount of thread synchronisation by
-introducing the NOWAIT qualifier wherever possible (blue lines and
-upwards-triangle symbols). For the Gnu compiler, this improves the
-performance of the executable on eight or more threads while, for the
-Intel compiler, it only gives an improvement for the 24-thread
-case. Moving the SINGLE region before a parallel loop is marginally
-beneficial for the Gnu- and Cray-compiled binaries and yet reduces the
-performance of the Intel binary (purple lines and right-pointing
-triangles).
-
-We have used different scales for the $y$-axes in each of the plots in
-Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
-and~\ref{FIG_cray_omp_256} in order to highlight the performance
-differences between the code versions with a given compiler. However,
-the best performance obtained on twelve threads (\textit{i.e.}  without
-hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
-Intel and Cray compilers, respectively. This serves to emphasise the
-democratisation that the introduction of OpenMP has had; for the
-serial case the Cray- and Intel-compiled exectables were a factor of
-two faster than the Gnu-compiled binary
-(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
-Gnu-compiled binary is only 6\% slower than that produced by the Intel
-compiler and is 13\% \textit{faster} that that of the Cray compiler. In
-part this situation comes about because of the effect that adding the
-OpenMP compiler flag has on the optimisations performed by the
-compiler.  In particular, the Cray compiler no longer vectorises the
-key Momentum section, despite the directive added during the serial
-optimisation work. This defficiency has been reported to Cray.
-
-Since NEMO and similar finite-difference codes tend to be
-memory-bandwidth bound, we checked the sensitivity of our performance
-results to this quantity by benchmarking using two sockets of Intel
-Ivy Bridge (\textit{i.e.} using a complete node of ARCHER, a Cray
-XC30). For this configuration we ensured that threads were evenly
-shared over the two sockets. The performance obtained for the $256^2$
-case with the `early SINGLE' version of the code is compared with the
-single-socket performance in
-Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the available
-memory bandwidth in this way has little effect on performance - the
-two-socket performance figures track those from a single socket very
-closely. The only significant difference in performance is at 24
-threads where, in addition to the difference in available memory
-bandwidth, the single-socket configuration is using hyperthreading
-while the two-socket case is not.  The discrepancy in the performance
-of the Cray- and Intel-compiled binaries at this thread count is under
-investigation by Cray.
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{1and2sockets}
-\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-  NEMOLite2D on one and two sockets of Intel Ivy Bridge. The 24-thread
-  runs on a single socket used hyperthreading and the two-socket runs
-  had the threads shared equally between the sockets.}
-\label{FIG_omp_2socks}
-\end{figure}
-
-A further complication is the choice of scheduling of the OpenMP
-threads.  We have investigated the performance of each of the
-executables (and thus the associated OpenMP run-time library) with the
-standard OpenMP \textit{static}, \textit{dynamic} and \textit{guided}
-scheduling policies~\citep{openmp_standard}. For the Intel
-compiler/run-time, static loop scheduling was found to be best for all
-versions apart from that where we have attempted to amortize the cost
-of the SINGLE section. This is to be expected since that strategy
-requires some form of dynamic loop scheduling in order to reduce the
-load imbalance introduced by the SINGLE section.
-
-In contrast, some form of dynamic scheduling gave a performance
-improvement with the Gnu compiler/run-time even for the `first-touch'
-version of the code. This is despite the fact that this version
-contains (implicit) thread synchronisation after every parallel loop.
-For the Cray compiler/run-time, some form of dynamic scheduling became
-optimal once inter-thread synchronisation was reduced using the NOWAIT
-qualifiers.
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
-%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-%  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
-%  socket. The corresponding parallel efficiencies are shown using open
-%  symbols and dashed lines. The 24-thread runs employed hyperthreading
-%  and the optimal OpenMP schedule is given in parentheses.}
-%\label{FIG_gnu_omp_32}
-%\end{figure}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\picwidth]{omp_scaling_32_intel}
-%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-%  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
-%  socket.  The corresponding parallel efficiencies are shown using
-%  open symbols and dashed lines. The 24-thread runs employed
-%  hyperthreading and the optimal OpenMP schedule is given in
-%  parentheses.}
-%\label{FIG_intel_omp_32}
-%\end{figure}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\picwidth]{omp_scaling_32_cray}
-%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
-%  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
-%  socket.  The corresponding parallel efficiencies are shown using
-%  open symbols and dashed lines. The 24-thread runs employed
-%  hyperthreading and the optimal OpenMP schedule is given in
-%  parentheses.}
-%\label{FIG_cray_omp_32}
-%\end{figure}
-
-\subsubsection{OpenACC Version}
-
-The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
-wish to run NEMOLite2D on different hardware, \textit{e.g.} a
-GPU. This is because the necessary code modifications are, by design,
-limited to the middle PSy layer. In order to demonstrate this and to
-check for any limitations imposed by the {PS}y{KA}l re-structuring, we
-had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
-used OpenACC directives for this porting as that approach is similar
-to the use of OpenMP directives and works well within the {PS}y{KA}l
-apprach. He also experimented with using CUDA directly within the
-original form of NEMOLite2D in order to quantify any performance
-penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
-
-The performance of these two versions of the code is plotted in
-Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring of
-NEMOLite2D did not prove to be a barrier to achieving a performant
-code when using OpenACC directives, we only report the performance of
-the {PS}y{KA}l OpenACC version. For problem sizes up to $256^2$ the
-difference in the performance of the two versions is less than 5\% and
-for the larger domains it is at most 11\%. This demonstrates that the
-performance cost of the {PS}y{KA}l approach in utilising a GPU for
-NEMOLite2D is minimal.
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{gpu_opt_stages}
-\caption{Performance of the {PS}y{KA}l (OpenACC) and CUDA GPU
-  implementations of NEMOLite2D. The performance for the OpenACC
-  version of the original code is not shown since it is identical to
-  that of the {PS}y{KA}l version.}
-\label{FIG_gpu_opt_stages}
-\end{figure}
-
-In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
-the OpenMP and OpenACC implementations of NEMOLite2D across the range
-of problem sizes considered. The OpenMP figures are the maximum
-performance obtained from a whole Intel Ivy Bridge socket by any
-version of the code on any number of threads for a given
-compiler/run-time. For the smallest domain size ($64^2$) the OpenMP
-version significantly outperforms the GPU because there is
-insufficient parallelism to fully utilise the GPU and one time-step
-takes only 80~$\mu$s. The execution of a single time-step is then
-dominated by the time taken to launch a kernel on the GPU rather than
-the execution of the kernel itself.
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{cpu_cf_gpu}
-\caption{Performance of the best OpenMP-parallel version of {PS}y{KA}l\
-  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the {PS}y{KA}l GPU implementation (using OpenACC).}
-\label{FIG_cpu_cf_gpu}
-\end{figure}
-
-Once the problem size is increased to $128^2$, a single time-step
-takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
-comparable in performance to the OpenACC version. For all of the
-larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
-version is considerably faster than the CPU.  For problem sizes of
-$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
-exhausted and performance becomes limited by the bandwidth to main
-memory. At this stage the OpenACC version of the code on the GPU is
-some 3.4 times faster than the best OpenMP version on the CPU.
-
-%Peak memory bandwidth on the K40 GPU is 288 GB/s and we measured
-%(using STREAM2) bandwidths of 46.0 and 20.5 GB/s to L3 and main
-%memory, respectively on the CPU.
-
-\subsection{Performance Analysis with the Roofline Model}
-
-Although we have investigated how the performance of the {PS}y{KA}l
-version of NEMOLite2D compares with that or the original, we have not
-addressed how efficient the original actually is. In order to do so we
-use the Roofline Model~\citep{roofline} which provides a relatively
-simple way of characterising the performance of a code in terms of
-whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\citep{para_pearls} and use the
-STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
-get the upper bounds on the memory bandwidth and floating-point
-operations per second (FLOPS) for the test hardware. Since we are
-using an Intel Ivybridge CPU we used the Intel Math Kernel Library
-implementation of LINPACK.
-
-A key component of the Roofline model is the Arithmetic Intensity of
-the code being executed:
-\begin{equation}
-AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
-\end{equation}
-We calculated this quantity manually by examining the
-source code and counting the number of memory references and
-arithmetic operations that it contained. In doing this counting we
-assume that any references to adjacent array elements are fetched in a
-single cache-line and thus only count once.
-
-\begin{figure}
-\centering
-\includegraphics[width=\picwidth]{roofline}
-\caption{Comparison of the performance achieved by kernels from
-  NEMOLite2D and Shallow on a roof-line plot for the E5-1620
-  CPU. Results for the former are for the $256^2$ domain since that
-  gave the best performance. See the text for a discussion of the
-  different CPU ceilings (dashed lines).}
-\label{FIG_roofline}
-\end{figure}
-
-In Figure~\ref{FIG_roofline} we show the performance of kernels from
-both Shallow and NEMOLite2D on the roofline model for an Intel E5-1620 CPU.
-This demonstrates that the Shallow kernel is achieving a performance
-roughly consistent with saturating the available memory bandwidth. In
-contrast, the kernel taken from NEMOLite2D is struggling to reach a
-performance consistent with saturating the bandwidth to main
-memory. We experimented with reducing the problem size (so as to
-ensure it fitted within cache) but that did not significantly improve
-the performance of the kernel. This then points to more fundamental issues
-with the way that the kernel is implemented which are not captured in the
-simple roofline model.
-
-In order to aid our understanding of kernel performance we have
-developed a tool capable of parsing Fortran code and generating a
-Directed Acyclic Graph (DAG) of the data flow. This tool eases the
-laborious and error-prone process of counting memory accesses and
-FLOPs as well as providing information on those operations that are
-rate-limiting or on the critical path. Using the details of the Intel
-Ivy Bridge microarchitecture published by Fog~\citep{fog_microarch,
-  fog_instructions} we have constructed performance estimates of the
-NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
-(ILP), \textit{i.e.} assuming that all nodes in the DAG are executed
-in serial, we get a lower-bound performance estimate of $0.6391 \times
-CLOCK\_SPEED$ FLOPS which gives 2.46~GFLOPS at a clock speed of
-3.85~GHz.
-
-Alternatively, we may construct an upper bound by assuming the
-out-of-order execution engine of the Ivy Bridge core is able to
-perfectly schedule and pipeline all operations such that those that go
-to different execution ports are always run in parallel. In the Ivy
-Bridge core, floating-point multiplication and division operations go
-to port 0 while addition and subtraction go to port
-1~\citep{fog_instructions}. Therefore we sum the cost of all
-multiplication and division operations in the DAG and compare that
-with the sum of all addition and subtraction operations. The greater
-of these two quantities is then taken to be the cost of executing the
-kernel; all of the operations on the other port are assumed to be done
-in parallel. This gives a performance estimate of $1.029 \times
-CLOCK\_SPEED$ FLOPS or 3.96 GFLOPS at 3.85~GHz.
-
-These performance estimates are plotted as CPU ceilings (horizontal
-dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
-Momentum kernel is seen to fall between these two bounds which
-demonstrates that its performance is not memory-bandwidth limited, as
-might have been assumed by its low AI. Therefore, although the
-performance of this kernel is clearly sub-optimal, this is due to the
-balance of floating-point operations that it contains. Improving this
-is outside the scope of the current work since here we are focused on
-the {PS}y{KA}l separation of concerns and the introduction of parallelism
-in the PSy layer.
-
-Enabling SIMD vectorisation for this kernel does not significantly
-improve its performance (Figure~\ref{FIG_roofline}) and in fact,
-limiting it to SSE instructions (vector width of two double-precision
-floating point numbers) rather than AVX (vector width of four) was
-found to produce a slightly more performant version. The poor
-efficiency of the SIMD version is highlighted by the fact that the
-corresponding kernel performance no longer falls within the bounds of
-our performance estimate. This is because we have incorporated the
-effect of SSE into that estimate by simply assuming perfect
-vectorisation which gives a performance increase of a factor of two.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\conclusions
-
-We have investigated the application of the {PS}y{KA}l separation of
-concerns approach to the domain of shared-memory parallel,
-finite-difference shallow-water models. This approach enables the
-computational science (performance) related aspects of a computer
-model to be kept separate from the natural (oceanographic) science
-aspects.
-
-We have used a new and un-optimised, two-dimensional model extracted
-from the NEMO ocean model for this work. As a consequence of this, the
-introduction of the {PS}y{KA}l separation of concerns followed by
-suitable transformations of the PSy Layer is actually found to improve
-performance. This is in contrast to our previous
-experience~\citep{shallow_psykal} with tackling the Shallow code which
-has been optimised over many years. In that case we were able to
-recover (to within a few percent) the performance of the original and
-in some cases exceed it, in spite of limiting ourselves to
-transformations which replicated the structure of the original,
-optimised code.
-
-Investigation of the absolute serial performance of the NEMOLite2D
-code using the roofline model revealed that it was still significantly
-below any of the traditional roofline ceilings. We have developed
-Habakkuk (https://github.com/arporter/habakkuk), a code-analysis tool
-that is capable of providing more realistic ceilings by analysing the
-nature of the floating point computations performed by a kernel. The
-bounds produced by Habakkuk are in good agreement with the measured
-performance of the principal (Momentum) kernel in NEMOLite2D. In
-future work we aim to extend this tool to account for SIMD operations
-and make it applicable to code parallelised using OpenMP.
-
-The application of code transformations to the middle/PSy layer is key
-to the performance of the {PS}y{KA}l version of a code. For both
-NEMOLite2D and Shallow we have found that for serial performance, the
-most important transformation is that of in-lining the kernel source
-at the call site, i.e. within the PSy layer. (Although we have done
-this in-lining manually for this work, our aim is that in future, such
-transformations will be performed automatically at compile-time and
-therefore do not affect the code that a scientist writes.)
-For the more complex NEMOLite2D code the Cray compiler also
-had to be coerced into performing SIMD vectorisation through the use
-of souce-code directives.
-
-In this work we have also demonstrated the introduction of parallelism
-into the PSy layer with both OpenMP and OpenACC directives. In both
-cases we were able to leave the natural-science parts of the code
-unchanged. For OpenMP we achieved a parallel efficiency of $\sim70\%$
-on 12 threads by enclosing the body of the (single) PSy layer routine
-within a single PARALLEL region. Removal of unecessary synchronisation
-points through use of the NOWAIT clause boosted 12-thread performance
-by approximately 10\% with the Gnu and Cray compilers. The
-{PS}y{KA}l re-structuring of this (admittedly small) code was not found
-to pose any problems for the introduction of performant OpenMP.
-Similarly, we have also demonstrated good GPU performance using OpenACC
-in a {PS}y{KA}l version of the code.
-
-This paper demonstrates that the {PS}y{KA}l separation of concerns may
-be applied to 2D finite-difference codes without loss of
-performance. We have also shown that the resulting code is amenable to
-efficient parallelisation on both GPU and shared-memory CPU systems.
-This then means that it is possible to achieve performance portability
-while maintaining single-source science code.
-
-Our next steps will be first, to consider the automatic generation of the
-PSy layer and second, to look at extending the approach to the full NEMO
-model (i.e. three dimensions). In future work we will analyse the
-performance of a domain-specific compiler that performs the automatic
-generation of the PSy layer. This compiler (which we have named
-`PSyclone', see https://github.com/stfc/psyclone) is currently under
-development.
-
-\codeavailability{The NEMOLite2D package is available as the GOcean
- 2.0 benchmark, available from https://puma.nerc.ac.uk/trac/GOcean.}
-
-\competinginterests{No competing interests are present.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{acknowledgements}
-
-This work made use of the ARCHER UK National Supercomputing Service
-(\url{http://www.archer.ac.uk}) and Emerald, a GPU-accelerated High
-Performance Computer, made available by the Science \& Engineering
-South Consortium operated in partnership with the STFC
-Rutherford-Appleton Laboratory
-(\url{http://www.ses.ac.uk/high-performance-computing/emerald/}).
-
-\end{acknowledgements}
-
-\bibliographystyle{copernicus}
-\bibliography{nemolite2d_perf}
-
 \end{document}

### 2017-07-10 nemolite2d_perf.tex
nemolite2d-paper: update to copernicus 4.6 package and add line-numbering

@@
 %% 2-column papers and discussion papers
-\documentclass[gmd, manuscript]{copernicus}
+\documentclass[gmdd, manuscript]{copernicus}
+
+\usepackage{lineno}
+\linenumbers
 
@@
 version of NEMOLite2D compares with that of the original, we have not
-addressed how efficient the original actually is. In order to do so we
-use the Roofline Model~\citep{roofline} which provides a relatively
-simple way of characterising the performance of a code in terms of
-whether it is memory-bandwidth bound or compute bound. To do so we
-follow the approach suggested in~\cite{para_pearls} and use the
-STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
-get the upper bounds on the memory bandwidth and floating-point
-operations per second (FLOPS) for the test hardware. Since we are
-using an Intel Ivybridge CPU we used the Intel Math Kernel Library
+addressed how efficient the original actually is. Without this
+information we have no way of knowing whether further optimisations
+might yield worthwhile performance improvements. Therefore, we
+consider the performance of the original, serial NEMOLIte2D code on
+the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
+which provides a relatively simple way of characterising the
+performance of a code in terms of whether it is memory-bandwidth bound
+or compute bound. To do so we follow the approach suggested
+in~\cite{para_pearls} and construct a roofline model using the
+STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order
+to obtain appropriate upper bounds on the memory bandwidth and
+floating-point operations per second (FLOPS), respectively. Since we
+are using an Intel Ivy Bridge CPU we used the Intel Math Kernel Library
 implementation of LINPACK.
@@
 A key component of the Roofline model is the Operational or Arithmetic
-Intensity of the code being executed:
+Intensity (AI) of the code being executed:
 \[
@@
 corresponding kernel performance no longer falls within the bounds of
-our performance estimate. This is because we have incorporated the
-effect of SSE into that estimate by simply assuming perfect
-vectorisation which gives a performance increase of a factor of two.
+the performance estimates produced by Habakkuk. This is because we
+have incorporated the effect of SSE into that estimate by simply
+assuming perfect vectorisation which gives a performance increase of a
+factor of two.  Further investigation of this issue revealed that
+several of the NEMOLite2D kernels make frequent use of floating-point
+division operations. Although SSE and AVX versions of the division
+operation are available, on Ivy Bridge they do not provide any
+performance benefit~\citep{fog_instructions}. A straightforward
+optimisation then would be to alter the kernels in order to reduce the
+number of division operations. However, again, kernel optimisation is
+outside the scope of this paper since it breaks the PSyKAl separation
+of concerns.
 

### 2017-09-06 docs/papers/roofline_habakkuk/habakkuk.tex
habakkuk-paper: add skeleton to conclusions, toc

@@
 
+\tableofcontents
+
 \section{Introduction}
 
-Shallow implementation of Sadourny~\cite{sadourny75}.
-NEMOLite2D.
+\begin{itemize}
+\item Shallow implementation of Sadourny~\cite{sadourny75}.
+\item NEMOLite2D.
+\item NEMO
+\end{itemize}
 
@@
 however be shared by the executing threads.
+
+\section{Building a Model of a CPU Core}
+
+We require various parameters that characterise the performance of a
+CPU core, the value of these are micro-architecture dependent.
+
+\begin{itemize}
+\item Vector width
+\item Size of a cache line
+\item Support for FMA
+\item Support for out-of-order execution
+\item Available Instruction-Level Parallelism
+\item Cost of arithmetic operations
+\end{itemize}
+
+\subsection{Micro-Benchmarks}
+
+\begin{itemize}
+\item Fortran for high-level, intrinsic ops (e.g. SIN())
+\item Likwid for low-level, individual arithmetic ops (e.g. floating-point division)
+\item Results from Agner Fog
+\end{itemize}
 
@@
 
+\begin{itemize}
+
+\item Constructing a traditional roofline model is not that straightforward.
+
+\item We have demonstrated that the upper bounds in such a model can be
+  unrealistically high and therefore not that useful in determining
+  the performance-limiting aspects of a kernel.
+
+\item We have developed a tool, Habakkuk, which parses Fortran source
+  code and generates various performance estimates using a simple
+  model of the microarchitecture of the target CPU.
+
+\item Using the resulting performance estimates as the ceilings in the
+  roofline model results in more accurate bounds and provides more
+  information to a developer seeking to optimise their source code.
+
+\end{itemize}
+
 \bibliographystyle{unsrt}

### 2017-09-13 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: WIP on new, Related Approaches section

@@
 a new Dynamical core for the UK Met Office, and its design has been
-influenced by earlier work on OP2~\citep{OP2, PYOP2}.
-
-In common with OP2, the {PS}y{KA}l approach separates out the science
-code and the performance-related code into distinct layers. The calls
-that specify parallelism in both approaches are similar in terms of
-where they are placed in the code and in their semantics. However, the
-{PS}y{KA}l approach supports the specification of more than one kernel
-in a parallel region of code, compared with one for OP2, giving more
-scope for optimisation. In addition, the metadata describing a kernel
-is included with the kernel code in the {PS}y{KA}l approach whereas it
-is provided as a part of the kernel call in OP2.
+influenced by earlier work on OP2~\citep{OP2, PYOP2} - see Section
+\ref{sec_other_approaches} for more details.
 
@@
 HPC expert and/or some coding rules to help ensure that the kernels
-compile into efficient code.  In an alternative
-approach~\citep{firedrake,fenics}, kernels are generated from a
-high-level specification, potentially allowing them to be optimised
-automatically~\citep{coffee}.
+compile into efficient code.
 
@@
 discussed in the remainder of this paper.
+
+\subsection{Related Approaches}
+\label{sec_other_approaches}
+
+As already mentioned, our approach is heavily influenced by the OP2
+system~\citep{OP2, PYOP2}.
+In common with OP2, the {PS}y{KA}l approach separates out the science
+code and the performance-related code into distinct layers. The calls
+that specify parallelism in both approaches are similar in terms of
+where they are placed in the code and in their semantics. However, the
+{PS}y{KA}l approach supports the specification of more than one kernel
+in a parallel region of code, compared with one for OP2, giving more
+scope for optimisation. In addition, the metadata describing a kernel
+is included with the kernel code in the {PS}y{KA}l approach whereas it
+is provided as a part of the kernel call in OP2.
+
+Although {PS}y{KA}l is a design for architecting code, rather than an
+implementation of a tool, there is an implicit assumption that the
+majority of the kernels in an application will be provided by the
+application developer. In the GridTools~\citep{grid_tools} and
+Firedrake~\citep{firedrake,fenics} approaches, the mathematical
+operations (finite-difference stencils and finite element operations,
+respectively) are specified in a high-level language by the
+user. Kernel code is then generated automatically. It is possible to
+optimise this generated code~\citep{coffee} and, in the case of
+GridTools, support both CPU and GPU architectures.
+
+% Put this later after we've discussed porting nemolite2d to GPU?
+%It is likely that the {PS}y{KA}l approach of constraining all
+%performance-related code to be within the middle, {PS}y, layer
+
+Lower-level abstractions, e.g. OCCA.
+
+H.
+C.
+Edwards, C.
+R.
+Trott, D.
+Sunderland, "Kokkos: Enabling manycore performance portability through polymorphic memory access patterns", J.
+Parallel Distrib.
+Comput., vol.
+74, no.
+12, pp.
+3202-3216, 2014.
 

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: complete first draught of new similar-approaches section

@@
 
+This paper is concerned with the implications of {PS}y{KA}l as a
+design for architecting code. The implementation of an associated tool
+for generating the middle, {PS}y, layer will be the subject of a future
+paper. However, comparison with other approaches necessarily involves
+tools rather than simply architectures.
+
 As already mentioned, our approach is heavily influenced by the OP2
@@
 
-Although {PS}y{KA}l is a design for architecting code, rather than an
-implementation of a tool, there is an implicit assumption that the
+In the {PS}y{KA}l approach there is an implicit assumption that the
 majority of the kernels in an application will be provided by the
@@
 optimise this generated code~\citep{coffee} and, in the case of
-GridTools, support both CPU and GPU architectures.
+GridTools, support both CPU and GPU architectures. This ability to
+generate kernels specific to a given computer architecture is a
+powerful feature. However, with this power comes the responsibility of
+providing domain scientists with the necessary functionality to
+describe all conceivable kernel operations as well as a programming
+interface with which they are comfortable.
+
+The {PS}y{KA}l, GridTools and Firedrake approaches are all based on
+the concept of (various flavours of) a Domain-Specific Language (DSL)
+for finite-difference and finite-element applications. This is
+distinct from other, lower-level abstractions such as
+Kokkos~\citep{kokkos} and OCCA~\citep{occa} where the aim is to
+provide a language that permits a user to implement a kernel just once
+and have it compile to performant code on a range of multi- and
+many-core devices. One could imagine using such abstractions to
+implement kernels for the {PS}y{KA}l, GridTools and Firedrake
+approaches rather than using e.g. OpenMP or CUDA directly.
 
@@
 
-Lower-level abstractions, e.g. OCCA.
-
-H.
-C.
-Edwards, C.
-R.
-Trott, D.
-Sunderland, "Kokkos: Enabling manycore performance portability through polymorphic memory access patterns", J.
-Parallel Distrib.
-Comput., vol.
-74, no.
-12, pp.
-3202-3216, 2014.
+\begin{table}
+\caption{An overview of the functionality of similar approaches. Static compilation here means that all code is compiled before program execution is begun.}
+\label{TABLE_dsl_compare}
+\centering
+\begin{tabular}{l|c|c|c|c|c|c|c}
+\hline
+Approach   & DSL & MPI & Threading & Data layout & Kernels & Language & Compilation  \\
+\hline
+{PS}y{KA}l & Yes & Yes & Yes & Fixed  & User-supplied & Fortran & Static \\
+GridTools  & Yes & No  & Yes & Flexible & Generated & C++  & Static \\
+Firedrake  & Yes & Yes & No  & Fixed  & Generated & C with Python interface & Run-time \\
+Kokkos     & No  & No  & Yes & Flexible & User-supplied & C++ & Static \\
+OCCA       & No  & No  & Yes & Fixed  & User-supplied & C (Python \& Fortran interfaces) & Run-time \\
+\hline
+\end{tabular}
+\end{table}
 

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: update title and code DOI

@@
 
-\title{Portable Multi- and Many-Core Performance for Finite Difference Codes;
-  Application to the Free-Surface Component of NEMO.}
+\title{Portable Multi- and Many-Core Performance for Finite-Difference/Element Codes;
+  Application to the Free-Surface Component of NEMO (NEMOLite2D 1.0).}
 
@@
 
-\codeavailability{The NEMOLite2D package is available as the GOcean
-  2.0 benchmark, available from
-  https://puma.nerc.ac.uk/trac/GOcean. Habakkuk is available from
+\codeavailability{The NEMOLite2D Benchmark Suite 1.0 is available from
+  the eData repository with DOI:
+  http://dx.doi.org/10.5286/edata/707. Habakkuk is available from
   https://github.com/arporter/habakkuk.}

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-paper: mention PSyclone by name in related-approaches section

@@
 design for architecting code. The implementation of an associated tool
-for generating the middle, {PS}y, layer will be the subject of a future
-paper. However, comparison with other approaches necessarily involves
-tools rather than simply architectures.
+(which we have named 'PSyclone') for generating the middle, {PS}y,
+layer will be the subject of a future paper. However, comparison with
+other approaches necessarily involves discussing other tools rather
+than simply architectures.
 

### 2017-09-14 docs/papers/nemolite2d_performance/referee_response.txt
nemolite-paper: add response to first referee

The authors thank the reviewers for their detailed feedback on our paper.
We address the points raised below, beginning with the first (anonymous)
referee and then moving on to those from Carlos Osuna.

Anonymous referee:

1/ The reviewer is right to be concerned with seeing how much
   performance may be lost by the re-structuring required for
   PSyKAl. However, the original form of the NEMOLite2D code that we
   started with for this work was unoptimised. This is emphasised by
   the results in Figure 4 which show that, for the majority of cases,
   the final PSyKAl version was significantly faster. In contrast, our
   previous paper (Porter et al, 2016) tackled the "Shallow" code
   which has had optimisations applied to it over a period of some
   twenty years and was therefore a much more demanding test. There,
   the initial PSyKAl re-structuring did significantly harm
   performance. However, we showed that it was possible to recover the
   performance of the optimised original by applying code
   transformations that were PSyKAl-compliant. We therefore do not
   think that showing results for the optimisations
   applied to the original form of NEMOLite2D is necessary.

2/ The PSyKAl approach is in fact applicable to both Finite Element and
   Finite Difference codes (and in fact, LFRic, the largest project making use
   of it is using Finite Elements). We have updated the title to make
   this clear.

3/ A new section, "Related Approaches", has been added to the paper. In this
   we bring together a comparison of various other approaches including OCCA.
   This makes explicit the differences and makes clearer the advantages of
   PSyKAl.

4/ We also use the new "Related Approaches" section to emphasise that
    although the tool to generate the PSy layer exists ("PSyclone"),
    the current paper is about code structuring.

### 2017-09-14 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: add mom kernel fragment

@@
     the current paper is about code structuring.
+
+Carlos Osuna's general review:
+
+We have added a few key lines of source code from the Momentum kernel in
+section 3.3. We have also added text that explains that the apparently poor
+performance of this kernel is down to the number of division operations.
+
+The suggestion of a "Related Work" section is a good one and we have added
+this just after the Introduction.

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: add mom kernel fragment

@@
 might yield worthwhile performance improvements. Therefore, we
-consider the performance of the original, serial NEMOLIte2D code on
+consider the performance of the original, serial NEMOLite2D code on
 the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
@@
 CPU, this is due to the balance of floating-point operations that it
-contains. Improving this is outside the scope of the current work
-since here we are focused on the {PS}y{KA}l separation of concerns and
-the introduction of parallelism in the PSy layer.
+contains and in particular, the number of division
+operations. (Division costs at least eight times as much as a
+multiplication in the Ivy Bridge core~\citep{fog_instructions}.) For
+instance, a fragment of the most costly part of the momentum kernel is
+shown below:
+\begin{verbatim}
+    ...
+    ! -pressure gradient
+    hpg = -g * (hu(ji,jj) + sshn_u(ji,jj)) * e2u(ji,jj) * &
+           (sshn(ji+1,jj) - sshn(ji,jj))
+
+    ! -linear bottom friction (implemented implicitly.)
+    ua(ji,jj) = (un(ji,jj) * (hu(ji,jj) + sshn_u(ji,jj)) + rdt * &
+                 (adv + vis + cor + hpg) / e12u(ji,jj)) / &
+                (hu(ji,jj) + ssha_u(ji,jj)) / (1.0_wp + cbfr * rdt) 
+\end{verbatim}
+The liberal use of the division operation is clearly something that
+can be improved upon. However, this is outside the scope of the
+current work since here we are focused on the {PS}y{KA}l separation of
+concerns and the introduction of parallelism in the PSy layer.
 

### 2017-09-14 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: explain why SSE > AVX

@@
 
-We have added a few key lines of source code from the Momentum kernel in
-section 3.3. We have also added text that explains that the apparently poor
-performance of this kernel is down to the number of division operations.
+We have added a few key lines of source code from the Momentum kernel
+in section 3.3. We have also added text that explains that the
+apparently poor performance of this kernel is down to the number of
+division operations.  On the Ivy Bridge, division operations require
+at least eight times as many cycles as a multiplication which can be
+very significant.
 
@@
 this just after the Introduction.
+
+We attribute the better performance of the SSE version of the code to the
+fact that the poor SIMD efficiency of the code is failing to amortize the
+larger cost of the peel and remainder loops associated with the greater
+vector width of AVX. We have expanded upon this in the text.

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: explain why SSE > AVX

@@
 floating point numbers) rather than AVX (vector width of four) was
-found to produce a slightly more performant version. The poor
-efficiency of the SIMD version is highlighted by the fact that the
-corresponding kernel performance no longer falls within the bounds of
-the performance estimates produced by Habakkuk. This is because we
-have incorporated the effect of SSE into that estimate by simply
-assuming perfect vectorisation which gives a performance increase of a
-factor of two.  Further investigation of this issue revealed that
-several of the NEMOLite2D kernels make frequent use of floating-point
-division operations. Although SSE and AVX versions of the division
-operation are available, on Ivy Bridge they do not provide any
-performance benefit~\citep{fog_instructions}. A straightforward
-optimisation then would be to alter the kernels in order to reduce the
-number of division operations. However, again, kernel optimisation is
-outside the scope of this paper since it breaks the PSyKAl separation
-of concerns.
+found to produce a slightly more performant version. We attribute this
+performance deficit to the low efficiency of the vectorised code
+combined with the higher trip-counts of the peel- and remainder loops
+required for the greater vector width of AVX.  The poor efficiency of
+the SIMD version is highlighted by the fact that the corresponding
+kernel performance no longer falls within the bounds of the
+performance estimates produced by Habakkuk. This is because we have
+incorporated the effect of SSE into that estimate by simply assuming
+perfect vectorisation which gives a performance increase of a factor
+of two.  Further investigation of this issue revealed that, as
+mentioned above, several of the NEMOLite2D kernels make frequent use
+of floating-point division. Although SSE and AVX versions
+of the division operation are available, on Ivy Bridge they do not
+provide any performance benefit~\citep{fog_instructions}. A
+straightforward optimisation then would be to alter the kernels in
+order to reduce the number of division operations. However, again,
+kernel optimisation is outside the scope of this paper since it breaks
+the PSyKAl separation of concerns.
 

### 2017-09-14 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: clarify safe_address directive and change to ...Intel can...

@@
 vector width of AVX. We have expanded upon this in the text.
+
+Carlos Osuna's specific comments:
+
+(Top of page 8.) We have clarified why allocating larger arrays was
+necessary in order to use the safe_address directive.
+
+Changed "Intel does..." to "Intel can..." as requested.
+
+Sec 2.1.5 - the grid is not attached to the field, rather each field
+keeps a pointer back to the grid upon which it is defined. This approach
+was chosen in order to support applications which use more than one grid.
+
+Sec 2.1.8 -

### 2017-09-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: clarify safe_address directive and change to ...Intel can...

@@
 
-Much of the optimisations we have performed have been informed by the
+Many of the optimisations we have performed have been informed by the
 diagnostic output produced by either the Cray or Intel compilers. Many
@@
 \end{verbatim}
-In order that we could safely add such directives we altered the
-GOcean infrastructure to allocate all field-data arrays with extents
-greater than strictly required.  We were then able to safely add the
-\texttt{safe\_address} before all of the loops where the Cray compiler
-indicated it might be useful (the Momentum loops and some of the BC
-loops).
+Originally, all fields were only allocated the bare minimum of storage
+and the conditional statements within kernels prevented out-of-bounds
+accesses, \textit{e.g.} at the edge of the simulation domain.  We
+subsequently altered the GOcean infrastructure to allocate all
+field-data arrays with extents greater than strictly required.  This
+enabled us to safely add the \texttt{safe\_address} before all of
+the loops where the Cray compiler indicated it might be useful (the
+Momentum loops and some of the BC loops).
 
@@
 
-Although the Intel compiler does do in-lining when routines are in
+Although the Intel compiler can perform in-lining when routines are in
 separate source files, we have found (both here and in our previous

### 2017-09-15 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: respond to more of Carlos comments

@@
 
-Sec 2.1.8 -
+Sec 2.1.8 - it is not clear why in-lining manually helps (especially since
+one would expect the Intel compiler in particular to do this). We suspect
+that the change in the PSy-layer code causes the compiler to make some
+different optimisation decisions and these happen to be beneficial. A
+proper investigation of this is outside the scope of this paper. We have
+added a sentence about this.
+
+Sec 2.2.1 - yes, other strategies might be beneficial, especially blocking.
+The key point is that adopting the PSyKAl structure does not limit these
+possibilities.
+
+Figure 3 - The Gnu results are much slower because gfortran is only SIMD
+vectorising the loop containing the Continuity kernel. For some reason
+it does not vectorise any others. Unlike Cray, there is no directive
+that can be inserted to mandate that a loop be SIMD vectorised. We have
+added text to this effect.

### 2017-09-15 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: respond to more of Carlos comments

@@
 the current time-step become the fields at the previous time-step.
-Initially we implemented these copies as 'built-in' kernels (in the
-GOcean infrastructure) as they are specified in the algorithm
-layer. However, we obtained better performance by simply in-lining
-these array copies into the PSy layer.
+Initially we implemented these copies as `built-in' kernels (in the
+GOcean infrastructure) as they are specified in the Algorithm
+layer. However, we obtained better performance (for the Gnu and Intel
+compilers) by simply manually in-lining these array copies into the
+PSy layer. As discussed in~\ref{sec_inline_mom}, it is unclear why
+in-lining should improve performance, particularly for the Intel
+compiler. However, a detailed investigation of this issue is outside
+the scope of this paper.
 
@@
 Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
-fastest {PS}y{KA}l version for each of the compiler/problem-size combinations.
-While the Intel compiler still produces the best-performing binary,
-the Cray-compiled binary is now a very close second. In fact, the
-performance of both the Gnu- and Cray-compiled {PS}y{KA}l versions is
-generally significantly greater than that of their respective original
-versions. We also note that best absolute performance (in terms of
-grid points processed per second) with any compiler is obtained with
-the $256^2$ domain.
+fastest {PS}y{KA}l version for each of the compiler/problem-size
+combinations.  While the Intel compiler still produces the
+best-performing binary, the Cray-compiled binary is now a very close
+second. In fact, the performance of both the Gnu- and Cray-compiled
+{PS}y{KA}l versions is generally significantly greater than that of
+their respective original versions. We also note that best absolute
+performance (in terms of grid points processed per second) with any
+compiler is obtained with the $256^2$ domain. The performance of the
+Gnu-compiled binary is consistently a factor of two slower than those
+of the other compilers. This is due to the fact that it only SIMD
+vectorises the loop performing the Continuity calculation and, unlike
+Cray and Intel, it does not have an intrinsic to make vectorisation of
+a loop mandatory. (When using OpenMP 4.0 the SIMD directive could be
+used but even then that is only taken as a hint.)
 
@@
 of the code performs better. The results for the Cray compiler are
-however somewhat sqewed by the fact that it did not SIMD vectorise key
+however somewhat skewed by the fact that it did not SIMD vectorise key
 parts of the original version (see below).

### 2017-09-18 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: respond to comments on serial and OMP performance

@@
 added text to this effect.
+
+Page 15, line 5 - we were referring to the L3 cache. This has been clarified.
+
+Figure 5 - it is a reflection of the fact that Gnu is free while the Intel and
+Cray compilers have performance of the generated code as a key part of their
+respective business models.
+
+Page 18, line 9 - the NEMOLite2D performance for the 32x32 domain is
+significantly greater than for the other domains because it largely
+fits within L2 cache. We have calculated that the working-set size for
+the momentum kernel is 136*n^2. For 32x32 this gives ~136KB while for
+64x64 this gives ~544KB. The former will fit within L2 cache while the
+latter will spill to L3 cache. In addition, recall that we only parallelise
+the outermost loop (Sec. 2.2.1) so, for the 32x32 domain, there are only
+32 items to be shared amongst the threads. We have added a note to the
+text regarding this.
+
+Figure 11 -

### 2017-09-18 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: respond to comments on serial and OMP performance

@@
 step is to examine the necessary code transformations in detail.  We
-do this for the $256^{2}$ case since this fits within cache on the Ivy
+do this for the $256^{2}$ case since this fits within (L3) cache on the Ivy
 Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
@@
 limited quantity of parallelism in the $32 \times 32$ domain inhibits
-scaling. Therefore, to fully test the performance of our OpenMP
-implementations we have chosen to examine results for the $256 \times
-256$ domain and these are shown in Figures~\ref{FIG_gnu_omp_256},
-\ref{FIG_intel_omp_256} and \ref{FIG_cray_omp_256}.
+scaling. (Recall that we are only parallelising the outer loop -
+Section~\ref{sec_sep_dos}.) Therefore, to fully test the performance
+of our OpenMP implementations we have chosen to examine results for
+the $256 \times 256$ domain and these are shown in
+Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256} and
+\ref{FIG_cray_omp_256}.
 

### 2017-09-18 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: address comment on small effect of COLLAPSE in OpenACC

@@
 
-Figure 11 -
+Figure 11 - the size of the Momentum kernels meant that
+their resource use limited the occupancy that could be achieved.
+Therefore COLLAPSE might only help the performance of the other kernels and
+they only account for some 30% of the run-time. In addition, the kernels
+had been parallelised by enclosing them within a !$acc kernels region
+which leaves the details to the compiler. It would seem that in this
+case the decisions made by the compiler were good and COLLAPSE provided
+little or no benefit. The text has been altered to cover both of these
+points.

### 2017-09-18 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: address comment on small effect of COLLAPSE in OpenACC

@@
 nests has a small but beneficial effect for the majority of problem
-sizes. Gains here are limited by the fact that this optimisation was
-only beneficial for the non-Momentum kernels and these only account
-for some 30\% of the run-time.
+sizes.  The gains here are small for two reasons; first, the automatic
+CUDA kernel generation performed by the compiler (as instructed by the
+\texttt{kernels} directive) is working well for this code and second,
+this optimisation was only beneficial for the non-Momentum kernels and
+these account for just 30\% of the run-time.
 

### 2017-09-18 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: improve figure 13 formatting
nemolite2d-paper: add text about GPU performance quantification

@@
 points.
+
+Page 22, line 11 - we agree with the reviewer's point that there was
+no proof that the CUDA version was performing well. However, it is
+also not a matter of comparing memory bandwidths on the CPU and GPU -
+our work on the roofline model has demonstrated that the Momentum
+kernel is not memory-bandwidth bound. Instead, we have added a
+paragraph that demonstrates we have used profiling to investigate the
+performance limitations on the GPU and are confident that we can do no
+better without changing kernel code.
+
+Figure 13 - font size and line widths have been increased. Labels have been
+re-arranged to improve clarity.

### 2017-09-18 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: add text about GPU performance quantification

@@
 the performance of the two versions is less than 5\% and for the
-larger domains it is at most 12\%. This demonstrates that the
-performance cost of the {PS}y{KA}l approach in utilising a GPU for
-NEMOLite2D is minimal.  The discrepancy in performance between the
-CUDA and OpenACC versions has been fed back to the PGI compiler team.
+larger domains it is at most 12\%.
+
+In order to check the efficiency of the CUDA implementation we
+profiled the code on the GPU.  For large problem sizes this
+showed that the non-Momentum kernels are memory-bandwidth bound and
+account for about 40\% of the runtime. However, the Momentum kernels
+were latency limited, getting $\sim$103 GB/s of memory bandwidth. Although
+fusing these kernels reduced the required memory bandwidth (and
+produced a performance improvement), doing so resulted in a fairly
+large kernel requiring a large number of registers. This in turn
+reduces the occupancy of the device which exposes memory latencies.
+
+All of this demonstrates that the performance cost of the {PS}y{KA}l
+approach in utilising a GPU for NEMOLite2D is minimal.  The
+discrepancy in performance between the CUDA and OpenACC versions has
+been fed back to the PGI compiler team.
 

### 2017-09-21 docs/papers/roofline_habakkuk/habakkuk.tex
habakkuk-paper: add/move words to Introduction

@@
 
+The Roofline Model (RM)~\cite{roofline} has become a very popular way
+of examining the performance of a given computational kernel and
+contrasting it with the various limits associated with the system upon
+which it is executing. However, its simplicity naturally means that
+it has limitations.
+
+In this work we examine those limitations in light of two different
+shallow-water, finite-difference models. We explore the reasons behind
+the limited utility of the RM for these two models and present a
+tool, Habakkuk, that uses static code analysis to provide more
+realistic performance predictions for Fortran kernels.
+
 \begin{itemize}
@@
 
+The construction of a Roofline Model requires the peak floating-point
+operations per second (FLOPS) and memory bandwidth (GB/s) that the
+computer hardware is capable of. Although theoretical values for these
+quantities may be computed (using the technical specification of
+the machine, e.g. the number of floating-point instructions that may
+be issued per clock cycle).
+
+information on both the
+computer being used and the code being run. For the former,
+
 Arithmetic intensity and MAX(memory bandwidth, FLOPS)
+
+A key component of the Roofline Model is the Arithmetic Intensity (AI)
+of the code being executed:
+\begin{equation}
+AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
+\end{equation}
 
@@
 \end{equation}
+
 Since HWPC data proved to be unreliable, we calculated this quantity

### 2017-10-11 docs/papers/nemolite2d_performance/referee_response.txt
nemolite2d-paper: update referee response to match copernicus requirements

@@
-The authors thank the reviewers for their detailed feedback on our paper.
-We address the points raised below, beginning with the first (anonymous)
-referee and then moving on to those from Carlos Osuna.
+A. Kerkweg:
+==========
+
+We have added the name of the model used/developed in this work to the title.
+
+This model has been published with a DOI and a reference to this has been
+added to the Code Availability section of the paper.
 
 Anonymous referee:
+=================
+
+"1. The optimization and performance sections are very well done,
+however, I would like to see one additional data point in the parallel
+performance sec- tion: all of the optimizations and directives applied
+to the original NEMOLite2D code.  This will allow the reader to see if
+much or any performance is being left on the table by using the PSyKAl
+approach"
+
+Our response:
 
@@
 
+--
+
+"2.  The title of the paper suggests the PSyKAl approach is limited to
+finite-difference models, yet there is barely any mention of this
+limitation in the paper.  Can you describe what causes this limitation
+or if there will be attempts to extend the idea to more complex
+finite-element or finite-volume models?"
+
+Our response:
+
 2/ The PSyKAl approach is in fact applicable to both Finite Element and
@@
    this clear.
+
+--
+
+"3.  I fully agree that a separation of concerns is necessary to
+achieve portable performance, however it is not clear to me how
+significantly different PSyKAl is from similar ideas, such as OCCA
+(http://libocca.org) where kernels are written in a simple C or
+Fortran based kernel language and translated into OpenMP, CUDA or
+OpenCL and invoked similarly to a CUDA kernel.  Also, time-stepping
+research has long assumed that the spatial domain is computed using a
+single function call which iterates over all of points and executes
+the appropriate kernels.  Could the authors be more specific about the
+contributions which the PSyKAl approach makes, citing specific
+differences from ex- isting programming models, like CUDA and OpenMP?"
+
+Our response:
 
@@
 
+--
+
+"4.  The paper briefly mentions that generation of the PSy layer will
+be automated in the future.  Perhaps the authors could emphasize this
+more (as part of any changes made #3), as this paper describes a
+number of lessons learned which will be applied to work on automation."
+
+Our response:
+
 4/ We also use the new "Related Approaches" section to emphasise that
@@
 Carlos Osuna's general review:
+=============================
+
+"While the mathematics and equations are well described, the reviewer
+is missing some snippet of code that is representative of the target
+model.  Showing some simplified codes, like the dominant momentum
+section, could help the reader to understand the computational
+patterns (see some floating point operations performed, extensively
+dis- cussed in section 3.3), and, in general, better reason about
+sections that discuss the relevant characteristics of the code, like
+AI, or ILP."
+
+Our response:
 
@@
 
+--
+
+"A small "Related Work" section where other ap- proaches are explained
+and compared would help and can be used to highlight the novelties of
+this work. Similarly a section or paragraph describing the limitations
+would be valuable. Particularly a discussion on the design choices of
+PSyKAI in three levels, where all the computing architecture dependent
+optimizations happen in the PSy layer.  Is it the right choice, could
+the author imagine other optimizations that would need to be applied
+to the Kernel layer?"
+
+Our response:
+
 The suggestion of a "Related Work" section is a good one and we have added
 this just after the Introduction.
+
+--
+
+"What is there in the implementation, in terms of floating point
+operations that makes this code perform badly?  Also in the comparison
+of SSE and AVX, it is said that SSE and AVX versions of the division
+vector instruction does not provide any benefit. That would explain
+why they do not perform well, but not why SSE performs better than
+AVX."
+
+Our response:
+
+We have added text in section 3.3. to explain that the poor performance
+is due to liberal use of the division operation.
 
@@
 Carlos Osuna's specific comments:
+================================
+
+"pag.  7, line 10: "In order that we could safely add such directives
+we altered the GOcean infrastructure to allocate all field-data arrays
+with extents greater than strictly required": It is not clear to the
+reader why these larger extents are required in order to add this
+directive."
+
+Our response:
 
@@
 
+--
+
+"pag. 7, line 22: "Although the Intel compiler does do" I would
+replace by "Although the Intel compiler can do" since it is not
+guarantee when the compiler will inline."
+
+Our response:
+
 Changed "Intel does..." to "Intel can..." as requested.
+
+--
+
+"section 2.1.5:  the discussion makes the reader wonder whether attaching the
+grid to the field, which can be read only or not is the right choice."
+
+Our response:
 
@@
 was chosen in order to support applications which use more than one grid.
+
+--
+
+"section 2.1.8: it is not clear why inlining these copies gives better
+performance. Is it due to data reuse of the same fields that might be
+in cache? If so, that will depend on the domain size? Maybe a sentence
+clarifying that would help."
+
+Our response:
 
@@
 
+--
+
+"section 2.2.1: Aside note: It looks like other omp strategies like
+blocking would help increasing the data locality of computations among
+cores computing different blocks, or parallel do simd that would allow
+collapsing the loops and increase parallelism?"
+
+Our response:
+
 Sec 2.2.1 - yes, other strategies might be beneficial, especially blocking.
@@
 possibilities.
+
+--
+
+"Figure 3: why is gnu still much slower?  We would expect that PSyKAI
+would help the compiler adding the corresponding directives so that
+all compilers could deliver similar performance. Still is a factor 2x
+away from others. Can the reason for this be clarified?"
+
+Our response:
 
@@
 
+--
+
+"page 15, line 5: Text says that 256ˆ2 fits in cache.  Which cache are
+the authors referring to? A single field with that domain size would
+take 512 KB (in double precision) which will be out of L1"
+
+Our response:
+
 Page 15, line 5 - we were referring to the L3 cache. This has been clarified.
+
+--
+
+"Figure 5: same comment I made before.  It seems that we can not
+attain same perf.  with all compilers. Would be interesting clarify
+what is the intrinsic reason for this, if this is a real limitation of
+the compiler, or future work can improve the gnu performance."
+
+Our response:
 
@@
 Cray compilers have performance of the generated code as a key part of their
-respective business models.
+respective business models. We have added text about the lack of a crucial
+directive to force the Gnu compiler to SIMD vectorise a given loop.
+
+--
+
+"page 18, line 9: text says that 32x32, due to limited amount of
+parallelism inhibits scaling, therefore focus is on 256x256 domain
+sizes.  Is it clear that is parallelism what limits scalability? The
+curve shown by 32x32 might have a reasonable shape for memory bound
+codes, where with few cores one can saturate the memory bandwidth of
+the Ivy Bridge, and adding more cores might not provide better
+performance. On the contrary the large gap between 32x32 and the other
+domains suggests that the code is not cache oblivious, where the
+updated points/s would not (so strongly) depend on the domain size?
+See for example: https://arxiv.org/pdf/1410.5010.pdf"
+
+Our response:
 
@@
 
+--
+
+"Figure 11: For domain size of 64, parallelism will be very small on
+the GPU, if col- lapse(2) is not used.  If used, 4096 GPU threads
+could be active, still not 100% occu- pancy but much better. So it is
+not clear why the collapse(2) didnt help in performance?"
+
+Our response:
+
 Figure 11 - the size of the Momentum kernels meant that
@@
 
+--
+
+"Page 22, line 11: The statement says that the comparison between
+PSyKAI approach and the CUDA fortran implementation proves little
+overhead introduced by PSyKAI.  While this is true, it is no proven
+that the implementation is efficient, since also a CUDA implementation
+could be inefficient.  Figure 12 suggests that actually the GPU imple-
+mentation provides an expected performance in comparison with memory
+bandwidths of the IB and K40. Adding the GPU numbers to the Roofline
+model of Figure 13 would help in proving efficency of the GPU code and
+make a more solid statements about the GPU performance."
+
+Our response:
+
 Page 22, line 11 - we agree with the reviewer's point that there was
@@
 
+--
+
+"Figure 13: font size and lines are small and hard to read. Increasing
+the font size and probably using markers would help"
+
+Our response:
+
 Figure 13 - font size and line widths have been increased. Labels have been

### 2017-10-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: update abstract to reflect more general title

@@
 
-We present an approach which we call {PS}y{KA}l that is designed to achieve
-portable performance for parallel, finite-difference Ocean models.  In
-{PS}y{KA}l the code related to the underlying science is formally
-separated from code related to parallelisation and single-core
-optimisations. This separation of concerns allows scientists to code
-their science independently of the underlying hardware architecture
-and for optimisation specialists to be able to tailor the code for a
+We present an approach which we call {PS}y{KA}l that is designed to
+achieve portable performance for parallel,
+finite-difference/volume/element earth-system models.  In {PS}y{KA}l
+the code related to the underlying science is formally separated from
+code related to parallelisation and single-core optimisations. This
+separation of concerns allows scientists to code their science
+independently of the underlying hardware architecture and for
+optimisation specialists to be able to tailor the code for a
 particular machine independently of the science code. We have taken
@@
 full ocean models (input/output, boundary conditions, \textit{etc.}).
-We have then manually constructed a {PS}y{KA}l version of this code and
-investigated the transformations that must be applied to the
+We have then manually constructed a {PS}y{KA}l version of this code
+and investigated the transformations that must be applied to the
 middle/PSy layer in order to achieve good performance, both serial and
@@
 or not the obtained performance is `good' we also consider the
-limitations of the basic roofline model and improve on it
-by generating kernel-specific CPU ceilings.
+limitations of the basic roofline model and improve on it by
+generating kernel-specific CPU ceilings.
 

### 2017-10-11 docs/papers/nemolite2d_performance/nemolite2d_perf_abstract.tex
nemolite2d-paper: update abstract to reflect more general title

@@
 
-We present an approach which we call {PS}y{KA}l that is designed to achieve
-portable performance for parallel, finite-difference Ocean models.  In
-{PS}y{KA}l the code related to the underlying science is formally
-separated from code related to parallelisation and single-core
-optimisations. This separation of concerns allows scientists to code
-their science independently of the underlying hardware architecture
-and for optimisation specialists to be able to tailor the code for a
+We present an approach which we call {PS}y{KA}l that is designed to
+achieve portable performance for parallel,
+finite-difference/ volume/element earth-system models.  In {PS}y{KA}l
+the code related to the underlying science is formally separated from
+code related to parallelisation and single-core optimisations. This
+separation of concerns allows scientists to code their science
+independently of the underlying hardware architecture and for
+optimisation specialists to be able to tailor the code for a
 particular machine independently of the science code. We have taken

### 2017-10-11 nemolite2d_psykal_ocl/nemolite2d.c
opencl - use CPP directives to get code to compile when using OCL 1.0 (Intel SDK)

@@ -243,9 +243,14 @@ int main(){
 				   &result_len);
     ret = clGetDeviceInfo(device_ids[idev], CL_DEVICE_VERSION,
 			  (size_t)128, &version_str, &result_len);
+#if __OPENCL_VERSION__ < 120
+    /* The Intel/Alter OpenCL SDK is only version 1.0 */
+    fp_config = 0;
+#else
     ret = clGetDeviceInfo(device_ids[idev], CL_DEVICE_DOUBLE_FP_CONFIG,
-			  (size_t)(sizeof(cl_device_fp_config)),
-			  &fp_config, &result_len);
+    			  (size_t)(sizeof(cl_device_fp_config)),
+    			  &fp_config, &result_len);
+#endif
     size_t wg_size;
     ret = clGetDeviceInfo(device_ids[idev], CL_DEVICE_MAX_WORK_GROUP_SIZE,
 			  (size_t)(sizeof(size_t)),
@@ -279,7 +284,10 @@ int main(){
   device = &(device_ids[idev]);
 
   /* Check what version of OpenCL is supported */
-  if(strstr(version_str, "OpenCL 1.2")){
+  if(strstr(version_str, "OpenCL 1.0")){
+    cl_version = 10;
+  }
+  else if(strstr(version_str, "OpenCL 1.2")){
     cl_version = 12;
   }
   else if(strstr(version_str, "OpenCL 2.0")){
@@ -300,21 +308,21 @@ int main(){
   check_status("clCreateContext", ret);
   
   /* Create Command Queue with properties set to NULL */
-  if(cl_version == 12){
-    /* NVIDIA only support OpenCL 1.2 so we get a seg. fault if we attempt
+#if __OPENCL_VERSION__ < 200
+  /* The Intel/Alter OpenCL SDK is only version 1.0 */
+  /* NVIDIA only support OpenCL 1.2 so we get a seg. fault if we attempt
      to call the ...WithProperties version of this routine */
-    command_queue = clCreateCommandQueue(context,
-					 *device,
-					 queue_properties, &ret);
-    check_status("clCreateCommandQueue", ret);
-  }
-  else if(cl_version == 20){
-    command_queue = clCreateCommandQueueWithProperties(context,
-						       *device,
-						       &queue_properties, &ret);
-    check_status("clCreateCommandQueueWithProperties", ret);
-  }
-
+  command_queue = clCreateCommandQueue(context,
+				       *device,
+				       queue_properties, &ret);
+  check_status("clCreateCommandQueue", ret);
[…]
